---
title: "ATTEND — 0. Methods"
author: "sceriff0"
date: "2026-08-11"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

> **Question.** Why is each analysis choice in this pipeline the choice it is?
> **Cohort.** None of its own; the fitted panels below use all patients on the master.
> **Reads.** attend_master_joined, plus data/nassar/ when a reference cohort is present.
> **Out of scope.** Results. Every figure that answers a biological question lives in a
> numbered report; this file carries only the justification behind how those figures are made.

```{r setup, include=FALSE}
# renv: tidyverse, here. Same dependency set as the numbered reports; this file fits the
# Nassar recalibration, so it loads the master like any other report.
library(tidyverse)
library(here)
# Headless-safe raster device. The cluster's compute nodes have no X11, so R's default
# png() fails with "unable to open connection to X11 display" and every figure dies. cairo
# needs no display; the guard keeps this working on a machine built without it.
if (capabilities("cairo")) options(bitmapType = "cairo")

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 5)

source(here("code", "attend_classes.R"))     # attend_tmb_*, ancestry_by_pid(), nassar helpers
source(here("code", "attend_plots.R"))       # attend_theme(), attend_box(), the shared palette
source(here("code", "attend_io.R"))          # read_intermediate()
source(here("code", "build_master.R"))   # get_master() — builds on demand
source(here("code", "load_clinical.R"))      # load_gianlu_clinical_data()
```

This file collects every **justification** in the pipeline: why a distance metric, why a
correction, why a threshold. It is read once, not once per report. Operational recipes —
how to run a tool on the cluster — live in runbooks/ instead, and the derivation of any
individual figure stays inline in the report that draws it.

The master is loaded here only because two of the justifications below are quantitative:
the ancestry correction is *fitted*, so its coefficients and effect are shown rather than
asserted.

```{r data}
master <- get_master() |> add_molecular_classes()
have_master <- !is.null(master)
```

## Copy-number clustering — why 1 − Pearson distance with Ward.D2 linkage

Kandoth et al. (*Nature* 2013) Suppl. Methods S2 names no metric or linkage for the
*copy-number* clustering, though it names 1 − Pearson / Ward for its mRNA and methylation
clusterings. This pipeline applies that same pair to the copy-number matrix, set via
cnv_method / cnv_dist in attend_classes.R and passed to cluster_arm_matrix(). It is the only distance/linkage combination the pipeline uses.

Euclidean with complete linkage is the obvious alternative. Comparing the two shows how
sensitive the copy-number groups are to a metric choice the source paper never fixed.

## How the serous-like cluster is chosen, and why it is directional

Clustering copy-number profiles produces k unlabelled clusters. TCGA never had to decide
which one was serous-like: with 373 tumours and histology in hand, cluster 4 was visibly
94% serous. ATTEND has no such label, so the call has to be **inferred from the copy-number
data itself** — and that inference step is **not part of TCGA's methodology**. Nothing
about the clustering below departs from TCGA; only this labelling step is ours.

The obvious rule — pick the cluster with the largest **mean |value|**, i.e. the most
altered one — turns out to answer the wrong question. "Most altered" is not "most
serous-like", and on *thresholded* features the two come apart: an undirected score
rewards a deletion-dominated cluster just as much as the amplification-driven profile
that actually characterises serous-like disease.

This report therefore scores clusters **directionally**: only *concordant* events count —
**gains at amplification peaks, losses at deletion peaks**. It is the same principle
attend_scna$panel already applies to the pre-specified locus panel in [07-mutation-and-cna](07-mutation-and-cna.html)
(*a deletion at MYC must not score as serous-like*).

⚠️ **Both calls are computed and printed, and a [DISAGREE] flag is raised when they
differ.** If they agree on ATTEND, the directional correction changed nothing here and
the result is a confirmation. Read that flag before drawing any conclusion from the
cluster labels.

## Where this reproduction follows TCGA, and where it departs

Each subsection names one step of [09-tcga-classification](09-tcga-classification.html)
and states what it takes from Kandoth et al. Suppl. Methods S2 and what is an ATTEND
addition. Read it once; the reports themselves carry only their own derivations.

### The clustering

*Follows:* Suppl. Methods S2 exactly — thresholded CN in GISTIC significant peaks,
hierarchical clustering. *Differs:* DRAGEN WES .seg rather than SNP6; 1 − Pearson /
Ward.D2 (S2 leaves the CN metric unspecified but uses this pair for mRNA/methylation); the
cascade reads a single cut (k_cnv in attend_classes.R), while the k-sweep heatmap
shows a range of cuts around it; and the integrated-class and aneuploidy-score annotation
bars are ATTEND additions, not Fig. 1a elements.

### The k-sweep

TCGA fixed a single k (4). This panel is an ATTEND diagnostic: it exposes the
whole cut spectrum on the Fig-1a canvas so the k = 4 parity cut used by the cascade can be
judged against its neighbours, under the Pearson/Ward metric.

### The aneuploidy-enrichment cross-check

This is an ATTEND **QC**, not a TCGA figure: TCGA *characterised* cluster 4
as the highest-SCNA/aneuploidy group but did not run this label-vs-independent-score check.
It exists to prove our burden-based CN-high label isn't circular with the aneuploidy score.

### The cascade

*Follows:* the per-sample priority stratification of their Fig. 2b —
substitution/POLE → MSI → copy-number cluster; CN-high = cluster membership. *Differs:*
**POLE omitted** (ATTEND has no POLE-ultramutated cases), so the cascade starts at MMRd;
and MMRd is called by **MMR IHC** alone (TCGA used MSI/MLH1).

### The TP53-enrichment QC

This *validates* the property Kandoth et al. reported (cluster 4 ≈ 90%
*TP53*-mutant) rather than assuming it. We do **not** gate the serous call on TP53 (that
would be the "and" rule); TP53 enrichment is the independent confirmation the cluster maps
to TCGA's serous-like group.

### The per-class panels

*Follows:* their finding that CN-high/serous-like carries the highest SCNA
burden and lowest mutation rate, MSI the highest TMB (Fig. 1–2). *Differs:* per-definition
TMB panels are an ATTEND addition (ancestry-aware recompute), absent from TCGA.

## Ancestry-corrected TMB — how it works

Tumour-only TMB is inflated by residual germline variants, and — because gnomAD's
reference is Euro-skewed — the inflation is **ancestry-dependent** (Nassar et al.,
*Cancer Cell* 2022). The ATTEND MAFs carry gnomAD v4.1 **per-sub-population** allele
frequencies, which is what makes a correction possible without matched normals.

ATTEND is sequenced tumour-only, so leftover germline variants inflate TMB, and because
germline allele frequencies differ between ancestries that inflation is ancestry-dependent
(Nassar et al., *Cancer Cell* 2022). The pipeline therefore carries two definitions and
never silently picks one:

| Reported name | Column | What it is |
|---|---|---|
| **TMB (normal)** | tmb__TMB_SCORE | the upstream tumour-only score, as delivered |
| **TMB (ancestry corrected)** | tmb__TMB_nassar | the same score after the ancestry recalibration |

Two further columns, tmb__TMB_robust and tmb__TMB_matched, exist on the master because
the recalibration is fitted on top of robust. They are intermediates: tmb_defs_present()
cannot return them, so no figure can show them. How the recalibration works is in
[00-methods](00-methods.html).

Every definition is re-binarised at the same `r attend_thresholds$tmb` mut/Mb cutoff by
tmb_class_of(), so panels that differ only in definition differ only in who crossed the
cut. The ancestry-corrected column stays all-NA until attend_ancestry$clinical_col is
set, and panels that would be empty are skipped rather than drawn blank.

Getting from the delivered score to the reported corrected one passes through two
intermediates, computed in [01-data-integration](01-data-integration.html) and living on
the master, but never plotted, faceted, or labelled anywhere:

- **robust** — drop variants common in **any** population (grpmax ≥
  `r attend_tmb_recompute$af_threshold`). This is Nassar's germline filter, and it needs
  no ancestry label.
- **matched** — drop variants common in the **patient's own** sub-population, from the
  self-reported clinical ancestry (attend_ancestry).

The reported ancestry-corrected column is then the paper's actual method: an
**ancestry-specific regression recalibration** TMB = m·TMB_robust + b (European vs
non-European), applied on top of robust. Its coefficients (attend_tmb_nassar$coef)
**default to identity**, so until values fit on TCGA WES are supplied this column equals
robust — the published Oncopanel numbers are panel-specific and do not transfer to
exome data.

tmb_defs_present() cannot return the intermediates, so no report can reintroduce them
into a figure by accident. An unfiltered recompute is also used internally to calibrate
the exome denominator; it is not emitted, because it would simply duplicate the baseline.
The whole recompute lives in R with no upstream re-run — see
attend_classes.R::tmb_variant_table() and apply_nassar_recalibration().

### How the Nassar et al. recalibration actually works

It helps to be precise about what Nassar et al. (*Cancer Cell* 2022) did, because our
nassar column reproduces it and the distinction from the other variants is subtle.

**The problem.** In tumor-only sequencing there is no matched normal, so inherited
(germline) variants are mixed in with somatic ones. TMB should count only somatic
mutations, so pipelines remove germline by filtering out variants that are **common in
the population** (looked up in gnomAD). But gnomAD is **ancestry-skewed** — its v2.1
exomes are ~57k European vs ~8k African. A variant that is common in African ancestry
but rare in the (Euro-heavy) reference is *not recognised as germline*, survives the
filter, and is **miscounted as a somatic mutation**. The net effect: tumor-only TMB is
**inflated, and the inflation is larger for non-European patients** — who were then
**over-classified as TMB-high**, a label that (in their data) did **not** predict
immunotherapy response outside European ancestry.

**Their method has two distinct steps.** Do not conflate them:

1. **Germline filter (variant level).** Exclude any variant whose **maximum population
   allele frequency across all gnomAD sub-populations** (maxPOPAF, i.e. grpmax)
   exceeds **0.1%**. This is a *fixed, ancestry-agnostic* filter — it drops a variant if
   it is common in *any* population. **This step is exactly our robust variant**
   (grpmax ≥ 0.001). It removes most, but not all, of the ancestry-driven inflation.

2. **Ancestry regression recalibration (sample level) — the headline contribution.**
   This is **not another filter**. Using cohorts that had *both* tumor-only and matched
   tumor-normal ("paired") TMB as ground truth, they fit a **linear regression of the
   true paired TMB on the tumor-only TMB, separately for European and non-European
   patients**:
   $$\text{TMB}_{\text{paired}} \;=\; m_{\text{group}} \cdot \text{TMB}_{\text{tumor-only}} \;+\; b_{\text{group}}$$
   The slope $m$ and intercept $b$ therefore encode, *per ancestry group*, how much the
   tumor-only estimate over- (or under-) shoots the truth. To recalibrate a new sample
   you apply its group's coefficients: $\text{TMB}_{\text{corrected}} = m_{\text{group}}
   \cdot \text{TMB}_{\text{tumor-only}} + b_{\text{group}}$. Because $(m, b)$ differ
   between European and non-European, the *same* tumor-only TMB is corrected by a *different*
   amount depending on ancestry — that is precisely how the method removes the residual
   ancestry bias that the fixed filter (step 1) leaves behind.

**Two supporting details.** (a) **Ancestry was inferred *genetically*** — principal-
component analysis of common germline SNPs called from the sequencing reads — not from
self-reported race. (b) The coefficients were **fit and validated against matched
tumor-normal** as the gold standard.

**How our nassar column maps to this — and where it differs.**

| Nassar step | Our implementation |
|---|---|
| germline filter maxPOPAF ≤ 0.1% | **robust** (grpmax ≥ 0.001) — identical in spirit |
| regression TMB = m·TMB + b, per ancestry, on top of the filtered TMB | **apply_nassar_recalibration()** on top of robust |
| ancestry via **genetic PCA** | **self-reported clinical** ancestry (attend_ancestry) — a coarser proxy; swap in EthSEQ/somalier genetic ancestry later |
| coefficients fit on **paired tumor-normal**, panel-specific (e.g. Oncopanel v3 European m=1.094, b=−1.94) | **default identity (m=1, b=0)** — those published numbers are **panel-specific and do NOT transfer to whole-exome**, so nassar == robust until you fit WES coefficients |
| gnomAD v2.1 + TOPMED | gnomAD **v4.1** (newer, larger, less Euro-biased) |

**To make nassar genuinely faithful for ATTEND (whole-exome):** fit the regression on
**TCGA WES** — compute paired-WES TMB and tumor-only-WES TMB (tumor-only simulated by
dropping the normal and applying the same robust filter), label European vs
non-European, then

```r
coefs <- fit_nassar_coefficients(tcga_ref)   # ref has paired_tmb, tumor_only_tmb, group
attend_tmb_nassar$coef <- coefs              # paste into code/attend_classes.R
```

Until then, nassar deliberately equals robust rather than showing a spurious
correction. (Sources: Nassar et al., *Cancer Cell* 2022,
[full text](https://pmc.ncbi.nlm.nih.gov/articles/PMC9559771/).)

```{r ancestry-tmb, fig.height=5}
tmb_recomp <- if (have_master) tmb_defs_present(master) else character(0)

if (have_master && length(tmb_recomp) >= 2) {
  # Per-patient TMB across recomputed definitions + reclassification vs the cutoff.
  long <- master |>
    select(pid, all_of(unname(tmb_recomp))) |>
    pivot_longer(all_of(unname(tmb_recomp)), names_to = "tmb_col", values_to = "tmb") |>
    mutate(def = factor(names(tmb_recomp)[match(tmb_col, tmb_recomp)], levels = names(tmb_recomp)),
           tmb = suppressWarnings(as.numeric(tmb))) |>
    filter(!is.na(tmb))

  print(
    ggplot(long, aes(def, tmb)) +
      # Deliberately NOT attend_box(): this is a paired design — each point must sit at the
      # same x as its patient's connecting line, so the points cannot be jittered. Every
      # definition is measured on the whole cohort, so the box is supported at any n.
      geom_boxplot(outlier.shape = NA, fill = attend_neutral, width = 0.6) +
      geom_line(aes(group = pid), alpha = 0.15) +
      geom_point(size = 0.7, alpha = 0.4) +
      geom_hline(yintercept = attend_thresholds$tmb, linetype = 2, colour = "grey40") +
      labs(title = "Per-patient TMB: baseline (TMB_SCORE) vs ancestry-aware definitions",
           subtitle = paste0("lines connect the same patient; dashed = TMB-high cutoff; n = ",
                             dplyr::n_distinct(long$pid), " patients"),
           x = "TMB definition", y = "TMB (mut/Mb)") +
      attend_theme()
  )

  # How many patients are TMB-high under each definition (the reclassification).
  reclass <- long |> group_by(def) |>
    summarise(n = sum(!is.na(tmb)),
              tmb_high = sum(tmb >= attend_thresholds$tmb, na.rm = TRUE),
              pct_high = round(100 * tmb_high / n, 1), .groups = "drop")
  knitr::kable(reclass, caption = paste0("TMB-high (>= ", attend_thresholds$tmb,
               " mut/Mb) count per definition — the ancestry-driven reclassification"))
} else {
  message("Ancestry-corrected TMB not on the master yet — rebuild report 1 with ",
          "the gnomAD-annotated MAFs synced into data/variant_annotations/.")
}
```

The diagnostic the whole correction exists for: **how much TMB the correction removes,
broken down by ancestry**. Each point is one patient's TMB (normal) − TMB (ancestry
corrected) in mut/Mb — the amount attributed to residual germline contamination.

This is the plot that shows whether the bias is real *in this cohort*. Nassar et al.
predict systematically **larger** removals in non-European groups, because gnomAD's
non-European allele-frequency reference is sparser and more germline variants survive an
undifferentiated filter. A flat distribution across sub-populations means either the
cohort is too homogeneous to show it, or the correction is not doing anything.

Sub-populations come from mapping the self-reported clinical ancestry column onto gnomAD
codes (attend_ancestry); patients whose label does not map are NA and dropped, so
group sizes here can be much smaller than the cohort.

```{r ancestry-tmb-by-group, fig.height=5}
anc <- if (have_master) ancestry_by_pid(master) else NULL
if (!is.null(anc) && all(c("tmb__TMB_SCORE","tmb__TMB_nassar") %in% names(master))) {
  g <- master |>
    transmute(pid,
              normal    = suppressWarnings(as.numeric(`tmb__TMB_SCORE`)),
              corrected = suppressWarnings(as.numeric(`tmb__TMB_nassar`)),
              removed   = normal - corrected) |>
    inner_join(anc, by = "pid") |>
    filter(!is.na(removed))
  if (nrow(g) > 0) {
    ggplot(g, aes(subpop, removed)) +
      # Sub-populations are the smallest groups in the pipeline — the note above says group
      # sizes here "can be much smaller than the cohort" — so the mark is chosen per group.
      attend_box(g, x = "subpop", y = "removed") +
      geom_hline(yintercept = 0, linetype = 2, colour = "grey50") +
      labs(title = "TMB removed by the ancestry correction, by ancestry",
           subtitle = "TMB (normal) - TMB (ancestry corrected); larger = more germline contamination removed",
           x = "gnomAD sub-population (from self-reported ancestry)",
           y = "TMB removed (mut/Mb)") +
      attend_theme()
  } else message("No patients with both TMB definitions and a mapped ancestry.")
} else {
  message("Need the clinical ancestry column (attend_ancestry$clinical_col = '",
          attend_ancestry$clinical_col, "') and tmb__TMB_nassar on the master ",
          "for the by-ancestry view.")
}
```

## Nassar recalibration coefficients

tmb__TMB_nassar = m_group · tmb__TMB_robust + b_group, with (m, b) **per ancestry
group** (European vs non-European). Which coefficients are applied is set by
attend_tmb_nassar$active_source:

- **"identity" (default)** — m=1, b=0, so nassar equals robust.
- **"oncopanel_v3" / "_v2" / "_v1"** — the **published Nassar et al. 2022**
  coefficients (STAR Methods, "TMB Recalibration"). ⚠️ These were fit on **OncoPanel**
  (targeted-panel) tumor-only TMB with TCGA WES as ground truth, **not** on WES input —
  so applied to ATTEND's exome robust TMB they are an **off-scale approximation** (the
  slope maps OncoPanel→WES-truth, not WES→WES-truth). Available on request; use knowingly.
- **"custom"** — your own values in attend_tmb_nassar$coef.

The published values (per OncoPanel version): European v3 m=1.094, b=−1.94
(v2 0.994, −2.12; v1 0.989, −2.18); Non-European v3 m=0.895, b=−1.29
(v2 0.840, −1.71; v1 0.821, −1.71). To use them, set
attend_tmb_nassar$active_source <- "oncopanel_v3" and re-knit.

**To fit your own (the WES-correct path):** build a reference cohort with matched
**paired (tumor-normal) TMB**, **tumor-only TMB** (simulated with the same robust/grpmax
0.1% filter), and an ancestry group column, drop it in data/nassar/ (any
*reference*.tsv/.csv), re-knit to fit, paste the printed coefficients into
attend_tmb_nassar$coef, and set active_source = "custom". The panels below render the
recalibration effect whenever the active coefficients are non-identity.

```{r nassar-fit}
nc      <- nassar_coef_table()               # active coefficients + identity flag
ref_cfg <- attend_tmb_nassar$reference
ref_dir <- here("data", ref_cfg$dir)
ref_f   <- if (dir.exists(ref_dir))
  list.files(ref_dir, pattern = utils::glob2rx(ref_cfg$glob), full.names = TRUE, recursive = TRUE) else character(0)

if (length(ref_f) > 0) {
  ref <- tryCatch(
    if (grepl("\\.csv$", ref_f[1])) readr::read_csv(ref_f[1], show_col_types = FALSE)
    else                            readr::read_tsv(ref_f[1], show_col_types = FALSE),
    error = function(e) { message("nassar reference unreadable: ", conditionMessage(e)); NULL })
  if (!is.null(ref)) {
    fitted <- fit_nassar_coefficients(ref, paired = ref_cfg$paired,
                                      tumor_only = ref_cfg$tumor_only, group = ref_cfg$group)
    cat("Fitted from", basename(ref_f[1]), "— paste into attend_tmb_nassar$coef:\n")
    cat(sprintf("  european     = list(m = %.4f, b = %.4f)\n", fitted$european$m, fitted$european$b))
    cat(sprintf("  non_european = list(m = %.4f, b = %.4f)\n", fitted$non_european$m, fitted$non_european$b))
  }
} else {
  message("No reference cohort in ", ref_dir, " — reporting the active coefficients only.")
}

if (grepl("^oncopanel", nc$source)) {
  cat("WARNING: active_source = '", nc$source, "' — OncoPanel-derived published coefficients\n",
      "applied to WES robust TMB. Off-scale approximation (slope maps OncoPanel->WES-truth,\n",
      "not WES->WES-truth); prefer fitting on WES via the reference path.\n", sep = "")
}
knitr::kable(nc$table, digits = 4,
             caption = paste0("Active Nassar coefficients — source: ", nc$source,
                              " (identity = ", nc$identity, "; identity means nassar == robust)"))
```

What the correction actually does to each patient, on the two reported definitions.
**TMB (normal)** on the x-axis, **TMB (ancestry corrected)** on the y-axis, one point per
patient, coloured by ancestry group.

How to read it: the **dashed diagonal** is no change — a point on it was unaffected. Points
**below** it had TMB removed. The **dotted lines** are the `r attend_thresholds$tmb` mut/Mb
TMB-high cutoff on each axis, so the off-diagonal quadrants they carve out are the patients
whose **TMB-high/low call changed**. Those are the only patients for whom the choice of
definition changes any downstream figure, and the table counts them per ancestry group.

If the recalibration coefficients are still identity, this plot still shows the effect of
the germline filter itself — the correction chain from the delivered score to the reported
corrected one — which is the comparison every other report is drawing.

```{r nassar-effect, fig.height=5}
need <- c("tmb__TMB_SCORE", "tmb__TMB_nassar")
anc  <- if (have_master) ancestry_by_pid(master) else NULL

if (have_master && all(need %in% names(master)) && !is.null(anc)) {
  grp <- anc |>
    mutate(group = ifelse(subpop %in% attend_tmb_nassar$european_subpops,
                          "european", "non_european")) |>
    distinct(pid, group)
  eff <- master |>
    transmute(pid,
              normal    = suppressWarnings(as.numeric(`tmb__TMB_SCORE`)),
              corrected = suppressWarnings(as.numeric(`tmb__TMB_nassar`))) |>
    inner_join(grp, by = "pid") |>
    filter(!is.na(normal), !is.na(corrected))

  if (nrow(eff) > 0) {
    print(
      ggplot(eff, aes(normal, corrected, colour = group)) +
        geom_abline(slope = 1, intercept = 0, linetype = 2, colour = "grey50") +
        geom_point(alpha = 0.6) +
        geom_hline(yintercept = attend_thresholds$tmb, linetype = 3, colour = "grey40") +
        geom_vline(xintercept = attend_thresholds$tmb, linetype = 3, colour = "grey40") +
        labs(title = "Effect of the ancestry correction, by ancestry group",
             subtitle = paste0("dashed = no change; dotted = the ", attend_thresholds$tmb,
                               " mut/Mb TMB-high cutoff on each axis",
                               if (nc$identity) " | recalibration coefficients are identity, so this is the germline filter alone" else ""),
             x = "TMB (normal), mut/Mb", y = "TMB (ancestry corrected), mut/Mb",
             colour = NULL) +
        attend_theme()
    )
    reclass_n <- eff |> group_by(group) |>
      summarise(n = n(),
                high_normal    = sum(normal    >= attend_thresholds$tmb),
                high_corrected = sum(corrected >= attend_thresholds$tmb),
                changed_call   = sum((normal >= attend_thresholds$tmb) !=
                                     (corrected >= attend_thresholds$tmb)),
                .groups = "drop")
    knitr::kable(reclass_n, caption = paste0(
      "TMB-high (>= ", attend_thresholds$tmb, " mut/Mb) reclassification between the two ",
      "definitions, by ancestry group. changed_call = patients whose high/low call differs."))
  } else message("No patients with both TMB definitions and a mapped ancestry group.")
} else {
  message("Need tmb__TMB_SCORE, tmb__TMB_nassar and an ancestry column on the master ",
          "to show the correction effect.")
}
```


## Where the configuration lives

Every threshold and column name the reports read is declared in code/attend_classes.R,
never inline in a report. The entries that govern the copy-number and TMB methods above:

| Object | Governs |
|---|---|
| attend_cnv | CNV sources, the arm table, and the pileup amplitude split |
| attend_tcga_levels | integrated-class labels and their display order |
| attend_tmb_recompute | germline-filter threshold, gnomAD columns, counted variant classes |
| attend_tmb_nassar | recalibration coefficients, the active source, the reference-cohort glob |
| attend_ancestry | self-reported ancestry to gnomAD sub-population map; set clinical_col |
| attend_pole | POLE hotspots and TMB cutoff, kept unwired (ATTEND has no POLE cases) |

The arm-boundary table is data/arm_boundaries_hg38.tsv, derived from UCSC hg38 cytoBand
centromeres; set attend_cnv$build and attend_cnv$arm_table to work in hg19. The clustering parameters
cnv_method, cnv_dist, k_cnv and k_sweep are declared there too rather than at the
top of report 08 — a parameter this file *discusses* and another file *applies* has to be
visible to both.

## The aneuploidy classes
Aneuploidy score is the ASCETS arm-level SCNA burden read from data/ascets/: the fraction
of evaluable chromosome arms called altered, with arms below ASCETS' breadth-of-coverage
threshold (LOWCOV) excluded from the denominator rather than counted as unaltered.

Patients are split at attend_thresholds$aneuploidy = `r attend_thresholds$aneuploidy`
into **aneuploidy-high** and **aneuploidy-low** by aneuploidy_class_of(). The cut is a
configured constant, not a cohort quantile, so it does not move when the cohort grows.
Patients without an ASCETS score are NA and are dropped from class comparisons.

## Highlighted patients
Two overlay groups are drawn on top of the base colouring by highlight_points(), which
partitions the data so no patient is plotted twice:

- **polipo** — sample 21S188, in blue.
- **cohort** — the multi-modal subset, in yellow: patients who also have IHC or imaging
  data, and so appear in the concordance and composition reports.

polipo is declared first in attend_highlight, and matching is first-match-wins, so
21S188 is never absorbed into cohort.

## What "MMR-deficient only" means
The cohort restriction is is_mmrd(), which reads the **IHC** MMR call: loss of one or
more mismatch-repair proteins on staining. That is the criterion the TCGA cascade puts at
the top, and the one used throughout this pipeline.

MSI-high by NGS is a **correlate, not a criterion**. It agrees with the IHC call in most
patients and is reported alongside it, but it never defines the cohort — restricting on it
instead would silently swap the population between reports. Patients with no MMR IHC call
are excluded rather than assumed proficient.

## Where the data comes from
Everything below is read from attend_master_joined, the one-row-per-patient table written
by [01-data-integration](01-data-integration.html) and loaded here through
read_intermediate() — never by a hard-coded .csv or .parquet path.

That report translates each modality's native identifier onto pid, collapses to one row
per patient (numeric columns averaged, others first-non-NA), and prefixes columns with
the source dataset: gianlu__ clinical, aneu__ ASCETS, tmb__ TMB, hrd__, msi__,
ihc__. Mutation status is the exception — the MAF is never mean-collapsed, so
TP53_pathogenic and panel_pathogenic are derived per patient by
pathogenic_by_patient() and joined on.

Column names and thresholds are read from attend_cols and attend_thresholds
(code/attend_classes.R); chunks guarded by have_cols() skip themselves when a name is
still a placeholder.

```{r session-info}
sessionInfo()
```
