---
title: "ATTEND — 10. Data integration: the join, and what it cost"
author: "sceriff0"
date: "2026-08-13"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
library(here)
library(fs)
library(UpSetR)
# Headless-safe raster device. The cluster's compute nodes have no X11, so R's default
# png() fails with "unable to open connection to X11 display" and every figure dies. cairo
# needs no display; the guard keeps this working on a machine built without it.
if (capabilities("cairo")) options(bitmapType = "cairo")

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5)

source(here("code", "build_master.R"))   # load_raw_inputs(), build_crosswalks(), build_master()
source(here("code", "attend_plots.R"))   # highlight_points() overlay

# The join lives in code/build_master.R, so this report narrates and audits it rather than
# being the only thing that can produce it. That is what lets any report be built first.
raw <- load_raw_inputs()
cwx <- build_crosswalks(raw)

clinical_data        <- raw$clinical_data
imaging_data         <- raw$imaging_data
ihc_data             <- raw$ihc_data
gianlu_clinical_data <- raw$gianlu_clinical_data
aneu_data            <- raw$aneu_data
hrd_data             <- raw$hrd_data
msi_data             <- raw$msi_data
tmb_data             <- raw$tmb_data
maf_data             <- raw$maf_data
conversion_data      <- raw$conversion_data

id_cfg         <- cwx$id_cfg
cw_barcode_pid <- cwx$cw_barcode_pid
cw_image_pid   <- cwx$cw_image_pid
pid_vector     <- cwx$pid_vector
tables         <- cwx$tables
sets           <- cwx$sets
```

> **Question.** Which measurement belongs to which patient, what did the join cost, and are the joined values usable?
> **Cohort.** Every patient reached by at least one modality; no restriction.
> **Reads.** The code/load_*.R loaders, straight from HPC storage, plus data/attend_barcodes.csv.
> **Out of scope.** Any biological comparison. The TMB definitions, the aneuploidy cut and the highlight groups are defined once in [00-methods](00-methods.html).

ATTEND measures the same patients several different ways, each recorded with a different
kind of identifier. This report puts them onto one patient key, writes the result, and then
reports what that cost: which records never reached a patient, which patients are missing
which modality, and whether the joined values are sane.

Genomics tables are keyed by sequencing **barcode**, imaging and IHC by **image** id,
clinical tables by **patient** id (pid). Two crosswalks translate everything onto pid;
that logic is in code/attend_harmonise.R and the join consuming it is in
code/build_master.R — one definition each, so no two reports can drift.

Optional and guarded, so the report knits without them: pointblank, reclin2,
naniar/skimr, arrow.

## Part 1 — building the master

The whole join in one call: union every patient any dataset reached, left_join each
collapsed dataset onto it, derive mutation status per patient as booleans (never averaged),
recompute the ancestry-aware TMB definitions, and write the result plus every raw table to
output/clean_data/.

```{r master-build}
master <- build_master(raw = raw, cw = cwx, write = TRUE, verbose = TRUE)
```

ATTEND measures the same patients several different ways — clinical records, radiology
imaging, IHC, and WES-derived genomics (TMB, MSI, HRD, aneuploidy, and a per-variant MAF).
Each modality was recorded with a different kind of identifier, so this report puts them
onto one patient key and writes the result.

Genomics tables are keyed by sequencing **barcode**, imaging and IHC by **image** id,
clinical tables by **patient** id (pid). Two crosswalks, barcode to pid and image to pid,
translate everything onto pid. That logic lives in code/attend_harmonise.R — one
definition, so no two reports can drift apart.

Four packages are optional and every use is guarded, so the report knits without them:
pointblank for the post-join gate, reclin2 for the fuzzy near-miss check, and arrow
for Parquet intermediates. Install with
renv::install(c("pointblank","reclin2","arrow")).

### The raw inputs

Every table is read straight from its loader in code/load_*.R, which are the seam
between HPC storage (/hpcnfs/scratch/P_DIMA_ATTEND) and this pipeline. Nothing is
filtered or transformed at this point — these are the raw shapes as delivered.


Row and column counts of each raw table as loaded above, before any ID translation.
The key column differs per table — clinical/MSI/TMB/MAF use ID, imaging and IHC use
image_id, HRD uses Sample, aneuploidy uses sequencing_id, the conversion table
uses barcode — and reconciling those is the whole job of this report.

```{r load-dims}
raw_inputs <- list(
  clinical = clinical_data, imaging = imaging_data, ihc = ihc_data,
  gianlu = gianlu_clinical_data, aneu = aneu_data, hrd = hrd_data,
  msi = msi_data, tmb = tmb_data, maf = maf_data, conversion = conversion_data
)

tibble(
  dataset = names(raw_inputs),
  rows    = map_int(raw_inputs, nrow),
  cols    = map_int(raw_inputs, ncol)
)
```

### Building the two crosswalks

The **barcode→pid** crosswalk is derived from gianlu_clinical_data, which is the only
table carrying both a sequencing barcode (TUMOR_BARCODE) and a patient id. The
**image→pid** crosswalk comes from imaging_data, which carries both image_id and a
patient id. Everything else is translated through one of these two.


Size of each crosswalk. These are the limiting reagent for the whole join: a genomic
sample can only reach a patient if its barcode appears in the barcode crosswalk, and an
IHC slide only if its image appears in the image crosswalk. Most of the unmapped
records counted in Part 2 are unmapped for this reason.

```{r crosswalk-coverage}
cat("barcode crosswalk:", nrow(cw_barcode_pid), "pairs,",
    n_distinct(cw_barcode_pid$pid), "patients\n")
cat("image   crosswalk:", nrow(cw_image_pid), "pairs,",
    n_distinct(cw_image_pid$pid), "patients\n")
```

#### Is the derived crosswalk right?

The barcode→pid crosswalk above is *reverse-engineered* from the clinical table, but
data/attend_barcodes.csv is the **canonical** map. The two are compared here on two
questions: which barcodes each one knows about, and — more seriously — whether they ever
**disagree about which patient a barcode belongs to**. A disagreement is not a coverage
gap; it would attach one patient's genomics to another patient's clinical record, and
every downstream result would be quietly wrong. It must be zero.

```{r crosswalk-validate}
derived_barcodes <- unique(cw_barcode_pid$barcode)
canon_barcodes   <- unique(as.character(conversion_data$barcode))

cat("barcodes in derived crosswalk:", length(derived_barcodes), "\n")
cat("barcodes in canonical file   :", length(canon_barcodes), "\n")
cat("in derived but NOT canonical :", length(setdiff(derived_barcodes, canon_barcodes)), "\n")
cat("in canonical but NOT derived :", length(setdiff(canon_barcodes, derived_barcodes)), "\n")

# The canonical file's patient-id column is auto-detected: the real file carries
# barcode / project / enrolment_id / tss / internal_id / patient_status, and
# enrolment_id is the study subject id. Override this if the detection is wrong.
canon_pid_candidates <- c("enrolment_id", "enrollment_id", "internal_id",
                          "pid", "PID", "PATIENT_ID", "patient_id", "Patient_ID",
                          "patient", "Patient", "ATTEND_ID", "attend_id", "ID")
canon_pid_col <- setdiff(intersect(canon_pid_candidates, names(conversion_data)),
                         "barcode")[1]
cat("using canonical patient-id column:",
    if (is.na(canon_pid_col)) "(none found)" else canon_pid_col, "\n")

if (!is.na(canon_pid_col)) {
  canon_cw <- conversion_data |>
    transmute(barcode   = as.character(barcode),
              pid_canon = as.character(.data[[canon_pid_col]])) |>
    filter(!is.na(barcode), !is.na(pid_canon)) |> distinct()

  disagree <- cw_barcode_pid |>
    inner_join(canon_cw, by = "barcode") |>
    filter(pid != pid_canon)

  cat("barcodes where derived pid DISAGREES with canonical:", nrow(disagree),
      "  <- must be 0\n")
  disagree
} else {
  message("No patient-id column detected in attend_barcodes.csv — ",
          "set canon_pid_col manually to enable the agreement check.")
}
```

The comparison above is character-for-character, so a canonical barcode differing only
by case, whitespace, or a -DNA suffix is reported as missing when it is really the
same sample. This fuzzy-matches the unmatched canonical barcodes against the derived set
(Jaro–Winkler ≥ 0.9 but < 1, i.e. close but not identical) and lists the near-misses, so
formatting drift can be told apart from genuine absence. Size-capped and guarded.

```{r crosswalk-reclin2}
only_canon <- setdiff(canon_barcodes, derived_barcodes)

if (requireNamespace("reclin2", quietly = TRUE) &&
    length(only_canon) > 0 && length(only_canon) <= 2000 && length(derived_barcodes) > 0) {
  near <- tryCatch({
    a <- tibble(barcode = only_canon)
    b <- tibble(barcode = derived_barcodes)
    pairs <- reclin2::pair(a, b)
    reclin2::compare_pairs(pairs, on = "barcode",
                           default_comparator = reclin2::cmp_jarowinkler(0.9),
                           inplace = TRUE)
    pp <- as.data.frame(pairs)
    pp <- pp[!is.na(pp$barcode) & pp$barcode >= 0.9 & pp$barcode < 1, , drop = FALSE]
    if (nrow(pp))
      tibble(canonical_barcode = a$barcode[pp$.x],
             nearest_derived   = b$barcode[pp$.y],
             jaro_winkler      = round(pp$barcode, 3)) |>
        arrange(desc(jaro_winkler))
    else tibble()
  }, error = function(e) { message("reclin2 near-miss check skipped: ", conditionMessage(e)); tibble() })

  if (nrow(near)) {
    cat("Canonical barcodes with a fuzzy near-miss (JW >= 0.9, not exact):\n")
    print(near, n = 50)
  } else cat("No fuzzy near-misses — the exact loss above is real, not formatting drift.\n")
} else {
  message("reclin2 probabilistic check skipped (package missing, nothing unmatched, ",
          "or unmatched set too large).")
}
```

#### Two duplication failure modes

These are opposite problems and only one of them is allowed.
**One ID → many patients** must be zero: a barcode pointing at two patients duplicates
rows on every join and corrupts counts.
**One patient → many IDs** is legitimate (a patient can have several samples or slides),
but it has to be quantified, because the collapse step below **averages those rows
together** and that hides within-patient variation.

```{r duplicates}
dup_barcode <- cw_barcode_pid |> count(barcode, name = "n_pid") |> filter(n_pid > 1)
dup_image   <- cw_image_pid   |> count(image,   name = "n_pid") |> filter(n_pid > 1)
cat("barcodes -> >1 patient:", nrow(dup_barcode), " <- must be 0\n")
cat("images   -> >1 patient:", nrow(dup_image),   " <- must be 0\n")

cw_barcode_pid |> semi_join(dup_barcode, by = "barcode") |> arrange(barcode)
cw_image_pid   |> semi_join(dup_image,   by = "image")   |> arrange(image)

multi_barcode <- cw_barcode_pid |> count(pid, name = "n_barcode") |> filter(n_barcode > 1)
multi_image   <- cw_image_pid   |> count(pid, name = "n_image")   |> filter(n_image > 1)
cat("\npatients with >1 barcode:", nrow(multi_barcode),
    "| patients with >1 image:", nrow(multi_image),
    " <- legitimate, but averaged by the collapse below\n")
```

### Collapsing to one row per patient

The master keeps exactly one row per patient. Where a patient contributed several rows,
numeric columns are **averaged** and non-numeric columns take the first non-missing
value (num_or_first() in attend_harmonise.R). That is harmless for a score measured
once and lossy for anything measured repeatedly, so the table below counts, per dataset,
how many patients were affected — that count is the size of the hidden heterogeneity.

```{r collapse-report}
collapse_report <- bind_rows(
  Map(function(df, cfg) {
    pid <- pid_vector(df, cfg); pid <- pid[!is.na(pid)]
    tibble(n_rows_mapped        = length(pid),
           n_patients           = dplyr::n_distinct(pid),
           n_multi_row_patients = sum(table(pid) > 1))
  }, tables, id_cfg[names(tables)]),
  .id = "dataset"
)
collapse_report
```

### Assembling the master table

Start from the **union** of every patient any dataset reached, left_join each collapsed
dataset onto it (so an absent modality becomes NA rather than dropping the patient),
prefix each dataset's columns with its name (tmb__, aneu__, …) so nothing collides,
and add an in_<dataset> flag per modality.

**The MAF is deliberately excluded from the averaging.** Mean-collapsing mutation rows is
meaningless — the average of a genomic position is not a genomic position. Instead
mutation status is derived per patient as booleans by pathogenic_by_patient()
(attend_classes.R), which is the same helper the downstream mutation reports call, so
they cannot drift from this table.


#### The two TMB definitions

The recompute reads the gnomAD-annotated long MAF (data/variant_annotations/, loaded by
load_maf_tmb()), counts the variant classes listed in attend_tmb_recompute, and anchors
the result to the upstream tmb__TMB_SCORE scale so the definitions differ only by the
germline filter. apply_nassar_recalibration() then produces tmb__TMB_nassar on top of
the robust recompute. If the annotated MAFs are not synced, load_maf_tmb() returns empty
and both recomputed columns are simply absent from the master.


#### Write the integration point

The master is written through write_intermediate(), which picks Parquet when arrow is
installed and CSV otherwise, so no report ever hard-codes an extension. This is the file
every report from 11 onward reads, and nothing downstream re-reads the raw loaders.


attend_master_joined is the contract every later report builds on, so a broken join
should fail **here and loudly** rather than surfacing as a strange survival curve three
reports later. This asserts the structural invariants of the join: pid exists, is never
null, and is unique — one row per patient. Guarded and warn-level, so it never halts a
knit.

```{r pointblank-gate}
if (requireNamespace("pointblank", quietly = TRUE)) {
  pointblank::create_agent(master, label = "ATTEND master-join QC") |>
    pointblank::col_exists("pid") |>
    pointblank::col_vals_not_null("pid") |>
    pointblank::rows_distinct(columns = "pid") |>
    pointblank::interrogate()
} else {
  message("pointblank not installed — renv::install('pointblank') to enable the QC gate.")
}
```

Each raw table is additionally written to its own intermediate next to the master, so
every input is preserved individually and downstream reports that need raw rows (rather
than the collapsed master) can read them back through read_intermediate().


## Part 2 — what the join cost

A join that silently drops a third of a modality is worse than no join, so the losses are
counted here rather than discovered downstream. "Unmapped" means three different things —
at record level, at patient level, and at cohort level — and each is reported separately.

Presence is not the same as usable, so the last section checks the *contents* of the table
report 01 wrote. That file is the contract every later report inherits, and a bad value is
cheapest to catch here.

### Records that never reached a patient

Native IDs that failed to resolve to any patient — orphaned records. The first table
counts them per dataset; the second lists them.

```{r unmapped}
mapres <- Map(function(df, cfg) {
  ids <- as.character(df[[cfg$col]])
  pid <- pid_vector(df, cfg)
  ok  <- !is.na(pid)
  list(n_ids      = length(unique(ids)),
       mapped     = unique(pid[ok]),
       unmapped   = unique(ids[!ok]),
       mapped_ids = unique(ids[ok]))
}, tables, id_cfg[names(tables)])

unmapped_tbl <- mapres |>
  lapply(\(x) tibble(unmapped_id = x$unmapped)) |>
  bind_rows(.id = "dataset")

unmapped_tbl |> count(dataset, name = "n_unmapped")
print(unmapped_tbl, n = 50)
```

A raw orphan count overstates real loss, because some orphans differ from a known key
only by case or whitespace. This splits the barcode-keyed datasets' orphans into
recoverable formatting drift and genuine absence.

```{r unmapped-nearmiss}
norm <- function(x) toupper(trimws(as.character(x)))
barcode_pool <- unique(norm(cw_barcode_pid$barcode))

unmapped_tbl |>
  filter(dataset %in% barcode_datasets) |>
  mutate(recoverable_by_norm = norm(unmapped_id) %in% barcode_pool) |>
  count(dataset, recoverable_by_norm)
```

#### What safe recovery actually reclaimed

The master was built with recover = TRUE, so the near-misses above were acted on, not
just counted. First table: records that are orphans under exact matching but resolve to a
single patient once normalised. Second table: normalised keys that map to more than one
patient, refused rather than guessed, because recovering them could attach a sample to the
wrong patient.

```{r recovery-ledger}
pid_vector_exact <- make_pid_vector(cw_barcode_pid, cw_image_pid, recover = FALSE)

recoverable <- names(tables)[vapply(names(tables),
  function(nm) id_cfg[[nm]]$space %in% c("barcode", "image"), logical(1))]
reclaimed <- lapply(recoverable, function(nm) {
  df  <- tables[[nm]]; cfg <- id_cfg[[nm]]
  if (is.null(df)) return(NULL)
  ex  <- pid_vector_exact(df, cfg)
  rc  <- pid_vector(df, cfg)          # recover = TRUE, the one used for the master
  got <- is.na(ex) & !is.na(rc)
  if (!any(got)) return(NULL)
  tibble(dataset = nm,
         space         = cfg$space,
         orphaned_id   = as.character(df[[cfg$col]])[got],
         norm_key      = norm_id(as.character(df[[cfg$col]])[got]),
         recovered_pid = rc[got])
}) |> bind_rows()

cat("Patients reclaimed by safe recovery:",
    dplyr::n_distinct(reclaimed$recovered_pid), "distinct pid(s) across",
    nrow(reclaimed), "reclaimed ID row(s)\n")
if (nrow(reclaimed)) print(reclaimed, n = 50)

collisions_of <- function(cw, key) cw |>
  mutate(nkey = norm_id(.data[[key]])) |>
  group_by(nkey) |>
  summarise(n_pid = dplyr::n_distinct(pid),
            pids  = paste(unique(pid), collapse = "; "), .groups = "drop") |>
  filter(n_pid > 1)
collisions_bc  <- collisions_of(cw_barcode_pid, "barcode")
collisions_img <- collisions_of(cw_image_pid,   "image")
cat("\nAmbiguous keys REFUSED (would risk the wrong patient) — barcode:",
    nrow(collisions_bc), "| image:", nrow(collisions_img), "\n")
if (nrow(collisions_bc))  { cat("-- barcode --\n"); print(collisions_bc,  n = 50) }
if (nrow(collisions_img)) { cat("-- image --\n");   print(collisions_img, n = 50) }
```

### Coverage per dataset

Per dataset: native IDs held, mapped, and lost, plus distinct patients reached in a
separate column because it is a different unit. A patient with three samples is three IDs
and one patient, so the two columns must not be read against each other.

```{r loss-table}
loss <- tibble(
  dataset         = names(mapres),
  n_native_ids    = vapply(mapres, `[[`, integer(1), "n_ids"),
  n_ids_mapped    = vapply(mapres, function(x) length(x$mapped_ids), integer(1)),
  n_ids_unmapped  = vapply(mapres, function(x) length(x$unmapped),  integer(1)),
  n_distinct_pids = vapply(mapres, function(x) length(x$mapped),     integer(1))
)
loss

cat("Patients present in ALL datasets:", length(Reduce(intersect, sets)), "\n")
cat("Patients present in ANY dataset:", length(Reduce(union, sets)), "\n")
```

The same table as a figure. Each bar is one dataset's **native IDs**, split into mapped
(grey) and unmapped (red) — both in the same unit, so the bar is genuinely additive
and its full length is the dataset's raw size. Red is where linkage is failing.

```{r plot-loss}
loss_long <- loss |>
  transmute(dataset, mapped = n_ids_mapped, unmapped = n_ids_unmapped) |>
  pivot_longer(-dataset, names_to = "status", values_to = "n")

ggplot(loss_long, aes(reorder(dataset, n), n, fill = status)) +
  geom_col() + coord_flip() +
  # Rule [A]: the state a reader should NOT look at (successfully mapped) is grey, so the
  # unmapped remainder is the only thing carrying colour. Was a green/orange pair from no
  # palette in particular, which made "mapped" the loudest element of a loss figure.
  scale_fill_manual(values = c(mapped = attend_neutral,
                               unmapped = unname(ATTEND_PALETTE["red"]))) +
  labs(x = NULL, y = "Native IDs (count)", fill = NULL,
       title    = "Per-dataset coverage and ID-conversion loss",
       subtitle = "Native-ID units; mapped + unmapped is additive") +
  attend_theme()
```

### Which patients have which modalities

Patients per combination of datasets, from the sets built in setup by build_sets():
horizontal bars are each dataset's patient count, the dot matrix names a combination, and
the top bars count it. The intersections are **exclusive** — a column with dots on only
{imaging, ihc} is patients with imaging and IHC and nothing else — so every patient falls
in exactly one column and the top bars sum to the distinct-patient total.

```{r plot-upset, fig.width=10, fig.height=6}
upset(fromList(sets),
      nsets = length(sets), nintersects = NA,
      order.by = "freq",
      mainbar.y.label = "patients in this EXACT combination",
      sets.x.label    = "patients per dataset",
      text.scale = c(1.3, 1.2, 1.1, 1.0, 1.2, 1.1))
```

UpSet answers *which* combination; this answers *how complete*. For each **k**, the
number of patients appearing in exactly *k* of the eight datasets. The right-most bar is
the fully complete cohort and the left-most are single-modality patients — usually the
number people actually want.

```{r plot-degree}
all_pids   <- Reduce(union, sets)
membership <- sapply(sets, function(s) all_pids %in% s)
degree_tbl <- tibble(pid = all_pids, k = rowSums(membership))

ggplot(degree_tbl, aes(factor(k))) +
  geom_bar(fill = attend_pal(1)) +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.3, size = 3.5) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(title = "Cohort completeness",
       subtitle = paste0("n = ", nrow(degree_tbl), " patients across ", length(sets), " datasets"),
       x = "Number of datasets a patient appears in (k)", y = "Patients (count)") +
  attend_theme()
```

The per-patient view of the same information, read off the master's in_<dataset> flags.
One row per patient, ordered by how many modalities they have (least complete at the
bottom); blue = present, pale grey = absent. Vertical pale stripes are the bottleneck
modalities; the solid blue block at the top is the complete cohort.

```{r plot-availability, fig.width=8, fig.height=7}
presence <- master |>
  select(pid, starts_with("in_")) |>
  pivot_longer(-pid, names_to = "dataset", values_to = "present") |>
  mutate(dataset = sub("^in_", "", dataset))

pid_order <- presence |> group_by(pid) |> summarise(k = sum(present), .groups = "drop") |>
  arrange(k) |> pull(pid)

ggplot(presence, aes(dataset, factor(pid, levels = pid_order), fill = present)) +
  geom_tile() +
  scale_fill_manual(values = c(`TRUE` = attend_pal(1), `FALSE` = "grey93")) +
  labs(title = "Patient x dataset availability",
       subtitle = paste0("n = ", dplyr::n_distinct(presence$pid), " patients"),
       x = NULL, y = "Patients (one row each)", fill = "present") +
  attend_theme() +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1), panel.grid = element_blank())
```

#### Two datasets that are loaded but never joined

clinical_data and conversion_data are saved but are not in tables, so neither is
harmonised into the master — only gianlu_clinical_data supplies clinical fields. This
checks whether clinical_data's IDs are even in pid space, which decides how much work
joining it would be.

```{r clinical-overlap}
clin_ids <- unique(as.character(clinical_data$ID))
cat("clinical_data IDs that are in the pid union:",
    sum(clin_ids %in% Reduce(union, sets)), "of", length(clin_ids), "\n")
```

### Are the joined values sane?

A patient can be flagged in_tmb == TRUE and still have a missing or impossible TMB
value, so presence and usability are checked separately.

Fraction of missing values per column of the master, worst first. A payload column near
1.0 is either an unset placeholder name or a modality that mostly failed to join.

```{r master-missingness}
col_missing <- master |>
  summarise(across(everything(), ~ mean(is.na(.)))) |>
  pivot_longer(everything(), names_to = "column", values_to = "frac_NA") |>
  arrange(desc(frac_NA))

print(col_missing, n = Inf)
```

The same picture with a ranked plot (naniar) and a typed per-column summary (skimr).
Both optional.

```{r eda-naniar-skimr}
if (requireNamespace("naniar", quietly = TRUE)) {
  print(naniar::miss_var_summary(master), n = 25)
  print(naniar::gg_miss_var(master) +
          ggplot2::labs(title = "Missingness per column (master)"))
} else message("naniar not installed — skipping missingness summary/plot.")

if (requireNamespace("skimr", quietly = TRUE)) skimr::skim(master) else
  message("skimr not installed — skipping typed EDA summary.")
```

Range sanity on the four genomic score tables (data/-side, read back from the
intermediates), computed *before* collapsing so a bad input value is not masked by
averaging. Per numeric column: missing/zero/negative rates, observed range, distinct-value
count, and a robust outlier count at median ± 5·MAD. The MAF is excluded — its numeric
columns are genomic coordinates, where "range" means nothing.

```{r value-audit}
score_tables <- list(tmb = tmb_data, msi = msi_data, hrd = hrd_data, aneu = aneu_data)

value_audit <- imap(score_tables, ~ numeric_audit(.x, .y)) |> bind_rows()
print(value_audit, n = Inf)
```

Optional **hard domain bounds** — anything outside [lower, upper] is flagged. Left
empty so it runs as-is; the commented rows are a template with plausible ranges to
confirm against the real definitions.

```{r value-bounds}
score_bounds <- tribble(
  ~dataset, ~column,       ~lower, ~upper,
  # "tmb",  "TMB",          0,      Inf,   # mutations/Mb, non-negative
  # "msi",  "msi_score",    0,      1,     # fraction of unstable loci
  # "hrd",  "HRD_score",    0,      100,   # composite HRD score
  # "aneu", "aneuploidy",   0,      1,     # fraction of genome altered
)

if (nrow(score_bounds) > 0) {
  pmap_dfr(score_bounds, function(dataset, column, lower, upper) {
    df <- score_tables[[dataset]]
    if (is.null(df) || !column %in% names(df))
      return(tibble(dataset, column, status = "column not found",
                    n_violations = NA_integer_))
    x <- suppressWarnings(as.numeric(df[[column]]))
    tibble(dataset, column, status = "checked",
           n_violations = sum(x < lower | x > upper, na.rm = TRUE))
  })
} else {
  message("score_bounds empty — fill in column names to enable hard-bound checks.")
}
```

One histogram per numeric score column of the master, so these are the post-collapse
values every downstream report sees. Look for impossible values, a spike at zero (a failed
measurement coded as 0 rather than missing), and bimodality, which the mean collapse would
blur.

```{r plot-score-dist, fig.width=9, fig.height=5}
score_long <- master |>
  select(pid, matches("^(tmb|msi|hrd|aneu)__")) |>
  select(pid, where(is.numeric)) |>
  pivot_longer(-pid, names_to = "metric", values_to = "value") |>
  filter(!is.na(value))

if (nrow(score_long) > 0) {
  ggplot(score_long, aes(value)) +
    geom_histogram(bins = 30, fill = attend_pal(1)) +
    facet_wrap(~ metric, scales = "free", ncol = 3) +
    labs(title = "Distributions of the genomic score columns (master)",
         subtitle = paste0("Check for impossible values, spikes at 0, and bimodality; n = ",
                           dplyr::n_distinct(score_long$pid), " patients"),
         x = "Score (units differ per panel; see 00-methods)", y = "Patients (count)") +
    attend_theme()
} else {
  message("No numeric score columns matched in master — check column prefixes.")
}
```

The continuous WES MSI score (msi__MSI_SCORE, percent unstable sites, from
data/msi.20250831.csv) against the clinical MSI status the clinical table calls
independently. The two arrive through different ID spaces — score through the barcode
crosswalk, status through the patient id — so clean separation is positive evidence that
the barcode-to-pid mapping is right, and overlap points at a linkage fault rather than
biology. Wilcoxon rank-sum on the two groups.

```{r msi-score-vs-status}
msi_score_col <- "msi__MSI_SCORE"   # continuous WES score; categorical = attend_cols$msi_status

if (have_cols(master, msi_score_col, attend_cols$msi_status)) {
  msi_df <- master |>
    transmute(pid,
              status = .data[[attend_cols$msi_status]],
              score  = suppressWarnings(as.numeric(.data[[msi_score_col]]))) |>
    filter(!is.na(status), !is.na(score))

  p_lab <- tryCatch({
    grp <- split(msi_df$score, msi_df$status)
    if (length(grp) == 2)
      sprintf("Wilcoxon p = %s",
              format.pval(wilcox.test(grp[[1]], grp[[2]])$p.value, digits = 2, eps = 1e-4))
    else NULL
  }, error = function(e) NULL)

  ggplot(msi_df, aes(status, score)) +
    # points = FALSE: highlight_points() below supplies the partitioned point layer, so
    # attend_box() contributes only the box (or, where a stratum is under n=10, the median
    # crossbar) and the per-group n. The old fill was #A6CEE3 — which attend_plots.R defines
    # as "aneuploidy-low", a meaning this figure does not have.
    attend_box(msi_df, x = "status", y = "score", points = FALSE) +
    highlight_points(msi_df, "status", "score", base_size = 0.7, base_alpha = 0.4, base_const = "black") +
    labs(title = "WES MSI score by clinical MSI status",
         subtitle = p_lab, x = "Clinical MSI status",
         y = "WES MSI score (% unstable sites)") +
    attend_theme()
} else {
  message("Both msi__MSI_SCORE and ", attend_cols$msi_status,
          " must be in the master — skipping MSI concordance plot.")
}
```

```{r session-info}
sessionInfo()
```
