---
title: "ATTEND — 20. FlowPath vs the pathologist"
author: "sceriff0"
date: "2026-08-11"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
# renv: tidyverse, here (+ optional gtsummary for the cohort table).
#   renv::install("gtsummary"); renv::snapshot()
library(tidyverse)
library(here)
# Headless-safe raster device. The cluster's compute nodes have no X11, so R's default
# png() fails with "unable to open connection to X11 display" and every figure dies. cairo
# needs no display; the guard keeps this working on a machine built without it.
if (capabilities("cairo")) options(bitmapType = "cairo")

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 6)

source(here("code", "load_phenotypes.R"))    # load_ihc_data()
source(here("code", "load_clinical.R"))      # load_imaging_data(), load_gianlu_clinical_data()
source(here("code", "attend_harmonise.R"))   # crosswalks + pid_vector
source(here("code", "attend_classes.R"))     # attend_cols, attend_highlight
source(here("code", "attend_plots.R"))       # highlight_points() overlay
source(here("code", "attend_ihc.R"))         # FlowPath IHC metric definitions

ihc_data     <- load_ihc_data()        # FlowPath per-image cell counts, keyed by image_id
imaging_data <- load_imaging_data()    # pathologist annotations, keyed by image_id
```

> **Question.** Does FlowPath's automated quantification agree with the pathologist's eye?
> **Cohort.** All patients with IHC data (`r dplyr::n_distinct(ihc_data$image_id)` images); no molecular restriction.
> **Reads.** The FlowPath export and the imaging annotation sheet, via load_ihc_data() and load_imaging_data().
> **Out of scope.** What composition means biologically -> [05_aneuploidy_mmrd](05_aneuploidy_mmrd.html). Survival by immune content -> [04_survival](04_survival.html).

FlowPath counts individual cells on each IHC slide and turns them into positivity and
composition fractions; a pathologist independently scored the same slides by eye. This is
a measurement-validation question, so it runs on every imaged patient — restricting it to
a molecular subgroup would only make it noisier.

## Loading and mapping to patients

The FlowPath export is keyed by image_id, the imaging annotation sheet carries both
image_id and the patient id, and the master is keyed by pid. Everything is put onto
pid through the image crosswalk built in code/attend_harmonise.R — the same one
[01_data_integration](01_data_integration.html) used, so this report's cohort cannot drift
from the master's. The two counts below are the ceiling for everything that follows: rows
that fail to map are invisible to every figure in this report.

```{r load}
id_cfg         <- make_id_cfg()
cw_image_pid   <- build_image_pid(imaging_data, id_cfg)   # image -> pid
cw_barcode_pid <- build_barcode_pid(load_gianlu_clinical_data(), id_cfg)
pid_vector     <- make_pid_vector(cw_barcode_pid, cw_image_pid)

ihc_pid     <- ihc_data     |> mutate(pid = pid_vector(ihc_data,     id_cfg$ihc))
imaging_pid <- imaging_data |> mutate(pid = pid_vector(imaging_data, id_cfg$imaging))

cat("IHC rows mapped to a patient    :", sum(!is.na(ihc_pid$pid)),     "/", nrow(ihc_pid), "\n")
cat("imaging rows mapped to a patient:", sum(!is.na(imaging_pid$pid)), "/", nrow(imaging_pid), "\n")
```

## The imaged cohort

Per-image FlowPath counts joined to the pathologist annotations, with the
positivity/composition ratios derived by ihc_image_metrics() (code/attend_ihc.R). The
summary table describes the cohort the IHC slides came from — not the whole ATTEND
cohort, which is the relevant denominator here.

```{r ihc-metrics}
ihc_img <- ihc_image_metrics(ihc_data, imaging_data)
cat("IHC images joined to imaging annotations:", sum(!is.na(ihc_img$ID)), "/", nrow(ihc_img), "\n")

if (requireNamespace("gtsummary", quietly = TRUE)) {
  ihc_img |>
    select(any_of(c("MSI_Status", "treatment", "ARID1A", "MMR_Status"))) |>
    gtsummary::tbl_summary()
}
```

## Concordance — continuous measures

For these three, the pathologist also gave a **number**, so the two methods can be
plotted against each other directly. Each point is one image. The dashed red line is
x = y, perfect agreement: points systematically above it mean the pathologist scored
higher than FlowPath, below it the reverse. Spearman's rho measures whether the two
**rank** images the same way, which is the weaker but more robust claim — a high rho with
points far off the diagonal means consistent but miscalibrated.

Pathologist percentages are divided by 100 so both axes are fractions.

### Neoplastic cellularity

FlowPath tumour fraction = all tumour cells / all cells, against the pathologist's
neoplastic-cellularity percentage.

```{r concordance-cellularity}
cc <- ihc_img |>
  transmute(ID,                                          # ID == pid; kept so the highlight can match
            flowpath    = tumor_over_all,
            pathologist = suppressWarnings(as.numeric(Neoplastic_cellularity)) / 100) |>
  filter(!is.na(flowpath), !is.na(pathologist))

if (nrow(cc) > 2) {
  rho <- cor(cc$flowpath, cc$pathologist, method = "spearman", use = "complete.obs")
  ggplot(cc, aes(flowpath, pathologist)) +
    highlight_points(cc, "flowpath", "pathologist", jitter_width = 0, base_size = 1.5, base_alpha = 0.6) +
    geom_abline(slope = 1, linetype = "dashed", colour = "red") +
    coord_equal() +
    annotate("text", x = -Inf, y = Inf, hjust = -0.2, vjust = 2,
             label = paste0("Spearman = ", round(rho, 2))) +
    labs(title = "Neoplastic cellularity", x = "FlowPath tumour fraction",
         y = "Pathologist cellularity") +
    theme_classic(base_size = 12)
}
```

### Immune infiltrate

FlowPath immune content = CD45+ cells / all cells inside the annotation, against the
pathologist's immune-infiltrate percentage.

```{r concordance-immune}
ii <- ihc_img |>
  transmute(ID,
            flowpath    = cd45_over_inside,
            pathologist = suppressWarnings(as.numeric(Immune_infiltrate)) / 100) |>
  filter(!is.na(flowpath), !is.na(pathologist))

if (nrow(ii) > 2) {
  rho <- cor(ii$flowpath, ii$pathologist, method = "spearman", use = "complete.obs")
  ggplot(ii, aes(flowpath, pathologist)) +
    highlight_points(ii, "flowpath", "pathologist", jitter_width = 0, base_size = 1.5, base_alpha = 0.6) +
    geom_abline(slope = 1, linetype = "dashed", colour = "red") +
    coord_equal() +
    annotate("text", x = -Inf, y = Inf, hjust = -0.2, vjust = 2,
             label = paste0("Spearman = ", round(rho, 2))) +
    labs(title = "Immune infiltrate", x = "FlowPath CD45 fraction (inside)",
         y = "Pathologist immune infiltrate") +
    theme_classic(base_size = 12)
}
```

### PD-1 on immune cells

FlowPath PD-1+ immune = PD-1+ CD45 cells / CD45 cells inside the annotation. A linear
trend line replaces the x = y line here, because the pathologist's PD-1 score is not on
the same scale as the FlowPath fraction — only the ranking is comparable.

```{r concordance-pd1}
pd1 <- ihc_img |>
  transmute(ID,
            flowpath    = cd45_pd1_over_cd45_inside,
            pathologist = suppressWarnings(as.numeric(Immune_infiltrate_PD1)) / 100) |>
  filter(!is.na(flowpath), !is.na(pathologist))

if (nrow(pd1) > 2) {
  rho <- cor(pd1$flowpath, pd1$pathologist, method = "spearman", use = "complete.obs")
  ggplot(pd1, aes(flowpath, pathologist)) +
    highlight_points(pd1, "flowpath", "pathologist", jitter_width = 0, base_size = 1.5, base_alpha = 0.6) +
    geom_smooth(method = "lm", se = FALSE, colour = "grey40", linewidth = 0.6) +
    annotate("text", x = -Inf, y = Inf, hjust = -0.2, vjust = 2,
             label = paste0("Spearman = ", round(rho, 2))) +
    labs(title = "PD-1 on immune cells", x = "FlowPath PD-1+ / CD45 (inside)",
         y = "Pathologist PD-1 infiltrate") +
    theme_classic(base_size = 12)
}
```

## Concordance — categorical calls

For these three the pathologist gave a **category**, not a number, so the comparison is
a boxplot: the FlowPath fraction distribution within each pathologist call. Concordance
looks like boxes that step monotonically across the categories. Overlapping boxes mean
the automated fraction does not reproduce the call.

### ARID1A

FlowPath ARID1A+ = ARID1A+ tumour cells / tumour cells inside the annotation. The
pathologist's categories run from complete loss through clonal loss to positive, so the
FlowPath fraction should **increase** left to right.

```{r concordance-aridia}
if ("ARID1A" %in% names(ihc_img)) {
  d <- ihc_img |>
    mutate(ARID1A = factor(ARID1A, levels = c("Complete loss", "Clonal loss", "Positive"))) |>
    filter(!is.na(ARID1A), !is.na(tumor_aridia_over_tumor_inside))
  ggplot(d, aes(ARID1A, tumor_aridia_over_tumor_inside, colour = ARID1A)) +
      geom_boxplot(outlier.shape = NA) +
      highlight_points(d, "ARID1A", "tumor_aridia_over_tumor_inside", base_colour = "ARID1A",
                       base_name = "Pathologist", jitter_width = 0.2, base_size = 1.5, base_alpha = 0.6) +
      labs(title = "ARID1A on tumour cells", x = NULL, y = "FlowPath positivity (fraction)") +
      theme_classic(base_size = 12)
}
```

### PD-L1 on immune cells

FlowPath PD-L1+ immune = PD-L1+ CD45 cells / CD45 cells inside. The pathologist's IC
categories are explicit percentage bands (IC0 < 1%, IC1 1–5%, IC2 5–10%), so this is the
strictest of the categorical checks — the FlowPath fraction should land inside the named
band, not merely increase across bands.

```{r concordance-pdl1}
if ("Immune_infiltrate_PDL1" %in% names(ihc_img)) {
  d <- ihc_img |>
    mutate(Immune_infiltrate_PDL1 = factor(Immune_infiltrate_PDL1,
             levels = c("", "IC0 (<1%)", "IC1 (>=1% and <5%)", "IC2 (>=5% and <10%)"))) |>
    filter(!is.na(Immune_infiltrate_PDL1), !is.na(cd45_pdl1_over_cd45_inside))
  ggplot(d, aes(Immune_infiltrate_PDL1, cd45_pdl1_over_cd45_inside, colour = Immune_infiltrate_PDL1)) +
      geom_boxplot(outlier.shape = NA) +
      highlight_points(d, "Immune_infiltrate_PDL1", "cd45_pdl1_over_cd45_inside",
                       base_colour = "Immune_infiltrate_PDL1", base_name = "Pathologist",
                       jitter_width = 0.2, base_size = 1.5, base_alpha = 0.6) +
      labs(title = "PD-L1 on immune cells", x = NULL, y = "FlowPath positivity (fraction)") +
      theme_classic(base_size = 12)
}
```

### PD-L1 on tumour cells

FlowPath PD-L1+ tumour = PD-L1+ tumour cells / tumour cells inside the annotation.

```{r concordance-tcpdl1}
if ("Tumor_cells_PDL1" %in% names(ihc_img)) {
  d <- ihc_img |>
    filter(!is.na(Tumor_cells_PDL1), Tumor_cells_PDL1 != "", !is.na(tumor_pdl1_over_tumor_inside))
  ggplot(d, aes(Tumor_cells_PDL1, tumor_pdl1_over_tumor_inside, colour = Tumor_cells_PDL1)) +
      geom_boxplot(outlier.shape = NA) +
      highlight_points(d, "Tumor_cells_PDL1", "tumor_pdl1_over_tumor_inside",
                       base_colour = "Tumor_cells_PDL1", base_name = "Pathologist",
                       jitter_width = 0.2, base_size = 1.5, base_alpha = 0.6) +
      labs(title = "PD-L1 on tumour cells", x = NULL, y = "FlowPath positivity (fraction)") +
      theme_classic(base_size = 12) +
      theme(axis.text.x = element_text(angle = 20, hjust = 1))
}
```

## Per-patient IHC metrics

Everything above is per **image**. A patient can contribute several images, so the
metrics are collapsed to one row per patient (mean across that patient's images) before
being used against patient-level variables. ihc_patient_metrics() is the entry point
[04_survival](04_survival.html) and [05_aneuploidy_mmrd](05_aneuploidy_mmrd.html) call to
get the same per-patient values.

```{r ihc-patient}
ihc_pt <- ihc_patient_metrics(ihc_data, imaging_data)
glimpse(ihc_pt)
```

```{r session-info}
sessionInfo()
```
