---
title: "ATTEND — 03. Covariate distributions by MMR status"
author: "sceriff0"
date: "2026-08-11"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
library(here)
# Headless-safe raster device. The cluster's compute nodes have no X11, so R's default
# png() fails with "unable to open connection to X11 display" and every figure dies. cairo
# needs no display; the guard keeps this working on a machine built without it.
if (capabilities("cairo")) options(bitmapType = "cairo")

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 10, fig.height = 7)

source(here("code", "attend_classes.R"))   # attend_cols / attend_levels / attend_thresholds
source(here("code", "attend_plots.R"))     # attend_theme(), attend_mmr_cols
source(here("code", "attend_io.R"))        # read_intermediate()
source(here("code", "build_master.R"))   # get_master() — builds on demand

master <- get_master()
```

> **Question.** Do the continuous molecular covariates differ in distribution between MMR-deficient and MMR-proficient tumours?
> **Cohort.** All `r nrow(master)` patients on the master; each panel drops its own missing values.
> **Reads.** attend_master_joined only.
> **Out of scope.** Anything conditioned on aneuploidy class -> reports [05](05-aneuploidy-mmrd.html) to [08](08-tmb.html). Survival -> [04-survival](04-survival.html).

Whole distributions, not group averages. The expected differences here are tail phenomena:
MMR-deficient tumours carry higher TMB and MSI through a hypermutated *subset* rather than
a uniform shift, which an ECDF shows plainly and a median hides.

### Why an ECDF rather than a boxplot

An empirical cumulative distribution function plots, for every value on the x-axis, the
fraction of patients at or below it. The curve rises from 0 to 1 left to right; where the
two curves separate says *where in the range* the groups differ, a curve further right
means systematically higher values, and the largest vertical gap between them is the KS
statistic D tested below.

## Data

Covariates are mapped to readable labels here, and only those actually present on the
master are kept, so a covariate that failed to join is dropped rather than drawn empty.
MMR status is attend_cols$mmr_status, collapsed to deficient
(`r attend_levels$mmr_deficient`) versus everything else non-missing.

```{r data}
# continuous covariates (prefixed master columns) -> readable labels
covar_map <- c(
  "aneuploidy score" = attend_cols$aneuploidy,   # aneu__aneuploidy_score
  "HRD score"        = attend_cols$hrd,          # hrd__HRD_Score
  "MSI (% unstable)" = "msi__MSI_SCORE"
)

# The two TMB definitions, under their reader-facing names. tmb_defs_present() returns
# normal first, then ancestry corrected (only if it is on the master), so this cannot
# drift from the TMB panels in any other report.
tmb_defs06 <- tmb_defs_present(master)
tmb_map    <- setNames(unname(tmb_defs06), paste0(tmb_def_labels(tmb_defs06), " (mut/Mb)"))
covar_map  <- c(covar_map, tmb_map)

covar_map <- covar_map[covar_map %in% names(master)]   # keep only those present

mmr_col <- attend_cols$mmr_status                       # gianlu__MMR_STATUS
cat("covariates available:", paste(names(covar_map), collapse = ", "), "\n")
```

## ECDF by MMR status

One panel per covariate. Each curve is a group's cumulative distribution; the x-axis is
the covariate's own scale (panels are free-scaled, so they are not comparable to each
other). Patients missing a value are dropped from that panel only, so panel denominators
differ — the KS table below reports each panel's group sizes.

```{r ecdf}
if (length(covar_map) > 0 && mmr_col %in% names(master)) {
  long <- master |>
    transmute(
      MMR = case_when(
        .data[[mmr_col]] == attend_levels$mmr_deficient ~ "MMR deficient",
        !is.na(.data[[mmr_col]])                         ~ "MMR proficient",
        TRUE                                            ~ NA_character_),
      across(all_of(unname(covar_map)), ~ suppressWarnings(as.numeric(.x)))
    ) |>
    pivot_longer(all_of(unname(covar_map)), names_to = "covariate", values_to = "value") |>
    mutate(covariate = names(covar_map)[match(covariate, covar_map)]) |>
    filter(!is.na(MMR), !is.na(value))

  ggplot(long, aes(value, colour = MMR)) +
    stat_ecdf(linewidth = 0.9) +
    facet_wrap(~ covariate, scales = "free_x") +
    # colours are the shared attend_mmr_cols (proficient grey / deficient orange)
    scale_colour_manual(values = attend_mmr_cols) +
    labs(title = "ECDF of molecular covariates by MMR status",
         x = "value", y = "F(x) — cumulative fraction of patients", colour = NULL) +
    attend_theme() +
    theme(legend.position = "top")
} else {
  message("No covariate columns or the MMR column are present in the master.")
}
```

## Per-covariate Kolmogorov–Smirnov test

The formal counterpart to the figure. The two-sample KS test asks whether the two
distributions differ **anywhere**, without assuming a shape or testing only the mean.
D is the largest vertical gap between the two curves above — so the test statistic is
literally the thing the eye picks out of the plot.

n_deficient and n_proficient are that covariate's own group sizes after dropping
missing values, which is why they vary between rows.

**No multiple-testing correction is applied** across the covariates listed. With a handful
of pre-specified covariates that is defensible, but the p-values are per-covariate and
should be read that way. Note also that the two TMB rows are **not independent** — they
are the same patients under two definitions — so they cannot be treated as two separate
pieces of evidence.

```{r ks}
if (length(covar_map) > 0 && mmr_col %in% names(master)) {
  mmr <- case_when(
    master[[mmr_col]] == attend_levels$mmr_deficient ~ "deficient",
    !is.na(master[[mmr_col]])                         ~ "proficient",
    TRUE                                             ~ NA_character_)

  imap_dfr(covar_map, function(col, label) {
    d <- tibble(x = suppressWarnings(as.numeric(master[[col]])), mmr = mmr) |>
      filter(!is.na(x), !is.na(mmr))
    g <- split(d$x, d$mmr)
    if (length(g) == 2) {
      kt <- suppressWarnings(ks.test(g[["deficient"]], g[["proficient"]]))
      tibble(covariate = label,
             n_deficient = length(g[["deficient"]]), n_proficient = length(g[["proficient"]]),
             D = round(unname(kt$statistic), 3), p = signif(kt$p.value, 3))
    } else {
      tibble(covariate = label, n_deficient = NA_integer_, n_proficient = NA_integer_,
             D = NA_real_, p = NA_real_)
    }
  })
}
```

## Aneuploidy score by MMR status

The ECDF above shows the whole aneuploidy distribution; this is the same comparison as a
two-group contrast with a test attached, because aneuploidy is this pipeline's subject and
the ECDF alone gives no effect size or p-value. Aneuploidy score is
aneu__aneuploidy_score (ASCETS arm-level SCNA burden from data/ascets/, LOWCOV arms
excluded from each sample's denominator); the dashed line is the
`r attend_thresholds$aneuploidy` class cutoff, drawn so the boxes can be read against the
split the later reports use. Wilcoxon rank-sum on the two groups, uncorrected.

```{r aneu-by-mmr, fig.width=6, fig.height=5}
aneu_col <- attend_cols$aneuploidy

if (aneu_col %in% names(master) && mmr_col %in% names(master)) {
  am <- master |>
    transmute(
      MMR = case_when(
        .data[[mmr_col]] == attend_levels$mmr_deficient ~ "MMR deficient",
        !is.na(.data[[mmr_col]])                        ~ "MMR proficient",
        TRUE                                            ~ NA_character_),
      aneuploidy = suppressWarnings(as.numeric(.data[[aneu_col]]))) |>
    filter(!is.na(MMR), !is.na(aneuploidy))

  p_lab <- tryCatch({
    grp <- split(am$aneuploidy, am$MMR)
    if (length(grp) == 2)
      sprintf("Wilcoxon p = %s | n = %d deficient, %d proficient",
              format.pval(wilcox.test(grp[[1]], grp[[2]])$p.value, digits = 2, eps = 1e-4),
              length(grp[["MMR deficient"]]), length(grp[["MMR proficient"]]))
    else NULL
  }, error = function(e) NULL)

  ggplot(am, aes(MMR, aneuploidy, fill = MMR)) +
    attend_box(am, x = "MMR", y = "aneuploidy", fill = NULL) +
    geom_hline(yintercept = attend_thresholds$aneuploidy, linetype = 2, colour = "grey40") +
    scale_fill_manual(values = attend_mmr_cols) +
    labs(title = "Aneuploidy score by MMR status",
         subtitle = p_lab, x = NULL, y = "Aneuploidy score (0-1)") +
    attend_theme() + theme(legend.position = "none")
} else {
  message("Need ", aneu_col, " and ", mmr_col, " on the master for the aneuploidy contrast.")
}
```

```{r session-info}
sessionInfo()
```
