---
title: "ATTEND — 3. Aneuploidy vs mutations, within MMR-deficient tumours"
author: "sceriff0"
date: "2026-08-09"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
# renv: needs ggpubr (+ tidyverse, here). Mutation status also uses attend_harmonise.R.
# If missing: renv::install("ggpubr"); renv::snapshot()
library(tidyverse)
library(here)
library(ggpubr)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 7)

source(here("code", "attend_harmonise.R"))   # build_barcode_pid(), make_id_cfg()
source(here("code", "attend_classes.R"))     # column config + class helpers + gene panels
source(here("code", "attend_plots.R"))       # highlight_points() + attend_highlight overlay
source(here("code", "attend_io.R"))          # read_intermediate() (Parquet/CSV)
```

## What this report does

**The entire report is restricted to MMR-deficient (MMRd) tumours.** Nothing below
includes an MMR-proficient patient.

That restriction is the point. MMR deficiency is itself a route to genomic instability,
so in the whole cohort any mutation that tracks MMR status will also appear to track
aneuploidy. Conditioning on MMRd removes that shortcut and asks a sharper question:
**among tumours that are already mismatch-repair deficient, does aneuploidy still vary
with which gene is mutated?**

The outcome variable in every figure is the same — the continuous **aneuploidy score**
(ASCETS, fraction of chromosome arms altered). What changes is the grouping on the
x-axis.

### The gene panel, spelled out

Genes are tested **one at a time**, each in its own panel, so a signal is attributable
to a specific gene rather than to a union that could be driven by whichever gene is most
frequently mutated. Two groups, kept deliberately separate:

| Group | Genes | Why they are grouped this way |
|---|---|---|
| **MMR panel** | `r paste(attend_gene_panel, collapse = ", ")` | Loss of any of these causes mismatch-repair deficiency. MSH3 pairs with MSH2 to form MutSβ. |
| **Wnt / β-catenin** | `r paste(attend_gene_panel_wnt, collapse = ", ")` | An independent oncogenic pathway, not an MMR mechanism. |
| **TP53** | TP53 | Handled separately — it is already a first-class column on the master. |

The **MMR union** (`panel-altered` = at least one pathogenic variant in any MMR panel
gene) is shown last, after the per-gene panels. `CTNNB1` and `APC` are deliberately
**excluded from that union**: they are not MMR genes, and folding them in would change
what "panel-altered" means everywhere else in the pipeline. There is no Wnt union — with
two genes, the per-gene panels already show everything a union would.

### How to read the figures

Each panel is one grouping variable. The **box** summarises the aneuploidy-score
distribution in each group. The **points** are individual patients, coloured by their
aneuploidy class (high vs low at `r attend_thresholds$aneuploidy`) so the split driving
the box is visible. Two overlays sit on top: **gold** marks patients who also have IHC /
imaging data, **magenta** marks sample 21S188 (*polipo*).

P-values come from `stat_compare_means()`, which uses Wilcoxon rank-sum for two groups
and Kruskal–Wallis for more. They are **not** corrected for multiple testing across
panels — with one test per gene, treat them as descriptive.

## Data

Two sources, joined on `pid`.

The **aneuploidy score and the score-derived classes** come from
`output/clean_data/attend_master_joined` (report 1), binarised by
`add_molecular_classes()` at the thresholds in `code/attend_classes.R`.

The **mutation status** is recomputed here directly from the long MAF, *not* taken from
the master's MAF columns — those are mean-collapsed, and the average of a variant row is
meaningless. `pathogenic_by_patient()` and `mutation_status_long()` are the same helpers
report 1 used to write `TP53_pathogenic` / `panel_pathogenic`, so the two cannot drift.
Anchoring on the crosswalk and crossing over the full gene list guarantees every gene
gets a column even if it never appears in the MAF, with unmutated patients set to `FALSE`
rather than `NA`.

```{r data}
master <- read_intermediate("attend_master_joined") |>
  add_molecular_classes()

source(here("code", "load_clinical.R"))
source(here("code", "load_wes_results.R"))

cw  <- build_barcode_pid(load_gianlu_clinical_data(), make_id_cfg())
maf <- read_intermediate("maf_data")

mut <- pathogenic_by_patient(maf, cw)   # pid, TP53_pathogenic, panel_pathogenic (MMR union)

# Per-gene status for the one-gene-at-a-time panels: MMR panel + Wnt genes.
gene_status <- cw |> distinct(pid) |>
  tidyr::crossing(gene = attend_genes_per_gene) |>
  left_join(mutation_status_long(maf, cw), by = c("pid", "gene")) |>
  mutate(mutated = replace_na(mutated, FALSE),
         class   = if_else(mutated, paste0(gene, "-mut"), paste0(gene, "-wt"))) |>
  select(pid, gene, class) |>
  pivot_wider(names_from = gene, values_from = class, names_glue = "{gene}_class")

dat_all <- master |>
  select(-any_of(c("TP53_pathogenic", "panel_pathogenic"))) |>  # recompute here
  left_join(mut, by = "pid") |>
  left_join(gene_status, by = "pid") |>
  mutate(
    aneuploidy_value = suppressWarnings(as.numeric(.data[[attend_cols$aneuploidy]])),
    TP53_class  = if_else(TP53_pathogenic,  "TP53-mut",     "TP53-wt"),
    panel_class = if_else(panel_pathogenic, "MMR-panel-altered", "MMR-panel-wt")
  )
```

### Restricting to MMR-deficient

The single filter the whole report rests on. The counts below are the cohort every
figure that follows is drawn from — if the MMRd group is small, every p-value in this
report inherits that.

```{r restrict-mmrd}
have_mmr <- have_cols(dat_all, "MMR_class", "aneuploidy_value")

dat <- if (have_mmr) dat_all |> filter(MMR_class == attend_levels$mmr_deficient) else dat_all[0, ]

cat("Whole cohort              :", nrow(dat_all), "patients\n")
cat("MMR-deficient (this report):", nrow(dat), "patients\n")
if (nrow(dat) > 0)
  cat("  with an aneuploidy score :", sum(!is.na(dat$aneuploidy_value)), "patients\n")

if (have_mmr && nrow(dat) == 0)
  message("No MMR-deficient patients (MMR_class == '", attend_levels$mmr_deficient,
          "') — every figure below is skipped.")
```

Which of the manually highlighted patients actually resolve to a `pid` in this restricted
cohort. Printed so that a silent ID mismatch shows up as a number here rather than as
"the gold points just aren't drawn".

```{r highlight-coverage}
if (nrow(dat) > 0) cat(highlight_coverage(dat))
```

```{r box-helper, include=FALSE}
# Shared boxplot: aneuploidy score on y, one facet per grouping variable. Boxes are
# neutral; points are coloured by aneuploidy class, with the highlight overlay drawn
# on top (partitioned, so no patient is plotted twice). highlight_points() owns both
# colour scales, so no scale_colour_manual is added here.
aneu_box <- function(d, vars, title, subtitle = NULL, ncol = 3) {
  long <- d |>
    select(pid, aneuploidy_value, aneuploidy_class, any_of(vars)) |>
    pivot_longer(any_of(vars), names_to = "comparison", values_to = "group") |>
    filter(!is.na(aneuploidy_value), !is.na(group)) |>
    mutate(comparison = factor(comparison, levels = vars))
  if (nrow(long) == 0) return(NULL)

  ggplot(long, aes(group, aneuploidy_value)) +
    geom_boxplot(outlier.shape = NA, fill = "#A6CEE3", width = 0.6) +
    highlight_points(long, "group", "aneuploidy_value", base_colour = "aneuploidy_class",
                     base_palette = attend_aneu_cols, base_name = "aneuploidy",
                     base_size = 0.7, base_alpha = 0.6) +
    facet_wrap(~ comparison, scales = "free_x", ncol = ncol) +
    stat_compare_means(label = "p.format", size = 3) +
    labs(title = title, subtitle = subtitle, x = NULL, y = "aneuploidy score") +
    attend_theme() +
    theme(axis.text.x = element_text(angle = 20, hjust = 1))
}

# Group sizes behind each panel — a two-group Wilcoxon on 3 vs 40 patients looks
# identical to one on 20 vs 23 unless the counts are printed.
group_counts <- function(d, vars) {
  d |>
    select(pid, aneuploidy_value, any_of(vars)) |>
    pivot_longer(any_of(vars), names_to = "comparison", values_to = "group") |>
    filter(!is.na(group), !is.na(aneuploidy_value)) |>
    count(comparison, group, name = "n") |>
    pivot_wider(names_from = group, values_from = n, values_fill = 0)
}
```

## 1. Aneuploidy by gene — MMR panel, one gene at a time

Each panel is one MMR gene: patients carrying a pathogenic variant in that gene
(`<GENE>-mut`) versus those who do not (`<GENE>-wt`), within the MMRd cohort only.
Pathogenicity is decided by `attend_pathogenic_classes` on the MAF's
`Variant_Classification`.

Read the group sizes first. Some of these genes are mutated in very few patients, and a
panel with a handful of `-mut` patients will show a wide box and an unstable p-value.

```{r mmr-gene-counts}
mmr_gene_vars <- paste0(attend_gene_panel, "_class")
if (nrow(dat) > 0)
  knitr::kable(group_counts(dat, mmr_gene_vars),
               caption = "Patients per group, MMR panel genes (MMR-deficient cohort)")
```

```{r mmr-gene-box, fig.height=7}
if (nrow(dat) > 0)
  aneu_box(dat, mmr_gene_vars,
           "Aneuploidy score by MMR panel gene (MMR-deficient only)",
           paste0("One panel per gene; pathogenic vs wild-type. n = ", nrow(dat),
                  " MMR-deficient patients"))
```

## 2. Aneuploidy by gene — Wnt / β-catenin pathway

The same comparison for `r paste(attend_gene_panel_wnt, collapse = " and ")`. These are
shown apart from the MMR panel because they are a different mechanism: they are not a
cause of mismatch-repair deficiency, so a signal here would mean something biologically
different from a signal in section 1.

```{r wnt-gene-counts}
wnt_gene_vars <- paste0(attend_gene_panel_wnt, "_class")
if (nrow(dat) > 0)
  knitr::kable(group_counts(dat, wnt_gene_vars),
               caption = "Patients per group, Wnt pathway genes (MMR-deficient cohort)")
```

```{r wnt-gene-box, fig.height=4.5}
if (nrow(dat) > 0)
  aneu_box(dat, wnt_gene_vars,
           "Aneuploidy score by Wnt pathway gene (MMR-deficient only)",
           "CTNNB1 and APC — shown separately from the MMR panel; not part of the MMR union",
           ncol = 2)
```

## 3. TP53 and the MMR union

`TP53` on its own, then the MMR panel **union** — at least one pathogenic variant in any
of `r paste(attend_gene_panel, collapse = ", ")`. The union is the last mutation view
rather than the first: it is the least specific, and the per-gene panels above are what
say which gene any union signal came from.

```{r tp53-union-counts}
summary_vars <- c("TP53_class", "panel_class")
if (nrow(dat) > 0)
  knitr::kable(group_counts(dat, summary_vars),
               caption = "Patients per group, TP53 and the MMR union (MMR-deficient cohort)")
```

```{r tp53-union-box, fig.height=4.5}
if (nrow(dat) > 0)
  aneu_box(dat, summary_vars,
           "Aneuploidy score by TP53 status and by the MMR panel union (MMR-deficient only)",
           ncol = 2)
```

## 4. Aneuploidy by the other molecular classes

The non-mutation axes, still within MMRd: TMB, HRD, and MSI status. `MMR_class` is
omitted because it is constant here by construction — every patient in this report is
MMR-deficient.

MSI is worth reading carefully: MMR loss is the usual *cause* of microsatellite
instability, so within an MMRd cohort the MSI-stable group is both small and unusual.

```{r other-classes-box, fig.height=4.5}
other_vars <- c("TMB_class", "HRD_class", "MSI_class")
if (nrow(dat) > 0) {
  knitr::kable(group_counts(dat, other_vars),
               caption = "Patients per group, score-derived classes (MMR-deficient cohort)")
  aneu_box(dat, other_vars,
           "Aneuploidy score by TMB / HRD / MSI class (MMR-deficient only)",
           ncol = 3)
}
```

## 5. Aneuploidy by TMB class — both TMB definitions

Section 4's TMB panel uses the upstream score. Because ATTEND is sequenced tumour-only,
residual germline variants inflate TMB in an ancestry-dependent way (Nassar et al.,
*Cancer Cell* 2022), so the comparison is redrawn here under **both** definitions the
pipeline carries:

- **TMB (normal)** — `tmb__TMB_SCORE`, the upstream tumour-only score.
- **TMB (ancestry corrected)** — `tmb__TMB_nassar`, after the ancestry recalibration.

Both are binarised at the same `r attend_thresholds$tmb` mut/Mb cutoff, so the only
difference between the two facets is which patients crossed the threshold once the
germline correction was applied. The ancestry-corrected facet is dropped rather than
drawn empty until `attend_ancestry$clinical_col` is set.

```{r tmb-definitions, fig.height=5}
tmb_defs03 <- tmb_defs_present(dat)          # normal + ancestry corrected
tmb_labs03 <- tmb_def_labels(tmb_defs03)

if (nrow(dat) > 0 && length(tmb_defs03) >= 1) {
  tlong <- purrr::map_dfr(seq_along(tmb_defs03), function(i) {
    dat |> transmute(
      pid, aneuploidy_value, aneuploidy_class,
      def   = tmb_labs03[i],
      group = ifelse(suppressWarnings(as.numeric(.data[[tmb_defs03[[i]]]])) >= attend_thresholds$tmb,
                     "TMB-high", "TMB-low"))
  }) |>
    filter(!is.na(aneuploidy_value), !is.na(group)) |>
    mutate(def = factor(def, levels = tmb_labs03))
  tlong <- tlong |> group_by(def) |> filter(!all(is.na(group))) |> ungroup() |>
    mutate(def = droplevels(def))

  if (nrow(tlong) > 0) {
    print(knitr::kable(count(tlong, def, group, name = "n") |>
                         pivot_wider(names_from = group, values_from = n, values_fill = 0),
                       caption = "Patients per TMB group, by definition (MMR-deficient cohort)"))

    ggplot(tlong, aes(group, aneuploidy_value)) +
      geom_boxplot(outlier.shape = NA, fill = "#A6CEE3", width = 0.6) +
      highlight_points(tlong, "group", "aneuploidy_value", base_colour = "aneuploidy_class",
                       base_palette = attend_aneu_cols, base_name = "aneuploidy",
                       base_size = 0.7, base_alpha = 0.6) +
      facet_wrap(~ def, nrow = 1) +
      stat_compare_means(label = "p.format", size = 3) +
      labs(title = "Aneuploidy score by TMB class, per TMB definition (MMR-deficient only)",
           subtitle = paste0("TMB-high cutoff = ", attend_thresholds$tmb, " mut/Mb"),
           x = NULL, y = "aneuploidy score") +
      attend_theme() +
      theme(axis.text.x = element_text(angle = 20, hjust = 1))
  }
}
```

## What to take away

- Every result here is **conditional on MMR deficiency**. It says nothing about
  MMR-proficient tumours, and it is not comparable to a whole-cohort analysis.
- Genes are tested **individually**. A per-gene panel is interpretable; the union at the
  end is not, on its own, because it cannot say which gene carried the signal.
- **CTNNB1 and APC are not in the MMR union.** They are a separate pathway and are
  reported separately throughout.
- Check the counts table before the p-value. Several genes are mutated in very few MMRd
  patients, and those panels are unstable.
- P-values are uncorrected across panels.

To change the panel membership, the pathogenicity definition, or the cutoffs, edit
`attend_gene_panel` / `attend_gene_panel_wnt`, `attend_pathogenic_classes`, and
`attend_thresholds` in `code/attend_classes.R`. If a real pathogenicity annotation
(ClinVar / OncoKB) becomes available, point `pathogenic_by_patient()` at that column
instead of relying on `Variant_Classification`.

```{r session-info}
sessionInfo()
```
