---
title: "ATTEND — 04. Aneuploidy in MMR-deficient tumours"
author: "sceriff0"
date: "2026-08-13"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
# renv: tidyverse, here, ggpubr.
library(tidyverse)
library(here)
library(ggpubr)
# Headless-safe raster device. The cluster's compute nodes have no X11, so R's default
# png() fails with "unable to open connection to X11 display" and every figure dies. cairo
# needs no display; the guard keeps this working on a machine built without it.
if (capabilities("cairo")) options(bitmapType = "cairo")

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 10, fig.height = 6)

source(here("code", "build_master.R"))       # get_master() — builds on demand, no build order
source(here("code", "attend_harmonise.R"))   # build_barcode_pid(), make_id_cfg()
source(here("code", "attend_plots.R"))       # highlight_points(), attend_theme()
source(here("code", "attend_ihc.R"))         # ihc_celltype_metrics(), ihc_content_metrics()
source(here("code", "load_phenotypes.R"))    # load_ihc_celltypes()

# No local re-factor: add_molecular_classes() now returns aneuploidy_class as an ordered
# factor, AS Low first, matching the palette's quiet -> altered direction and reports 06/11.
# This line used to force the opposite order, which is why this report's boxes read
# right-to-left against every other page on the site.
master <- get_master() |> add_molecular_classes()

# The same patients once per aneuploidy cut, with aneuploidy_class re-derived at each.
# Every CUT-DEPENDENT figure below — every figure whose x axis, fill or model term is the
# aneuploidy class — facets on as_cut_lab rather than being written twice. The figures that
# plot the raw score are NOT faceted: their boxes, their n and their p-values are identical
# at either cut, so they carry the two cuts as dashed lines instead. See 00-methods.
master_cuts <- add_aneuploidy_cuts(master, scna = TRUE)
cuts        <- aneu_cuts()

# mmr_group does not depend on the aneuploidy cut, so the MMRd restriction is still one
# vector of pids — applied to the LONG frames below with the same filter(pid %in% mmrd_pids).
mmrd_pids <- add_scna_group(master) |> filter(mmr_group == "MMRd") |> pull(pid)
n_mmrd <- length(mmrd_pids)

# The polipo highlight (21S188) is drawn here, as it is on every per-patient figure in the
# pipeline -- there is no per-report opt-out any more. Its ids are resolved through the
# barcode -> pid crosswalk by highlight_group_of(), so the overlay lands on this report's
# pid-keyed frames; before that resolution existed it matched nothing here and drew nothing.
```

> **Question.** Within MMR-deficient tumours, what does aneuploidy track — mutation status, or cell-type composition — and does either answer move when the aneuploidy cut does?
> **Cohort.** MMR-deficient only (n = `r n_mmrd`), every aneuploidy-class figure drawn at both cuts (`r paste(cuts, collapse = " and ")`), with one marked exception: the interaction model, which cannot be estimated inside a single stratum.
> **Reads.** The master (built on demand by get_master()), the long MAF, and the FlowPath annotation tree.
> **Out of scope.** Anything on the full cohort. Aneuploidy against TMB across both MMR strata -> [06-tmb](06-tmb.html). The same composition panels keyed on treatment response -> [05-response](05-response.html). Why there are two cuts and why both are reported -> [00-methods](00-methods.html).

The restriction is the point. MMR deficiency is itself a route to genomic instability, so
in the whole cohort any variable that tracks MMR status will also appear to track
aneuploidy. Conditioning on MMRd removes that shortcut. The outcome in every figure is the
same continuous aneuploidy score; what changes is the grouping on the x-axis — mutation
status in Part 1, cell-type composition in Part 2. The cohort definition, the two aneuploidy
cuts and the highlight colours are in [00-methods](00-methods.html).

The two cuts, `r paste(cuts, collapse = " and ")`, split this report in two ways. Part 1
puts the **score** on the y axis and mutation status on the x axis, so its panels are the
same at either cut and both cuts are drawn as dashed lines across them. Part 2 puts the
**class** on the x axis, so every one of its figures is drawn at both cuts — the two cuts
are the columns of each figure, and each column carries its own test. Raising the cut can
only move a patient from the high box to the low box, so the columns are nested rather than
independent: only patients scoring between the cuts move.

## Part 1 — aneuploidy by mutation status

The restriction is the point. MMR deficiency is itself a route to genomic instability, so
in the whole cohort any mutation that tracks MMR status will also appear to track
aneuploidy. Conditioning on MMRd removes that shortcut. The outcome variable in every
figure is the same continuous aneuploidy score; what changes is the grouping on the x-axis.

#### The gene panel, spelled out

Genes are tested one at a time, each in its own panel, so a signal is attributable to a
specific gene rather than to a union that could be driven by whichever gene is most
frequently mutated. Two groups, kept deliberately separate:

| Group | Genes | Why they are grouped this way |
|---|---|---|
| **MMR panel** | `r paste(attend_gene_panel, collapse = ", ")` | Loss of any of these causes mismatch-repair deficiency. MSH3 pairs with MSH2 to form MutSb. |
| **Wnt / b-catenin** | `r paste(attend_gene_panel_wnt, collapse = ", ")` | An independent oncogenic pathway, not an MMR mechanism. |
| **TP53** | TP53 | Handled separately: it is already a first-class column on the master. |

The MMR union (panel-altered = at least one pathogenic variant in any MMR panel gene) is
shown last, after the per-gene panels. CTNNB1 and APC are deliberately excluded from
that union, because folding them in would change what "panel-altered" means everywhere else
in the pipeline. There is no Wnt union: with two genes the per-gene panels already show
everything a union would.

P-values come from stat_compare_means(), Wilcoxon rank-sum for two groups and
Kruskal-Wallis for more, uncorrected across panels. With one test per gene, treat them as
descriptive.

### Data

Two sources, joined on pid. The aneuploidy score and its derived classes come from
attend_master_joined. The mutation status is recomputed here directly from the long MAF
rather than taken from the master's collapsed columns, because the average of a variant row
is meaningless; pathogenic_by_patient() and mutation_status_long() are the same helpers
[01-data-integration](01-data-integration.html) used, so the two cannot drift. Crossing the
crosswalk against the full gene list guarantees every gene gets a column even when it never
appears in the MAF, with unmutated patients FALSE rather than NA.

```{r data}
cw  <- build_barcode_pid(load_gianlu_clinical_data(), make_id_cfg())
maf <- read_intermediate("maf_data")

mut <- pathogenic_by_patient(maf, cw)   # pid, TP53_pathogenic, panel_pathogenic (MMR union)

# Per-gene status for the one-gene-at-a-time panels: MMR panel + Wnt genes.
#
# ⚠️ NO replace_na(mutated, FALSE) HERE. It used to sit on the line below, and it turned
# every patient the MAF never reached — and every gene that was never statused — into
# "<GENE>-wt". Combined with the loader statusing only TP53, that made all five MMR panels
# and both Wnt panels a single wild-type box, captioned as though the comparison had been
# made. An uncalled gene is NOT a wild-type gene; the class stays NA so the panel drops the
# patient, and gene_called below reports how many were dropped and why.
gene_long <- cw |> distinct(pid) |>
  tidyr::crossing(gene = attend_genes_per_gene) |>
  left_join(mutation_status_long(maf, cw), by = c("pid", "gene")) |>
  mutate(class = if_else(mutated, paste0(gene, "-mut"), paste0(gene, "-wt")))  # NA stays NA

gene_status <- gene_long |>
  select(pid, gene, class) |>
  pivot_wider(names_from = gene, values_from = class, names_glue = "{gene}_class")

# The denominator behind every per-gene panel below: how many patients this gene could
# actually be called in. A gene with 0 called is reported as a skip, never drawn.
gene_called <- gene_long |>
  group_by(gene) |>
  summarise(called    = sum(!is.na(class)),
            uncalled  = sum(is.na(class)),
            mutated_n = sum(class == paste0(gene, "-mut"), na.rm = TRUE),
            .groups = "drop")

dat_all <- master |>
  select(-any_of(c("TP53_pathogenic", "panel_pathogenic"))) |>  # recompute here
  left_join(mut, by = "pid") |>
  left_join(gene_status, by = "pid") |>
  mutate(
    aneuploidy_value = suppressWarnings(as.numeric(.data[[attend_cols$aneuploidy]])),
    TP53_class  = if_else(TP53_pathogenic,  "TP53-mut",     "TP53-wt"),
    panel_class = if_else(panel_pathogenic, "MMR-panel-altered", "MMR-panel-wt")
  )
```
#### Restricting to MMR-deficient

The single filter the whole report rests on. The counts below are the cohort every
figure that follows is drawn from — if the MMRd group is small, every p-value in this
report inherits that.

```{r restrict-mmrd}
have_mmr <- have_cols(dat_all, "MMR_class", "aneuploidy_value")

dat <- if (have_mmr) dat_all |> filter(MMR_class == attend_levels$mmr_deficient) else dat_all[0, ]

cat("Whole cohort              :", nrow(dat_all), "patients\n")
cat("MMR-deficient (this report):", nrow(dat), "patients\n")
if (nrow(dat) > 0)
  cat("  with an aneuploidy score :", sum(!is.na(dat$aneuploidy_value)), "patients\n")

if (have_mmr && nrow(dat) == 0)
  note_skip("No MMR-deficient patients (MMR_class == '", attend_levels$mmr_deficient,
          "') — every figure below is skipped.")
```

Coverage of the manual highlight overlay on this report's cohort. `attend_highlight` holds
one group, `polipo` (sample 21S188, blue), and 21S188 is a sequencing barcode while `dat` is
keyed on `pid`; `highlight_coverage()` scores the match after resolving the id through the
barcode -> pid crosswalk, so `1/1 matched` means the point really is drawn below and not that
the string happened to appear somewhere. `0/1` means the patient is outside the MMRd
restriction, which is a fact about this cohort, not a broken overlay.

```{r highlight-coverage}
if (nrow(dat) > 0) cat(highlight_coverage(dat))
```

```{r box-helper, include=FALSE}
# Shared boxplot: aneuploidy score on y, one facet per grouping variable. Boxes are
# neutral; points are coloured by aneuploidy class, with the highlight overlay drawn
# on top (partitioned, so no patient is plotted twice). highlight_points() owns both
# colour scales, so no scale_colour_manual is added here.
aneu_box <- function(d, vars, title, subtitle = NULL, ncol = 3) {
  long <- d |>
    select(pid, aneuploidy_value, aneuploidy_class, any_of(vars)) |>
    pivot_longer(any_of(vars), names_to = "comparison", values_to = "group") |>
    filter(!is.na(aneuploidy_value), !is.na(group)) |>
    mutate(comparison = factor(comparison, levels = vars))
  if (nrow(long) == 0) return(NULL)

  ggplot(long, aes(group, aneuploidy_value)) +
    # Was fill = #A6CEE3, which attend_plots.R defines as "aneuploidy-low" — so every box in
    # a figure ABOUT aneuploidy was tinted with one of its two class colours. Neutral now, and
    # the class colour survives where it belongs: on the points, via base_colour below.
    attend_box(long, x = "group", y = "aneuploidy_value", by = "comparison", points = FALSE) +
    highlight_points(long, "group", "aneuploidy_value", base_colour = "aneuploidy_class",
                     base_palette = attend_aneu_cols, base_name = attend_as_legend,
                     base_size = 0.9, base_alpha = 0.65) +
    facet_wrap(~ comparison, scales = "free_x", ncol = ncol) +
    # Raw 0-1 scale, Wilcoxon. The arcsine transform stays in the COMPOSITION panels of Part 2,
    # where the fractions span orders of magnitude; the aneuploidy score is the outcome here
    # and is read directly as a fraction of altered arms.
    stat_compare_means(label = "p.format", size = 3) +
    # NOT faceted by aneuploidy cut, deliberately. The score is the OUTCOME here and the
    # class is not the grouping, so every box, every n and every p-value is identical at
    # either cut — a cut facet would print the same figure twice. Both cuts are drawn as
    # dashed lines instead; the point colour is the PRIMARY cut, so a salmon point below the
    # upper line is a patient the higher cut demotes.
    geom_hline(yintercept = cuts, linetype = 2, colour = "grey40") +
    labs(title = title, subtitle = subtitle, x = NULL, y = attend_as_axis) +
    attend_theme() +
    theme(axis.text.x = element_text(angle = 20, hjust = 1))
}

# Group sizes behind each panel — a two-group Wilcoxon on 3 vs 40 patients looks
# identical to one on 20 vs 23 unless the counts are printed.
group_counts <- function(d, vars) {
  d |>
    select(pid, aneuploidy_value, any_of(vars)) |>
    pivot_longer(any_of(vars), names_to = "comparison", values_to = "group") |>
    filter(!is.na(group), !is.na(aneuploidy_value)) |>
    count(comparison, group, name = "n") |>
    pivot_wider(names_from = group, values_from = n, values_fill = 0)
}
```

### 1. Aneuploidy by gene — MMR panel, one gene at a time

Each panel is one MMR gene: patients carrying a pathogenic variant in that gene
(<GENE>-mut) versus those who do not (<GENE>-wt), within the MMRd cohort only.
Pathogenicity is decided by attend_pathogenic_classes on the MAF's
Variant_Classification.

Read the group sizes first. Some of these genes are mutated in very few patients, and a
panel with a handful of -mut patients will show a wide box and an unstable p-value.

```{r mmr-gene-counts}
mmr_gene_vars <- paste0(attend_gene_panel, "_class")
if (nrow(dat) > 0) {
  # Called vs uncalled FIRST, because it is the denominator the group counts sit inside.
  # A gene with 0 called produces no panel at all, and that has to be visible as an absent
  # measurement rather than inferred from a missing figure.
  print(knitr::kable(
    gene_called |> dplyr::filter(gene %in% attend_gene_panel),
    caption = paste("Patients in whom each MMR panel gene could be called at all.",
                    "`uncalled` = no MAF for that patient, or the gene was never statused —",
                    "these are DROPPED from the panels below, not counted as wild-type")))
  print(knitr::kable(group_counts(dat, mmr_gene_vars),
               caption = "Patients per group, MMR panel genes (MMR-deficient cohort)"))
}
```

Aneuploidy score (aneu__aneuploidy_score, ASCETS arm-level burden from data/ascets/)
in mutated versus wild-type patients, one panel per MMR gene, over the dat cohort above.
The score is the outcome and the class is not the grouping, so this figure is the same at
either cut and is drawn once: the two dashed lines are
`r paste(cuts, collapse = " and ")`, and moving the cut moves only the point colours, never
a box or a p-value.

```{r mmr-gene-box, fig.height=7}
if (nrow(dat) > 0)
  aneu_box(dat, mmr_gene_vars,
           "Aneuploidy score by MMR panel gene (MMR-deficient only)",
           paste0("One panel per gene; pathogenic vs wild-type. n = ", nrow(dat),
                  " MMR-deficient patients"))
```

### 2. Aneuploidy by gene — Wnt / β-catenin pathway

The same comparison for `r paste(attend_gene_panel_wnt, collapse = " and ")`. These are
shown apart from the MMR panel because they are a different mechanism: they are not a
cause of mismatch-repair deficiency, so a signal here would mean something biologically
different from a signal in the MMR panel above.

```{r wnt-gene-counts}
wnt_gene_vars <- paste0(attend_gene_panel_wnt, "_class")
if (nrow(dat) > 0)
  knitr::kable(group_counts(dat, wnt_gene_vars),
               caption = "Patients per group, Wnt pathway genes (MMR-deficient cohort)")
```

The same aneuploidy score against Wnt-pathway mutation status, two panels. Cut-independent
for the same reason as the MMR panels above — the dashed lines carry both cuts.

```{r wnt-gene-box, fig.height=4.5}
if (nrow(dat) > 0)
  aneu_box(dat, wnt_gene_vars,
           "Aneuploidy score by Wnt pathway gene (MMR-deficient only)",
           "CTNNB1 and APC — shown separately from the MMR panel; not part of the MMR union",
           ncol = 2)
```

### 3. TP53 and the MMR union

TP53 on its own, then the MMR panel **union** — at least one pathogenic variant in any
of `r paste(attend_gene_panel, collapse = ", ")`. The union is the last mutation view
rather than the first: it is the least specific, and the per-gene panels above are what
say which gene any union signal came from.

```{r tp53-union-counts}
summary_vars <- c("TP53_class", "panel_class")
if (nrow(dat) > 0)
  knitr::kable(group_counts(dat, summary_vars),
               caption = "Patients per group, TP53 and the MMR union (MMR-deficient cohort)")
```

Aneuploidy score by TP53 status and by the MMR union, side by side, so a union signal can
be read against the per-gene panels above that would have to explain it. Both dashed lines
are the two cuts; neither box nor test depends on which one is chosen.

```{r tp53-union-box, fig.height=4.5}
if (nrow(dat) > 0)
  aneu_box(dat, summary_vars,
           "Aneuploidy score by TP53 status and by the MMR panel union (MMR-deficient only)",
           ncol = 2)
```

### 4. Aneuploidy by the other molecular classes

The non-mutation axes, still within MMRd: TMB, HRD, and MSI status. MMR_class is
omitted because it is constant here by construction — every patient in this report is
MMR-deficient.

MSI is worth reading carefully: MMR loss is the usual *cause* of microsatellite
instability, so within an MMRd cohort the MSI-stable group is both small and unusual.
None of these three classes is derived from the aneuploidy score, so the figure is again
one figure with the two cuts drawn as dashed lines across it.

```{r other-classes-box, fig.height=4.5}
other_vars <- c("TMB_class", "HRD_class", "MSI_class")
if (nrow(dat) > 0) {
  knitr::kable(group_counts(dat, other_vars),
               caption = "Patients per group, score-derived classes (MMR-deficient cohort)")
  aneu_box(dat, other_vars,
           "Aneuploidy score by TMB / HRD / MSI class (MMR-deficient only)",
           ncol = 3)
}
```

## Part 2 — aneuploidy by cell-type composition

#### Cell types and normalization

"Cell type" is phenotype_clean, the parenthetical phenotype label in the FlowPath export
(changeable in code/load_phenotypes.R). Every count is taken inside the annotated region
and expressed as a fraction, but the denominator changes the question, so all three are
shown:

| Normalization | Denominator | The question it answers |
|---|---|---|
| **all cells inside** | every cell in the annotation | what fraction of the tissue is this cell type? |
| **tumour cells inside** | tumour cells only | how much of this cell type per unit of tumour? |
| **CD45+ cells inside** | leukocytes only | what is the composition *of the immune compartment*? |

Each is drawn twice: the raw fraction with a Wilcoxon test, and the
arcsine-square-root-transformed fraction with a two-sample t-test. Agreement between the
two is reassurance; disagreement means the result depends on the test rather than the data.
No multiple-testing correction is applied anywhere in this report, and with one test per
cell type per normalization that is many comparisons, so every p-value here is descriptive.

### Building the restricted cohort

Three things are joined on pid: the aneuploidy class from the master, the per-patient
cell-type composition read from the FlowPath tree by ihc_celltype_metrics(), and the
continuous covariates. MMR status comes from add_scna_group(), so it is the same
is_mmrd() call the copy-number and classification reports use. The class is joined from the
LONG frame, so every patient appears once per cut carrying the class that cut gives them —
the table below prints how many patients each cut calls high, and the difference between
its two rows is exactly the patients the higher cut demotes.

```{r composition-data}
master_grp <- add_scna_group(master)        # adds mmr_group ("MMRd"/"MMRp") + scna_group

# One row per patient PER CUT, carrying the class that cut derives. Every cut-dependent
# figure in Part 2 joins against this rather than re-deriving a class locally, which is how
# aneuploidy_class came to read in two directions on one site (see add_molecular_classes()).
cuts_pt <- master_cuts |> select(pid, as_cut, as_cut_lab, aneuploidy_class)

# many-to-many is DECLARED, not silenced by accident: one metric row per patient meets two
# cut rows per patient, and doubling the frame is exactly the intended shape.
with_cuts <- function(d)
  d |>
    left_join(cuts_pt, by = "pid", relationship = "many-to-many") |>
    filter(!is.na(aneuploidy_class))

# per-patient cell-type composition inside the annotation (reads the FlowPath tree)
ct_base <- ihc_celltype_metrics(load_ihc_celltypes(), load_imaging_data())
ct_all  <- with_cuts(ct_base)
ct      <- ct_all |> filter(pid %in% mmrd_pids)          # <- the MMRd restriction
# The SINGLE-ROW MMRd frame. ihc_series_metrics() builds the content panels from raw counts
# and unique()s its constants, so it must see each patient once — handing it the long frame
# would double every celltype- and immune-sourced series.
ct_pt   <- ct_base |> filter(pid %in% mmrd_pids)

# The leukocyte metrics are built HERE, beside ct, not in the immune section further down:
# the tissue-content panels above that section need imm_metrics, and a chunk that reads a
# variable a LATER chunk assigns knits fine in an interactive session (where it is already
# in the global env) and dies on a clean build with "object not found".
imm_base    <- ihc_immune_metrics(load_ihc_celltypes(), load_imaging_data())
imm_all     <- with_cuts(imm_base)
imm_metrics <- imm_all |> filter(pid %in% mmrd_pids)      # <- the MMRd restriction
imm_pt      <- imm_base |> filter(pid %in% mmrd_pids)

# per-patient continuous covariates + TP53 status, restricted the same way
numcol <- function(col) if (col %in% names(master)) suppressWarnings(as.numeric(master[[col]])) else NA_real_
pt_all <- tibble(
  pid              = master$pid,
  aneuploidy_class = master$aneuploidy_class,
  aneuploidy_score = numcol(attend_cols$aneuploidy),
  HRD              = numcol(attend_cols$hrd),
  MSI              = numcol("msi__MSI_SCORE"),
  TP53             = if ("TP53_pathogenic" %in% names(master))
                       if_else(master$TP53_pathogenic, "TP53-abnormal", "TP53-normal") else NA_character_
)

# The two TMB definitions, added under their reader-facing names so every facet reads
# "TMB (normal)" / "TMB (ancestry corrected)" rather than an internal column name.
tmb_defs04 <- tmb_defs_present(master)
tmb_labs04 <- tmb_def_labels(tmb_defs04)
for (i in seq_along(tmb_defs04)) pt_all[[tmb_labs04[i]]] <- numcol(tmb_defs04[[i]])
covar_cols04 <- c(tmb_labs04, "HRD", "MSI")

pt <- pt_all |> filter(pid %in% mmrd_pids)          # <- the MMRd restriction

# The same covariates once per cut, for the binarised panel further down. The scatter and
# the TP53 panel plot the raw score and stay on the single-row frame above, or every patient
# would be drawn twice at the same coordinates.
pt_cuts <- pt_all |> select(-aneuploidy_class) |> with_cuts() |> filter(pid %in% mmrd_pids)

cat("MMR-deficient patients                       :", length(mmrd_pids), "\n")
cat("  ... with cell-type metrics + aneuploidy class:", n_distinct(ct$pid), "\n")
cat("  ... cell types measured                      :", n_distinct(ct$cell_type), "\n")
cat("TMB definitions plotted:", paste(tmb_labs04, collapse = " | "), "\n")
cat("aneuploidy cuts drawn side by side:", paste(cuts, collapse = " and "), "\n\n")
if (nrow(ct) > 0) {
  cls_by_cut <- ct |> distinct(pid, as_cut_lab, aneuploidy_class)
  print(table(cut        = droplevels(cls_by_cut$as_cut_lab),
              aneuploidy = droplevels(factor(cls_by_cut$aneuploidy_class))))
}
```

```{r box-helpers, include=FALSE}
# The builder is composition_box() in attend_plots.R, shared with report 05 so the two
# composition figure sets cannot drift apart again (they had: 9 x 20 inches here, 14 x 17
# there, two copies of the code). This wrapper fixes what is 04's own: the grouping is the
# aneuploidy class, and THE TWO CUTS ARE THE COLUMNS, the cell type the row. That way round
# because `free_y` gives each ROW its own axis: with the cuts on the rows a cell type would be
# drawn against two different y ranges and the two halves of the comparison could not be read
# against each other.
box_by_aneu <- function(df, denom_label, arcsin = FALSE, scope = NULL, ...)
  composition_box(df, "aneuploidy_class", attend_aneu_cols, attend_as_legend,
                  by_label = "aneuploidy — MMRd only, both AS cuts", denom_label = denom_label,
                  arcsin = arcsin, scope = scope, cols = "as_cut_lab", width_in = 9, ...)

# Row counts for the chunk canvases, known BEFORE the figures exist: one row per cell type.
# A fixed fig.height over a variable row count is what squashed the overview.
n_focus   <- sum(attend_ihc_focus %in% ct$cell_type)
n_overall <- n_distinct(ct$cell_type[!grepl(attend_ihc_tumor_pattern, ct$cell_type)])
comp_h    <- function(n) composition_height(composition_rows(n, cols = "as_cut_lab"))

# ONE normalization, one of its two figures, raw/Wilcoxon then arcsine/t-test. The transform
# happens here rather than in the builder because the axis has to be computed on the same
# values the panels plot.
draw_composition <- function(frac_col, denom_label, which = c("focus", "all")) {
  which <- match.arg(which)
  for (asin in c(FALSE, TRUE)) {
    d <- ct |>
      mutate(.v = if (asin) arcsin_sqrt(.data[[frac_col]]) else .data[[frac_col]]) |>
      filter(is.finite(.v))
    # The focus set sits under ~0.6 on the arcsin scale, so it gets an axis of its own. The
    # overview drops the tumour phenotypes: at ~1.0 one of them flattens every immune panel,
    # and tumour content has its own panel below.
    d <- if (which == "focus") filter(d, cell_type %in% attend_ihc_focus)
         else filter(d, !grepl(attend_ihc_tumor_pattern, cell_type))
    p <- if (nrow(d) > 0)
           box_by_aneu(d, denom_label, asin,
                       scope = if (which == "focus") "focus set" else "all cell types",
                       free_y = which == "all" && attend_ihc_all_free_y)
    if (is.null(p)) {
      note_skip("Skipped: ", denom_label, " / ", which, " — no finite fractions",
                if (which == "focus") paste0(", or none of attend_ihc_focus (",
                                             paste(attend_ihc_focus, collapse = ", "),
                                             ") found in cell_type") else "", ".")
      return(invisible(NULL))
    }
    print(p)
  }
}
```

### Cell-type composition by aneuploidy class

One row per cell type and **one column per aneuploidy cut**, aneuploidy-high vs -low within
each, over MMRd only, from the ct table built above. Each normalization is shown as a
raw-fraction/Wilcoxon figure followed by its arcsine/t-test companion, and every test is
computed inside its own panel, so the left and right columns are separate tests and are
never pooled. Moving the cut moves patients from the high box to the low box and nothing
else: the fractions on the y axis are the same numbers in both columns, so a contrast that
survives the move is one the threshold is not making.

Each normalization is drawn **twice**: first the cell types in `attend_ihc_focus`
(`T cytotoxic`, `T helper`, `Macrophages`, `M1`) on an axis of their own, then every cell
type on one axis. The focus figure exists because `Tumor` reaches ~1.0 on the arcsine scale
while all four of these sit under ~0.6, so an axis wide enough for `Tumor` leaves them
unreadable — and these four are what the figure is for.

Every cell type is drawn in **one figure per normalization, on a single shared y axis**.
A panel's height therefore reads as that cell type's abundance relative to every other one,
which is the comparison the figure exists to support: `Macrophages` is read against
`T cytotoxic` on the same scale, not against a rescaled axis of its own. `scales = "free_y"`
would give each panel its own range and make a cell type at a fraction of a percent of the
tissue draw the same box, in the same place, as one at half the tissue. The cost is accepted
deliberately — on the tumour-cell denominator these cell types span orders of magnitude, so
the rarest panels sit low against a scale set by the most abundant.

The two figures differ in one respect on purpose: the **focus set shares one y axis**, so
those four cell types are directly comparable, while the **overview gives each cell-type row
its own** — a row, not a panel, so the two cut columns of one cell type always stay on one
axis and the only difference between them is which patients sit in which box. Over every
measured cell type a single abundant one sets a ceiling the rest never
approach, and panels that are boxes in the focus figure flatten to lines. Abundance is still
readable from each overview panel's own axis, and the content panels below carry it
directly. The policy is `attend_ihc_all_free_y`.

The **tumour phenotypes are excluded** from that figure. The export carries them separately
(`PanCK+ Tumor`, `VIM+ Tumor`), and on one shared axis a tumour phenotype near 1.0 on the
arcsine scale compresses every immune panel under 0.35 into a flat line. Tumour content is
reported on its own panel below, against the same denominator, so nothing is lost.

The axis stops at the **largest upper whisker** in the figure (Q3 + 1.5 IQR over every class
x cell-type cell), not at the largest value, and the p-value label is placed inside that
range by the same call. Both follow from one measurement: ggpubr derives its default label
height from the plot-wide maximum, so a single extreme patient simultaneously flattened
every box and parked every p-value at the ceiling, where it read as no p-value at all. The
range is a **zoom** (`coord_cartesian()`), so every patient is still in the box statistics
and in the test — nothing is dropped, and a point above the axis is counted in the subtitle.
Groups smaller than `attend_min_box_n` still show a median crossbar instead of a box.

Each cell type as a fraction of every cell in the annotation, so a panel reads "how much
of the tissue is this". The focus set first, raw fractions with a
Wilcoxon and then the arcsine/t-test companion, one row per cell type.

```{r box-inside, fig.width=9, fig.height=comp_h(n_focus), out.width="100%"}
draw_composition("frac_inside", "all cells inside", "focus")
```

Every non-tumour cell type against the same denominator, each row on its own axis,
raw then arcsine.

```{r box-inside-all, fig.width=9, fig.height=comp_h(n_overall), out.width="100%"}
draw_composition("frac_inside", "all cells inside", "all")
```

#### Normalized by tumour cells inside

Each cell type per unit of tumour: the denominator is tumour cells only, so a shift here
survives a change in how much tumour the annotation caught. The focus set first, raw fractions with a
Wilcoxon and then the arcsine/t-test companion, one row per cell type.

```{r box-tumor, fig.width=9, fig.height=comp_h(n_focus), out.width="100%"}
draw_composition("frac_tumor", "tumour cells inside", "focus")
```

Every non-tumour cell type against the same denominator, each row on its own axis,
raw then arcsine.

```{r box-tumor-all, fig.width=9, fig.height=comp_h(n_overall), out.width="100%"}
draw_composition("frac_tumor", "tumour cells inside", "all")
```

#### Normalized by CD45+ cells inside

Composition *of the immune compartment*: the denominator is leukocytes, so tumour-content
differences between patients cannot drive the contrast. The focus set first, raw fractions with a
Wilcoxon and then the arcsine/t-test companion, one row per cell type.

```{r box-cd45, fig.width=9, fig.height=comp_h(n_focus), out.width="100%"}
draw_composition("frac_cd45", "CD45+ cells inside", "focus")
```

Every non-tumour cell type against the same denominator, each row on its own axis,
raw then arcsine.

```{r box-cd45-all, fig.width=9, fig.height=comp_h(n_overall), out.width="100%"}
draw_composition("frac_cd45", "CD45+ cells inside", "all")
```


#### Tissue content: tumour, leukocyte and PD-L1+, on one axis

How much of the annotation is tumour, how much is leukocyte, and how much is PD-L1-positive,
as fractions of **the same denominator** so the three are directly comparable. All three are
per-image constants pooled per patient by `ihc_celltype_metrics()` — `n_tumor_inside`,
`n_cd45_inside` and `n_pdl1_inside`, each a count of cells inside the annotation in the raw
FlowPath per-cell export, over `n_inside`. They are read as constants rather than as
cell-type rows because the export splits tumour across several phenotypes (`PanCK+ Tumor`,
`VIM+ Tumor`, ...) and carries no `PD-L1` phenotype at all, so there is no single row to
select. **`all cells inside` is the only denominator on which all three are meaningful** —
`Tumor` over tumour cells and CD45+ over CD45+ cells are both identically 1 — so this panel
is drawn once, not once per normalization. Raw fractions with a Wilcoxon, then the
arcsine/t-test companion, each with the two aneuploidy cuts as its columns and its own
Wilcoxon per column — raising the cut cannot change any of these three fractions, only which
box a patient's fraction is counted in.

⚠️ **The PD-L1 column has a smaller cohort than the other two.** PD-L1 was stained in only
some batches, so a patient from an unstained batch is NA for PD-L1 and present in tumour and
CD45+. Those patients drop out of the third box only, which means the three boxes on this one
axis are not all drawn from the same patients. **Read the n printed under each box before
comparing their heights.** PD-L1+ per unit of tumour, the other denominator, is in the
immune-content section below.

```{r box-content, fig.width=9, fig.height=comp_h(length(attend_ihc_content)), out.width="100%"}
# Built from the SINGLE-ROW frames: ihc_series_metrics() reads raw counts, and a patient
# arriving twice (once per cut) would be counted twice in every series. The cuts are joined
# on afterwards, which is what makes the two columns of the figure.
ct_content <- ihc_series_metrics(ct_pt, imm_pt, attend_ihc_content) |> with_cuts()

for (asin in c(FALSE, TRUE)) {
  d <- ct_content |> mutate(.v = if (asin) arcsin_sqrt(frac) else frac)
  p <- box_by_aneu(d, "all cells inside", asin, scope = "tissue content",
                   unit_override = "")
  if (!is.null(p)) print(p)
}
```

#### Tumour burden beside immune infiltration per unit of tumour

Three populations against **different** denominators: `Tumor` over all cells inside (how
much of the tissue is tumour) beside CD45+ and PD-L1+ over tumour cells inside (how many
leukocytes, and how many PD-L1-positive cells, per tumour cell). Each panel strip names its
own denominator, because the y axis carries one unit and cannot. The two right-hand columns
share a denominator, so they **are** directly comparable to each other; neither is
comparable to the tumour-content column on the left.

Counts come from the raw FlowPath per-cell export as per-image constants pooled per patient
(`n_tumor_inside`, `n_cd45_inside`, `n_pdl1_inside` over `n_inside` / `n_tumor_inside`).
As in the panel above, PD-L1 was stained in only some batches, so the PD-L1 column is drawn
from fewer patients than the other two — read the n under each box.

CD45+ / tumour cells and PD-L1+ / tumour cells are **ratios, not proportions** — they exceed
1 wherever those cells outnumber tumour cells — and `arcsin_sqrt()` clamps its input to
[0, 1]. Both versions are
drawn, and the count of values above 1 is printed above the figures **and** in the
transformed figure's own subtitle: those patients all land on the same value there, so a
clamped group is a wall of identical numbers and its t-test is not interpretable. Read the
raw/Wilcoxon panel whenever that count is non-zero; when it is zero the transform is exact
and the two panels are the usual pair. The two aneuploidy cuts are again the columns, and
the clamp count is a property of the ratios, so it is identical in both.

```{r box-content-mixed, fig.width=9, fig.height=comp_h(length(attend_ihc_content_mixed)), out.width="100%"}
# Same single-row input as the panel above, and the clamp count is taken BEFORE the cut join
# so it counts patients, not patient-cut rows — the two cuts do not change a ratio.
mix_pt <- ihc_series_metrics(ct_pt, imm_pt, attend_ihc_content_mixed)
mix    <- mix_pt |> with_cuts() |> mutate(.v = frac)

n_over <- sum(mix_pt$frac > 1, na.rm = TRUE)
cat("patients:", n_distinct(mix$pid), "| series:",
    paste(gsub("\n", " ", levels(mix$cell_type)), collapse = " | "),
    "\nvalues above 1 (a ratio, not a proportion):", n_over,
    if (n_over > 0) " <- clamped by arcsin_sqrt() in the transformed figure below" else "", "\n")

# Raw, then the arcsine/t-test companion. The companion carries a CLAMP COUNT because
# arcsin_sqrt() bounds its input to [0, 1] and BOTH per-tumour columns are ratios that can exceed it:
# every value above 1 lands on pi/2, so a clamped group is a wall of identical numbers and
# its t-test is not interpretable. When the count is 0 the transform is exact and the note
# does not appear, which is the case this figure is drawn for.
for (asin in c(FALSE, TRUE)) {
  note <- if (asin && n_over > 0)
            sprintf("; %d value%s > 1 CLAMPED to pi/2 by arcsin(sqrt()) — read the raw panel",
                    n_over, if (n_over == 1L) "" else "s") else ""
  d <- mix |> mutate(.v = if (asin) arcsin_sqrt(frac) else frac)
  p <- box_by_aneu(d, "the denominator in each panel", arcsin = asin,
                   scope = "tissue content, mixed denominators",
                   unit_override = "", extra_note = note)
  if (!is.null(p)) print(p)
}
```

### Immune content — the headline contrast

The question this report exists for: among MMR-deficient tumours, is aneuploidy-high
associated with an immune-cold phenotype?

Two leukocyte metrics counted inside the annotation — **CD45+ total** (all leukocytes)
and **CD3+CD45+** (double-positive T cells) — under each normalization. CD3+CD45+
requires a CD3_sign column in the FlowPath export; without it those rows are NA and
their facets drop out on their own.

```{r immune-data}
cat("MMR-deficient patients with immune metrics:", n_distinct(imm_metrics$pid),
    "| metrics present:", paste(sort(unique(imm_metrics$metric)), collapse = ", "), "\n")
```

```{r immune-box-helper, include=FALSE}
# One figure PER DENOMINATOR, faceted by metric. The five immune rows carry three different
# denominators, and on one shared axis CD3+CD45+ over CD45+ sits an order of magnitude above
# the same numerator over all cells inside — so the panels that are not being compared set
# the scale for the ones that are. Splitting on denom_label means every panel in a figure is
# a fraction OF THE SAME THING, which is the only case where a shared axis is honest.
#
# The builder is composition_box(), the one every composition figure uses: metrics become its
# rows (renamed onto `cell_type`) and the two aneuploidy cuts its columns, so a free y axis is
# free per ROW and the two cuts of one metric keep one scale.
box_immune <- function(df, denom_, arcsin = FALSE) {
  d <- df |>
    filter(is.finite(frac), denom_label == denom_) |>
    transmute(pid, cell_type = metric, aneuploidy_class, as_cut_lab,
              .v = if (arcsin) arcsin_sqrt(frac) else frac)
  if (nrow(d) == 0) return(NULL)
  composition_box(d, "aneuploidy_class", attend_aneu_cols, attend_as_legend,
                  by_label = "aneuploidy — MMRd only, both AS cuts", denom_label = denom_,
                  arcsin = arcsin, scope = "immune content", cols = "as_cut_lab",
                  width_in = 10)
}

# Rows per figure = the most metrics any one denominator carries.
n_imm_rows <- if (nrow(imm_metrics) > 0)
                max(table(distinct(imm_metrics, metric, denom_label)$denom_label)) else 1L
```

CD45+ total and CD3+CD45+ fractions from imm_metrics, by aneuploidy class, one row per
metric and one column per aneuploidy cut, within MMRd. One figure per denominator rather
than one figure with all three: a shared axis is only honest between panels that are
fractions of the same thing, and CD3+CD45+ over CD45+ sits an order of magnitude above the
same numerator over all cells inside. The denominator is named in each subtitle. Each column
carries its own test on the same fractions, so a contrast present in only one of them is one
the threshold is producing.

```{r immune-box, fig.width=10, fig.height=comp_h(n_imm_rows), out.width="100%"}
if (nrow(filter(imm_metrics, is.finite(frac))) > 2) {
  # Denominator order is the order ihc_immune_metrics() declares its specs in, so the
  # figures come out widest-denominator first however the table happened to be sorted.
  for (dn in unique(imm_metrics$denom_label))
    for (asin in c(FALSE, TRUE)) {
      p <- box_immune(imm_metrics, dn, asin)
      if (!is.null(p)) print(p)
    }
} else note_skip("No finite immune metrics in the MMRd cohort (CD3/CD45 counts missing) — skipping.")
```

#### The pre-specified primary contrast

Of everything above, one comparison is designated **primary**: **CD45+ / all cells
inside**, tested with the **arcsine-square-root transform and a two-sample t-test**. The
Wilcoxon on the raw fraction is reported next to it as a companion, not a substitute.
The other two normalizations, and CD3+CD45+, are **secondary** — they can and do degrade
to a skip when counts are missing.

Naming one primary endpoint in advance is what keeps this from being a search across six
panels for the smallest p-value. The endpoint is named once and tested **once per cut**: two
cuts is two tests of one pre-specified contrast, not two chances at it, and a primary that
holds at one cut and not the other is reported as exactly that.

```{r immune-mmrd-test}
mmrd_arcsin_t_wilcox <- function(df, metric_, denom_) {
  base <- df |> dplyr::filter(metric == metric_, denom_label == denom_, is.finite(frac))
  rows <- list()
  # One test per cut, never pooled: the two blocks hold the same patients under two
  # classifications, so a test across both would count every patient twice.
  for (lab in levels(droplevels(base$as_cut_lab))) {
    d   <- base |> dplyr::filter(as_cut_lab == lab)
    tab <- table(d$aneuploidy_class)   # a factor, so an empty arm is a 0, not a dropped name
    if (length(tab) < 2 || any(tab < 2)) {
      note_skip("MMRd ", metric_, " / ", denom_, " at ", lab,
                ": insufficient n per aneuploidy arm — no test at this cut.")
      next
    }
    t_res <- t.test(arcsin_sqrt(frac) ~ aneuploidy_class, data = d)     # primary
    w_res <- wilcox.test(frac ~ aneuploidy_class, data = d)             # companion
    cat(sprintf(paste0("MMRd %-10s / %-20s %-11s n=%d (high=%d, low=%d)  ",
                       "arcsin-t p=%.3g  Wilcoxon p=%.3g\n"),
        metric_, denom_, lab, nrow(d), tab[["aneuploidy-high"]], tab[["aneuploidy-low"]],
        t_res$p.value, w_res$p.value))
    rows[[lab]] <- data.frame(as_cut_lab = lab, n = nrow(d),
                              n_high = tab[["aneuploidy-high"]],
                              n_low  = tab[["aneuploidy-low"]],
                              t_p = t_res$p.value, w_p = w_res$p.value,
                              stringsAsFactors = FALSE)
  }
  if (!length(rows)) return(invisible(NULL))
  list(d = base, stats = do.call(rbind, rows))
}
res_primary <- mmrd_arcsin_t_wilcox(imm_metrics, "CD45+", "all cells inside")      # PRIMARY
res_tumor   <- mmrd_arcsin_t_wilcox(imm_metrics, "CD45+", "tumour cells inside")   # secondary
res_cd45n   <- mmrd_arcsin_t_wilcox(imm_metrics, "CD3+CD45+", "CD45+ cells inside") # secondary
```

The primary contrast on its own, one panel per cut. The subtitle carries each cut's two
p-values and its group size — at this n a single patient moving between groups can change
the result, and moving the cut is precisely what moves patients between these two groups,
so both counts are on the figure rather than in a caption.

```{r immune-mmrd-box, fig.width=8, fig.height=5}
if (!is.null(res_primary)) {
  d  <- res_primary$d
  st <- res_primary$stats
  ggplot(d, aes(aneuploidy_class, frac, fill = aneuploidy_class)) +
    attend_box(d, x = "aneuploidy_class", y = "frac", by = "as_cut_lab",
               fill = NULL, points = FALSE) +
    highlight_points(d, "aneuploidy_class", "frac", base_size = 1.2, base_alpha = 0.5, base_const = "black") +
    # Per panel, so the Wilcoxon on the figure is the one from that cut's own test above.
    stat_compare_means(method = "wilcox.test", label = "p.format", size = 3) +
    facet_wrap(~ as_cut_lab) +
    scale_x_discrete(labels = as_label) +
    scale_fill_manual(values = attend_aneu_cols, labels = as_label,
                      name = attend_as_legend) +
    labs(title = "PRIMARY: CD45+ immune fraction (all cells inside) — MMR-deficient only",
         subtitle = paste(sprintf("%s: arcsin(sqrt) t p = %s | Wilcoxon p = %s (n = %d)",
                                  st$as_cut_lab, signif(st$t_p, 3), signif(st$w_p, 3), st$n),
                          collapse = "   "),
         x = NULL, y = "CD45+ fraction of all cells inside") +
    attend_theme() + theme(legend.position = "none")
} else note_skip("Primary within-MMRd immune contrast skipped at every cut — see above.")
```

#### The one model that needs MMR-proficient patients

The single exception to this report's MMRd restriction, and it has to be: an interaction
asks whether the aneuploidy effect *differs between* MMRd and MMRp, which cannot be
estimated inside one stratum. The model is
arcsin(sqrt(CD45+/all cells inside)) ~ aneuploidy_class * mmr_group, fitted on both
groups from imm_all. The interaction row is the one of interest: a large coefficient
there is what justifies restricting everything else to MMRd. The model has aneuploidy_class
as a term, so it is **fitted once per cut** and the table gains a cut column — a single fit
over the stacked frame would use every patient twice under two contradictory labels. No
correction is applied, and nothing is pooled across the two fits.

```{r immune-interaction}
imm_int <- imm_all |>
  dplyr::filter(metric == "CD45+", denom_label == "all cells inside", is.finite(frac)) |>
  dplyr::left_join(master_grp |> dplyr::select(pid, mmr_group), by = "pid") |>
  dplyr::filter(!is.na(mmr_group), !is.na(aneuploidy_class)) |>
  dplyr::mutate(y = arcsin_sqrt(frac))

fit_one_cut <- function(d) {
  if (!(nrow(d) > 4 && dplyr::n_distinct(d$aneuploidy_class) == 2 &&
        dplyr::n_distinct(d$mmr_group) == 2)) {
    note_skip("AS cut ", d$as_cut[1], ": not enough patients across both MMR groups and ",
              "both aneuploidy classes for the interaction model.")
    return(NULL)
  }
  fit <- lm(y ~ aneuploidy_class * mmr_group, data = d)
  tb  <- if (requireNamespace("broom", quietly = TRUE)) broom::tidy(fit) else {
    m <- as.data.frame(summary(fit)$coefficients); m$term <- rownames(m); m
  }
  cbind(`AS cut` = as.character(d$as_cut_lab[1]), tb)
}

coef_tbl <- if (nrow(imm_int) > 0)
  do.call(rbind, lapply(split(imm_int, droplevels(imm_int$as_cut_lab)), fit_one_cut)) else NULL

if (!is.null(coef_tbl)) {
  print(knitr::kable(coef_tbl, digits = 3, row.names = FALSE,
        caption = paste("arcsin(sqrt(CD45+/all-cells-inside)) ~ aneuploidy_class * mmr_group",
                        "— BOTH MMR groups, fitted separately at each aneuploidy cut,",
                        "no correction")))
} else note_skip("The interaction model could not be fitted at either aneuploidy cut.")
```

### Aneuploidy against the continuous covariates

Aneuploidy score against TMB (both definitions), HRD, and MSI, within MMRd. One
point per patient, coloured by aneuploidy class, with Spearman's rho per panel. Spearman
rather than Pearson because these scores are bounded and skewed.

The two TMB panels are the same patients under the two definitions, so a visible
difference between them is the germline correction moving values, not new data. This figure
is **not** faceted by cut: x is the raw score and rho is computed on it, so both cuts give
the identical panel. They appear as the two dashed vertical lines instead, which is what
lets a reader judge whether either falls anywhere the points suggest a break.

```{r aneu-scatter, fig.width=12, fig.height=7}
scat <- pt |>
  pivot_longer(all_of(covar_cols04), names_to = "covariate", values_to = "value") |>
  filter(!is.na(aneuploidy_score), !is.na(value)) |>
  mutate(covariate = factor(covariate, levels = covar_cols04))

if (nrow(scat) > 2) {
  rho <- scat |>
    group_by(covariate) |>
    summarise(label = paste0("rho = ",
                round(suppressWarnings(cor(aneuploidy_score, value, method = "spearman",
                                           use = "complete.obs")), 2)), .groups = "drop")
  ggplot(scat, aes(aneuploidy_score, value, colour = aneuploidy_class)) +
    # Both cuts, on a figure that is otherwise cut-independent: the points, the fit and rho
    # are the same at either, and the point colour is the primary cut.
    geom_vline(xintercept = cuts, linetype = 2, colour = "grey40") +
    highlight_points(scat, "aneuploidy_score", "value", base_colour = "aneuploidy_class",
                     base_palette = attend_aneu_cols,
                     base_name = attend_as_legend, jitter_width = 0, base_size = 1.5, base_alpha = 0.6) +
    geom_smooth(method = "lm", se = FALSE, colour = "grey40", linewidth = 0.6) +
    geom_text(data = rho, aes(x = Inf, y = Inf, label = label),
              inherit.aes = FALSE, hjust = 1.05, vjust = 1.5, size = 3) +
    facet_wrap(~ covariate, scales = "free_y", ncol = 3) +
    labs(title = "Aneuploidy score vs continuous covariates (MMR-deficient only)",
         subtitle = paste("Dashed lines: the two AS cuts,", paste(cuts, collapse = " and ")),
         x = "aneuploidy score", y = "covariate value", colour = NULL) +
    attend_theme() + theme(legend.position = "top")
} else note_skip("Not enough MMR-deficient patients with aneuploidy score + a covariate.")
```

The same covariates against the **binarised** aneuploidy class rather than the continuous
score — the categorical counterpart to the scatter above, with a Wilcoxon per covariate and
per cut. Rows are the covariates, columns the two cuts (`r paste(cuts, collapse = " and ")`),
drawn from pt_cuts so each patient carries the class its column's cut gives them. Comparing
the two columns is what the scatter above cannot show: a correlation visible there but absent
in both columns is being discarded by binarising at all, while one that appears in a single
column is being produced by that cutpoint.

```{r aneu-class-box, fig.width=9, fig.height=10}
cb <- pt_cuts |>
  pivot_longer(all_of(covar_cols04), names_to = "covariate", values_to = "value") |>
  filter(!is.na(aneuploidy_class), !is.na(value)) |>
  mutate(covariate = factor(covariate, levels = covar_cols04))

if (nrow(cb) > 2) {
  ggplot(cb, aes(aneuploidy_class, value, fill = aneuploidy_class)) +
    attend_box(cb, x = "aneuploidy_class", y = "value", by = c("covariate", "as_cut_lab"),
               fill = NULL, points = FALSE) +
    highlight_points(cb, "aneuploidy_class", "value", base_size = 0.6, base_alpha = 0.4,
                     base_const = "black") +
    stat_compare_means(method = "wilcox.test", label = "p.format", size = 3) +
    # Cut on the COLUMNS so "free_y" stays free per covariate: a covariate must keep one
    # axis across the two cuts, or the columns cannot be read against each other.
    facet_grid(covariate ~ as_cut_lab, scales = "free_y") +
    scale_x_discrete(labels = as_label) +
    scale_fill_manual(values = attend_aneu_cols, labels = as_label,
                      name = attend_as_legend) +
    labs(title = "Continuous covariates by aneuploidy class (MMR-deficient only, both AS cuts)",
         x = NULL, y = "covariate value") +
    attend_theme() + theme(legend.position = "none")
} else note_skip("Not enough MMR-deficient patients with aneuploidy class + a covariate.")
```

### Aneuploidy by TP53 status

Within MMRd, aneuploidy score in TP53-abnormal versus TP53-normal tumours. TP53 status is
TP53_pathogenic from the master, derived from the MAF in report 01. TP53 loss is the
canonical driver of chromosomal instability, so this is the expected-positive control for
the aneuploidy score itself: if aneuploidy does not separate on TP53 here, the score is
suspect before any of the immune results are interpreted. The control is on the raw score
and TP53 status does not depend on the cut, so this is one figure with both cuts drawn
across it as dashed lines — where they fall relative to the two boxes is how much of this
separation either cut can capture.

```{r mmrd-tp53, fig.width=5, fig.height=5}
tp <- pt |> filter(!is.na(TP53), !is.na(aneuploidy_score))

if (nrow(tp) > 2 && dplyr::n_distinct(tp$TP53) == 2) {
  ggplot(tp, aes(TP53, aneuploidy_score, fill = TP53)) +
    attend_box(tp, x = "TP53", y = "aneuploidy_score", fill = NULL, points = FALSE) +
    highlight_points(tp, "TP53", "aneuploidy_score", base_size = 1.5, base_alpha = 0.5,
                     base_const = "black") +
    stat_compare_means(method = "wilcox.test", label = "p.format") +
    geom_hline(yintercept = cuts, linetype = 2, colour = "grey40") +
    scale_x_discrete(labels = as_label) +
    scale_fill_manual(values = attend_tp53_cols, labels = as_label,
                      name = attend_tp53_legend) +
    labs(title = "Aneuploidy by TP53 status (MMR-deficient only)",
         subtitle = paste("Dashed lines: the two AS cuts,", paste(cuts, collapse = " and ")),
         x = NULL, y = attend_as_axis) +
    attend_theme() + theme(legend.position = "none")
} else {
  note_skip("Not enough MMR-deficient patients with both TP53 classes, or TP53_pathogenic ",
          "is missing from the master (rebuild report 01).")
}
```

### Aneuploidy against the pathologist's imaging calls

The pathologist's per-patient scores compared between aneuploidy high and low, within
MMRd. These come from the imaging sheet keyed by image → pid, averaged over a patient's
images for the continuous measures and taking the first non-missing value for the
categorical calls. Both figures below are keyed on the aneuploidy class, so the frame is
built once per cut here and each patient appears twice — once under each classification.

```{r imaging-covars}
first_nonNA <- function(x) { x <- x[!is.na(x)]; if (length(x)) x[1] else NA }

imaging_pt <- load_imaging_data() |>
  filter(!is.na(ID)) |>
  group_by(pid = as.character(ID)) |>
  summarise(
    Immune_infiltrate      = mean(suppressWarnings(as.numeric(Immune_infiltrate)),     na.rm = TRUE),
    Immune_infiltrate_PD1  = mean(suppressWarnings(as.numeric(Immune_infiltrate_PD1)), na.rm = TRUE),
    ARID1A                 = first_nonNA(ARID1A),
    Tumor_cells_PDL1       = first_nonNA(Tumor_cells_PDL1),
    Immune_infiltrate_PDL1 = first_nonNA(Immune_infiltrate_PDL1),
    .groups = "drop"
  ) |>
  filter(pid %in% mmrd_pids) |>   # <- the MMRd restriction
  with_cuts()                     # <- one row per patient per aneuploidy cut

cat("MMR-deficient patients with imaging covariates:", n_distinct(imaging_pt$pid), "\n")
cat("  ... drawn once per aneuploidy cut:", paste(cuts, collapse = " and "), "\n")
```

Continuous pathologist scores (immune infiltrate and PD-1 infiltrate, as percentages) by
aneuploidy class, one row per score and one column per cut, with its own Wilcoxon in each
panel. These are the pathologist's independent read of the same immune question the FlowPath
contrast above asked; the pathologist's numbers do not move with the cut, so the only thing
that differs between the two columns is which patients are in which box.

```{r imaging-covars-cont, fig.width=8, fig.height=7}
icc <- imaging_pt |>
  pivot_longer(c(Immune_infiltrate, Immune_infiltrate_PD1),
               names_to = "covariate", values_to = "value") |>
  filter(is.finite(value))

if (nrow(icc) > 2) {
  ggplot(icc, aes(aneuploidy_class, value, fill = aneuploidy_class)) +
    attend_box(icc, x = "aneuploidy_class", y = "value", by = c("covariate", "as_cut_lab"),
               fill = NULL, points = FALSE) +
    highlight_points(icc, "aneuploidy_class", "value", base_size = 0.6, base_alpha = 0.4,
                     base_const = "black") +
    stat_compare_means(method = "wilcox.test", label = "p.format", size = 3) +
    # Cut on the columns, so "free_y" stays free per pathologist score rather than per cut.
    facet_grid(covariate ~ as_cut_lab, scales = "free_y") +
    scale_x_discrete(labels = as_label) +
    scale_fill_manual(values = attend_aneu_cols, labels = as_label,
                      name = attend_as_legend) +
    labs(title = "Pathologist immune scores by aneuploidy class (MMR-deficient only, both AS cuts)",
         x = NULL, y = "pathologist %") +
    attend_theme() + theme(legend.position = "none")
}
```

Categorical calls as stacked proportions — how each call distributes within aneuploidy
high vs low, one row per call and one column per cut. Bars are proportions, so the **group
sizes are not visible in this figure**; read them from the count printed above, and read
them per column, because moving the cut is exactly what changes how many patients each bar
is drawn from. MMR_Status is omitted because it is constant here by construction.

```{r imaging-covars-cat, fig.width=9, fig.height=8}
icat <- imaging_pt |>
  select(as_cut_lab, aneuploidy_class, ARID1A, Tumor_cells_PDL1, Immune_infiltrate_PDL1) |>
  pivot_longer(-c(as_cut_lab, aneuploidy_class), names_to = "covariate", values_to = "level") |>
  filter(!is.na(level), level != "")

if (nrow(icat) > 0) {
  ggplot(icat, aes(aneuploidy_class, fill = level)) +
    geom_bar(position = "fill") +
    # The pathologist's category sets differ per covariate and have no semantic palette of
    # their own, so they take the project's qualitative one POSITIONALLY. Without it these
    # bars used ggplot's default hue wheel. A stacked bar carries no per-patient highlight
    # point, so the reserved blue at attend_pal()[1] cannot hide one here (rule [C], the
    # same exemption the KM and ECDF palettes take).
    scale_fill_manual(values = attend_pal(dplyr::n_distinct(icat$level))) +
    facet_grid(covariate ~ as_cut_lab) +
    labs(title = "Pathologist categorical calls by aneuploidy class (MMR-deficient only, both AS cuts)",
         x = NULL, y = "proportion of patients", fill = NULL) +
    attend_theme() +
    theme(axis.text.x = element_text(angle = 20, hjust = 1))
}
```

```{r session-info}
sessionInfo()
```
