---
title: "ATTEND — 4. IHC, imaging, and cell-type composition in MMR-deficient tumours"
author: "sceriff0"
date: "2026-08-09"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
# renv: needs tidyverse, here, survival, survminer (KM via km_facet()), ggpubr (boxplot
# p-values), gtsummary (tbl_summary).
#   renv::install(c("survival","survminer","ggpubr","gtsummary")); renv::snapshot()
#   # richer spatial stack, if/when point-pattern statistics are added (research/ §5):
#   renv::install(c("bioc::SpatialExperiment","bioc::imcRtools","spatstat.geom","sf","dbscan"))
library(tidyverse)
library(here)
library(survival)
library(survminer)
library(ggpubr)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 6)

source(here("code", "load_phenotypes.R"))    # load_ihc_data(), load_ihc_celltypes()
source(here("code", "load_clinical.R"))      # load_imaging_data(), load_gianlu_clinical_data()
source(here("code", "attend_harmonise.R"))   # crosswalks + pid_vector
source(here("code", "attend_classes.R"))     # add_molecular_classes(), km_facet()
source(here("code", "attend_plots.R"))       # highlight_points() + attend_highlight overlay
source(here("code", "attend_io.R"))          # read_intermediate() (Parquet/CSV)
source(here("code", "attend_ihc.R"))         # FlowPath IHC metric definitions
```

## What this report does

Two halves, with different cohorts, and it matters which is which.

**Part 1 — does the automated quantification agree with the pathologist?**
FlowPath counts individual cells on each IHC slide and turns them into positivity and
composition fractions. A pathologist independently scored the same slides by eye. Part 1
compares the two, on **all patients with IHC data**. This is a measurement-validation
question, so restricting it to a subgroup would only make it noisier.

**Part 2 — does cell-type composition track aneuploidy?**
Using the same FlowPath counts, Part 2 asks whether the mix of cells inside the annotated
tumour region differs between aneuploidy-high and aneuploidy-low tumours. **This half is
restricted to MMR-deficient (MMRd) tumours.** MMR deficiency drives both immune
infiltration and genomic instability, so in a mixed cohort any aneuploidy–immune
association could simply be MMR status showing through. Conditioning on MMRd removes that
confound, at the cost of a much smaller cohort — the MMRd aneuploidy-high group is
roughly **n ≈ 9**, so every p-value in Part 2 is suggestive, not confirmatory.

### What the highlighted points mean

Most scatter and box plots carry two overlays drawn on top of the ordinary points:

| Overlay | What it marks |
|---|---|
| **Gold** | Patients who **have IHC / imaging data** — the multi-modal subset. On a genomics figure this shows where the patients in *this* report sit within the wider cohort. |
| **Magenta** | Sample **21S188** (*polipo*), always drawn in its own colour so it never blends into the gold group or the plot's own encoding. |

The groups are defined once in `attend_highlight` (`code/attend_classes.R`) and matched
by `pid` **or** sequencing barcode, so the same patients are highlighted identically in
every report. `polipo` is listed before `cohort` so 21S188 can never be absorbed into the
gold group.

### Cell types and normalization

"Cell type" is `phenotype_clean`, the parenthetical phenotype label in the FlowPath
export (changeable in `code/load_phenotypes.R`). Every count is taken **inside the
annotated region** and expressed as a fraction — but the denominator changes the
question, so all three are shown:

| Normalization | Denominator | The question it answers |
|---|---|---|
| **all cells inside** | every cell in the annotation | what fraction of the tissue is this cell type? |
| **tumour cells inside** | tumour cells only | how much of this cell type per unit of tumour? |
| **CD45+ cells inside** | leukocytes only | what is the composition *of the immune compartment*? |

Each is drawn twice: the **raw fraction with a Wilcoxon** test, and the
**arcsine-square-root transformed** fraction with a **two-sample t-test**. The arcsine
transform is the standard variance-stabiliser for proportions — fractions near 0 or 1
have compressed variance, which the t-test otherwise assumes away. The two are companions,
not a choice: agreement between them is reassurance, disagreement is a signal that the
result depends on the test rather than the data.

**No multiple-testing correction is applied anywhere in this report.** With one test per
cell type per normalization that is a lot of comparisons, so read them as descriptive.

---

# Part 1 — IHC and imaging: does FlowPath agree with the pathologist?

## Loading and mapping to patients

The FlowPath export is keyed by `image_id`, the imaging annotation sheet carries both
`image_id` and the patient id, and the master is keyed by `pid`. Everything is put onto
`pid` through the image crosswalk built in `code/attend_harmonise.R` — the same one report
1 used, so this report's cohort cannot drift from the master's.

The two counts below are the ceiling for everything in Part 1: rows that fail to map are
invisible to every figure that follows.

```{r load}
ihc_data     <- load_ihc_data()        # FlowPath per-image cell counts, keyed by image_id
imaging_data <- load_imaging_data()    # pathologist annotations, keyed by image_id

id_cfg         <- make_id_cfg()
cw_image_pid   <- build_image_pid(imaging_data, id_cfg)   # image -> pid
cw_barcode_pid <- build_barcode_pid(load_gianlu_clinical_data(), id_cfg)
pid_vector     <- make_pid_vector(cw_barcode_pid, cw_image_pid)

ihc_pid     <- ihc_data     |> mutate(pid = pid_vector(ihc_data,     id_cfg$ihc))
imaging_pid <- imaging_data |> mutate(pid = pid_vector(imaging_data, id_cfg$imaging))

cat("IHC rows mapped to a patient    :", sum(!is.na(ihc_pid$pid)),     "/", nrow(ihc_pid), "\n")
cat("imaging rows mapped to a patient:", sum(!is.na(imaging_pid$pid)), "/", nrow(imaging_pid), "\n")
```

Survival and MMR columns are attached from the master so Part 1 can end with a
survival figure without re-deriving anything.

```{r survival-join}
master <- read_intermediate("attend_master_joined")

surv_cols <- intersect(c(attend_cols$surv_time, attend_cols$surv_event,
                         attend_cols$km_group, attend_cols$mmr_status),
                       names(master))

clinical_min <- master |> select(pid, all_of(surv_cols))

ihc_clin     <- ihc_pid     |> filter(!is.na(pid)) |> left_join(clinical_min, by = "pid")
imaging_clin <- imaging_pid |> filter(!is.na(pid)) |> left_join(clinical_min, by = "pid")
```

## The imaged cohort

Per-image FlowPath counts joined to the pathologist annotations, with the
positivity/composition ratios derived by `ihc_image_metrics()` (`code/attend_ihc.R`). The
summary table describes the cohort the IHC slides came from — not the whole ATTEND
cohort, which is the relevant denominator for everything in Part 1.

```{r ihc-metrics}
ihc_img <- ihc_image_metrics(ihc_data, imaging_data)
cat("IHC images joined to imaging annotations:", sum(!is.na(ihc_img$ID)), "/", nrow(ihc_img), "\n")

if (requireNamespace("gtsummary", quietly = TRUE)) {
  ihc_img |>
    select(any_of(c("MSI_Status", "treatment", "ARID1A", "MMR_Status"))) |>
    gtsummary::tbl_summary()
}
```

## Concordance — continuous measures

For these three, the pathologist also gave a **number**, so the two methods can be
plotted against each other directly. Each point is one image. The dashed red line is
`x = y`, perfect agreement: points systematically above it mean the pathologist scored
higher than FlowPath, below it the reverse. Spearman's rho measures whether the two
**rank** images the same way, which is the weaker but more robust claim — a high rho with
points far off the diagonal means consistent but miscalibrated.

Pathologist percentages are divided by 100 so both axes are fractions.

### Neoplastic cellularity

FlowPath tumour fraction = all tumour cells / all cells, against the pathologist's
neoplastic-cellularity percentage.

```{r concordance-cellularity}
cc <- ihc_img |>
  transmute(ID,                                          # ID == pid; kept so the highlight can match
            flowpath    = tumor_over_all,
            pathologist = suppressWarnings(as.numeric(Neoplastic_cellularity)) / 100) |>
  filter(!is.na(flowpath), !is.na(pathologist))

if (nrow(cc) > 2) {
  rho <- cor(cc$flowpath, cc$pathologist, method = "spearman", use = "complete.obs")
  ggplot(cc, aes(flowpath, pathologist)) +
    highlight_points(cc, "flowpath", "pathologist", jitter_width = 0, base_size = 1.5, base_alpha = 0.6) +
    geom_abline(slope = 1, linetype = "dashed", colour = "red") +
    coord_equal() +
    annotate("text", x = -Inf, y = Inf, hjust = -0.2, vjust = 2,
             label = paste0("Spearman = ", round(rho, 2))) +
    labs(title = "Neoplastic cellularity", x = "FlowPath tumour fraction",
         y = "Pathologist cellularity") +
    theme_classic(base_size = 12)
}
```

### Immune infiltrate

FlowPath immune content = CD45+ cells / all cells inside the annotation, against the
pathologist's immune-infiltrate percentage.

```{r concordance-immune}
ii <- ihc_img |>
  transmute(ID,
            flowpath    = cd45_over_inside,
            pathologist = suppressWarnings(as.numeric(Immune_infiltrate)) / 100) |>
  filter(!is.na(flowpath), !is.na(pathologist))

if (nrow(ii) > 2) {
  rho <- cor(ii$flowpath, ii$pathologist, method = "spearman", use = "complete.obs")
  ggplot(ii, aes(flowpath, pathologist)) +
    highlight_points(ii, "flowpath", "pathologist", jitter_width = 0, base_size = 1.5, base_alpha = 0.6) +
    geom_abline(slope = 1, linetype = "dashed", colour = "red") +
    coord_equal() +
    annotate("text", x = -Inf, y = Inf, hjust = -0.2, vjust = 2,
             label = paste0("Spearman = ", round(rho, 2))) +
    labs(title = "Immune infiltrate", x = "FlowPath CD45 fraction (inside)",
         y = "Pathologist immune infiltrate") +
    theme_classic(base_size = 12)
}
```

### PD-1 on immune cells

FlowPath PD-1+ immune = PD-1+ CD45 cells / CD45 cells inside the annotation. A linear
trend line replaces the `x = y` line here, because the pathologist's PD-1 score is not on
the same scale as the FlowPath fraction — only the ranking is comparable.

```{r concordance-pd1}
pd1 <- ihc_img |>
  transmute(ID,
            flowpath    = cd45_pd1_over_cd45_inside,
            pathologist = suppressWarnings(as.numeric(Immune_infiltrate_PD1)) / 100) |>
  filter(!is.na(flowpath), !is.na(pathologist))

if (nrow(pd1) > 2) {
  rho <- cor(pd1$flowpath, pd1$pathologist, method = "spearman", use = "complete.obs")
  ggplot(pd1, aes(flowpath, pathologist)) +
    highlight_points(pd1, "flowpath", "pathologist", jitter_width = 0, base_size = 1.5, base_alpha = 0.6) +
    geom_smooth(method = "lm", se = FALSE, colour = "grey40", linewidth = 0.6) +
    annotate("text", x = -Inf, y = Inf, hjust = -0.2, vjust = 2,
             label = paste0("Spearman = ", round(rho, 2))) +
    labs(title = "PD-1 on immune cells", x = "FlowPath PD-1+ / CD45 (inside)",
         y = "Pathologist PD-1 infiltrate") +
    theme_classic(base_size = 12)
}
```

## Concordance — categorical calls

For these three the pathologist gave a **category**, not a number, so the comparison is
a boxplot: the FlowPath fraction distribution within each pathologist call. Concordance
looks like boxes that step monotonically across the categories. Overlapping boxes mean
the automated fraction does not reproduce the call.

### ARID1A

FlowPath ARID1A+ = ARID1A+ tumour cells / tumour cells inside the annotation. The
pathologist's categories run from complete loss through clonal loss to positive, so the
FlowPath fraction should **increase** left to right.

```{r concordance-aridia}
if ("ARID1A" %in% names(ihc_img)) {
  d <- ihc_img |>
    mutate(ARID1A = factor(ARID1A, levels = c("Complete loss", "Clonal loss", "Positive"))) |>
    filter(!is.na(ARID1A), !is.na(tumor_aridia_over_tumor_inside))
  ggplot(d, aes(ARID1A, tumor_aridia_over_tumor_inside, colour = ARID1A)) +
      geom_boxplot(outlier.shape = NA) +
      highlight_points(d, "ARID1A", "tumor_aridia_over_tumor_inside", base_colour = "ARID1A",
                       base_name = "Pathologist", jitter_width = 0.2, base_size = 1.5, base_alpha = 0.6) +
      labs(title = "ARID1A on tumour cells", x = NULL, y = "FlowPath positivity (fraction)") +
      theme_classic(base_size = 12)
}
```

### PD-L1 on immune cells

FlowPath PD-L1+ immune = PD-L1+ CD45 cells / CD45 cells inside. The pathologist's IC
categories are explicit percentage bands (IC0 < 1%, IC1 1–5%, IC2 5–10%), so this is the
strictest of the categorical checks — the FlowPath fraction should land inside the named
band, not merely increase across bands.

```{r concordance-pdl1}
if ("Immune_infiltrate_PDL1" %in% names(ihc_img)) {
  d <- ihc_img |>
    mutate(Immune_infiltrate_PDL1 = factor(Immune_infiltrate_PDL1,
             levels = c("", "IC0 (<1%)", "IC1 (>=1% and <5%)", "IC2 (>=5% and <10%)"))) |>
    filter(!is.na(Immune_infiltrate_PDL1), !is.na(cd45_pdl1_over_cd45_inside))
  ggplot(d, aes(Immune_infiltrate_PDL1, cd45_pdl1_over_cd45_inside, colour = Immune_infiltrate_PDL1)) +
      geom_boxplot(outlier.shape = NA) +
      highlight_points(d, "Immune_infiltrate_PDL1", "cd45_pdl1_over_cd45_inside",
                       base_colour = "Immune_infiltrate_PDL1", base_name = "Pathologist",
                       jitter_width = 0.2, base_size = 1.5, base_alpha = 0.6) +
      labs(title = "PD-L1 on immune cells", x = NULL, y = "FlowPath positivity (fraction)") +
      theme_classic(base_size = 12)
}
```

### PD-L1 on tumour cells

FlowPath PD-L1+ tumour = PD-L1+ tumour cells / tumour cells inside the annotation.

```{r concordance-tcpdl1}
if ("Tumor_cells_PDL1" %in% names(ihc_img)) {
  d <- ihc_img |>
    filter(!is.na(Tumor_cells_PDL1), Tumor_cells_PDL1 != "", !is.na(tumor_pdl1_over_tumor_inside))
  ggplot(d, aes(Tumor_cells_PDL1, tumor_pdl1_over_tumor_inside, colour = Tumor_cells_PDL1)) +
      geom_boxplot(outlier.shape = NA) +
      highlight_points(d, "Tumor_cells_PDL1", "tumor_pdl1_over_tumor_inside",
                       base_colour = "Tumor_cells_PDL1", base_name = "Pathologist",
                       jitter_width = 0.2, base_size = 1.5, base_alpha = 0.6) +
      labs(title = "PD-L1 on tumour cells", x = NULL, y = "FlowPath positivity (fraction)") +
      theme_classic(base_size = 12) +
      theme(axis.text.x = element_text(angle = 20, hjust = 1))
}
```

## Per-patient IHC metrics

Everything above is per **image**. A patient can contribute several images, so the
metrics are collapsed to one row per patient (mean across that patient's images) before
being used against patient-level variables like survival or aneuploidy.

```{r ihc-patient}
ihc_pt <- ihc_patient_metrics(ihc_data, imaging_data)
glimpse(ihc_pt)
```

Survival split by immune content, on the **whole imaged cohort**. Immune content is the
CD45 fraction inside the annotation, binarised at its **median** — a data-driven split,
not a pre-registered threshold, so it is descriptive. Panels are MMR status; curves are
immune-high vs immune-low.

```{r survival-by-ihc}
d <- ihc_pt |>
  left_join(master |> select(pid, any_of(c(attend_cols$surv_time, attend_cols$surv_event,
                                           attend_cols$mmr_status))), by = "pid") |>
  mutate(immune_class = if_else(cd45_over_inside >= median(cd45_over_inside, na.rm = TRUE),
                                "immune-high", "immune-low"))

if (have_cols(d, attend_cols$surv_time, attend_cols$surv_event,
              attend_cols$mmr_status, "immune_class")) {
  km_facet(d, group = "immune_class", facets = attend_cols$mmr_status)
} else message("survival columns not set in attend_cols — skipping KM by IHC feature.")
```

---

# Part 2 — Cell-type composition vs aneuploidy, in MMR-deficient tumours

Everything from here on is **MMR-deficient only**, with one clearly-marked exception
(the interaction model, which cannot be estimated inside a single stratum).

## Building the restricted cohort

Three things are joined on `pid`: the aneuploidy class from the master, the per-patient
cell-type composition from the FlowPath tree, and the continuous covariates. MMR status
comes from `add_scna_group()`, which calls `is_mmrd()` — **IHC-based, with MSI
deliberately not a criterion** — so this report's MMRd definition is identical to the one
the copy-number and classification reports use.

The counts printed here are the cohort for the whole of Part 2.

```{r data}
master <- read_intermediate("attend_master_joined") |>
  add_molecular_classes() |>
  mutate(aneuploidy_class = factor(aneuploidy_class,
                                   levels = c("aneuploidy-high", "aneuploidy-low")))

master_grp <- add_scna_group(master)        # adds mmr_group ("MMRd"/"MMRp") + scna_group
mmrd_pids  <- master_grp |> filter(mmr_group == "MMRd") |> pull(pid)

# per-patient cell-type composition inside the annotation (reads the FlowPath tree)
ct_all <- ihc_celltype_metrics(load_ihc_celltypes(), load_imaging_data()) |>
  left_join(master |> select(pid, aneuploidy_class), by = "pid") |>
  filter(!is.na(aneuploidy_class))
ct <- ct_all |> filter(pid %in% mmrd_pids)          # <- the MMRd restriction

# per-patient continuous covariates + TP53 status, restricted the same way
numcol <- function(col) if (col %in% names(master)) suppressWarnings(as.numeric(master[[col]])) else NA_real_
pt_all <- tibble(
  pid              = master$pid,
  aneuploidy_class = master$aneuploidy_class,
  aneuploidy_score = numcol(attend_cols$aneuploidy),
  HRD              = numcol(attend_cols$hrd),
  MSI              = numcol("msi__MSI_SCORE"),
  TP53             = if ("TP53_pathogenic" %in% names(master))
                       if_else(master$TP53_pathogenic, "TP53-abnormal", "TP53-normal") else NA_character_
)

# The two TMB definitions, added under their reader-facing names so every facet reads
# "TMB (normal)" / "TMB (ancestry corrected)" rather than an internal column name.
tmb_defs04 <- tmb_defs_present(master)
tmb_labs04 <- tmb_def_labels(tmb_defs04)
for (i in seq_along(tmb_defs04)) pt_all[[tmb_labs04[i]]] <- numcol(tmb_defs04[[i]])
covar_cols04 <- c(tmb_labs04, "HRD", "MSI")

pt <- pt_all |> filter(pid %in% mmrd_pids)          # <- the MMRd restriction

cat("MMR-deficient patients                       :", length(mmrd_pids), "\n")
cat("  ... with cell-type metrics + aneuploidy class:", n_distinct(ct$pid), "\n")
cat("  ... cell types measured                      :", n_distinct(ct$cell_type), "\n")
cat("TMB definitions plotted:", paste(tmb_labs04, collapse = " | "), "\n\n")
if (nrow(ct) > 0) print(table(aneuploidy = droplevels(factor(
  ct |> distinct(pid, aneuploidy_class) |> pull(aneuploidy_class)))))
```

```{r box-helpers, include=FALSE}
# One boxplot per cell type: aneuploidy-high vs -low. Boxes are filled by aneuploidy
# class; the highlight overlay is drawn on top in its own colour scale (colour != fill,
# so the two legends do not collide), partitioned so no patient is plotted twice.
box_by_aneu <- function(df, frac_col, denom_label) {
  d <- df |> filter(is.finite(.data[[frac_col]]))
  if (nrow(d) == 0) return(NULL)
  ggplot(d, aes(aneuploidy_class, .data[[frac_col]], fill = aneuploidy_class)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.5) +
    highlight_points(d, "aneuploidy_class", frac_col, base_size = 0.5, base_alpha = 0.35,
                     base_const = "black") +
    facet_wrap(~ cell_type, scales = "free_y") +
    stat_compare_means(method = "wilcox.test", label = "p.format", size = 3) +
    scale_fill_manual(values = attend_aneu_cols) +
    labs(title = paste0("Cell-type presence by aneuploidy (MMRd only) — normalized by ", denom_label),
         x = NULL, y = paste0("fraction of ", denom_label)) +
    attend_theme() +
    theme(legend.position = "none", axis.text.x = element_text(angle = 20, hjust = 1))
}

# Variance-stabilised companion: same panels, arcsine-square-root fraction + t-test.
box_by_aneu_arcsin <- function(df, frac_col, denom_label) {
  d <- df |> filter(is.finite(.data[[frac_col]])) |>
    mutate(.y = arcsin_sqrt(.data[[frac_col]]))
  if (nrow(d) == 0) return(NULL)
  ggplot(d, aes(aneuploidy_class, .y, fill = aneuploidy_class)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.5) +
    highlight_points(d, "aneuploidy_class", ".y", base_size = 0.5, base_alpha = 0.35,
                     base_const = "black") +
    facet_wrap(~ cell_type, scales = "free_y") +
    stat_compare_means(method = "t.test", label = "p.format", size = 3) +
    scale_fill_manual(values = attend_aneu_cols) +
    labs(title = paste0("Cell-type presence by aneuploidy (MMRd only) — arcsin(sqrt), by ", denom_label),
         subtitle = "arcsine-square-root transformed fraction; two-sample t-test",
         x = NULL, y = paste0("arcsin(sqrt(fraction of ", denom_label, "))")) +
    attend_theme() +
    theme(legend.position = "none", axis.text.x = element_text(angle = 20, hjust = 1))
}
```

## Cell-type composition by aneuploidy class

One panel per cell type, aneuploidy-high vs -low, within MMRd. Each normalization is
shown as a raw-fraction/Wilcoxon figure followed by its arcsine/t-test companion — see
the normalization table at the top for what each denominator asks.

### Normalized by all cells inside

```{r box-inside, fig.width=15, fig.height=11, out.width="100%"}
box_by_aneu(ct, "frac_inside", "all cells inside")
```

```{r box-inside-arcsin, fig.width=15, fig.height=11, out.width="100%"}
box_by_aneu_arcsin(ct, "frac_inside", "all cells inside")
```

### Normalized by tumour cells inside

```{r box-tumor, fig.width=15, fig.height=11, out.width="100%"}
box_by_aneu(ct, "frac_tumor", "tumour cells inside")
```

```{r box-tumor-arcsin, fig.width=15, fig.height=11, out.width="100%"}
box_by_aneu_arcsin(ct, "frac_tumor", "tumour cells inside")
```

### Normalized by CD45+ cells inside

```{r box-cd45, fig.width=15, fig.height=11, out.width="100%"}
box_by_aneu(ct, "frac_cd45", "CD45+ cells inside")
```

```{r box-cd45-arcsin, fig.width=15, fig.height=11, out.width="100%"}
box_by_aneu_arcsin(ct, "frac_cd45", "CD45+ cells inside")
```

## Immune content — the headline contrast

The biological question Part 2 exists for: **among MMR-deficient tumours, is
aneuploidy-high associated with an immune-cold phenotype?**

Two leukocyte metrics counted inside the annotation — **CD45+ total** (all leukocytes)
and **CD3+CD45+** (double-positive T cells) — under each normalization. CD3+CD45+
requires a `CD3_sign` column in the FlowPath export; without it those rows are `NA` and
their facets drop out on their own.

```{r immune-data}
imm_all <- ihc_immune_metrics(load_ihc_celltypes(), load_imaging_data()) |>
  left_join(master |> select(pid, aneuploidy_class), by = "pid") |>
  filter(!is.na(aneuploidy_class))
imm_metrics <- imm_all |> filter(pid %in% mmrd_pids)      # <- the MMRd restriction

cat("MMR-deficient patients with immune metrics:", n_distinct(imm_metrics$pid),
    "| metrics present:", paste(sort(unique(imm_metrics$metric)), collapse = ", "), "\n")
```

```{r immune-box-helper, include=FALSE}
box_immune <- function(df, arcsin = FALSE) {
  d <- df |> filter(is.finite(frac))
  if (nrow(d) == 0) return(NULL)
  if (arcsin) d <- d |> mutate(frac = arcsin_sqrt(frac))
  method <- if (arcsin) "t.test"                 else "wilcox.test"
  ylab   <- if (arcsin) "arcsin(sqrt(fraction))" else "fraction"
  ttl    <- if (arcsin) "arcsin(sqrt) + t-test"  else "raw fraction + Wilcoxon"
  ggplot(d, aes(aneuploidy_class, frac, fill = aneuploidy_class)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.5) +
    highlight_points(d, "aneuploidy_class", "frac", base_size = 0.6, base_alpha = 0.4,
                     base_const = "black") +
    facet_wrap(~ metric + denom_label, scales = "free_y") +
    stat_compare_means(method = method, label = "p.format", size = 3) +
    scale_fill_manual(values = attend_aneu_cols) +
    labs(title = paste0("Immune content by aneuploidy (MMRd only) — ", ttl),
         x = NULL, y = ylab) +
    attend_theme() +
    theme(legend.position = "none", axis.text.x = element_text(angle = 20, hjust = 1))
}
```

```{r immune-box, fig.width=12, fig.height=8, out.width="100%"}
if (nrow(filter(imm_metrics, is.finite(frac))) > 2) {
  print(box_immune(imm_metrics, arcsin = FALSE))
  print(box_immune(imm_metrics, arcsin = TRUE))
} else message("No finite immune metrics in the MMRd cohort (CD3/CD45 counts missing) — skipping.")
```

### The pre-specified primary contrast

Of everything above, one comparison is designated **primary**: **CD45+ / all cells
inside**, tested with the **arcsine-square-root transform and a two-sample t-test**. The
Wilcoxon on the raw fraction is reported next to it as a companion, not a substitute.
The other two normalizations, and CD3+CD45+, are **secondary** — they can and do degrade
to a skip when counts are missing.

Naming one primary endpoint in advance is what keeps this from being a search across six
panels for the smallest p-value.

```{r immune-mmrd-test}
mmrd_arcsin_t_wilcox <- function(df, metric_, denom_) {
  d <- df |> dplyr::filter(metric == metric_, denom_label == denom_, is.finite(frac))
  tab <- table(d$aneuploidy_class)
  if (length(tab) < 2 || any(tab < 2)) {
    message("MMRd ", metric_, " / ", denom_, ": insufficient n per aneuploidy arm — skipping.")
    return(invisible(NULL))
  }
  t_res <- t.test(arcsin_sqrt(frac) ~ aneuploidy_class, data = d)     # primary
  w_res <- wilcox.test(frac ~ aneuploidy_class, data = d)             # companion
  cat(sprintf("MMRd %-10s / %-20s  n=%d (high=%d, low=%d)  arcsin-t p=%.3g  Wilcoxon p=%.3g\n",
      metric_, denom_, nrow(d), tab[["aneuploidy-high"]], tab[["aneuploidy-low"]],
      t_res$p.value, w_res$p.value))
  list(d = d, t = t_res, w = w_res)
}
res_primary <- mmrd_arcsin_t_wilcox(imm_metrics, "CD45+", "all cells inside")      # PRIMARY
res_tumor   <- mmrd_arcsin_t_wilcox(imm_metrics, "CD45+", "tumour cells inside")   # secondary
res_cd45n   <- mmrd_arcsin_t_wilcox(imm_metrics, "CD3+CD45+", "CD45+ cells inside") # secondary
```

The primary contrast on its own. The subtitle carries both p-values and the group size —
at this n a single patient moving between groups can change the result, which is why the
count is on the figure rather than in a caption.

```{r immune-mmrd-box, fig.width=6, fig.height=5}
if (!is.null(res_primary)) {
  d <- res_primary$d
  ggplot(d, aes(aneuploidy_class, frac, fill = aneuploidy_class)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.5) +
    highlight_points(d, "aneuploidy_class", "frac", base_size = 1.2, base_alpha = 0.5, base_const = "black") +
    stat_compare_means(method = "wilcox.test", label = "p.format", size = 3) +
    scale_fill_manual(values = attend_aneu_cols) +
    labs(title = "PRIMARY: CD45+ immune fraction (all cells inside) — MMR-deficient only",
         subtitle = paste0("arcsin(sqrt) t p = ", signif(res_primary$t$p.value, 3),
                           " | Wilcoxon p = ", signif(res_primary$w$p.value, 3),
                           " (n = ", nrow(d), "; small-n — suggestive only)"),
         x = NULL, y = "CD45+ fraction of all cells inside") +
    attend_theme() + theme(legend.position = "none")
} else message("Primary within-MMRd immune contrast skipped — see message above.")
```

### The one model that needs MMR-proficient patients

This is the single exception to Part 2's MMRd restriction, and it has to be: an
**interaction** asks whether the aneuploidy effect *differs between* MMRd and MMRp, and
that cannot be estimated inside one stratum. The model is
`arcsin(sqrt(CD45+/all cells inside)) ~ aneuploidy_class * mmr_group`, fitted on both
groups.

The interaction term is the row of interest — a large coefficient there is what would
justify the MMRd-only restriction the rest of Part 2 applies. No correction is applied.

```{r immune-interaction}
imm_int <- imm_all |>
  dplyr::filter(metric == "CD45+", denom_label == "all cells inside", is.finite(frac)) |>
  dplyr::left_join(master_grp |> dplyr::select(pid, mmr_group), by = "pid") |>
  dplyr::filter(!is.na(mmr_group), !is.na(aneuploidy_class)) |>
  dplyr::mutate(y = arcsin_sqrt(frac))

if (nrow(imm_int) > 4 && dplyr::n_distinct(imm_int$aneuploidy_class) == 2 &&
    dplyr::n_distinct(imm_int$mmr_group) == 2) {
  fit <- lm(y ~ aneuploidy_class * mmr_group, data = imm_int)
  coef_tbl <- if (requireNamespace("broom", quietly = TRUE)) broom::tidy(fit) else
    as.data.frame(summary(fit)$coefficients)
  print(knitr::kable(coef_tbl, digits = 3,
        caption = "arcsin(sqrt(CD45+/all-cells-inside)) ~ aneuploidy_class * mmr_group — BOTH MMR groups, no correction"))
} else message("Not enough patients across both MMR groups for the interaction model.")
```

## Aneuploidy against the continuous covariates

Aneuploidy score against TMB (**both definitions**), HRD, and MSI — within MMRd. One
point per patient, coloured by aneuploidy class, with Spearman's rho per panel. Spearman
rather than Pearson because these scores are bounded and skewed.

The two TMB panels are the same patients under the two definitions, so a visible
difference between them is the germline correction moving values, not new data.

```{r aneu-scatter, fig.width=12, fig.height=7}
scat <- pt |>
  pivot_longer(all_of(covar_cols04), names_to = "covariate", values_to = "value") |>
  filter(!is.na(aneuploidy_score), !is.na(value)) |>
  mutate(covariate = factor(covariate, levels = covar_cols04))

if (nrow(scat) > 2) {
  rho <- scat |>
    group_by(covariate) |>
    summarise(label = paste0("rho = ",
                round(suppressWarnings(cor(aneuploidy_score, value, method = "spearman",
                                           use = "complete.obs")), 2)), .groups = "drop")
  ggplot(scat, aes(aneuploidy_score, value, colour = aneuploidy_class)) +
    highlight_points(scat, "aneuploidy_score", "value", base_colour = "aneuploidy_class",
                     base_palette = attend_aneu_cols,
                     base_name = "aneuploidy", jitter_width = 0, base_size = 1.5, base_alpha = 0.6) +
    geom_smooth(method = "lm", se = FALSE, colour = "grey40", linewidth = 0.6) +
    geom_text(data = rho, aes(x = Inf, y = Inf, label = label),
              inherit.aes = FALSE, hjust = 1.05, vjust = 1.5, size = 3) +
    facet_wrap(~ covariate, scales = "free_y", ncol = 3) +
    labs(title = "Aneuploidy score vs continuous covariates (MMR-deficient only)",
         x = "aneuploidy score", y = "covariate value", colour = NULL) +
    attend_theme() + theme(legend.position = "top")
} else message("Not enough MMR-deficient patients with aneuploidy score + a covariate.")
```

The same covariates against the **binarised** aneuploidy class rather than the continuous
score — the categorical counterpart to the scatter above, with a per-covariate Wilcoxon.
Comparing the two is a check on the `r attend_thresholds$aneuploidy` cutpoint: a
correlation visible in the scatter but absent here means the cutpoint is discarding it.

```{r aneu-class-box, fig.width=12, fig.height=7}
cb <- pt |>
  pivot_longer(all_of(covar_cols04), names_to = "covariate", values_to = "value") |>
  filter(!is.na(aneuploidy_class), !is.na(value)) |>
  mutate(covariate = factor(covariate, levels = covar_cols04))

if (nrow(cb) > 2) {
  ggplot(cb, aes(aneuploidy_class, value, fill = aneuploidy_class)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.5) +
    highlight_points(cb, "aneuploidy_class", "value", base_size = 0.6, base_alpha = 0.4,
                     base_const = "black") +
    stat_compare_means(method = "wilcox.test", label = "p.format", size = 3) +
    facet_wrap(~ covariate, scales = "free_y", ncol = 3) +
    scale_fill_manual(values = attend_aneu_cols) +
    labs(title = "Continuous covariates by aneuploidy class (MMR-deficient only)",
         x = NULL, y = "covariate value") +
    attend_theme() + theme(legend.position = "none")
} else message("Not enough MMR-deficient patients with aneuploidy class + a covariate.")
```

## Aneuploidy by TP53 status

Within MMRd, aneuploidy score in TP53-abnormal versus TP53-normal tumours. TP53 status is
`TP53_pathogenic` from the master, derived from the MAF in report 1. TP53 loss is the
canonical driver of chromosomal instability, so this is the expected-positive control for
the aneuploidy score itself: if aneuploidy does not separate on TP53 here, the score is
suspect before any of the immune results are interpreted.

```{r mmrd-tp53, fig.width=5, fig.height=5}
tp <- pt |> filter(!is.na(TP53), !is.na(aneuploidy_score))

if (nrow(tp) > 2 && dplyr::n_distinct(tp$TP53) == 2) {
  ggplot(tp, aes(TP53, aneuploidy_score, fill = TP53)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.5) +
    highlight_points(tp, "TP53", "aneuploidy_score", base_size = 1.5, base_alpha = 0.5,
                     base_const = "black") +
    stat_compare_means(method = "wilcox.test", label = "p.format") +
    scale_fill_manual(values = attend_tp53_cols) +
    labs(title = "Aneuploidy by TP53 status (MMR-deficient only)",
         x = NULL, y = "aneuploidy score") +
    attend_theme() + theme(legend.position = "none")
} else {
  message("Not enough MMR-deficient patients with both TP53 classes, or TP53_pathogenic ",
          "is missing from the master (rebuild report 1).")
}
```

## Aneuploidy against the pathologist's imaging calls

The pathologist's per-patient scores compared between aneuploidy high and low, within
MMRd. These come from the imaging sheet keyed by image → pid, averaged over a patient's
images for the continuous measures and taking the first non-missing value for the
categorical calls.

```{r imaging-covars}
first_nonNA <- function(x) { x <- x[!is.na(x)]; if (length(x)) x[1] else NA }

imaging_pt <- load_imaging_data() |>
  filter(!is.na(ID)) |>
  group_by(pid = as.character(ID)) |>
  summarise(
    Immune_infiltrate      = mean(suppressWarnings(as.numeric(Immune_infiltrate)),     na.rm = TRUE),
    Immune_infiltrate_PD1  = mean(suppressWarnings(as.numeric(Immune_infiltrate_PD1)), na.rm = TRUE),
    ARID1A                 = first_nonNA(ARID1A),
    Tumor_cells_PDL1       = first_nonNA(Tumor_cells_PDL1),
    Immune_infiltrate_PDL1 = first_nonNA(Immune_infiltrate_PDL1),
    .groups = "drop"
  ) |>
  left_join(master |> select(pid, aneuploidy_class), by = "pid") |>
  filter(!is.na(aneuploidy_class), pid %in% mmrd_pids)   # <- the MMRd restriction

cat("MMR-deficient patients with imaging covariates:", n_distinct(imaging_pt$pid), "\n")
```

Continuous pathologist scores (immune infiltrate and PD-1 infiltrate, as percentages) by
aneuploidy class, Wilcoxon per panel. These are the pathologist's independent read of the
same immune question the FlowPath contrast above asked.

```{r imaging-covars-cont, fig.width=8, fig.height=4}
icc <- imaging_pt |>
  pivot_longer(c(Immune_infiltrate, Immune_infiltrate_PD1),
               names_to = "covariate", values_to = "value") |>
  filter(is.finite(value))

if (nrow(icc) > 2) {
  ggplot(icc, aes(aneuploidy_class, value, fill = aneuploidy_class)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.5) +
    highlight_points(icc, "aneuploidy_class", "value", base_size = 0.6, base_alpha = 0.4,
                     base_const = "black") +
    stat_compare_means(method = "wilcox.test", label = "p.format", size = 3) +
    facet_wrap(~ covariate, scales = "free_y") +
    scale_fill_manual(values = attend_aneu_cols) +
    labs(title = "Pathologist immune scores by aneuploidy class (MMR-deficient only)",
         x = NULL, y = "pathologist %") +
    attend_theme() + theme(legend.position = "none")
}
```

Categorical calls as stacked proportions — how each call distributes within aneuploidy
high vs low. Bars are proportions, so the **group sizes are not visible in this figure**;
read them from the count printed above. `MMR_Status` is omitted because it is constant
here by construction.

```{r imaging-covars-cat, fig.width=11, fig.height=4}
icat <- imaging_pt |>
  select(aneuploidy_class, ARID1A, Tumor_cells_PDL1, Immune_infiltrate_PDL1) |>
  pivot_longer(-aneuploidy_class, names_to = "covariate", values_to = "level") |>
  filter(!is.na(level), level != "")

if (nrow(icat) > 0) {
  ggplot(icat, aes(aneuploidy_class, fill = level)) +
    geom_bar(position = "fill") +
    facet_wrap(~ covariate, scales = "free_x", nrow = 1) +
    labs(title = "Pathologist categorical calls by aneuploidy class (MMR-deficient only)",
         x = NULL, y = "proportion of patients", fill = NULL) +
    attend_theme() +
    theme(axis.text.x = element_text(angle = 20, hjust = 1))
}
```

## Survival by aneuploidy x immune content

Crossing the two axes of Part 2, within MMRd: curves are immune-high vs immune-low (CD45+
fraction of all cells inside, split at the **median** — data-driven, not pre-registered),
panelled by aneuploidy class. With the MMRd cohort split four ways, panel sizes here are
very small.

```{r survival-aneu-immune}
imm <- ct |>
  distinct(pid, n_inside, n_cd45_inside) |>
  mutate(cd45_frac = n_cd45_inside / n_inside) |>
  left_join(master |> select(pid, aneuploidy_class,
                             any_of(c(attend_cols$surv_time, attend_cols$surv_event))),
            by = "pid") |>
  mutate(immune_class = if_else(cd45_frac >= median(cd45_frac, na.rm = TRUE),
                                "immune-high", "immune-low"))

if (have_cols(imm, attend_cols$surv_time, attend_cols$surv_event,
              "immune_class", "aneuploidy_class")) {
  km_facet(imm, group = "immune_class", facets = "aneuploidy_class")
} else message("survival columns not set in attend_cols — skipping KM.")
```

## What to take away

- **Part 1 is measurement validation, Part 2 is biology.** Part 1 runs on every imaged
  patient; Part 2 runs on MMR-deficient patients only. They are not the same cohort and
  should not be quoted together.
- **Concordance first.** If FlowPath and the pathologist disagree on a marker in Part 1,
  any Part 2 result built on that marker inherits the disagreement.
- **One primary endpoint.** CD45+ / all cells inside, arcsine + t-test. Everything else in
  the immune section is secondary or exploratory.
- **The n is small and no correction is applied.** MMRd aneuploidy-high is around 9
  patients. Treat every p-value in Part 2 as suggestive.
- **The median split is not a threshold.** Immune-high/low is defined by this cohort's own
  median, so it moves if the cohort does.

```{r session-info}
sessionInfo()
```
