---
title: "ATTEND — 30. Survival"
author: "sceriff0"
date: "2026-08-11"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
# renv: needs survival, survminer (faceted KM), gtsummary (survival table), tidyverse, here.
# km_facet() matches the reference attend_wes_analysis figures (ggsurvplot_facet). If missing:
#   renv::install(c("survival","survminer","gtsummary")); renv::snapshot()
library(tidyverse)
library(here)
library(survival)
library(survminer)
library(gtsummary)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 7)

source(here("code", "attend_classes.R"))     # column config + km_facet() / surv_table()
source(here("code", "attend_io.R"))          # read_intermediate() (Parquet/CSV)
source(here("code", "build_master.R"))   # get_master() — builds on demand
source(here("code", "attend_ihc.R"))         # ihc_patient_metrics(), for the immune split
source(here("code", "load_phenotypes.R"))    # load_ihc_data()
source(here("code", "load_clinical.R"))      # load_imaging_data()

master <- get_master() |> add_molecular_classes()
```

> **Question.** Does the treatment comparison behave differently in different molecular subgroups?
> **Cohort.** All `r nrow(master)` patients on the master; the immune-content panels are restricted to the imaged subset.
> **Reads.** `attend_master_joined`, plus the FlowPath/imaging join for `immune_class`.
> **Out of scope.** Everything that is not a survival model. Composition itself -> [05_aneuploidy_mmrd](05_aneuploidy_mmrd.html); concordance -> [02_flowpath_concordance](02_flowpath_concordance.html).

Every figure below has the same structure. The two coloured curves are always the same
comparison, `attend_cols$km_group` (the treatment arm); what changes between figures is how
the cohort is split into panels — by aneuploidy and MMR status first, then TMB, HRD, MSI,
and immune content. Each panel carries its own log-rank p-value, comparing the two arms
*within* that panel only. A gap that is similar in every panel means the molecular axis
does not modify the treatment effect; a gap that opens in one panel and closes in another
might, and the interaction test after the headline figure puts a number on it.

Continuous scores are binarised by `add_molecular_classes()` at the configured thresholds:
aneuploidy at `r attend_thresholds$aneuploidy`, TMB at `r attend_thresholds$tmb`, HRD at
`r attend_thresholds$hrd`. Every chunk is guarded by `have_cols()`, so the report knits to a
legible skeleton before those column names are set.

## Data

The master table, with the binarised molecular classes added. The table below is a
cross-tabulation of those five class variables — it shows how the cohort actually
distributes across the panels used later, which is the fastest way to spot a panel that
will be too small to interpret.

```{r data}
# quick visibility on the binarised classes
master |>
  count(aneuploidy_class, MMR_class, TMB_class, HRD_class, MSI_class) |>
  arrange(desc(n)) |> head(20)
```

### Survival summary table

A publication-style survival table (`gtsummary::tbl_survfit`) for the comparison
group — survival at 1/2/3/5 years with confidence intervals. Guarded on the same
columns as the curves below.

```{r surv-table}
if (have_cols(master, attend_cols$surv_time, attend_cols$surv_event, attend_cols$km_group)) {
  tryCatch(surv_table(master, group = attend_cols$km_group),
           error = function(e) message("surv_table skipped (gtsummary/cardx): ", conditionMessage(e)))
}
```

## Headline figure — aneuploidy x MMR status

The main result. Curves are the two treatment arms; **columns** are MMR status
(proficient / deficient, from `MMR_class`) and **rows** are aneuploidy high vs low
(split at `r attend_thresholds$aneuploidy` on the ASCETS aneuploidy score). Each panel's
log-rank p compares the two arms within that panel alone — it is *not* a test of
aneuploidy or MMR.

The table underneath gives each panel's N and event count, with a `small_n` flag
(n < 10 or fewer than 5 events). A panel carrying that flag can produce a dramatic-looking
curve separation from a handful of patients; the flag is there so that is visible rather
than inferred.

```{r km-reference}
if (have_cols(master, attend_cols$surv_time, attend_cols$surv_event,
              attend_cols$km_group, "aneuploidy_class", "MMR_class")) {
  print(km_facet(master, group = attend_cols$km_group,
           facets = c("aneuploidy_class", "MMR_class"),
           title = "Survival by group — faceted by aneuploidy × MMR status"))
  cnt <- km_panel_counts(master, facets = c("aneuploidy_class", "MMR_class"))
  if (nrow(cnt) > 0)
    knitr::kable(cnt, caption = "Per-panel N / events across both arms (small_n = n<10 or events<5 — interpret with caution).")
}
```

### Supporting statistics — subgroup Cox and 3-way interaction

The Kaplan–Meier grid above is the headline result. The subgroup Cox model and the
3-way interaction test below are **supporting** evidence only — they quantify effects
the KM panels already show and stress-test whether the aneuploidy × MMR split
modifies the treatment effect. At this cohort's size the 3-way interaction test is
expected to be underpowered; a non-significant or NA result there is reported
honestly (flagged, never suppressed) and should be read as **suggestive**, not
confirmatory.

```{r cox-mmrd-aneuhigh}
mmrd_aneuhigh <- master |>
  dplyr::filter(MMR_class == attend_levels$mmr_deficient,
                aneuploidy_class == "aneuploidy-high")
eff <- logrank_cox_effect(mmrd_aneuhigh, group = attend_cols$km_group)
if (have_cols(master, attend_cols$surv_time, attend_cols$surv_event, attend_cols$km_group,
              "aneuploidy_class", "MMR_class") && isTRUE(eff$n_groups == 2)) {
  small_flag <- eff$n < 20 || eff$n_events < 10
  cat(sprintf("\n\n**MMRd + aneuploidy-high subgroup: n = %d, events = %d%s**\n\n",
              eff$n, eff$n_events,
              if (small_flag) " — SMALL SAMPLE, treat as suggestive only" else ""))
  ct <- tryCatch(cox_table(mmrd_aneuhigh, covariates = attend_cols$km_group),
                 error = function(e) { message("cox_table skipped: ", conditionMessage(e)); NULL })
  if (!is.null(ct))
    knitr::kable(ct$tidy, digits = 3,
                 caption = "MMRd + aneuploidy-high — Cox PH: PFS ~ treatment arm (HR, 95% CI). n/events via logrank_cox_effect().")
} else message("MMRd × aneuploidy-high Cox subgroup skipped — <2 treatment arms present in this cell.")
```

A three-way Cox model on the master, `treatment arm x aneuploidy class x MMR status`, so
the subgroup Cox above can be read against a single formal test of whether the arms differ
by subgroup rather than against a series of per-panel log-ranks.

```{r interaction-3way}
d_int <- master |>
  dplyr::mutate(dplyr::across(dplyr::all_of(attend_cols$surv_event), recode_event)) |>
  dplyr::filter(dplyr::if_all(dplyr::all_of(c(attend_cols$surv_time, attend_cols$surv_event,
                              attend_cols$km_group, "aneuploidy_class", "MMR_class")), ~ !is.na(.))) |>
  dplyr::mutate(arm  = droplevels(as.factor(.data[[attend_cols$km_group]])),
                aneu = droplevels(as.factor(aneuploidy_class)),
                mmr  = droplevels(as.factor(MMR_class)))
n_int <- nrow(d_int); events_int <- sum(d_int[[attend_cols$surv_event]] == 1, na.rm = TRUE)
levels_ok <- have_cols(master, attend_cols$surv_time, attend_cols$surv_event, attend_cols$km_group,
                       "aneuploidy_class", "MMR_class") &&
             dplyr::n_distinct(d_int$arm) == 2 && dplyr::n_distinct(d_int$aneu) == 2 &&
             dplyr::n_distinct(d_int$mmr) == 2
if (levels_ok && n_int >= 20 && events_int >= 10) {
  ft <- attend_cols$surv_time; fe <- attend_cols$surv_event
  full <- tryCatch(survival::coxph(stats::as.formula(sprintf(
            "survival::Surv(`%s`, `%s`) ~ arm * aneu * mmr", ft, fe)), data = d_int),
          error = function(e) NULL)
  null <- tryCatch(survival::coxph(stats::as.formula(sprintf(
            "survival::Surv(`%s`, `%s`) ~ arm + aneu + mmr", ft, fe)), data = d_int),
          error = function(e) NULL)
  if (!is.null(full) && !is.null(null)) {
    # count only ESTIMATED (non-NA) coefficients: coxph keeps an aliased empty-cell
    # term as an NA entry, so length() would inflate df and mis-calibrate the LRT.
    ddf <- sum(!is.na(stats::coef(full))) - sum(!is.na(stats::coef(null)))
    chi <- 2 * (as.numeric(stats::logLik(full)) - as.numeric(stats::logLik(null)))
    p   <- if (ddf > 0) stats::pchisq(chi, df = ddf, lower.tail = FALSE) else NA_real_
    under <- n_int < 60 || events_int < 30
    cat(sprintf("\n\n**arm × aneuploidy × MMR interaction LRT**: n = %d, events = %d, df = %s, chi-sq = %.2f, p = %s%s\n\n",
        n_int, events_int, ddf, chi,
        if (is.na(p)) "NA (rank-deficient — an empty cell)" else formatC(p, digits = 3, format = "g"),
        if (under) " — **UNDERPOWERED: interpret as suggestive, not confirmatory**" else ""))
  } else message("3-way interaction Cox models failed to fit (rank-deficient/empty cell).")
} else message(sprintf("3-way interaction skipped — need arm/aneu/MMR each 2 levels and n>=20/events>=10 (got n=%d, events=%d).", n_int, events_int))
```

## Split by TMB (cutoff `r attend_thresholds$tmb`) — both TMB definitions

The same arm comparison, now panelled by TMB-high/low against MMR status — **drawn
twice, once per TMB definition**.

Why twice: ATTEND's sequencing is tumour-only, with no matched normal, so leftover
germline variants inflate TMB. Because germline allele frequencies differ between
ancestries, that inflation is ancestry-dependent (Nassar et al., *Cancer Cell* 2022).
The pipeline therefore carries two definitions and never silently picks one:

- **TMB (normal)** — `tmb__TMB_SCORE`, the upstream tumour-only score as delivered.
- **TMB (ancestry corrected)** — `tmb__TMB_nassar`, the same score after the ancestry
  recalibration.

Each definition is re-binarised at the same `r attend_thresholds$tmb` mut/Mb cutoff by
`tmb_class_of()`, so the only thing differing between the two figures is the TMB value
itself. If a patient sits near the cutoff, the correction can move them across it — so
comparing the two figures shows whether any survival signal is robust to the germline
correction or is an artefact of it.

The ancestry-corrected figure is skipped entirely (not drawn empty) until
`attend_ancestry$clinical_col` is set, because the column is all-`NA` before then.

```{r km-tmb, results='asis'}
tmb_defs02 <- tmb_defs_present(master)          # normal + ancestry corrected
tmb_labs02 <- tmb_def_labels(tmb_defs02)

if (have_cols(master, attend_cols$surv_time, attend_cols$surv_event,
              attend_cols$km_group, "MMR_class") && length(tmb_defs02) >= 1) {
  for (i in seq_along(tmb_defs02)) {
    d <- master |> mutate(TMB_class_def = tmb_class_of(.data[[tmb_defs02[[i]]]]))
    if (all(is.na(d$TMB_class_def))) next       # ancestry corrected before clinical_col is set
    cat("\n\n#### ", tmb_labs02[i], " (`", tmb_defs02[[i]], "`)\n\n", sep = "")
    p <- km_facet(d, group = attend_cols$km_group,
                  facets = c("TMB_class_def", "MMR_class"),
                  title = paste0("Survival by arm — ", tmb_labs02[i], " x MMR status"))
    if (!is.null(p)) print(p)
  }
}
```

## Split by HRD score (cutoff `r attend_thresholds$hrd`)

Arms compared within panels of homologous-recombination-deficiency score (high vs low at
`r attend_thresholds$hrd`, from `hrd__`) crossed with MMR status.

```{r km-hrd}
if (have_cols(master, attend_cols$surv_time, attend_cols$surv_event,
              attend_cols$km_group, "HRD_class", "MMR_class")) {
  km_facet(master, group = attend_cols$km_group,
           facets = c("HRD_class", "MMR_class"),
           title = "Survival by group — faceted by HRD × MMR status")
}
```

## Split by MSI status

Arms compared within panels of microsatellite-instability status crossed with MMR status.
MSI and MMR are biologically correlated — MMR loss is the usual cause of MSI — so the
off-diagonal panels here are small by construction and often carry the `small_n`
caveat.

```{r km-msi}
if (have_cols(master, attend_cols$surv_time, attend_cols$surv_event,
              attend_cols$km_group, "MSI_class", "MMR_class")) {
  km_facet(master, group = attend_cols$km_group,
           facets = c("MSI_class", "MMR_class"),
           title = "Survival by group — faceted by MSI × MMR status")
}
```

## Split by immune content

Survival by immune content on the imaged cohort, which is why this section carries a
different denominator from every other section above. Immune content is the CD45+ fraction
inside the annotated region, read per patient from the FlowPath tree by
`ihc_patient_metrics()` (the same helper
[02_flowpath_concordance](02_flowpath_concordance.html) validates) and binarised at its
**median** — a data-driven split, not a pre-registered threshold, so it is descriptive.
Panels are MMR status; curves are immune-high vs immune-low.

```{r ihc-immune-build}
ihc_pt <- tryCatch(ihc_patient_metrics(load_ihc_data(), load_imaging_data()),
                   error = function(e) NULL)
have_ihc <- !is.null(ihc_pt) && nrow(ihc_pt) > 0 && "cd45_over_inside" %in% names(ihc_pt)

if (have_ihc) {
  ihc_surv <- ihc_pt |>
    left_join(master |> select(pid, any_of(c(attend_cols$surv_time, attend_cols$surv_event,
                                             attend_cols$mmr_status, attend_cols$aneuploidy,
                                             "aneuploidy_class"))), by = "pid") |>
    mutate(immune_class = if_else(cd45_over_inside >= median(cd45_over_inside, na.rm = TRUE),
                                  "immune-high", "immune-low"))
  cat("imaged patients with an immune split:", sum(!is.na(ihc_surv$immune_class)), "\n")
} else {
  ihc_surv <- tibble()
  cat("No FlowPath IHC data — the immune-content panels are skipped.\n")
}
```

Treatment arm within each MMR stratum, split by immune content.

```{r survival-by-ihc}
if (have_ihc && have_cols(ihc_surv, attend_cols$surv_time, attend_cols$surv_event,
                          attend_cols$mmr_status, "immune_class")) {
  km_facet(ihc_surv, group = "immune_class", facets = attend_cols$mmr_status)
} else message("survival columns not set in attend_cols — skipping KM by IHC feature.")
```

Aneuploidy class crossed with immune content. This panel keeps the **MMR-deficient**
restriction it carried in its previous home: it was computed on the MMRd cell-type table,
so widening it to every imaged patient would change the result rather than relocate it.
Immune content is the same median split as above, recomputed inside the MMRd subset so the
median is that subset's own.

```{r survival-aneu-immune}
mmrd_imaged <- if (have_ihc && attend_cols$mmr_status %in% names(ihc_surv))
  ihc_surv |> filter(is_mmrd(.data[[attend_cols$mmr_status]])) else tibble()

if (nrow(mmrd_imaged) > 0) {
  imm <- mmrd_imaged |>
    mutate(immune_class = if_else(cd45_over_inside >= median(cd45_over_inside, na.rm = TRUE),
                                  "immune-high", "immune-low"))
  if (have_cols(imm, attend_cols$surv_time, attend_cols$surv_event,
                "immune_class", "aneuploidy_class")) {
    km_facet(imm, group = "immune_class", facets = "aneuploidy_class")
  } else message("survival columns not set in attend_cols — skipping KM.")
} else message("No MMR-deficient imaged patients — aneuploidy x immune KM skipped.")
```

```{r session-info}
sessionInfo()
```
