---
title: "ATTEND — 05. Aneuploidy in MMR-deficient tumours"
author: "sceriff0"
date: "2026-08-13"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
# renv: tidyverse, here, ggpubr.
library(tidyverse)
library(here)
library(ggpubr)
# Headless-safe raster device. The cluster's compute nodes have no X11, so R's default
# png() fails with "unable to open connection to X11 display" and every figure dies. cairo
# needs no display; the guard keeps this working on a machine built without it.
if (capabilities("cairo")) options(bitmapType = "cairo")

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 10, fig.height = 6)

source(here("code", "build_master.R"))       # get_master() — builds on demand, no build order
source(here("code", "attend_harmonise.R"))   # build_barcode_pid(), make_id_cfg()
source(here("code", "attend_plots.R"))       # highlight_points(), attend_theme()
source(here("code", "attend_ihc.R"))         # ihc_celltype_metrics(), ihc_content_metrics()
source(here("code", "load_phenotypes.R"))    # load_ihc_celltypes()

master <- get_master() |> add_molecular_classes() |>
  mutate(aneuploidy_class = factor(aneuploidy_class,
                                   levels = c("aneuploidy-high", "aneuploidy-low")))
mmrd_pids <- add_scna_group(master) |> filter(mmr_group == "MMRd") |> pull(pid)
n_mmrd <- length(mmrd_pids)

# The polipo highlight (21S188) is drawn here, as it is on every per-patient figure in the
# pipeline -- there is no per-report opt-out any more. Its ids are resolved through the
# barcode -> pid crosswalk by highlight_group_of(), so the overlay lands on this report's
# pid-keyed frames; before that resolution existed it matched nothing here and drew nothing.
```

> **Question.** Within MMR-deficient tumours, what does aneuploidy track — mutation status, or cell-type composition?
> **Cohort.** MMR-deficient only (n = `r n_mmrd`), with one marked exception: the interaction model, which cannot be estimated inside a single stratum.
> **Reads.** The master (built on demand by get_master()), the long MAF, and the FlowPath annotation tree.
> **Out of scope.** Anything on the full cohort. Aneuploidy against TMB across both MMR strata -> [08-tmb](08-tmb.html). The same composition panels keyed on treatment response -> [06-response](06-response.html).

The restriction is the point. MMR deficiency is itself a route to genomic instability, so
in the whole cohort any variable that tracks MMR status will also appear to track
aneuploidy. Conditioning on MMRd removes that shortcut. The outcome in every figure is the
same continuous aneuploidy score; what changes is the grouping on the x-axis — mutation
status in Part 1, cell-type composition in Part 2. The cohort definition, the aneuploidy
cut and the highlight colours are in [00-methods](00-methods.html).

## Part 1 — aneuploidy by mutation status

The restriction is the point. MMR deficiency is itself a route to genomic instability, so
in the whole cohort any mutation that tracks MMR status will also appear to track
aneuploidy. Conditioning on MMRd removes that shortcut. The outcome variable in every
figure is the same continuous aneuploidy score; what changes is the grouping on the x-axis.

#### The gene panel, spelled out

Genes are tested one at a time, each in its own panel, so a signal is attributable to a
specific gene rather than to a union that could be driven by whichever gene is most
frequently mutated. Two groups, kept deliberately separate:

| Group | Genes | Why they are grouped this way |
|---|---|---|
| **MMR panel** | `r paste(attend_gene_panel, collapse = ", ")` | Loss of any of these causes mismatch-repair deficiency. MSH3 pairs with MSH2 to form MutSb. |
| **Wnt / b-catenin** | `r paste(attend_gene_panel_wnt, collapse = ", ")` | An independent oncogenic pathway, not an MMR mechanism. |
| **TP53** | TP53 | Handled separately: it is already a first-class column on the master. |

The MMR union (panel-altered = at least one pathogenic variant in any MMR panel gene) is
shown last, after the per-gene panels. CTNNB1 and APC are deliberately excluded from
that union, because folding them in would change what "panel-altered" means everywhere else
in the pipeline. There is no Wnt union: with two genes the per-gene panels already show
everything a union would.

P-values come from stat_compare_means(), Wilcoxon rank-sum for two groups and
Kruskal-Wallis for more, uncorrected across panels. With one test per gene, treat them as
descriptive.

### Data

Two sources, joined on pid. The aneuploidy score and its derived classes come from
attend_master_joined. The mutation status is recomputed here directly from the long MAF
rather than taken from the master's collapsed columns, because the average of a variant row
is meaningless; pathogenic_by_patient() and mutation_status_long() are the same helpers
[01-data-integration](01-data-integration.html) used, so the two cannot drift. Crossing the
crosswalk against the full gene list guarantees every gene gets a column even when it never
appears in the MAF, with unmutated patients FALSE rather than NA.

```{r data}
cw  <- build_barcode_pid(load_gianlu_clinical_data(), make_id_cfg())
maf <- read_intermediate("maf_data")

mut <- pathogenic_by_patient(maf, cw)   # pid, TP53_pathogenic, panel_pathogenic (MMR union)

# Per-gene status for the one-gene-at-a-time panels: MMR panel + Wnt genes.
gene_status <- cw |> distinct(pid) |>
  tidyr::crossing(gene = attend_genes_per_gene) |>
  left_join(mutation_status_long(maf, cw), by = c("pid", "gene")) |>
  mutate(mutated = replace_na(mutated, FALSE),
         class   = if_else(mutated, paste0(gene, "-mut"), paste0(gene, "-wt"))) |>
  select(pid, gene, class) |>
  pivot_wider(names_from = gene, values_from = class, names_glue = "{gene}_class")

dat_all <- master |>
  select(-any_of(c("TP53_pathogenic", "panel_pathogenic"))) |>  # recompute here
  left_join(mut, by = "pid") |>
  left_join(gene_status, by = "pid") |>
  mutate(
    aneuploidy_value = suppressWarnings(as.numeric(.data[[attend_cols$aneuploidy]])),
    TP53_class  = if_else(TP53_pathogenic,  "TP53-mut",     "TP53-wt"),
    panel_class = if_else(panel_pathogenic, "MMR-panel-altered", "MMR-panel-wt")
  )
```
#### Restricting to MMR-deficient

The single filter the whole report rests on. The counts below are the cohort every
figure that follows is drawn from — if the MMRd group is small, every p-value in this
report inherits that.

```{r restrict-mmrd}
have_mmr <- have_cols(dat_all, "MMR_class", "aneuploidy_value")

dat <- if (have_mmr) dat_all |> filter(MMR_class == attend_levels$mmr_deficient) else dat_all[0, ]

cat("Whole cohort              :", nrow(dat_all), "patients\n")
cat("MMR-deficient (this report):", nrow(dat), "patients\n")
if (nrow(dat) > 0)
  cat("  with an aneuploidy score :", sum(!is.na(dat$aneuploidy_value)), "patients\n")

if (have_mmr && nrow(dat) == 0)
  message("No MMR-deficient patients (MMR_class == '", attend_levels$mmr_deficient,
          "') — every figure below is skipped.")
```

Coverage of the manual highlight overlay on this report's cohort. `attend_highlight` holds
one group, `polipo` (sample 21S188, blue), and 21S188 is a sequencing barcode while `dat` is
keyed on `pid`; `highlight_coverage()` scores the match after resolving the id through the
barcode -> pid crosswalk, so `1/1 matched` means the point really is drawn below and not that
the string happened to appear somewhere. `0/1` means the patient is outside the MMRd
restriction, which is a fact about this cohort, not a broken overlay.

```{r highlight-coverage}
if (nrow(dat) > 0) cat(highlight_coverage(dat))
```

```{r box-helper, include=FALSE}
# Shared boxplot: aneuploidy score on y, one facet per grouping variable. Boxes are
# neutral; points are coloured by aneuploidy class, with the highlight overlay drawn
# on top (partitioned, so no patient is plotted twice). highlight_points() owns both
# colour scales, so no scale_colour_manual is added here.
aneu_box <- function(d, vars, title, subtitle = NULL, ncol = 3) {
  long <- d |>
    select(pid, aneuploidy_value, aneuploidy_class, any_of(vars)) |>
    pivot_longer(any_of(vars), names_to = "comparison", values_to = "group") |>
    filter(!is.na(aneuploidy_value), !is.na(group)) |>
    mutate(comparison = factor(comparison, levels = vars))
  if (nrow(long) == 0) return(NULL)

  ggplot(long, aes(group, aneuploidy_value)) +
    # Was fill = #A6CEE3, which attend_plots.R defines as "aneuploidy-low" — so every box in
    # a figure ABOUT aneuploidy was tinted with one of its two class colours. Neutral now, and
    # the class colour survives where it belongs: on the points, via base_colour below.
    attend_box(long, x = "group", y = "aneuploidy_value", by = "comparison", points = FALSE) +
    highlight_points(long, "group", "aneuploidy_value", base_colour = "aneuploidy_class",
                     base_palette = attend_aneu_cols, base_name = "aneuploidy",
                     base_size = 0.7, base_alpha = 0.6) +
    facet_wrap(~ comparison, scales = "free_x", ncol = ncol) +
    stat_compare_means(label = "p.format", size = 3) +
    labs(title = title, subtitle = subtitle, x = NULL, y = "aneuploidy score") +
    attend_theme() +
    theme(axis.text.x = element_text(angle = 20, hjust = 1))
}

# Group sizes behind each panel — a two-group Wilcoxon on 3 vs 40 patients looks
# identical to one on 20 vs 23 unless the counts are printed.
group_counts <- function(d, vars) {
  d |>
    select(pid, aneuploidy_value, any_of(vars)) |>
    pivot_longer(any_of(vars), names_to = "comparison", values_to = "group") |>
    filter(!is.na(group), !is.na(aneuploidy_value)) |>
    count(comparison, group, name = "n") |>
    pivot_wider(names_from = group, values_from = n, values_fill = 0)
}
```

### 1. Aneuploidy by gene — MMR panel, one gene at a time

Each panel is one MMR gene: patients carrying a pathogenic variant in that gene
(<GENE>-mut) versus those who do not (<GENE>-wt), within the MMRd cohort only.
Pathogenicity is decided by attend_pathogenic_classes on the MAF's
Variant_Classification.

Read the group sizes first. Some of these genes are mutated in very few patients, and a
panel with a handful of -mut patients will show a wide box and an unstable p-value.

```{r mmr-gene-counts}
mmr_gene_vars <- paste0(attend_gene_panel, "_class")
if (nrow(dat) > 0)
  knitr::kable(group_counts(dat, mmr_gene_vars),
               caption = "Patients per group, MMR panel genes (MMR-deficient cohort)")
```

Aneuploidy score (aneu__aneuploidy_score, ASCETS arm-level burden from data/ascets/)
in mutated versus wild-type patients, one panel per MMR gene, over the dat cohort above.

```{r mmr-gene-box, fig.height=7}
if (nrow(dat) > 0)
  aneu_box(dat, mmr_gene_vars,
           "Aneuploidy score by MMR panel gene (MMR-deficient only)",
           paste0("One panel per gene; pathogenic vs wild-type. n = ", nrow(dat),
                  " MMR-deficient patients"))
```

### 2. Aneuploidy by gene — Wnt / β-catenin pathway

The same comparison for `r paste(attend_gene_panel_wnt, collapse = " and ")`. These are
shown apart from the MMR panel because they are a different mechanism: they are not a
cause of mismatch-repair deficiency, so a signal here would mean something biologically
different from a signal in the MMR panel above.

```{r wnt-gene-counts}
wnt_gene_vars <- paste0(attend_gene_panel_wnt, "_class")
if (nrow(dat) > 0)
  knitr::kable(group_counts(dat, wnt_gene_vars),
               caption = "Patients per group, Wnt pathway genes (MMR-deficient cohort)")
```

The same aneuploidy score against Wnt-pathway mutation status, two panels.

```{r wnt-gene-box, fig.height=4.5}
if (nrow(dat) > 0)
  aneu_box(dat, wnt_gene_vars,
           "Aneuploidy score by Wnt pathway gene (MMR-deficient only)",
           "CTNNB1 and APC — shown separately from the MMR panel; not part of the MMR union",
           ncol = 2)
```

### 3. TP53 and the MMR union

TP53 on its own, then the MMR panel **union** — at least one pathogenic variant in any
of `r paste(attend_gene_panel, collapse = ", ")`. The union is the last mutation view
rather than the first: it is the least specific, and the per-gene panels above are what
say which gene any union signal came from.

```{r tp53-union-counts}
summary_vars <- c("TP53_class", "panel_class")
if (nrow(dat) > 0)
  knitr::kable(group_counts(dat, summary_vars),
               caption = "Patients per group, TP53 and the MMR union (MMR-deficient cohort)")
```

Aneuploidy score by TP53 status and by the MMR union, side by side, so a union signal can
be read against the per-gene panels above that would have to explain it.

```{r tp53-union-box, fig.height=4.5}
if (nrow(dat) > 0)
  aneu_box(dat, summary_vars,
           "Aneuploidy score by TP53 status and by the MMR panel union (MMR-deficient only)",
           ncol = 2)
```

### 4. Aneuploidy by the other molecular classes

The non-mutation axes, still within MMRd: TMB, HRD, and MSI status. MMR_class is
omitted because it is constant here by construction — every patient in this report is
MMR-deficient.

MSI is worth reading carefully: MMR loss is the usual *cause* of microsatellite
instability, so within an MMRd cohort the MSI-stable group is both small and unusual.

```{r other-classes-box, fig.height=4.5}
other_vars <- c("TMB_class", "HRD_class", "MSI_class")
if (nrow(dat) > 0) {
  knitr::kable(group_counts(dat, other_vars),
               caption = "Patients per group, score-derived classes (MMR-deficient cohort)")
  aneu_box(dat, other_vars,
           "Aneuploidy score by TMB / HRD / MSI class (MMR-deficient only)",
           ncol = 3)
}
```

## Part 2 — aneuploidy by cell-type composition

#### Cell types and normalization

"Cell type" is phenotype_clean, the parenthetical phenotype label in the FlowPath export
(changeable in code/load_phenotypes.R). Every count is taken inside the annotated region
and expressed as a fraction, but the denominator changes the question, so all three are
shown:

| Normalization | Denominator | The question it answers |
|---|---|---|
| **all cells inside** | every cell in the annotation | what fraction of the tissue is this cell type? |
| **tumour cells inside** | tumour cells only | how much of this cell type per unit of tumour? |
| **CD45+ cells inside** | leukocytes only | what is the composition *of the immune compartment*? |

Each is drawn twice: the raw fraction with a Wilcoxon test, and the
arcsine-square-root-transformed fraction with a two-sample t-test. Agreement between the
two is reassurance; disagreement means the result depends on the test rather than the data.
No multiple-testing correction is applied anywhere in this report, and with one test per
cell type per normalization that is many comparisons, so every p-value here is descriptive.

### Building the restricted cohort

Three things are joined on pid: the aneuploidy class from the master, the per-patient
cell-type composition read from the FlowPath tree by ihc_celltype_metrics(), and the
continuous covariates. MMR status comes from add_scna_group(), so it is the same
is_mmrd() call the copy-number and classification reports use. The counts printed here
are the cohort for every figure below.

```{r composition-data}
master_grp <- add_scna_group(master)        # adds mmr_group ("MMRd"/"MMRp") + scna_group

# per-patient cell-type composition inside the annotation (reads the FlowPath tree)
ct_all <- ihc_celltype_metrics(load_ihc_celltypes(), load_imaging_data()) |>
  left_join(master |> select(pid, aneuploidy_class), by = "pid") |>
  filter(!is.na(aneuploidy_class))
ct <- ct_all |> filter(pid %in% mmrd_pids)          # <- the MMRd restriction

# per-patient continuous covariates + TP53 status, restricted the same way
numcol <- function(col) if (col %in% names(master)) suppressWarnings(as.numeric(master[[col]])) else NA_real_
pt_all <- tibble(
  pid              = master$pid,
  aneuploidy_class = master$aneuploidy_class,
  aneuploidy_score = numcol(attend_cols$aneuploidy),
  HRD              = numcol(attend_cols$hrd),
  MSI              = numcol("msi__MSI_SCORE"),
  TP53             = if ("TP53_pathogenic" %in% names(master))
                       if_else(master$TP53_pathogenic, "TP53-abnormal", "TP53-normal") else NA_character_
)

# The two TMB definitions, added under their reader-facing names so every facet reads
# "TMB (normal)" / "TMB (ancestry corrected)" rather than an internal column name.
tmb_defs04 <- tmb_defs_present(master)
tmb_labs04 <- tmb_def_labels(tmb_defs04)
for (i in seq_along(tmb_defs04)) pt_all[[tmb_labs04[i]]] <- numcol(tmb_defs04[[i]])
covar_cols04 <- c(tmb_labs04, "HRD", "MSI")

pt <- pt_all |> filter(pid %in% mmrd_pids)          # <- the MMRd restriction

cat("MMR-deficient patients                       :", length(mmrd_pids), "\n")
cat("  ... with cell-type metrics + aneuploidy class:", n_distinct(ct$pid), "\n")
cat("  ... cell types measured                      :", n_distinct(ct$cell_type), "\n")
cat("TMB definitions plotted:", paste(tmb_labs04, collapse = " | "), "\n\n")
if (nrow(ct) > 0) print(table(aneuploidy = droplevels(factor(
  ct |> distinct(pid, aneuploidy_class) |> pull(aneuploidy_class)))))
```

```{r box-helpers, include=FALSE}
# One boxplot per cell type, aneuploidy-high vs -low, all cell types on one axis. Boxes are
# filled by aneuploidy class; the highlight overlay is drawn on top in its own colour scale
# (colour != fill, so the two legends do not collide), partitioned so no patient is plotted
# twice. `arcsin` switches BOTH the y variable and the test, so the raw/Wilcoxon and the
# arcsin/t-test versions cannot drift apart the way two copied functions did.
# The y variable is always `.v`, prepared by draw_composition() — the transform happens there
# because the axis has to be computed on the same values the panels plot.
box_by_aneu <- function(df, denom_label, arcsin = FALSE, scope = NULL,
                        width_in = 13, height_in = 11, unit_override = NULL,
                        extra_note = "") {
  d <- df |> filter(is.finite(.v)) |> droplevels()
  if (nrow(d) == 0) return(NULL)
  method <- if (arcsin) "t.test" else "wilcox.test"
  unit   <- if (!is.null(unit_override)) unit_override
            else if (arcsin) paste0("arcsin(sqrt(fraction of ", denom_label, "))")
            else        paste0("fraction of ", denom_label)
  sub    <- if (arcsin) "arcsine-square-root transformed fraction; two-sample t-test"
            else        "raw fraction; Wilcoxon rank-sum"
  # The ceiling and the p-value height come from ONE call: ggpubr's default label.y is the
  # PLOT-WIDE maximum, so on this shared axis a single extreme patient parks every label at
  # the ceiling and squashes every box. See frac_display_y() in attend_plots.R.
  # A title fragment per scope. The fallback capitalises rather than passing the scope
  # through verbatim -- an unmapped scope produced a title starting in lower case.
  what <- if (is.null(scope)) "Cell types"
          else switch(scope,
                      `focus set`      = "Cell types (focus)",
                      `all cell types` = "Cell types (all)",
                      `tissue content` = "Tissue content",
                      `tissue content, mixed denominators` = "Tissue content (mixed denominators)",
                      sub("^(.)", "\\U\\1", scope, perl = TRUE))
  sty <- ihc_fig_style()
  yy  <- frac_display_y(d, ".v", c("aneuploidy_class", "cell_type"))
  ggplot(d, aes(aneuploidy_class, .v, fill = aneuploidy_class)) +
    attend_box(d, x = "aneuploidy_class", y = ".v", by = "cell_type", fill = NULL,
               points = FALSE, expand_y = FALSE,
               linewidth = sty$linewidth, label_size = sty$label_size) +
    highlight_points(d, "aneuploidy_class", ".v", base_size = 0.5, base_alpha = 0.35,
                     base_const = "black") +
    facet_wrap(~ cell_type) +
    stat_compare_means(method = method, label = "p.format", size = sty$p_size,
                       label.y = yy$label_y) +
    shared_frac_y(yy) +
    scale_fill_manual(values = attend_aneu_cols) +
    labs(title = wrap_fig_text(paste0(what, " by aneuploidy — MMRd only"),
                               width_in, sty$title_size, sty$gutter_in),
         subtitle = wrap_fig_text(paste0(sub, yy$note, extra_note),
                                  width_in, sty$sub_size, sty$gutter_in),
         x = NULL,
         y = if (!nzchar(unit)) NULL else wrap_fig_text(unit, height_in, sty$sub_size)) +
    attend_theme(base_size = sty$base_size) +
    # attend_theme() sizes the title off base_size, so these two must be restated after it:
    # the plot type is scaled up, the labelling deliberately is not.
    theme(legend.position = "none", axis.text.x = element_text(angle = 20, hjust = 1),
          plot.title    = element_text(size = sty$title_size, face = "bold"),
          plot.subtitle = element_text(size = sty$sub_size, colour = "grey30"))
}

# ONE figure per normalization, every cell type in it, on one axis. The transform happens
# here rather than in box_by_aneu() because the axis has to be computed on the same values
# the panels plot.
draw_composition <- function(frac_col, denom_label, arcsin = FALSE) {
  d <- ct |>
    mutate(.v = if (arcsin) arcsin_sqrt(.data[[frac_col]]) else .data[[frac_col]]) |>
    filter(is.finite(.v))
  if (nrow(d) == 0) {
    message("Skipped: ", denom_label, " — no finite fractions."); return(invisible(NULL))
  }
  # The focus set first, on an axis of its own: these four sit under ~0.6 on the arcsin
  # scale, so sharing an axis with `Tumor` (~1.0) leaves them unreadable. Then every cell
  # type on one axis, which is the comparison the shared scale exists for.
  foc <- filter(d, cell_type %in% attend_ihc_focus)
  if (nrow(foc) > 0) {
    p <- box_by_aneu(foc, denom_label, arcsin, scope = "focus set")
    if (!is.null(p)) print(p)
  } else {
    message("Focus panel skipped: none of attend_ihc_focus found in cell_type (",
            paste(attend_ihc_focus, collapse = ", "), ").")
  }
  p <- box_by_aneu(d, denom_label, arcsin, scope = "all cell types")
  if (!is.null(p)) print(p)
}
```

### Cell-type composition by aneuploidy class

One panel per cell type, aneuploidy-high vs -low, within MMRd, from the ct table built
above. Each normalization is shown as a raw-fraction/Wilcoxon figure followed by its
arcsine/t-test companion.

Each normalization is drawn **twice**: first the cell types in `attend_ihc_focus`
(`T cytotoxic`, `T helper`, `Macrophages`, `M1`) on an axis of their own, then every cell
type on one axis. The focus figure exists because `Tumor` reaches ~1.0 on the arcsine scale
while all four of these sit under ~0.6, so an axis wide enough for `Tumor` leaves them
unreadable — and these four are what the figure is for.

Every cell type is drawn in **one figure per normalization, on a single shared y axis**.
A panel's height therefore reads as that cell type's abundance relative to every other one,
which is the comparison the figure exists to support: `Macrophages` is read against
`T cytotoxic` on the same scale, not against a rescaled axis of its own. `scales = "free_y"`
would give each panel its own range and make a cell type at a fraction of a percent of the
tissue draw the same box, in the same place, as one at half the tissue. The cost is accepted
deliberately — on the tumour-cell denominator these cell types span orders of magnitude, so
the rarest panels sit low against a scale set by the most abundant.

The axis stops at the **largest upper whisker** in the figure (Q3 + 1.5 IQR over every class
x cell-type cell), not at the largest value, and the p-value label is placed inside that
range by the same call. Both follow from one measurement: ggpubr derives its default label
height from the plot-wide maximum, so a single extreme patient simultaneously flattened
every box and parked every p-value at the ceiling, where it read as no p-value at all. The
range is a **zoom** (`coord_cartesian()`), so every patient is still in the box statistics
and in the test — nothing is dropped, and a point above the axis is counted in the subtitle.
Groups smaller than `attend_min_box_n` still show a median crossbar instead of a box.

#### Normalized by all cells inside

Each cell type as a fraction of every cell in the annotation, so a panel reads "how much
of the tissue is this". Raw fractions, Wilcoxon.

```{r box-inside, fig.width=13, fig.height=11, out.width="100%"}
draw_composition("frac_inside", "all cells inside")
```

The same panels on the arcsine-square-root scale with a two-sample t-test, which is the
variance-stabilised companion rather than an alternative answer.

```{r box-inside-arcsin, fig.width=13, fig.height=11, out.width="100%"}
draw_composition("frac_inside", "all cells inside", arcsin = TRUE)
```

#### Normalized by tumour cells inside

Each cell type per unit of tumour: the denominator is tumour cells only, so a shift here
survives a change in how much tumour the annotation caught. Raw fractions, Wilcoxon.

```{r box-tumor, fig.width=13, fig.height=11, out.width="100%"}
draw_composition("frac_tumor", "tumour cells inside")
```

The arcsine/t-test companion of the panels above.

```{r box-tumor-arcsin, fig.width=13, fig.height=11, out.width="100%"}
draw_composition("frac_tumor", "tumour cells inside", arcsin = TRUE)
```

#### Normalized by CD45+ cells inside

Composition *of the immune compartment*: the denominator is leukocytes, so tumour-content
differences between patients cannot drive the contrast. Raw fractions, Wilcoxon.

```{r box-cd45, fig.width=13, fig.height=11, out.width="100%"}
draw_composition("frac_cd45", "CD45+ cells inside")
```

The arcsine/t-test companion of the panels above.

```{r box-cd45-arcsin, fig.width=13, fig.height=11, out.width="100%"}
draw_composition("frac_cd45", "CD45+ cells inside", arcsin = TRUE)
```

#### Tissue content: tumour and leukocyte, on one axis

How much of the annotation is tumour, and how much is leukocyte, as fractions of the same
denominator so the two are directly comparable. `Tumor` is a cell-type row from
`ihc_celltype_metrics()`; CD45+ is a per-image denominator constant that only
`ihc_immune_metrics()` surfaces, so `ihc_content_metrics()` unions the two tables onto one
`cell_type` column. **`all cells inside` is the only denominator on which both are
meaningful** — `Tumor` over tumour cells and CD45+ over CD45+ cells are both identically 1
— so this panel is drawn once, not once per normalization. Raw fractions with a Wilcoxon,
then the arcsine/t-test companion.

```{r box-content, fig.width=13, fig.height=7, out.width="100%"}
ct_content <- ihc_series_metrics(ct, imm_metrics, attend_ihc_content) |>
  left_join(master |> select(pid, aneuploidy_class), by = "pid") |>
  filter(!is.na(aneuploidy_class))

for (asin in c(FALSE, TRUE)) {
  d <- ct_content |> mutate(.v = if (asin) arcsin_sqrt(frac) else frac)
  p <- box_by_aneu(d, "all cells inside", asin, scope = "tissue content",
                   width_in = 13, height_in = 7, unit_override = "")
  if (!is.null(p)) print(p)
}
```

#### Tumour burden beside immune infiltration per unit of tumour

The same two populations against **different** denominators: `Tumor` over all cells inside
(how much of the tissue is tumour) beside CD45+ over tumour cells inside (how many
leukocytes per tumour cell). Each panel strip names its own denominator, because the y axis
carries one unit and cannot.

CD45+ / tumour cells is a **ratio, not a proportion** — it exceeds 1 wherever leukocytes
outnumber tumour cells — and `arcsin_sqrt()` clamps its input to [0, 1]. Both versions are
drawn, and the count of values above 1 is printed above the figures **and** in the
transformed figure's own subtitle: those patients all land on the same value there, so a
clamped group is a wall of identical numbers and its t-test is not interpretable. Read the
raw/Wilcoxon panel whenever that count is non-zero; when it is zero the transform is exact
and the two panels are the usual pair.

```{r box-content-mixed, fig.width=13, fig.height=7, out.width="100%"}
mix <- ihc_series_metrics(ct, imm_metrics, attend_ihc_content_mixed) |>
  left_join(master |> select(pid, aneuploidy_class), by = "pid") |>
  filter(!is.na(aneuploidy_class)) |>
  mutate(.v = frac)

n_over <- sum(mix$frac > 1, na.rm = TRUE)
cat("patients:", n_distinct(mix$pid), "| series:",
    paste(gsub("\n", " ", levels(mix$cell_type)), collapse = " | "),
    "\nvalues above 1 (a ratio, not a proportion):", n_over,
    if (n_over > 0) " <- clamped by arcsin_sqrt() in the transformed figure below" else "", "\n")

# Raw, then the arcsine/t-test companion. The companion carries a CLAMP COUNT because
# arcsin_sqrt() bounds its input to [0, 1] and CD45+/tumour is a ratio that can exceed it:
# every value above 1 lands on pi/2, so a clamped group is a wall of identical numbers and
# its t-test is not interpretable. When the count is 0 the transform is exact and the note
# does not appear, which is the case this figure is drawn for.
for (asin in c(FALSE, TRUE)) {
  note <- if (asin && n_over > 0)
            sprintf("; %d value%s > 1 CLAMPED to pi/2 by arcsin(sqrt()) — read the raw panel",
                    n_over, if (n_over == 1L) "" else "s") else ""
  d <- mix |> mutate(.v = if (asin) arcsin_sqrt(frac) else frac)
  p <- box_by_aneu(d, "the denominator in each panel", arcsin = asin,
                   scope = "tissue content, mixed denominators",
                   width_in = 13, height_in = 7, unit_override = "", extra_note = note)
  if (!is.null(p)) print(p)
}
```

### Immune content — the headline contrast

The question this report exists for: among MMR-deficient tumours, is aneuploidy-high
associated with an immune-cold phenotype?

Two leukocyte metrics counted inside the annotation — **CD45+ total** (all leukocytes)
and **CD3+CD45+** (double-positive T cells) — under each normalization. CD3+CD45+
requires a CD3_sign column in the FlowPath export; without it those rows are NA and
their facets drop out on their own.

```{r immune-data}
imm_all <- ihc_immune_metrics(load_ihc_celltypes(), load_imaging_data()) |>
  left_join(master |> select(pid, aneuploidy_class), by = "pid") |>
  filter(!is.na(aneuploidy_class))
imm_metrics <- imm_all |> filter(pid %in% mmrd_pids)      # <- the MMRd restriction

cat("MMR-deficient patients with immune metrics:", n_distinct(imm_metrics$pid),
    "| metrics present:", paste(sort(unique(imm_metrics$metric)), collapse = ", "), "\n")
```

```{r immune-box-helper, include=FALSE}
# One figure PER DENOMINATOR, faceted by metric. The five immune rows carry three different
# denominators, and on one shared axis CD3+CD45+ over CD45+ sits an order of magnitude above
# the same numerator over all cells inside — so the panels that are not being compared set
# the scale for the ones that are. Splitting on denom_label means every panel in a figure is
# a fraction OF THE SAME THING, which is the only case where a shared axis is honest.
box_immune <- function(df, denom_, arcsin = FALSE, width_in = 13, height_in = 11) {
  d <- df |> filter(is.finite(frac), denom_label == denom_) |> droplevels()
  if (nrow(d) == 0) return(NULL)
  if (arcsin) d <- d |> mutate(frac = arcsin_sqrt(frac))
  method <- if (arcsin) "t.test"                 else "wilcox.test"
  ylab   <- if (arcsin) "arcsin(sqrt(fraction))" else "fraction"
  ttl    <- if (arcsin) "arcsin(sqrt) + t-test"  else "raw fraction + Wilcoxon"
  sty <- ihc_fig_style()
  yy  <- frac_display_y(d, "frac", c("aneuploidy_class", "metric"))
  ggplot(d, aes(aneuploidy_class, frac, fill = aneuploidy_class)) +
    attend_box(d, x = "aneuploidy_class", y = "frac", by = "metric", fill = NULL,
               points = FALSE, expand_y = FALSE,
               linewidth = sty$linewidth, label_size = sty$label_size) +
    highlight_points(d, "aneuploidy_class", "frac", base_size = 0.6, base_alpha = 0.4,
                     base_const = "black") +
    facet_wrap(~ metric) +
    stat_compare_means(method = method, label = "p.format", size = sty$p_size,
                       label.y = yy$label_y) +
    shared_frac_y(yy) +
    scale_fill_manual(values = attend_aneu_cols) +
    labs(title = wrap_fig_text(paste0("Immune content by aneuploidy (MMRd only) — ", ttl), width_in, sty$title_size, sty$gutter_in),
         subtitle = wrap_fig_text(paste0("normalized by ", denom_, yy$note), width_in, sty$sub_size, sty$gutter_in),
         x = NULL,
         y = if (!nzchar(ylab)) NULL else wrap_fig_text(ylab, height_in, sty$sub_size)) +
    attend_theme(base_size = sty$base_size) +
    # attend_theme() sizes the title off base_size, so these two must be restated after it:
    # the plot type is scaled up, the labelling deliberately is not.
    theme(legend.position = "none", axis.text.x = element_text(angle = 20, hjust = 1),
          plot.title    = element_text(size = sty$title_size, face = "bold"),
          plot.subtitle = element_text(size = sty$sub_size, colour = "grey30"))
}
```

CD45+ total and CD3+CD45+ fractions from imm_metrics, by aneuploidy class, one facet per
normalization, within MMRd. One figure per denominator rather than one figure with all
three: a shared axis is only honest between panels that are fractions of the same thing,
and CD3+CD45+ over CD45+ sits an order of magnitude above the same numerator over all cells
inside. The denominator is named in each subtitle.

```{r immune-box, fig.width=12, fig.height=8, out.width="100%"}
if (nrow(filter(imm_metrics, is.finite(frac))) > 2) {
  # Denominator order is the order ihc_immune_metrics() declares its specs in, so the
  # figures come out widest-denominator first however the table happened to be sorted.
  for (dn in unique(imm_metrics$denom_label))
    for (asin in c(FALSE, TRUE)) {
      p <- box_immune(imm_metrics, dn, asin); if (!is.null(p)) print(p)
    }
} else message("No finite immune metrics in the MMRd cohort (CD3/CD45 counts missing) — skipping.")
```

#### The pre-specified primary contrast

Of everything above, one comparison is designated **primary**: **CD45+ / all cells
inside**, tested with the **arcsine-square-root transform and a two-sample t-test**. The
Wilcoxon on the raw fraction is reported next to it as a companion, not a substitute.
The other two normalizations, and CD3+CD45+, are **secondary** — they can and do degrade
to a skip when counts are missing.

Naming one primary endpoint in advance is what keeps this from being a search across six
panels for the smallest p-value.

```{r immune-mmrd-test}
mmrd_arcsin_t_wilcox <- function(df, metric_, denom_) {
  d <- df |> dplyr::filter(metric == metric_, denom_label == denom_, is.finite(frac))
  tab <- table(d$aneuploidy_class)
  if (length(tab) < 2 || any(tab < 2)) {
    message("MMRd ", metric_, " / ", denom_, ": insufficient n per aneuploidy arm — skipping.")
    return(invisible(NULL))
  }
  t_res <- t.test(arcsin_sqrt(frac) ~ aneuploidy_class, data = d)     # primary
  w_res <- wilcox.test(frac ~ aneuploidy_class, data = d)             # companion
  cat(sprintf("MMRd %-10s / %-20s  n=%d (high=%d, low=%d)  arcsin-t p=%.3g  Wilcoxon p=%.3g\n",
      metric_, denom_, nrow(d), tab[["aneuploidy-high"]], tab[["aneuploidy-low"]],
      t_res$p.value, w_res$p.value))
  list(d = d, t = t_res, w = w_res)
}
res_primary <- mmrd_arcsin_t_wilcox(imm_metrics, "CD45+", "all cells inside")      # PRIMARY
res_tumor   <- mmrd_arcsin_t_wilcox(imm_metrics, "CD45+", "tumour cells inside")   # secondary
res_cd45n   <- mmrd_arcsin_t_wilcox(imm_metrics, "CD3+CD45+", "CD45+ cells inside") # secondary
```

The primary contrast on its own. The subtitle carries both p-values and the group size —
at this n a single patient moving between groups can change the result, which is why the
count is on the figure rather than in a caption.

```{r immune-mmrd-box, fig.width=6, fig.height=5}
if (!is.null(res_primary)) {
  d <- res_primary$d
  ggplot(d, aes(aneuploidy_class, frac, fill = aneuploidy_class)) +
    attend_box(d, x = "aneuploidy_class", y = "frac", fill = NULL, points = FALSE) +
    highlight_points(d, "aneuploidy_class", "frac", base_size = 1.2, base_alpha = 0.5, base_const = "black") +
    stat_compare_means(method = "wilcox.test", label = "p.format", size = 3) +
    scale_fill_manual(values = attend_aneu_cols) +
    labs(title = "PRIMARY: CD45+ immune fraction (all cells inside) — MMR-deficient only",
         subtitle = paste0("arcsin(sqrt) t p = ", signif(res_primary$t$p.value, 3),
                           " | Wilcoxon p = ", signif(res_primary$w$p.value, 3),
                           " (n = ", nrow(d), "; small-n — suggestive only)"),
         x = NULL, y = "CD45+ fraction of all cells inside") +
    attend_theme() + theme(legend.position = "none")
} else message("Primary within-MMRd immune contrast skipped — see message above.")
```

#### The one model that needs MMR-proficient patients

The single exception to this report's MMRd restriction, and it has to be: an interaction
asks whether the aneuploidy effect *differs between* MMRd and MMRp, which cannot be
estimated inside one stratum. The model is
arcsin(sqrt(CD45+/all cells inside)) ~ aneuploidy_class * mmr_group, fitted on both
groups from imm_all. The interaction row is the one of interest: a large coefficient
there is what justifies restricting everything else to MMRd. No correction is applied.

```{r immune-interaction}
imm_int <- imm_all |>
  dplyr::filter(metric == "CD45+", denom_label == "all cells inside", is.finite(frac)) |>
  dplyr::left_join(master_grp |> dplyr::select(pid, mmr_group), by = "pid") |>
  dplyr::filter(!is.na(mmr_group), !is.na(aneuploidy_class)) |>
  dplyr::mutate(y = arcsin_sqrt(frac))

if (nrow(imm_int) > 4 && dplyr::n_distinct(imm_int$aneuploidy_class) == 2 &&
    dplyr::n_distinct(imm_int$mmr_group) == 2) {
  fit <- lm(y ~ aneuploidy_class * mmr_group, data = imm_int)
  coef_tbl <- if (requireNamespace("broom", quietly = TRUE)) broom::tidy(fit) else
    as.data.frame(summary(fit)$coefficients)
  print(knitr::kable(coef_tbl, digits = 3,
        caption = "arcsin(sqrt(CD45+/all-cells-inside)) ~ aneuploidy_class * mmr_group — BOTH MMR groups, no correction"))
} else message("Not enough patients across both MMR groups for the interaction model.")
```

### Aneuploidy against the continuous covariates

Aneuploidy score against TMB (both definitions), HRD, and MSI, within MMRd. One
point per patient, coloured by aneuploidy class, with Spearman's rho per panel. Spearman
rather than Pearson because these scores are bounded and skewed.

The two TMB panels are the same patients under the two definitions, so a visible
difference between them is the germline correction moving values, not new data.

```{r aneu-scatter, fig.width=12, fig.height=7}
scat <- pt |>
  pivot_longer(all_of(covar_cols04), names_to = "covariate", values_to = "value") |>
  filter(!is.na(aneuploidy_score), !is.na(value)) |>
  mutate(covariate = factor(covariate, levels = covar_cols04))

if (nrow(scat) > 2) {
  rho <- scat |>
    group_by(covariate) |>
    summarise(label = paste0("rho = ",
                round(suppressWarnings(cor(aneuploidy_score, value, method = "spearman",
                                           use = "complete.obs")), 2)), .groups = "drop")
  ggplot(scat, aes(aneuploidy_score, value, colour = aneuploidy_class)) +
    highlight_points(scat, "aneuploidy_score", "value", base_colour = "aneuploidy_class",
                     base_palette = attend_aneu_cols,
                     base_name = "aneuploidy", jitter_width = 0, base_size = 1.5, base_alpha = 0.6) +
    geom_smooth(method = "lm", se = FALSE, colour = "grey40", linewidth = 0.6) +
    geom_text(data = rho, aes(x = Inf, y = Inf, label = label),
              inherit.aes = FALSE, hjust = 1.05, vjust = 1.5, size = 3) +
    facet_wrap(~ covariate, scales = "free_y", ncol = 3) +
    labs(title = "Aneuploidy score vs continuous covariates (MMR-deficient only)",
         x = "aneuploidy score", y = "covariate value", colour = NULL) +
    attend_theme() + theme(legend.position = "top")
} else message("Not enough MMR-deficient patients with aneuploidy score + a covariate.")
```

The same covariates against the **binarised** aneuploidy class rather than the continuous
score — the categorical counterpart to the scatter above, with a per-covariate Wilcoxon.
Comparing the two is a check on the `r attend_thresholds$aneuploidy` cutpoint: a
correlation visible in the scatter but absent here means the cutpoint is discarding it.

```{r aneu-class-box, fig.width=12, fig.height=7}
cb <- pt |>
  pivot_longer(all_of(covar_cols04), names_to = "covariate", values_to = "value") |>
  filter(!is.na(aneuploidy_class), !is.na(value)) |>
  mutate(covariate = factor(covariate, levels = covar_cols04))

if (nrow(cb) > 2) {
  ggplot(cb, aes(aneuploidy_class, value, fill = aneuploidy_class)) +
    attend_box(cb, x = "aneuploidy_class", y = "value", by = "covariate", fill = NULL, points = FALSE) +
    highlight_points(cb, "aneuploidy_class", "value", base_size = 0.6, base_alpha = 0.4,
                     base_const = "black") +
    stat_compare_means(method = "wilcox.test", label = "p.format", size = 3) +
    facet_wrap(~ covariate, scales = "free_y", ncol = 3) +
    scale_fill_manual(values = attend_aneu_cols) +
    labs(title = "Continuous covariates by aneuploidy class (MMR-deficient only)",
         x = NULL, y = "covariate value") +
    attend_theme() + theme(legend.position = "none")
} else message("Not enough MMR-deficient patients with aneuploidy class + a covariate.")
```

### Aneuploidy by TP53 status

Within MMRd, aneuploidy score in TP53-abnormal versus TP53-normal tumours. TP53 status is
TP53_pathogenic from the master, derived from the MAF in report 01. TP53 loss is the
canonical driver of chromosomal instability, so this is the expected-positive control for
the aneuploidy score itself: if aneuploidy does not separate on TP53 here, the score is
suspect before any of the immune results are interpreted.

```{r mmrd-tp53, fig.width=5, fig.height=5}
tp <- pt |> filter(!is.na(TP53), !is.na(aneuploidy_score))

if (nrow(tp) > 2 && dplyr::n_distinct(tp$TP53) == 2) {
  ggplot(tp, aes(TP53, aneuploidy_score, fill = TP53)) +
    attend_box(tp, x = "TP53", y = "aneuploidy_score", fill = NULL, points = FALSE) +
    highlight_points(tp, "TP53", "aneuploidy_score", base_size = 1.5, base_alpha = 0.5,
                     base_const = "black") +
    stat_compare_means(method = "wilcox.test", label = "p.format") +
    scale_fill_manual(values = attend_tp53_cols) +
    labs(title = "Aneuploidy by TP53 status (MMR-deficient only)",
         x = NULL, y = "aneuploidy score") +
    attend_theme() + theme(legend.position = "none")
} else {
  message("Not enough MMR-deficient patients with both TP53 classes, or TP53_pathogenic ",
          "is missing from the master (rebuild report 01).")
}
```

### Aneuploidy against the pathologist's imaging calls

The pathologist's per-patient scores compared between aneuploidy high and low, within
MMRd. These come from the imaging sheet keyed by image → pid, averaged over a patient's
images for the continuous measures and taking the first non-missing value for the
categorical calls.

```{r imaging-covars}
first_nonNA <- function(x) { x <- x[!is.na(x)]; if (length(x)) x[1] else NA }

imaging_pt <- load_imaging_data() |>
  filter(!is.na(ID)) |>
  group_by(pid = as.character(ID)) |>
  summarise(
    Immune_infiltrate      = mean(suppressWarnings(as.numeric(Immune_infiltrate)),     na.rm = TRUE),
    Immune_infiltrate_PD1  = mean(suppressWarnings(as.numeric(Immune_infiltrate_PD1)), na.rm = TRUE),
    ARID1A                 = first_nonNA(ARID1A),
    Tumor_cells_PDL1       = first_nonNA(Tumor_cells_PDL1),
    Immune_infiltrate_PDL1 = first_nonNA(Immune_infiltrate_PDL1),
    .groups = "drop"
  ) |>
  left_join(master |> select(pid, aneuploidy_class), by = "pid") |>
  filter(!is.na(aneuploidy_class), pid %in% mmrd_pids)   # <- the MMRd restriction

cat("MMR-deficient patients with imaging covariates:", n_distinct(imaging_pt$pid), "\n")
```

Continuous pathologist scores (immune infiltrate and PD-1 infiltrate, as percentages) by
aneuploidy class, Wilcoxon per panel. These are the pathologist's independent read of the
same immune question the FlowPath contrast above asked.

```{r imaging-covars-cont, fig.width=8, fig.height=4}
icc <- imaging_pt |>
  pivot_longer(c(Immune_infiltrate, Immune_infiltrate_PD1),
               names_to = "covariate", values_to = "value") |>
  filter(is.finite(value))

if (nrow(icc) > 2) {
  ggplot(icc, aes(aneuploidy_class, value, fill = aneuploidy_class)) +
    attend_box(icc, x = "aneuploidy_class", y = "value", by = "covariate", fill = NULL, points = FALSE) +
    highlight_points(icc, "aneuploidy_class", "value", base_size = 0.6, base_alpha = 0.4,
                     base_const = "black") +
    stat_compare_means(method = "wilcox.test", label = "p.format", size = 3) +
    facet_wrap(~ covariate, scales = "free_y") +
    scale_fill_manual(values = attend_aneu_cols) +
    labs(title = "Pathologist immune scores by aneuploidy class (MMR-deficient only)",
         x = NULL, y = "pathologist %") +
    attend_theme() + theme(legend.position = "none")
}
```

Categorical calls as stacked proportions — how each call distributes within aneuploidy
high vs low. Bars are proportions, so the **group sizes are not visible in this figure**;
read them from the count printed above. MMR_Status is omitted because it is constant
here by construction.

```{r imaging-covars-cat, fig.width=11, fig.height=4}
icat <- imaging_pt |>
  select(aneuploidy_class, ARID1A, Tumor_cells_PDL1, Immune_infiltrate_PDL1) |>
  pivot_longer(-aneuploidy_class, names_to = "covariate", values_to = "level") |>
  filter(!is.na(level), level != "")

if (nrow(icat) > 0) {
  ggplot(icat, aes(aneuploidy_class, fill = level)) +
    geom_bar(position = "fill") +
    facet_wrap(~ covariate, scales = "free_x", nrow = 1) +
    labs(title = "Pathologist categorical calls by aneuploidy class (MMR-deficient only)",
         x = NULL, y = "proportion of patients", fill = NULL) +
    attend_theme() +
    theme(axis.text.x = element_text(angle = 20, hjust = 1))
}
```

```{r session-info}
sessionInfo()
```
