---
title: "ATTEND — 32. Treatment response and cell-type composition"
author: "sceriff0"
date: "2026-08-13"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
# renv: tidyverse, here, ggpubr.
library(tidyverse)
library(here)
library(ggpubr)
# Headless-safe raster device. The cluster's compute nodes have no X11, so R's default
# png() fails with "unable to open connection to X11 display" and every figure dies. cairo
# needs no display; the guard keeps this working on a machine built without it.
if (capabilities("cairo")) options(bitmapType = "cairo")

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 10, fig.height = 6)

source(here("code", "build_master.R"))       # get_master() — builds on demand, no build order
source(here("code", "attend_classes.R"))     # add_response_class(), response_counts()
source(here("code", "attend_harmonise.R"))   # build_barcode_pid(), make_id_cfg()
source(here("code", "attend_plots.R"))       # highlight_points(), attend_theme(), attend_resp_cols
source(here("code", "attend_ihc.R"))         # ihc_celltype_metrics(), ihc_immune_metrics()
source(here("code", "load_clinical.R"))      # load_imaging_data()
source(here("code", "load_phenotypes.R"))    # load_ihc_celltypes()

master     <- get_master() |> add_molecular_classes() |> add_response_class()
master_grp <- add_scna_group(master)         # adds mmr_group ("MMRd"/"MMRp")
mmrd_pids  <- master_grp |> filter(mmr_group == "MMRd") |> pull(pid)

resp_cut  <- attend_thresholds$response_months
n_resp    <- sum(master$response_class == "responder",     na.rm = TRUE)
n_nonresp <- sum(master$response_class == "non-responder", na.rm = TRUE)
n_class   <- n_resp + n_nonresp
n_lost    <- sum(is.na(master$response_class))
n_mmrd_cl <- sum(master$pid %in% mmrd_pids & !is.na(master$response_class))

# The polipo highlight (21S188) is drawn here, as it is on every per-patient figure in the
# pipeline -- there is no per-report opt-out any more. Its ids are resolved through the
# barcode -> pid crosswalk by highlight_group_of(), so the overlay lands on this report's
# pid-keyed frames; before that resolution existed it matched nothing here and drew nothing.
```

> **Question.** Does cell-type composition differ between responders and non-responders, and is the answer inside MMR-deficient tumours the same as across the whole cohort?
> **Cohort.** Two, drawn side by side in every figure: all classifiable patients (n = `r n_class`) and the MMR-deficient subset of them (n = `r n_mmrd_cl`).
> **Reads.** The master (built on demand by get_master()), the FlowPath annotation tree, and the survival endpoint the master carries.
> **Out of scope.** Survival modelling itself -> [04-survival](04-survival.html). The same composition panels keyed on aneuploidy -> [05-aneuploidy-mmrd](05-aneuploidy-mmrd.html).

This report is the response-keyed counterpart of Part 2 of
[05-aneuploidy-mmrd](05-aneuploidy-mmrd.html). Every panel, normalization and test is the
one used there; the only thing that changes is the grouping variable on the x-axis, which
is treatment response rather than aneuploidy class. Keeping the two reports structurally
identical is deliberate — it is what makes the two groupings comparable rather than two
separate analyses that happen to use the same cells.

Both cohorts are drawn for every figure, always in the same order (all patients first,
MMR-deficient second). Report 05 restricts to MMRd because MMR deficiency is itself a route
to genomic instability and would otherwise confound an aneuploidy contrast. Response has no
such built-in confound with MMR status, so the whole cohort is the primary cohort here and
the MMRd panel is the stratified check.

## Defining response

Response is a **landmark** on the survival endpoint the pipeline is already configured
with: `attend_cols$surv_time` / `attend_cols$surv_event`, which currently point at
progression-free survival in months. A patient still progression-free past
`r resp_cut` months is a **responder**; one who progressed or died at or before
`r resp_cut` months is a **non-responder**. The cut lives in
`attend_thresholds$response_months`, so changing it changes every figure below and nothing
else, and repointing `attend_cols$surv_time` at overall survival changes what "response"
means here without touching this report.

The third group is the one that decides whether the definition is honest. A patient whose
follow-up simply stopped before the landmark — censored, no event recorded — has no
observable response status at all. Counting them as non-responders would score lost
follow-up as treatment failure and inflate the responder arm by comparison. `add_response_class()`
returns them as NA instead, and the table below is where they are reported rather than
quietly dropped: `r n_lost` of `r nrow(master)` patients cannot be classified.

An event *after* the landmark does not demote a responder. The question is what had
happened by month `r resp_cut`, not eventual outcome, so this is a binary contrast and not
a survival model — the survival models themselves are all in
[04-survival](04-survival.html).

### How the cohort splits

Every patient on the master, sorted into the two response arms and the two reasons a
patient can fail to get an arm. The bottom two rows are the denominator this report cannot
speak about; if they dominate, every p-value below is describing a minority of the cohort.

```{r response-counts}
have_resp <- have_cols(master, attend_cols$surv_time, attend_cols$surv_event)

knitr::kable(response_counts(master),
             caption = paste0("Response classification across the whole master (landmark = ",
                              resp_cut, " months on ", attend_cols$surv_time, ")"))

if (!have_resp)
  message("Survival columns are still placeholders in attend_cols — every figure below ",
          "is skipped. Set surv_time / surv_event and re-knit.")
```

The same split inside the MMR-deficient stratum, which is the second cohort every figure
draws. MMR status comes from `add_scna_group()`, so it is the same `is_mmrd()` call the
copy-number and classification reports use.

```{r response-counts-mmrd}
resp_mmrd <- master |> filter(pid %in% mmrd_pids)
knitr::kable(response_counts(resp_mmrd),
             caption = "Response classification within the MMR-deficient stratum")

if (nrow(resp_mmrd) > 0)
  cat("MMR-deficient patients:", nrow(resp_mmrd),
      "| classifiable:", sum(!is.na(resp_mmrd$response_class)), "\n")
```

Coverage of the manual highlight overlay on the response-classifiable cohort. The one
configured group is `polipo` (sample 21S188, blue); the count is scored after the barcode ->
pid crosswalk resolves it onto this pid-keyed frame, so `1/1 matched` means the overlay is
really drawn. A `0/1` here means 21S188 has no observable 6-month response status — read it
against the four-reason partition above, not as a plotting failure.

```{r highlight-coverage}
cat(highlight_coverage(master |> filter(!is.na(response_class))))
```

## Cell-type composition by response

"Cell type" is `phenotype_clean`, the parenthetical phenotype label in the FlowPath export
(changeable in `code/load_phenotypes.R`). Every count is taken inside the annotated region
and expressed as a fraction, but the denominator changes the question, so all three are
shown — this is the same three-way normalization report 05 uses:

| Normalization | Denominator | The question it answers |
|---|---|---|
| **all cells inside** | every cell in the annotation | what fraction of the tissue is this cell type? |
| **tumour cells inside** | tumour cells only | how much of this cell type per unit of tumour? |
| **CD45+ cells inside** | leukocytes only | what is the composition *of the immune compartment*? |

Each is drawn twice per cohort: the raw fraction with a Wilcoxon rank-sum test, and the
arcsine-square-root-transformed fraction with a two-sample t-test. Agreement between the two
is reassurance; disagreement means the result depends on the test rather than on the data.
No multiple-testing correction is applied anywhere in this report, and with one test per
cell type per normalization per cohort that is many comparisons, so every p-value here is
descriptive.

### Building the two cohorts

Per-patient cell-type composition is read from the FlowPath tree by
`ihc_celltype_metrics()` (which pools a patient's images by summing counts and denominators)
and joined on pid to the response class derived above. `ct_all` keeps every classifiable
patient; `ct_mmrd` is the same table restricted to the MMR-deficient pids. Patients with no
response class are dropped here, which is why the counts below can be smaller than the
imaged cohort.

```{r composition-data}
ct_all <- ihc_celltype_metrics(load_ihc_celltypes(), load_imaging_data()) |>
  left_join(master |> select(pid, response_class), by = "pid") |>
  filter(!is.na(response_class))
ct_mmrd <- ct_all |> filter(pid %in% mmrd_pids)

ct_cohorts <- list(`all patients` = ct_all, `MMR-deficient` = ct_mmrd)

cat("patients with cell-type metrics + a response class:", n_distinct(ct_all$pid), "\n")
cat("  ... of them MMR-deficient                       :", n_distinct(ct_mmrd$pid), "\n")
cat("cell types measured                               :", n_distinct(ct_all$cell_type), "\n\n")
# distinct(pid, ...) first: ct is long (one row per patient x cell type), so tabulating it
# directly would count every patient once per cell type. The cohort name is a caption and
# NOT a table() argument -- table() cross-tabulates its arguments and requires them to be
# equal length, so a length-1 label next to the vector is an error, not a heading.
for (nm in names(ct_cohorts)) {
  d <- ct_cohorts[[nm]] |> distinct(pid, response_class)
  cat("\n", nm, " (n = ", nrow(d), " patients):\n", sep = "")
  if (nrow(d) > 0) print(table(response = droplevels(d$response_class)))
}
```

```{r box-helpers, include=FALSE}
# One boxplot per cell type, responder vs non-responder. Boxes are filled by response
# class; the highlight overlay is drawn on top in its own colour scale (colour != fill, so
# the two legends do not collide), partitioned so no patient is plotted twice. `arcsin`
# switches BOTH the y variable and the test, so the raw/Wilcoxon and arcsin/t-test versions
# can never drift apart the way two copied functions would.
box_by_resp <- function(df, frac_col, denom_label, cohort_label, arcsin = FALSE) {
  d <- df |> filter(is.finite(.data[[frac_col]]), !is.na(response_class)) |>
    mutate(.y = if (arcsin) arcsin_sqrt(.data[[frac_col]]) else .data[[frac_col]]) |>
    droplevels()
  # stat_compare_means() errors on a single-level grouping, so both arms must be populated.
  if (nrow(d) == 0 || n_distinct(d$response_class) < 2) return(NULL)
  method <- if (arcsin) "t.test"                             else "wilcox.test"
  ylab   <- if (arcsin) paste0("arcsin(sqrt(fraction of ", denom_label, "))")
            else        paste0("fraction of ", denom_label)
  sub    <- if (arcsin) "arcsine-square-root transformed fraction; two-sample t-test"
            else        "raw fraction; Wilcoxon rank-sum"
  ggplot(d, aes(response_class, .y, fill = response_class)) +
    attend_box(d, x = "response_class", y = ".y", by = "cell_type", fill = NULL, points = FALSE) +
    highlight_points(d, "response_class", ".y", base_size = 0.5, base_alpha = 0.35,
                     base_const = "black") +
    facet_wrap(~ cell_type, scales = "free_y") +
    stat_compare_means(method = method, label = "p.format", size = 3) +
    scale_fill_manual(values = attend_resp_cols) +
    labs(title = paste0("Cell-type presence by response (", cohort_label,
                        ") — normalized by ", denom_label),
         subtitle = sub, x = NULL, y = ylab) +
    attend_theme() +
    theme(legend.position = "none", axis.text.x = element_text(angle = 20, hjust = 1))
}

# Draw one normalization for BOTH cohorts, in a fixed order, so the two are always adjacent
# in the knitted page and a cohort that degrades to a skip says so instead of vanishing.
draw_both <- function(frac_col, denom_label, arcsin = FALSE) {
  for (nm in names(ct_cohorts)) {
    p <- box_by_resp(ct_cohorts[[nm]], frac_col, denom_label, nm, arcsin)
    if (is.null(p)) message("Skipped: ", nm, " / ", denom_label,
                            " — no finite fractions, or only one response arm present.")
    else print(p)
  }
}
```

### Normalized by all cells inside

Each cell type as a fraction of every cell in the annotation, so a panel reads "how much of
the tissue is this". Raw fractions with a Wilcoxon, drawn for all patients and then for the
MMR-deficient subset.

```{r box-inside, fig.width=15, fig.height=11, out.width="100%"}
draw_both("frac_inside", "all cells inside")
```

The same panels on the arcsine-square-root scale with a two-sample t-test, which is the
variance-stabilised companion rather than an alternative answer.

```{r box-inside-arcsin, fig.width=15, fig.height=11, out.width="100%"}
draw_both("frac_inside", "all cells inside", arcsin = TRUE)
```

### Normalized by tumour cells inside

Each cell type per unit of tumour: the denominator is tumour cells only, so a shift here
survives a change in how much tumour the annotation caught. Raw fractions, Wilcoxon, both
cohorts.

```{r box-tumor, fig.width=15, fig.height=11, out.width="100%"}
draw_both("frac_tumor", "tumour cells inside")
```

The arcsine/t-test companion of the panels above.

```{r box-tumor-arcsin, fig.width=15, fig.height=11, out.width="100%"}
draw_both("frac_tumor", "tumour cells inside", arcsin = TRUE)
```

### Normalized by CD45+ cells inside

Composition *of the immune compartment*: the denominator is leukocytes, so differences in
tumour content between patients cannot drive the contrast. Raw fractions, Wilcoxon, both
cohorts.

```{r box-cd45, fig.width=15, fig.height=11, out.width="100%"}
draw_both("frac_cd45", "CD45+ cells inside")
```

The arcsine/t-test companion of the panels above.

```{r box-cd45-arcsin, fig.width=15, fig.height=11, out.width="100%"}
draw_both("frac_cd45", "CD45+ cells inside", arcsin = TRUE)
```

## Immune content — the headline contrast

Two leukocyte metrics counted inside the annotation by `ihc_immune_metrics()` — **CD45+
total** (all leukocytes) and **CD3+CD45+** (double-positive T cells) — under each
normalization, joined to the response class on pid. CD3+CD45+ requires a `CD3_sign` column
in the FlowPath export; without it those rows are NA and their facets drop out on their own.

```{r immune-data}
imm_all <- ihc_immune_metrics(load_ihc_celltypes(), load_imaging_data()) |>
  left_join(master |> select(pid, response_class), by = "pid") |>
  filter(!is.na(response_class))
imm_mmrd <- imm_all |> filter(pid %in% mmrd_pids)

imm_cohorts <- list(`all patients` = imm_all, `MMR-deficient` = imm_mmrd)

cat("patients with immune metrics + a response class:", n_distinct(imm_all$pid),
    "| of them MMR-deficient:", n_distinct(imm_mmrd$pid), "\n")
cat("metrics present:", paste(sort(unique(imm_all$metric)), collapse = ", "), "\n")
```

```{r immune-box-helper, include=FALSE}
# Same construction as box_by_resp(), faceted by metric x normalization instead of by
# cell type, so all six immune panels sit on one figure per cohort.
box_immune_resp <- function(df, cohort_label, arcsin = FALSE) {
  d <- df |> filter(is.finite(frac), !is.na(response_class)) |> droplevels()
  if (nrow(d) == 0 || n_distinct(d$response_class) < 2) return(NULL)
  if (arcsin) d <- d |> mutate(frac = arcsin_sqrt(frac))
  method <- if (arcsin) "t.test"                 else "wilcox.test"
  ylab   <- if (arcsin) "arcsin(sqrt(fraction))" else "fraction"
  ttl    <- if (arcsin) "arcsin(sqrt) + t-test"  else "raw fraction + Wilcoxon"
  ggplot(d, aes(response_class, frac, fill = response_class)) +
    attend_box(d, x = "response_class", y = "frac", by = c("metric", "denom_label"), fill = NULL, points = FALSE) +
    highlight_points(d, "response_class", "frac", base_size = 0.6, base_alpha = 0.4,
                     base_const = "black") +
    facet_wrap(~ metric + denom_label, scales = "free_y") +
    stat_compare_means(method = method, label = "p.format", size = 3) +
    scale_fill_manual(values = attend_resp_cols) +
    labs(title = paste0("Immune content by response (", cohort_label, ") — ", ttl),
         x = NULL, y = ylab) +
    attend_theme() +
    theme(legend.position = "none", axis.text.x = element_text(angle = 20, hjust = 1))
}
```

CD45+ total and CD3+CD45+ fractions by response class, one facet per metric ×
normalization, raw then arcsine, for each cohort in turn.

```{r immune-box, fig.width=12, fig.height=8, out.width="100%"}
for (nm in names(imm_cohorts)) {
  for (tr in c(FALSE, TRUE)) {
    p <- box_immune_resp(imm_cohorts[[nm]], nm, arcsin = tr)
    if (is.null(p)) message("Skipped immune panel: ", nm, if (tr) " (arcsin)" else " (raw)",
                            " — no finite metrics, or only one response arm present.")
    else print(p)
  }
}
```

### The pre-specified primary contrast

Of everything above, one comparison is designated **primary**: **CD45+ / all cells inside**,
tested with the **arcsine-square-root transform and a two-sample t-test**, on the **whole
cohort**. The Wilcoxon on the raw fraction is reported next to it as a companion, not a
substitute, and the MMR-deficient version of the same contrast is **secondary** — it is the
stratified check, and it can and does degrade to a skip when counts are missing.

Naming one primary endpoint in advance is what keeps this from being a search across
twelve panels for the smallest p-value. It mirrors the primary contrast in
[05-aneuploidy-mmrd](05-aneuploidy-mmrd.html), which is the same metric under the same test
against the other grouping.

```{r immune-primary-test}
resp_arcsin_t_wilcox <- function(df, cohort_label, metric_, denom_) {
  d <- df |> dplyr::filter(metric == metric_, denom_label == denom_, is.finite(frac),
                           !is.na(response_class)) |>
    droplevels()
  tab <- table(d$response_class)
  if (length(tab) < 2 || any(tab < 2)) {
    message(cohort_label, " ", metric_, " / ", denom_,
            ": insufficient n per response arm — skipping.")
    return(invisible(NULL))
  }
  t_res <- t.test(arcsin_sqrt(frac) ~ response_class, data = d)     # primary
  w_res <- wilcox.test(frac ~ response_class, data = d)             # companion
  cat(sprintf("%-14s %-10s / %-20s  n=%d (resp=%d, non=%d)  arcsin-t p=%.3g  Wilcoxon p=%.3g\n",
      cohort_label, metric_, denom_, nrow(d), tab[["responder"]], tab[["non-responder"]],
      t_res$p.value, w_res$p.value))
  list(d = d, t = t_res, w = w_res)
}
res_primary <- resp_arcsin_t_wilcox(imm_all,  "all patients",  "CD45+", "all cells inside")
res_mmrd    <- resp_arcsin_t_wilcox(imm_mmrd, "MMR-deficient", "CD45+", "all cells inside")
res_cd3     <- resp_arcsin_t_wilcox(imm_all,  "all patients",  "CD3+CD45+", "CD45+ cells inside")
```

The primary contrast on its own. The subtitle carries both p-values and the group size — at
this n a single patient moving between arms can change the result, which is why the count is
on the figure rather than in a caption.

```{r immune-primary-box, fig.width=6, fig.height=5}
if (!is.null(res_primary)) {
  d <- res_primary$d
  ggplot(d, aes(response_class, frac, fill = response_class)) +
    attend_box(d, x = "response_class", y = "frac", fill = NULL, points = FALSE) +
    highlight_points(d, "response_class", "frac", base_size = 1.2, base_alpha = 0.5,
                     base_const = "black") +
    stat_compare_means(method = "wilcox.test", label = "p.format", size = 3) +
    scale_fill_manual(values = attend_resp_cols) +
    labs(title = "PRIMARY: CD45+ immune fraction (all cells inside) by response",
         subtitle = paste0("landmark ", resp_cut, " mo | arcsin(sqrt) t p = ",
                           signif(res_primary$t$p.value, 3),
                           " | Wilcoxon p = ", signif(res_primary$w$p.value, 3),
                           " (n = ", nrow(d), "; small-n — suggestive only)"),
         x = NULL, y = "CD45+ fraction of all cells inside") +
    attend_theme() + theme(legend.position = "none")
} else message("Primary immune contrast skipped — see message above.")
```

### Does the response effect differ between MMR strata

The two cohorts above are drawn side by side but never compared to each other, and looking
at two p-values is not a test of whether they differ. This model is:
`arcsin(sqrt(CD45+/all cells inside)) ~ response_class * mmr_group`, fitted on both MMR
groups at once from `imm_all`. The interaction row is the one of interest — a large
coefficient there means the response effect is not the same in MMRd and MMRp, which is what
would justify reading the stratified panels separately rather than as one result. No
correction is applied, and at this cohort size the interaction is expected to be
underpowered.

```{r immune-interaction}
imm_int <- imm_all |>
  dplyr::filter(metric == "CD45+", denom_label == "all cells inside", is.finite(frac)) |>
  dplyr::left_join(master_grp |> dplyr::select(pid, mmr_group), by = "pid") |>
  dplyr::filter(!is.na(mmr_group), !is.na(response_class)) |>
  dplyr::mutate(y = arcsin_sqrt(frac)) |>
  droplevels()

if (nrow(imm_int) > 4 && dplyr::n_distinct(imm_int$response_class) == 2 &&
    dplyr::n_distinct(imm_int$mmr_group) == 2) {
  fit <- lm(y ~ response_class * mmr_group, data = imm_int)
  coef_tbl <- if (requireNamespace("broom", quietly = TRUE)) broom::tidy(fit) else
    as.data.frame(summary(fit)$coefficients)
  print(knitr::kable(coef_tbl, digits = 3,
        caption = "arcsin(sqrt(CD45+/all-cells-inside)) ~ response_class * mmr_group — BOTH MMR groups, no correction"))
} else message("Not enough patients across both MMR groups and both response arms for the interaction model.")
```

## Response against the molecular covariates

The aneuploidy score, both TMB definitions, HRD and MSI, compared between response arms.
All are numeric columns on the master (`attend_cols`), read straight from it — the two TMB
columns are added under their reader-facing labels so a facet never shows an internal column
name. This is the check on whether response is confounded with the molecular axes the rest
of the pipeline splits on.

```{r covariate-data}
numcol <- function(col) if (col %in% names(master)) suppressWarnings(as.numeric(master[[col]])) else NA_real_

pt_all <- tibble(
  pid              = master$pid,
  response_class   = master$response_class,
  aneuploidy_class = master$aneuploidy_class,
  aneuploidy_score = numcol(attend_cols$aneuploidy),
  HRD              = numcol(attend_cols$hrd),
  MSI              = numcol("msi__MSI_SCORE")
)

tmb_defs06 <- tmb_defs_present(master)
tmb_labs06 <- tmb_def_labels(tmb_defs06)
for (i in seq_along(tmb_defs06)) pt_all[[tmb_labs06[i]]] <- numcol(tmb_defs06[[i]])
covar_cols06 <- c("aneuploidy_score", tmb_labs06, "HRD", "MSI")

pt_cohorts <- list(`all patients`  = pt_all |> filter(!is.na(response_class)),
                   `MMR-deficient` = pt_all |> filter(!is.na(response_class), pid %in% mmrd_pids))

cat("covariates plotted:", paste(covar_cols06, collapse = " | "), "\n")
cat("classifiable patients — all:", nrow(pt_cohorts[["all patients"]]),
    "| MMRd:", nrow(pt_cohorts[["MMR-deficient"]]), "\n")
```

One panel per covariate, responder vs non-responder, with a per-panel Wilcoxon; both
cohorts in turn. The two TMB panels are the same patients under the two definitions, so a
visible difference between them is the ancestry correction moving values, not new data.

```{r covariate-box, fig.width=12, fig.height=7}
for (nm in names(pt_cohorts)) {
  cb <- pt_cohorts[[nm]] |>
    pivot_longer(all_of(covar_cols06), names_to = "covariate", values_to = "value") |>
    filter(!is.na(value)) |>
    mutate(covariate = factor(covariate, levels = covar_cols06)) |>
    droplevels()
  if (nrow(cb) > 2 && n_distinct(cb$response_class) == 2) {
    print(
      ggplot(cb, aes(response_class, value, fill = response_class)) +
        attend_box(cb, x = "response_class", y = "value", by = "covariate", fill = NULL, points = FALSE) +
        highlight_points(cb, "response_class", "value", base_size = 0.6, base_alpha = 0.4,
                         base_const = "black") +
        stat_compare_means(method = "wilcox.test", label = "p.format", size = 3) +
        facet_wrap(~ covariate, scales = "free_y", ncol = 3) +
        scale_fill_manual(values = attend_resp_cols) +
        labs(title = paste0("Molecular covariates by response (", nm, ")"),
             x = NULL, y = "covariate value") +
        attend_theme() + theme(legend.position = "none"))
  } else message("Not enough ", nm, " with a response class and a covariate — skipping.")
}
```

### Response against aneuploidy class

The direct cross-tabulation of the two groupings — the one in
[05-aneuploidy-mmrd](05-aneuploidy-mmrd.html) against the one used here — with a Fisher
exact test on the 2×2. Everything in this report is drawn on the same cells report 05 draws,
so if these two splits largely coincide then the two reports are describing one contrast
under two names rather than two independent ones. The test is on patients, not images.

```{r response-vs-aneuploidy}
xt <- pt_all |> filter(!is.na(response_class), !is.na(aneuploidy_class)) |> droplevels()

if (nrow(xt) > 0 && n_distinct(xt$response_class) == 2 && n_distinct(xt$aneuploidy_class) == 2) {
  tab <- table(response = xt$response_class, aneuploidy = xt$aneuploidy_class)
  print(knitr::kable(as.data.frame.matrix(tab),
        caption = paste0("Response class vs aneuploidy class, all patients (n = ", nrow(xt), ")")))
  cat("Fisher exact p =", signif(fisher.test(tab)$p.value, 3), "\n")
} else message("Not enough patients with both a response class and an aneuploidy class.")
```

## Response against the pathologist's imaging calls

The pathologist's per-patient scores compared between response arms. These come from the
imaging sheet keyed by image -> pid, averaged over a patient's images for the continuous
measures and taking the first non-missing value for the categorical calls — the same
summarisation [05-aneuploidy-mmrd](05-aneuploidy-mmrd.html) applies before its aneuploidy
contrast.

```{r imaging-covars}
first_nonNA <- function(x) { x <- x[!is.na(x)]; if (length(x)) x[1] else NA }

imaging_pt <- load_imaging_data() |>
  filter(!is.na(ID)) |>
  group_by(pid = as.character(ID)) |>
  summarise(
    Immune_infiltrate      = mean(suppressWarnings(as.numeric(Immune_infiltrate)),     na.rm = TRUE),
    Immune_infiltrate_PD1  = mean(suppressWarnings(as.numeric(Immune_infiltrate_PD1)), na.rm = TRUE),
    ARID1A                 = first_nonNA(ARID1A),
    Tumor_cells_PDL1       = first_nonNA(Tumor_cells_PDL1),
    Immune_infiltrate_PDL1 = first_nonNA(Immune_infiltrate_PDL1),
    .groups = "drop"
  ) |>
  left_join(master |> select(pid, response_class), by = "pid") |>
  filter(!is.na(response_class))

cat("patients with imaging covariates + a response class:", n_distinct(imaging_pt$pid),
    "| of them MMR-deficient:", n_distinct(filter(imaging_pt, pid %in% mmrd_pids)$pid), "\n")
```

Continuous pathologist scores (immune infiltrate and PD-1 infiltrate, as percentages) by
response arm, Wilcoxon per panel. This is the pathologist's independent read of the same
immune question the FlowPath contrast above asked, on the same patients.

```{r imaging-covars-cont, fig.width=8, fig.height=4}
icc <- imaging_pt |>
  pivot_longer(c(Immune_infiltrate, Immune_infiltrate_PD1),
               names_to = "covariate", values_to = "value") |>
  filter(is.finite(value)) |> droplevels()

if (nrow(icc) > 2 && n_distinct(icc$response_class) == 2) {
  ggplot(icc, aes(response_class, value, fill = response_class)) +
    attend_box(icc, x = "response_class", y = "value", by = "covariate", fill = NULL, points = FALSE) +
    highlight_points(icc, "response_class", "value", base_size = 0.6, base_alpha = 0.4,
                     base_const = "black") +
    stat_compare_means(method = "wilcox.test", label = "p.format", size = 3) +
    facet_wrap(~ covariate, scales = "free_y") +
    scale_fill_manual(values = attend_resp_cols) +
    labs(title = "Pathologist immune scores by response class (all patients)",
         x = NULL, y = "pathologist %") +
    attend_theme() + theme(legend.position = "none")
} else message("Not enough patients with a pathologist score in both response arms.")
```

Categorical calls as stacked proportions — how each call distributes within the two
response arms. Bars are proportions, so the **group sizes are not visible in this figure**;
read them from the count printed above.

```{r imaging-covars-cat, fig.width=11, fig.height=4}
icat <- imaging_pt |>
  select(response_class, ARID1A, Tumor_cells_PDL1, Immune_infiltrate_PDL1) |>
  pivot_longer(-response_class, names_to = "covariate", values_to = "level") |>
  filter(!is.na(level), level != "") |> droplevels()

if (nrow(icat) > 0) {
  ggplot(icat, aes(response_class, fill = level)) +
    geom_bar(position = "fill") +
    facet_wrap(~ covariate, scales = "free_x", nrow = 1) +
    labs(title = "Pathologist categorical calls by response class (all patients)",
         x = NULL, y = "proportion of patients", fill = NULL) +
    attend_theme() +
    theme(axis.text.x = element_text(angle = 20, hjust = 1))
}
```

```{r session-info}
sessionInfo()
```
