---
title: "ATTEND — 06. Tumour mutational burden and the ancestry correction"
author: "sceriff0"
date: "2026-08-13"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
library(here)
# Headless-safe raster device. The cluster's compute nodes have no X11, so R's default
# png() fails with "unable to open connection to X11 display" and every figure dies. cairo
# needs no display; the guard keeps this working on a machine built without it.
if (capabilities("cairo")) options(bitmapType = "cairo")

# renv: tidyverse, here. Optional and guarded: diptest, mclust.
# The TCGA part needs data/tcga_2013/ — Rscript code/fetch_tcga_ucec_2013.R
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 10, fig.height = 7)

source(here("code", "build_master.R"))       # get_master()
source(here("code", "attend_bimodality.R"))  # tmb_long(), tmb_battery(), .gmm_modes()
source(here("code", "attend_plots.R"))       # attend_aneu_cols, attend_theme()
source(here("code", "load_wes_results.R"))   # load_tcga_ucec_2013()

master <- get_master()

# The same patients once per aneuploidy cut, with aneuploidy_class (and the MMR x aneuploidy
# group) RE-DERIVED at each, so Part 2's panels gain a facet dimension instead of a second
# copy of their code. The TMB-definition loop is a separate dimension and stays separate.
master_cuts <- add_aneuploidy_cuts(master, scna = TRUE)
cuts        <- aneu_cuts()

# Shared derivations, computed here from the master rather than read from a file another
# report had to write first. Same values, no build order.
long       <- tmb_long(master)
finite_n   <- attr(long, "finite_n")          # per definition, BEFORE the min_finite filter
def_cols   <- attr(long, "def_cols")          # key -> column, AFTER it
def_labels <- levels(long$definition)         # labels for def_cols, same order
```

> **Question.** Does TMB split the cohort into two populations, does that split track aneuploidy within each MMR stratum, and does it reproduce in a paired-sequencing cohort?
> **Cohort.** All `r nrow(master)` patients; Part 2 is stratified by MMR, Part 3 compares against TCGA UCEC.
> **Reads.** The master (built on demand), plus data/tcga_2013/ for Part 3.
> **Out of scope.** How the ancestry correction works, why there are exactly two TMB definitions, and why every aneuploidy-keyed figure is drawn at two cuts -> [00-methods](00-methods.html).

A genuine second mode at high TMB *is* the hypermutator population: MMR-deficient tumours
whose failed mismatch repair produces an order of magnitude more mutations. ATTEND has no
POLE cases, the other hypermutator route, so the question is whether TMB by itself
identifies the hypermutators.

Every distribution and every test is on the **raw mut/Mb scale**, with no log transform. On
that scale a long tail and a genuine second mode look alike, which is why Part 1 triangulates
three independent criteria rather than trusting any one of them.

## Part 1 — is the distribution split?

Four steps: describe each distribution, test for bimodality under three independent
criteria, validate any second mode against MMR status, and — in Part 3
below — reproduce the whole thing in TCGA UCEC, where the answer is externally
known.

Comparing the two definitions is a second question this report answers. Tumour-only TMB
carries ancestry-dependent germline inflation, which is noise added to the signal, so
removing it should deepen the valley between the modes. If the corrected definition is
more clearly bimodal than the uncorrected one, the correction is doing something real; if
not, it is not helping the split, and the battery below says so either way.

### Data

The tmb__* columns of the master, reshaped to one row per patient and definition.
Definitions with fewer than 10 finite values are dropped: the ancestry-corrected column is
all-NA until attend_ancestry$clinical_col is set, and a distribution test needs values.



### 1. Distribution overview

Per-definition summary of the raw TMB. %high is the fraction at or above the
pipeline TMB-high cut (attend_thresholds$tmb = `r attend_thresholds$tmb` mut/Mb),
via the shared tmb_class_of() — the same high/low split every TMB-keyed figure uses.

```{r summary-table}
if (nrow(long) > 0) {
  summ <- long |>
    group_by(definition) |>
    summarise(
      n       = dplyr::n(),
      median  = median(tmb),
      IQR     = IQR(tmb),
      mean    = mean(tmb),
      sd      = sd(tmb),
      min     = min(tmb),
      max     = max(tmb),
      skew    = unname(.skew_kurt(tmb)["skew"]),
      exkurt  = unname(.skew_kurt(tmb)["kurt"]),
      `%high` = 100 * mean(tmb_class_of(tmb) == "TMB-high"),
      .groups = "drop"
    )
  knitr::kable(summ, digits = 2,
               caption = "Raw TMB (mut/Mb) by definition. High right-skew (large positive `skew`) is expected and is itself a hint of a heavy hypermutator tail.")
} else {
  cat("No TMB values present — nothing to summarise.\n")
}
```

Kernel density of the raw mut/Mb values in long, one curve per definition, with the
`r attend_thresholds$tmb` mut/Mb cutoff marked. A valley between two humps is the shape the
battery below puts numbers on.

```{r density, fig.height=6}
if (nrow(long) > 0) {
  ggplot(long, aes(tmb)) +
    geom_histogram(aes(y = after_stat(density)), bins = 40,
                   fill = "grey85", colour = "grey65", linewidth = 0.2) +
    geom_density(colour = attend_pal(1), linewidth = 0.8, adjust = 0.9, na.rm = TRUE) +
    geom_rug(alpha = 0.25) +
    geom_vline(xintercept = attend_thresholds$tmb, linetype = "dashed",
               colour = "grey40") +
    facet_wrap(~ definition, scales = "free") +
    labs(title = "TMB distribution by definition (raw mut/Mb)",
         subtitle = paste0("dashed = TMB-high cut (", attend_thresholds$tmb,
                           " mut/Mb). A distinct high-TMB hump = hypermutator mode."),
         x = "TMB (mut/Mb)", y = "density") +
    attend_theme()
} else {
  note_skip("No TMB values to plot.")
}
```

### 2. Bimodality battery

Three independent criteria per definition, because on the raw scale no single test is
decisive:

- **Hartigan's dip test** (dip_bimodality(), reused from attend_classes.R) — the
  scale-invariant departure-from-unimodality statistic; small **p** ⇒ bimodal.
- **Sarle's bimodality coefficient (BC)** — from skewness/kurtosis; **b > 0.555** leans
  bimodal. (Inflated by a heavy tail, so corroborative only.)
- **Gaussian mixture (GMM)** — mclust::Mclust(G = 1:3), BIC-selected component count;
  **≥ 2** components ⇒ the data prefer more than one Gaussian.

The **verdict** calls a definition *bimodal* when **≥ 2 of the 3** criteria agree —
a deliberately conservative majority vote.

```{r battery}
if (nrow(long) > 0) {
  battery <- tmb_battery(long, by = "definition")
  knitr::kable(
    battery |>
      transmute(definition, n,
                dip_D = round(dip_D, 4), dip_p = signif(dip_p, 3), dip_bimodal,
                BC = round(BC, 3), BC_bimodal,
                gmm_G, gmm_bimodal,
                votes, verdict),
    caption = "Bimodality battery per TMB definition. verdict = bimodal when >= 2 of {dip, BC, GMM} agree."
  )
} else {
  battery <- tibble()
  cat("No TMB values — bimodality battery skipped.
")
}
```

The same table read back as sentences, one line per definition, so the vote and its
verdict are stated rather than inferred from the columns.

```{r battery-note, results='asis'}
if (exists("battery") && nrow(battery) > 0) {
  bm <- battery |> filter(verdict == "bimodal") |> pull(definition) |> as.character()
  if (length(bm)) {
    cat("**Bimodal by majority vote:** ", paste(bm, collapse = ", "), ".\n\n", sep = "")
  } else {
    cat("**No definition reaches the ≥2-of-3 bimodality threshold** on the raw scale. ",
        "The high-TMB signal may be a heavy tail rather than a resolved second mode — ",
        "revisit on a log scale if a hypermutator subpopulation is expected.\n\n", sep = "")
  }
}
```

### 3. Mode validation — does the high-TMB mode capture MMR-deficient tumours?

A statistical second mode is only interesting if it is the **biological** one. For every
definition the mixture split into ≥ 2 components, we assign each patient to a component by
posterior, label the **highest-mean component** as the high-TMB mode, and cross-tabulate
that membership against **MMR status** (gianlu__MMR_STATUS, via the shared is_mmrd()).
A Fisher test asks whether the high-TMB mode is enriched for MMR-deficient tumours — the
expected hypermutators.

```{r mode-validation, results='asis'}
mmr_col <- attend_cols$mmr_status   # gianlu__MMR_STATUS

validate_mode <- function(def_key, def_lab) {
  col <- def_cols[[def_key]]
  d <- master |>
    transmute(pid,
              tmb  = suppressWarnings(as.numeric(.data[[col]])),
              mmrd = is_mmrd(.data[[mmr_col]])) |>
    filter(is.finite(tmb))
  if (nrow(d) < 6) return(NULL)
  mc <- tryCatch(mclust::Mclust(d$tmb, G = 1:3, verbose = FALSE), error = function(e) NULL)
  if (is.null(mc) || mc$G < 2) return(NULL)
  hi_comp  <- which.max(as.numeric(mc$parameters$mean))         # highest-mean component
  d$hi_mode <- factor(if_else(mc$classification == hi_comp, "high-TMB mode", "bulk"),
                      levels = c("bulk", "high-TMB mode"))
  d2 <- d |> filter(!is.na(mmrd))
  if (nrow(d2) < 4 || dplyr::n_distinct(d2$hi_mode) < 2 || dplyr::n_distinct(d2$mmrd) < 2)
    return(list(lab = def_lab, tab = table(d$hi_mode), p = NA_real_))
  tab <- table(MMR = if_else(d2$mmrd, "MMRd", "MMRp"), mode = d2$hi_mode)
  p   <- tryCatch(fisher.test(tab)$p.value, error = function(e) NA_real_)
  list(lab = def_lab, tab = tab, p = p)
}

if (!have_mclust) {
  cat("_`mclust` not installed — mode validation skipped._\n\n")
} else if (!mmr_col %in% names(master)) {
  cat("_MMR status column (`", mmr_col, "`) absent from the master — mode validation skipped._\n\n", sep = "")
} else if (!exists("battery") || nrow(battery) == 0) {
  cat("_No definitions to validate._\n\n")
} else {
  qualifying <- battery |> filter(is.finite(gmm_G) & gmm_G >= 2) |> pull(definition) |> as.character()
  key_by_lab <- setNames(names(def_cols), unname(def_labels))
  if (!length(qualifying)) {
    cat("_No definition produced a ≥2-component mixture — no high-TMB mode to validate._\n\n")
  } else {
    for (lab in qualifying) {
      res <- validate_mode(key_by_lab[[lab]], lab)
      if (is.null(res)) next
      cat("### ", res$lab, "\n\n", sep = "")
      print(knitr::kable(as.data.frame.matrix(res$tab),
                         caption = paste0("High-TMB mode membership vs MMR status — ", res$lab)))
      cat("\n\nFisher exact p = ",
          if (is.finite(res$p)) signif(res$p, 3) else "NA (a margin was degenerate)",
          if (is.finite(res$p) && res$p < 0.05) " — the high-TMB mode is significantly enriched for MMRd." else "",
          "\n\n", sep = "")
    }
  }
}
```

### 4. Cross-definition synthesis — does correction sharpen the split?

Removing ancestry-dependent germline inflation should make the hypermutator mode *stand
out more*: as
ancestry-inflated germline counts are stripped, the bulk tightens and the valley
deepens. Two comparators, read left-to-right along the chain:

- **Dip statistic D** (scale-invariant) — higher = further from unimodal.
- **SD-normalised mixture gap** — the distance between the extreme component means in
  units of the definition's own SD (comparable across definitions; the raw mut/Mb gap
  is not, because the definitions live on shifted scales).

```{r synthesis, fig.height=4.5}
have_syn <- exists("battery") && nrow(battery) > 0 &&
  (any(is.finite(battery$dip_D)) || any(is.finite(battery$gmm_gap_sd)))
if (have_syn) {
  syn <- battery |>
    select(definition, `dip statistic D` = dip_D, `mixture gap (÷SD)` = gmm_gap_sd) |>
    pivot_longer(-definition, names_to = "metric", values_to = "value") |>
    filter(is.finite(value))
  if (nrow(syn) > 0) {
    ggplot(syn, aes(definition, value, group = 1)) +
      geom_line(colour = "grey60") +
      geom_point(size = 3, colour = attend_pal(1)) +
      facet_wrap(~ metric, scales = "free_y") +
      labs(title = "Separation strength along the germline-correction chain",
           subtitle = "left to right = normal -> ancestry corrected; up = more clearly bimodal",
           x = NULL, y = NULL) +
      attend_theme() +
      theme(axis.text.x = element_text(angle = 20, hjust = 1))
  } else {
    note_skip("No finite separation metrics to plot.")
  }
} else {
  note_skip("Battery unavailable — synthesis skipped.")
}
```

The cross-definition comparison in words: whether moving from the delivered score to the
ancestry-corrected one deepens the valley, holds it, or closes it.

```{r synthesis-note, results='asis'}
if (exists("battery") && nrow(battery) > 0 && sum(is.finite(battery$dip_D)) >= 2) {
  b <- battery |> filter(is.finite(dip_D))
  first_lab <- as.character(b$definition[1]); last_lab <- as.character(b$definition[nrow(b)])
  d_first <- b$dip_D[1]; d_last <- b$dip_D[nrow(b)]
  cat("Dip D moves from ", signif(d_first, 3), " (", first_lab, ") to ",
      signif(d_last, 3), " (", last_lab, "). ",
      if (d_last > d_first)
        "Separation **increases** after correction — consistent with the correction sharpening the hypermutator mode."
      else if (d_last < d_first)
        "Separation **decreases** after correction — it does not visibly sharpen the split on the raw scale (or the corrected column is too sparse to resolve it)."
      else "Separation is unchanged after correction.",
      "\n\n", sep = "")
}
```

The battery is written to an intermediate for Part 3 below, which appends
TCGA as one more row of this same table.


## Part 2 — TMB by aneuploidy, within MMR strata

Splitting by MMR *first* is the point. The two routes to a high number are biologically
distinct and must not be pooled: MMR-deficient tumours are hypermutators, high TMB with
often-modest aneuploidy, whereas the CN-high / serous-like route drives high aneuploidy
with typically low TMB. Pooling the two would let one route's signal mask the other, so the
aneuploidy-high-vs-low contrast is drawn once per MMR group and repeated across every TMB
definition present.

### Building the strata

Each point is one patient. TMB (mut/Mb, raw) comes from the tmb__* master columns via
the long table above. Aneuploidy status is aneu__aneuploidy_score, the ASCETS arm-level
SCNA burden from data/ascets/, classified **at both cuts** —
`r paste(cuts, collapse = " and ")`, re-derived per cut by add_aneuploidy_cuts() — so every
panel below carries the two classifications side by side rather than one. Raising the cut
can only move a patient from AS High to AS Low, so the AS High arm shrinks and the AS Low
arm absorbs exactly those patients; the count table is the cost of that choice. MMR group is
gianlu__MMR_STATUS collapsed to MMRd/MMRp by the shared is_mmrd(), so the split matches
every other report. Patients missing an aneuploidy score or an MMR call are dropped, at both
cuts. Why two cuts are reported rather than one -> [00-methods](00-methods.html).

```{r aneu-tmb-build}
aneu_col <- attend_cols$aneuploidy    # aneu__aneuploidy_score
mmr_col  <- attend_cols$mmr_status    # gianlu__MMR_STATUS

have_aneu_tmb <- nrow(long) > 0 &&
  aneu_col %in% names(master) && mmr_col %in% names(master)

if (have_aneu_tmb) {
  # The LONG per-cut master: one row per (patient, cut), carrying the aneuploidy_class and
  # mmr_group THAT cut gives them. Joined onto the per-(patient, definition) TMB rows from
  # tmb_long(), so `ta` is one row per patient x TMB definition x cut — the definition and
  # the cut are independent facet dimensions and stay that way in every figure below.
  meta <- master_cuts |>
    select(pid, as_cut, as_cut_lab, aneu_class = aneuploidy_class, mmr_group)

  ta <- long |>
    # many-to-many DECLARED: several TMB definitions per patient meet two cuts per patient,
    # and that product is exactly the intended shape, not an accidental fan-out.
    inner_join(meta, by = "pid", relationship = "many-to-many") |>
    filter(!is.na(aneu_class), !is.na(mmr_group))

  # Facet rows = TMB definitions actually present; drives the plot's fig.height below
  # so each row keeps a legible height whether 1 or 4 definitions render. The cuts go in
  # the COLUMNS, so adding one does not change the row height.
  n_aneu_defs <- max(1L, nlevels(droplevels(ta$definition)))

  cat("patients per MMR group × aneuploidy status, at each cut (baseline definition):\n")
  print(ta |>
          filter(definition == levels(definition)[1]) |>
          count(as_cut_lab, mmr_group, aneu_class) |>
          tidyr::pivot_wider(names_from = aneu_class, values_from = n, values_fill = 0))
} else {
  ta <- tibble()
  n_aneu_defs <- 1L
  cat("Aneuploidy score (`", aneu_col, "`) or MMR status (`", mmr_col,
      "`) absent from the master, or no TMB values — section skipped.\n", sep = "")
}
```

Both two-group tests on the **continuous** TMB, one pair per (aneuploidy cut × MMR group ×
TMB definition) cell, computed here so the figure below carries no statistics of its own.
The cut is part of the grouping, never pooled across: the two classifications are nested, so
a pooled test would count the patients who move between the cuts twice. The **Welch
t-test** runs on **log10 TMB**, not on raw mut/Mb: TMB spans orders of magnitude and the
MMRd arm is hypermutated, so a t on the raw scale is set by whichever single patient sits
highest and its p is not the p it claims to be. The **Mann-Whitney** (`wilcox.test`) runs on
the raw values as the companion, not a substitute — it is rank-based, so the log would not
move it, and it asks about stochastic ordering rather than a mean difference. A cell needs
at least two patients in each aneuploidy arm to be tested; below that both p-values are NA.
A TMB of exactly zero has no log and is dropped from the t-test only, so `n_zero` is
reported beside the counts. Neither p is adjusted across the cells shown.

```{r aneu-tmb-pvals}
if (nrow(ta) > 0) {
  aneu_tmb_p <- ta |>
    group_by(definition, as_cut_lab, mmr_group) |>
    group_modify(function(.x, ...) {
      n_lo <- sum(.x$aneu_class == "aneuploidy-low")
      n_hi <- sum(.x$aneu_class == "aneuploidy-high")
      ok   <- n_lo >= 2 && n_hi >= 2
      # The t-test's own frame: zeros removed, because log10(0) is -Inf and would otherwise
      # silently take the whole group's mean with it. Counted, never absorbed.
      lg <- .x[is.finite(.x$tmb) & .x$tmb > 0, , drop = FALSE]
      t_p <- if (ok && sum(lg$aneu_class == "aneuploidy-low")  >= 2 &&
                       sum(lg$aneu_class == "aneuploidy-high") >= 2) {
        tryCatch(stats::t.test(log10(tmb) ~ aneu_class, data = lg)$p.value,
                 error = function(e) NA_real_)
      } else NA_real_
      w_p <- if (ok) {
        tryCatch(suppressWarnings(stats::wilcox.test(tmb ~ aneu_class, data = .x)$p.value),
                 error = function(e) NA_real_)
      } else NA_real_
      tibble(n_low = n_lo, n_high = n_hi,
             n_zero = sum(.x$tmb <= 0, na.rm = TRUE), t_p = t_p, mw_p = w_p)
    }) |>
    ungroup() |>
    # One two-line label per facet, drawn at the panel ceiling by the figure below.
    mutate(.lab = paste0("t p = ",    ifelse(is.na(t_p),  "NA", signif(t_p,  2)),
                         "\nMW p = ", ifelse(is.na(mw_p), "NA", signif(mw_p, 2))))
} else {
  aneu_tmb_p <- tibble()
  cat("No aneuploidy/TMB data — the two-group tests are skipped.\n")
}
```

The two boxplots the analysis calls for: TMB (raw mut/Mb) in AS High vs AS Low tumours,
faceted **by TMB definition** (rows) and **by aneuploidy cut then MMR group** (columns, so
the same MMR group sits once per cut). The two dimensions are deliberately not collapsed:
the cut changes which patients are AS High, the definition changes what TMB *is*, and one
column pair moving while the other does not is the thing to read. Rows share a y scale
across cuts for that reason. Boxes are filled by aneuploidy status (pale blue = low, salmon
= high — the shared attend_aneu_cols palette every other report uses); the
dashed grey line is the TMB-high cut
(attend_thresholds$tmb = `r attend_thresholds$tmb` mut/Mb). Points are individual
patients (jittered), not group means. Each panel carries the Welch-t (on log10 TMB) and
Mann-Whitney p from the chunk above; at these group sizes they are suggestive, not
confirmatory.

```{r aneu-tmb-box, fig.width=12, fig.height=4.2 * n_aneu_defs + 1.8}
if (nrow(ta) > 0) {
  # fill rides the TOP-LEVEL aes and attend_box(fill = NULL) inherits it — the idiom every
  # other semantic-fill box in the pipeline uses. Without it geom_boxplot falls back to its
  # own default (white) and the scale_fill_manual below binds to nothing and is DISCARDED,
  # silently: the boxes rendered with no colour and no legend, and nothing errored.
  ggplot(ta, aes(aneu_class, tmb, fill = aneu_class)) +
    # points = FALSE: highlight_points() supplies the point layer, partitioned so no patient
    # is drawn twice, and carries the polipo overlay this figure used to be missing.
    # "as_cut_lab" is in `by` so the n-aware mark is decided per PANEL: a group that keeps
    # ten patients at one cut and four at the other must draw a box in one and a median
    # crossbar in the other, which is the cost of the cut made visible.
    attend_box(ta, x = "aneu_class", y = "tmb",
               by = c("definition", "as_cut_lab", "mmr_group"),
               fill = NULL, points = FALSE) +
    highlight_points(ta, "aneu_class", "tmb", base_size = 0.9, base_alpha = 0.5,
                     base_const = "black") +
    # "firebrick" is a base R colour name from no palette, and every other cutoff line in
    # the pipeline is a neutral dashed grey. A threshold is a reference, not a result.
    geom_hline(yintercept = attend_thresholds$tmb, linetype = "dashed",
               colour = "grey40") +
    geom_text(data = aneu_tmb_p, inherit.aes = FALSE,
              aes(x = 1.5, y = Inf, label = .lab),
              vjust = 1.1, size = 2.7, lineheight = 0.95, colour = "grey25") +
    scale_fill_manual(values = attend_aneu_cols, labels = as_label,
                      name = attend_as_legend) +
    # Headroom at the top for the two-line p label; attend_box(label_n = FALSE) adds no
    # y scale of its own here, so this one is not replacing anything.
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.20))) +
    # Short horizontal tick labels — the fill legend already spells out the full
    # aneuploidy status, so the x-axis needn't repeat it (frees per-panel width).
    scale_x_discrete(labels = as_label) +
    # Rows stay the TMB definitions, so free_y scales per definition and NOT per cut: the
    # two cuts of one definition must be read on one axis. The cut leads the column pair.
    facet_grid(definition ~ as_cut_lab + mmr_group, scales = "free_y",
               labeller = as_labeller) +
    labs(title = "TMB by aneuploidy status, split by MMR group, at both aneuploidy cuts",
         subtitle = paste0("dashed = TMB-high cut (", attend_thresholds$tmb,
                           " mut/Mb); AS High = score ≥ the cut named in the strip (",
                           paste(cuts, collapse = " or "), "); t on log10 TMB, MW on raw"),
         x = NULL, y = "TMB (mut/Mb)") +
    attend_theme() +
    theme(axis.text.x = element_text(angle = 0),
          legend.position = "top")
} else {
  note_skip("No aneuploidy/TMB data to plot.")
}
```

Formalising each panel at the **10 mut/Mb** cut rather than by median: within every
(aneuploidy cut × MMR group × TMB definition) cell — one table row per panel of the figure
above, so the two aneuploidy cuts are tested separately and never pooled — patients are
binarised **TMB-high vs TMB-low** at
attend_thresholds$tmb = `r attend_thresholds$tmb` mut/Mb via the shared
tmb_class_of() (the same cut every TMB-keyed figure uses), and the table reports the
**count and fraction of TMB-high tumours in the aneuploidy-low vs -high groups**. A
**Fisher exact test** on the 2×2 (aneuploidy status × TMB-high/low) asks whether the
TMB-high fraction differs by aneuploidy status. Cells with an aneuploidy group of
fewer than 3 patients report NA. The Welch-t and Mann-Whitney p-values computed above are
joined on as the last two columns, so one table carries all three: Fisher asks whether the
*proportion over the 10 mut/Mb cut* differs, the other two ask whether the *distribution*
does, and they can disagree — a shifted distribution that never crosses the cut moves the
latter two and not Fisher.

```{r aneu-tmb-test}
if (nrow(ta) > 0) {
  aneu_tmb_tab <- ta |>
    group_by(definition, as_cut_lab, mmr_group) |>
    group_modify(function(.x, ...) {
      # Binarise TMB at the pipeline 10 mut/Mb cut (shared tmb_class_of()), then
      # count TMB-high tumours within each aneuploidy group.
      tmb_hi <- tmb_class_of(.x$tmb) == "TMB-high"
      is_lo  <- .x$aneu_class == "aneuploidy-low"
      is_hi  <- .x$aneu_class == "aneuploidy-high"
      n_lo <- sum(is_lo); n_hi <- sum(is_hi)
      hi_in_lo <- sum(is_lo & tmb_hi); hi_in_hi <- sum(is_hi & tmb_hi)
      p <- if (n_lo >= 3 && n_hi >= 3) {
        tab <- table(aneu = factor(.x$aneu_class,
                                   levels = c("aneuploidy-low", "aneuploidy-high")),
                     tmb  = factor(tmb_hi, levels = c(FALSE, TRUE)))
        tryCatch(fisher.test(tab)$p.value, error = function(e) NA_real_)
      } else NA_real_
      tibble(
        n_aneu_low    = n_lo,
        n_aneu_high   = n_hi,
        tmbhigh_low   = hi_in_lo,
        pct_low       = if (n_lo > 0) 100 * hi_in_lo / n_lo else NA_real_,
        tmbhigh_high  = hi_in_hi,
        pct_high      = if (n_hi > 0) 100 * hi_in_hi / n_hi else NA_real_,
        fisher_p      = p
      )
    }) |>
    ungroup() |>
    # The two continuous tests were computed above on the SAME (definition, cut, mmr_group)
    # grouping, so this is a key-for-key join, not a re-derivation.
    left_join(select(aneu_tmb_p, definition, as_cut_lab, mmr_group, t_p, mw_p),
              by = c("definition", "as_cut_lab", "mmr_group"))

  knitr::kable(
    aneu_tmb_tab |>
      transmute(definition, `AS cut` = as_cut_lab, mmr_group,
                `n aneu-low`  = n_aneu_low,
                `n aneu-high` = n_aneu_high,
                `TMB-high | aneu-low`  = ifelse(is.finite(pct_low),
                    sprintf("%d (%.0f%%)", tmbhigh_low,  pct_low),  "—"),
                `TMB-high | aneu-high` = ifelse(is.finite(pct_high),
                    sprintf("%d (%.0f%%)", tmbhigh_high, pct_high), "—"),
                `Fisher p` = signif(fisher_p, 3),
                `t p (log10)` = signif(t_p, 3),
                `Mann-Whitney p` = signif(mw_p, 3)),
    caption = paste0("TMB-high tumours (≥ ", attend_thresholds$tmb,
                     " mut/Mb) as count (%) within the aneuploidy-low vs -high groups, ",
                     "with Fisher's exact p on the 2×2 and the two continuous-TMB tests ",
                     "(Welch t on log10, Mann-Whitney on raw), per aneuploidy cut, MMR ",
                     "group and TMB definition.")
  )
} else {
  cat("No aneuploidy/TMB data — test skipped.\n")
}
```

The table above read back as one line per TMB definition **and cut**, giving each MMR
group's TMB-high fraction in the AS Low and AS High groups with its Fisher p. Two lines
that agree mean the contrast survives moving the threshold; two that disagree mean the
result is a statement about `r paste(cuts, collapse = " or ")`, not about aneuploidy.

```{r aneu-tmb-note, results='asis'}
if (exists("aneu_tmb_tab") && nrow(aneu_tmb_tab) > 0) {
  # One MMRp/MMRd line per (TMB definition, cut) present, definitions in correction order —
  # the same cells the boxplot panels and the table above show, in the same order.
  say <- function(b, grp) {
    r <- b |> filter(mmr_group == grp)
    if (nrow(r) == 0) return(paste0("**", grp, "**: no patients."))
    paste0("**", grp, "**: TMB-high fraction ",
           if (is.finite(r$pct_low))  sprintf("%.0f%%", r$pct_low)  else "NA",
           " in ", as_label("aneuploidy-low"), " vs ",
           if (is.finite(r$pct_high)) sprintf("%.0f%%", r$pct_high) else "NA",
           " in ", as_label("aneuploidy-high"), " (Fisher p = ",
           if (is.finite(r$fisher_p)) signif(r$fisher_p, 3) else "NA", ")")
  }
  for (lab in levels(ta$definition)) {
    for (ck in levels(droplevels(aneu_tmb_tab$as_cut_lab))) {
      b <- aneu_tmb_tab |> filter(definition == lab, as_cut_lab == ck)
      if (nrow(b) == 0) next
      cat("- _", lab, "_, ", ck, " — ", say(b, "MMRp"), "; ", say(b, "MMRd"), ".\n", sep = "")
    }
  }
  cat("\n_Read against the two-routes framing at the top: within MMRp the CN-high/serous-like ",
      "signature is a **lower** TMB-high fraction among AS High tumours (high ",
      "aneuploidy, low TMB), while MMRd tumours stay TMB-high largely independent of ",
      "aneuploidy. Comparing the lines across definitions shows whether stripping ",
      "ancestry-inflated germline (normal -> ancestry corrected) changes that within-stratum ",
      "contrast; comparing them across cuts shows whether the contrast is a property of the ",
      "data or of the threshold._\n\n", sep = "")
}
```

## Part 3 — does the split reproduce in TCGA UCEC?

TCGA UCEC 2013 (Kandoth et al., *Nature* 2013) is the same disease sequenced tumour with
matched normal, so germline variants were subtracted at source and its TMB carries none of
the tumour-only inflation the ancestry correction is trying to undo. That makes it a
reference rather than just another cohort: if TCGA is clearly bimodal and ATTEND is not,
the missing split is a measurement problem in tumour-only TMB; and if the ancestry-corrected
definition looks more TCGA-like than the uncorrected one, the correction is moving ATTEND
toward the paired-sequencing ground truth.

TCGA UCEC is **primary** disease and ATTEND is **metastatic**, so any difference between
the cohorts is confounded with disease setting and cannot be attributed to sequencing
strategy alone. TCGA is a reference for what a paired-sequencing TMB distribution looks
like, not a matched control.

### Loading the reference cohort

TCGA's TMB_NONSYNONYMOUS (non-synonymous mutations per Mb, available on the 248 of 373
tumours that were exome-sequenced) on the same raw mut/Mb scale as ATTEND. POLE-ultramutated
tumours are excluded: TCGA has 17 at roughly 232 mut/Mb, an order of magnitude above even
the MSI hypermutators, and ATTEND structurally cannot have them, so leaving them in would
make every comparison detect that rather than the MMRd split. TCGA's MMRd group is its own
MSI (Hyper-mutated) integrative cluster (mmrd_rule = "tcga_exact"), i.e. the paper's
published classification rather than a marker call.

```{r tcga-load}
tcga_raw <- tryCatch(load_tcga_ucec_2013(), error = function(e) tibble())
have_tcga <- nrow(tcga_raw) > 0 && "tmb" %in% names(tcga_raw) &&
             sum(is.finite(tcga_raw$tmb)) >= 10

if (!have_tcga) {
  cat("**TCGA UCEC 2013 not available** (`data/tcga_2013/`) — run",
      "`Rscript code/fetch_tcga_ucec_2013.R`. This report is skipped, not silently empty.\n")
} else {
  n_all  <- sum(is.finite(tcga_raw$tmb))
  tcga <- tcga_raw |> filter(is.finite(tmb), !pole)      # <- POLE excluded
  cat("TCGA tumours with a TMB value :", n_all, "\n")
  cat("POLE-ultramutated, excluded   :", n_all - nrow(tcga), "\n")
  cat("TCGA cohort used here         :", nrow(tcga), "\n")
  cat("  of which MMRd (MSI cluster) :", sum(tcga$mmrd, na.rm = TRUE),
      " | MMRp:", sum(!tcga$mmrd, na.rm = TRUE),
      " | unclassified:", sum(is.na(tcga$mmrd)), "\n")
}
```

Densities of the ATTEND definitions and the TCGA reference on one axis. The ATTEND
distributions are tumour-only, TCGA's is paired — so a systematic rightward shift of the
ATTEND curves relative to TCGA is the germline inflation, visible directly. The dashed
line is the pipeline's `r attend_thresholds$tmb` mut/Mb TMB-high cutoff.

The x-axis is trimmed at the 99th percentile so the shape near the modes stays readable;
the extreme tail is still in every test below, just not in the picture.

```{r tcga-density, fig.height=5}
if (have_tcga && nrow(long) > 0) {
  cmp <- bind_rows(
    long |> transmute(cohort = paste0("ATTEND — ", as.character(definition)), tmb),
    tibble(cohort = "TCGA UCEC (paired, POLE excluded)", tmb = tcga$tmb)
  ) |> filter(is.finite(tmb))

  xmax <- quantile(cmp$tmb, 0.99, na.rm = TRUE)

  ggplot(cmp, aes(tmb, colour = cohort)) +
    geom_density(linewidth = 0.9, na.rm = TRUE) +
    geom_vline(xintercept = attend_thresholds$tmb, linetype = 2, colour = "grey50") +
    coord_cartesian(xlim = c(0, xmax)) +
    labs(title = "TMB distribution: ATTEND (tumour-only) vs TCGA UCEC (tumour-normal paired)",
         subtitle = paste0("raw mut/Mb, x-axis trimmed at the 99th percentile; dashed line = ",
                           attend_thresholds$tmb, " mut/Mb TMB-high cutoff"),
         x = "TMB (mut/Mb)", y = "density", colour = NULL) +
    attend_theme() +
    theme(legend.position = "top") +
    guides(colour = guide_legend(ncol = 1))
}
```

The same three-criterion battery Part 1 runs — literally the same tmb_battery() call —
now with TCGA as an extra row. Read the TCGA row first: it is the benchmark. If TCGA is bimodal and an ATTEND definition is not,
the split exists in this disease and that definition is failing to resolve it.

```{r tcga-battery}
if (have_tcga) {
  # Same tmb_battery() Part 1 runs, on one ungrouped vector — so the TCGA row and the
  # ATTEND rows cannot be computed by two subtly different routes.
  tcga_row <- tmb_battery(tibble(tmb = tcga$tmb), by = NULL) |>
    mutate(definition = "TCGA UCEC (paired, POLE excl.)", .before = 1)

  combined <- if (exists("battery") && nrow(battery) > 0)
    bind_rows(battery |> mutate(definition = as.character(definition)), tcga_row) else tcga_row

  knitr::kable(
    combined |>
      transmute(definition, n,
                dip_D = round(dip_D, 4), dip_p = signif(dip_p, 3), dip_bimodal,
                BC = round(BC, 3), BC_bimodal,
                gmm_G, gmm_bimodal,
                gmm_gap_sd = round(gmm_gap_sd, 2),
                votes, verdict),
    caption = "Bimodality battery — ATTEND definitions plus the TCGA paired-sequencing reference. verdict = bimodal when >= 2 of {dip, BC, GMM} agree."
  )
}
```

Does TCGA's high-TMB mode contain TCGA's MMR-deficient tumours? The same mode validation
Part 1 runs on ATTEND, applied to the reference cohort. A strong enrichment
here establishes that the high mode really is the hypermutator population *in this
disease* — which is what licenses reading ATTEND's high mode the same way.

```{r tcga-mode-validation, results='asis'}
if (have_tcga && have_mclust) {
  mc <- tryCatch(mclust::Mclust(tcga$tmb, G = 1:3, verbose = FALSE), error = function(e) NULL)
  if (is.null(mc) || mc$G < 2) {
    cat("_TCGA TMB resolves to a single mixture component — no high mode to validate._\n\n")
  } else {
    hi_comp <- which.max(as.numeric(mc$parameters$mean))
    d <- tcga |>
      mutate(hi_mode = factor(if_else(mc$classification == hi_comp, "high-TMB mode", "bulk"),
                              levels = c("bulk", "high-TMB mode"))) |>
      filter(!is.na(mmrd))
    if (nrow(d) >= 4 && dplyr::n_distinct(d$hi_mode) == 2 && dplyr::n_distinct(d$mmrd) == 2) {
      tab <- table(MMR = if_else(d$mmrd, "MMRd", "MMRp"), mode = d$hi_mode)
      pv  <- tryCatch(fisher.test(tab)$p.value, error = function(e) NA_real_)
      cat("**TCGA — high-TMB mode vs MMR status** (", mc$G, " mixture components):\n\n", sep = "")
      print(knitr::kable(tab))
      cat("\nFisher p = ", signif(pv, 3), "\n\n", sep = "")
    } else {
      cat("_Not enough classified TCGA tumours across both mode and MMR groups._\n\n")
    }
  }
} else if (have_tcga) {
  cat("_`mclust` not installed — TCGA mode validation skipped._\n\n")
}
```

How each cohort's TMB separates MMR-deficient from MMR-proficient tumours, side by side.
This is the comparison the whole section is for: the **effect size** of MMR status on TMB
in a paired cohort versus in ATTEND under each definition. If the ancestry correction is
working, ATTEND's corrected separation should look more like TCGA's than the uncorrected
one does.

Wilcoxon rather than t-test, because TMB is heavily skewed on the raw scale.

```{r tcga-mmr-contrast, fig.height=5}
if (have_tcga) {
  mmr_col_l <- attend_cols$mmr_status
  att <- if (mmr_col_l %in% names(master)) {
    master |>
      select(pid, all_of(unname(def_cols)), all_of(mmr_col_l)) |>
      pivot_longer(all_of(unname(def_cols)), names_to = "col", values_to = "tmb") |>
      mutate(cohort = paste0("ATTEND — ", def_labels[match(col, def_cols)]),
             tmb    = suppressWarnings(as.numeric(tmb)),
             mmrd   = is_mmrd(.data[[mmr_col_l]])) |>
      filter(is.finite(tmb), !is.na(mmrd)) |>
      select(pid, cohort, tmb, mmrd)     # pid kept so the polipo overlay can resolve
  } else tibble()

  both <- bind_rows(att,
    tcga |> filter(!is.na(mmrd)) |>
      transmute(pid = NA_character_, cohort = "TCGA UCEC (paired)", tmb, mmrd))

  if (nrow(both) > 0) {
    both <- both |> mutate(MMR = if_else(mmrd, "MMRd", "MMRp"))

    print(knitr::kable(
      both |> group_by(cohort, MMR) |>
        summarise(n = n(), median = round(median(tmb), 2),
                  IQR = paste0(round(quantile(tmb, .25), 2), "-", round(quantile(tmb, .75), 2)),
                  .groups = "drop"),
      caption = "TMB by MMR status in each cohort / definition"))

    pv <- both |> group_by(cohort) |>
      summarise(p = tryCatch(signif(wilcox.test(tmb ~ MMR)$p.value, 3),
                             error = function(e) NA_real_), .groups = "drop")
    print(knitr::kable(pv, caption = "MMRd vs MMRp Wilcoxon p, per cohort / definition"))

    ggplot(both, aes(MMR, tmb, fill = MMR)) +
      # expand_y = FALSE: this panel sets its own scale_y_log10() below, which would replace
      # attend_box()'s expansion scale and emit a "Scale for y is already present" message.
      # points = FALSE + highlight_points(): the ATTEND facets carry the polipo overlay like
      # every other per-patient figure. `both` has no pid column (it is cohort/tmb/mmrd), so
      # the id is carried through explicitly — see the cohort build above.
      attend_box(both, x = "MMR", y = "tmb", by = "cohort", fill = NULL,
                 points = FALSE, expand_y = FALSE) +
      highlight_points(both, "MMR", "tmb", base_size = 0.7, base_alpha = 0.45,
                       base_const = "black") +
      geom_hline(yintercept = attend_thresholds$tmb, linetype = 2, colour = "grey50") +
      scale_y_log10() +
      facet_wrap(~ cohort, nrow = 1) +
      scale_fill_manual(values = attend_mmr_cols, labels = as_label,
                        name = attend_mmr_legend) +
      labs(title = "TMB by MMR status — ATTEND definitions vs the TCGA paired reference",
           subtitle = "log10 y-axis for display only; all tests above are on the raw scale",
           x = NULL, y = "TMB (mut/Mb, log10)") +
      attend_theme() + theme(legend.position = "none")
  }
}
```

```{r session-info}
sessionInfo()
```
