---
title: "ATTEND — 6. Covariate distributions by MMR status"
author: "sceriff0"
date: "2026-08-09"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
library(here)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 10, fig.height = 7)

source(here("code", "attend_classes.R"))   # attend_cols / attend_levels
source(here("code", "attend_io.R"))        # read_intermediate()
```

## What this report does

Compares the **whole distribution** of each continuous molecular covariate between
MMR-deficient and MMR-proficient tumours, rather than comparing group averages.

### Why an ECDF rather than a boxplot

An empirical cumulative distribution function plots, for every value on the x-axis, the
fraction of patients at or below it. Reading one:

- The curve always rises from 0 to 1, left to right.
- **Where the two curves separate** tells you *where in the range* the groups differ — a
  gap only at the right-hand end means the groups differ in their high tail, not overall.
- **A curve further right** means systematically higher values in that group.
- The **largest vertical gap** between the curves is the KS statistic `D` tested below.

This matters here because the expected differences are tail phenomena. MMR-deficient
tumours are expected to carry higher TMB and MSI, but through a *hypermutated subset*
rather than a uniform shift — a difference an ECDF shows plainly and a boxplot's median
can hide.

### The two TMB definitions

As everywhere in this pipeline, TMB appears exactly twice: **TMB (normal)**
(`tmb__TMB_SCORE`, the upstream tumour-only score) and **TMB (ancestry corrected)**
(`tmb__TMB_nassar`, after the ancestry recalibration). They get their own ECDF panels, so
whether the germline correction changes the MMR-deficient tail is directly visible.

## Data

Everything is read from `output/clean_data/attend_master_joined` (report 1). Covariates
are mapped to readable labels here; only those actually present on the master are kept,
so a covariate that failed to join is dropped rather than drawn empty.

MMR status is `attend_cols$mmr_status`, collapsed to two groups: deficient
(`r attend_levels$mmr_deficient`) versus everything else non-missing.

```{r data}
master <- read_intermediate("attend_master_joined")

# continuous covariates (prefixed master columns) -> readable labels
covar_map <- c(
  "aneuploidy score" = attend_cols$aneuploidy,   # aneu__aneuploidy_score
  "HRD score"        = attend_cols$hrd,          # hrd__HRD_Score
  "MSI (% unstable)" = "msi__MSI_SCORE"
)

# The two TMB definitions, under their reader-facing names. tmb_defs_present() returns
# normal first, then ancestry corrected (only if it is on the master), so this cannot
# drift from the TMB panels in any other report.
tmb_defs06 <- tmb_defs_present(master)
tmb_map    <- setNames(unname(tmb_defs06), paste0(tmb_def_labels(tmb_defs06), " (mut/Mb)"))
covar_map  <- c(covar_map, tmb_map)

covar_map <- covar_map[covar_map %in% names(master)]   # keep only those present

mmr_col <- attend_cols$mmr_status                       # gianlu__MMR_STATUS
cat("covariates available:", paste(names(covar_map), collapse = ", "), "\n")
```

## ECDF by MMR status

One panel per covariate. Each curve is a group's cumulative distribution; the x-axis is
the covariate's own scale (panels are free-scaled, so they are not comparable to each
other). Patients missing a value are dropped from that panel only, so panel denominators
differ — the KS table below reports each panel's group sizes.

```{r ecdf}
if (length(covar_map) > 0 && mmr_col %in% names(master)) {
  long <- master |>
    transmute(
      MMR = case_when(
        .data[[mmr_col]] == attend_levels$mmr_deficient ~ "MMR deficient",
        !is.na(.data[[mmr_col]])                         ~ "MMR proficient",
        TRUE                                            ~ NA_character_),
      across(all_of(unname(covar_map)), ~ suppressWarnings(as.numeric(.x)))
    ) |>
    pivot_longer(all_of(unname(covar_map)), names_to = "covariate", values_to = "value") |>
    mutate(covariate = names(covar_map)[match(covariate, covar_map)]) |>
    filter(!is.na(MMR), !is.na(value))

  ggplot(long, aes(value, colour = MMR)) +
    stat_ecdf(linewidth = 0.9) +
    facet_wrap(~ covariate, scales = "free_x") +
    # colours are the shared attend_mmr_cols (proficient grey / deficient orange)
    scale_colour_manual(values = attend_mmr_cols) +
    labs(title = "ECDF of molecular covariates by MMR status",
         x = "value", y = "F(x) — cumulative fraction of patients", colour = NULL) +
    attend_theme() +
    theme(legend.position = "top")
} else {
  message("No covariate columns or the MMR column are present in the master.")
}
```

## Per-covariate Kolmogorov–Smirnov test

The formal counterpart to the figure. The two-sample KS test asks whether the two
distributions differ **anywhere**, without assuming a shape or testing only the mean.
`D` is the largest vertical gap between the two curves above — so the test statistic is
literally the thing the eye picks out of the plot.

`n_deficient` and `n_proficient` are that covariate's own group sizes after dropping
missing values, which is why they vary between rows.

**No multiple-testing correction is applied** across the covariates listed. With a handful
of pre-specified covariates that is defensible, but the p-values are per-covariate and
should be read that way. Note also that the two TMB rows are **not independent** — they
are the same patients under two definitions — so they cannot be treated as two separate
pieces of evidence.

```{r ks}
if (length(covar_map) > 0 && mmr_col %in% names(master)) {
  mmr <- case_when(
    master[[mmr_col]] == attend_levels$mmr_deficient ~ "deficient",
    !is.na(master[[mmr_col]])                         ~ "proficient",
    TRUE                                             ~ NA_character_)

  imap_dfr(covar_map, function(col, label) {
    d <- tibble(x = suppressWarnings(as.numeric(master[[col]])), mmr = mmr) |>
      filter(!is.na(x), !is.na(mmr))
    g <- split(d$x, d$mmr)
    if (length(g) == 2) {
      kt <- suppressWarnings(ks.test(g[["deficient"]], g[["proficient"]]))
      tibble(covariate = label,
             n_deficient = length(g[["deficient"]]), n_proficient = length(g[["proficient"]]),
             D = round(unname(kt$statistic), 3), p = signif(kt$p.value, 3))
    } else {
      tibble(covariate = label, n_deficient = NA_integer_, n_proficient = NA_integer_,
             D = NA_real_, p = NA_real_)
    }
  })
}
```

## What to take away

- **Where the curves separate matters as much as whether they separate.** A gap confined
  to the right-hand tail is a hypermutated-subset story, not a shift in the typical
  patient.
- **TMB appears under both definitions.** If the MMR-deficient tail moves between the two
  panels, that is the germline correction acting, not new biology.
- **The KS test is uncorrected and the two TMB rows are the same patients twice.**

```{r session-info}
sessionInfo()
```
