---
title: "ATTEND — 33. Mutation landscape and copy number"
author: "sceriff0"
date: "2026-08-13"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
# renv: bioc::maftools (+ tidyverse, here, fs, data.table).
library(tidyverse)
library(here)
library(fs)
# Headless-safe raster device. The cluster's compute nodes have no X11, so R's default
# png() fails with "unable to open connection to X11 display" and every figure dies. cairo
# needs no display; the guard keeps this working on a machine built without it.
if (capabilities("cairo")) options(bitmapType = "cairo")

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 11, fig.height = 8)

source(here("code", "build_master.R"))       # get_master()
source(here("code", "attend_harmonise.R"))   # build_barcode_pid(), make_id_cfg()
source(here("code", "attend_scna.R"))        # peak union, panel score, LOO stability
source(here("code", "attend_plots.R"))       # scna_mirror_plot(), panel_score_plot(), attend_theme()

MAFTOOLS_OK <- requireNamespace("maftools", quietly = TRUE)
master <- get_master() |> add_molecular_classes() |> add_scna_group()
n_mmrd_high <- sum(master$scna_group == "MMRd-high", na.rm = TRUE)
```

> **Question.** What is altered across this cohort — in genes, in chromosome arms, and along the genome — and does recurrence differ between aneuploidy-high MMRd and MMRp tumours?
> **Cohort.** All `r nrow(master)` patients for Parts 1 and 2; Part 3 is a contrast, MMRd-high (n = `r n_mmrd_high`) against MMRp-high.
> **Reads.** The per-sample MAFs, ASCETS arm calls, ClassifyCNV segments, and GISTIC peaks.
> **Out of scope.** The TCGA cascade -> [09-tcga-classification](09-tcga-classification.html). Aneuploidy within MMRd -> [05-aneuploidy-mmrd](05-aneuploidy-mmrd.html).

Three resolutions of one question, coarsest last. Plot conventions follow TCGA UCEC
(Kandoth et al., *Nature* 2013); the rationale for each is in
[00-methods](00-methods.html). Parts 1 and 2 are descriptive frequencies with **no
significance testing** — GISTIC 2.0 q-values are the field's standard for recurrence
significance and are not applied to these figures.

```{r maftools-missing, eval=!MAFTOOLS_OK, echo=FALSE, results='asis'}
cat('`maftools` is not installed, so the oncoplots are skipped. Install with',
    '`renv::install("bioc::maftools")` and re-knit. The copy-number parts do not need it.\n')
```

## Part 1 — the mutation landscape

Oncoplots follow the TCGA UCEC conventions (Kandoth et al., *Nature* 2013,
papers/nature12113.pdf): patients on the horizontal axis, genes stacked vertically,
covariates as annotation bars, and a per-subgroup frequency comparison matching their
Fig. 2d. The rationale for each convention is in [00-methods](00-methods.html).

The oncoplot annotation carries exactly those two TMB tracks, so reading the two rows
against each other shows directly whether the germline correction moves any patient across
the TMB-high/low boundary.

```{r maftools-missing-part1, eval=!MAFTOOLS_OK, echo=FALSE, results='asis'}
cat('`maftools` is not installed, so the mutation plots below are skipped. Install with',
    '`renv::install("bioc::maftools")` (Bioconductor 3.18) and re-knit.\n')
```

### Building the annotated MAF

Every per-sample MAF in data/variant_annotations/ is read into one object.
Tumor_Sample_Barcode is taken from the file name and translated to pid through the
same barcode crosswalk [01-data-integration](01-data-integration.html) used, then the
per-patient covariate classes are attached
as maftools clinicalData so they can be drawn as annotation tracks.

Watch the mapped-sample count printed at the end: samples that fail to map carry no
covariates and will appear with blank annotation bars.

```{r build-maf, eval=MAFTOOLS_OK}
cw       <- build_barcode_pid(load_gianlu_clinical_data(), make_id_cfg())
maf_long <- load_maf_long()

# TMB annotation tracks: the two definitions the pipeline plots, nothing else.
# All-NA definitions (ancestry corrected before clinical_col is set) are dropped.
tmb_defs05 <- tmb_defs_present(master)
tmb_defs05 <- tmb_defs05[vapply(tmb_defs05, function(col) any(!is.na(master[[col]])), logical(1))]
tmb_labs05 <- tmb_def_labels(tmb_defs05)
# Track names, not internal names: maftools prints clinicalFeatures column names onto
# the oncoplot itself, so these become visible axis labels.
tmb_cls05  <- tmb_def_track_names(tmb_defs05)
for (i in seq_along(tmb_defs05))
  master[[tmb_cls05[i]]] <- tmb_class_of(master[[tmb_defs05[[i]]]])

# Recurrent arm-level CNV as annotation tracks: the top `attend_cnv$oncoplot_arms`
# most-altered chromosome arms (ASCETS calls) become gain/loss/neutral rows per patient.
# Knit-safe: no ASCETS data -> cnv_cls05 stays empty and the oncoplot is mutation-only.
cnv_cls05 <- character(0)
arm_rec   <- recurrent_arm_calls(load_ascets_calls(), n_top = attend_cnv$oncoplot_arms)
if (!is.null(arm_rec) && nrow(arm_rec$wide_top) > 0) {
  arm_by_pid <- arm_rec$wide_top |>
    mutate(pid = cw$pid[match(ID, cw$barcode)]) |>
    filter(!is.na(pid)) |>
    select(pid, starts_with("CNV_"))
  master    <- left_join(master, arm_by_pid, by = "pid")
  cnv_cls05 <- intersect(paste0("CNV_", arm_rec$top_arms), names(master))
}

clin <- maf_long |>
  distinct(Tumor_Sample_Barcode) |>
  mutate(pid = cw$pid[match(Tumor_Sample_Barcode, cw$barcode)]) |>
  left_join(master |> select(pid, aneuploidy_class, all_of(tmb_cls05),
                             MSI_class, MMR_class, all_of(cnv_cls05)),
            by = "pid") |>
  select(Tumor_Sample_Barcode, aneuploidy_class, all_of(tmb_cls05),
         MSI_class, MMR_class, all_of(cnv_cls05))

# GISTIC2 focal peaks, if present, overlay gene-level amp/del directly in the mutation
# grid — the TCGA-standard focal layer. maftools::readGistic REQUIRES all four files and
# hard-stops if any is missing, so only take the overlay branch when all four are found.
gx        <- find_gistic_files()
gistic_ok <- !is.null(gx) && !any(vapply(gx, is.null, logical(1)))
maf <- if (gistic_ok) {
  maftools::read.maf(maf = maf_long, clinicalData = clin,
                     gisticAllLesionsFile = gx$all_lesions,
                     gisticAmpGenesFile   = gx$amp_genes,
                     gisticDelGenesFile   = gx$del_genes,
                     gisticScoresFile     = gx$scores, verbose = FALSE)
} else {
  maftools::read.maf(maf = maf_long, clinicalData = clin, verbose = FALSE)
}
cat("samples mapped to a pid:", sum(!is.na(clin$pid)), "/", nrow(clin),
    "\nTMB tracks:", paste(tmb_labs05, collapse = " | "),
    "\nrecurrent CNV arms annotated:", length(cnv_cls05),
    "\nGISTIC focal overlay:", gistic_ok, "\n")
maf
```

### Cohort mutation summary

The standard maftools dashboard: variant classification and type counts, the
transition/transversion split, per-sample mutation burden, and the most frequently
mutated genes. This is the descriptive baseline — read it before the oncoplot to know
which variant classes dominate, since that drives what the oncoplot colours show.

```{r maf-summary, eval=MAFTOOLS_OK}
maftools::plotmafSummary(maf, addStat = "median", dashboard = TRUE)
```

### Oncoplot — mutation landscape (memo-sorted)

Columns are patients, rows are genes, and the coloured cells are mutation types. The
default maftools memo sort: samples are ordered so co-occurring and mutually
exclusive alterations form the characteristic staircase, the primary mutation-landscape
layout readers expect from an OncoPrint. The annotation-sorted panel below answers a
different question — where alterations concentrate within a covariate — not this one.

```{r oncoplot-memo, eval=MAFTOOLS_OK, fig.height=9}
covar <- c("aneuploidy_class", tmb_cls05, "MSI_class", "MMR_class", cnv_cls05)
maftools::oncoplot(maf, top = 20, clinicalFeatures = covar,
                   sortByAnnotation = FALSE, removeNonMutated = FALSE)
```

### Oncoplot — sorted by annotation (enrichment view)

Samples are **sorted by the annotation tracks** rather than by mutation burden, so enrichment
shows up as a block of colour lining up with an annotation band — TP53 concentrating in
aneuploidy-high, or hypermutation concentrating in MMR-deficient, for example.

removeNonMutated = FALSE keeps unmutated patients in the figure, so the percentages
down the side are out of the **whole cohort**, not out of mutated patients only.

```{r oncoplot, eval=MAFTOOLS_OK, fig.height=9}
covar <- c("aneuploidy_class", tmb_cls05, "MSI_class", "MMR_class", cnv_cls05)
maftools::oncoplot(maf, top = 20, clinicalFeatures = covar,
                   sortByAnnotation = TRUE, removeNonMutated = FALSE)
```

### Oncoplot — the configured gene panels

The same figure restricted to the genes this pipeline tracks by name: the **MMR panel**
(`r paste(attend_gene_panel, collapse = ", ")`), the **Wnt / β-catenin** genes
(`r paste(attend_gene_panel_wnt, collapse = ", ")`), and **TP53**. Genes absent from the
MAF entirely are dropped by the intersection below rather than drawn as empty rows.

```{r oncoplot-panel, eval=MAFTOOLS_OK}
panel_genes <- unique(c(attend_genes_per_gene, "TP53"))
panel <- intersect(panel_genes, as.character(maftools::getGeneSummary(maf)$Hugo_Symbol))
if (length(panel) > 0) {
  maftools::oncoplot(maf, genes = panel, clinicalFeatures = covar,
                     sortByAnnotation = TRUE, removeNonMutated = FALSE)
} else {
  message("None of the configured panel genes are mutated in this cohort.")
}
```

### Mutation frequency by subgroup — the TCGA Fig. 2d view

**This is the figure the oncoplot cannot give you.** An oncoplot shows which patients are
mutated; it does not say whether a gene is mutated at a *different rate* between
subgroups, and it offers no multiple-testing control.

Following TCGA Fig. 2d, each gene's mutation frequency is computed **within each
subgroup** and the subgroups are shown side by side. The subgroups are the four
scna_group strata — MMR status crossed with aneuploidy class — which is the split the
rest of this report and the classification report use.

**The test.** For each gene, a Fisher exact test across the subgroups, then
Benjamini–Hochberg FDR across the genes tested. Genes reaching **FDR < 0.05** are marked
with an asterisk, exactly as TCGA marks them. Only genes mutated in at least
min_mut patients are tested — testing a gene mutated once produces noise, not a result.

Frequencies are *within-group*: the denominator is that subgroup's own patient count, so
a small subgroup shows large jumps per patient. Group sizes are printed with the table.

```{r mutfreq-groups, eval=MAFTOOLS_OK, fig.width=11, fig.height=7}
min_mut  <- 3    # a gene must be mutated in at least this many patients to be tested
top_freq <- 25   # genes shown, ranked by overall mutation frequency

master_grp <- add_scna_group(master)      # mmr_group + scna_group (MMR x aneuploidy)

mut_by_pid <- maf_long |>
  mutate(pid = cw$pid[match(Tumor_Sample_Barcode, cw$barcode)]) |>
  filter(!is.na(pid)) |>
  distinct(pid, Hugo_Symbol)

grp_tbl <- master_grp |>
  filter(!is.na(scna_group)) |>
  select(pid, scna_group) |>
  mutate(scna_group = factor(scna_group, levels = attend_scna$group_levels)) |>
  droplevels()

n_by_grp <- count(grp_tbl, scna_group, name = "n_group")

if (nrow(grp_tbl) > 0 && dplyr::n_distinct(grp_tbl$scna_group) >= 2) {
  print(knitr::kable(n_by_grp, caption = "Patients per subgroup (the denominators below)"))

  genes_keep <- mut_by_pid |>
    semi_join(grp_tbl, by = "pid") |>
    count(Hugo_Symbol, name = "n_mut") |>
    filter(n_mut >= min_mut) |>
    slice_max(n_mut, n = top_freq, with_ties = FALSE) |>
    pull(Hugo_Symbol)

  freq_tbl <- tidyr::crossing(pid = grp_tbl$pid, Hugo_Symbol = genes_keep) |>
    left_join(mut_by_pid |> mutate(mutated = TRUE), by = c("pid", "Hugo_Symbol")) |>
    mutate(mutated = replace_na(mutated, FALSE)) |>
    left_join(grp_tbl, by = "pid") |>
    group_by(Hugo_Symbol, scna_group) |>
    summarise(n_mut = sum(mutated), n = n(), freq = n_mut / n, .groups = "drop")

  # Fisher across subgroups per gene, then BH across genes (TCGA marks FDR < 0.05).
  fdr_tbl <- freq_tbl |>
    group_by(Hugo_Symbol) |>
    summarise(p = {
      m <- rbind(n_mut, n - n_mut)
      tryCatch(stats::fisher.test(m, simulate.p.value = TRUE, B = 1e4)$p.value,
               error = function(e) NA_real_)
    }, .groups = "drop") |>
    mutate(fdr = stats::p.adjust(p, method = "BH"),
           star = if_else(!is.na(fdr) & fdr < 0.05, "*", ""))

  plot_tbl <- freq_tbl |>
    left_join(fdr_tbl, by = "Hugo_Symbol") |>
    mutate(Hugo_Symbol = fct_reorder(paste0(Hugo_Symbol, star), freq, .fun = max))

  print(ggplot(plot_tbl, aes(Hugo_Symbol, freq, fill = scna_group)) +
    geom_col(position = position_dodge(width = 0.8), width = 0.75) +
    coord_flip() +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                       expand = expansion(mult = c(0, 0.05))) +
    labs(title = "Mutation frequency by subgroup (TCGA Fig. 2d convention)",
         subtitle = paste0("Fisher across subgroups, BH-adjusted across ",
                           nrow(fdr_tbl), " genes; * = FDR < 0.05. ",
                           "Frequencies are within-subgroup."),
         x = NULL, y = "Patients mutated (% of subgroup)", fill = NULL) +
    attend_theme() +
    theme(legend.position = "top"))

  print(knitr::kable(fdr_tbl |> arrange(fdr) |> mutate(across(c(p, fdr), ~ signif(.x, 3))),
                     caption = "Per-gene Fisher test across subgroups, BH-adjusted"))
} else {
  message("Fewer than two non-empty scna_group strata — subgroup frequency comparison skipped.")
}
```

## Part 2 — recurrent copy number across the cohort

Two resolutions of the same question, coarsest first. Both are **descriptive frequencies
with no significance testing**: GISTIC 2.0 q-values are the field's standard for
recurrence significance and are not applied to these figures.

### Arm level

Cohort frequency of whole-arm gains and losses from the **ASCETS arm-level calls** — the
recurrent events that define endometrial subtypes, such as 1q and 8q gain in serous-like
tumours. Gains extend right, losses left, and arms are shown in **genome order**
(order_arms_genomic()), not sorted by frequency. Each arm's fraction is out of its own
**evaluable** samples — calls of LOWCOV/NC/NA are excluded from that arm's denominator,
not counted as neutral — and that count is printed on the axis label (arm (n=…)) so a
reduced denominator is visible rather than hidden. The bars below show the
`r attend_cnv$arm_calls$n_barplot` most-altered *eligible* arms out of roughly 44 in the
genome — not every arm — where eligible means evaluable in at least
`r 100 * attend_cnv$arm_calls$min_evaluable_frac`% of samples, the identical rule
top_arms applies to the arm annotation tracks on the oncoplots in
[07-mutation-and-cna](07-mutation-and-cna.html), so neither figure can rank a
poorly-covered arm ahead of a well-covered one. An uncallable arm's CNV_<arm> value on
those oncoplots is NA, not "neutral": missing data is not evidence of no alteration.

```{r recurrent-cnv, fig.width=9, fig.height=6}
arm_freq <- recurrent_arm_calls(load_ascets_calls(), n_top = attend_cnv$oncoplot_arms)

if (is.null(arm_freq)) {
  message("No ASCETS arm-level calls (data/ascets/) — skipping recurrent-CNV summary.")
} else {
  # Same eligibility gate recurrent_arm_calls() applies to top_arms (freq$eligible), so
  # the barplot and the oncoplot annotation tracks can never disagree about which arms
  # are well-covered enough to rank. Applied BEFORE slice_head(), so a poorly-covered arm
  # (e.g. an acrocentric p-arm ASCETS marks LOWCOV in most samples) cannot be promoted into
  # the figure just because it scores 100% altered on a 2-sample denominator.
  fr <- arm_freq$freq |>
    filter(eligible) |>
    slice_head(n = attend_cnv$arm_calls$n_barplot) |>
    mutate(arm = factor(arm, levels = rev(order_arms_genomic(arm))))

  # Axis labels carry each arm's evaluable-sample count, so a reader can see when a
  # frequency rests on a reduced denominator (matched by factor level, not row order).
  arm_labels <- setNames(paste0(as.character(fr$arm), " (n=", fr$n_evaluable, ")"),
                         as.character(fr$arm))

  fr_long <- fr |>
    pivot_longer(c(gain, loss), names_to = "direction", values_to = "frac") |>
    mutate(frac = if_else(direction == "loss", -frac, frac))

  ggplot(fr_long, aes(arm, frac, fill = direction)) +
    geom_col(width = 0.8) +
    geom_hline(yintercept = 0, colour = "grey40", linewidth = 0.3) +
    coord_flip() +
    scale_fill_manual(values = c(gain = unname(ATTEND_OKABE_ITO["vermillion"]),
                                 loss = unname(ATTEND_OKABE_ITO["blue"]))) +
    scale_x_discrete(labels = arm_labels) +
    scale_y_continuous(labels = function(x) paste0(abs(round(100 * x)), "%")) +
    labs(title = "Recurrent arm-level SCNAs across the cohort",
         subtitle = "gain (right) vs loss (left), by fraction of evaluable samples; arms in genome order",
         x = NULL, y = "fraction of evaluable samples", fill = NULL) +
    attend_theme()
}
```

### Genome-wide, by chromosomal location

The TCGA Fig. 1a convention: alterations plotted **by position along the genome** rather
than summarised per arm, split into two amplitude panels — **low-level** and **high-level**
(|log2 ratio| >=  `r attend_cnv$pileup$high_abs_logratio`, the conventional
amplification/homozygous-deletion cutoff) — gain up/red, loss down/blue. Grey vertical
lines are chromosome boundaries; dashed lines mark centromeres, read from the arm-boundary
table. ChrY is dropped from the axis — the cohort is uterine and therefore all female, so
it carries no calls and only an empty tail.

Built from the ClassifyCNV altered segments (data/cnv_annotations/) in
`r attend_cnv$pileup$bin / 1e6` Mb bins; the figure caption states the sample denominator
and whether it came from an explicit count, the profiled-sample attribute, or the segments
observed. **These are descriptive frequencies with no significance testing** — GISTIC
q-values are the field's standard for recurrence significance and are not applied here.
This is the figure that shows *focal* recurrence — a narrow tall spike is a focal peak, a
broad plateau is an arm-level event and also appears in the arm barplot above.

```{r segment-pileup, fig.width=12, fig.height=6}
arm_bounds <- load_arm_boundaries()
pile <- segment_pileup(load_cnv_data(), arm_bounds)

if (is.null(pile)) {
  message("No ClassifyCNV segments (data/cnv_annotations/) or arm-boundary table — ",
          "skipping the genome-wide pileup.")
} else {
  # Capture the denominator + its source before any filter()/mutate() that could drop
  # these custom attributes.
  n_samples        <- attr(pile, "n_samples")
  n_samples_source <- attr(pile, "n_samples_source")
  offs             <- attr(pile, "chrom_offsets")

  # Readable phrase for the subtitle, not the internal token (argument/profiled/observed)
  # segment_pileup() records in n_samples_source — a subtitle should read as English.
  n_samples_desc <- c(argument = "an explicitly supplied cohort size",
                      profiled = "the number of samples profiled",
                      observed = "the samples present in the segment data")[[n_samples_source]]

  # ChrY dropped: the cohort is uterine and therefore all female, so chrY carries no
  # calls and would only draw an empty axis tail. ChrX is kept.
  pile <- pile |> filter(chrom != "chrY")
  offs <- offs  |> filter(chrom != "chrY")

  # Centromere = each chromosome's p/q boundary, read from the arm-boundary table
  # (never hard-coded) and converted to genome coordinates with the same offsets used
  # for the chromosome-boundary lines and axis breaks.
  centromeres <- arm_bounds |>
    filter(str_detect(arm, "p$"), chrom != "chrY") |>
    mutate(chrom = as.character(chrom)) |>
    inner_join(offs |> select(chrom, offset), by = "chrom") |>
    transmute(gpos = offset + end)

  pl <- pile |>
    select(gpos, gain_low, gain_high, loss_low, loss_high) |>
    pivot_longer(-gpos, names_to = c("direction", "tier"), names_sep = "_",
                values_to = "frac") |>
    mutate(frac = if_else(direction == "loss", -frac, frac),
           tier = factor(tier, levels = c("low", "high"),
                        labels = c("low-level", "high-level")))

  ggplot(pl, aes(gpos, frac, fill = direction)) +
    geom_col(width = attend_cnv$pileup$bin) +
    geom_hline(yintercept = 0, colour = "grey40", linewidth = 0.3) +
    geom_vline(xintercept = offs$offset[-1], colour = "grey85", linewidth = 0.2) +
    geom_vline(data = centromeres, aes(xintercept = gpos), colour = "grey60",
              linewidth = 0.3, linetype = "dashed") +
    facet_grid(tier ~ ., scales = "free_y") +
    scale_fill_manual(values = c(gain = unname(ATTEND_OKABE_ITO["vermillion"]),
                                 loss = unname(ATTEND_OKABE_ITO["blue"]))) +
    scale_x_continuous(breaks = offs$mid, labels = str_remove(offs$chrom, "^chr"),
                       expand = c(0, 0)) +
    scale_y_continuous(labels = function(x) paste0(abs(round(100 * x)), "%")) +
    labs(title = "Genome-wide recurrent SCNA pileup (TCGA Fig. 1a convention)",
         subtitle = paste0("gain (up) vs loss (down), fraction of ", n_samples,
                           " samples (", n_samples_desc, "); ",
                           attend_cnv$pileup$bin / 1e6,
                           " Mb bins; dashed lines = centromeres; no significance testing applied"),
         x = "chromosome", y = "fraction of samples", fill = NULL) +
    attend_theme() +
    theme(panel.grid.major.x = element_blank(),
          panel.grid.minor.x = element_blank())
}
```

---

## Part 3 — the aneuploidy x MMR contrast

Either the MMRd-high group reaches aneuploidy the same way MMR-proficient serous-like
tumours do, in which case it should carry the same recurrent peaks, or it gets there by a
different route and the peak profile should differ. With n = `r n_mmrd_high` in the
MMRd-high group, the endpoint is a single pre-specified composite score rather than a
per-peak test.

### Setup and the two ID spaces

GISTIC and .seg sample ids are sequencing **barcodes**; the master is keyed on **pid**.
Joining them directly gives an all-NA result *with no error*, so every lookup below goes
through group_of_barcode(), built on the same crosswalk as everything else.

The grouping variable scna_group composes is_mmrd() (IHC-based, MSI deliberately not a
criterion) with aneuploidy_class at the pre-registered cut
`r attend_thresholds$aneuploidy`, the same definition
[07-mutation-and-cna](07-mutation-and-cna.html)'s frequency comparison uses.

```{r scna-gate}
have_master <- exists("master") && !is.null(master) && nrow(master) > 0
if (have_master && !"scna_group" %in% names(master)) master <- add_scna_group(master)

cw_scna <- tryCatch(build_barcode_pid(load_gianlu_clinical_data(), make_id_cfg()),
                    error = function(e) NULL)
have_cw <- !is.null(cw_scna) && nrow(cw_scna) > 0

group_of_barcode <- function(ids) {
  if (!have_master || !have_cw)
    return(factor(rep(NA_character_, length(ids)), levels = attend_scna$group_levels))
  barcode_scna_group(ids, master, cw_scna)
}

gx_dir   <- here("data", attend_cnv$gistic$dir)
run_dirs <- c(all       = file.path(gx_dir, "all"),
              mmrp_high = file.path(gx_dir, "mmrp_high"),
              mmrp_low  = file.path(gx_dir, "mmrp_low"),
              mmrd_low  = file.path(gx_dir, "mmrd_low"),
              mmrd_high = file.path(gx_dir, "mmrd_high"))
have_gistic <- dir.exists(run_dirs[["all"]])

if (!have_master) {
  cat("Master table absent — run report 01 first.\n")
} else if (!have_gistic) {
  cat("**GISTIC outputs absent** (`", run_dirs[["all"]], "`).\n\n",
      "Run `MODE=groups sbatch code/run_gistic.sh`, then `MODE=loo`. ",
      "this report is skipped, not silently empty.\n", sep = "")
}
```

Peaks come from five runs; per-sample calls come from one. Five GISTIC2 runs feed this
report: the pooled cohort and the four scna_group strata. GISTIC's background model is
cohort-specific, so the same tumour gets a different thresholded call depending on which
cohort it was run in, and comparing calls across runs would compare five different null
models. The two roles are therefore kept apart:

- **Peak definition** uses the union of significant peaks across all five runs
  (gistic_peak_union()), so a peak private to the small MMRd-high stratum, which the
  pooled run's genome-wide threshold would bury, is still discoverable.
- **Per-sample calls** are read only from the pooled run's all_lesions.conf_*.txt. No
  per-sample call here comes from a stratified run.

The cohort this contrast conditions on. Fisher tests whether MMR status and aneuploidy class are
associated at all — if they are strongly associated, the MMRd-high cell is small by
construction and everything below is underpowered.

```{r scna-contingency, eval=have_master}
if (all(c("mmr_group", "aneuploidy_class") %in% names(master))) {
  tab2 <- table(MMR = master$mmr_group, Aneuploidy = master$aneuploidy_class)
  print(knitr::kable(tab2, caption = "MMR x aneuploidy-class contingency (patient level)"))
  ft2 <- tryCatch(fisher.test(tab2), error = function(e) NULL)
  if (!is.null(ft2)) cat("Fisher p =", signif(ft2$p.value, 3),
                        ", OR =", signif(unname(ft2$estimate), 3), "\n\n")
}
```

### The primary endpoint — a pre-specified serous-like panel score

**Why a composite score rather than testing peaks one at a time.** At n ≈ 9 no individual
peak has the power to show a difference, and testing hundreds of peaks in a group that
small produces false positives, not findings. A single composite score over a *fixed,
externally-specified* set of loci concentrates the evidence into one test.

**Pre-specification is what makes this confirmatory.** The 12-locus panel
(attend_scna$panel) is fixed *a priori* from the TCGA UCEC recurrent-peak list
(Kandoth et al. 2013). No locus was added, removed, or re-directioned after looking at
ATTEND's own GISTIC output. Each locus is **directional** — a deletion at *MYC* does not
count as serous-like — matched by match_panel_peaks().

Loci with no matching pooled-run peak of the right direction **leave the denominator**
rather than counting as unaltered, so the score is a fraction of *resolvable* loci.

The test is a **label permutation** (B = `r attend_scna$perm_B`), which makes no
distributional assumption — appropriate at this n. A Wilcoxon is reported alongside.
A permutation p is bounded below by 1/(B+1), so a floor-level result is reported as
p < … rather than as an exact tiny value.

```{r scna-panel-score, eval=have_master && have_gistic, fig.height=5}
pooled <- load_gistic_lesions_at(run_dirs[["all"]])
if (is.null(pooled)) {
  cat("Pooled GISTIC run (`", run_dirs[["all"]], "`) not readable — panel score skipped.\n", sep = "")
} else {
  scores <- panel_score(pooled$mat, pooled$peaks)
  grp    <- group_of_barcode(rownames(pooled$mat))

  cat("Panel resolved", attr(scores, "n_loci"), "of 12 pre-specified loci to a pooled-run ",
      "peak of the matching direction",
      if (attr(scores, "n_loci") > 0) paste0(" (", paste(attr(scores, "loci"), collapse = ", "), ").")
      else ".", "\n\n")

  p1 <- panel_score_plot(scores, grp,
                         title = "Composite TCGA UCEC serous-like panel score, by subgroup")
  if (!is.null(p1)) print(p1)

  keep2 <- !is.na(grp) & grp %in% c("MMRd-high", "MMRp-high")
  if (sum(keep2) >= 3 && dplyr::n_distinct(droplevels(grp[keep2])) == 2) {
    g2 <- droplevels(grp[keep2])
    pt <- perm_test_two_group(scores[keep2], g2, B = attend_scna$perm_B, seed = 1)
    wt <- tryCatch(stats::wilcox.test(scores[keep2] ~ g2), error = function(e) NULL)

    floor_p <- 1 / (pt$B + 1)
    p_txt <- if (pt$p <= floor_p + 1e-12) {
      sprintf("p < %s (bounded at the B = %s resolution floor)",
              format(floor_p, scientific = TRUE, digits = 1), format(pt$B, big.mark = ","))
    } else sprintf("p = %.4g", pt$p)

    cat("**MMRd-high (n =", sum(g2 == "MMRd-high"), ") vs MMRp-high (n =", sum(g2 == "MMRp-high"),
        ")**: observed difference in mean panel score =", signif(pt$stat, 3),
        "; label-permutation ", p_txt, ".",
        if (!is.null(wt)) paste0(" Wilcoxon p = ", signif(wt$p.value, 3), ".") else "",
        "\n\nIf the 9 reach aneuploidy by a different route, the score separates from ",
        "MMRp-high; if they phenocopy copy-number-high tumours, it does not.\n\n")
  } else {
    cat("Fewer than 3 samples resolved to both MMRd-high and MMRp-high — permutation test skipped.\n\n")
  }
}
```

#### Which loci drive the score

The composite above is the primary endpoint; this breaks it into its 12 parts as a
**secondary, more granular** view. Each locus is tested one at a time and **Holm-corrected
within the 12-locus family**.

Holm is applied over **all 12 pre-specified loci**, including any that did not resolve to
a pooled-run peak (p = NA) — not just the testable ones. Shrinking the family to the
loci that happened to resolve would inflate significance on the survivors.

```{r scna-locus-table, eval=have_master && have_gistic}
if (exists("pooled") && !is.null(pooled)) {
  mp <- match_panel_peaks(pooled$peaks, attend_scna$panel)
  ft_all <- scna_freq_table(pooled$mat, grp)
  rows <- lapply(seq_len(nrow(mp)), function(i) {
    pk <- mp$peak_id[i]
    if (is.na(pk)) return(data.frame(locus = mp$locus[i], direction = mp$direction[i],
                                     freq_mmrd_high = NA_real_, freq_mmrp_high = NA_real_,
                                     p = NA_real_, stringsAsFactors = FALSE))
    a <- ft_all[ft_all$peak_id == pk & ft_all$scna_group == "MMRd-high", ]
    b <- ft_all[ft_all$peak_id == pk & ft_all$scna_group == "MMRp-high", ]
    p <- NA_real_
    if (nrow(a) && nrow(b)) {
      tt <- matrix(c(a$n_altered, a$n - a$n_altered, b$n_altered, b$n - b$n_altered), nrow = 2)
      p <- tryCatch(stats::fisher.test(tt)$p.value, error = function(e) NA_real_)
    }
    data.frame(locus = mp$locus[i], direction = mp$direction[i],
               freq_mmrd_high = if (nrow(a)) a$freq else NA_real_,
               freq_mmrp_high = if (nrow(b)) b$freq else NA_real_,
               p = p, stringsAsFactors = FALSE)
  })
  locus_tbl <- do.call(rbind, rows)
  locus_tbl$p_holm <- family_adjust(locus_tbl$p, "A", n = nrow(locus_tbl))
  knitr::kable(locus_tbl, digits = 3,
               caption = "The 12 pre-specified TCGA UCEC loci, Holm-adjusted within family (secondary to the composite score)")
}
```

### Where each subgroup's calls sit on the genome

Peak-call frequency from the pooled run, amplifications above the axis and deletions
below, split by subgroup and chromosome. The visual counterpart to the composite score:
if MMRd-high really phenocopies MMRp-high, their two panels should have peaks in the same
places.

```{r scna-mirror, eval=have_master && have_gistic}
if (exists("pooled") && !is.null(pooled)) {
  grp3      <- group_of_barcode(rownames(pooled$mat))
  freq_tbl3 <- scna_freq_table(pooled$mat, grp3)
  p3 <- scna_mirror_plot(freq_tbl3, pooled$peaks,
                         title = "Pooled-run peak-call frequency by subgroup (amp above / del below)")
  if (!is.null(p3)) print(p3)
}
```

How much the five runs' significant-peak sets overlap, as pairwise Jaccard. Low overlap
between a stratum and the pooled run means that stratum has genuinely private peaks —
which is the case the union-of-five design exists to catch. Peaks private to a single run
are listed separately.

```{r scna-peakset, eval=have_master && have_gistic}
union_peaks <- gistic_peak_union(run_dirs)
if (is.null(union_peaks)) {
  cat("No GISTIC lesions parsed from any of the five run folders — skipped.\n")
} else {
  pss <- peak_set_structure(union_peaks)
  print(knitr::kable(round(pss$jaccard, 2),
                     caption = "Jaccard overlap of significant-peak sets between the five GISTIC runs"))
  if (nrow(pss$private))
    print(knitr::kable(pss$private,
                       caption = "Peaks private to a single run (no overlapping peak in any other)"))
}
```

### Leave-one-out stability of the MMRd-high peaks

**The asymmetric rule this rests on.** A peak *present* in the MMRd-high list, supported by
several of the 9 tumours and stable to leave-one-out, is **informative**. A peak *absent*
from that list is **uninformative** — at n = 9 absence is indistinguishable from low power.
Presence carries evidence; absence does not. That is why the table reports support counts
and retention fractions for MMRd-high peaks rather than a between-group presence/absence
verdict.

The retention fraction is the proportion of the 9 leave-one-out GISTIC re-runs in which a
peak survives. A peak retained in 9/9 does not depend on any single tumour; a peak
retained in 3/9 is one or two tumours' worth of signal.

**This is a descriptive robustness diagnostic, not a calibrated error bound** — informal
stability selection (Meinshausen & Bühlmann 2010) adapted to n = 9. Canonical stability
selection uses ⌊n/2⌋ subsamples and an exchangeability assumption that leave-one-out at
this n only approximates. It is not an FDR statement.

Support counts carry Wilson 95% confidence intervals, which behave sensibly at small n
and at frequencies near 0 or 1 where a normal-approximation interval does not.

```{r scna-loo, eval=have_master && have_gistic}
if (exists("union_peaks") && !is.null(union_peaks)) {
  mmrd_high <- load_gistic_lesions_at(run_dirs[["mmrd_high"]])
  if (is.null(mmrd_high)) {
    cat("MMRd-high GISTIC run (`", run_dirs[["mmrd_high"]], "`) absent — cannot list its peaks.\n", sep = "")
  } else {
    # Soft check: the GISTIC run's sample composition may predate a later threshold
    # change on the master, so a mismatch is a warning, not an error.
    gb_supp <- group_of_barcode(rownames(mmrd_high$mat))
    if (any(!is.na(gb_supp) & gb_supp != "MMRd-high")) {
      warning("MMRd-high GISTIC run contains ", sum(!is.na(gb_supp) & gb_supp != "MMRd-high"),
              " sample(s) no longer classified MMRd-high under the current master — ",
              "the GISTIC run may predate a threshold change")
    }

    supp <- scna_freq_table(mmrd_high$mat,
                            factor(rep("MMRd-high", nrow(mmrd_high$mat)),
                                   levels = attend_scna$group_levels))
    supp <- supp[supp$scna_group == "MMRd-high", ]

    loo_dir <- file.path(gx_dir, "loo")
    loo_folders <- if (dir.exists(loo_dir))
      list.dirs(loo_dir, full.names = TRUE, recursive = FALSE) else character(0)
    loo <- if (length(loo_folders)) loo_stability(mmrd_high$peaks, loo_folders) else NULL

    tbl <- merge(supp[, c("peak_id", "n_altered", "n", "freq", "ci_lo", "ci_hi")],
                 mmrd_high$peaks[, c("peak_id", "descriptor", "chrom", "direction")],
                 by = "peak_id")
    if (!is.null(loo)) {
      tbl <- merge(tbl, loo, by = "peak_id", all.x = TRUE)
    } else {
      tbl$n_loo <- NA_integer_; tbl$n_retained <- NA_integer_; tbl$retained_frac <- NA_real_
    }
    tbl <- tbl[order(-tbl$freq), ]
    print(knitr::kable(tbl, digits = 2,
          caption = "Every MMRd-high peak: support count with Wilson 95% CI, and leave-one-out retention fraction (descriptive, not a calibrated bound)"))
    if (is.null(loo)) cat("\nnote: no leave-one-out GISTIC output under `", loo_dir,
                          "` — retention columns are NA; the support counts stand alone.\n\n", sep = "")
  }
}
```

```{r session-info}
sessionInfo()
```
