---
title: "ATTEND — 07. Mutation landscape and copy number"
author: "sceriff0"
date: "2026-08-13"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
# renv: bioc::maftools (+ tidyverse, here, fs, data.table).
library(tidyverse)
library(here)
library(fs)
# Headless-safe raster device. The cluster's compute nodes have no X11, so R's default
# png() fails with "unable to open connection to X11 display" and every figure dies. cairo
# needs no display; the guard keeps this working on a machine built without it.
if (capabilities("cairo")) options(bitmapType = "cairo")

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 11, fig.height = 8)

source(here("code", "build_master.R"))       # get_master()
source(here("code", "attend_harmonise.R"))   # build_barcode_pid(), make_id_cfg()
source(here("code", "attend_scna.R"))        # peak union, panel score, LOO stability
source(here("code", "attend_plots.R"))       # scna_mirror_plot(), panel_score_plot(), attend_theme()

# Two INDEPENDENT preconditions for the oncoplots, kept apart on purpose. A single
# MAFTOOLS_OK gate reported "maftools is not installed" whichever one failed, so a
# missing data/variant_annotations/ sync was misdiagnosed as a missing package — and
# with maftools present but the folder absent the knit died inside the loader instead.
MAFTOOLS_OK <- requireNamespace("maftools", quietly = TRUE)
MAF_DATA_OK <- has_maf_data()                    # code/load_wes_results.R
ONCO_OK     <- MAFTOOLS_OK && MAF_DATA_OK
master <- get_master() |> add_molecular_classes() |> add_scna_group()
n_mmrd_high <- sum(master$scna_group == "MMRd-high", na.rm = TRUE)
```

> **Question.** What is altered across this cohort — in genes, in chromosome arms, and along the genome — and does recurrence differ between aneuploidy-high MMRd and MMRp tumours?
> **Cohort.** All `r nrow(master)` patients for Parts 1 and 2; Part 3 is a contrast, MMRd-high (n = `r n_mmrd_high`) against MMRp-high.
> **Reads.** The per-sample MAFs, ASCETS arm calls, ClassifyCNV segments, and GISTIC peaks.
> **Out of scope.** The TCGA cascade -> [09-tcga-classification](09-tcga-classification.html). Aneuploidy within MMRd -> [05-aneuploidy-mmrd](05-aneuploidy-mmrd.html).

Three resolutions of one question, coarsest last. Plot conventions follow TCGA UCEC
(Kandoth et al., *Nature* 2013); the rationale for each is in
[00-methods](00-methods.html). Parts 1 and 2 are descriptive frequencies with **no
significance testing** — GISTIC 2.0 q-values are the field's standard for recurrence
significance and are not applied to these figures.

```{r maftools-missing, eval=!ONCO_OK, echo=FALSE, results='asis'}
cat("### Part 1 is not in this build\n\n",
    "Everything in Part 1 needs both `maftools` **and** the per-sample MAFs. ",
    "What is missing here:\n\n", sep = "")
if (!MAFTOOLS_OK)
  cat("- **`maftools` is not installed.** `renv::install(\"bioc::maftools\")`, ",
      "then `renv::snapshot()`.\n", sep = "")
if (!MAF_DATA_OK)
  cat("- **No MAFs in `", maf_dir(), "`.** Sync them from the cluster ",
      "(`/hpcnfs/scratch/P_DIMA_ATTEND/.../variant_annotation`) into that folder.\n", sep = "")
cat("\nParts 2 and 3 need neither and are built below. Every figure in Part 1 is ",
    "skipped rather than drawn empty, so an absent input can never be read as ",
    "an absent alteration.\n", sep = "")
```

## Part 1 — the mutation landscape

Oncoplots follow the TCGA UCEC conventions (Kandoth et al., *Nature* 2013,
papers/nature12113.pdf): patients on the horizontal axis, genes stacked vertically,
covariates as annotation bars, and a per-subgroup frequency comparison matching their
Fig. 2d. The rationale for each convention is in [00-methods](00-methods.html).

The oncoplot annotation carries exactly those two TMB tracks, so reading the two rows
against each other shows directly whether the germline correction moves any patient across
the TMB-high/low boundary.

### Building the annotated MAF

Every per-sample MAF in data/variant_annotations/ is read into one object.
Tumor_Sample_Barcode is taken from the file name and translated to pid through the
same barcode crosswalk [01-data-integration](01-data-integration.html) used, then the
per-patient covariate classes are attached
as maftools clinicalData so they can be drawn as annotation tracks.

Watch the mapped-sample count printed at the end: samples that fail to map carry no
covariates and will appear with blank annotation bars.

```{r build-maf, eval=ONCO_OK}
cw       <- build_barcode_pid(load_gianlu_clinical_data(), make_id_cfg())
maf_long <- load_maf_long()

# TMB annotation tracks: the two definitions the pipeline plots, nothing else.
# All-NA definitions (ancestry corrected before clinical_col is set) are dropped.
tmb_defs05 <- tmb_defs_present(master)
tmb_defs05 <- tmb_defs05[vapply(tmb_defs05, function(col) any(!is.na(master[[col]])), logical(1))]
tmb_labs05 <- tmb_def_labels(tmb_defs05)
# Track names, not internal names: maftools prints clinicalFeatures column names onto
# the oncoplot itself, so these become visible axis labels.
tmb_cls05  <- tmb_def_track_names(tmb_defs05)
for (i in seq_along(tmb_defs05))
  master[[tmb_cls05[i]]] <- tmb_class_of(master[[tmb_defs05[[i]]]])

# Recurrent arm-level CNV as annotation tracks: the top `attend_cnv$oncoplot_arms`
# most-altered chromosome arms (ASCETS calls) become gain/loss/neutral rows per patient.
# Knit-safe: no ASCETS data -> cnv_cls05 stays empty and the oncoplot is mutation-only.
cnv_cls05 <- character(0)
arm_rec   <- recurrent_arm_calls(load_ascets_calls(), n_top = attend_cnv$oncoplot_arms)
if (!is.null(arm_rec) && nrow(arm_rec$wide_top) > 0) {
  arm_by_pid <- arm_rec$wide_top |>
    mutate(pid = cw$pid[match(ID, cw$barcode)]) |>
    filter(!is.na(pid)) |>
    select(pid, starts_with("CNV_"))
  master    <- left_join(master, arm_by_pid, by = "pid")
  cnv_cls05 <- intersect(paste0("CNV_", arm_rec$top_arms), names(master))
}

clin <- maf_long |>
  distinct(Tumor_Sample_Barcode) |>
  mutate(pid = cw$pid[match(Tumor_Sample_Barcode, cw$barcode)]) |>
  left_join(master |> select(pid, aneuploidy_class, all_of(tmb_cls05),
                             MSI_class, MMR_class, all_of(cnv_cls05)),
            by = "pid") |>
  select(Tumor_Sample_Barcode, aneuploidy_class, all_of(tmb_cls05),
         MSI_class, MMR_class, all_of(cnv_cls05))

# GISTIC2 focal peaks, if present, overlay gene-level amp/del directly in the mutation
# grid — the TCGA-standard focal layer. maftools::readGistic REQUIRES all four files and
# hard-stops if any is missing, so only take the overlay branch when all four are found.
gx        <- find_gistic_files()
gistic_ok <- !is.null(gx) && !any(vapply(gx, is.null, logical(1)))
maf <- if (gistic_ok) {
  maftools::read.maf(maf = maf_long, clinicalData = clin,
                     gisticAllLesionsFile = gx$all_lesions,
                     gisticAmpGenesFile   = gx$amp_genes,
                     gisticDelGenesFile   = gx$del_genes,
                     gisticScoresFile     = gx$scores, verbose = FALSE)
} else {
  maftools::read.maf(maf = maf_long, clinicalData = clin, verbose = FALSE)
}
cat("samples mapped to a pid:", sum(!is.na(clin$pid)), "/", nrow(clin),
    "\nTMB tracks:", paste(tmb_labs05, collapse = " | "),
    "\nrecurrent CNV arms annotated:", length(cnv_cls05),
    "\nGISTIC focal overlay:", gistic_ok, "\n")
maf
```

### Cohort mutation summary

The standard maftools dashboard: variant classification and type counts, the
transition/transversion split, per-sample mutation burden, and the most frequently
mutated genes. This is the descriptive baseline — read it before the oncoplot to know
which variant classes dominate, since that drives what the oncoplot colours show.

```{r maf-summary, eval=ONCO_OK}
maftools::plotmafSummary(maf, addStat = "median", dashboard = TRUE)
```

### Oncoplot — mutation landscape (memo-sorted)

Columns are patients, rows are genes, and the coloured cells are mutation types. The
default maftools memo sort: samples are ordered so co-occurring and mutually
exclusive alterations form the characteristic staircase, the primary mutation-landscape
layout readers expect from an OncoPrint. The annotation-sorted panel below answers a
different question — where alterations concentrate within a covariate — not this one.

```{r oncoplot-memo, eval=ONCO_OK, fig.height=9}
covar <- c("aneuploidy_class", tmb_cls05, "MSI_class", "MMR_class", cnv_cls05)
maftools::oncoplot(maf, top = 20, clinicalFeatures = covar,
                   sortByAnnotation = FALSE, removeNonMutated = FALSE)
```

### Oncoplot — sorted by annotation (enrichment view)

Samples are **sorted by the annotation tracks** rather than by mutation burden, so enrichment
shows up as a block of colour lining up with an annotation band — TP53 concentrating in
aneuploidy-high, or hypermutation concentrating in MMR-deficient, for example.

removeNonMutated = FALSE keeps unmutated patients in the figure, so the percentages
down the side are out of the **whole cohort**, not out of mutated patients only.

```{r oncoplot, eval=ONCO_OK, fig.height=9}
covar <- c("aneuploidy_class", tmb_cls05, "MSI_class", "MMR_class", cnv_cls05)
maftools::oncoplot(maf, top = 20, clinicalFeatures = covar,
                   sortByAnnotation = TRUE, removeNonMutated = FALSE)
```

### Oncoplot — the configured gene panels

The same figure restricted to the genes this pipeline tracks by name: the **MMR panel**
(`r paste(attend_gene_panel, collapse = ", ")`), the **Wnt / β-catenin** genes
(`r paste(attend_gene_panel_wnt, collapse = ", ")`), and **TP53**. Genes absent from the
MAF entirely are dropped by the intersection below rather than drawn as empty rows.

```{r oncoplot-panel, eval=ONCO_OK}
panel_genes <- unique(c(attend_genes_per_gene, "TP53"))
panel <- intersect(panel_genes, as.character(maftools::getGeneSummary(maf)$Hugo_Symbol))
if (length(panel) > 0) {
  maftools::oncoplot(maf, genes = panel, clinicalFeatures = covar,
                     sortByAnnotation = TRUE, removeNonMutated = FALSE)
} else {
  message("None of the configured panel genes are mutated in this cohort.")
}
```

### Positional clustering within genes — OncodriveCLUST

Which genes carry mutations piled onto a few codons rather than spread along their
length. Built from the same per-sample MAFs in data/variant_annotations/ as the
oncoplots, using the amino-acid column the annotator emitted
(attend_maf$protein_col, resolved by maf_protein_col()) — positions come from that
column alone, so the chunk is skipped entirely when no annotator supplied one rather
than run on a fallback that would mean something else.

**Why this and not MutSigCV.** MutSigCV asks whether a gene is mutated above a
*background rate* it estimates from the cohort's own mutations plus covariates. Two
properties of this cohort break that: ~40 samples make the per-gene rate estimate
unstable, and the data are **tumour-only**, so residual germline inflates the counts
in the ancestry-dependent way [00-methods](00-methods.html) documents for TMB — and
MutSigCV has no germline model, so it would absorb that leakage into the background.
OncodriveCLUST instead compares each gene's mutations against *its own* positional
distribution, needing no cohort-wide rate model, which is why it degrades gracefully
here. The two are not substitutes: this identifies hotspot-like genes, not drivers.
For a driver list, TCGA's published significantly-mutated genes are the external
reference, the same role attend_scna$panel plays for peaks.

Exploratory, and the q-values are its own background estimate over the top genes,
not a genome-wide claim. Read it as a shortlist, not a result.

```{r oncodrive, eval=ONCO_OK, fig.height=5}
prot <- maf_protein_col(maf_long)
if (!length(prot)) {
  cat("No protein-change column in the MAFs (looked for: ",
      paste(attend_maf$protein_col, collapse = ", "),
      "). OncodriveCLUST needs codon positions, so this is skipped.\n", sep = "")
} else {
  # minMut = 3 matches the Fisher comparison above: a gene mutated once or twice has no
  # positional distribution to test, and oncodrive() errors rather than skips on some
  # such inputs, so the filter is set here rather than left to the default.
  od <- tryCatch(maftools::oncodrive(maf = maf, AACol = prot[1], minMut = 3,
                                     pvalMethod = "zscore"),
                 error = function(e) { message("oncodrive(): ", conditionMessage(e)); NULL })
  if (is.null(od) || !nrow(od)) {
    cat("OncodriveCLUST returned nothing at minMut = 3 — no gene has enough mutations ",
        "to test for positional clustering.\n", sep = "")
  } else {
    cat("Protein-change column used:", prot[1],
        "| genes tested:", nrow(od),
        "| fdr < 0.1:", sum(od$fdr < 0.1, na.rm = TRUE), "\n\n")
    print(knitr::kable(head(as.data.frame(od), 15), digits = 4,
                       caption = "OncodriveCLUST: positional clustering per gene (exploratory)"))
    if (any(od$fdr < 0.1, na.rm = TRUE))
      maftools::plotOncodrive(res = od, fdrCutOff = 0.1, useFraction = TRUE)
  }
}
```

### Mutation frequency by subgroup — the TCGA Fig. 2d view

**This is the figure the oncoplot cannot give you.** An oncoplot shows which patients are
mutated; it does not say whether a gene is mutated at a *different rate* between
subgroups, and it offers no multiple-testing control.

Following TCGA Fig. 2d, each gene's mutation frequency is computed **within each
subgroup** and the subgroups are shown side by side. The subgroups are the four
scna_group strata — MMR status crossed with aneuploidy class — which is the split the
rest of this report and the classification report use.

**The test.** For each gene, a Fisher exact test across the subgroups, then
Benjamini–Hochberg FDR across the genes tested. Genes reaching **FDR < 0.05** are marked
with an asterisk, exactly as TCGA marks them. Only genes mutated in at least
min_mut patients are tested — testing a gene mutated once produces noise, not a result.

Frequencies are *within-group*: the denominator is that subgroup's own patient count, so
a small subgroup shows large jumps per patient. Group sizes are printed with the table.

```{r mutfreq-groups, eval=ONCO_OK, fig.width=11, fig.height=7}
min_mut  <- 3    # a gene must be mutated in at least this many patients to be tested
top_freq <- 25   # genes shown, ranked by overall mutation frequency

master_grp <- add_scna_group(master)      # mmr_group + scna_group (MMR x aneuploidy)

mut_by_pid <- maf_long |>
  mutate(pid = cw$pid[match(Tumor_Sample_Barcode, cw$barcode)]) |>
  filter(!is.na(pid)) |>
  distinct(pid, Hugo_Symbol)

grp_tbl <- master_grp |>
  filter(!is.na(scna_group)) |>
  select(pid, scna_group) |>
  mutate(scna_group = factor(scna_group, levels = attend_scna$group_levels)) |>
  droplevels()

n_by_grp <- count(grp_tbl, scna_group, name = "n_group")

if (nrow(grp_tbl) > 0 && dplyr::n_distinct(grp_tbl$scna_group) >= 2) {
  print(knitr::kable(n_by_grp, caption = "Patients per subgroup (the denominators below)"))

  genes_keep <- mut_by_pid |>
    semi_join(grp_tbl, by = "pid") |>
    count(Hugo_Symbol, name = "n_mut") |>
    filter(n_mut >= min_mut) |>
    slice_max(n_mut, n = top_freq, with_ties = FALSE) |>
    pull(Hugo_Symbol)

  freq_tbl <- tidyr::crossing(pid = grp_tbl$pid, Hugo_Symbol = genes_keep) |>
    left_join(mut_by_pid |> mutate(mutated = TRUE), by = c("pid", "Hugo_Symbol")) |>
    mutate(mutated = replace_na(mutated, FALSE)) |>
    left_join(grp_tbl, by = "pid") |>
    group_by(Hugo_Symbol, scna_group) |>
    summarise(n_mut = sum(mutated), n = n(), freq = n_mut / n, .groups = "drop")

  # Fisher across subgroups per gene, then BH across genes (TCGA marks FDR < 0.05).
  fdr_tbl <- freq_tbl |>
    group_by(Hugo_Symbol) |>
    summarise(p = {
      m <- rbind(n_mut, n - n_mut)
      tryCatch(stats::fisher.test(m, simulate.p.value = TRUE, B = 1e4)$p.value,
               error = function(e) NA_real_)
    }, .groups = "drop") |>
    mutate(fdr = stats::p.adjust(p, method = "BH"),
           star = if_else(!is.na(fdr) & fdr < 0.05, "*", ""))

  plot_tbl <- freq_tbl |>
    left_join(fdr_tbl, by = "Hugo_Symbol") |>
    mutate(Hugo_Symbol = fct_reorder(paste0(Hugo_Symbol, star), freq, .fun = max))

  print(ggplot(plot_tbl, aes(Hugo_Symbol, freq, fill = scna_group)) +
    geom_col(position = position_dodge(width = 0.8), width = 0.75) +
    coord_flip() +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                       expand = expansion(mult = c(0, 0.05))) +
    labs(title = "Mutation frequency by subgroup (TCGA Fig. 2d convention)",
         subtitle = paste0("Fisher across subgroups, BH-adjusted across ",
                           nrow(fdr_tbl), " genes; * = FDR < 0.05. ",
                           "Frequencies are within-subgroup."),
         x = NULL, y = "Patients mutated (% of subgroup)", fill = NULL) +
    attend_theme() +
    theme(legend.position = "top"))

  print(knitr::kable(fdr_tbl |> arrange(fdr) |> mutate(across(c(p, fdr), ~ signif(.x, 3))),
                     caption = "Per-gene Fisher test across subgroups, BH-adjusted"))
} else {
  message("Fewer than two non-empty scna_group strata — subgroup frequency comparison skipped.")
}
```

## Part 2 — recurrent copy number across the cohort

Two resolutions of the same question, coarsest first. Both are **descriptive frequencies
with no significance testing**: GISTIC 2.0 q-values are the field's standard for
recurrence significance and are not applied to these figures.

### Arm level

Cohort frequency of whole-arm gains and losses from the **ASCETS arm-level calls** — the
recurrent events that define endometrial subtypes, such as 1q and 8q gain in serous-like
tumours. Gains extend right, losses left, and arms are shown in **genome order**
(order_arms_genomic()), not sorted by frequency. Each arm's fraction is out of its own
**evaluable** samples — calls of LOWCOV/NC/NA are excluded from that arm's denominator,
not counted as neutral — and that count is printed on the axis label (arm (n=…)) so a
reduced denominator is visible rather than hidden. The bars below show the
`r attend_cnv$arm_calls$n_barplot` most-altered *eligible* arms out of roughly 44 in the
genome — not every arm — where eligible means evaluable in at least
`r 100 * attend_cnv$arm_calls$min_evaluable_frac`% of samples, the identical rule
top_arms applies to the arm annotation tracks on the oncoplots in Part 1
above, so neither figure can rank a
poorly-covered arm ahead of a well-covered one. An uncallable arm's CNV_<arm> value on
those oncoplots is NA, not "neutral": missing data is not evidence of no alteration.

```{r recurrent-cnv, fig.width=9, fig.height=6}
arm_freq <- recurrent_arm_calls(load_ascets_calls(), n_top = attend_cnv$oncoplot_arms)

if (is.null(arm_freq)) {
  message("No ASCETS arm-level calls (data/ascets/) — skipping recurrent-CNV summary.")
} else {
  # Same eligibility gate recurrent_arm_calls() applies to top_arms (freq$eligible), so
  # the barplot and the oncoplot annotation tracks can never disagree about which arms
  # are well-covered enough to rank. Applied BEFORE slice_head(), so a poorly-covered arm
  # (e.g. an acrocentric p-arm ASCETS marks LOWCOV in most samples) cannot be promoted into
  # the figure just because it scores 100% altered on a 2-sample denominator.
  fr <- arm_freq$freq |>
    filter(eligible) |>
    slice_head(n = attend_cnv$arm_calls$n_barplot) |>
    mutate(arm = factor(arm, levels = rev(order_arms_genomic(arm))))

  # Axis labels carry each arm's evaluable-sample count, so a reader can see when a
  # frequency rests on a reduced denominator (matched by factor level, not row order).
  arm_labels <- setNames(paste0(as.character(fr$arm), " (n=", fr$n_evaluable, ")"),
                         as.character(fr$arm))

  fr_long <- fr |>
    pivot_longer(c(gain, loss), names_to = "direction", values_to = "frac") |>
    mutate(frac = if_else(direction == "loss", -frac, frac))

  ggplot(fr_long, aes(arm, frac, fill = direction)) +
    geom_col(width = 0.8) +
    geom_hline(yintercept = 0, colour = "grey40", linewidth = 0.3) +
    coord_flip() +
    scale_fill_manual(values = c(gain = unname(ATTEND_PALETTE["red"]),
                                 loss = unname(ATTEND_PALETTE["blue"]))) +
    scale_x_discrete(labels = arm_labels) +
    scale_y_continuous(labels = function(x) paste0(abs(round(100 * x)), "%")) +
    labs(title = "Recurrent arm-level SCNAs across the cohort",
         subtitle = "gain (right) vs loss (left), by fraction of evaluable samples; arms in genome order",
         x = NULL, y = "fraction of evaluable samples", fill = NULL) +
    attend_theme()
}
```

### Genome-wide, by chromosomal location

The TCGA Fig. 1a convention: alterations plotted **by position along the genome** rather
than summarised per arm, split into two amplitude panels — **low-level** and **high-level**
(|log2 ratio| >=  `r attend_cnv$pileup$high_abs_logratio`, the conventional
amplification/homozygous-deletion cutoff) — gain up/red, loss down/blue. Grey vertical
lines are chromosome boundaries; dashed lines mark centromeres, read from the arm-boundary
table. ChrY is dropped from the axis — the cohort is uterine and therefore all female, so
it carries no calls and only an empty tail.

Built from the ClassifyCNV altered segments (data/cnv_annotations/) in
`r attend_cnv$pileup$bin / 1e6` Mb bins; the figure caption states the sample denominator
and whether it came from an explicit count, the profiled-sample attribute, or the segments
observed. **These are descriptive frequencies with no significance testing** — GISTIC
q-values are the field's standard for recurrence significance and are not applied here.
This is the figure that shows *focal* recurrence — a narrow tall spike is a focal peak, a
broad plateau is an arm-level event and also appears in the arm barplot above.

```{r segment-pileup, fig.width=12, fig.height=6}
arm_bounds <- load_arm_boundaries()
pile <- segment_pileup(load_cnv_data(), arm_bounds)

if (is.null(pile)) {
  message("No ClassifyCNV segments (data/cnv_annotations/) or arm-boundary table — ",
          "skipping the genome-wide pileup.")
} else {
  # Capture the denominator + its source before any filter()/mutate() that could drop
  # these custom attributes.
  n_samples        <- attr(pile, "n_samples")
  n_samples_source <- attr(pile, "n_samples_source")
  offs             <- attr(pile, "chrom_offsets")

  # Readable phrase for the subtitle, not the internal token (argument/profiled/observed)
  # segment_pileup() records in n_samples_source — a subtitle should read as English.
  n_samples_desc <- c(argument = "an explicitly supplied cohort size",
                      profiled = "the number of samples profiled",
                      observed = "the samples present in the segment data")[[n_samples_source]]

  # ChrY dropped: the cohort is uterine and therefore all female, so chrY carries no
  # calls and would only draw an empty axis tail. ChrX is kept.
  pile <- pile |> filter(chrom != "chrY")
  offs <- offs  |> filter(chrom != "chrY")

  # Centromere = each chromosome's p/q boundary, read from the arm-boundary table
  # (never hard-coded) and converted to genome coordinates with the same offsets used
  # for the chromosome-boundary lines and axis breaks.
  centromeres <- arm_bounds |>
    filter(str_detect(arm, "p$"), chrom != "chrY") |>
    mutate(chrom = as.character(chrom)) |>
    inner_join(offs |> select(chrom, offset), by = "chrom") |>
    transmute(gpos = offset + end)

  pl <- pile |>
    select(gpos, gain_low, gain_high, loss_low, loss_high) |>
    pivot_longer(-gpos, names_to = c("direction", "tier"), names_sep = "_",
                values_to = "frac") |>
    mutate(frac = if_else(direction == "loss", -frac, frac),
           tier = factor(tier, levels = c("low", "high"),
                        labels = c("low-level", "high-level")))

  ggplot(pl, aes(gpos, frac, fill = direction)) +
    geom_col(width = attend_cnv$pileup$bin) +
    geom_hline(yintercept = 0, colour = "grey40", linewidth = 0.3) +
    geom_vline(xintercept = offs$offset[-1], colour = "grey85", linewidth = 0.2) +
    geom_vline(data = centromeres, aes(xintercept = gpos), colour = "grey60",
              linewidth = 0.3, linetype = "dashed") +
    facet_grid(tier ~ ., scales = "free_y") +
    scale_fill_manual(values = c(gain = unname(ATTEND_PALETTE["red"]),
                                 loss = unname(ATTEND_PALETTE["blue"]))) +
    scale_x_continuous(breaks = offs$mid, labels = str_remove(offs$chrom, "^chr"),
                       expand = c(0, 0)) +
    scale_y_continuous(labels = function(x) paste0(abs(round(100 * x)), "%")) +
    labs(title = "Genome-wide recurrent SCNA pileup (TCGA Fig. 1a convention)",
         subtitle = paste0("gain (up) vs loss (down), fraction of ", n_samples,
                           " samples (", n_samples_desc, "); ",
                           attend_cnv$pileup$bin / 1e6,
                           " Mb bins; dashed lines = centromeres; no significance testing applied"),
         x = "chromosome", y = "fraction of samples", fill = NULL) +
    attend_theme() +
    theme(panel.grid.major.x = element_blank(),
          panel.grid.minor.x = element_blank())
}
```

---

## Part 3 — the aneuploidy x MMR contrast

Either the MMRd-high group reaches aneuploidy the same way MMR-proficient serous-like
tumours do, in which case it should carry the same recurrent peaks, or it gets there by a
different route and the peak profile should differ. With n = `r n_mmrd_high` in the
MMRd-high group, the endpoint is a single pre-specified composite score rather than a
per-peak test.

### Setup and the two ID spaces

GISTIC and .seg sample ids are sequencing **barcodes**; the master is keyed on **pid**.
Joining them directly gives an all-NA result *with no error*, so every lookup below goes
through group_of_barcode(), built on the same crosswalk as everything else.

The grouping variable scna_group composes is_mmrd() (IHC-based, MSI deliberately not a
criterion) with aneuploidy_class at the pre-registered cut
`r attend_thresholds$aneuploidy`, the same definition Part 1's
frequency comparison uses.

```{r scna-gate}
have_master <- exists("master") && !is.null(master) && nrow(master) > 0
if (have_master && !"scna_group" %in% names(master)) master <- add_scna_group(master)

cw_scna <- tryCatch(build_barcode_pid(load_gianlu_clinical_data(), make_id_cfg()),
                    error = function(e) NULL)
have_cw <- !is.null(cw_scna) && nrow(cw_scna) > 0

group_of_barcode <- function(ids) {
  if (!have_master || !have_cw)
    return(factor(rep(NA_character_, length(ids)), levels = attend_scna$group_levels))
  barcode_scna_group(ids, master, cw_scna)
}

gx_dir   <- here("data", attend_cnv$gistic$dir)
run_dirs <- c(all       = file.path(gx_dir, "all"),
              mmrp_high = file.path(gx_dir, "mmrp_high"),
              mmrp_low  = file.path(gx_dir, "mmrp_low"),
              mmrd_low  = file.path(gx_dir, "mmrd_low"),
              mmrd_high = file.path(gx_dir, "mmrd_high"))
have_gistic <- dir.exists(run_dirs[["all"]])

if (!have_master) {
  cat("Master table absent — run report 01 first.\n")
} else if (!have_gistic) {
  cat("**GISTIC outputs absent** (`", run_dirs[["all"]], "`).\n\n",
      "Run `MODE=groups sbatch code/run_gistic.sh`, then `MODE=loo`. ",
      "this report is skipped, not silently empty.\n", sep = "")
}
```

Peaks come from five runs; per-sample calls come from one. Five GISTIC2 runs feed this
report: the pooled cohort and the four scna_group strata. GISTIC's background model is
cohort-specific, so the same tumour gets a different thresholded call depending on which
cohort it was run in, and comparing calls across runs would compare five different null
models. The two roles are therefore kept apart:

- **Peak definition** uses the union of significant peaks across all five runs
  (gistic_peak_union()), so a peak private to the small MMRd-high stratum, which the
  pooled run's genome-wide threshold would bury, is still discoverable.
- **Per-sample calls** are read only from the pooled run's all_lesions.conf_*.txt. No
  per-sample call here comes from a stratified run.

The cohort this contrast conditions on. Fisher tests whether MMR status and aneuploidy class are
associated at all — if they are strongly associated, the MMRd-high cell is small by
construction and everything below is underpowered.

```{r scna-contingency, eval=have_master}
if (all(c("mmr_group", "aneuploidy_class") %in% names(master))) {
  tab2 <- table(MMR = master$mmr_group, Aneuploidy = master$aneuploidy_class)
  print(knitr::kable(tab2, caption = "MMR x aneuploidy-class contingency (patient level)"))
  ft2 <- tryCatch(fisher.test(tab2), error = function(e) NULL)
  if (!is.null(ft2)) cat("Fisher p =", signif(ft2$p.value, 3),
                        ", OR =", signif(unname(ft2$estimate), 3), "\n\n")
}
```

### The primary endpoint — a pre-specified serous-like panel score

**Why a composite score rather than testing peaks one at a time.** At n ≈ 9 no individual
peak has the power to show a difference, and testing hundreds of peaks in a group that
small produces false positives, not findings. A single composite score over a *fixed,
externally-specified* set of loci concentrates the evidence into one test.

**What the panel is.** The 12 loci (attend_scna$panel) are fixed *a priori* from the
TCGA UCEC recurrent-peak list (Kandoth et al. 2013). No locus was added, removed, or
re-directioned after looking at ATTEND's own GISTIC output. Each locus is
**directional** — a deletion at *MYC* does not count as serous-like — matched by
match_panel_peaks().

**Why the panel is fixed in advance, and what that costs.** A data-driven panel — the loci
that best separate MMRd-high from MMRp-high *in ATTEND* — is a legitimate analysis, but it
is not this one, and it cannot be tested the way this one is. Selecting loci uses the group
labels, so the resulting score is no longer independent of those labels under the null;
permuting the labels afterwards does not undo the selection, because the selection was made
on the unpermuted labels. The permutation null then understates the variability actually
present and the nominal p is not the type-I error rate (Kriegeskorte et al., *Nat Neurosci*
2009, on circular analysis; the general problem is post-selection inference). At n ≈ 9
against a few hundred candidate peaks, selection alone can manufacture near-complete
separation from noise, so this is not a small correction.

Three routes make a data-driven panel valid, and each is a different study: **hold-out**,
where loci are chosen on one split and scored on an untouched one — impossible at this n;
**selection nested inside the resampling**, where every permutation re-derives its own
panel from scratch so the null absorbs the selection — valid here, and strictly less
powerful than the fixed panel, because the null it builds is wider; or **a global test that
selects nothing**, such as a permutational multivariate test across the whole peak matrix,
which is data-driven in that it uses every locus but answers "is the profile different at
all?" rather than "is it serous-like?".

So pre-specification is not a claim to the better analysis. It buys one thing: a single
nominal p that can be read at face value. It costs power against any difference the panel
is not aimed at — if these tumours reach aneuploidy by a route TCGA's serous tumours do not
use, a fixed serous panel is pointed the wrong way and returns a null result that means
"not serous-like", never "not different". Choosing the fixed panel here is choosing to
spend the little power n ≈ 9 affords on the one hypothesis that has external support,
rather than to spread it across every peak or to buy apparent significance with selection.
A data-driven panel found in these data would be reported as **hypothesis-generating**,
alongside this test rather than in place of it.

Loci with no matching pooled-run peak of the right direction **leave the denominator**
rather than counting as unaltered, so the score is a fraction of *resolvable* loci.

The test is a **label permutation** (B = `r attend_scna$perm_B`), which makes no
distributional assumption — appropriate at this n. A Wilcoxon is reported alongside.
A permutation p is bounded below by 1/(B+1), so a floor-level result is reported as
p < … rather than as an exact tiny value.

```{r scna-panel-score, eval=have_master && have_gistic, fig.height=5}
pooled <- load_gistic_lesions_at(run_dirs[["all"]])
if (is.null(pooled)) {
  cat("Pooled GISTIC run (`", run_dirs[["all"]], "`) not readable — panel score skipped.\n", sep = "")
} else {
  scores <- panel_score(pooled$mat, pooled$peaks)
  grp    <- group_of_barcode(rownames(pooled$mat))

  cat("Panel resolved", attr(scores, "n_loci"), "of 12 pre-specified loci to a pooled-run ",
      "peak of the matching direction",
      if (attr(scores, "n_loci") > 0) paste0(" (", paste(attr(scores, "loci"), collapse = ", "), ").")
      else ".", "\n\n")

  p1 <- panel_score_plot(scores, grp,
                         title = "Composite TCGA UCEC serous-like panel score, by subgroup")
  if (!is.null(p1)) print(p1)

  keep2 <- !is.na(grp) & grp %in% c("MMRd-high", "MMRp-high")
  if (sum(keep2) >= 3 && dplyr::n_distinct(droplevels(grp[keep2])) == 2) {
    g2 <- droplevels(grp[keep2])
    pt <- perm_test_two_group(scores[keep2], g2, B = attend_scna$perm_B, seed = 1)
    wt <- tryCatch(stats::wilcox.test(scores[keep2] ~ g2), error = function(e) NULL)

    floor_p <- 1 / (pt$B + 1)
    p_txt <- if (pt$p <= floor_p + 1e-12) {
      sprintf("p < %s (bounded at the B = %s resolution floor)",
              format(floor_p, scientific = TRUE, digits = 1), format(pt$B, big.mark = ","))
    } else sprintf("p = %.4g", pt$p)

    cat("**MMRd-high (n =", sum(g2 == "MMRd-high"), ") vs MMRp-high (n =", sum(g2 == "MMRp-high"),
        ")**: observed difference in mean panel score =", signif(pt$stat, 3),
        "; label-permutation ", p_txt, ".",
        if (!is.null(wt)) paste0(" Wilcoxon p = ", signif(wt$p.value, 3), ".") else "",
        "\n\nIf the 9 reach aneuploidy by a different route, the score separates from ",
        "MMRp-high; if they phenocopy copy-number-high tumours, it does not.\n\n")
  } else {
    cat("Fewer than 3 samples resolved to both MMRd-high and MMRp-high — permutation test skipped.\n\n")
  }
}
```

### The exploratory counterpart — a data-driven panel

The panel above is aimed at the TCGA serous phenotype, so a null result there means "not
serous-like", never "not different". This asks the larger question the fixed panel cannot:
is there **any** directional peak set that separates MMRd-high from MMRp-high? The loci are
chosen from the same pooled-run all_lesions.conf_*.txt matrix, by the signed difference in
mean thresholded call between the two groups — magnitude ranks them, sign fixes each one's
direction, so the same concordant-events-only rule applies as for the pre-specified panel.

**This is Family B and can never be Family A.** The loci are chosen using the group labels,
so permuting those labels afterwards would not undo the selection and the resulting null
would be far too narrow. perm_test_selected_panel() therefore **re-runs the selection inside
every permutation replicate**, on the permuted labels, which is what restores the p-value's
meaning. Measured on pure noise, the hoisted-selection version returns the minimum possible
p on 100% of datasets while the nested version is uniform —
`code/tests/test_nested_selection_permutation.R` pins that calibration.

Three numbers make this readable, and all three are printed: the nested p, the selected
loci, and each locus's **leave-one-out selection frequency**. At n ≈ 9 per group a panel
whose membership turns over when one patient is dropped is not a finding, and a low LOO
fraction says so directly. The global burden test below it selects nothing at all, so it
stays interpretable when both panels come back null.

```{r scna-selected-panel, eval=have_master && have_gistic, fig.height=5}
if (exists("pooled") && !is.null(pooled)) {
  grp_sel <- group_of_barcode(rownames(pooled$mat))
  keep_s  <- !is.na(grp_sel) & grp_sel %in% c("MMRd-high", "MMRp-high")
  if (sum(keep_s) >= 6 && dplyr::n_distinct(droplevels(grp_sel[keep_s])) == 2) {
    m_sel <- pooled$mat[keep_s, , drop = FALSE]
    g_sel <- droplevels(grp_sel[keep_s])
    ds    <- perm_test_selected_panel(m_sel, g_sel, n_select = attend_scna$select_n,
                                      B = attend_scna$select_perm_B, seed = 1)
    floor_s <- 1 / (ds$B + 1)
    cat("**Data-driven panel (Family B, exploratory).** ", attend_scna$select_n,
        " loci selected from ", ncol(m_sel), " pooled-run peaks; nested permutation B = ",
        format(ds$B, big.mark = ","), ". Observed difference in mean score = ",
        signif(ds$stat, 3), "; ",
        if (ds$p <= floor_s + 1e-12) paste0("p < ", signif(floor_s, 2), " (resolution floor)")
        else paste0("p = ", signif(ds$p, 4)), ".\n\n", sep = "")
    print(knitr::kable(
      data.frame(locus = ds$loci, direction = ds$dir,
                 loo_selection_frac = round(ds$loo_frac, 2)),
      caption = paste("Selected loci and how often each survives leave-one-out;",
                      "a low fraction means the panel is an artefact of one patient")))
    # Same plot type as the pre-specified score, relabelled: asserting "TCGA UCEC panel"
    # on this figure would claim provenance this panel does not have.
    sel <- select_panel_loci(m_sel, g_sel, attend_scna$select_n)
    ps  <- rowMeans(.directional_altered(m_sel, sel$idx, sel$dir), na.rm = TRUE)
    p2 <- panel_score_plot(ps, g_sel,
            title = "Data-driven panel score (exploratory, Family B)",
            ylab  = "Fraction of selected loci altered (0-1)")
    if (!is.null(p2)) print(p2)

    gb <- perm_test_global_burden(m_sel, g_sel, B = attend_scna$perm_B, seed = 1)
    fg <- 1 / (gb$B + 1)
    cat("\n**Global burden, no selection.** Mean fraction of all ", ncol(m_sel),
        " peaks altered in either direction; difference = ", signif(gb$stat, 3), "; ",
        if (gb$p <= fg + 1e-12) paste0("p < ", signif(fg, 2)) else paste0("p = ", signif(gb$p, 4)),
        ".\n\n", sep = "")
  } else {
    cat("Fewer than 6 samples across both MMRd-high and MMRp-high — the data-driven ",
        "panel is skipped; selection needs more samples than the fixed panel, not fewer.\n",
        sep = "")
  }
}
```

#### Which loci drive the score

The composite above is the primary endpoint; this breaks it into its 12 parts as a
**secondary, more granular** view. Each locus is tested one at a time and **Holm-corrected
within the 12-locus family**.

Holm is applied over **all 12 pre-specified loci**, including any that did not resolve to
a pooled-run peak (p = NA) — not just the testable ones. Shrinking the family to the
loci that happened to resolve would inflate significance on the survivors.

```{r scna-locus-table, eval=have_master && have_gistic}
if (exists("pooled") && !is.null(pooled)) {
  mp <- match_panel_peaks(pooled$peaks, attend_scna$panel)
  ft_all <- scna_freq_table(pooled$mat, grp)
  rows <- lapply(seq_len(nrow(mp)), function(i) {
    pk <- mp$peak_id[i]
    if (is.na(pk)) return(data.frame(locus = mp$locus[i], direction = mp$direction[i],
                                     freq_mmrd_high = NA_real_, freq_mmrp_high = NA_real_,
                                     p = NA_real_, stringsAsFactors = FALSE))
    a <- ft_all[ft_all$peak_id == pk & ft_all$scna_group == "MMRd-high", ]
    b <- ft_all[ft_all$peak_id == pk & ft_all$scna_group == "MMRp-high", ]
    p <- NA_real_
    if (nrow(a) && nrow(b)) {
      tt <- matrix(c(a$n_altered, a$n - a$n_altered, b$n_altered, b$n - b$n_altered), nrow = 2)
      p <- tryCatch(stats::fisher.test(tt)$p.value, error = function(e) NA_real_)
    }
    data.frame(locus = mp$locus[i], direction = mp$direction[i],
               freq_mmrd_high = if (nrow(a)) a$freq else NA_real_,
               freq_mmrp_high = if (nrow(b)) b$freq else NA_real_,
               p = p, stringsAsFactors = FALSE)
  })
  locus_tbl <- do.call(rbind, rows)
  locus_tbl$p_holm <- family_adjust(locus_tbl$p, "A", n = nrow(locus_tbl))
  knitr::kable(locus_tbl, digits = 3,
               caption = "The 12 pre-specified TCGA UCEC loci, Holm-adjusted within family (secondary to the composite score)")
}
```

### Where each subgroup's calls sit on the genome

Peak-call frequency from the pooled run, amplifications above the axis and deletions
below, split by subgroup and chromosome. The visual counterpart to the composite score:
if MMRd-high really phenocopies MMRp-high, their two panels should have peaks in the same
places.

```{r scna-mirror, eval=have_master && have_gistic}
if (exists("pooled") && !is.null(pooled)) {
  grp3      <- group_of_barcode(rownames(pooled$mat))
  freq_tbl3 <- scna_freq_table(pooled$mat, grp3)
  p3 <- scna_mirror_plot(freq_tbl3, pooled$peaks,
                         title = "Pooled-run peak-call frequency by subgroup (amp above / del below)")
  if (!is.null(p3)) print(p3)
}
```

How much the five runs' significant-peak sets overlap, as pairwise Jaccard. Low overlap
between a stratum and the pooled run means that stratum has genuinely private peaks —
which is the case the union-of-five design exists to catch. Peaks private to a single run
are listed separately.

```{r scna-peakset, eval=have_master && have_gistic}
union_peaks <- gistic_peak_union(run_dirs)
if (is.null(union_peaks)) {
  cat("No GISTIC lesions parsed from any of the five run folders — skipped.\n")
} else {
  pss <- peak_set_structure(union_peaks)
  print(knitr::kable(round(pss$jaccard, 2),
                     caption = "Jaccard overlap of significant-peak sets between the five GISTIC runs"))
  if (nrow(pss$private))
    print(knitr::kable(pss$private,
                       caption = "Peaks private to a single run (no overlapping peak in any other)"))
}
```

### Leave-one-out stability of the MMRd-high peaks

**The asymmetric rule this rests on.** A peak *present* in the MMRd-high list, supported by
several of the 9 tumours and stable to leave-one-out, is **informative**. A peak *absent*
from that list is **uninformative** — at n = 9 absence is indistinguishable from low power.
Presence carries evidence; absence does not. That is why the table reports support counts
and retention fractions for MMRd-high peaks rather than a between-group presence/absence
verdict.

The retention fraction is the proportion of the 9 leave-one-out GISTIC re-runs in which a
peak survives. A peak retained in 9/9 does not depend on any single tumour; a peak
retained in 3/9 is one or two tumours' worth of signal.

**This is a descriptive robustness diagnostic, not a calibrated error bound** — informal
stability selection (Meinshausen & Bühlmann 2010) adapted to n = 9. Canonical stability
selection uses ⌊n/2⌋ subsamples and an exchangeability assumption that leave-one-out at
this n only approximates. It is not an FDR statement.

Support counts carry Wilson 95% confidence intervals, which behave sensibly at small n
and at frequencies near 0 or 1 where a normal-approximation interval does not.

```{r scna-loo, eval=have_master && have_gistic}
if (exists("union_peaks") && !is.null(union_peaks)) {
  mmrd_high <- load_gistic_lesions_at(run_dirs[["mmrd_high"]])
  if (is.null(mmrd_high)) {
    cat("MMRd-high GISTIC run (`", run_dirs[["mmrd_high"]], "`) absent — cannot list its peaks.\n", sep = "")
  } else {
    # Soft check: the GISTIC run's sample composition may predate a later threshold
    # change on the master, so a mismatch is a warning, not an error.
    gb_supp <- group_of_barcode(rownames(mmrd_high$mat))
    if (any(!is.na(gb_supp) & gb_supp != "MMRd-high")) {
      warning("MMRd-high GISTIC run contains ", sum(!is.na(gb_supp) & gb_supp != "MMRd-high"),
              " sample(s) no longer classified MMRd-high under the current master — ",
              "the GISTIC run may predate a threshold change")
    }

    supp <- scna_freq_table(mmrd_high$mat,
                            factor(rep("MMRd-high", nrow(mmrd_high$mat)),
                                   levels = attend_scna$group_levels))
    supp <- supp[supp$scna_group == "MMRd-high", ]

    loo_dir <- file.path(gx_dir, "loo")
    loo_folders <- if (dir.exists(loo_dir))
      list.dirs(loo_dir, full.names = TRUE, recursive = FALSE) else character(0)
    loo <- if (length(loo_folders)) loo_stability(mmrd_high$peaks, loo_folders) else NULL

    tbl <- merge(supp[, c("peak_id", "n_altered", "n", "freq", "ci_lo", "ci_hi")],
                 mmrd_high$peaks[, c("peak_id", "descriptor", "chrom", "direction")],
                 by = "peak_id")
    if (!is.null(loo)) {
      tbl <- merge(tbl, loo, by = "peak_id", all.x = TRUE)
    } else {
      tbl$n_loo <- NA_integer_; tbl$n_retained <- NA_integer_; tbl$retained_frac <- NA_real_
    }
    tbl <- tbl[order(-tbl$freq), ]
    print(knitr::kable(tbl, digits = 2,
          caption = "Every MMRd-high peak: support count with Wilson 95% CI, and leave-one-out retention fraction (descriptive, not a calibrated bound)"))
    if (is.null(loo)) cat("\nnote: no leave-one-out GISTIC output under `", loo_dir,
                          "` — retention columns are NA; the support counts stand alone.\n\n", sep = "")
  }
}
```

```{r session-info}
sessionInfo()
```
