---
title: "ATTEND — 7. TCGA integrated classification (Pearson/Ward clustering)"
author: "sceriff0"
date: "2026-08-09"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
# renv: tidyverse, here, ggpubr (+ optional bioc::ComplexHeatmap, circlize for the
# faithful Fig-1a heatmap; cluster for the silhouette; survival/ggsurvfit for the
# Fig-1b KM). If missing: renv::install(c("ggpubr","bioc::ComplexHeatmap","circlize",
# "cluster","survival","ggsurvfit")); renv::snapshot()
library(tidyverse)
library(here)
library(ggpubr)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 7)

source(here("code", "attend_harmonise.R"))   # make_id_cfg(), build_barcode_pid()
source(here("code", "attend_classes.R"))     # attend_cnv/_pole, arm + cascade helpers
source(here("code", "attend_plots.R"))       # fig1a_heatmap(), feature_positions(), km_by_cluster()
source(here("code", "attend_io.R"))          # read_intermediate()
source(here("code", "load_clinical.R"))      # load_gianlu_clinical_data()
source(here("code", "load_wes_results.R"))   # load_cnv_data(), load_ascets_*(), load_tmb_scores()

# --- CNV clustering distance and linkage -----------------------------------------
# 1 - Pearson correlation distance with Ward.D2 linkage. TCGA's Suppl. Methods S2 leaves
# the *copy-number* clustering metric unstated, but its mRNA and methylation clusterings
# both use 1 - Pearson / Ward, so the same pair is applied here to the CN matrix. This is
# the only distance/linkage combination the pipeline uses.
cnv_method <- "ward.D2"       # Ward linkage
cnv_dist   <- "correlation"   # 1 - Pearson correlation (cluster_arm_matrix() correlation path)

# Cut height for the DOWNSTREAM cascade / CN-high-burden label. TCGA found 4 SCNA clusters
# (Fig. 1a); ATTEND has no POLE and is smaller, but k = 4 keeps parity. Change freely.
k_cnv <- 4

# The chromosome heatmap is drawn ONCE and annotated with a cluster bar for every k in
# this range, so k = 2..8 can be compared in one figure (fig1a_heatmap_ksweep()).
k_sweep <- 2:8
```

## What this report does

Reproduces the **TCGA integrated genomic classification** of endometrial carcinoma
(Kandoth et al., *Nature* 2013 — `papers/nature12113.pdf`) on the ATTEND cohort, and asks
**which copy-number cluster is enriched for aneuploidy** — the hallmark of the
*copy-number-high / serous-like* group.

> ### How the serous-like cluster is chosen, and why it is directional
>
> Clustering copy-number profiles produces k unlabelled clusters. TCGA never had to decide
> which one was serous-like: with 373 tumours and histology in hand, cluster 4 was visibly
> 94% serous. ATTEND has no such label, so the call has to be **inferred from the copy-number
> data itself** — and that inference step is **not part of TCGA's methodology**. Nothing
> about the clustering below departs from TCGA; only this labelling step is ours.
>
> The obvious rule — pick the cluster with the largest **mean |value|**, i.e. the most
> altered one — turns out to answer the wrong question. "Most altered" is not "most
> serous-like", and on *thresholded* features the two come apart: an undirected score
> rewards a deletion-dominated cluster just as much as the amplification-driven profile
> that actually characterises serous-like disease.
>
> This report therefore scores clusters **directionally**: only *concordant* events count —
> **gains at amplification peaks, losses at deletion peaks**. It is the same principle
> `attend_scna$panel` already applies to the pre-specified locus panel in report 5
> (*a deletion at MYC must not score as serous-like*).
>
> ⚠️ **Both calls are computed and printed, and a `[DISAGREE]` flag is raised when they
> differ.** If they agree on ATTEND, the directional correction changed nothing here and
> the result is a confirmation. Read that flag before drawing any conclusion from the
> cluster labels.

This report reproduces the **TCGA integrated genomic classification** of endometrial
carcinoma (Kandoth et al., *Nature* 2013 — `papers/nature12113.pdf`) on the ATTEND
cohort, and asks **which copy-number cluster is enriched for aneuploidy** — the
hallmark of the *copy-number-high / serous-like* group.

The TCGA classification is a **priority cascade** (their Fig. 2b). **POLE is omitted
here** — ATTEND has no POLE-ultramutated cases — so the cascade **starts at MMRd**:

1. **MMRd** — MMR-protein loss by **IHC** (`gianlu__MMR_STATUS` Deficient). This is the
   top of the cascade and overwrites the CN split. The split is **MMRd vs MMRp by IHC
   alone**; MSI-high by the NGS caller is a **correlate**, not a criterion.
2. **Copy-number high (serous-like)** — many broad SCNAs, **extensive aneuploidy**.
   How the call is made is set by `attend_tcga_serous_rule`, defaulting to the **faithful
   TCGA rule `"cluster_only"`**: serous-like = membership in the high-aneuploidy CNV
   cluster, exactly as TCGA defined it (Fig. 2b, "serous-like ... copy-number cluster 4").
   *TP53* is **not** required here — it is a **property** of the cluster (~90% mutant),
   which we **validate** below rather than gate on. Alternatives: `"and"` (also require
   TP53 — higher precision), `"or"` (union), `"tp53_only"` (portable).
3. **Copy-number low (endometrioid)** — MMR-proficient, few broad SCNAs, everything not
   in the high-aneuploidy cluster.

The copy-number split comes from **clustering copy-number profiles** (their Fig. 1a;
cluster 4 = serous-like), then resolving the serous-like call per `attend_tcga_serous_rule`.
Because the default is `"cluster_only"`, the serous call is earned from copy number alone;
the **TP53-enrichment QC** below then confirms the cluster is genuinely ~90% *TP53*-mutant,
i.e. that it corresponds to TCGA's serous-like group — a non-circular validation.

## Data

```{r data}
master <- tryCatch(
  read_intermediate("attend_master_joined") |> add_molecular_classes(),
  error = function(e) { message("master unavailable: ", conditionMessage(e)); NULL }
)

# barcode -> pid crosswalk (the same derivation every other report uses)
cw <- tryCatch(
  build_barcode_pid(load_gianlu_clinical_data(), make_id_cfg()),
  error = function(e) { message("crosswalk unavailable: ", conditionMessage(e)); NULL }
)

have_master <- !is.null(master)
have_cw     <- !is.null(cw)

# TMB definitions available on the master: upstream first, then the recomputed set.
tmb_defs <- c(upstream = attend_cols$tmb, attend_cols$tmb_set)
if (have_master) tmb_defs <- tmb_defs_present(master)
```

## Copy-number ingestion — GISTIC thresholded peaks (the only clustering input)

This report clusters **exactly what TCGA clustered** (Kandoth et al., Suppl. Methods S2):
*"thresholded relative copy number data in significantly reoccurring amplifications or
deletions regions identified by GISTIC 2.0."* There is **one** clustering feature and no
fallbacks — the arm-level / bin-level proxies have been removed. The matrix is built by
`load_gistic_thresholded()` from GISTIC output in `data/gistic/`:

- input `all_thresholded.by_genes.txt` — genes × samples, values in **{−2,−1,0,+1,+2}**;
- restricted to the genes inside the **significant amp/del peaks** (`amp_genes`/`del_genes.conf`);
- result: a sample × peak-gene **thresholded** matrix — TCGA's exact feature space.

**Producing it requires running GISTIC on the DRAGEN `.seg`** (see *"Running GISTIC on the
cluster"* at the bottom). Until `data/gistic/` is populated, every CNV/cascade section
renders as scaffolding.

> **Computational steps.** (1) `load_gistic_thresholded()` reads GISTIC's
> `all_thresholded.by_genes.txt` and keeps only genes in the significant peaks. (2) The
> per-sample aneuploidy **score** (external, for the enrichment check only) is loaded
> separately (ASCETS if present, else the master). The clustering never sees the aneuploidy
> score — that keeps the enrichment plot non-circular.

```{r cnv-load}
cnv_mat    <- tryCatch(load_gistic_thresholded(), error = function(e) NULL)  # ID x peak-gene {-2..+2}
cnv_source <- "GISTIC 2.0 thresholded peaks"

# per-sample aneuploidy SCORE — used ONLY for the enrichment cross-check, not for clustering
ascets_aneu <- tryCatch(load_ascets_aneuploidy(), error = function(e) NULL)

have_cnv <- !is.null(cnv_mat) && nrow(cnv_mat) >= k_cnv
# cat(), not message(): opts_chunk sets message = FALSE, so a message() here is invisible on
# the knitted page and a skipped CNV section then looks like nothing happened at all.
if (!have_cnv) {
  cat("No GISTIC matrix with >=", k_cnv, "samples — run GISTIC (see the bottom of this",
      "report) so data/gistic/all_thresholded.by_genes.txt exists.",
      "CNV sections are scaffolding until then.\n")
} else {
  cat("Clustering input:", cnv_source, "—", nrow(cnv_mat), "samples x",
      ncol(cnv_mat) - 1, "peak genes.\n")
}
```

## The clustering feature — GISTIC thresholded peaks, and how it matches TCGA

The clustering feature is a **sample × peak-gene** matrix of **thresholded** GISTIC calls
({−2..+2}), restricted to the significantly recurrent amp/del regions — **identical in
kind to TCGA's** (Suppl. Methods S2). This variant hierarchically clusters it with
**1 − Pearson correlation distance and Ward.D2 linkage** (rather than Euclidean +
complete). S2 names no metric for the *copy-number* clustering, but its mRNA/methylation
clusterings use exactly this **1 − Pearson / Ward** pair — so this report asks what the CN
groups look like under a correlation-shaped, variance-minimising linkage rather than the
Euclidean/complete default. Instead of committing to one cut, the chromosome heatmap below
is annotated with the cluster assignment at **every k from 2 to 8** in a single figure.
What this gets right vs the old arm/bin proxies (now removed):

- **Thresholded, not continuous** — {−2..+2} discretised calls, exactly S2.
- **GISTIC significant peaks, not genome-wide** — the feature space is the recurrent
  amp/del peak genes (MYC 8q24, ERBB2 17q12, CCNE1 19q12, 1q, …), exactly S2.
- **Residual differences from TCGA** — platform (DRAGEN WES `.seg` vs SNP 6.0), cohort
  (metastatic vs primary), N, and the unspecified CN metric/linkage (here **1 − Pearson /
  Ward.D2**, S2's mRNA/methylation choice; switch via `cluster_arm_matrix(..., method=,
  dist_method=)`).

## CNV clustering (GISTIC thresholded peaks)

> **Computational steps.** (1) Cluster the sample × peak-gene **thresholded** GISTIC matrix
> `cnv_mat` with `cluster_arm_matrix(k = `r k_cnv`, method = "`r cnv_method`", dist_method =
> "`r cnv_dist`")` — **1 − Pearson correlation** distance, **Ward.D2** linkage (S2's
> mRNA/methylation choice). (2) Label the **copy-number-high**
> cluster by intrinsic SCNA burden = mean |thresholded value| per cluster
> (`cnv_high_cluster_burden()`), independent of the aneuploidy score. (3) Heatmap: peak
> genes (rows) × samples (columns by dendrogram), annotated by CN cluster, the **integrated
> TCGA class** (POLE [no cases] > MMRd > CN-high > CN-low, from the `add_tcga_class()` cascade
> shown below — the same call, drawn here on the Fig-1a clustering), MMR, TP53, **and the
> per-sample aneuploidy score**. That aneuploidy bar is the external ASCETS aneuploidy value
> (`data/ascets/`, `load_ascets_aneuploidy()`) or — when ASCETS is absent — the master's
> `aneu__aneuploidy_score` (from `data/aneu_scores.csv` via report 1) mapped from `pid` back
> onto sequencing barcodes through the report-01 crosswalk. It is a **continuous annotation,
> not a clustering feature**: the clustering never sees it, so visible alignment between the
> high-burden cluster and the high-aneuploidy samples is corroboration, not circularity.
>
> **vs TCGA.** *Follows:* Suppl. Methods S2 exactly — thresholded CN in GISTIC significant
> peaks, hierarchical clustering. *Differs:* DRAGEN WES `.seg` vs SNP6; **1 − Pearson / Ward.D2**
> (S2 leaves the CN metric unspecified but uses this pair for mRNA/methylation); the cascade
> below reads a single k = `r k_cnv`, while the k-sweep heatmap shows k = 2..8; the
> integrated-class and aneuploidy-score annotation bars are ATTEND additions, not Fig. 1a elements.

```{r cnv-cluster, fig.height=8}
cl <- if (have_cnv) cluster_arm_matrix(cnv_mat, k = k_cnv,
                                       method = cnv_method, dist_method = cnv_dist) else NULL

if (!is.null(cl)) {
  # Copy-number-high cluster from the matrix's OWN SCNA burden — derived only from the
  # clustering features, so the aneuploidy-by-cluster boxplot below is an independent check,
  # not a tautology.
  #
  # ⚠️ DIRECTIONAL (this is the ONLY change from the sibling report). The burden is scored
  # over CONCORDANT events only — gains at GISTIC amplification peaks, losses at deletion
  # peaks — instead of mean |value|. On thresholded GISTIC calls the two disagree: |value|
  # rewards a deletion-dominated cluster as strongly as an amplification-dominated one, and
  # on thresholded features it can pick the wrong cluster outright. `direction`
  # comes from this run's own amp_genes/del_genes lists; if they are unavailable the helper
  # falls back to |value| and says so, i.e. this report degrades to the sibling's behaviour.
  feat_dir     <- gistic_feature_direction()
  high_cluster <- cnv_high_cluster_burden(cl, direction = feat_dir)
  burden_tbl   <- cluster_scna_burden(cl, direction = feat_dir)
  high_absolute <- cnv_high_cluster_burden(cl)          # the undirected call, for comparison
  message("Copy-number-high cluster (DIRECTIONAL burden): ", high_cluster,
          " | undirected |value| would pick: ", high_absolute,
          if (identical(high_cluster, high_absolute)) "  [agree]" else "  [DISAGREE]")

  # Per-barcode aneuploidy, kept ONLY for the enrichment check + heatmap annotation:
  # prefer ASCETS, else map the master's aneuploidy score back onto barcodes via the crosswalk.
  aneu_id <- NULL
  if (!is.null(ascets_aneu)) {
    aneu_id <- ascets_aneu |> rename(aneuploidy_value = ascets_aneuploidy)
  } else if (have_master && have_cw) {
    aneu_id <- cw |>
      left_join(master |> transmute(pid, aneuploidy_value =
                  suppressWarnings(as.numeric(.data[[attend_cols$aneuploidy]]))), by = "pid") |>
      transmute(ID = barcode, aneuploidy_value)
  }

  # Heatmap of the GISTIC thresholded peak matrix (TCGA Fig. 1a). Peak genes (rows) x
  # samples (columns by dendrogram), annotated by CN cluster + INTEGRATED TCGA class + MMR +
  # TP53 + the external aneuploidy SCORE. The integrated-class and MMR/TP53 bars show the
  # molecular class distributing ACROSS the copy-number clusters — these are the *copy-number*
  # clusters (Fig. 1a), NOT the integrated four groups (the cascade sits on top). The
  # aneuploidy bar (aneu_id, built above from ASCETS or the master) is a CONTINUOUS annotation
  # only — never a clustering feature.
  ids <- rownames(cl$matrix)
  ann <- data.frame(CN_cluster = factor(cl$cluster[ids]), row.names = ids)
  if (have_master && have_cw) {
    pid_v <- cw$pid[match(ids, cw$barcode)]
    grab  <- function(col) if (col %in% names(master)) master[[col]][match(pid_v, master$pid)] else rep(NA, length(ids))

    # Integrated TCGA class per column (barcode), from the SAME cascade as the "Integrated
    # classification" section below (add_tcga_class): POLE omitted (no cases) > MMRd (MSI-high
    # NGS OR MMR-loss IHC) > CN-high (serous-like) > CN-low. Derived from each patient's
    # dominant CN cluster, so the Fig-1a heatmap shows where the integrated call lands on the
    # copy-number clustering. POLE is kept as an (empty) legend level to document "no cases".
    if (!is.na(high_cluster)) {
      cnv_pid_hm <- cl$assignment |>
        inner_join(cw, by = c("ID" = "barcode")) |>
        count(pid, cnv_cluster) |>
        group_by(pid) |> slice_max(n, n = 1, with_ties = FALSE) |> ungroup() |>
        transmute(pid, cnv_high = cnv_cluster == high_cluster)
      tcga_hm  <- master |> left_join(cnv_pid_hm, by = "pid") |> add_tcga_class()
      tcga_lab <- setNames(as.character(tcga_hm$tcga_class), tcga_hm$pid)[pid_v]
      ann$TCGA_class <- factor(tcga_lab,
        levels = c("POLE (ultramutated)", attend_tcga_levels$msi,
                   attend_tcga_levels$cn_high, attend_tcga_levels$cn_low))
    }

    mmrd  <- is_mmrd(grab("MMR_class"))   # MMRd = MMR IHC loss ONLY (shared is_mmrd; MSI not a criterion)
    ann$MMR <- factor(ifelse(is.na(mmrd), NA, ifelse(mmrd, "MMRd", "MMRp")), levels = c("MMRp", "MMRd"))
    tp <- grab("TP53_pathogenic")
    if (!all(is.na(tp))) ann$TP53 <- factor(ifelse(as.logical(tp), "mut", "wt"), levels = c("wt", "mut"))
  }
  # Aneuploidy score as a continuous annotation bar (same aneu_id used by the enrichment
  # boxplot below; keyed by barcode ID = heatmap column). Kept out of the clustering.
  if (!is.null(aneu_id))
    ann$Aneuploidy <- aneu_id$aneuploidy_value[match(ids, aneu_id$ID)]

  # Feature positions: GISTIC genes -> cytoband (from the loader attribute).
  fpos <- attr(cnv_mat, "feature_pos")
  cyto <- if (!is.null(fpos)) setNames(fpos$cytoband, fpos$feature) else NULL
  feat_pos <- feature_positions(colnames(cl$matrix), cytoband = cyto)

  # WHY-IS-THERE-NO-HEATMAP gate. fig1a_heatmap() returns invisible(NULL) on four different
  # conditions, each announced only with message() — which opts_chunk (message = FALSE)
  # swallows, so the figure just silently fails to appear. Print the preconditions instead:
  # whichever line below reads FALSE / 0 is the reason the heatmap is missing.
  cat("Fig-1a heatmap preconditions",
      "\n  ComplexHeatmap installed :", requireNamespace("ComplexHeatmap", quietly = TRUE),
      "\n  circlize installed       :", requireNamespace("circlize", quietly = TRUE),
      "\n  matrix (samples x peaks) :", nrow(cl$matrix), "x", ncol(cl$matrix),
      "\n  samples with a cluster   :", sum(rownames(cl$matrix) %in% names(cl$cluster)),
      "\n  feature positions mapped :", sum(feat_pos$feature %in% colnames(cl$matrix)),
      "\n  annotation bars          :", if (is.null(ann)) 0L else ncol(ann), "\n")

  fig1a_heatmap(cl$matrix, feat_pos, cl$cluster, ann_col = ann,
                value_type = "thresholded",
                main = paste0("GISTIC thresholded peaks — TCGA Fig. 1a (Suppl. S2); cluster ",
                              high_cluster, " = copy-number high"))
} else {
  cat("CNV clustering skipped — run GISTIC so data/gistic/ exists (see bottom of report).\n")
  high_cluster <- NA_integer_
}
```

## The k-sweep — k = 2…8 on one chromosome heatmap

This is the headline of this variant. The **Pearson/Ward dendrogram is computed once**, then
**cut at every k from 2 to 8**. Because a single hierarchical tree is cut repeatedly, the
partitions are **nested** (k = 8 is a refinement of k = 2), so they can share one figure:
the GISTIC thresholded matrix is drawn once — peak genes (rows) grouped by chromosome,
tumours (columns) ordered by the **actual Pearson/Ward dendrogram** — with a **stack of
cluster bars beneath, one per k**. Reading a column downward shows how that tumour's cluster
label subdivides as k grows; reading a bar left-to-right shows the k-cluster partition
against the same chromosomal SCNA background. This makes the choice of k a **visual
question** — at which k do the chromosome blocks stop corresponding to coherent SCNA
patterns — rather than a number fixed in advance.

> **Computational steps.** (1) Reuse the Pearson/Ward `cl$hclust` from the clustering chunk.
> (2) `fig1a_heatmap_ksweep(cl$matrix, feat_pos, cl$hclust, k_range = `r paste0(min(k_sweep), ":", max(k_sweep))`)`
> draws the matrix once (columns in dendrogram order, no per-k re-clustering) and adds one
> `cutree(hc, k)` cluster bar per k, plus the same top annotation bars (integrated class,
> MMR, TP53, aneuploidy score) as the single-k Fig-1a heatmap above. (3) The average
> silhouette across k (next chunk) is the quantitative companion to this visual sweep.
>
> **vs TCGA.** TCGA fixed a single k (4). This panel is an ATTEND diagnostic: it exposes the
> whole cut spectrum on the Fig-1a canvas so the k = 4 parity cut used by the cascade can be
> judged against its neighbours, under the Pearson/Ward metric.

```{r cnv-ksweep, fig.height=9}
if (!is.null(cl) && exists("feat_pos") && exists("ann")) {
  fig1a_heatmap_ksweep(cl$matrix, feat_pos, cl$hclust, k_range = k_sweep, ann_col = ann,
                       value_type = "thresholded",
                       main = paste0("GISTIC thresholded peaks (Pearson/Ward) — k = ",
                                     min(k_sweep), "..", max(k_sweep), " cluster bars on one Fig-1a canvas"))
} else cat("k-sweep heatmap skipped — need the Pearson/Ward clustering (run GISTIC first).\n")
```

```{r cnv-ksweep-silhouette, fig.height=5}
# Average silhouette width vs k over the SAME Pearson/Ward dendrogram — the quantitative
# read on "how many clusters". Peaks flag the most internally-coherent cut(s); the cascade's
# k = `r k_cnv` is marked for reference. Optional dependency `cluster`.
if (!is.null(cl) && requireNamespace("cluster", quietly = TRUE)) {
  D  <- if (identical(cnv_dist, "correlation"))
          stats::as.dist(1 - stats::cor(t(cl$matrix), use = "pairwise.complete.obs"))
        else stats::dist(cl$matrix, method = cnv_dist)
  ks <- k_sweep[k_sweep >= 2 & k_sweep <= (nrow(cl$matrix) - 1)]
  avg_sil <- vapply(ks, function(k) {
    s <- cluster::silhouette(stats::cutree(cl$hclust, k = k), D)
    if (is.matrix(s)) mean(s[, "sil_width"]) else NA_real_
  }, numeric(1))
  sil_df <- data.frame(k = ks, avg_sil = avg_sil)
  ggplot(sil_df, aes(k, avg_sil)) +
    geom_line(colour = "grey50") + geom_point(size = 2) +
    geom_vline(xintercept = k_cnv, linetype = 2, colour = "#D62728") +
    scale_x_continuous(breaks = ks) +
    labs(title = "Average silhouette width vs k (Pearson/Ward)",
         subtitle = paste0("dashed = cascade cut k = ", k_cnv, "; higher = more coherent partition"),
         x = "number of clusters (k)", y = "mean silhouette width") +
    theme_minimal(base_size = 12)
} else message("Silhouette-vs-k skipped — install the optional `cluster` package.")
```

> **TCGA Fig. 1b — progression-free survival by copy-number cluster.** KM of clinical
> PFS (`gianlu__PFS_MONTHS`/`_EVENT` via `recode_event()`) for the raw copy-number clusters,
> barcodes mapped to `pid` through the report-01 crosswalk. Log-rank P as in the paper.

```{r cnv-km-pfs, fig.height=5}
if (!is.null(cl) && have_master && have_cw) {
  km_by_cluster(cl$assignment, cw, master)
} else message("KM skipped — need CNV clusters + master + crosswalk.")
```

```{r cnv-silhouette, fig.height=5}
# Per-sample silhouette at the cascade cut k = k_cnv, using the SAME 1 - Pearson distance the
# clustering used (not Euclidean) so the widths are consistent with the Pearson/Ward tree.
if (!is.null(cl) && requireNamespace("cluster", quietly = TRUE)) {
  D_sil <- if (identical(cnv_dist, "correlation"))
             stats::as.dist(1 - stats::cor(t(cl$matrix), use = "pairwise.complete.obs"))
           else stats::dist(cl$matrix, method = cnv_dist)
  sil <- cluster::silhouette(cl$cluster[rownames(cl$matrix)], D_sil)
  plot(sil, main = paste0("Silhouette (Pearson/Ward), k = ", k_cnv), col = "grey40")
}
```

> **These are the *copy-number* clusters, not the four integrated molecular groups.** The
> heatmap above is Fig. 1a — unsupervised clustering of *copy number alone*, cluster 4 =
> serous-like. The four *integrated groups* (Fig. 2b: POLE → MMRd → CN-high → CN-low)
> are a **separate object** — a cascade layered on top, where MMRd is set by MMR IHC (not
> copy number) and the SCNA clusters collapse to high (cluster 4) vs low (1–3). The
> integrated-class, MMR and TP53 annotation bars show exactly this: the molecular class
> distributes *across* the CN clusters rather than defining them — the integrated-class bar is
> the cascade's per-patient call painted onto the copy-number clustering, not a re-clustering.

## Aneuploidy enrichment across clusters

The copy-number-high cluster was labelled **above** from the arm matrix's own **SCNA
burden** (mean |arm copy number|) — a signal built only from the clustering features.
This section is therefore an **independent check**: does that same cluster also carry
the highest **external aneuploidy score**? Because the label and the plotted score come
from *different* signals, a positive result here is genuine enrichment, not the
tautology you'd get by labelling the cluster with the very score you then plot. The
burden table shows why the cluster was chosen; the boxplot (Kruskal–Wallis,
red/turquoise palette) is the cross-check.

> **Computational steps.** (1) `cluster_scna_burden()` — mean |arm log2| per sample,
> averaged per cluster (the table). (2) Join the **external** per-sample aneuploidy score
> (ASCETS if present, else the master's `aneu__aneuploidy_score` via the crosswalk) to the
> cluster assignment. (3) Boxplot aneuploidy by cluster; **Kruskal–Wallis** across clusters.
>
> **vs TCGA.** This is an ATTEND **QC**, not a TCGA figure: TCGA *characterised* cluster 4
> as the highest-SCNA/aneuploidy group but did not run this label-vs-independent-score check.
> It exists to prove our burden-based CN-high label isn't circular with the aneuploidy score.

```{r cnv-burden-table}
if (!is.null(cl) && !is.null(burden_tbl)) {
  knitr::kable(burden_tbl, digits = 3,
    caption = paste0("Intrinsic SCNA burden per cluster (mean |arm copy number|) — ",
                     "cluster ", high_cluster, " is copy-number-high. This is the label ",
                     "signal; the aneuploidy boxplot below is the independent check."))
}
```

```{r cnv-aneu-enrich}
if (!is.null(cl) && exists("aneu_id") && !is.null(aneu_id)) {
  d <- cl$assignment |>
    inner_join(aneu_id, by = "ID") |>
    filter(!is.na(aneuploidy_value)) |>
    mutate(cluster = factor(cnv_cluster),
           is_high = cnv_cluster == high_cluster)
  if (exists("cw") && !is.null(cw)) d <- dplyr::left_join(d, cw, by = c("ID" = "barcode"))

  if (nrow(d) > 0) {
    ggplot(d, aes(cluster, aneuploidy_value)) +
      geom_boxplot(outlier.shape = NA, aes(fill = is_high), width = 0.6, alpha = 0.5) +
      aneu_point_layers(d, value_col = "aneuploidy_value") +
      stat_compare_means(label = "p.format", size = 3) +
      scale_fill_manual(values = c(`TRUE` = "red", `FALSE` = "turquoise3"),
                        name = "copy-number high", labels = c("other", "serous-like")) +
      labs(title = "Aneuploidy score by CNV cluster (independent check)",
           subtitle = paste0("source: ", cnv_source, "; cluster ", high_cluster,
                             " = copy-number high (labelled by SCNA burden, not this score)"),
           x = "CNV cluster", y = "aneuploidy score") +
      theme_minimal(base_size = 12)
  }
}
```

## Integrated classification cascade

> **Computational steps.** (1) Per patient, collapse barcode → `pid` and take the
> **dominant** CNV cluster (multi-sample patients), then flag `cnv_high` = membership in
> the CN-high cluster. (2) Join to the master and apply `add_tcga_class()`, a **priority
> cascade**: **MMRd** (MSI-high by NGS **OR** MMR-protein loss by IHC) > **CN-high**
> (serous rule = `r attend_tcga_serous_rule`) > **CN-low**; the first positive test wins.
> (3) Tabulate patients per class.
>
> **vs TCGA.** *Follows:* the per-sample priority stratification of their Fig. 2b —
> substitution/POLE → MSI → copy-number cluster; CN-high = cluster membership. *Differs:*
> **POLE omitted** (ATTEND has no POLE-ultramutated cases), so the cascade starts at MMRd;
> and MMRd is called by **MMR IHC** alone (TCGA used MSI/MLH1).

```{r cascade}
tcga_summary <- NULL

if (have_master && have_cw) {
  # Cascade (POLE removed): MMRd (MMR IHC) > CN-high (serous-like) > CN-low. MMRd = MMR-IHC-deficient
  # (MMR IHC only; MSI-high is a correlate); the serous-like call follows attend_tcga_serous_rule (default
  # "cluster_only": high-aneuploidy CNV cluster membership = TCGA's serous-like cluster 4,
  # TP53 NOT required — validated separately below). add_tcga_class(), attend_classes.R §12.

  # CNV cluster per patient (barcode -> pid; dominant cluster if multi-sample),
  # then the copy-number-high flag.
  if (!is.null(cl) && !is.na(high_cluster)) {
    cnv_pid <- cl$assignment |>
      inner_join(cw, by = c("ID" = "barcode")) |>
      count(pid, cnv_cluster) |>
      group_by(pid) |> slice_max(n, n = 1, with_ties = FALSE) |> ungroup() |>
      transmute(pid, cnv_high = cnv_cluster == high_cluster)
  } else {
    cnv_pid <- tibble(pid = character(), cnv_high = logical())
  }

  classed <- master |>
    left_join(cnv_pid, by = "pid") |>
    add_tcga_class()   # serous = high-aneuploidy cluster ("cluster_only"); TP53 validated, not gated

  tcga_summary <- classed |>
    count(tcga_class, .drop = FALSE) |>
    mutate(pct = round(100 * n / sum(n), 1))
  knitr::kable(tcga_summary, caption = paste0("Patients per TCGA integrated class (MMRd > serous-like > endometrioid; serous rule = '", attend_tcga_serous_rule, "')"))
}
```

### TP53-enrichment QC — is the CN-high cluster really serous-like?

With the faithful `"cluster_only"` rule, the serous-like call is made **from copy number
alone**. TCGA's serous-like cluster 4 was **~90% *TP53*-mutant** — so if our CN-high
cluster genuinely corresponds to it, it should come out **independently** enriched for
*TP53* pathogenic mutations. This is the validation that replaces the TP53 *gate*: TP53
is checked as a **property**, not required as a **criterion**. A high fraction here
confirms the cluster is serous-like; a low fraction is a red flag that the copy-number
cluster does **not** map to TCGA's group (reconsider `k_cnv`, the CNV input, or purity).

> **Computational steps.** (1) Split patients by `cnv_high` (CN-high cluster vs the rest).
> (2) Per group, count `TP53_pathogenic` and compute the mutant fraction. (3) **Fisher exact
> test** on the 2×2 (cluster × TP53) for enrichment; stacked bar with a dashed **90%** line.
>
> **vs TCGA.** This *validates* the property Kandoth et al. reported (cluster 4 ≈ 90%
> *TP53*-mutant) rather than assuming it. We do **not** gate the serous call on TP53 (that
> would be the `"and"` rule); TP53 enrichment is the independent confirmation the cluster maps
> to TCGA's serous-like group.

```{r tp53-enrichment}
if (exists("classed") && all(c("cnv_high", "TP53_pathogenic") %in% names(classed))) {
  tp <- classed |>
    filter(!is.na(cnv_high)) |>
    mutate(cluster = ifelse(cnv_high, "CN-high cluster", "other clusters"),
           tp53    = suppressWarnings(as.logical(TP53_pathogenic))) |>
    filter(!is.na(tp53))

  if (nrow(tp) > 0 && dplyr::n_distinct(tp$cluster) >= 1) {
    tab <- tp |>
      group_by(cluster) |>
      summarise(n = dplyr::n(), TP53_mut = sum(tp53),
                pct_TP53 = round(100 * mean(tp53), 1), .groups = "drop")
    knitr::kable(tab, caption = "TP53-pathogenic fraction: CN-high cluster vs the rest (TCGA serous-like cluster 4 was ~90% TP53-mut)")

    ct <- table(factor(tp$cluster, levels = c("other clusters", "CN-high cluster")), tp$tp53)
    if (all(dim(ct) == c(2, 2)) && min(rowSums(ct)) > 0) {
      ft <- fisher.test(ct)
      message("TP53 enrichment in the CN-high cluster: Fisher p = ", signif(ft$p.value, 3),
              ", OR = ", signif(unname(ft$estimate), 3),
              if (is.finite(ft$p.value) && ft$p.value < 0.05) "  (enriched — cluster is serous-like)"
              else "  (NOT clearly enriched — cluster may not map to TCGA serous-like)")
    }

    ggplot(tp, aes(cluster, fill = tp53)) +
      geom_bar(position = "fill", width = 0.6) +
      scale_y_continuous(labels = scales::percent) +
      scale_fill_manual(values = c(`TRUE` = "red", `FALSE` = "grey80"),
                        name = "TP53 pathogenic", labels = c("wild-type", "mutant")) +
      geom_hline(yintercept = 0.90, linetype = 2, colour = "grey30") +
      labs(title = "TP53 mutation fraction by copy-number cluster",
           subtitle = "dashed = TCGA serous-like ~90% TP53-mut benchmark",
           x = NULL, y = "% TP53 pathogenic") +
      theme_minimal(base_size = 12)
  } else message("TP53-enrichment QC: no CN-high assignments with TP53 status yet.")
} else {
  message("TP53-enrichment QC skipped — needs cnv_high (from clustering) + TP53_pathogenic on the master.")
}
```

> **Computational steps (next two panels).** (1) Take the per-patient `tcga_class` from the
> cascade. (2) **Aneuploidy panel:** aneuploidy score by class, Kruskal–Wallis. (3) **TMB
> panel:** loop over every TMB definition on the master (`tmb_defs_present()` — baseline
> `tmb__TMB_SCORE`, then robust/matched/nassar) and box TMB by class, one facet per definition.
>
> **vs TCGA.** *Follows:* their finding that CN-high/serous-like carries the highest SCNA
> burden and lowest mutation rate, MSI the highest TMB (Fig. 1–2). *Differs:* per-definition
> TMB panels are an ATTEND addition (ancestry-aware recompute), absent from TCGA.

```{r cascade-panels, fig.height=6}
# Aneuploidy by class (paper-style panel).
if (exists("classed")) {
  pd <- classed |>
    mutate(tcga_class = forcats::fct_na_value_to_level(tcga_class, "unclassified"),
           aneuploidy_value = suppressWarnings(as.numeric(.data[[attend_cols$aneuploidy]])))

  ggplot(filter(pd, !is.na(aneuploidy_value)), aes(tcga_class, aneuploidy_value)) +
    geom_boxplot(outlier.shape = NA, fill = "#A6CEE3") +
    aneu_point_layers(filter(pd, !is.na(aneuploidy_value)), value_col = "aneuploidy_value") +
    stat_compare_means(label = "p.format", size = 3) +
    labs(title = "Aneuploidy by integrated class", x = NULL, y = "aneuploidy score") +
    theme_minimal(base_size = 11) + theme(axis.text.x = element_text(angle = 20, hjust = 1))
}
```

**TMB by integrated class, under both TMB definitions** — **TMB (normal)**, the upstream
tumour-only score, and **TMB (ancestry corrected)**, the same score after the ancestry
recalibration explained in the "Ancestry-corrected TMB" section below. One panel per
definition, so each class's TMB can be read under both. The dashed line is the
`r attend_thresholds$tmb` mut/Mb TMB-high cutoff.

```{r cascade-panels-tmb, fig.height=6}
if (exists("classed") && length(tmb_defs) > 0) {
  tl <- classed |>
    mutate(tcga_class = forcats::fct_na_value_to_level(tcga_class, "unclassified")) |>
    select(pid, tcga_class, all_of(unname(tmb_defs))) |>
    pivot_longer(all_of(unname(tmb_defs)), names_to = "tmb_col", values_to = "tmb_value") |>
    mutate(tmb_def   = factor(tmb_def_labels(tmb_defs)[match(tmb_col, tmb_defs)],
                              levels = tmb_def_labels(tmb_defs)),
           tmb_value = suppressWarnings(as.numeric(tmb_value))) |>
    filter(!is.na(tmb_value))

  ggplot(tl, aes(tcga_class, tmb_value)) +
    geom_boxplot(outlier.shape = NA, fill = "#B2DF8A") +
    geom_jitter(width = 0.15, size = 0.6, alpha = 0.4) +
    geom_hline(yintercept = attend_thresholds$tmb, linetype = 2, colour = "grey50") +
    facet_wrap(~ tmb_def, scales = "free_y", nrow = 1) +
    labs(title = "TMB by integrated class, per TMB definition",
         subtitle = paste0("dashed line = TMB-high cutoff (", attend_thresholds$tmb, " mut/Mb)"),
         x = NULL, y = "TMB (mut/Mb)") +
    theme_minimal(base_size = 11) + theme(axis.text.x = element_text(angle = 25, hjust = 1))
}
```

## Ancestry-corrected TMB — how it works

Tumour-only TMB is inflated by residual germline variants, and — because gnomAD's
reference is Euro-skewed — the inflation is **ancestry-dependent** (Nassar et al.,
*Cancer Cell* 2022). The ATTEND MAFs carry gnomAD v4.1 **per-sub-population** allele
frequencies, which is what makes a correction possible without matched normals.

**The pipeline reports two TMB definitions and no others:**

| Reported | Column | What it is |
|---|---|---|
| **TMB (normal)** | `tmb__TMB_SCORE` | the upstream tumour-only score, uncorrected — the baseline |
| **TMB (ancestry corrected)** | `tmb__TMB_nassar` | the endpoint of the correction chain below |

Getting from one to the other passes through **two intermediates**, computed in report 1
and living on the master, but **never plotted, faceted, or labelled anywhere**:

- **robust** — drop variants common in **any** population (`grpmax` ≥
  `r attend_tmb_recompute$af_threshold`). This is Nassar's germline filter, and it needs
  no ancestry label.
- **matched** — drop variants common in the **patient's own** sub-population, from the
  self-reported clinical ancestry (`attend_ancestry`).

The reported ancestry-corrected column is then the paper's actual method: an
**ancestry-specific regression recalibration** `TMB = m·TMB_robust + b` (European vs
non-European), applied on top of `robust`. Its coefficients (`attend_tmb_nassar$coef`)
**default to identity**, so until values fit on TCGA WES are supplied this column equals
`robust` — the published Oncopanel numbers are panel-specific and do not transfer to
exome data.

`tmb_defs_present()` cannot return the intermediates, so no report can reintroduce them
into a figure by accident. An unfiltered recompute is also used internally to calibrate
the exome denominator; it is not emitted, because it would simply duplicate the baseline.
The whole recompute lives in R with no upstream re-run — see
`attend_classes.R::tmb_variant_table()` and `apply_nassar_recalibration()`.

### How the Nassar et al. recalibration actually works

It helps to be precise about what Nassar et al. (*Cancer Cell* 2022) did, because our
`nassar` column reproduces it and the distinction from the other variants is subtle.

**The problem.** In tumor-only sequencing there is no matched normal, so inherited
(germline) variants are mixed in with somatic ones. TMB should count only somatic
mutations, so pipelines remove germline by filtering out variants that are **common in
the population** (looked up in gnomAD). But gnomAD is **ancestry-skewed** — its v2.1
exomes are ~57k European vs ~8k African. A variant that is common in African ancestry
but rare in the (Euro-heavy) reference is *not recognised as germline*, survives the
filter, and is **miscounted as a somatic mutation**. The net effect: tumor-only TMB is
**inflated, and the inflation is larger for non-European patients** — who were then
**over-classified as TMB-high**, a label that (in their data) did **not** predict
immunotherapy response outside European ancestry.

**Their method has two distinct steps.** Do not conflate them:

1. **Germline filter (variant level).** Exclude any variant whose **maximum population
   allele frequency across all gnomAD sub-populations** (`maxPOPAF`, i.e. `grpmax`)
   exceeds **0.1%**. This is a *fixed, ancestry-agnostic* filter — it drops a variant if
   it is common in *any* population. **This step is exactly our `robust` variant**
   (`grpmax ≥ 0.001`). It removes most, but not all, of the ancestry-driven inflation.

2. **Ancestry regression recalibration (sample level) — the headline contribution.**
   This is **not another filter**. Using cohorts that had *both* tumor-only and matched
   tumor-normal ("paired") TMB as ground truth, they fit a **linear regression of the
   true paired TMB on the tumor-only TMB, separately for European and non-European
   patients**:
   $$\text{TMB}_{\text{paired}} \;=\; m_{\text{group}} \cdot \text{TMB}_{\text{tumor-only}} \;+\; b_{\text{group}}$$
   The slope $m$ and intercept $b$ therefore encode, *per ancestry group*, how much the
   tumor-only estimate over- (or under-) shoots the truth. To recalibrate a new sample
   you apply its group's coefficients: $\text{TMB}_{\text{corrected}} = m_{\text{group}}
   \cdot \text{TMB}_{\text{tumor-only}} + b_{\text{group}}$. Because $(m, b)$ differ
   between European and non-European, the *same* tumor-only TMB is corrected by a *different*
   amount depending on ancestry — that is precisely how the method removes the residual
   ancestry bias that the fixed filter (step 1) leaves behind.

**Two supporting details.** (a) **Ancestry was inferred *genetically*** — principal-
component analysis of common germline SNPs called from the sequencing reads — not from
self-reported race. (b) The coefficients were **fit and validated against matched
tumor-normal** as the gold standard.

**How our `nassar` column maps to this — and where it differs.**

| Nassar step | Our implementation |
|---|---|
| germline filter `maxPOPAF ≤ 0.1%` | **`robust`** (`grpmax ≥ 0.001`) — identical in spirit |
| regression `TMB = m·TMB + b`, per ancestry, on top of the filtered TMB | **`apply_nassar_recalibration()`** on top of `robust` |
| ancestry via **genetic PCA** | **self-reported clinical** ancestry (`attend_ancestry`) — a coarser proxy; swap in EthSEQ/somalier genetic ancestry later |
| coefficients fit on **paired tumor-normal**, panel-specific (e.g. Oncopanel v3 European `m=1.094, b=−1.94`) | **default identity `(m=1, b=0)`** — those published numbers are **panel-specific and do NOT transfer to whole-exome**, so `nassar` == `robust` until you fit WES coefficients |
| gnomAD v2.1 + TOPMED | gnomAD **v4.1** (newer, larger, less Euro-biased) |

**To make `nassar` genuinely faithful for ATTEND (whole-exome):** fit the regression on
**TCGA WES** — compute paired-WES TMB and tumor-only-WES TMB (tumor-only simulated by
dropping the normal and applying the same `robust` filter), label European vs
non-European, then

```r
coefs <- fit_nassar_coefficients(tcga_ref)   # ref has paired_tmb, tumor_only_tmb, group
attend_tmb_nassar$coef <- coefs              # paste into code/attend_classes.R
```

Until then, `nassar` deliberately equals `robust` rather than showing a spurious
correction. (Sources: Nassar et al., *Cancer Cell* 2022,
[full text](https://pmc.ncbi.nlm.nih.gov/articles/PMC9559771/).)

```{r ancestry-tmb, fig.height=5}
tmb_recomp <- if (have_master) tmb_defs_present(master) else character(0)

if (have_master && length(tmb_recomp) >= 2) {
  # Per-patient TMB across recomputed definitions + reclassification vs the cutoff.
  long <- master |>
    select(pid, all_of(unname(tmb_recomp))) |>
    pivot_longer(all_of(unname(tmb_recomp)), names_to = "tmb_col", values_to = "tmb") |>
    mutate(def = factor(names(tmb_recomp)[match(tmb_col, tmb_recomp)], levels = names(tmb_recomp)),
           tmb = suppressWarnings(as.numeric(tmb))) |>
    filter(!is.na(tmb))

  print(
    ggplot(long, aes(def, tmb)) +
      geom_boxplot(outlier.shape = NA, fill = "#FDBF6F", width = 0.6) +
      geom_line(aes(group = pid), alpha = 0.15) +
      geom_point(size = 0.7, alpha = 0.4) +
      geom_hline(yintercept = attend_thresholds$tmb, linetype = 2, colour = "grey40") +
      labs(title = "Per-patient TMB: baseline (TMB_SCORE) vs ancestry-aware definitions",
           subtitle = "lines connect the same patient; dashed = TMB-high cutoff",
           x = "TMB definition", y = "TMB (mut/Mb)") +
      theme_minimal(base_size = 12)
  )

  # How many patients are TMB-high under each definition (the reclassification).
  reclass <- long |> group_by(def) |>
    summarise(n = sum(!is.na(tmb)),
              tmb_high = sum(tmb >= attend_thresholds$tmb, na.rm = TRUE),
              pct_high = round(100 * tmb_high / n, 1), .groups = "drop")
  knitr::kable(reclass, caption = paste0("TMB-high (>= ", attend_thresholds$tmb,
               " mut/Mb) count per definition — the ancestry-driven reclassification"))
} else {
  message("Ancestry-corrected TMB not on the master yet — rebuild report 1 with ",
          "the gnomAD-annotated MAFs synced into data/variant_annotations/.")
}
```

The diagnostic the whole correction exists for: **how much TMB the correction removes,
broken down by ancestry**. Each point is one patient's `TMB (normal) − TMB (ancestry
corrected)` in mut/Mb — the amount attributed to residual germline contamination.

This is the plot that shows whether the bias is real *in this cohort*. Nassar et al.
predict systematically **larger** removals in non-European groups, because gnomAD's
non-European allele-frequency reference is sparser and more germline variants survive an
undifferentiated filter. A flat distribution across sub-populations means either the
cohort is too homogeneous to show it, or the correction is not doing anything.

Sub-populations come from mapping the self-reported clinical ancestry column onto gnomAD
codes (`attend_ancestry`); patients whose label does not map are `NA` and dropped, so
group sizes here can be much smaller than the cohort.

```{r ancestry-tmb-by-group, fig.height=5}
anc <- if (have_master) ancestry_by_pid(master) else NULL
if (!is.null(anc) && all(c("tmb__TMB_SCORE","tmb__TMB_nassar") %in% names(master))) {
  g <- master |>
    transmute(pid,
              normal    = suppressWarnings(as.numeric(`tmb__TMB_SCORE`)),
              corrected = suppressWarnings(as.numeric(`tmb__TMB_nassar`)),
              removed   = normal - corrected) |>
    inner_join(anc, by = "pid") |>
    filter(!is.na(removed))
  if (nrow(g) > 0) {
    ggplot(g, aes(subpop, removed)) +
      geom_boxplot(outlier.shape = NA, fill = "#CAB2D6") +
      geom_jitter(width = 0.15, size = 0.8, alpha = 0.5) +
      geom_hline(yintercept = 0, linetype = 2, colour = "grey50") +
      labs(title = "TMB removed by the ancestry correction, by ancestry",
           subtitle = "TMB (normal) - TMB (ancestry corrected), mut/Mb; larger = more germline contamination removed",
           x = "gnomAD sub-population (from self-reported ancestry)",
           y = "mut/Mb removed") +
      theme_minimal(base_size = 12)
  } else message("No patients with both TMB definitions and a mapped ancestry.")
} else {
  message("Need the clinical ancestry column (attend_ancestry$clinical_col = '",
          attend_ancestry$clinical_col, "') and tmb__TMB_nassar on the master ",
          "for the by-ancestry view.")
}
```

## Nassar recalibration coefficients

`tmb__TMB_nassar = m_group · tmb__TMB_robust + b_group`, with `(m, b)` **per ancestry
group** (European vs non-European). Which coefficients are applied is set by
`attend_tmb_nassar$active_source`:

- **`"identity"` (default)** — `m=1, b=0`, so `nassar` equals `robust`.
- **`"oncopanel_v3"` / `"_v2"` / `"_v1"`** — the **published Nassar et al. 2022**
  coefficients (STAR Methods, "TMB Recalibration"). ⚠️ These were fit on **OncoPanel**
  (targeted-panel) tumor-only TMB with TCGA WES as ground truth, **not** on WES input —
  so applied to ATTEND's exome `robust` TMB they are an **off-scale approximation** (the
  slope maps OncoPanel→WES-truth, not WES→WES-truth). Available on request; use knowingly.
- **`"custom"`** — your own values in `attend_tmb_nassar$coef`.

The published values (per OncoPanel version): European `v3 m=1.094, b=−1.94`
(`v2 0.994, −2.12`; `v1 0.989, −2.18`); Non-European `v3 m=0.895, b=−1.29`
(`v2 0.840, −1.71`; `v1 0.821, −1.71`). To use them, set
`attend_tmb_nassar$active_source <- "oncopanel_v3"` and re-knit.

**To fit your own (the WES-correct path):** build a reference cohort with matched
**paired (tumor-normal) TMB**, **tumor-only TMB** (simulated with the same robust/grpmax
0.1% filter), and an ancestry `group` column, drop it in `data/nassar/` (any
`*reference*.tsv/.csv`), re-knit to fit, paste the printed coefficients into
`attend_tmb_nassar$coef`, and set `active_source = "custom"`. The panels below render the
recalibration effect whenever the active coefficients are non-identity.

```{r nassar-fit}
nc      <- nassar_coef_table()               # active coefficients + identity flag
ref_cfg <- attend_tmb_nassar$reference
ref_dir <- here("data", ref_cfg$dir)
ref_f   <- if (dir.exists(ref_dir))
  list.files(ref_dir, pattern = utils::glob2rx(ref_cfg$glob), full.names = TRUE, recursive = TRUE) else character(0)

if (length(ref_f) > 0) {
  ref <- tryCatch(
    if (grepl("\\.csv$", ref_f[1])) readr::read_csv(ref_f[1], show_col_types = FALSE)
    else                            readr::read_tsv(ref_f[1], show_col_types = FALSE),
    error = function(e) { message("nassar reference unreadable: ", conditionMessage(e)); NULL })
  if (!is.null(ref)) {
    fitted <- fit_nassar_coefficients(ref, paired = ref_cfg$paired,
                                      tumor_only = ref_cfg$tumor_only, group = ref_cfg$group)
    cat("Fitted from", basename(ref_f[1]), "— paste into attend_tmb_nassar$coef:\n")
    cat(sprintf("  european     = list(m = %.4f, b = %.4f)\n", fitted$european$m, fitted$european$b))
    cat(sprintf("  non_european = list(m = %.4f, b = %.4f)\n", fitted$non_european$m, fitted$non_european$b))
  }
} else {
  message("No reference cohort in ", ref_dir, " — reporting the active coefficients only.")
}

if (grepl("^oncopanel", nc$source)) {
  cat("WARNING: active_source = '", nc$source, "' — OncoPanel-derived published coefficients\n",
      "applied to WES robust TMB. Off-scale approximation (slope maps OncoPanel->WES-truth,\n",
      "not WES->WES-truth); prefer fitting on WES via the reference path.\n", sep = "")
}
knitr::kable(nc$table, digits = 4,
             caption = paste0("Active Nassar coefficients — source: ", nc$source,
                              " (identity = ", nc$identity, "; identity means nassar == robust)"))
```

What the correction actually does to each patient, on the two reported definitions.
**TMB (normal)** on the x-axis, **TMB (ancestry corrected)** on the y-axis, one point per
patient, coloured by ancestry group.

How to read it: the **dashed diagonal** is no change — a point on it was unaffected. Points
**below** it had TMB removed. The **dotted lines** are the `r attend_thresholds$tmb` mut/Mb
TMB-high cutoff on each axis, so the off-diagonal quadrants they carve out are the patients
whose **TMB-high/low call changed**. Those are the only patients for whom the choice of
definition changes any downstream figure, and the table counts them per ancestry group.

If the recalibration coefficients are still identity, this plot still shows the effect of
the germline filter itself — the correction chain from the delivered score to the reported
corrected one — which is the comparison every other report is drawing.

```{r nassar-effect, fig.height=5}
need <- c("tmb__TMB_SCORE", "tmb__TMB_nassar")
anc  <- if (have_master) ancestry_by_pid(master) else NULL

if (have_master && all(need %in% names(master)) && !is.null(anc)) {
  grp <- anc |>
    mutate(group = ifelse(subpop %in% attend_tmb_nassar$european_subpops,
                          "european", "non_european")) |>
    distinct(pid, group)
  eff <- master |>
    transmute(pid,
              normal    = suppressWarnings(as.numeric(`tmb__TMB_SCORE`)),
              corrected = suppressWarnings(as.numeric(`tmb__TMB_nassar`))) |>
    inner_join(grp, by = "pid") |>
    filter(!is.na(normal), !is.na(corrected))

  if (nrow(eff) > 0) {
    print(
      ggplot(eff, aes(normal, corrected, colour = group)) +
        geom_abline(slope = 1, intercept = 0, linetype = 2, colour = "grey50") +
        geom_point(alpha = 0.6) +
        geom_hline(yintercept = attend_thresholds$tmb, linetype = 3, colour = "grey40") +
        geom_vline(xintercept = attend_thresholds$tmb, linetype = 3, colour = "grey40") +
        labs(title = "Effect of the ancestry correction, by ancestry group",
             subtitle = paste0("dashed = no change; dotted = the ", attend_thresholds$tmb,
                               " mut/Mb TMB-high cutoff on each axis",
                               if (nc$identity) " | recalibration coefficients are identity, so this is the germline filter alone" else ""),
             x = "TMB (normal), mut/Mb", y = "TMB (ancestry corrected), mut/Mb",
             colour = NULL) +
        theme_minimal(base_size = 12)
    )
    reclass_n <- eff |> group_by(group) |>
      summarise(n = n(),
                high_normal    = sum(normal    >= attend_thresholds$tmb),
                high_corrected = sum(corrected >= attend_thresholds$tmb),
                changed_call   = sum((normal >= attend_thresholds$tmb) !=
                                     (corrected >= attend_thresholds$tmb)),
                .groups = "drop")
    knitr::kable(reclass_n, caption = paste0(
      "TMB-high (>= ", attend_thresholds$tmb, " mut/Mb) reclassification between the two ",
      "definitions, by ancestry group. changed_call = patients whose high/low call differs."))
  } else message("No patients with both TMB definitions and a mapped ancestry group.")
} else {
  message("Need tmb__TMB_SCORE, tmb__TMB_nassar and an ancestry column on the master ",
          "to show the correction effect.")
}
```

## Running GISTIC — the required input, locally and on the HPC cluster

The clustering above **requires GISTIC output in `data/gistic/`** (there is no fallback).
GISTIC is a compiled MATLAB binary that runs **outside R**, so you produce its output once,
then knit this report. The recipe reproduces Kandoth et al. Suppl. Methods S2 exactly:
SNP6→tangent-norm→CBS→GISTIC 2.0 there, **DRAGEN `.seg`→GISTIC 2.0** here.

**Step 1 — build the GISTIC input (R, on the cluster where the `.seg` live):**

```bash
Rscript -e 'source("code/load_wes_results.R"); write_gistic_seg()'
# -> output/gistic_input/attend_all_segments.seg  (Sample/Chr/Start/End/Num_Markers/Seg.CN)
```

**Step 2 — run GISTIC via `code/run_gistic.sh`** (runs both locally and under SLURM):

```bash
# get GISTIC (pick what your HPC has):
singularity pull gistic2.sif docker://genepattern/docker-gistic   # or: module load gistic/2.0.23
# grab the hg38 refgene .mat (stock binary ships hg19 only — using it MISPLACES every peak):
#   hg38.UCSC.add_mir.160920.refgene.mat  (GISTIC2 GitHub / bzhanglab/GISTIC2_example)

# submit to the scheduler (or run ./code/run_gistic.sh interactively):
GISTIC_SIF=gistic2.sif REFGENE=/path/hg38.UCSC.add_mir.160920.refgene.mat \
  sbatch code/run_gistic.sh
```

The script carries `#SBATCH` directives (4 h, 32 GB, 4 cores), auto-wraps a Singularity
image when `GISTIC_SIF` is set (else uses a `module`/conda `gistic2`), `cd`s to the repo
root, and checks the seg + refgene exist. It writes `data/gistic/`:
`all_thresholded.by_genes.txt` (the clustering input), `amp_genes`/`del_genes.conf_*.txt`
(the significant peaks), `all_lesions.conf_*.txt` (report-07 oncoplot overlay).

**Step 3 — knit this report.** `load_gistic_thresholded()` reads
`all_thresholded.by_genes.txt`, restricts to the significant-peak genes, and the CNV
sections above populate automatically.

> **Distance/linkage note.** S2 names no metric or linkage for the *copy-number* clustering.
> This variant uses **1 − Pearson correlation distance + Ward.D2 linkage** — the pair S2 does
> name for its mRNA/methylation clusterings — set via `cnv_method`/`cnv_dist` at the top and
> passed to `cluster_arm_matrix()`. Euclidean + complete is the obvious alternative; compare the
> two to see how sensitive the CN groups are to the (unspecified) metric choice.

## Better classification options given your tools (DRAGEN + ASCETS + variantalker)

Flagging where this could go beyond the current GISTIC reproduction, ordered by
effort-to-payoff:

1. **GISTIC thresholded-peak clustering — DONE (the only clustering).** Reproduces Kandoth
   Suppl. S2 exactly. The **ASCETS** aneuploidy score (`data/ascets/`) is still consumed as
   an **independent** per-sample aneuploidy value for the enrichment cross-check — it does
   not enter the clustering.
2. **Metric/linkage — this variant uses 1 − Pearson / Ward.D2.** S2 leaves the CN
   metric/linkage unstated; this report applies the mRNA/methylation pair (`cnv_dist =
   "correlation"`, `cnv_method = "ward.D2"`). The k = 2..8 sweep + silhouette-vs-k above make
   the cut choice inspectable; compare against base 09 (Euclidean/complete) for robustness.
3. **POLE branch removed.** ATTEND has no POLE-ultramutated cases, so the cascade now
   starts at MSI. `pole_ultramutated()` / `attend_pole` are kept (unwired) — if a POLE
   branch is ever wanted, feed a MAF with a protein-change column to `pole_ultramutated()`
   (matches P286R/V411L directly) and reinstate the branch in `add_tcga_class()`.
4. **Purity/ploidy-correct the copy number (medium effort).** If DRAGEN ran
   tumor-only, log2 ratios aren't purity-corrected, blurring low-purity samples.
   Re-call with **ASCAT/FACETS** (matched normal) for integer allele-specific CN,
   whole-genome-doubling and LOH — sharpening the CN-high boundary and enabling a
   real **HRD** cross-check against `hrd__HRD_Score`.
5. **Serous-like = SCNA cluster membership (faithful TCGA), TP53 validated not gated.**
   `add_tcga_class()` sets the CN-high call via `attend_tcga_serous_rule`, now defaulting
   to **`"cluster_only"`** — serous-like = the high-aneuploidy CNV cluster, exactly as TCGA
   defined it (Fig. 2b, cluster 4). *TP53* is checked as a **property** (the TP53-enrichment
   QC above should show ~90% mutant), not required. Alternatives remain: `"and"`
   (also require TP53, higher precision), `"or"`, `"tp53_only"`. The top of the cascade is
   **MMRd = MMR-loss by IHC (MMRd vs MMRp)**; MSI-high (NGS) is a correlate, not a criterion.
   Remaining focal-CNV upgrade is note 2 (GISTIC).
6. **Ancestry-aware TMB — robust/matched DONE; `nassar` scaffolded.** The recompute
   (`tmb_variant_table()`) produces ancestry-robust (grpmax) / ancestry-matched (on the
   upstream TMB_SCORE baseline)
   TMB from the gnomAD-annotated MAFs (matched activates once
   `attend_ancestry$clinical_col` points at the real ancestry field; swap in
   EthSEQ/somalier genetic ancestry later for accuracy). The **`nassar`** regression
   variant (`apply_nassar_recalibration()`) is wired with identity coefficients — the
   remaining step is to **fit `(m, b)` per ancestry on TCGA WES** (paired vs
   tumor-only-simulated TMB) and paste them into `attend_tmb_nassar$coef`.

## Notes

Config lives in `code/attend_classes.R`: `attend_cnv` (CNV sources + arm table),
`attend_pole` (POLE hotspots/TMB cutoff), `attend_tcga_levels` (class labels),
`attend_tmb_recompute` (germline-filter threshold, gnomAD columns, counted classes),
`attend_ancestry` (self-reported → sub-population map; **set `clinical_col`**), and
`k_cnv` at the top of this report. The arm-boundary table is
`data/arm_boundaries_hg38.tsv` (UCSC hg38 cytoBand centromeres); set
`attend_cnv$build`/`arm_table` for hg19. All sections are knit-safe — the CNV sections
render as scaffolding until `data/ascets/` or `data/cnv_annotations/` is populated, and
the ancestry-corrected TMB column appears once report 1 is rebuilt with the gnomAD-annotated
MAFs synced into `data/variant_annotations/`.

```{r session-info}
sessionInfo()
```
