---
title: "ATTEND — 8. Is the TMB distribution split? (and does TCGA agree)"
author: "bolt3x"
date: "2026-08-09"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
library(here)

# renv: tidyverse, here. Optional (requireNamespace-guarded, like the rest of the
# pipeline): diptest (Hartigan's dip test) + mclust (Gaussian mixture). If missing:
#   renv::install(c("diptest","mclust")); renv::snapshot()
# The report still knits without them — those panels degrade to a scaffolding note.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 10, fig.height = 7)

source(here("code", "attend_classes.R"))   # tmb_defs_present(), tmb_class_of(),
                                            # dip_bimodality(), is_mmrd(), attend_cols/…
source(here("code", "attend_plots.R"))     # attend_aneu_cols, attend_theme()
source(here("code", "attend_io.R"))        # read_intermediate()
```

## What this report does

**One question: does TMB split the cohort into two populations, or is it one continuous
distribution with a long tail?**

That is not a statistical curiosity. A genuine second mode at high TMB *is* the
**hypermutator population** — MMR-deficient tumours whose failed mismatch repair produces
an order-of-magnitude more mutations. (POLE-ultramutated tumours are the other
hypermutator route, but ATTEND has no POLE cases.) So "is the TMB distribution bimodal?"
is really "does TMB, by itself, cleanly identify the hypermutators?"

The report answers it in four steps:

1. **Describe** each TMB distribution.
2. **Test for bimodality** with three independent criteria, because no single test is
   trustworthy on a skewed distribution.
3. **Validate any second mode against MMR status** — if the high mode is the hypermutator
   population, it should be full of MMR-deficient tumours. A second mode that is *not*
   enriched for MMRd is a statistical artefact, not a biological population.
4. **Reproduce the whole thing in TCGA UCEC**, where the answer is externally known.

### The two TMB definitions

As everywhere in this pipeline, exactly two:

| Definition | Column | What it is |
|---|---|---|
| **TMB (normal)** | `tmb__TMB_SCORE` | the upstream tumour-only score, uncorrected |
| **TMB (ancestry corrected)** | `tmb__TMB_nassar` | after the Nassar ancestry recalibration |

The comparison between them is a second question this report can answer: tumour-only TMB
carries ancestry-dependent germline inflation, which is **noise added to the signal**.
Removing it should **deepen the valley** between the two modes. If the corrected
definition is *more* clearly bimodal than the uncorrected one, that is direct evidence
the correction is doing something real. If it is not, the correction is not helping the
split — and this report says so either way.

The intermediates (`robust`, `matched`) are not analysed here; they are computed only as
steps toward the corrected column and are never reported as definitions.

### Why TCGA is the right external check

TCGA UCEC 2013 was sequenced **tumour with matched normal**. Germline variants were
subtracted at source, so its TMB needs no ancestry correction at all — it is what an
uninflated TMB distribution looks like in the same disease. That makes it a genuine
reference rather than just another cohort:

- If TCGA's TMB is clearly bimodal and ATTEND's is not, the missing split is likely a
  measurement problem in tumour-only TMB, not a property of the ATTEND cohort.
- If **TMB (ancestry corrected)** looks more TCGA-like than **TMB (normal)**, the
  correction is moving ATTEND toward the paired-sequencing ground truth.

**POLE-ultramutated tumours are excluded from the TCGA comparison.** TCGA has them and
ATTEND does not; leaving them in would give TCGA a third, extreme mode that ATTEND cannot
have, and any comparison would be measuring that difference rather than the MMRd split.

> **Scale.** Every distribution and every bimodality test is on the **raw mut/Mb** scale,
> with no log transform. TMB is heavily right-skewed, and on the raw scale a long tail and
> a genuine second mode can look alike — which is exactly why three independent criteria
> are triangulated below rather than trusting any one of them.

## Data

```{r data}
master <- read_intermediate("attend_master_joined")

# The two TMB definitions the pipeline plots: normal first, then ancestry corrected.
# Same helper every other report uses, so this one cannot drift from them.
def_cols <- tmb_defs_present(master)

# Drop definitions that are absent or effectively empty (the ancestry-corrected column
# stays all-NA until attend_ancestry$clinical_col is set) — fewer than 10 finite values
# cannot support a distribution test.
finite_n <- vapply(def_cols,
                   function(cc) sum(is.finite(suppressWarnings(as.numeric(master[[cc]])))),
                   integer(1))
def_cols <- def_cols[finite_n >= 10]

# Reader-facing labels from the one shared map, so "nassar" never reaches the page.
def_labels <- tmb_def_labels(def_cols)

# One long row per (patient, definition), raw mut/Mb, NA dropped. The factor keeps
# correction order so every facet and axis reads normal -> ancestry corrected.
long <- master |>
  select(pid, all_of(unname(def_cols))) |>
  pivot_longer(all_of(unname(def_cols)), names_to = "col", values_to = "tmb") |>
  mutate(
    tmb        = suppressWarnings(as.numeric(tmb)),
    definition = factor(names(def_cols)[match(col, def_cols)],
                        levels = names(def_cols), labels = unname(def_labels))
  ) |>
  filter(is.finite(tmb)) |>
  select(pid, definition, tmb)

cat("TMB definitions analysed:",
    if (length(def_cols)) paste(def_labels, collapse = ", ") else "none present", "\n")
cat("finite values per definition (full master):\n")
print(finite_n)
```

```{r bimod-helpers}
# --- Bimodality machinery used only in this report --------------------------------
# (Hartigan's dip test is reused from attend_classes.R::dip_bimodality(); the two
#  measures below are not reused elsewhere yet, so they live here.)

# Bias-corrected sample skewness / excess kurtosis (SAS/`moments` "type 2" convention,
# so the bimodality coefficient matches its published 5/9 benchmark). Computed inline
# to avoid adding a `moments` dependency.
.skew_kurt <- function(x) {
  x <- x[is.finite(x)]; n <- length(x)
  if (n < 4) return(c(skew = NA_real_, kurt = NA_real_))
  s <- sd(x); if (!is.finite(s) || s == 0) return(c(skew = NA_real_, kurt = NA_real_))
  z  <- (x - mean(x)) / s
  g1 <- (n / ((n - 1) * (n - 2))) * sum(z^3)                                  # skewness
  g2 <- ((n * (n + 1)) / ((n - 1) * (n - 2) * (n - 3))) * sum(z^4) -
        (3 * (n - 1)^2) / ((n - 2) * (n - 3))                                 # excess kurtosis
  c(skew = g1, kurt = g2)
}

# Sarle's bimodality coefficient b = (g1^2 + 1) / (g2 + 3(n-1)^2/((n-2)(n-3))).
# Benchmark: a uniform distribution gives b = 5/9 ≈ 0.555; b > 0.555 leans bimodal,
# b < 0.555 leans unimodal. (A heavy right tail alone inflates skewness and can push
# b up without a true second mode — hence we never rely on b alone.)
.bimodality_coef <- function(x) {
  x <- x[is.finite(x)]; n <- length(x)
  sk <- .skew_kurt(x)
  if (any(!is.finite(sk))) return(NA_real_)
  (sk["skew"]^2 + 1) / (sk["kurt"] + (3 * (n - 1)^2) / ((n - 2) * (n - 3)))
}

# BIC-selected Gaussian mixture (1..3 components). Returns the chosen component count,
# the raw gap between the extreme component means, and that gap normalised by the
# vector's SD (the raw gap is NOT comparable across definitions, because the correction
# shifts the mut/Mb values; the SD-normalised gap is). Knit-safe without `mclust`.
.gmm_modes <- function(x) {
  x <- x[is.finite(x)]
  na <- list(G = NA_integer_, gap = NA_real_, gap_sd = NA_real_, mc = NULL)
  if (length(x) < 6 || !requireNamespace("mclust", quietly = TRUE)) return(na)
  mc <- tryCatch(mclust::Mclust(x, G = 1:3, verbose = FALSE), error = function(e) NULL)
  if (is.null(mc)) return(na)
  mns <- sort(as.numeric(mc$parameters$mean))
  gap <- if (length(mns) >= 2) mns[length(mns)] - mns[1] else 0
  s   <- sd(x)
  list(G = mc$G, gap = gap, gap_sd = if (is.finite(s) && s > 0) gap / s else NA_real_, mc = mc)
}

have_diptest <- requireNamespace("diptest", quietly = TRUE)
have_mclust  <- requireNamespace("mclust",  quietly = TRUE)
if (!have_diptest) message("`diptest` not installed — dip-test columns render as NA.")
if (!have_mclust)  message("`mclust` not installed — mixture columns render as NA.")
```

## 1. Distribution overview

Per-definition summary of the raw TMB. `%high` is the fraction at or above the
pipeline TMB-high cut (`attend_thresholds$tmb` = `r attend_thresholds$tmb` mut/Mb),
via the shared `tmb_class_of()` — the same high/low split every TMB-keyed figure uses.

```{r summary-table}
if (nrow(long) > 0) {
  summ <- long |>
    group_by(definition) |>
    summarise(
      n       = dplyr::n(),
      median  = median(tmb),
      IQR     = IQR(tmb),
      mean    = mean(tmb),
      sd      = sd(tmb),
      min     = min(tmb),
      max     = max(tmb),
      skew    = unname(.skew_kurt(tmb)["skew"]),
      exkurt  = unname(.skew_kurt(tmb)["kurt"]),
      `%high` = 100 * mean(tmb_class_of(tmb) == "TMB-high"),
      .groups = "drop"
    )
  knitr::kable(summ, digits = 2,
               caption = "Raw TMB (mut/Mb) by definition. High right-skew (large positive `skew`) is expected and is itself a hint of a heavy hypermutator tail.")
} else {
  cat("No TMB values present — nothing to summarise.\n")
}
```

```{r density, fig.height=6}
if (nrow(long) > 0) {
  ggplot(long, aes(tmb)) +
    geom_histogram(aes(y = after_stat(density)), bins = 40,
                   fill = "grey85", colour = "grey65", linewidth = 0.2) +
    geom_density(colour = "steelblue", linewidth = 0.8, adjust = 0.9, na.rm = TRUE) +
    geom_rug(alpha = 0.25) +
    geom_vline(xintercept = attend_thresholds$tmb, linetype = "dashed",
               colour = "firebrick") +
    facet_wrap(~ definition, scales = "free") +
    labs(title = "TMB distribution by definition (raw mut/Mb)",
         subtitle = paste0("dashed red = TMB-high cut (", attend_thresholds$tmb,
                           " mut/Mb). A distinct high-TMB hump = hypermutator mode."),
         x = "TMB (mut/Mb)", y = "density") +
    attend_theme()
} else {
  message("No TMB values to plot.")
}
```

## 2. Bimodality battery

Three independent criteria per definition, because on the raw scale no single test is
decisive:

- **Hartigan's dip test** (`dip_bimodality()`, reused from `attend_classes.R`) — the
  scale-invariant departure-from-unimodality statistic; small **p** ⇒ bimodal.
- **Sarle's bimodality coefficient (BC)** — from skewness/kurtosis; **b > 0.555** leans
  bimodal. (Inflated by a heavy tail, so corroborative only.)
- **Gaussian mixture (GMM)** — `mclust::Mclust(G = 1:3)`, BIC-selected component count;
  **≥ 2** components ⇒ the data prefer more than one Gaussian.

The **verdict** calls a definition *bimodal* when **≥ 2 of the 3** criteria agree —
a deliberately conservative majority vote.

```{r battery}
if (nrow(long) > 0) {
  battery <- long |>
    group_by(definition) |>
    group_modify(function(.x, ...) {
      x   <- .x$tmb
      dp  <- dip_bimodality(x)          # list(dip, p, n, bimodal)
      bc  <- as.numeric(.bimodality_coef(x))
      gmm <- .gmm_modes(x)
      tibble(
        n           = length(x),
        dip_D       = dp$dip,
        dip_p       = dp$p,
        dip_bimodal = isTRUE(dp$bimodal),
        BC          = bc,
        BC_bimodal  = is.finite(bc) && bc > 0.555,
        gmm_G       = gmm$G,
        gmm_bimodal = is.finite(gmm$G) && gmm$G >= 2,
        gmm_gap_sd  = gmm$gap_sd
      )
    }) |>
    ungroup() |>
    mutate(
      votes   = dip_bimodal + BC_bimodal + gmm_bimodal,
      verdict = if_else(votes >= 2, "bimodal", "unimodal")
    )

  knitr::kable(
    battery |>
      transmute(definition, n,
                dip_D = round(dip_D, 4), dip_p = signif(dip_p, 3), dip_bimodal,
                BC = round(BC, 3), BC_bimodal,
                gmm_G, gmm_bimodal,
                votes, verdict),
    caption = "Bimodality battery per TMB definition. verdict = bimodal when ≥ 2 of {dip, BC, GMM} agree."
  )
} else {
  battery <- tibble()
  cat("No TMB values — bimodality battery skipped.\n")
}
```

```{r battery-note, results='asis'}
if (exists("battery") && nrow(battery) > 0) {
  bm <- battery |> filter(verdict == "bimodal") |> pull(definition) |> as.character()
  if (length(bm)) {
    cat("**Bimodal by majority vote:** ", paste(bm, collapse = ", "), ".\n\n", sep = "")
  } else {
    cat("**No definition reaches the ≥2-of-3 bimodality threshold** on the raw scale. ",
        "The high-TMB signal may be a heavy tail rather than a resolved second mode — ",
        "revisit on a log scale if a hypermutator subpopulation is expected.\n\n", sep = "")
  }
}
```

## 3. Mode validation — does the high-TMB mode capture MMR-deficient tumours?

A statistical second mode is only interesting if it is the **biological** one. For every
definition the mixture split into ≥ 2 components, we assign each patient to a component by
posterior, label the **highest-mean component** as the high-TMB mode, and cross-tabulate
that membership against **MMR status** (`gianlu__MMR_STATUS`, via the shared `is_mmrd()`).
A Fisher test asks whether the high-TMB mode is enriched for MMR-deficient tumours — the
expected hypermutators.

```{r mode-validation, results='asis'}
mmr_col <- attend_cols$mmr_status   # gianlu__MMR_STATUS

validate_mode <- function(def_key, def_lab) {
  col <- def_cols[[def_key]]
  d <- master |>
    transmute(pid,
              tmb  = suppressWarnings(as.numeric(.data[[col]])),
              mmrd = is_mmrd(.data[[mmr_col]])) |>
    filter(is.finite(tmb))
  if (nrow(d) < 6) return(NULL)
  mc <- tryCatch(mclust::Mclust(d$tmb, G = 1:3, verbose = FALSE), error = function(e) NULL)
  if (is.null(mc) || mc$G < 2) return(NULL)
  hi_comp  <- which.max(as.numeric(mc$parameters$mean))         # highest-mean component
  d$hi_mode <- factor(if_else(mc$classification == hi_comp, "high-TMB mode", "bulk"),
                      levels = c("bulk", "high-TMB mode"))
  d2 <- d |> filter(!is.na(mmrd))
  if (nrow(d2) < 4 || dplyr::n_distinct(d2$hi_mode) < 2 || dplyr::n_distinct(d2$mmrd) < 2)
    return(list(lab = def_lab, tab = table(d$hi_mode), p = NA_real_))
  tab <- table(MMR = if_else(d2$mmrd, "MMRd", "MMRp"), mode = d2$hi_mode)
  p   <- tryCatch(fisher.test(tab)$p.value, error = function(e) NA_real_)
  list(lab = def_lab, tab = tab, p = p)
}

if (!have_mclust) {
  cat("_`mclust` not installed — mode validation skipped._\n\n")
} else if (!mmr_col %in% names(master)) {
  cat("_MMR status column (`", mmr_col, "`) absent from the master — mode validation skipped._\n\n", sep = "")
} else if (!exists("battery") || nrow(battery) == 0) {
  cat("_No definitions to validate._\n\n")
} else {
  qualifying <- battery |> filter(is.finite(gmm_G) & gmm_G >= 2) |> pull(definition) |> as.character()
  key_by_lab <- setNames(names(def_cols), unname(def_labels))
  if (!length(qualifying)) {
    cat("_No definition produced a ≥2-component mixture — no high-TMB mode to validate._\n\n")
  } else {
    for (lab in qualifying) {
      res <- validate_mode(key_by_lab[[lab]], lab)
      if (is.null(res)) next
      cat("### ", res$lab, "\n\n", sep = "")
      print(knitr::kable(as.data.frame.matrix(res$tab),
                         caption = paste0("High-TMB mode membership vs MMR status — ", res$lab)))
      cat("\n\nFisher exact p = ",
          if (is.finite(res$p)) signif(res$p, 3) else "NA (a margin was degenerate)",
          if (is.finite(res$p) && res$p < 0.05) " — the high-TMB mode is significantly enriched for MMRd." else "",
          "\n\n", sep = "")
    }
  }
}
```

## 4. Cross-definition synthesis — does correction sharpen the split?

Removing ancestry-dependent germline inflation should make the hypermutator mode *stand
out more*: as
ancestry-inflated germline counts are stripped, the bulk tightens and the valley
deepens. Two comparators, read left-to-right along the chain:

- **Dip statistic D** (scale-invariant) — higher = further from unimodal.
- **SD-normalised mixture gap** — the distance between the extreme component means in
  units of the definition's own SD (comparable across definitions; the raw mut/Mb gap
  is not, because the definitions live on shifted scales).

```{r synthesis, fig.height=4.5}
have_syn <- exists("battery") && nrow(battery) > 0 &&
  (any(is.finite(battery$dip_D)) || any(is.finite(battery$gmm_gap_sd)))
if (have_syn) {
  syn <- battery |>
    select(definition, `dip statistic D` = dip_D, `mixture gap (÷SD)` = gmm_gap_sd) |>
    pivot_longer(-definition, names_to = "metric", values_to = "value") |>
    filter(is.finite(value))
  if (nrow(syn) > 0) {
    ggplot(syn, aes(definition, value, group = 1)) +
      geom_line(colour = "grey60") +
      geom_point(size = 3, colour = "steelblue") +
      facet_wrap(~ metric, scales = "free_y") +
      labs(title = "Separation strength along the germline-correction chain",
           subtitle = "left to right = normal -> ancestry corrected; up = more clearly bimodal",
           x = NULL, y = NULL) +
      attend_theme() +
      theme(axis.text.x = element_text(angle = 20, hjust = 1))
  } else {
    message("No finite separation metrics to plot.")
  }
} else {
  message("Battery unavailable — synthesis skipped.")
}
```

```{r synthesis-note, results='asis'}
if (exists("battery") && nrow(battery) > 0 && sum(is.finite(battery$dip_D)) >= 2) {
  b <- battery |> filter(is.finite(dip_D))
  first_lab <- as.character(b$definition[1]); last_lab <- as.character(b$definition[nrow(b)])
  d_first <- b$dip_D[1]; d_last <- b$dip_D[nrow(b)]
  cat("Dip D moves from ", signif(d_first, 3), " (", first_lab, ") to ",
      signif(d_last, 3), " (", last_lab, "). ",
      if (d_last > d_first)
        "Separation **increases** after correction — consistent with the correction sharpening the hypermutator mode."
      else if (d_last < d_first)
        "Separation **decreases** after correction — it does not visibly sharpen the split on the raw scale (or the corrected column is too sparse to resolve it)."
      else "Separation is unchanged after correction.",
      "\n\n", sep = "")
}
```

## 5. Does the split reproduce in TCGA UCEC?

The external check. TCGA UCEC 2013 (Kandoth et al., *Nature* 2013) is the same disease,
sequenced **tumour with matched normal** — so germline variants were removed at source and
its TMB carries none of the tumour-only inflation the correction above is trying to undo.
It is the closest thing available to a ground-truth TMB distribution for endometrial
carcinoma.

**What is compared.** TCGA's `TMB_NONSYNONYMOUS` (non-synonymous mutations per Mb, from
`data/tcga_2013/data_clinical_sample.txt`, available on the 248 of 373 tumours that were
exome-sequenced) against each ATTEND definition, on the same raw mut/Mb scale and through
the identical bimodality battery used in section 2.

**POLE-ultramutated tumours are excluded.** TCGA has 17 of them at ~232 mut/Mb — an order
of magnitude above even the MSI hypermutators — and ATTEND has none. Leaving them in would
hand TCGA a third extreme mode that ATTEND structurally cannot have, so every comparison
would be detecting that, not the MMRd split. The exclusion count is printed below.

**TCGA's MMRd group** is its own MSI (Hyper-mutated) integrative cluster
(`mmrd_rule = "tcga_exact"`), which is the paper's classification rather than a marker
call — so "is the high mode enriched for MMRd?" is asked against TCGA's published labels.

```{r tcga-load}
source(here("code", "load_wes_results.R"))   # load_tcga_ucec_2013()

tcga_raw <- tryCatch(load_tcga_ucec_2013(), error = function(e) tibble())
have_tcga <- nrow(tcga_raw) > 0 && "tmb" %in% names(tcga_raw) &&
             sum(is.finite(tcga_raw$tmb)) >= 10

if (!have_tcga) {
  cat("**TCGA UCEC 2013 not available** (`data/tcga_2013/`) — run",
      "`Rscript code/fetch_tcga_ucec_2013.R`. This section is skipped, not silently empty.\n")
} else {
  n_all  <- sum(is.finite(tcga_raw$tmb))
  tcga <- tcga_raw |> filter(is.finite(tmb), !pole)      # <- POLE excluded
  cat("TCGA tumours with a TMB value :", n_all, "\n")
  cat("POLE-ultramutated, excluded   :", n_all - nrow(tcga), "\n")
  cat("TCGA cohort used here         :", nrow(tcga), "\n")
  cat("  of which MMRd (MSI cluster) :", sum(tcga$mmrd, na.rm = TRUE),
      " | MMRp:", sum(!tcga$mmrd, na.rm = TRUE),
      " | unclassified:", sum(is.na(tcga$mmrd)), "\n")
}
```

Densities of the ATTEND definitions and the TCGA reference on one axis. The ATTEND
distributions are tumour-only, TCGA's is paired — so a systematic rightward shift of the
ATTEND curves relative to TCGA is the germline inflation, visible directly. The dashed
line is the pipeline's `r attend_thresholds$tmb` mut/Mb TMB-high cutoff.

The x-axis is trimmed at the 99th percentile so the shape near the modes stays readable;
the extreme tail is still in every test below, just not in the picture.

```{r tcga-density, fig.height=5}
if (have_tcga && nrow(long) > 0) {
  cmp <- bind_rows(
    long |> transmute(cohort = paste0("ATTEND — ", as.character(definition)), tmb),
    tibble(cohort = "TCGA UCEC (paired, POLE excluded)", tmb = tcga$tmb)
  ) |> filter(is.finite(tmb))

  xmax <- quantile(cmp$tmb, 0.99, na.rm = TRUE)

  ggplot(cmp, aes(tmb, colour = cohort)) +
    geom_density(linewidth = 0.9, na.rm = TRUE) +
    geom_vline(xintercept = attend_thresholds$tmb, linetype = 2, colour = "grey50") +
    coord_cartesian(xlim = c(0, xmax)) +
    labs(title = "TMB distribution: ATTEND (tumour-only) vs TCGA UCEC (tumour-normal paired)",
         subtitle = paste0("raw mut/Mb, x-axis trimmed at the 99th percentile; dashed line = ",
                           attend_thresholds$tmb, " mut/Mb TMB-high cutoff"),
         x = "TMB (mut/Mb)", y = "density", colour = NULL) +
    attend_theme() +
    theme(legend.position = "top") +
    guides(colour = guide_legend(ncol = 1))
}
```

The same three-criterion battery from section 2, now with TCGA as an extra row. Read the
TCGA row first: it is the benchmark. If TCGA is bimodal and an ATTEND definition is not,
the split exists in this disease and that definition is failing to resolve it.

```{r tcga-battery}
if (have_tcga) {
  tcga_row <- local({
    x   <- tcga$tmb
    dp  <- dip_bimodality(x)
    bc  <- as.numeric(.bimodality_coef(x))
    gmm <- .gmm_modes(x)
    tibble(definition  = "TCGA UCEC (paired, POLE excl.)",
           n           = length(x),
           dip_D       = dp$dip,
           dip_p       = dp$p,
           dip_bimodal = isTRUE(dp$bimodal),
           BC          = bc,
           BC_bimodal  = is.finite(bc) && bc > 0.555,
           gmm_G       = gmm$G,
           gmm_bimodal = is.finite(gmm$G) && gmm$G >= 2,
           gmm_gap_sd  = gmm$gap_sd) |>
      mutate(votes   = dip_bimodal + BC_bimodal + gmm_bimodal,
             verdict = if_else(votes >= 2, "bimodal", "unimodal"))
  })

  combined <- if (exists("battery") && nrow(battery) > 0)
    bind_rows(battery |> mutate(definition = as.character(definition)), tcga_row) else tcga_row

  knitr::kable(
    combined |>
      transmute(definition, n,
                dip_D = round(dip_D, 4), dip_p = signif(dip_p, 3), dip_bimodal,
                BC = round(BC, 3), BC_bimodal,
                gmm_G, gmm_bimodal,
                gmm_gap_sd = round(gmm_gap_sd, 2),
                votes, verdict),
    caption = "Bimodality battery — ATTEND definitions plus the TCGA paired-sequencing reference. verdict = bimodal when >= 2 of {dip, BC, GMM} agree."
  )
}
```

Does TCGA's high-TMB mode contain TCGA's MMR-deficient tumours? This is the same mode
validation section 3 runs on ATTEND, applied to the reference cohort. A strong enrichment
here establishes that the high mode really is the hypermutator population *in this
disease* — which is what licenses reading ATTEND's high mode the same way.

```{r tcga-mode-validation, results='asis'}
if (have_tcga && have_mclust) {
  mc <- tryCatch(mclust::Mclust(tcga$tmb, G = 1:3, verbose = FALSE), error = function(e) NULL)
  if (is.null(mc) || mc$G < 2) {
    cat("_TCGA TMB resolves to a single mixture component — no high mode to validate._\n\n")
  } else {
    hi_comp <- which.max(as.numeric(mc$parameters$mean))
    d <- tcga |>
      mutate(hi_mode = factor(if_else(mc$classification == hi_comp, "high-TMB mode", "bulk"),
                              levels = c("bulk", "high-TMB mode"))) |>
      filter(!is.na(mmrd))
    if (nrow(d) >= 4 && dplyr::n_distinct(d$hi_mode) == 2 && dplyr::n_distinct(d$mmrd) == 2) {
      tab <- table(MMR = if_else(d$mmrd, "MMRd", "MMRp"), mode = d$hi_mode)
      pv  <- tryCatch(fisher.test(tab)$p.value, error = function(e) NA_real_)
      cat("**TCGA — high-TMB mode vs MMR status** (", mc$G, " mixture components):\n\n", sep = "")
      print(knitr::kable(tab))
      cat("\nFisher p = ", signif(pv, 3), "\n\n", sep = "")
    } else {
      cat("_Not enough classified TCGA tumours across both mode and MMR groups._\n\n")
    }
  }
} else if (have_tcga) {
  cat("_`mclust` not installed — TCGA mode validation skipped._\n\n")
}
```

How each cohort's TMB separates MMR-deficient from MMR-proficient tumours, side by side.
This is the comparison the whole section is for: the **effect size** of MMR status on TMB
in a paired cohort versus in ATTEND under each definition. If the ancestry correction is
working, ATTEND's corrected separation should look more like TCGA's than the uncorrected
one does.

Wilcoxon rather than t-test, because TMB is heavily skewed on the raw scale.

```{r tcga-mmr-contrast, fig.height=5}
if (have_tcga) {
  mmr_col_l <- attend_cols$mmr_status
  att <- if (mmr_col_l %in% names(master)) {
    master |>
      select(pid, all_of(unname(def_cols)), all_of(mmr_col_l)) |>
      pivot_longer(all_of(unname(def_cols)), names_to = "col", values_to = "tmb") |>
      mutate(cohort = paste0("ATTEND — ", def_labels[match(col, def_cols)]),
             tmb    = suppressWarnings(as.numeric(tmb)),
             mmrd   = is_mmrd(.data[[mmr_col_l]])) |>
      filter(is.finite(tmb), !is.na(mmrd)) |>
      select(cohort, tmb, mmrd)
  } else tibble()

  both <- bind_rows(att,
    tcga |> filter(!is.na(mmrd)) |>
      transmute(cohort = "TCGA UCEC (paired)", tmb, mmrd))

  if (nrow(both) > 0) {
    both <- both |> mutate(MMR = if_else(mmrd, "MMRd", "MMRp"))

    print(knitr::kable(
      both |> group_by(cohort, MMR) |>
        summarise(n = n(), median = round(median(tmb), 2),
                  IQR = paste0(round(quantile(tmb, .25), 2), "-", round(quantile(tmb, .75), 2)),
                  .groups = "drop"),
      caption = "TMB by MMR status in each cohort / definition"))

    pv <- both |> group_by(cohort) |>
      summarise(p = tryCatch(signif(wilcox.test(tmb ~ MMR)$p.value, 3),
                             error = function(e) NA_real_), .groups = "drop")
    print(knitr::kable(pv, caption = "MMRd vs MMRp Wilcoxon p, per cohort / definition"))

    ggplot(both, aes(MMR, tmb, fill = MMR)) +
      geom_boxplot(outlier.shape = NA, alpha = 0.6) +
      geom_jitter(width = 0.15, size = 0.5, alpha = 0.35) +
      geom_hline(yintercept = attend_thresholds$tmb, linetype = 2, colour = "grey50") +
      scale_y_log10() +
      facet_wrap(~ cohort, nrow = 1) +
      scale_fill_manual(values = attend_mmr_cols) +
      labs(title = "TMB by MMR status — ATTEND definitions vs the TCGA paired reference",
           subtitle = "log10 y-axis for display only; all tests above are on the raw scale",
           x = NULL, y = "TMB (mut/Mb, log10)") +
      attend_theme() + theme(legend.position = "none")
  }
}
```

> **What this section cannot settle.** TCGA UCEC is **primary** disease; ATTEND is
> **metastatic**. A difference between the cohorts is therefore confounded with disease
> setting and cannot be attributed to sequencing strategy alone. TCGA is a useful reference
> for *what a paired-sequencing TMB distribution looks like*, not a matched control.

## 6. TMB by aneuploidy status, within MMR strata

§3 asked whether the *high-TMB* mode is the MMR-deficient population. This section asks
the complementary question the other way round: **within each MMR stratum, does TMB
differ between aneuploidy-high and aneuploidy-low tumours?** Splitting by MMR *first*
is the point — the two routes to a high number are biologically distinct and must not
be pooled: **MMR-deficient** tumours are hypermutators (high TMB) with often-modest
aneuploidy, whereas the **CN-high / serous-like** route drives high aneuploidy with
typically *low* TMB. Pooling MMRd and MMRp would let one route's signal mask the other
(a Simpson's-paradox trap), so we draw the aneuploidy-high-vs-low TMB contrast **once
per MMR group**: one boxplot for MMR-proficient (MMRp) patients, one for MMR-deficient
(MMRd), repeated across every TMB definition present.

Each point is one patient. **TMB (mut/Mb, raw)** is the `tmb__*` master columns already
assembled in §Data (`long`). **Aneuploidy status** is the per-patient
`aneu__aneuploidy_score` (ASCETS-derived arm-level SCNA burden), split at the pipeline
cutoff `attend_thresholds$aneuploidy` = `r attend_thresholds$aneuploidy`
(≥ cutoff → aneuploidy-high). **MMR group** is `gianlu__MMR_STATUS` collapsed to
MMRd/MMRp by the shared `is_mmrd()`, so the split matches every other report exactly.
Patients missing an aneuploidy score or an MMR call are dropped.

```{r aneu-tmb-build}
aneu_col <- attend_cols$aneuploidy    # aneu__aneuploidy_score
mmr_col  <- attend_cols$mmr_status    # gianlu__MMR_STATUS
aneu_cut <- attend_thresholds$aneuploidy

have_aneu_tmb <- nrow(long) > 0 &&
  aneu_col %in% names(master) && mmr_col %in% names(master)

if (have_aneu_tmb) {
  # Per-patient aneuploidy class (fixed 0.1 cut) + MMR group; joined onto the
  # per-(patient, definition) TMB rows already built in §Data.
  meta <- master |>
    transmute(
      pid,
      aneu_class = factor(
        if_else(suppressWarnings(as.numeric(.data[[aneu_col]])) >= aneu_cut,
                "aneuploidy-high", "aneuploidy-low"),
        levels = c("aneuploidy-low", "aneuploidy-high")),
      mmr_group  = dplyr::case_when(
        is_mmrd(.data[[mmr_col]])   ~ "MMRd",
        !is_mmrd(.data[[mmr_col]])  ~ "MMRp",
        TRUE                        ~ NA_character_) |>
        factor(levels = c("MMRp", "MMRd"))
    )

  ta <- long |>
    inner_join(meta, by = "pid") |>
    filter(!is.na(aneu_class), !is.na(mmr_group))

  # Facet rows = TMB definitions actually present; drives the plot's fig.height below
  # so each row keeps a legible height whether 1 or 4 definitions render.
  n_aneu_defs <- max(1L, nlevels(droplevels(ta$definition)))

  cat("patients per MMR group × aneuploidy status (baseline definition):\n")
  print(ta |>
          filter(definition == levels(definition)[1]) |>
          count(mmr_group, aneu_class) |>
          tidyr::pivot_wider(names_from = aneu_class, values_from = n, values_fill = 0))
} else {
  ta <- tibble()
  n_aneu_defs <- 1L
  cat("Aneuploidy score (`", aneu_col, "`) or MMR status (`", mmr_col,
      "`) absent from the master, or no TMB values — section skipped.\n", sep = "")
}
```

The two boxplots the analysis calls for: TMB (raw mut/Mb) in aneuploidy-high vs
aneuploidy-low tumours, faceted **by MMR group** (columns: MMRp, MMRd) and **by TMB
definition** (rows). Boxes are filled by aneuploidy status (light purple = low, deep
purple = high — the shared `attend_aneu_cols` palette the other reports use); the
dashed red line is the TMB-high cut
(`attend_thresholds$tmb` = `r attend_thresholds$tmb` mut/Mb). Points are individual
patients (jittered), not group means.

```{r aneu-tmb-box, fig.height=2.5 * n_aneu_defs + 1.5}
if (nrow(ta) > 0) {
  aneu_pal <- attend_aneu_cols
  ggplot(ta, aes(aneu_class, tmb)) +
    geom_boxplot(aes(fill = aneu_class), outlier.shape = NA, width = 0.6,
                 alpha = 0.5) +
    geom_jitter(width = 0.15, height = 0, alpha = 0.4, size = 1) +
    geom_hline(yintercept = attend_thresholds$tmb, linetype = "dashed",
               colour = "firebrick") +
    scale_fill_manual(values = aneu_pal, name = "aneuploidy status") +
    # Short horizontal tick labels — the fill legend already spells out the full
    # aneuploidy status, so the x-axis needn't repeat it (frees per-panel width).
    scale_x_discrete(labels = c("aneuploidy-low" = "low",
                                "aneuploidy-high" = "high")) +
    facet_grid(definition ~ mmr_group, scales = "free_y") +
    labs(title = "TMB by aneuploidy status, split by MMR group",
         subtitle = paste0("dashed red = TMB-high cut (", attend_thresholds$tmb,
                           " mut/Mb); aneuploidy-high = score ≥ ", aneu_cut),
         x = NULL, y = "TMB (mut/Mb)") +
    attend_theme() +
    theme(axis.text.x = element_text(angle = 0),
          legend.position = "top")
} else {
  message("No aneuploidy/TMB data to plot.")
}
```

Formalising each panel at the **10 mut/Mb** cut rather than by median: within every
(MMR group × TMB definition) cell, patients are binarised **TMB-high vs TMB-low** at
`attend_thresholds$tmb` = `r attend_thresholds$tmb` mut/Mb via the shared
`tmb_class_of()` (the same cut every TMB-keyed figure uses), and the table reports the
**count and fraction of TMB-high tumours in the aneuploidy-low vs -high groups**. A
**Fisher exact test** on the 2×2 (aneuploidy status × TMB-high/low) asks whether the
TMB-high fraction differs by aneuploidy status. Cells with an aneuploidy group of
fewer than 3 patients report `NA`.

```{r aneu-tmb-test}
if (nrow(ta) > 0) {
  aneu_tmb_tab <- ta |>
    group_by(definition, mmr_group) |>
    group_modify(function(.x, ...) {
      # Binarise TMB at the pipeline 10 mut/Mb cut (shared tmb_class_of()), then
      # count TMB-high tumours within each aneuploidy group.
      tmb_hi <- tmb_class_of(.x$tmb) == "TMB-high"
      is_lo  <- .x$aneu_class == "aneuploidy-low"
      is_hi  <- .x$aneu_class == "aneuploidy-high"
      n_lo <- sum(is_lo); n_hi <- sum(is_hi)
      hi_in_lo <- sum(is_lo & tmb_hi); hi_in_hi <- sum(is_hi & tmb_hi)
      p <- if (n_lo >= 3 && n_hi >= 3) {
        tab <- table(aneu = factor(.x$aneu_class,
                                   levels = c("aneuploidy-low", "aneuploidy-high")),
                     tmb  = factor(tmb_hi, levels = c(FALSE, TRUE)))
        tryCatch(fisher.test(tab)$p.value, error = function(e) NA_real_)
      } else NA_real_
      tibble(
        n_aneu_low    = n_lo,
        n_aneu_high   = n_hi,
        tmbhigh_low   = hi_in_lo,
        pct_low       = if (n_lo > 0) 100 * hi_in_lo / n_lo else NA_real_,
        tmbhigh_high  = hi_in_hi,
        pct_high      = if (n_hi > 0) 100 * hi_in_hi / n_hi else NA_real_,
        fisher_p      = p
      )
    }) |>
    ungroup()

  knitr::kable(
    aneu_tmb_tab |>
      transmute(definition, mmr_group,
                `n aneu-low`  = n_aneu_low,
                `n aneu-high` = n_aneu_high,
                `TMB-high | aneu-low`  = ifelse(is.finite(pct_low),
                    sprintf("%d (%.0f%%)", tmbhigh_low,  pct_low),  "—"),
                `TMB-high | aneu-high` = ifelse(is.finite(pct_high),
                    sprintf("%d (%.0f%%)", tmbhigh_high, pct_high), "—"),
                `Fisher p` = signif(fisher_p, 3)),
    caption = paste0("TMB-high tumours (≥ ", attend_thresholds$tmb,
                     " mut/Mb) as count (%) within the aneuploidy-low vs -high groups, ",
                     "and Fisher's exact p, per MMR group and TMB definition.")
  )
} else {
  cat("No aneuploidy/TMB data — test skipped.\n")
}
```

```{r aneu-tmb-note, results='asis'}
if (exists("aneu_tmb_tab") && nrow(aneu_tmb_tab) > 0) {
  # One MMRp/MMRd line per TMB definition present (normal + ancestry corrected,
  # whichever survived the finite-value filter in §Data), in correction
  # order — the same set the boxplot rows and the table above show.
  say <- function(b, grp) {
    r <- b |> filter(mmr_group == grp)
    if (nrow(r) == 0) return(paste0("**", grp, "**: no patients."))
    paste0("**", grp, "**: TMB-high fraction ",
           if (is.finite(r$pct_low))  sprintf("%.0f%%", r$pct_low)  else "NA",
           " in aneuploidy-low vs ",
           if (is.finite(r$pct_high)) sprintf("%.0f%%", r$pct_high) else "NA",
           " in aneuploidy-high (Fisher p = ",
           if (is.finite(r$fisher_p)) signif(r$fisher_p, 3) else "NA", ")")
  }
  for (lab in levels(ta$definition)) {
    b <- aneu_tmb_tab |> filter(definition == lab)
    if (nrow(b) == 0) next
    cat("- _", lab, "_ — ", say(b, "MMRp"), "; ", say(b, "MMRd"), ".\n", sep = "")
  }
  cat("\n_Read against the two-routes framing above: within MMRp the CN-high/serous-like ",
      "signature is a **lower** TMB-high fraction among aneuploidy-high tumours (high ",
      "aneuploidy, low TMB), while MMRd tumours stay TMB-high largely independent of ",
      "aneuploidy. Comparing the lines across definitions shows whether stripping ",
      "ancestry-inflated germline (normal -> ancestry corrected) changes that within-stratum ",
      "contrast._\n\n", sep = "")
}
```

## Interpretation

- The **battery table (§2)** is the headline: for each TMB definition, is the raw
  distribution one hump or two? Read `verdict` with the caveat that on the raw scale a
  heavy right tail inflates both skewness (→ BC) and, sometimes, the dip — which is why
  the majority vote across three independent criteria is the call, not any single test.
- **§3** is the check that keeps §2 honest: a second statistical mode only *matters* if
  it is the MMR-deficient hypermutator population. A non-significant Fisher test there
  means the mixture found a mode that is **not** the biological one — treat that
  definition's "bimodality" cautiously.
- **§4** answers the design question directly — whether peeling ancestry-inflated
  germline off the tumour-only count actually deepens the valley. Until
  `attend_ancestry$clinical_col` is set, the ancestry-corrected column may be sparse or
  all-NA and simply drops out of every panel above (knit-safe), leaving only TMB (normal).
- **§5** is the external check. TCGA UCEC is tumour-normal **paired**, so its TMB is what
  the distribution looks like with germline already removed. If TCGA is bimodal and an
  ATTEND definition is not, the split is real in this disease and that definition is
  failing to resolve it. If the ancestry-corrected definition tracks TCGA more closely
  than the uncorrected one, the correction is moving ATTEND toward the paired-sequencing
  reference. **POLE tumours are excluded** so the comparison is not dominated by a mode
  ATTEND cannot have — and remember TCGA is **primary** disease while ATTEND is
  **metastatic**, so it is a reference, not a matched control.
- **§6** disentangles the two hypermutation routes by asking whether TMB tracks
  aneuploidy *within* each MMR stratum. The expected signature is a **negative** or flat
  association in MMRp (the CN-high/serous-like route: high aneuploidy, low TMB) and a
  weak association in MMRd (high TMB driven by mismatch-repair loss, not SCNA burden) —
  the two boxplots make that contrast visible, and the TMB-high-fraction (≥ 10 mut/Mb)
  Fisher table quantifies it.
  This complements §3: §3 checks that the high-TMB *mode* is MMRd; §6 checks that
  aneuploidy does **not** secretly drive TMB once MMR status is fixed.
```
