---
title: "ATTEND — 40. TCGA integrated classification"
author: "sceriff0"
date: "2026-08-11"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---
```{r setup, include=FALSE}
# renv: tidyverse, here, ggpubr (+ optional bioc::ComplexHeatmap, circlize for the
# faithful Fig-1a heatmap; cluster for the silhouette; survival/ggsurvfit for the
# Fig-1b KM). If missing: renv::install(c("ggpubr","bioc::ComplexHeatmap","circlize",
# "cluster","survival","ggsurvfit")); renv::snapshot()
library(tidyverse)
library(here)
library(ggpubr)
# Headless-safe raster device. The cluster's compute nodes have no X11, so R's default
# png() fails with "unable to open connection to X11 display" and every figure dies. cairo
# needs no display; the guard keeps this working on a machine built without it.
if (capabilities("cairo")) options(bitmapType = "cairo")

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 7)

source(here("code", "attend_harmonise.R"))   # make_id_cfg(), build_barcode_pid()
source(here("code", "attend_classes.R"))     # attend_cnv/_pole, arm + cascade helpers
source(here("code", "attend_plots.R"))       # fig1a_heatmap(), feature_positions(), km_by_cluster()
source(here("code", "attend_io.R"))          # read_intermediate()
source(here("code", "build_master.R"))   # get_master() — builds on demand
source(here("code", "load_clinical.R"))      # load_gianlu_clinical_data()
source(here("code", "load_wes_results.R"))   # load_cnv_data(), load_ascets_*(), load_tmb_scores()

# cnv_method, cnv_dist, k_cnv and k_sweep are configuration and live in attend_classes.R
# with every other method parameter, so 00-methods.Rmd can name the cut it justifies.
```
> **Question.** Which copy-number cluster is enriched for aneuploidy, i.e. which one is the copy-number-high / serous-like group?
> **Cohort.** All patients with GISTIC output; the cascade itself is drawn on the full master.
> **Reads.** GISTIC thresholded peaks in data/gistic/ (produced per runbooks/gistic.md), ASCETS scores in data/ascets/, and attend_master_joined.
> **Out of scope.** Why 1 - Pearson / Ward.D2, and how the serous-like call is made directional -> [00-methods](00-methods.html). Options beyond this reproduction -> runbooks/classification_options.md.


Reproduces the TCGA integrated genomic classification of endometrial carcinoma (Kandoth et
al., *Nature* 2013, papers/nature12113.pdf) on the ATTEND cohort, and asks which
copy-number cluster is enriched for aneuploidy — the hallmark of the **copy-number-high /
serous-like** group. Turning an unlabelled cluster into a serous-like call is inferred from
the copy-number data and is not part of TCGA's methodology; how, and why it is directional,
is in [00-methods](00-methods.html).

## The cascade

A priority cascade (TCGA Fig. 2b). POLE is omitted — ATTEND has no POLE-ultramutated
cases — so it starts at MMRd:

1. **MMRd** — MMR-protein loss by **IHC** (gianlu__MMR_STATUS Deficient). This is the
   top of the cascade and overwrites the CN split. The split is **MMRd vs MMRp by IHC
   alone**; MSI-high by the NGS caller is a **correlate**, not a criterion.
2. **Copy-number high (serous-like)** — many broad SCNAs, **extensive aneuploidy**.
   How the call is made is set by attend_tcga_serous_rule, defaulting to the **faithful
   TCGA rule "cluster_only"**: serous-like = membership in the high-aneuploidy CNV
   cluster, exactly as TCGA defined it (Fig. 2b, "serous-like ... copy-number cluster 4").
   *TP53* is **not** required here — it is a **property** of the cluster (~90% mutant),
   which we **validate** below rather than gate on. Alternatives: "and" (also require
   TP53 — higher precision), "or" (union), "tp53_only" (portable).
3. **Copy-number low (endometrioid)** — MMR-proficient, few broad SCNAs, everything not
   in the high-aneuploidy cluster.

The copy-number split comes from **clustering copy-number profiles** (their Fig. 1a;
cluster 4 = serous-like), then resolving the serous-like call per attend_tcga_serous_rule.
Because the default is "cluster_only", the serous call is earned from copy number alone;
the **TP53-enrichment QC** below then confirms the cluster is genuinely ~90% *TP53*-mutant,
i.e. that it corresponds to TCGA's serous-like group — a non-circular validation.

## Data

The master for the cascade and its covariates, and the barcode crosswalk, because GISTIC
and .seg sample ids are sequencing barcodes while the master is keyed on pid. Both are
wrapped so a missing input degrades to a skipped section rather than an error.

```{r data}
master <- get_master() |> add_molecular_classes()

# barcode -> pid crosswalk (the same derivation every other report uses)
cw <- tryCatch(
  build_barcode_pid(load_gianlu_clinical_data(), make_id_cfg()),
  error = function(e) { message("crosswalk unavailable: ", conditionMessage(e)); NULL }
)

have_master <- !is.null(master)
have_cw     <- !is.null(cw)

# TMB definitions available on the master: upstream first, then the recomputed set.
tmb_defs <- c(upstream = attend_cols$tmb, attend_cols$tmb_set)
if (have_master) tmb_defs <- tmb_defs_present(master)
```

## Copy-number ingestion — two GISTIC matrices, because the paper uses two

TCGA **clusters** one matrix and **draws** another, and both come from the same GISTIC run.
Keeping them separate is what makes this report's heatmap the paper's heatmap.

**The clustering input** is exactly what TCGA clustered (Kandoth et al., Suppl. Methods S2):
*"thresholded relative copy number data in significantly reoccurring amplifications or
deletions regions identified by GISTIC 2.0."* There is **one** clustering feature and no
fallbacks — the arm-level / bin-level proxies have been removed. load_gistic_thresholded()
builds it from data/gistic/:

- input all_thresholded.by_genes.txt — genes × samples, values in **{−2,−1,0,+1,+2}**;
- restricted to the genes inside the **significant amp/del peaks** (amp_genes/del_genes.conf);
- result: a sample × peak-gene **thresholded** matrix — TCGA's exact feature space.

**The figure input** is what the Fig. 1a legend describes: *"The heat map shows SCNAs in each
tumour (horizontal axis) plotted by chromosomal location (vertical axis)"* — the **whole
genome**, on a **continuous** colour scale. That is a different matrix, and
load_gistic_continuous() reads it from all_data_by_genes.txt in the same data/gistic/:
genome-wide, continuous copy number per gene, not peak-restricted. Drawing the *clustering*
matrix instead is what previously made this heatmap something other than Fig. 1a — a few
hundred peak genes on a five-step colour scale, genome-ordered but not the genomic landscape.

So: **cluster the peaks, display the genome.** The columns of the heatmap are split and
coloured by the peak-based clusters, and the body underneath them is the continuous
genome-wide landscape those clusters were learned from a subset of.

**Both matrices need GISTIC**, on the DRAGEN .seg — the recipe is runbooks/gistic.md. Until
data/gistic/ is populated, every CNV/cascade section renders as scaffolding. If
all_data_by_genes.txt is missing but all_thresholded.by_genes.txt is present, the report says
so and falls back to drawing the clustering matrix, which is the old, non-faithful figure.

**Computational steps.** (1) load_gistic_thresholded() reads GISTIC's
all_thresholded.by_genes.txt and keeps only genes in the significant peaks — the clustering
feature. (2) load_gistic_continuous() reads all_data_by_genes.txt genome-wide with no peak
filter and no integer coercion — the display feature. (3) The per-sample aneuploidy **score**
(external, for the enrichment check only) is loaded separately (ASCETS if present, else the
master). The clustering never sees the aneuploidy score — that keeps the enrichment plot
non-circular, and the display matrix never enters the clustering either.

```{r cnv-load}
cnv_mat    <- tryCatch(load_gistic_thresholded(), error = function(e) NULL)  # ID x peak-gene {-2..+2}
cnv_source <- "GISTIC 2.0 thresholded peaks"

# The Fig-1a DISPLAY matrix: genome-wide, continuous, same GISTIC run. Never clustered.
disp_mat   <- tryCatch(load_gistic_continuous(), error = function(e) NULL)   # ID x gene, continuous

# per-sample aneuploidy SCORE — used ONLY for the enrichment cross-check, not for clustering
ascets_aneu <- tryCatch(load_ascets_aneuploidy(), error = function(e) NULL)

have_cnv <- !is.null(cnv_mat) && nrow(cnv_mat) >= k_cnv
# cat(), not message(): opts_chunk sets message = FALSE, so a message() here is invisible on
# the knitted page and a skipped CNV section then looks like nothing happened at all.
if (!have_cnv) {
  cat("No GISTIC matrix with >=", k_cnv, "samples — run GISTIC (see the bottom of this",
      "report) so data/gistic/all_thresholded.by_genes.txt exists.",
      "CNV sections are scaffolding until then.\n")
} else {
  cat("Clustering input:", cnv_source, "—", nrow(cnv_mat), "samples x",
      ncol(cnv_mat) - 1, "peak genes.\n")
}
if (!is.null(disp_mat)) {
  cat("Fig-1a display input: GISTIC all_data_by_genes (continuous, genome-wide) —",
      nrow(disp_mat), "samples x", ncol(disp_mat) - 1, "genes.\n")
} else {
  cat("Fig-1a display input: NOT AVAILABLE (no all_data_by_genes.txt under data/gistic/).",
      "The heatmaps below fall back to the peak-restricted thresholded matrix, which is",
      "genome-ORDERED but is not the paper's genome-wide continuous landscape.\n")
}
```

## The clustering feature — GISTIC thresholded peaks, and how it matches TCGA

The clustering feature is a **sample × peak-gene** matrix of **thresholded** GISTIC calls
({−2..+2}), restricted to the significantly recurrent amp/del regions — **identical in
kind to TCGA's** (Suppl. Methods S2). This variant hierarchically clusters it with
**1 − Pearson correlation distance and Ward.D2 linkage** (rather than Euclidean +
complete). S2 names no metric for the *copy-number* clustering, but its mRNA/methylation
clusterings use exactly this **1 − Pearson / Ward** pair — so this report asks what the CN
groups look like under a correlation-shaped, variance-minimising linkage rather than the
Euclidean/complete default. Instead of committing to one cut, the chromosome heatmap below
is annotated with the cluster assignment at **every k from 2 to 8** in a single figure.
What this gets right vs the old arm/bin proxies (now removed):

- **Thresholded, not continuous** — {−2..+2} discretised calls, exactly S2.
- **GISTIC significant peaks, not genome-wide** — the feature space is the recurrent
  amp/del peak genes (MYC 8q24, ERBB2 17q12, CCNE1 19q12, 1q, …), exactly S2.

Both of those are properties of the **clustering**, and neither describes the **figure**. The
heatmap body is the genome-wide continuous matrix (see the ingestion section): peak-restricted
discretised values are the right thing to cluster and the wrong thing to draw, because Fig. 1a
is a picture of the whole copy-number landscape. The clusters are still the peak-based ones —
they set the column split — so nothing about the method moves.
- **Residual differences from TCGA** — platform (DRAGEN WES .seg vs SNP 6.0), cohort
  (metastatic vs primary), N, and the unspecified CN metric/linkage (here **1 − Pearson /
  Ward.D2**, S2's mRNA/methylation choice; switch via cluster_arm_matrix(..., method=,
  dist_method=)).

## CNV clustering (GISTIC thresholded peaks)

**Computational steps.** (1) Cluster the sample × peak-gene **thresholded** GISTIC matrix
cnv_mat with cluster_arm_matrix(), taking k, method and distance from the configured
values in attend_classes.R — **1 − Pearson correlation** distance, **Ward.D2** linkage (S2's
mRNA/methylation choice). (2) Label the **copy-number-high**
cluster by intrinsic SCNA burden = mean |thresholded value| per cluster
(cnv_high_cluster_burden()), independent of the aneuploidy score — computed on the CLUSTERING
matrix, so the burden rule never sees the genome-wide display values. (3) Heatmap: the
genome-wide continuous matrix, genes (rows, ordered by chromosomal location) × samples
(columns split by the peak-based clusters and ordered within each by dendrogram), annotated by
CN cluster, the **integrated
TCGA class** (POLE [no cases] > MMRd > CN-high > CN-low, from the add_tcga_class() cascade
shown below — the same call, drawn here on the Fig-1a clustering), MMR, TP53, **and the
per-sample aneuploidy score**. That aneuploidy bar is the external ASCETS aneuploidy value
(data/ascets/, load_ascets_aneuploidy()) or — when ASCETS is absent — the master's
aneu__aneuploidy_score (from data/aneu_scores.csv via report 1) mapped from pid back
onto sequencing barcodes through the report-01 crosswalk. It is a **continuous annotation,
not a clustering feature**: the clustering never sees it, so visible alignment between the
high-burden cluster and the high-aneuploidy samples is corroboration, not circularity.

```{r cnv-cluster, fig.height=8}
cl <- if (have_cnv) cluster_arm_matrix(cnv_mat, k = k_cnv,
                                       method = cnv_method, dist_method = cnv_dist) else NULL

if (!is.null(cl)) {
  # Copy-number-high cluster from the matrix's OWN SCNA burden — derived only from the
  # clustering features, so the aneuploidy-by-cluster boxplot below is an independent check,
  # not a tautology.
  #
  # ⚠️ DIRECTIONAL (this is the ONLY change from the sibling report). The burden is scored
  # over CONCORDANT events only — gains at GISTIC amplification peaks, losses at deletion
  # peaks — instead of mean |value|. On thresholded GISTIC calls the two disagree: |value|
  # rewards a deletion-dominated cluster as strongly as an amplification-dominated one, and
  # on thresholded features it can pick the wrong cluster outright. `direction`
  # comes from this run's own amp_genes/del_genes lists; if they are unavailable the helper
  # falls back to |value| and says so, i.e. this report degrades to the sibling's behaviour.
  feat_dir     <- gistic_feature_direction()
  high_cluster <- cnv_high_cluster_burden(cl, direction = feat_dir)
  burden_tbl   <- cluster_scna_burden(cl, direction = feat_dir)
  high_absolute <- cnv_high_cluster_burden(cl)          # the undirected call, for comparison
  message("Copy-number-high cluster (DIRECTIONAL burden): ", high_cluster,
          " | undirected |value| would pick: ", high_absolute,
          if (identical(high_cluster, high_absolute)) "  [agree]" else "  [DISAGREE]")

  # Per-barcode aneuploidy, kept ONLY for the enrichment check + heatmap annotation:
  # prefer ASCETS, else map the master's aneuploidy score back onto barcodes via the crosswalk.
  aneu_id <- NULL
  if (!is.null(ascets_aneu)) {
    aneu_id <- ascets_aneu |> rename(aneuploidy_value = ascets_aneuploidy)
  } else if (have_master && have_cw) {
    aneu_id <- cw |>
      left_join(master |> transmute(pid, aneuploidy_value =
                  suppressWarnings(as.numeric(.data[[attend_cols$aneuploidy]]))), by = "pid") |>
      transmute(ID = barcode, aneuploidy_value)
  }

  # Heatmap of the GISTIC thresholded peak matrix (TCGA Fig. 1a). Peak genes (rows) x
  # samples (columns by dendrogram), annotated by CN cluster + INTEGRATED TCGA class + MMR +
  # TP53 + the external aneuploidy SCORE. The integrated-class and MMR/TP53 bars show the
  # molecular class distributing ACROSS the copy-number clusters — these are the *copy-number*
  # clusters (Fig. 1a), NOT the integrated four groups (the cascade sits on top). The
  # aneuploidy bar (aneu_id, built above from ASCETS or the master) is a CONTINUOUS annotation
  # only — never a clustering feature.
  ids <- rownames(cl$matrix)
  ann <- data.frame(CN_cluster = factor(cl$cluster[ids]), row.names = ids)
  if (have_master && have_cw) {
    pid_v <- cw$pid[match(ids, cw$barcode)]
    grab  <- function(col) if (col %in% names(master)) master[[col]][match(pid_v, master$pid)] else rep(NA, length(ids))

    # Integrated TCGA class per column (barcode), from the SAME cascade as the "Integrated
    # classification" section below (add_tcga_class): POLE omitted (no cases) > MMRd (MSI-high
    # NGS OR MMR-loss IHC) > CN-high (serous-like) > CN-low. Derived from each patient's
    # dominant CN cluster, so the Fig-1a heatmap shows where the integrated call lands on the
    # copy-number clustering. POLE is kept as an (empty) legend level to document "no cases".
    if (!is.na(high_cluster)) {
      cnv_pid_hm <- cl$assignment |>
        inner_join(cw, by = c("ID" = "barcode")) |>
        count(pid, cnv_cluster) |>
        group_by(pid) |> slice_max(n, n = 1, with_ties = FALSE) |> ungroup() |>
        transmute(pid, cnv_high = cnv_cluster == high_cluster)
      tcga_hm  <- master |> left_join(cnv_pid_hm, by = "pid") |> add_tcga_class()
      tcga_lab <- setNames(as.character(tcga_hm$tcga_class), tcga_hm$pid)[pid_v]
      ann$TCGA_class <- factor(tcga_lab,
        levels = c("POLE (ultramutated)", attend_tcga_levels$msi,
                   attend_tcga_levels$cn_high, attend_tcga_levels$cn_low))
    }

    mmrd  <- is_mmrd(grab("MMR_class"))   # MMRd = MMR IHC loss ONLY (shared is_mmrd; MSI not a criterion)
    ann$MMR <- factor(ifelse(is.na(mmrd), NA, ifelse(mmrd, "MMRd", "MMRp")), levels = c("MMRp", "MMRd"))
    tp <- grab("TP53_pathogenic")
    if (!all(is.na(tp))) ann$TP53 <- factor(ifelse(as.logical(tp), "mut", "wt"), levels = c("wt", "mut"))
  }
  # Aneuploidy score as a continuous annotation bar (same aneu_id used by the enrichment
  # boxplot below; keyed by barcode ID = heatmap column). Kept out of the clustering.
  if (!is.null(aneu_id))
    ann$Aneuploidy <- aneu_id$aneuploidy_value[match(ids, aneu_id$ID)]

  # THE DISPLAY MATRIX, which is not the clustering matrix. Fig. 1a shows the whole genome on
  # a continuous scale; the clustering ran on peak-restricted {-2..+2} calls. Prefer the
  # genome-wide continuous matrix and fall back to the clustering matrix only if GISTIC's
  # all_data_by_genes.txt is absent — in which case the figure is the old, non-faithful one and
  # hm_label says so on the page rather than leaving the reader to guess.
  to_sample_matrix <- function(tb) {
    m <- as.matrix(tb[, setdiff(names(tb), "ID"), drop = FALSE])
    rownames(m) <- tb$ID
    m
  }
  hm_mat <- NULL
  if (!is.null(disp_mat)) {
    dm <- to_sample_matrix(disp_mat)
    dm <- dm[rownames(dm) %in% names(cl$cluster), , drop = FALSE]   # only clustered tumours
    if (nrow(dm) >= 2) hm_mat <- dm else
      cat("all_data_by_genes.txt shares", nrow(dm), "samples with the clustering —",
          "falling back to the thresholded matrix.\n")
  }
  if (!is.null(hm_mat)) {
    hm_src   <- disp_mat
    hm_vtype <- "continuous"
    hm_label <- "genome-wide continuous SCNA (all_data_by_genes) — the Fig-1a display"
  } else {
    hm_mat   <- cl$matrix
    hm_src   <- cnv_mat
    hm_vtype <- "thresholded"
    hm_label <- "FALLBACK: peak-restricted thresholded calls (no all_data_by_genes.txt)"
  }

  # Feature positions: GISTIC genes -> cytoband, from whichever matrix is being DISPLAYED.
  # Taking it from the display matrix matters: the two files carry different gene sets, and
  # cytoband keys looked up from the wrong one resolve to nothing and drop every row.
  fpos <- attr(hm_src, "feature_pos")
  cyto <- if (!is.null(fpos)) setNames(fpos$cytoband, fpos$feature) else NULL
  feat_pos <- feature_positions(colnames(hm_mat), cytoband = cyto)

  # WHY-IS-THERE-NO-HEATMAP gate. fig1a_heatmap() returns invisible(NULL) on four different
  # conditions, each announced only with message() — which opts_chunk (message = FALSE)
  # swallows, so the figure just silently fails to appear. Print the preconditions instead:
  # whichever line below reads FALSE / 0 is the reason the heatmap is missing.
  cat("Fig-1a heatmap preconditions",
      "\n  displaying               :", hm_label,
      "\n  ComplexHeatmap installed :", requireNamespace("ComplexHeatmap", quietly = TRUE),
      "\n  circlize installed       :", requireNamespace("circlize", quietly = TRUE),
      "\n  display (samples x feat) :", nrow(hm_mat), "x", ncol(hm_mat),
      "\n  clustered on             :", nrow(cl$matrix), "samples x", ncol(cl$matrix), "peak genes",
      "\n  samples with a cluster   :", sum(rownames(hm_mat) %in% names(cl$cluster)),
      "\n  feature positions mapped :", sum(feat_pos$feature %in% colnames(hm_mat)),
      "\n  annotation bars          :", if (is.null(ann)) 0L else ncol(ann), "\n")

  fig1a_heatmap(hm_mat, feat_pos, cl$cluster, ann_col = ann,
                value_type = hm_vtype,
                main = paste0("TCGA Fig. 1a — SCNAs by tumour x chromosomal location; ",
                              "columns split by the GISTIC-peak clusters (Suppl. S2), cluster ",
                              high_cluster, " = copy-number high"))
} else {
  cat("CNV clustering skipped — run GISTIC so data/gistic/ exists (see bottom of report).\n")
  high_cluster <- NA_integer_
}
```

## The k-sweep — k = 2…8 on one chromosome heatmap

This is the headline of this variant. The **Pearson/Ward dendrogram is computed once**, then
**cut at every k from 2 to 8**. Because a single hierarchical tree is cut repeatedly, the
partitions are **nested** (k = 8 is a refinement of k = 2), so they can share one figure:
the display matrix is drawn once — genome-wide genes (rows) grouped by chromosome,
tumours (columns) ordered by the **actual Pearson/Ward dendrogram** — with a **stack of
cluster bars beneath, one per k**. Reading a column downward shows how that tumour's cluster
label subdivides as k grows; reading a bar left-to-right shows the k-cluster partition
against the same chromosomal SCNA background. This makes the choice of k a **visual
question** — at which k do the chromosome blocks stop corresponding to coherent SCNA
patterns — rather than a number fixed in advance.

**Computational steps.** (1) Reuse the Pearson/Ward cl$hclust from the clustering chunk.
(2) fig1a_heatmap_ksweep(cl$matrix, feat_pos, cl$hclust, k_range = `r paste0(min(k_sweep), ":", max(k_sweep))`)
draws the matrix once (columns in dendrogram order, no per-k re-clustering) and adds one
cutree(hc, k) cluster bar per k, plus the same top annotation bars (integrated class,
MMR, TP53, aneuploidy score) as the single-k Fig-1a heatmap above. (3) The average
silhouette across k (next chunk) is the quantitative companion to this visual sweep.

```{r cnv-ksweep, fig.height=9}
if (!is.null(cl) && exists("hm_mat") && exists("feat_pos") && exists("ann")) {
  # Same display matrix as the single-k panel, same peak-based dendrogram. ksweep intersects
  # rownames(mat) with hc$labels itself, so the genome-wide matrix and the peak-learned tree
  # line up without any extra handling here.
  fig1a_heatmap_ksweep(hm_mat, feat_pos, cl$hclust, k_range = k_sweep, ann_col = ann,
                       value_type = hm_vtype,
                       main = paste0("Fig-1a canvas (", hm_label, ") — k = ",
                                     min(k_sweep), "..", max(k_sweep),
                                     " Pearson/Ward cluster bars"))
} else cat("k-sweep heatmap skipped — need the Pearson/Ward clustering (run GISTIC first).\n")
```

Average silhouette width against k over the same Pearson/Ward dendrogram, which is the
quantitative companion to the visual sweep above: a peak flags the most internally coherent
cut, and the cascade's configured cut k_cnv is marked for reference. Needs the optional cluster
package.

```{r cnv-ksweep-silhouette, fig.height=5}
# Average silhouette width vs k over the SAME Pearson/Ward dendrogram — the quantitative
# read on "how many clusters". Peaks flag the most internally-coherent cut(s); the cascade's
# the cascade's configured cut k_cnv is marked for reference. Optional dependency `cluster`.
if (!is.null(cl) && requireNamespace("cluster", quietly = TRUE)) {
  D  <- if (identical(cnv_dist, "correlation"))
          stats::as.dist(1 - stats::cor(t(cl$matrix), use = "pairwise.complete.obs"))
        else stats::dist(cl$matrix, method = cnv_dist)
  ks <- k_sweep[k_sweep >= 2 & k_sweep <= (nrow(cl$matrix) - 1)]
  avg_sil <- vapply(ks, function(k) {
    s <- cluster::silhouette(stats::cutree(cl$hclust, k = k), D)
    if (is.matrix(s)) mean(s[, "sil_width"]) else NA_real_
  }, numeric(1))
  sil_df <- data.frame(k = ks, avg_sil = avg_sil)
  ggplot(sil_df, aes(k, avg_sil)) +
    geom_line(colour = "grey50") + geom_point(size = 2) +
    geom_vline(xintercept = k_cnv, linetype = 2,
               colour = unname(ATTEND_PALETTE["red"])) +
    scale_x_continuous(breaks = ks) +
    labs(title = "Average silhouette width vs k (Pearson/Ward)",
         subtitle = paste0("dashed = cascade cut k = ", k_cnv, "; higher = more coherent partition"),
         x = "number of clusters (k)", y = "mean silhouette width") +
    attend_theme()
} else message("Silhouette-vs-k skipped — install the optional `cluster` package.")
```

**TCGA Fig. 1b — progression-free survival by copy-number cluster.** KM of clinical
PFS (gianlu__PFS_MONTHS/_EVENT via recode_event()) for the raw copy-number clusters,
barcodes mapped to pid through the report-01 crosswalk. Log-rank P as in the paper.

```{r cnv-km-pfs, fig.height=5}
if (!is.null(cl) && have_master && have_cw) {
  km_by_cluster(cl$assignment, cw, master)
} else message("KM skipped — need CNV clusters + master + crosswalk.")
```

Per-sample silhouette at the cascade cut k_cnv, computed on the same 1 - Pearson
distance the clustering used rather than Euclidean, so the widths are consistent with the
tree the clusters came from.

```{r cnv-silhouette, fig.height=5}
# Per-sample silhouette at the cascade cut k = k_cnv, using the SAME 1 - Pearson distance the
# clustering used (not Euclidean) so the widths are consistent with the Pearson/Ward tree.
if (!is.null(cl) && requireNamespace("cluster", quietly = TRUE)) {
  D_sil <- if (identical(cnv_dist, "correlation"))
             stats::as.dist(1 - stats::cor(t(cl$matrix), use = "pairwise.complete.obs"))
           else stats::dist(cl$matrix, method = cnv_dist)
  sil <- cluster::silhouette(cl$cluster[rownames(cl$matrix)], D_sil)
  plot(sil, main = paste0("Silhouette (Pearson/Ward), k = ", k_cnv), col = "grey40")
}
```

**These are the *copy-number* clusters, not the four integrated molecular groups.** The
heatmap above is Fig. 1a — unsupervised clustering of *copy number alone*, cluster 4 =
serous-like. The four *integrated groups* (Fig. 2b: POLE → MMRd → CN-high → CN-low)
are a **separate object** — a cascade layered on top, where MMRd is set by MMR IHC (not
copy number) and the SCNA clusters collapse to high (cluster 4) vs low (1–3). The
integrated-class, MMR and TP53 annotation bars show exactly this: the molecular class
distributes *across* the CN clusters rather than defining them — the integrated-class bar is
the cascade's per-patient call painted onto the copy-number clustering, not a re-clustering.

## Aneuploidy enrichment across clusters

The copy-number-high cluster was labelled **above** from the arm matrix's own **SCNA
burden** (mean |arm copy number|) — a signal built only from the clustering features.
This section is therefore an **independent check**: does that same cluster also carry
the highest **external aneuploidy score**? Because the label and the plotted score come
from *different* signals, a positive result here is genuine enrichment, not the
tautology you'd get by labelling the cluster with the very score you then plot. The
burden table shows why the cluster was chosen; the boxplot (Kruskal–Wallis,
red/turquoise palette) is the cross-check.

**Computational steps.** (1) cluster_scna_burden() — mean |arm log2| per sample,
averaged per cluster (the table). (2) Join the **external** per-sample aneuploidy score
(ASCETS if present, else the master's aneu__aneuploidy_score via the crosswalk) to the
cluster assignment. (3) Boxplot aneuploidy by cluster; **Kruskal–Wallis** across clusters.

```{r cnv-burden-table}
if (!is.null(cl) && !is.null(burden_tbl)) {
  knitr::kable(burden_tbl, digits = 3,
    caption = paste0("Intrinsic SCNA burden per cluster (mean |arm copy number|) — ",
                     "cluster ", high_cluster, " is copy-number-high. This is the label ",
                     "signal; the aneuploidy boxplot below is the independent check."))
}
```

The independent check on that label: the external per-sample aneuploidy score
(data/ascets/ when present, else the master's aneu__aneuploidy_score) by cluster. The
clustering never saw this value, so alignment between the high-burden cluster and the
high-aneuploidy samples is corroboration rather than circularity.

```{r cnv-aneu-enrich}
if (!is.null(cl) && exists("aneu_id") && !is.null(aneu_id)) {
  d <- cl$assignment |>
    inner_join(aneu_id, by = "ID") |>
    filter(!is.na(aneuploidy_value)) |>
    mutate(cluster = factor(cnv_cluster),
           is_high = cnv_cluster == high_cluster)
  if (exists("cw") && !is.null(cw)) d <- dplyr::left_join(d, cw, by = c("ID" = "barcode"))

  if (nrow(d) > 0) {
    ggplot(d, aes(cluster, aneuploidy_value)) +
      # fill kept (is_high names which cluster is serous-like — the x tick does not), but
      # alpha dropped: blending the fill against white means the rendered box never matches
      # its own legend swatch. CN-high takes the aneuploidy-high red, since that is
      # the same "altered" end of the same biology one panel over.
      geom_boxplot(outlier.shape = NA, aes(fill = is_high), width = 0.6) +
      aneu_point_layers(d, value_col = "aneuploidy_value") +
      stat_compare_means(label = "p.format", size = 3) +
      scale_fill_manual(values = c(`TRUE` = attend_aneu_high, `FALSE` = attend_neutral),
                        name = "copy-number high", labels = c("other", "serous-like")) +
      labs(title = "Aneuploidy score by CNV cluster (independent check)",
           subtitle = paste0("source: ", cnv_source, "; cluster ", high_cluster,
                             " = copy-number high (labelled by SCNA burden, not this score)"),
           x = "CNV cluster", y = "aneuploidy score") +
      attend_theme()
  }
}
```

## Integrated classification cascade

**Computational steps.** (1) Per patient, collapse barcode → pid and take the
**dominant** CNV cluster (multi-sample patients), then flag cnv_high = membership in
the CN-high cluster. (2) Join to the master and apply add_tcga_class(), a **priority
cascade**: **MMRd** (MSI-high by NGS **OR** MMR-protein loss by IHC) > **CN-high**
(serous rule = `r attend_tcga_serous_rule`) > **CN-low**; the first positive test wins.
(3) Tabulate patients per class.

```{r cascade}
tcga_summary <- NULL

if (have_master && have_cw) {
  # Cascade (POLE removed): MMRd (MMR IHC) > CN-high (serous-like) > CN-low. MMRd = MMR-IHC-deficient
  # (MMR IHC only; MSI-high is a correlate); the serous-like call follows attend_tcga_serous_rule (default
  # "cluster_only": high-aneuploidy CNV cluster membership = TCGA's serous-like cluster 4,
  # TP53 NOT required — validated separately below). add_tcga_class(), attend_classes.R §12.

  # CNV cluster per patient (barcode -> pid; dominant cluster if multi-sample),
  # then the copy-number-high flag.
  if (!is.null(cl) && !is.na(high_cluster)) {
    cnv_pid <- cl$assignment |>
      inner_join(cw, by = c("ID" = "barcode")) |>
      count(pid, cnv_cluster) |>
      group_by(pid) |> slice_max(n, n = 1, with_ties = FALSE) |> ungroup() |>
      transmute(pid, cnv_high = cnv_cluster == high_cluster)
  } else {
    cnv_pid <- tibble(pid = character(), cnv_high = logical())
  }

  classed <- master |>
    left_join(cnv_pid, by = "pid") |>
    add_tcga_class()   # serous = high-aneuploidy cluster ("cluster_only"); TP53 validated, not gated

  tcga_summary <- classed |>
    count(tcga_class, .drop = FALSE) |>
    mutate(pct = round(100 * n / sum(n), 1))
  knitr::kable(tcga_summary, caption = paste0("Patients per TCGA integrated class (MMRd > serous-like > endometrioid; serous rule = '", attend_tcga_serous_rule, "')"))
}
```

### TP53-enrichment QC — is the CN-high cluster really serous-like?

With the faithful "cluster_only" rule, the serous-like call is made **from copy number
alone**. TCGA's serous-like cluster 4 was **~90% *TP53*-mutant** — so if our CN-high
cluster genuinely corresponds to it, it should come out **independently** enriched for
*TP53* pathogenic mutations. This is the validation that replaces the TP53 *gate*: TP53
is checked as a **property**, not required as a **criterion**. A high fraction here
confirms the cluster is serous-like; a low fraction is a red flag that the copy-number
cluster does **not** map to TCGA's group (reconsider k_cnv, the CNV input, or purity).

**Computational steps.** (1) Split patients by cnv_high (CN-high cluster vs the rest).
(2) Per group, count TP53_pathogenic and compute the mutant fraction. (3) **Fisher exact
test** on the 2×2 (cluster × TP53) for enrichment; stacked bar with a dashed **90%** line.

```{r tp53-enrichment}
if (exists("classed") && all(c("cnv_high", "TP53_pathogenic") %in% names(classed))) {
  tp <- classed |>
    filter(!is.na(cnv_high)) |>
    mutate(cluster = ifelse(cnv_high, "CN-high cluster", "other clusters"),
           tp53    = suppressWarnings(as.logical(TP53_pathogenic))) |>
    filter(!is.na(tp53))

  if (nrow(tp) > 0 && dplyr::n_distinct(tp$cluster) >= 1) {
    tab <- tp |>
      group_by(cluster) |>
      summarise(n = dplyr::n(), TP53_mut = sum(tp53),
                pct_TP53 = round(100 * mean(tp53), 1), .groups = "drop")
    knitr::kable(tab, caption = "TP53-pathogenic fraction: CN-high cluster vs the rest (TCGA serous-like cluster 4 was ~90% TP53-mut)")

    ct <- table(factor(tp$cluster, levels = c("other clusters", "CN-high cluster")), tp$tp53)
    if (all(dim(ct) == c(2, 2)) && min(rowSums(ct)) > 0) {
      ft <- fisher.test(ct)
      message("TP53 enrichment in the CN-high cluster: Fisher p = ", signif(ft$p.value, 3),
              ", OR = ", signif(unname(ft$estimate), 3),
              if (is.finite(ft$p.value) && ft$p.value < 0.05) "  (enriched — cluster is serous-like)"
              else "  (NOT clearly enriched — cluster may not map to TCGA serous-like)")
    }

    ggplot(tp, aes(cluster, fill = tp53)) +
      geom_bar(position = "fill", width = 0.6) +
      scale_y_continuous(labels = scales::percent) +
      scale_fill_manual(values = c(`TRUE` = unname(attend_tp53_cols[["mut"]]),
                                   `FALSE` = unname(attend_tp53_cols[["wt"]])),
                        name = "TP53 pathogenic", labels = c("wild-type", "mutant")) +
      geom_hline(yintercept = 0.90, linetype = 2, colour = "grey30") +
      labs(title = "TP53 mutation fraction by copy-number cluster",
           subtitle = "dashed = TCGA serous-like ~90% TP53-mut benchmark",
           x = NULL, y = "% TP53 pathogenic") +
      attend_theme()
  } else message("TP53-enrichment QC: no CN-high assignments with TP53 status yet.")
} else {
  message("TP53-enrichment QC skipped — needs cnv_high (from clustering) + TP53_pathogenic on the master.")
}
```

**Computational steps (next two panels).** (1) Take the per-patient tcga_class from the
cascade. (2) **Aneuploidy panel:** aneuploidy score by class, Kruskal–Wallis. (3) **TMB
panel:** loop over every TMB definition on the master (tmb_defs_present() — baseline
tmb__TMB_SCORE and the ancestry-corrected column) and box TMB by class, one facet per
definition.

```{r cascade-panels, fig.height=6}
# Aneuploidy by class (paper-style panel).
if (exists("classed")) {
  pd <- classed |>
    mutate(tcga_class = forcats::fct_na_value_to_level(tcga_class, "unclassified"),
           aneuploidy_value = suppressWarnings(as.numeric(.data[[attend_cols$aneuploidy]])))

  ggplot(filter(pd, !is.na(aneuploidy_value)), aes(tcga_class, aneuploidy_value)) +
    # Was filled #A6CEE3 = attend_aneu_low, i.e. this figure told the reader every class was
    # aneuploidy-low. The integrated classes are the smallest strata in the cascade, so the
    # mark is chosen per class; points come from aneu_point_layers().
    attend_box(filter(pd, !is.na(aneuploidy_value)), x = "tcga_class", y = "aneuploidy_value",
               points = FALSE) +
    aneu_point_layers(filter(pd, !is.na(aneuploidy_value)), value_col = "aneuploidy_value") +
    stat_compare_means(label = "p.format", size = 3) +
    labs(title = "Aneuploidy by integrated class", x = NULL, y = "Aneuploidy score (0-1)") +
    attend_theme() + theme(axis.text.x = element_text(angle = 20, hjust = 1))
}
```

**TMB by integrated class, under both TMB definitions** — **TMB (normal)**, the upstream
tumour-only score, and **TMB (ancestry corrected)**, the same score after the ancestry
recalibration explained in the "Ancestry-corrected TMB" section below. One panel per
definition, so each class's TMB can be read under both. The dashed line is the
`r attend_thresholds$tmb` mut/Mb TMB-high cutoff.

```{r cascade-panels-tmb, fig.height=6}
if (exists("classed") && length(tmb_defs) > 0) {
  tl <- classed |>
    mutate(tcga_class = forcats::fct_na_value_to_level(tcga_class, "unclassified")) |>
    select(pid, tcga_class, all_of(unname(tmb_defs))) |>
    pivot_longer(all_of(unname(tmb_defs)), names_to = "tmb_col", values_to = "tmb_value") |>
    mutate(tmb_def   = factor(tmb_def_labels(tmb_defs)[match(tmb_col, tmb_defs)],
                              levels = tmb_def_labels(tmb_defs)),
           tmb_value = suppressWarnings(as.numeric(tmb_value))) |>
    filter(!is.na(tmb_value))

  ggplot(tl, aes(tcga_class, tmb_value)) +
    attend_box(tl, x = "tcga_class", y = "tmb_value", by = "tmb_def") +
    geom_hline(yintercept = attend_thresholds$tmb, linetype = 2, colour = "grey50") +
    facet_wrap(~ tmb_def, scales = "free_y", nrow = 1) +
    labs(title = "TMB by integrated class, per TMB definition",
         subtitle = paste0("dashed line = TMB-high cutoff (", attend_thresholds$tmb, " mut/Mb)"),
         x = NULL, y = "TMB (mut/Mb)") +
    attend_theme() + theme(axis.text.x = element_text(angle = 25, hjust = 1))
}
```


```{r session-info}
sessionInfo()
```
