---
title: "ATTEND — 10. Clustering on TCGA's published GISTIC peaks"
author: "sceriff0"
date: "2026-08-17"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---
```{r setup, include=FALSE}
# renv: tidyverse, here (+ optional bioc::ComplexHeatmap, circlize for the peak heatmaps).
# If missing: renv::install(c("bioc::ComplexHeatmap","circlize")); renv::snapshot()
library(tidyverse)
library(here)
# Headless-safe raster device. The cluster's compute nodes have no X11, so R's default
# png() fails with "unable to open connection to X11 display" and every figure dies. cairo
# needs no display; the guard keeps this working on a machine built without it.
if (capabilities("cairo")) options(bitmapType = "cairo")

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 11, fig.height = 8)

source(here("code", "attend_classes.R"))     # attend_tcga_ref_2013, k_cnv, cluster_arm_matrix(),
                                             # adjusted_rand_index(), cnv_high_cluster_burden()
source(here("code", "attend_plots.R"))       # fig1a_heatmap(), feature_positions(), attend_theme()
source(here("code", "attend_io.R"))          # read_intermediate()
source(here("code", "build_master.R"))       # get_master() — builds on demand
source(here("code", "load_wes_results.R"))   # load_tcga_published_peaks(), the two peak-matrix
                                             # builders, load_gistic_thresholded()

ref <- attend_tcga_ref_2013
# k_cnv and cnv_method/cnv_dist are configuration and live in attend_classes.R with every
# other method parameter, so 00-methods.Rmd can name the cut it justifies. This report
# deliberately clusters with the SAME call report 09 uses, so any difference between the two
# is the feature set and not the method.
```
> **Question.** Does clustering on TCGA's *published* GISTIC peaks — an external feature set fitted to neither cohort — recover TCGA's own copy-number partition, and do ATTEND's clusters correspond to TCGA's in that shared space?
> **Cohort.** TCGA UCEC 2013 (all classified tumours) and all ATTEND patients with GISTIC output, in one 79-peak space.
> **Reads.** data/tcga_gistic_peaks_2013.tsv (Kandoth et al. Suppl. Data File S2.1), data/tcga_2013/ (clinical + hg19 segments), and ATTEND's own all_thresholded.by_genes.txt in data/gistic/.
> **Out of scope.** The ATTEND cascade itself, and which ATTEND cluster is serous-like -> [09-tcga-classification](09-tcga-classification.html). Why 1 - Pearson / Ward.D2 is the pipeline's only distance/linkage pair -> [00-methods](00-methods.html). Producing the GISTIC input -> runbooks/gistic.md.

[09-tcga-classification](09-tcga-classification.html) clusters ATTEND on peaks that GISTIC
learned **from ATTEND itself**. That is circular in a specific, quantifiable way: the
features are fitted to the same cohort they then partition, so peak membership shifts when
samples are added or removed. A TCGA run of our own has the mirror-image problem — it uses
**our** parameters, because TCGA never published theirs (the Supplementary Methods say only
*"identified from segmented data using GISTIC 2.0"*, with no `-ta`/`-td`/`-conf`/reference
build anywhere).

This report removes both problems by using an **external, pre-registered feature set**:
TCGA's own published peak list, Supplementary Data File S2.1, "all tumor" sheets, converted
once to `data/tcga_gistic_peaks_2013.tsv`. The peaks were defined before either analysis and
are not a function of either cohort's samples, so there is nothing to leave one out of; the
TCGA arm needs no GISTIC run at all, because the regions are given and only each sample's
copy number inside them has to be read; and both cohorts land in **one shared feature
space**, which is what makes an ATTEND-vs-TCGA cluster comparison a comparison between like
and like.

It is **not** a reproduction of the paper's *clustering*. The Methods say only
*"hierarchical clustering... using R"*, never the distance or linkage, so those remain this
pipeline's choice.

## Part 1 — the external feature set

One row per peak × gene as delivered by `load_tcga_published_peaks()`, which reads
`data/tcga_gistic_peaks_2013.tsv`: `peak_id`, `direction` (amp/del), `cytoband`, GISTIC
`q_value` and `residual_q`, hg19 `chrom`/`start`/`end`, and `gene`. The table below counts
distinct peaks and distinct genes per direction. Three cytobands carry two distinct peaks
each (`amp:8q24.21`, `amp:3q26.2`, `amp:19p13.2`) and are suffixed `#2` in the source file so
they stay separate rather than silently merging into one column.

```{r peaks}
peaks <- tryCatch(load_tcga_published_peaks(), error = function(e) tibble())
have_peaks <- nrow(peaks) > 0
if (!have_peaks) {
  cat("**data/tcga_gistic_peaks_2013.tsv missing** — every panel below is skipped.\n")
} else {
  knitr::kable(
    peaks |>
      group_by(direction) |>
      summarise(peaks = dplyr::n_distinct(peak_id),
                genes = dplyr::n_distinct(gene), .groups = "drop"),
    caption = "TCGA 2013 published GISTIC peaks (Suppl. Data File S2.1, 'all tumor' sheets).")
}
```

The strongest peaks by GISTIC q-value, with representative member genes. This is a check
that the file is the one it claims to be rather than a result: ERBB2/17q12 and CCNE1/19q12
amplifications and the PTEN/10q23.31 deletion are the canonical endometrial hits, so their
absence from the top of this table would mean the conversion went wrong.

```{r peaks-top}
if (have_peaks) {
  knitr::kable(
    peaks |> distinct(peak_id, direction, cytoband, q_value, chrom, start, end) |>
      group_by(direction) |> slice_min(q_value, n = 5, with_ties = FALSE) |> ungroup() |>
      left_join(peaks |> group_by(peak_id) |>
                  summarise(n_genes = n(),
                            example = paste(head(gene, 3), collapse = ", "), .groups = "drop"),
                by = "peak_id") |>
      arrange(direction, q_value),
    caption = "Top 5 amplification and deletion peaks by q-value, with representative genes.")
}
```

## Part 2 — the two cohorts, on one feature set

TCGA's segments (`data/tcga_2013/data_cna_hg19.seg`, via `load_tcga_2013_seg()`) are hg19,
the same build as the published peak boundaries, so the peaks apply directly:
`published_peak_matrix_from_seg()` takes, for each sample × peak, the **length-weighted mean
log2 seg-mean across the peak's wide boundaries**, and sets a peak with no overlapping
segment to 0 (copy-neutral).

Those values are then **thresholded**, because the Supplementary Methods are explicit that
clustering ran on *"**thresholded** relative copy number data in significantly reoccurring
amplifications or deletions regions"* — not on continuous ratios. Discretising to
{−2,−1,0,+1,+2} also matches the ATTEND route below, whose GISTIC gene calls are already
thresholded, so the two cohorts share a representation as well as a feature set. The paper
does not state its cut points, so **both forms are built** and scored against each other in
Part 3 rather than one being assumed correct.

```{r tcga-matrix}
tcga_seg  <- tryCatch(load_tcga_2013_seg(),  error = function(e) tibble())
tcga_clin <- tryCatch(load_tcga_ucec_2013(), error = function(e) tibble())

# |x| < 0.1 -> 0 ; 0.1 <= |x| < 0.9 -> +-1 (low-level) ; |x| >= 0.9 -> +-2 (high-level).
# Not TCGA's cut points (never published); the continuous matrix below is the control that
# says what this choice is worth.
thr_cuts <- c(0.1, 0.9)

tcga_mat_cont <- if (have_peaks && nrow(tcga_seg) > 0)
  published_peak_matrix_from_seg(tcga_seg, peaks) else NULL
tcga_mat <- if (have_peaks && nrow(tcga_seg) > 0)
  published_peak_matrix_from_seg(tcga_seg, peaks, threshold = thr_cuts) else NULL

have_tcga <- !is.null(tcga_mat) && nrow(tcga_mat) >= k_cnv
# cat(), not message(): opts_chunk sets message = FALSE, so a message() here is invisible on
# the knitted page and a skipped section then looks like nothing happened at all.
if (have_tcga) {
  cat("TCGA peak matrix:", nrow(tcga_mat), "samples x", ncol(tcga_mat) - 1,
      "peaks (thresholded at", paste(thr_cuts, collapse = "/"), ")\n")
} else cat("**TCGA peak matrix unavailable** — run `Rscript code/fetch_tcga_ucec_2013.R`.\n")
```

ATTEND is **hg38**, so the hg19 peak *coordinates* do not transfer — but gene symbols do.
`load_gistic_thresholded()` reads ATTEND's own `all_thresholded.by_genes.txt` (genes ×
samples, values already in {−2,−1,0,+1,+2}), and `published_peak_matrix_from_genes()` subsets
it to the genes TCGA reported inside its peaks and **averages the per-gene calls within each
peak**, putting ATTEND on the same peak columns as TCGA. No liftover is involved.

The two routes produce comparable *structure*, not comparable *units*: TCGA's values are
length-weighted mean log2 ratios, ATTEND's are thresholded GISTIC calls averaged per peak.
Signs and relative magnitudes correspond, absolute values do not — which is why the
cross-cohort comparison in Part 4 z-scores within cohort before correlating anything. Two
further limits belong here rather than beneath the figure: gene-symbol matching depends on
nomenclature, so genes renamed since 2013 drop out silently (the loader reports how many of
the ~5,000 matched), and a peak with no member gene present in ATTEND is filled with 0,
which is a conservative assumption and not a measurement.

```{r attend-matrix}
attend_thr <- tryCatch(load_gistic_thresholded(), error = function(e) NULL)
attend_mat <- if (have_peaks && !is.null(attend_thr))
  published_peak_matrix_from_genes(attend_thr, peaks, collapse = TRUE) else NULL
have_attend <- !is.null(attend_mat) && nrow(attend_mat) >= k_cnv

if (have_attend) {
  cat("ATTEND peak matrix:", nrow(attend_mat), "samples x", ncol(attend_mat) - 1, "peaks\n")
} else cat("**No ATTEND GISTIC output** — populate data/gistic/ (see runbooks/gistic.md).",
           "The TCGA panels still render.\n")
```

Both cohorts are clustered with one identical call — the `cnv_dist` / `cnv_method` pair
configured in `attend_classes.R` and used by [09-tcga-classification](09-tcga-classification.html),
cut at the configured `k_cnv` — so any difference between the cohorts below is the data and
not the method. The peak "position" used for the heatmap rows comes from each peak's
cytoband, so the vertical axis stays genomic even though the features are peaks rather than
bins.

```{r cluster}
cl_tcga   <- if (have_tcga)
  cluster_arm_matrix(tcga_mat,   k = k_cnv, method = cnv_method, dist_method = cnv_dist) else NULL
cl_attend <- if (have_attend)
  cluster_arm_matrix(attend_mat, k = k_cnv, method = cnv_method, dist_method = cnv_dist) else NULL

peak_cyto <- if (have_peaks)
  peaks |> distinct(peak_id, cytoband) |> tibble::deframe() else NULL

# Annotation bars for the TCGA heatmap come from TCGA's own published fields, so the
# heatmap can be read against the paper: integrative subtype, the MSI-derived MMR call, and
# the retrofitted PanCancer-Atlas aneuploidy score.
ann_tcga <- if (nrow(tcga_clin) > 0) {
  a <- tcga_clin |>
    transmute(ID, TCGA_class = subtype_short,
              MMR = factor(ifelse(mmrd, "MMRd", "MMRp"), levels = c("MMRp", "MMRd")),
              Aneuploidy = aneuploidy) |>
    as.data.frame()
  rownames(a) <- a$ID; a$ID <- NULL; a
} else NULL
```

TCGA UCEC on TCGA's published peaks. Rows are the published peaks placed by cytoband,
columns are tumours ordered by the clustering above, and the body is the length-weighted
mean log2 ratio inside each peak — red for gain, blue for loss, the same direction encoding
the ATTEND SCNA panels use in [07-mutation-and-cna](07-mutation-and-cna.html). The
annotation bars are TCGA's published subtype, its MSI-derived MMR call and its aneuploidy
score, none of which entered the clustering.

```{r hm-tcga, fig.height=8}
if (!is.null(cl_tcga)) {
  fp <- feature_positions(colnames(cl_tcga$matrix), cytoband = peak_cyto)
  fig1a_heatmap(cl_tcga$matrix, fp, cl_tcga$cluster, ann_col = ann_tcga,
                value_type = "continuous",
                main = paste0("TCGA UCEC 2013 on TCGA's published GISTIC peaks — k = ", k_cnv))
} else cat("TCGA peak heatmap skipped.\n")
```

ATTEND on the same peaks, same clustering call, same row order. There are no annotation bars
here because the point of the panel is the copy-number pattern itself; the ATTEND
annotation stack is in [09-tcga-classification](09-tcga-classification.html).

```{r hm-attend, fig.height=8}
if (!is.null(cl_attend)) {
  fp <- feature_positions(colnames(cl_attend$matrix), cytoband = peak_cyto)
  fig1a_heatmap(cl_attend$matrix, fp, cl_attend$cluster, ann_col = NULL,
                value_type = "continuous",
                main = paste0("ATTEND on TCGA's published GISTIC peaks — k = ", k_cnv))
} else cat("ATTEND peak heatmap skipped — needs data/gistic/.\n")
```

## Part 3 — does the external feature set recover TCGA's own answer?

The validation that licenses everything else: clustering TCGA on TCGA's published peaks
should recover TCGA's published cluster assignment (`CNA_CLUSTER_K4`, carried by
`load_tcga_ucec_2013()`). Scored with the **adjusted Rand index**, which is chance-corrected
so it is not inflated by cluster-size imbalance, on the tumours that carry a published
assignment; both the thresholded and continuous representations are scored so the difference
that thresholding makes is reported rather than assumed.

Three caveats belong with the numbers rather than after them. **First**, `burden_pick` is
what `cnv_high_cluster_burden()` chooses as the serous-like cluster and `best_clust` is the
cluster that actually overlaps published cluster 4 most; the burden rule is a *heuristic*
(greatest mean |value|), and on thresholded features mean |value| counts a deep deletion and
a high amplification identically, so a small deletion-heavy cluster can outrank the
genuinely serous-like one. **When the two columns disagree, the heuristic has failed** — and
that failure would otherwise propagate silently into any burden-based serous call. **Second**,
the contingency table is printed underneath precisely so a high ARI with a mislabelled
CN-high cluster cannot pass unnoticed. **Third**, a comparison against a same-cohort GISTIC
run of our own is not available here, because this repository has no such TCGA run; the
question of whether published peaks beat a locally-fitted set is therefore left open rather
than answered.

```{r validate}
score_matrix <- function(mat, label) {
  cl <- cluster_arm_matrix(mat, k = k_cnv, method = cnv_method, dist_method = cnv_dist)
  if (is.null(cl)) return(NULL)
  pub <- tcga_clin |> filter(!is.na(cn_cluster_k4)) |>
    transmute(ID, published = as.character(cn_cluster_k4))
  j  <- cl$assignment |> inner_join(pub, by = "ID")
  if (nrow(j) == 0) return(NULL)
  hi <- cnv_high_cluster_burden(cl)
  tp <- sum(j$cnv_cluster == hi & j$published == "4")
  fp <- sum(j$cnv_cluster == hi & j$published != "4")
  fn <- sum(j$cnv_cluster != hi & j$published == "4")
  # ORACLE, for the heuristic-failure check described above: the cluster with the largest
  # overlap with published 4. Reporting both makes a wrong burden pick visible.
  best <- j |> filter(published == "4") |> count(cnv_cluster) |>
    slice_max(n, n = 1, with_ties = FALSE)
  bo   <- best$cnv_cluster[1]
  btp  <- sum(j$cnv_cluster == bo & j$published == "4")
  bfp  <- sum(j$cnv_cluster == bo & j$published != "4")
  bfn  <- sum(j$cnv_cluster != bo & j$published == "4")
  list(cl = cl, j = j, row = tibble(
    representation = label, n_scored = nrow(j),
    ARI = round(adjusted_rand_index(j$cnv_cluster, j$published), 3),
    burden_pick = as.character(hi),
    burden_F1   = round(2 * tp / max(1, 2 * tp + fp + fn), 3),
    best_clust  = as.character(bo),
    best_prec   = round(btp / max(1, btp + bfp), 3),
    best_recall = round(btp / max(1, btp + bfn), 3),
    best_F1     = round(2 * btp / max(1, 2 * btp + bfp + bfn), 3)))
}

s_thr <- s_cont <- NULL
if (!is.null(cl_tcga) && nrow(tcga_clin) > 0 && "cn_cluster_k4" %in% names(tcga_clin)) {
  s_thr  <- score_matrix(tcga_mat,
                         paste0("thresholded {-2..+2} @ ", paste(thr_cuts, collapse = "/")))
  s_cont <- score_matrix(tcga_mat_cont, "continuous log2")
}
if (!is.null(s_thr)) {
  knitr::kable(bind_rows(s_thr$row, s_cont$row), caption = paste0(
    "Clustering TCGA on TCGA's own published peaks, scored against its published ",
    "CNA_CLUSTER_K4. The paper specifies THRESHOLDED data; the continuous row shows what ",
    "that instruction is worth. burden_pick vs best_clust disagreeing means the ",
    "greatest-mean-|value| serous heuristic has failed on this representation."))
} else cat("Validation skipped — needs data/tcga_2013/ with CNA_CLUSTER_K4.\n")
```

The contingency behind the thresholded row: recomputed clusters as rows, TCGA's published
clusters as columns. A concentrated off-diagonal block is the recovery the ARI summarises;
mass spread across a row means that recomputed cluster has no published counterpart.

```{r validate-contingency}
if (!is.null(s_thr)) {
  knitr::kable(as.data.frame.matrix(table(recomputed = s_thr$j$cnv_cluster,
                                          published  = s_thr$j$published)),
               caption = "Thresholded (faithful) representation: recomputed rows vs published columns.")
} else cat("Contingency skipped.\n")
```

## Part 4 — do the two cohorts' clusters mean the same thing?

Because both matrices span the same peaks, each cluster can be summarised as a **centroid**
in that space and the cohorts' centroids correlated. Values are **z-scored within cohort
first**, since the two routes differ in units (Part 2), so the correlation is about the
*pattern* of peaks and not amplitude. Each ATTEND cluster correlating strongly with exactly
one TCGA cluster is what would let an ATTEND cluster be named using TCGA's definition rather
than an internal guess; a row with no strong partner is a copy-number configuration with no
primary-disease counterpart, which is the observation
[11-mmrd-crosscohort](11-mmrd-crosscohort.html) pursues on the aneuploidy axis. With four
clusters per side the correlation matrix is 4 × 4 and every entry is estimated from a handful
of tumours, so read the *ranking* within a row, not the absolute coefficient.

```{r cross-cohort, fig.height=5}
if (!is.null(cl_tcga) && !is.null(cl_attend)) {
  shared <- intersect(colnames(cl_tcga$matrix), colnames(cl_attend$matrix))
  zc <- function(cl, tag) {
    m <- scale(cl$matrix[, shared, drop = FALSE])          # z-score within cohort
    m[!is.finite(m)] <- 0
    t(sapply(split(seq_len(nrow(m)), cl$cluster[rownames(cl$matrix)]),
             function(i) colMeans(m[i, , drop = FALSE]))) |>
      as.data.frame() |> tibble::rownames_to_column("cluster") |> mutate(cohort = tag)
  }
  cen <- bind_rows(zc(cl_tcga, "TCGA"), zc(cl_attend, "ATTEND"))
  M   <- as.matrix(cen[, shared, drop = FALSE])
  rownames(M) <- paste(cen$cohort, cen$cluster)
  cm  <- stats::cor(t(M))
  ta  <- cm[grep("^TCGA", rownames(cm)), grep("^ATTEND", colnames(cm)), drop = FALSE]
  knitr::kable(round(ta, 2), caption = paste0(
    "Pearson correlation between TCGA cluster centroids (rows) and ATTEND cluster centroids ",
    "(columns) across the ", length(shared), " shared published peaks, z-scored within ",
    "cohort. A high off-diagonal value pairs an ATTEND cluster with its TCGA counterpart."))
} else cat("Cross-cohort centroid comparison needs BOTH cohorts' matrices.\n")
```

```{r session-info}
sessionInfo()
```
