---
title: "ATTEND — 10. TCGA UCEC 2013 — Fig. 1a reproduced on TCGA's own data"
author: "sceriff0"
date: "2026-08-17"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---
```{r setup, include=FALSE}
# renv: tidyverse, here (+ optional bioc::ComplexHeatmap, circlize for the Fig-1a heatmaps).
# If missing: renv::install(c("bioc::ComplexHeatmap","circlize")); renv::snapshot()
# Needs data/tcga_2013/ — Rscript code/fetch_tcga_ucec_2013.R
library(tidyverse)
library(here)
# Headless-safe raster device. The cluster's compute nodes have no X11, so R's default
# png() fails with "unable to open connection to X11 display" and every figure dies. cairo
# needs no display; the guard keeps this working on a machine built without it.
if (capabilities("cairo")) options(bitmapType = "cairo")

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 11, fig.height = 8)

source(here("code", "attend_classes.R"))     # k_cnv, k_sweep, cnv_method, cnv_dist,
                                             # segments_to_bin_matrix(), cluster_arm_matrix(),
                                             # adjusted_rand_index(), cnv_high_cluster_burden()
source(here("code", "attend_plots.R"))       # fig1a_heatmap(), feature_positions(), attend_box()
source(here("code", "attend_io.R"))          # read_intermediate()
source(here("code", "load_wes_results.R"))   # load_tcga_ucec_2013(), load_tcga_2013_seg(),
                                             # load_tcga_published_peaks(), load_tcga_2013_ascets(),
                                             # published_peak_matrix_from_seg()

# The clustering cut and metric are configuration and live in attend_classes.R with every
# other method parameter, so this report and 00-methods.Rmd cannot disagree about them. Using
# the SAME pair as report 09 is deliberate: any difference between the two reports is then the
# cohort and the platform, never the method.
```
> **Question.** Run on TCGA's own data, does this pipeline's Fig-1a reproduction recover TCGA's published copy-number partition — and where do MMR status and aneuploidy sit on that figure?
> **Cohort.** TCGA UCEC 2013, every tumour carrying SNP6 segments.
> **Reads.** data/tcga_2013/ (data_cna_hg19.seg, the two clinical tables, pancan_aneuploidy.tsv), data/tcga_gistic_peaks_2013.tsv, and data/arm_boundaries_hg19.tsv.
> **Out of scope.** The ATTEND cohort and its cascade -> [09-tcga-classification](09-tcga-classification.html). Why 1 - Pearson / Ward.D2, and why the heatmap draws a different matrix from the one it clusters -> [00-methods](00-methods.html).

This is [09-tcga-classification](09-tcga-classification.html)'s figure, built the same way,
on **TCGA's own tumours**. It exists because here **the right answer is published**: the 2013
study ships its own copy-number cluster assignment (`CNA_CLUSTER_K4`), so the pipeline can be
scored rather than merely inspected. If it cannot recover TCGA's partition on TCGA's data,
report 09's ATTEND clusters need a second look before they are interpreted.

It also draws something the paper could not. TCGA 2013 has **no per-sample aneuploidy score**
— the score used here is retrofitted from the later PanCancer-Atlas work — so the
relationship between the four SCNA clusters and aneuploidy was never plotted. Every heatmap
below carries an aneuploidy bar, and Part 4 makes the comparison explicit.

Two properties make this a cleaner test than report 09 can be:

* **The peaks are published.** Report 09 clusters peaks GISTIC learned *from ATTEND*, and a
  TCGA run of our own would need parameters TCGA never published (the Supplementary Methods
  say only *"identified from segmented data using GISTIC 2.0"* — no `-ta`/`-td`/`-conf`, no
  reference build). Here the peak list is TCGA's own, from Supplementary Data File S2.1, so
  the clustering feature involves no parameter guess and no GISTIC run.
* **No liftover.** TCGA's segments and TCGA's published peak boundaries are both hg19, so the
  peaks apply directly and the genome bins use the bundled hg19 arm table. Binning hg19
  coordinates against hg38 lengths would shift every bin and misplace every centromere.

What remains ours, and cannot be taken from the paper: the **distance and linkage**. The
Supplementary Methods name none for the copy-number clustering, so this report uses the
pipeline's configured pair, exactly as report 09 does.

## Part 1 — the two matrices

TCGA use **two different matrices** for this figure, and keeping them apart is what makes it
the paper's figure rather than a picture of a clustering input —
[00-methods](00-methods.html) sets out the rule. Both are built here from the same
`data/tcga_2013/data_cna_hg19.seg`.

**The clustering feature** is TCGA's stated one (Suppl. Methods S2): *"thresholded relative
copy number data in significantly reoccurring amplifications or deletions regions identified
by GISTIC 2.0."* `published_peak_matrix_from_seg()` takes, for each sample × published peak,
the length-weighted mean log2 seg-mean inside the peak's wide boundaries (a peak with no
overlapping segment becomes 0, copy-neutral), then **thresholds** it, because S2 is explicit
that the clustering ran on thresholded and not continuous values. The cut points are not
published, so they are configuration here and stated in the table below.

**The display feature** is what the Fig. 1a legend describes: *"SCNAs in each tumour
(horizontal axis) plotted by chromosomal location (vertical axis)"* — the whole genome, on a
continuous colour scale. `segments_to_bin_matrix()` tiles the same segments into 1 Mb bins
genome-wide using the hg19 arm table for chromosome lengths and order, taking the
length-weighted mean `Segment_Mean` per sample × bin, uncovered genome = 0.

```{r tcga-load}
# |x| < 0.1 -> 0 ; 0.1 <= |x| < 0.9 -> +-1 (low-level) ; |x| >= 0.9 -> +-2 (high-level).
# NOT TCGA's cut points, which were never published — hence stated on the page, not buried.
thr_cuts <- c(0.1, 0.9)

tcga_clin <- tryCatch(load_tcga_ucec_2013(),   error = function(e) tibble())
tcga_seg  <- tryCatch(load_tcga_2013_seg(),    error = function(e) tibble())
peaks     <- tryCatch(load_tcga_published_peaks(), error = function(e) tibble())
# hg19, NOT the pipeline default: these are SNP6 segments on hg19 coordinates.
arms19    <- tryCatch(load_arm_boundaries(build = "hg19"), error = function(e) NULL)

have_seg   <- nrow(tcga_seg) > 0
have_peaks <- nrow(peaks) > 0

# CLUSTERING matrix — thresholded, peak-restricted (S2's feature space).
clust_mat <- if (have_seg && have_peaks)
  published_peak_matrix_from_seg(tcga_seg, peaks, threshold = thr_cuts) else NULL
# DISPLAY matrix — continuous, genome-wide (the Fig-1a body).
disp_mat  <- if (have_seg && !is.null(arms19))
  segments_to_bin_matrix(tcga_seg, arms19) else NULL

# Drop bins on chromosomes this cohort has NO segments for at all.
# segments_to_bin_matrix() fills uncovered genome with 0 = copy-neutral, which is the right
# assumption for a gap inside a chromosome that WAS assayed and the wrong one for a whole
# chromosome that never was: the hg19 arm table carries chrY, this is a uterine cohort with
# zero chrY segments, and its all-zero chrY bins would render as a white "no alteration" block
# instead of "not applicable". Driven by the cohort's own coverage, not a chrY special case.
if (!is.null(disp_mat)) {
  covered <- unique(paste0("chr", str_remove(as.character(tcga_seg$Chromosome), "^chr")))
  bin_chr <- str_extract(setdiff(names(disp_mat), "ID"), "^chr[0-9XY]+")
  drop_b  <- setdiff(names(disp_mat), "ID")[!bin_chr %in% covered]
  if (length(drop_b)) {
    disp_mat <- disp_mat |> select(-all_of(drop_b))
    cat("Dropped", length(drop_b), "genome bins on chromosome(s) with no segments in this",
        "cohort:", paste(sort(unique(str_extract(drop_b, "^chr[0-9XY]+"))), collapse = ", "),
        "— all-zero bins there would read as 'no alteration' rather than 'not assayed'.\n")
  }
}

# cat(), not message(): opts_chunk sets message = FALSE, so a message() here is invisible on
# the knitted page and a skipped section then looks like nothing happened at all.
if (!have_seg)
  cat("**No TCGA segments** (data/tcga_2013/data_cna_hg19.seg) —",
      "run `Rscript code/fetch_tcga_ucec_2013.R`. Every panel below is skipped.\n")
if (have_seg && !have_peaks)
  cat("**No published peak list** (data/tcga_gistic_peaks_2013.tsv) —",
      "the clustering feature cannot be built.\n")
if (have_seg && is.null(arms19))
  cat("**No hg19 arm table** (data/arm_boundaries_hg19.tsv) —",
      "the genome-wide display matrix cannot be built; the heatmap will fall back.\n")
```

Both matrices side by side, so the difference between what is clustered and what is drawn is
a number on the page rather than a claim in prose. The row counts are the point: a few dozen
peaks against a few thousand genome bins.

```{r matrix-table}
if (!is.null(clust_mat) || !is.null(disp_mat)) {
  knitr::kable(tibble(
    role     = c("clustered", "displayed"),
    matrix   = c("published GISTIC peaks, THRESHOLDED",
                 "1 Mb genome bins, CONTINUOUS"),
    samples  = c(if (is.null(clust_mat)) NA_integer_ else nrow(clust_mat),
                 if (is.null(disp_mat))  NA_integer_ else nrow(disp_mat)),
    features = c(if (is.null(clust_mat)) NA_integer_ else ncol(clust_mat) - 1L,
                 if (is.null(disp_mat))  NA_integer_ else ncol(disp_mat)  - 1L),
    provenance = c(paste0("Suppl. Data File S2.1 peaks; cut at ",
                          paste(thr_cuts, collapse = "/")),
                   "data_cna_hg19.seg tiled on arm_boundaries_hg19.tsv"),
    caption_note = c("Suppl. Methods S2 feature space", "Fig. 1a legend")),
    caption = paste0("The two matrices. Only the first is clustered; only the second is ",
                     "drawn. Both come from the same hg19 segment file."))
} else cat("Matrix summary skipped — no segments.\n")
```

The cohort as TCGA classified it, which is also the annotation stack for the heatmap. Every
column here is **TCGA's own published field**, not a call of ours: the integrative subtype,
the MSI-derived MMR status, and the published copy-number cluster. Aneuploidy is the one
exception and is flagged as such — it is retrofitted from the PanCancer-Atlas table
(`pancan_aneuploidy.tsv`, a *count* of altered arms, 0–39), or from ASCETS run on these same
segments when `code/run_ascets_tcga.sh` has been run, which puts it on ATTEND's own 0–1
scale. Whichever is used is named in the table.

**TP53 is absent from this report's annotation stack**, unlike report 09's. `data/tcga_2013/`
carries clinical tables and segments only — no mutation calls — so there is nothing to build
a TP53 bar from, and an all-NA bar is dropped rather than drawn as an empty row.

```{r cohort-table}
tcga_ascets <- tryCatch(load_tcga_2013_ascets(), error = function(e) tibble())
use_ascets  <- nrow(tcga_ascets) > 0

aneu_tbl <- if (use_ascets) {
  tcga_ascets |> transmute(ID, aneuploidy_value = ascets_aneuploidy)
} else if (nrow(tcga_clin) > 0) {
  tcga_clin |> transmute(ID, aneuploidy_value = aneuploidy)
} else NULL
aneu_label <- if (use_ascets)
  "ASCETS fraction of evaluable arms (0-1) — same statistic as ATTEND" else
  "PanCancer-Atlas arm COUNT (0-39, Taylor 2018) — NOT ATTEND's scale"

if (nrow(tcga_clin) > 0) {
  knitr::kable(tibble(
    field = c("tumours with clinical data", "with a published CN cluster",
              "with an integrative subtype", "MMR-deficient (TCGA's MSI cluster)",
              "POLE-ultramutated", "with an aneuploidy value"),
    n = c(nrow(tcga_clin),
          sum(!is.na(tcga_clin$cn_cluster_k4)),
          sum(!is.na(tcga_clin$subtype_short)),
          sum(tcga_clin$mmrd %in% TRUE),
          sum(tcga_clin$pole %in% TRUE),
          if (is.null(aneu_tbl)) 0L else sum(is.finite(aneu_tbl$aneuploidy_value)))),
    caption = paste0("TCGA UCEC 2013 as published. Aneuploidy source: ", aneu_label,
                     ". A third of the cohort has no integrative subtype — it was assigned ",
                     "to the core set only — which is why the heatmap legends the missing ",
                     "category explicitly instead of leaving it an unlabelled grey."))
} else cat("Cohort table skipped — no clinical data.\n")
```

## Part 2 — the Fig-1a heatmap

The clustering runs on the **thresholded peak** matrix with the pipeline's configured cut,
distance and linkage, and the copy-number-high cluster is named from that matrix's own SCNA
burden (`cnv_high_cluster_burden()`, mean |value| per cluster) so it never sees the aneuploidy
score. The annotation frame is keyed by sample barcode, which is what `fig1a_heatmap()`
indexes its columns by.

```{r cluster}
cl <- if (!is.null(clust_mat) && nrow(clust_mat) >= k_cnv)
  cluster_arm_matrix(clust_mat, k = k_cnv, method = cnv_method, dist_method = cnv_dist) else NULL

high_cluster <- if (!is.null(cl)) cnv_high_cluster_burden(cl) else NA_integer_
ann <- NULL
if (!is.null(cl)) {
  ids <- rownames(cl$matrix)
  ann <- data.frame(CN_cluster = factor(cl$cluster[ids]), row.names = ids)
  if (nrow(tcga_clin) > 0) {
    grab <- function(col) tcga_clin[[col]][match(ids, tcga_clin$ID)]
    ann$TCGA_class <- factor(as.character(grab("subtype_short")),
                             levels = levels(tcga_clin$subtype_short))
    mm <- grab("mmrd")
    ann$MMR <- factor(ifelse(is.na(mm), NA_character_, ifelse(mm, "MMRd", "MMRp")),
                      levels = c("MMRp", "MMRd"))
    # TCGA's OWN partition as a bar, so agreement with the recomputed clusters beneath is
    # readable by eye. Same cluster colours as the bottom bar (see .fig1a_covariate_cols):
    # "recomputed 3" and "published 3" are the same colour on purpose.
    ann$Published_CN_cluster <- factor(as.character(grab("cn_cluster_k4")),
                                       levels = as.character(1:4))
  }
  if (!is.null(aneu_tbl))
    ann$Aneuploidy <- aneu_tbl$aneuploidy_value[match(ids, aneu_tbl$ID)]
  cat("Clustered", nrow(cl$matrix), "tumours on", ncol(cl$matrix),
      "published peaks;", cnv_dist, "distance +", cnv_method, "linkage, k =", k_cnv,
      "\nHighest-burden (copy-number-high) cluster:", high_cluster, "\n")
} else cat("Clustering skipped — need the thresholded peak matrix.\n")
```

The figure. Rows are the 1 Mb genome bins ordered by chromosomal location and split by
chromosome; columns are tumours, split by the peak-based clusters and ordered within each
split by dendrogram; the body is the continuous log2 landscape, red for gain and blue for
loss. The colour limit is the 98th percentile of |value|, so a handful of extreme
amplifications cannot wash the whole panel out. The annotation bars, top to bottom, are
TCGA's integrative subtype, its MMR status, its published copy-number cluster, and aneuploidy
as a binary split plus the continuous score. If the hg19 arm table is missing this falls back
to drawing the thresholded peak matrix and says so — that fallback is *not* Fig. 1a, and the
label on the panel states it rather than leaving the reader to infer it.

```{r hm, fig.height=8}
hm_mat <- NULL
if (!is.null(cl) && !is.null(disp_mat)) {
  dm <- as.matrix(disp_mat[, setdiff(names(disp_mat), "ID"), drop = FALSE])
  rownames(dm) <- disp_mat$ID
  dm <- dm[rownames(dm) %in% names(cl$cluster), , drop = FALSE]   # clustered tumours only
  if (nrow(dm) >= 2) hm_mat <- dm else
    cat("The bin matrix shares", nrow(dm), "samples with the clustering — falling back.\n")
}
if (!is.null(cl)) {
  if (!is.null(hm_mat)) {
    hm_vtype <- "continuous"
    hm_label <- "genome-wide continuous SCNA (1 Mb bins) — the Fig-1a display"
    feat_pos <- feature_positions(colnames(hm_mat))     # "chr1:0" bins carry their own coords
  } else {
    hm_mat   <- cl$matrix
    hm_vtype <- "thresholded"
    hm_label <- "FALLBACK: thresholded published peaks (no hg19 bin matrix) — not Fig. 1a"
    pk_cyto  <- if (have_peaks) peaks |> distinct(peak_id, cytoband) |> tibble::deframe() else NULL
    feat_pos <- feature_positions(colnames(hm_mat), cytoband = pk_cyto)
  }

  # WHY-IS-THERE-NO-HEATMAP gate. fig1a_heatmap() returns invisible(NULL) on several
  # conditions, each announced only with message() — which opts_chunk (message = FALSE)
  # swallows, so the figure just silently fails to appear. Whichever line reads FALSE / 0
  # below is the reason.
  cat("Fig-1a preconditions",
      "\n  displaying               :", hm_label,
      "\n  ComplexHeatmap installed :", requireNamespace("ComplexHeatmap", quietly = TRUE),
      "\n  circlize installed       :", requireNamespace("circlize", quietly = TRUE),
      "\n  display (samples x feat) :", nrow(hm_mat), "x", ncol(hm_mat),
      "\n  clustered on             :", nrow(cl$matrix), "x", ncol(cl$matrix), "peaks",
      "\n  feature positions mapped :", sum(feat_pos$feature %in% colnames(hm_mat)),
      "\n  annotation bars          :", if (is.null(ann)) 0L else ncol(ann), "\n")

  fig1a_heatmap(hm_mat, feat_pos, cl$cluster, ann_col = ann, value_type = hm_vtype,
                main = paste0("TCGA UCEC 2013 — Fig. 1a; columns split by clusters from ",
                              "TCGA's published peaks, cluster ", high_cluster,
                              " = copy-number high"))
} else cat("Heatmap skipped — no clustering.\n")
```

The same canvas with the cluster assignment at every k from 2 to 8 stacked beneath it. The
dendrogram is computed once and cut repeatedly, so the partitions are nested and can share
one figure: reading a column downward shows how that tumour's label subdivides as k grows.
Here it answers a question report 09 cannot — TCGA fixed k = 4 and published the result, so
the `Published_CN_cluster` bar above is the reference the sweep can be read against.

```{r hm-ksweep, fig.height=9}
if (!is.null(cl) && exists("hm_mat") && exists("feat_pos")) {
  # ksweep intersects rownames(mat) with hc$labels itself, so the genome-wide display matrix
  # and the peak-learned dendrogram line up with no extra handling.
  fig1a_heatmap_ksweep(hm_mat, feat_pos, cl$hclust, k_range = k_sweep, ann_col = ann,
                       value_type = hm_vtype,
                       main = paste0("TCGA UCEC 2013 — k = ", min(k_sweep), "..", max(k_sweep),
                                     " cluster bars on one Fig-1a canvas"))
} else cat("k-sweep skipped — no clustering.\n")
```

## Part 3 — does it recover TCGA's published partition?

The test the whole report exists for. Agreement is scored with the **adjusted Rand index**,
which is chance-corrected and so is not inflated by cluster-size imbalance, over the tumours
carrying a published assignment.

Two columns qualify the answer and belong beside it rather than after it. `burden_pick` is
the cluster `cnv_high_cluster_burden()` names as copy-number-high; `best_overlap` is the
cluster that actually overlaps published cluster 4 (TCGA's serous-like group) most. The
burden rule is a **heuristic** — greatest mean |value| — and on thresholded features mean
|value| counts a deep deletion and a high amplification identically, so a small
deletion-heavy cluster can outrank the genuinely serous-like one. **When the two disagree the
heuristic has failed**, and since report 09 uses that same rule on ATTEND, where no published
answer exists to catch it, this is the only place the failure is visible at all.

```{r ari}
scored <- NULL
if (!is.null(cl) && nrow(tcga_clin) > 0 && any(!is.na(tcga_clin$cn_cluster_k4))) {
  pub <- tcga_clin |> filter(!is.na(cn_cluster_k4)) |>
    transmute(ID, published = as.character(cn_cluster_k4))
  j <- cl$assignment |> inner_join(pub, by = "ID")
  if (nrow(j) > 0) {
    f1 <- function(tp, fp, fn) round(2 * tp / max(1, 2 * tp + fp + fn), 3)
    best <- j |> filter(published == "4") |> count(cnv_cluster) |>
      slice_max(n, n = 1, with_ties = FALSE)
    bo <- if (nrow(best)) as.character(best$cnv_cluster[1]) else NA_character_
    scored <- tibble(
      n_scored     = nrow(j),
      ARI          = round(adjusted_rand_index(j$cnv_cluster, j$published), 3),
      burden_pick  = as.character(high_cluster),
      burden_F1_vs_pub4 = f1(sum(j$cnv_cluster == high_cluster & j$published == "4"),
                             sum(j$cnv_cluster == high_cluster & j$published != "4"),
                             sum(j$cnv_cluster != high_cluster & j$published == "4")),
      best_overlap = bo,
      best_F1_vs_pub4   = if (is.na(bo)) NA_real_ else
        f1(sum(j$cnv_cluster == bo & j$published == "4"),
           sum(j$cnv_cluster == bo & j$published != "4"),
           sum(j$cnv_cluster != bo & j$published == "4")),
      heuristic_ok = !is.na(bo) && identical(as.character(high_cluster), bo))
    knitr::kable(scored, caption = paste0(
      "Recomputed clusters scored against TCGA's published CNA_CLUSTER_K4. ",
      "heuristic_ok = FALSE means the greatest-mean-|value| serous rule picked the wrong ",
      "cluster on this cohort, which is a warning about report 09's use of the same rule."))
  }
}
if (is.null(scored)) cat("Scoring skipped — needs the clustering and CNA_CLUSTER_K4.\n")
```

The contingency behind that index: recomputed clusters as rows, TCGA's published clusters as
columns. A concentrated off-diagonal block is the recovery the ARI summarises as one number;
mass spread across a row means that recomputed cluster has no published counterpart, which a
single index cannot show.

```{r contingency}
if (!is.null(scored)) {
  pub <- tcga_clin |> filter(!is.na(cn_cluster_k4)) |>
    transmute(ID, published = as.character(cn_cluster_k4))
  j <- cl$assignment |> inner_join(pub, by = "ID")
  knitr::kable(as.data.frame.matrix(table(recomputed = j$cnv_cluster, published = j$published)),
               caption = "Recomputed (rows) vs TCGA's published CNA_CLUSTER_K4 (columns).")
} else cat("Contingency skipped.\n")
```

## Part 4 — MMR and aneuploidy across the clusters

Aneuploidy per recomputed cluster. This is the panel the 2013 paper could not draw: it had no
per-sample aneuploidy score, so the relationship between its SCNA clusters and aneuploidy was
never shown. The value is whichever source Part 1 named, and it is **not** a clustering
feature — the clustering saw only thresholded peak calls — so alignment between the
high-burden cluster and the high-aneuploidy tumours is corroboration rather than circularity.
`attend_box()` picks the mark per cluster from the data, box and points where a cluster holds
at least ten tumours and a median crossbar below that, and prints n in every case.

```{r aneu-by-cluster, fig.height=4.5}
aneu_cl <- if (!is.null(cl) && !is.null(aneu_tbl)) {
  cl$assignment |>
    left_join(aneu_tbl, by = "ID") |>
    filter(is.finite(aneuploidy_value)) |>
    mutate(cluster = factor(cnv_cluster))
} else NULL

if (!is.null(aneu_cl) && nrow(aneu_cl) > 0 && dplyr::n_distinct(aneu_cl$cluster) > 1) {
  kw <- suppressWarnings(stats::kruskal.test(aneuploidy_value ~ cluster, data = aneu_cl))
  ggplot(aneu_cl, aes(cluster, aneuploidy_value)) +
    attend_box(aneu_cl, x = "cluster", y = "aneuploidy_value") +
    labs(title = "TCGA UCEC 2013 — aneuploidy by recomputed copy-number cluster",
         subtitle = paste0(aneu_label, "; Kruskal-Wallis p = ", signif(kw$p.value, 3),
                           "; cluster ", high_cluster, " = highest intrinsic SCNA burden"),
         x = "copy-number cluster", y = "aneuploidy") +
    attend_theme()
} else cat("Aneuploidy-by-cluster skipped — needs clusters and aneuploidy values.\n")
```

MMR status across the clusters, as counts and as a test. TCGA's MMR call here is its MSI
(hyper-mutated) integrative cluster, so this asks whether the *copy-number* clustering
separates MMR-deficient tumours — and the expected answer is that it largely does not, since
MMR deficiency and broad copy-number change are different routes to a genomic extreme. Fisher
is used rather than chi-square because a 4 × 2 table on this cohort has thin cells, and it is
simulated when an exact computation would be too large.

```{r mmr-by-cluster}
mmr_cl <- if (!is.null(cl) && nrow(tcga_clin) > 0) {
  cl$assignment |>
    left_join(tcga_clin |> transmute(ID, mmrd, pole, subtype_short), by = "ID") |>
    filter(!is.na(mmrd)) |>
    mutate(cluster = factor(cnv_cluster),
           MMR = factor(ifelse(mmrd, "MMRd", "MMRp"), levels = c("MMRp", "MMRd")))
} else NULL

if (!is.null(mmr_cl) && nrow(mmr_cl) > 0 && dplyr::n_distinct(mmr_cl$cluster) > 1) {
  tab <- table(cluster = mmr_cl$cluster, MMR = mmr_cl$MMR)
  ft  <- suppressWarnings(tryCatch(stats::fisher.test(tab, simulate.p.value = TRUE, B = 1e5),
                                   error = function(e) NULL))
  knitr::kable(as.data.frame.matrix(tab), caption = paste0(
    "MMR status (TCGA's MSI integrative cluster) by recomputed copy-number cluster",
    if (!is.null(ft)) paste0(". Fisher exact p = ", signif(ft$p.value, 3),
                             " (simulated, B = 1e5)") else "."))
} else cat("MMR-by-cluster skipped — needs clusters and MMR calls.\n")
```

The same cross-tabulation against TCGA's full integrative subtype, which is the direct
statement of what the copy-number clustering does and does not recover: the CN-high and
CN-low subtypes were *defined* from copy number and should separate, while the MSI and POLE
subtypes were defined from mutation and need not. Rows are the recomputed clusters, so
reading across a row shows what that cluster is made of. Tumours with no published subtype
(about a third of the cohort — it was assigned to the core set only) appear as their own
column rather than being dropped, so the denominators stay honest.

```{r subtype-by-cluster}
if (!is.null(mmr_cl) && nrow(mmr_cl) > 0) {
  st <- cl$assignment |>
    left_join(tcga_clin |> transmute(ID, subtype_short), by = "ID") |>
    mutate(cluster  = factor(cnv_cluster),
           subtype  = factor(ifelse(is.na(as.character(subtype_short)), "not classified",
                                    as.character(subtype_short)),
                             levels = c(levels(tcga_clin$subtype_short), "not classified")))
  knitr::kable(as.data.frame.matrix(table(recomputed = st$cluster, subtype = st$subtype)),
               caption = paste0("Recomputed copy-number clusters (rows) against TCGA's ",
                                "published integrative subtype (columns). CN-high and CN-low ",
                                "were defined from copy number and should separate; MSI and ",
                                "POLE were not and need not."))
} else cat("Subtype cross-tabulation skipped.\n")
```

```{r session-info}
sessionInfo()
```
