---
title: "ATTEND — 10. Data integration: building the master table"
author: "sceriff0"
date: "2026-08-11"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
library(here)
library(fs)
options(bitmapType = "cairo")

knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 8, fig.height = 5
)

source(here("code", "load_clinical.R"))
source(here("code", "load_phenotypes.R"))
source(here("code", "load_wes_results.R"))
source(here("code", "attend_harmonise.R"))   # ID spaces, crosswalks, collapse rule
source(here("code", "attend_classes.R"))     # pathogenic_by_patient(), TMB recompute
source(here("code", "attend_io.R"))          # write_intermediate() (Parquet/CSV)
```

> **Question.** Which measurement belongs to which patient, and what does the joined table look like?
> **Cohort.** Every patient reached by at least one modality; no restriction is applied here.
> **Reads.** The `code/load_*.R` loaders, straight from HPC storage, plus `data/attend_barcodes.csv`.
> **Out of scope.** What the join cost, and whether the joined values are sane -> [11_join_audit](11_join_audit.html).

ATTEND measures the same patients several different ways — clinical records, radiology
imaging, IHC, and WES-derived genomics (TMB, MSI, HRD, aneuploidy, and a per-variant MAF).
Each modality was recorded with a different kind of identifier, so this report puts them
onto one patient key and writes the result.

Genomics tables are keyed by sequencing **barcode**, imaging and IHC by **image** id,
clinical tables by **patient** id (`pid`). Two crosswalks, barcode to pid and image to pid,
translate everything onto `pid`. That logic lives in `code/attend_harmonise.R` — one
definition, so no two reports can drift apart.

Four packages are optional and every use is guarded, so the report knits without them:
`pointblank` for the post-join gate, `reclin2` for the fuzzy near-miss check, and `arrow`
for Parquet intermediates. Install with
`renv::install(c("pointblank","reclin2","arrow"))`.

## The raw inputs

Every table is read straight from its loader in `code/load_*.R`, which are the seam
between HPC storage (`/hpcnfs/scratch/P_DIMA_ATTEND`) and this pipeline. Nothing is
filtered or transformed at this point — these are the raw shapes as delivered.

```{r load}
clinical_data        <- load_clinical_data()         # ATTEND summary file (ClinData)
imaging_data         <- load_imaging_data()          # ATTEND summary file (Imaging)
ihc_data             <- load_ihc_data()              # FlowPath
gianlu_clinical_data <- load_gianlu_clinical_data()  # clinical_with_mutation_status.tsv
aneu_data            <- load_aneuploidy_scores()     # WES: as_ASCET_attend.csv
hrd_data             <- load_hrd_scores()            # WES: hrd_scores.csv
msi_data             <- load_msi_scores()            # WES: msi.20250831.csv
tmb_data             <- load_tmb_scores()            # WES: tmb.20250831.csv
maf_data             <- load_maf_data()              # WES: glob *.maf
conversion_data      <- read.csv(here("data", "attend_barcodes.csv"))
```

Row and column counts of each raw table as loaded above, before any ID translation.
The key column differs per table — clinical/MSI/TMB/MAF use `ID`, imaging and IHC use
`image_id`, HRD uses `Sample`, aneuploidy uses `sequencing_id`, the conversion table
uses `barcode` — and reconciling those is the whole job of this report.

```{r load-dims}
raw_inputs <- list(
  clinical = clinical_data, imaging = imaging_data, ihc = ihc_data,
  gianlu = gianlu_clinical_data, aneu = aneu_data, hrd = hrd_data,
  msi = msi_data, tmb = tmb_data, maf = maf_data, conversion = conversion_data
)

tibble(
  dataset = names(raw_inputs),
  rows    = map_int(raw_inputs, nrow),
  cols    = map_int(raw_inputs, ncol)
)
```

## Building the two crosswalks

The **barcode→pid** crosswalk is derived from `gianlu_clinical_data`, which is the only
table carrying both a sequencing barcode (`TUMOR_BARCODE`) and a patient id. The
**image→pid** crosswalk comes from `imaging_data`, which carries both `image_id` and a
patient id. Everything else is translated through one of these two.

```{r harmonise}
id_cfg <- make_id_cfg()

cw_barcode_pid <- build_barcode_pid(gianlu_clinical_data, id_cfg)
cw_image_pid   <- build_image_pid(imaging_data, id_cfg)

# recover = TRUE promotes an orphaned id when its normalised key (upper-cased,
# trimmed) maps to exactly ONE patient. Ambiguous keys are refused, never guessed —
# the ledger in Part 2 shows precisely who is reclaimed and who is refused.
pid_vector <- make_pid_vector(cw_barcode_pid, cw_image_pid, recover = TRUE)

tables <- list(
  tmb = tmb_data, msi = msi_data, maf = maf_data, hrd = hrd_data,
  aneu = aneu_data, gianlu = gianlu_clinical_data,
  imaging = imaging_data, ihc = ihc_data
)

sets <- build_sets(tables, id_cfg, pid_vector)   # patients reached, per dataset
```

Size of each crosswalk. These are the limiting reagent for the whole join: a genomic
sample can only reach a patient if its barcode appears in the barcode crosswalk, and an
IHC slide only if its image appears in the image crosswalk. Most of the unmapped
records counted in Part 2 are unmapped for this reason.

```{r crosswalk-coverage}
cat("barcode crosswalk:", nrow(cw_barcode_pid), "pairs,",
    n_distinct(cw_barcode_pid$pid), "patients\n")
cat("image   crosswalk:", nrow(cw_image_pid), "pairs,",
    n_distinct(cw_image_pid$pid), "patients\n")
```

### Is the derived crosswalk right?

The barcode→pid crosswalk above is *reverse-engineered* from the clinical table, but
`data/attend_barcodes.csv` is the **canonical** map. The two are compared here on two
questions: which barcodes each one knows about, and — more seriously — whether they ever
**disagree about which patient a barcode belongs to**. A disagreement is not a coverage
gap; it would attach one patient's genomics to another patient's clinical record, and
every downstream result would be quietly wrong. It must be zero.

```{r crosswalk-validate}
derived_barcodes <- unique(cw_barcode_pid$barcode)
canon_barcodes   <- unique(as.character(conversion_data$barcode))

cat("barcodes in derived crosswalk:", length(derived_barcodes), "\n")
cat("barcodes in canonical file   :", length(canon_barcodes), "\n")
cat("in derived but NOT canonical :", length(setdiff(derived_barcodes, canon_barcodes)), "\n")
cat("in canonical but NOT derived :", length(setdiff(canon_barcodes, derived_barcodes)), "\n")

# The canonical file's patient-id column is auto-detected: the real file carries
# barcode / project / enrolment_id / tss / internal_id / patient_status, and
# enrolment_id is the study subject id. Override this if the detection is wrong.
canon_pid_candidates <- c("enrolment_id", "enrollment_id", "internal_id",
                          "pid", "PID", "PATIENT_ID", "patient_id", "Patient_ID",
                          "patient", "Patient", "ATTEND_ID", "attend_id", "ID")
canon_pid_col <- setdiff(intersect(canon_pid_candidates, names(conversion_data)),
                         "barcode")[1]
cat("using canonical patient-id column:",
    if (is.na(canon_pid_col)) "(none found)" else canon_pid_col, "\n")

if (!is.na(canon_pid_col)) {
  canon_cw <- conversion_data |>
    transmute(barcode   = as.character(barcode),
              pid_canon = as.character(.data[[canon_pid_col]])) |>
    filter(!is.na(barcode), !is.na(pid_canon)) |> distinct()

  disagree <- cw_barcode_pid |>
    inner_join(canon_cw, by = "barcode") |>
    filter(pid != pid_canon)

  cat("barcodes where derived pid DISAGREES with canonical:", nrow(disagree),
      "  <- must be 0\n")
  disagree
} else {
  message("No patient-id column detected in attend_barcodes.csv — ",
          "set canon_pid_col manually to enable the agreement check.")
}
```

The comparison above is character-for-character, so a canonical barcode differing only
by case, whitespace, or a `-DNA` suffix is reported as missing when it is really the
same sample. This fuzzy-matches the unmatched canonical barcodes against the derived set
(Jaro–Winkler ≥ 0.9 but < 1, i.e. close but not identical) and lists the near-misses, so
formatting drift can be told apart from genuine absence. Size-capped and guarded.

```{r crosswalk-reclin2}
only_canon <- setdiff(canon_barcodes, derived_barcodes)

if (requireNamespace("reclin2", quietly = TRUE) &&
    length(only_canon) > 0 && length(only_canon) <= 2000 && length(derived_barcodes) > 0) {
  near <- tryCatch({
    a <- tibble(barcode = only_canon)
    b <- tibble(barcode = derived_barcodes)
    pairs <- reclin2::pair(a, b)
    reclin2::compare_pairs(pairs, on = "barcode",
                           default_comparator = reclin2::cmp_jarowinkler(0.9),
                           inplace = TRUE)
    pp <- as.data.frame(pairs)
    pp <- pp[!is.na(pp$barcode) & pp$barcode >= 0.9 & pp$barcode < 1, , drop = FALSE]
    if (nrow(pp))
      tibble(canonical_barcode = a$barcode[pp$.x],
             nearest_derived   = b$barcode[pp$.y],
             jaro_winkler      = round(pp$barcode, 3)) |>
        arrange(desc(jaro_winkler))
    else tibble()
  }, error = function(e) { message("reclin2 near-miss check skipped: ", conditionMessage(e)); tibble() })

  if (nrow(near)) {
    cat("Canonical barcodes with a fuzzy near-miss (JW >= 0.9, not exact):\n")
    print(near, n = 50)
  } else cat("No fuzzy near-misses — the exact loss above is real, not formatting drift.\n")
} else {
  message("reclin2 probabilistic check skipped (package missing, nothing unmatched, ",
          "or unmatched set too large).")
}
```

### Two duplication failure modes

These are opposite problems and only one of them is allowed.
**One ID → many patients** must be zero: a barcode pointing at two patients duplicates
rows on every join and corrupts counts.
**One patient → many IDs** is legitimate (a patient can have several samples or slides),
but it has to be quantified, because the collapse step below **averages those rows
together** and that hides within-patient variation.

```{r duplicates}
dup_barcode <- cw_barcode_pid |> count(barcode, name = "n_pid") |> filter(n_pid > 1)
dup_image   <- cw_image_pid   |> count(image,   name = "n_pid") |> filter(n_pid > 1)
cat("barcodes -> >1 patient:", nrow(dup_barcode), " <- must be 0\n")
cat("images   -> >1 patient:", nrow(dup_image),   " <- must be 0\n")

cw_barcode_pid |> semi_join(dup_barcode, by = "barcode") |> arrange(barcode)
cw_image_pid   |> semi_join(dup_image,   by = "image")   |> arrange(image)

multi_barcode <- cw_barcode_pid |> count(pid, name = "n_barcode") |> filter(n_barcode > 1)
multi_image   <- cw_image_pid   |> count(pid, name = "n_image")   |> filter(n_image > 1)
cat("\npatients with >1 barcode:", nrow(multi_barcode),
    "| patients with >1 image:", nrow(multi_image),
    " <- legitimate, but averaged by the collapse below\n")
```

## Collapsing to one row per patient

The master keeps exactly one row per patient. Where a patient contributed several rows,
numeric columns are **averaged** and non-numeric columns take the first non-missing
value (`num_or_first()` in `attend_harmonise.R`). That is harmless for a score measured
once and lossy for anything measured repeatedly, so the table below counts, per dataset,
how many patients were affected — that count is the size of the hidden heterogeneity.

```{r collapse-report}
collapse_report <- bind_rows(
  Map(function(df, cfg) {
    pid <- pid_vector(df, cfg); pid <- pid[!is.na(pid)]
    tibble(n_rows_mapped        = length(pid),
           n_patients           = dplyr::n_distinct(pid),
           n_multi_row_patients = sum(table(pid) > 1))
  }, tables, id_cfg[names(tables)]),
  .id = "dataset"
)
collapse_report
```

## Assembling the master table

Start from the **union** of every patient any dataset reached, `left_join` each collapsed
dataset onto it (so an absent modality becomes `NA` rather than dropping the patient),
prefix each dataset's columns with its name (`tmb__`, `aneu__`, …) so nothing collides,
and add an `in_<dataset>` flag per modality.

**The MAF is deliberately excluded from the averaging.** Mean-collapsing mutation rows is
meaningless — the average of a genomic position is not a genomic position. Instead
mutation status is derived per patient as booleans by `pathogenic_by_patient()`
(`attend_classes.R`), which is the same helper the downstream mutation reports call, so
they cannot drift from this table.

```{r master-build}
spine_name <- names(which.max(lengths(sets)))
cat("Largest single dataset:", spine_name, "—", length(sets[[spine_name]]), "patients\n")

master <- tibble(pid = Reduce(union, sets)) |>
  left_join(collapse_pid(gianlu_clinical_data, id_cfg$gianlu,  "gianlu",  pid_vector), by = "pid") |>
  left_join(collapse_pid(imaging_data,         id_cfg$imaging, "imaging", pid_vector), by = "pid") |>
  left_join(collapse_pid(ihc_data,             id_cfg$ihc,     "ihc",     pid_vector), by = "pid") |>
  left_join(collapse_pid(tmb_data,             id_cfg$tmb,     "tmb",     pid_vector), by = "pid") |>
  left_join(collapse_pid(msi_data,             id_cfg$msi,     "msi",     pid_vector), by = "pid") |>
  left_join(collapse_pid(hrd_data,             id_cfg$hrd,     "hrd",     pid_vector), by = "pid") |>
  left_join(collapse_pid(aneu_data,            id_cfg$aneu,    "aneu",    pid_vector), by = "pid")

# Mutation status from the MAF, as booleans (never averaged) — see note above.
maf_mut <- pathogenic_by_patient(maf_data, cw_barcode_pid)  # pid, TP53_pathogenic, panel_pathogenic
master  <- master |> left_join(maf_mut, by = "pid")

for (s in names(sets)) master[[paste0("in_", s)]] <- master$pid %in% sets[[s]]
```

### The two TMB definitions

```{r child='_explainers/tmb-two-definitions.Rmd'}
```

The recompute reads the gnomAD-annotated long MAF (`data/variant_annotations/`, loaded by
`load_maf_tmb()`), counts the variant classes listed in `attend_tmb_recompute`, and anchors
the result to the upstream `tmb__TMB_SCORE` scale so the definitions differ only by the
germline filter. `apply_nassar_recalibration()` then produces `tmb__TMB_nassar` on top of
the robust recompute. If the annotated MAFs are not synced, `load_maf_tmb()` returns empty
and both recomputed columns are simply absent from the master.

```{r master-tmb}
# Recompute robust/matched TMB from the gnomAD-annotated long MAF, anchored to the
# upstream tmb__TMB_SCORE scale (attend_classes.R §13). Knit-safe: if the annotated
# MAFs are not synced, load_maf_tmb() returns empty and these columns are skipped.
maf_tmb <- tryCatch(load_maf_tmb(), error = function(e) tibble())
if (nrow(maf_tmb) > 0) {
  anchor <- if (attend_cols$tmb %in% names(master))
    master |> transmute(pid, tmb_anchor = suppressWarnings(as.numeric(.data[[attend_cols$tmb]]))) else NULL
  tmb_vars <- tmb_variant_table(maf_tmb, cw_barcode_pid,
                                anc_by_pid = ancestry_by_pid(master),
                                anchor_tmb = anchor)
  if (!is.null(tmb_vars)) master <- master |> left_join(tmb_vars, by = "pid")
}

# The ancestry recalibration itself (attend_classes.R §14), applied on top of robust.
master <- apply_nassar_recalibration(master, ancestry_by_pid(master))

# Report which of the two plotted definitions actually made it onto the table.
present_defs <- tmb_defs_present(master)
cat("TMB definitions available downstream:\n")
print(tibble(definition = tmb_def_labels(present_defs), column = unname(present_defs)))
```

### Write the integration point

The master is written through `write_intermediate()`, which picks Parquet when `arrow` is
installed and CSV otherwise, so no report ever hard-codes an extension. This is the file
every report from 11 onward reads, and nothing downstream re-reads the raw loaders.

```{r master-write}
master_path <- write_intermediate(master, "attend_master_joined")  # Parquet if arrow, else CSV
cat("Wrote", basename(master_path), ":", nrow(master), "patients x", ncol(master), "columns\n")
```

`attend_master_joined` is the contract every later report builds on, so a broken join
should fail **here and loudly** rather than surfacing as a strange survival curve three
reports later. This asserts the structural invariants of the join: `pid` exists, is never
null, and is unique — one row per patient. Guarded and warn-level, so it never halts a
knit.

```{r pointblank-gate}
if (requireNamespace("pointblank", quietly = TRUE)) {
  pointblank::create_agent(master, label = "ATTEND master-join QC") |>
    pointblank::col_exists("pid") |>
    pointblank::col_vals_not_null("pid") |>
    pointblank::rows_distinct(columns = "pid") |>
    pointblank::interrogate()
} else {
  message("pointblank not installed — renv::install('pointblank') to enable the QC gate.")
}
```

Each raw table is additionally written to its own intermediate next to the master, so
every input is preserved individually and downstream reports that need raw rows (rather
than the collapsed master) can read them back through `read_intermediate()`.

```{r save-each}
datasets <- list(
  clinical_data        = clinical_data,
  imaging_data         = imaging_data,
  ihc_data             = ihc_data,
  gianlu_clinical_data = gianlu_clinical_data,
  aneu_data            = aneu_data,
  hrd_data             = hrd_data,
  msi_data             = msi_data,
  tmb_data             = tmb_data,
  maf_data             = maf_data,
  conversion_data      = conversion_data
)
# gnomAD-annotated long MAF, needed by the ancestry-aware TMB recompute.
if (exists("maf_tmb") && nrow(maf_tmb) > 0) datasets$maf_tmb <- maf_tmb

iwalk(datasets, ~ write_intermediate(.x, .y))
```


```{r session-info}
sessionInfo()
```
