---
title: "ATTEND — 11. MMRd aneuploidy: metastatic (ATTEND) vs primary (TCGA)"
author: "sceriff0"
date: "2026-08-17"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---
```{r setup, include=FALSE}
# renv: tidyverse, here. Optional and guarded: diptest (Hartigan's dip), survival +
# ggsurvfit (the TCGA KM). If missing:
# renv::install(c("diptest","survival","ggsurvfit")); renv::snapshot()
# The TCGA arm needs data/tcga_2013/ — Rscript code/fetch_tcga_ucec_2013.R — and the
# shared-scale panels additionally need ASCETS run on TCGA's segments: bash code/run_ascets_tcga.sh
library(tidyverse)
library(here)
# Headless-safe raster device. The cluster's compute nodes have no X11, so R's default
# png() fails with "unable to open connection to X11 display" and every figure dies. cairo
# needs no display; the guard keeps this working on a machine built without it.
if (capabilities("cairo")) options(bitmapType = "cairo")

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 5)

source(here("code", "attend_classes.R"))     # mmrd_aneuploidy_split(), dip_bimodality(),
                                             # aneu_cuts(), add_aneuploidy_cuts(),
                                             # attend_tcga_ref, attend_tcga_ref_2013, thresholds
source(here("code", "attend_plots.R"))       # attend_theme(), attend_box(), attend_cohort_cols
source(here("code", "attend_io.R"))          # read_intermediate()
source(here("code", "build_master.R"))       # get_master() — builds on demand
source(here("code", "load_wes_results.R"))   # load_tcga_ucec_2013(), load_tcga_2013_ascets(),
                                             # load_tcga_ucec() (legacy fallback)

# The configured aneuploidy cuts, as ONE ordered pair for the whole report. Every
# cut-dependent panel below facets on the label this pair generates rather than being
# written twice — see add_aneuploidy_cuts() for why that is a helper and not two call sites.
cuts <- aneu_cuts()
```
> **Question.** Does aneuploidy substructure MMR-deficient tumours in metastatic disease (ATTEND) in a way that is absent from primary disease (TCGA UCEC)?
> **Cohort.** MMR-deficient only, in both cohorts, side by side.
> **Reads.** The master (built on demand by get_master()) for the ATTEND arm; data/tcga_2013/ plus ASCETS run on its segments for the TCGA arm.
> **Out of scope.** The ATTEND MMRd split itself, against mutations and composition -> [04-aneuploidy-mmrd](04-aneuploidy-mmrd.html). ATTEND TMB bimodality and its own TCGA replication -> [06-tmb](06-tmb.html). Every ATTEND survival model -> [02-survival](02-survival.html). Cluster-level correspondence between the two cohorts -> [09-tcga2013-fig1a](09-tcga2013-fig1a.html). The aneuploidy classes, the two cuts and why both are reported -> [00-methods](00-methods.html).

ATTEND is a cohort of **metastatic** endometrial tumours; TCGA UCEC is **primary**.
[04-aneuploidy-mmrd](04-aneuploidy-mmrd.html) works inside ATTEND's MMR-deficient stratum,
where aneuploidy appears to carve the group in two. This report asks whether that is specific
to metastatic disease, on four measurements that fail in different ways, so no single one has
to carry the answer:

* **Shape** — is the aneuploidy distribution within MMRd bimodal in ATTEND and unimodal in
  TCGA? Bimodality is **scale-invariant**, so this works even when the two cohorts' scores
  are on different axes.
* **Level** — is MMRd aneuploidy *absolutely* higher in metastatic disease? This needs a
  shared axis and is skipped without one.
* **Prevalence** — what fraction of MMRd tumours cross the aneuploidy line in each cohort?
  Two integers per cohort and an exact test: no distributional assumption, no survival event,
  and the only arm whose precision improves as the cohorts grow.
* **Orthogonality** — does TMB track aneuploidy *within* an MMR stratum in TCGA? If it did,
  the aneuploidy split would be re-discovering the hypermutator phenotype rather than adding
  an axis.

**Every aneuploidy-keyed figure and table here is drawn at both cuts**,
`r paste(cuts, collapse = " and ")` — attend_thresholds$aneuploidy and $aneuploidy_alt,
returned as a pair by aneu_cuts(); [00-methods](00-methods.html) holds the justification and
is not restated. The cut is load-bearing for this report's central question, so a result that
survives only one of the two is a result about the threshold. Panels that plot the
*continuous* score carry no class and are therefore drawn once, with both cuts as reference
lines instead of as a facet.

The shared axis is the load-bearing piece. ATTEND's aneuploidy score is ASCETS (fraction of
evaluable arms altered, 0–1); the cBioPortal column is Taylor et al. 2018, a *count* of
altered arms (0–39). No constant converts one into the other, because ASCETS's denominator is
the per-sample number of evaluable arms. `code/run_ascets_tcga.sh` removes the obstacle by
running **ASCETS on TCGA's own segments**, producing the same statistic on both sides;
without that output every scale-dependent panel below is skipped rather than computed on
incomparable numbers.

Calling stringency is matched where it can be: ATTEND runs `ascets(min_boc = 0.5,
alteration_threshold = 0.7)` and leaves `threshold` and `noise` at their defaults — a fixed
±0.2 log2 cutoff with no LCR noise modelling — and `run_ascets_tcga.sh` uses those same
values. (ATTEND runs ASCETS per sample and the TCGA script runs it pooled; with a fixed
threshold that is equivalent, since the only cohort-level step in ASCETS is the
noise-threshold estimation, which is not in use.) **One difference cannot be removed here:**
ATTEND is tumour-only DRAGEN WES with no purity/ploidy correction, TCGA is purity-corrected
SNP6. Low-purity ATTEND tumours have compressed log2 ratios and will *under*-call arms, so an
ATTEND-higher result is conservative — but the assay difference also changes which arms clear
the coverage cut, so a *small* level difference is suggestive, not decisive.

## Part 1 — is the split present in each cohort?

The ATTEND arm is the MMR-deficient stratum of the master, taken through
`mmrd_aneuploidy_split()`, whose MMRd criterion is MMR-protein loss by **IHC** —
[00-methods](00-methods.html) explains why MSI-high is a correlate and not a criterion. Its
aneuploidy values are the ASCETS scores loaded from `data/ascets/`.

The TCGA arm prefers the **2013 publication cohort** (`data/tcga_2013/`, via
`load_tcga_ucec_2013()`) joined to ASCETS run on its segments (`load_tcga_2013_ascets()`),
which is what creates the shared scale. Its MMRd rule is `attend_tcga_ref_2013$mmrd_rule`,
whose default `"tcga_exact"` makes MMRd **the paper's own MSI (hyper-mutated) integrative
cluster and nothing else**: every other classified tumour is MMRp, and the tumours TCGA never
classified are NA — unclassified, not proficient. That is the only rule whose output can be
quoted as "TCGA's MSI group". If the ASCETS output is absent the report falls back to the
legacy PanCanAtlas export (`load_tcga_ucec()`, Taylor arm count) and drops every
scale-dependent panel, saying so in the table.

Both cohorts are classified by **one helper over one cut vector**: `add_aneuploidy_cuts()`
re-derives `aneuploidy_class` at each cut from `attend_cols$aneuploidy`, for the ATTEND rows
(`master_cuts`) and for the TCGA rows (`tcga_cuts_df`) alike. `same_cut` records whether the
two vectors are identical and gates every cross-cohort panel below, because ATTEND at
`r min(cuts)` drawn beside TCGA at `r max(cuts)` would be a comparison of two different
questions. The last table is the cost of the choice in ATTEND MMRd patients: raising a cut
can only demote, so the two classifications are nested and the movers score between them.

```{r cohorts}
master      <- get_master() |> add_molecular_classes()
# The same patients once per cut, with aneuploidy_class and the MMR x AS group re-derived at
# each. Every cut-dependent panel facets on as_cut_lab instead of being written twice.
master_cuts <- add_aneuploidy_cuts(master, scna = TRUE)
attend_mmrd <- mmrd_aneuploidy_split(master)
attend_aneu <- if (!is.null(attend_mmrd))
  attend_mmrd$aneu_value[is.finite(attend_mmrd$aneu_value)] else numeric(0)

t13 <- tryCatch(load_tcga_ucec_2013(),   error = function(e) tibble())
tas <- tryCatch(load_tcga_2013_ascets(), error = function(e) tibble())
use_ascets <- nrow(t13) > 0 && nrow(tas) > 0

if (use_ascets) {
  tcga <- t13 |>
    left_join(tas, by = "ID") |>
    distinct(pid, .keep_all = TRUE) |>          # survival is a per-PATIENT analysis
    transmute(pid, mmrd, pole, tmb, aneuploidy = ascets_aneuploidy, cn_cluster_k4,
              os_months, os_event, pfs_months, pfs_event, stage, grade)
  tcga_scale  <- "ASCETS (fraction of evaluable arms, 0-1) — SAME statistic as ATTEND"
  tcga_source <- "data/tcga_2013/ascets/ (code/run_ascets_tcga.sh)"
} else {
  tcga <- tryCatch(load_tcga_ucec(), error = function(e) tibble())
  if (nrow(tcga) > 0) tcga <- tcga |>
    mutate(pfs_months = NA_real_, pfs_event = NA_integer_,
           stage = NA_character_, grade = NA_character_, cn_cluster_k4 = NA_character_,
           pole = NA, tmb = suppressWarnings(as.numeric(tmb)))
  tcga_scale  <- paste0("cBioPortal ", attend_tcga_ref$aneuploidy_col,
                        " (Taylor 2018 arm COUNT, 0-39) — NOT ATTEND's scale")
  tcga_source <- "data/tcga/ (code/fetch_tcga_ucec.R)"
}

tcga_mmrd <- if (nrow(tcga) > 0) filter(tcga, mmrd) else tcga
tcga_aneu <- if (nrow(tcga_mmrd) > 0) tcga_mmrd$aneuploidy[is.finite(tcga_mmrd$aneuploidy)] else numeric(0)

lab_attend <- "ATTEND (metastatic)"
lab_tcga   <- "TCGA UCEC (primary)"
have_attend <- length(attend_aneu) >= 4
have_tcga   <- length(tcga_aneu)   >= 4
# Every LEVEL comparison below requires both cohorts on the ASCETS [0,1] scale. Shape
# comparisons do not, because bimodality is scale-invariant.
comparable  <- use_ascets && have_attend && have_tcga &&
               all(is.finite(range(attend_aneu))) && max(attend_aneu, na.rm = TRUE) <= 1

# ONE cut vector per cohort, and the cross-cohort panels run only when the two are
# IDENTICAL. ATTEND is always on the ASCETS [0,1] scale, so it always carries the configured
# pair. TCGA carries the SAME pair when ASCETS has been run on its segments; on the
# arm-COUNT fallback 0.1 and 0.2 are not its scale's numbers at all — a 0.1 line there sits
# against the axis and would call every tumour aneuploidy-high — so it falls back to a
# single within-cohort median, labelled with its own value. same_cut is the guard that makes
# "ATTEND at 0.1 beside TCGA at 0.2" impossible rather than merely unintended.
tcga_all_aneu <- if (nrow(tcga) > 0) tcga$aneuploidy[is.finite(tcga$aneuploidy)] else numeric(0)
tcga_cuts <- if (comparable) cuts else
  if (length(tcga_all_aneu) > 0) stats::median(tcga_all_aneu) else numeric(0)
same_cut  <- comparable && identical(as.numeric(tcga_cuts), as.numeric(cuts))

# ⚠️ TWO fallback medians, on purpose. Part 4 compares MMRd against MMRp, so its fallback cut
# is the WHOLE cohort's median. Part 5 asks whether aneuploidy substructures MMRd — a
# WITHIN-stratum question — so its fallback is the MMRd subset's own median. They are the
# same number only by accident: MMRd tumours are quieter than UCEC as a whole (which carries
# the serous, copy-number-high arm), so splitting MMRd at the cohort median can leave an
# "aneuploidy-high" arm of two patients and answer the substructure question with an empty
# cell. When ASCETS has been run on TCGA both collapse to the configured pair and this is
# moot. Each part labels the cut it used.
#
# The TCGA frame once per cut goes through the SAME helper on the SAME column name the ATTEND
# rows go through: one derivation, so the two halves cannot drift to different rules.
tcga_cuts_df <- if (length(tcga_cuts) > 0 && nrow(tcga) > 0) {
  tcga |>
    mutate(!!attend_cols$aneuploidy := aneuploidy) |>
    add_aneuploidy_cuts(cuts = tcga_cuts) |>
    tibble::as_tibble()
} else tibble()

# Part 5's frame: MMRd only, cut at MMRd's own median on the fallback scale.
tcga_mmrd_cuts <- if (comparable) cuts else
  if (length(tcga_aneu) > 0) stats::median(tcga_aneu) else numeric(0)
tcga_mmrd_cuts_df <- if (length(tcga_mmrd_cuts) > 0 && nrow(tcga_mmrd) > 0) {
  tcga_mmrd |>
    mutate(!!attend_cols$aneuploidy := aneuploidy) |>
    add_aneuploidy_cuts(cuts = tcga_mmrd_cuts) |>
    tibble::as_tibble()
} else tibble()

knitr::kable(tibble(
  cohort  = c(lab_attend, lab_tcga),
  MMRd_n  = c(length(attend_aneu), length(tcga_aneu)),
  scale   = c("ASCETS (fraction of evaluable arms, 0-1)", tcga_scale),
  source  = c("data/ascets/ (DRAGEN .seg)", tcga_source)),
  caption = "MMR-deficient tumours per cohort, and the aneuploidy statistic each uses.")

cat("\nAneuploidy cuts — ATTEND: ", paste(cuts, collapse = ", "),
    " | TCGA: ", if (length(tcga_cuts) > 0) paste(signif(tcga_cuts, 3), collapse = ", ") else "none",
    if (same_cut) " — IDENTICAL, so the cross-cohort panels run.\n" else
      " — NOT identical, so every cross-cohort class panel is skipped.\n", sep = "")

# The ATTEND MMRd arm at each cut, from master_cuts. mmrd_aneuploidy_split() above and
# add_aneuploidy_cuts() here select MMRd through the same is_mmrd() IHC criterion and read
# the same attend_cols$aneuploidy column, so these are the same patients as attend_aneu,
# classified at the fixed cut rather than at the split's optionally GMM-refined boundary.
attend_by_cut <- master_cuts |>
  filter(is_mmrd(MMR_class) %in% TRUE, !is.na(aneuploidy_class)) |>
  count(as_cut_lab, class = as_label(aneuploidy_class)) |>
  tidyr::pivot_wider(names_from = class, values_from = n, values_fill = 0)
knitr::kable(attend_by_cut, caption = paste0(
  "ATTEND MMRd tumours per aneuploidy class at each cut. Raising the cut can only demote, ",
  "so the two classifications are nested and the difference between the AS High counts is ",
  "the number of patients the choice of threshold moves."))

# cat(), not message() — see note_skip() in attend_plots.R for why (message = FALSE).
if (!have_tcga)
  cat("\n**No TCGA MMRd data** — run `Rscript code/fetch_tcga_ucec_2013.R`, then re-knit.\n")
if (nrow(t13) > 0 && !use_ascets)
  cat("\n**TCGA 2013 clinical present but no ASCETS output** — run `bash code/run_ascets_tcga.sh`",
      "to unlock the shared-scale panels (Parts 2 and 3).\n")
```

Aneuploidy density within MMRd, one panel per cohort, each annotated with **Hartigan's
dip-test** p from `dip_bimodality()` — a small p means the distribution is multimodal, i.e. a
real split rather than a continuum. The panels are **free-scaled on purpose**: this figure
compares shape only, and putting two different statistics on one axis would invite exactly
the level comparison that Part 2 gates behind the shared scale. TCGA is drawn in the
reference grey and only ATTEND carries a hue, because TCGA is the external cohort the ATTEND
result is measured against. The dip test needs at least four finite values and the optional
`diptest` package; where either is missing the annotation says so instead of printing a
number. The score is plotted **raw**, so no class and no cut enters this panel: both cuts
appear only as dashed reference lines, and only on a panel whose axis is the ASCETS [0,1]
scale — on the arm-count fallback the TCGA panel carries no line, because
`r paste(cuts, collapse = " and ")` are not numbers on that axis.

```{r density-contrast, fig.height=5}
dip_a <- dip_bimodality(attend_aneu)
dip_t <- dip_bimodality(tcga_aneu)

# Braces required: a function body of the bare form `if (x) a` with `else` on the NEXT line
# is an R syntax error, the same trap the survival-cut label below carries.
dip_lab <- function(d) {
  if (is.finite(d$p)) {
    sprintf("dip p = %.3g%s", d$p,
            ifelse(isTRUE(d$bimodal), " (multimodal)", " (unimodal)"))
  } else "diptest unavailable"
}

dens_df <- bind_rows(
  if (have_attend) tibble(cohort = lab_attend, aneu = attend_aneu) else NULL,
  if (have_tcga)   tibble(cohort = lab_tcga,   aneu = tcga_aneu)   else NULL)

if (nrow(dens_df) > 0) {
  labs_df <- tibble(cohort = c(lab_attend, lab_tcga),
                    lab    = c(dip_lab(dip_a), dip_lab(dip_t))) |>
    filter(cohort %in% dens_df$cohort)

  # Reference lines per PANEL, not per figure: ATTEND is always ASCETS, TCGA only when
  # run_ascets_tcga.sh has been run. Drawing 0.1 on an arm-count axis would put the line at
  # the origin and read as "every TCGA tumour is aneuploidy-high".
  cut_df <- bind_rows(
    if (have_attend)               tibble(cohort = lab_attend, xc = cuts) else NULL,
    if (have_tcga && use_ascets)   tibble(cohort = lab_tcga,   xc = cuts) else NULL)

  p <- ggplot(dens_df, aes(aneu)) +
    geom_density(aes(fill = cohort), colour = "grey30", alpha = 0.5, adjust = 0.9) +
    geom_rug(alpha = 0.4) +
    geom_text(data = labs_df, aes(x = Inf, y = Inf, label = lab),
              hjust = 1.05, vjust = 1.5, size = 3.3) +
    facet_wrap(~ cohort, scales = "free") +
    scale_fill_manual(values = attend_cohort_cols, guide = "none") +
    labs(title = "Aneuploidy within MMRd — metastatic vs primary",
         subtitle = paste0("free axes: shape only. Multimodal = a real split; unimodal = a ",
                           "continuum. Dashed = the two AS cuts, ",
                           paste(cuts, collapse = " and "), " (ASCETS panels only)"),
         x = "aneuploidy score (cohort-specific scale)", y = "density") +
    attend_theme()
  if (nrow(cut_df) > 0)
    p <- p + geom_vline(data = cut_df, aes(xintercept = xc), linetype = "dashed",
                        colour = "grey40", linewidth = 0.4)
  print(p)
} else cat("No aneuploidy data for either cohort yet.\n")
```

The same dip test as a table, so the statistic and n are legible rather than read off a
panel annotation. Read `n` first: with a handful of tumours per cohort the dip test has very
little power, so a non-significant p is weak evidence of unimodality and not evidence of it.
One row per cohort and not per cut: the dip statistic is computed on the raw score vector and
never sees a class, so both cuts would return the identical number.

```{r bimodality-table}
knitr::kable(tibble(
  cohort   = c(lab_attend, lab_tcga),
  n        = c(dip_a$n, dip_t$n),
  dip_stat = round(c(dip_a$dip, dip_t$dip), 4),
  dip_p    = signif(c(dip_a$p, dip_t$p), 3),
  multimodal = c(dip_a$bimodal, dip_t$bimodal)),
  caption = paste0("Hartigan dip test per cohort, on the MMRd aneuploidy values. ",
                   "Scale-invariant, so it is valid even on the arm-count fallback."))
```

## Part 2 — is MMRd aneuploidy absolutely higher in metastatic disease?

Part 1 compared shape; this compares **level**, which the shared ASCETS scale makes
legitimate. Both axes are now "fraction of evaluable chromosome arms altered", so the two
densities can be drawn on one axis and the medians compared directly. The dashed lines are
the two aneuploidy-high cutoffs, `r paste(cuts, collapse = " and ")`, drawn in the
aneuploidy-high colour because that is what they mark — and they now mean the same thing in
both cohorts. The figure is **not** faceted by cut: it plots the raw score with no class, so
the cut changes nothing in it except where the lines fall, which is exactly the comparison a
reader needs in order to judge either cut.

The test is a **Mann–Whitney** (rank-based, no distributional assumption) on the two MMRd
aneuploidy vectors, reported with the rank-biserial effect size, because a p-value alone is
uninformative when the group sizes differ by an order of magnitude. It is one test and not
one per cut, for the same reason: it ranks the scores themselves and never thresholds them. The whole part is
**skipped** unless ASCETS has been run on TCGA: on mismatched scales the comparison is
meaningless rather than merely imprecise. Interpret the size of any shift against the
purity/platform asymmetry stated at the top — the two uncorrectable biases run in opposite
directions, so a small shift is not interpretable in either direction.

```{r absolute-level, fig.height=4.5}
if (comparable) {
  both <- bind_rows(tibble(cohort = lab_attend, aneu = attend_aneu),
                    tibble(cohort = lab_tcga,   aneu = tcga_aneu))
  w  <- suppressWarnings(stats::wilcox.test(attend_aneu, tcga_aneu))
  # rank-biserial r = 1 - 2U/(n1*n2); negated below so positive = ATTEND higher.
  U  <- unname(w$statistic); rbc <- 1 - 2 * U / (length(attend_aneu) * length(tcga_aneu))
  md <- both |> group_by(cohort) |>
    summarise(n = n(), median = round(median(aneu), 3),
              IQR = paste0(round(quantile(aneu, .25), 3), "-", round(quantile(aneu, .75), 3)),
              .groups = "drop")
  knitr::kable(md, caption = paste0(
    "MMRd aneuploidy on the SHARED ASCETS scale. Mann-Whitney p = ", signif(w$p.value, 3),
    "; rank-biserial r = ", round(-rbc, 3), " (positive = ATTEND higher)."))

  ggplot(both, aes(aneu, fill = cohort)) +
    geom_density(colour = "grey30", alpha = 0.45, adjust = 0.9) +
    # BOTH cuts, because this panel carries no class: the lines are the only thing the
    # choice of threshold changes here.
    geom_vline(xintercept = cuts, linetype = "dashed",
               colour = attend_aneu_high, linewidth = 0.4) +
    scale_fill_manual(values = attend_cohort_cols, name = attend_cohort_legend) +
    labs(title = "MMRd aneuploidy, metastatic vs primary — one shared axis",
         subtitle = paste0("ASCETS fraction of evaluable arms altered; dashed = the two ",
                           "aneuploidy-high cutoffs, ", paste(cuts, collapse = " and ")),
         x = "ASCETS aneuploidy score (0-1)", y = "density") +
    attend_theme()
} else if (!use_ascets) {
  cat("Level comparison skipped — needs ASCETS on TCGA. Run `bash code/run_ascets_tcga.sh`.\n")
} else cat("Level comparison skipped — ATTEND aneuploidy is not on the [0,1] ASCETS scale.\n")
```

## Part 3 — how many MMRd tumours are aneuploidy-high in each cohort?

Part 1 asks whether a split exists and Part 2 whether the whole distribution shifts. This
asks the simplest question of the three: **what fraction of MMRd tumours cross the
aneuploidy line, in each cohort?** A 2×2 and a Fisher exact test, with exact
Clopper–Pearson intervals on each proportion. It needs no distributional assumption and no
survival event, which is why the intervals are the part to read: with a small
aneuploidy-high count on either side the interval stays wide even when the point estimates
differ substantially, and a wide interval is the honest summary.

This is the arm the cut moves most, so it is computed at **both** cuts,
`r paste(cuts, collapse = " and ")`, one block of rows each, with its **own** Fisher test per
cut — the two cuts are never pooled into one test, because they are two different questions
about the same patients. Inside each block a single `k` is applied to both cohorts in the
same expression, which is what makes "ATTEND at one cut, TCGA at another" impossible rather
than merely unintended; the whole table is additionally gated on `same_cut`. The arm is
skipped on the arm-count fallback because neither cut is transferable there. The **fixed**
cutoffs are used deliberately, not the GMM-refined boundary `mmrd_aneuploidy_split()` can
compute: that boundary is fitted to ATTEND's own distribution, so applying it to TCGA would
beg the question this part asks.

```{r prop-shared}
prop_ci <- function(x, n) {                       # Clopper-Pearson exact interval
  if (n == 0) return("n/a")
  ci <- stats::binom.test(x, n)$conf.int
  sprintf("%.1f%% (%.1f-%.1f)", 100 * x / n, 100 * ci[1], 100 * ci[2])
}

if (same_cut) {
  # ONE k per block, applied to both cohorts in the SAME expression and with the same
  # `>= k` rule add_molecular_classes() uses, so the two halves of a row cannot be counted
  # at different thresholds. The test is inside the block: cuts are never pooled.
  prop_at <- function(k) {
    a_hi <- sum(attend_aneu >= k); a_n <- length(attend_aneu)
    t_hi <- sum(tcga_aneu   >= k); t_n <- length(tcga_aneu)
    ft   <- stats::fisher.test(matrix(c(a_hi, a_n - a_hi, t_hi, t_n - t_hi), nrow = 2))
    tibble(cut             = as.character(as_cut_label(k, cuts)),
           cohort          = c(lab_attend, lab_tcga),
           MMRd_n          = c(a_n, t_n),
           aneuploidy_high = c(a_hi, t_hi),
           percent         = c(prop_ci(a_hi, a_n), prop_ci(t_hi, t_n)),
           fisher_p        = signif(ft$p.value, 3),
           odds_ratio      = sprintf("%.2f (%.2f-%.2f)", ft$estimate,
                                     ft$conf.int[1], ft$conf.int[2]))
  }
  knitr::kable(bind_rows(lapply(cuts, prop_at)), caption = paste0(
    "MMRd tumours at or above each shared ASCETS cutoff (>= ",
    paste(cuts, collapse = ", "), "). Fisher exact and the odds ratio are computed ",
    "SEPARATELY within each cut and repeat down its two rows; percentages carry exact ",
    "Clopper-Pearson intervals. ATTEND is the first row of each pair, so the odds ratio ",
    "reads ATTEND against TCGA at that cut."))
} else cat("Skipped — needs ASCETS on TCGA so both cohorts share the same cutoffs.",
           "Run `bash code/run_ascets_tcga.sh`.\n")
```

A second reference point taken **entirely from TCGA's published data**: how many of TCGA's
MMRd tumours were assigned to copy-number **cluster 4**, its serous-like SCNA cluster? Both
fields are TCGA's own (integrative cluster and `CNA_CLUSTER_K4`, Suppl. Data File S1.1), so
this proportion involves no cutoff, no ASCETS and no choice of ours — and so is one table,
not one per cut. It is **not the same criterion** as the table above — cluster membership rather than an aneuploidy threshold — so
the two are not interchangeable and only the shared-cutoff table supports a like-for-like
test. This is a sanity anchor: if either shared cutoff puts a wildly different fraction of
TCGA MMRd above the line than TCGA's own clustering did, that cutoff is drawing a different
line than the clustering, and that needs explaining before either number is quoted. The
cross-tabulation is a derived observation — the paper reports cluster 4's histology
composition but never its subtype composition — so it is not quotable from the publication.

```{r prop-published}
if (have_tcga && "cn_cluster_k4" %in% names(tcga_mmrd) && any(!is.na(tcga_mmrd$cn_cluster_k4))) {
  d  <- tcga_mmrd |> filter(!is.na(cn_cluster_k4))
  n4 <- sum(as.character(d$cn_cluster_k4) == "4")
  knitr::kable(tibble(
    definition  = "TCGA MMRd in copy-number cluster 4 (serous-like)",
    MMRd_n      = nrow(d),
    in_cluster4 = n4,
    percent     = prop_ci(n4, nrow(d))),
    caption = "TCGA's own published assignment — no threshold, no ASCETS, no choice of ours.")
} else cat("Published-anchor table skipped — needs the 2013 cohort with CNA_CLUSTER_K4.\n")
```

## Part 4 — does TMB track aneuploidy within an MMR stratum?

[06-tmb](06-tmb.html) Part 2 asks, in ATTEND: within each MMR stratum, does TMB differ
between aneuploidy-high and aneuploidy-low tumours? Splitting by MMR **first** matters
because the two routes to a genomic extreme are distinct — MMR-deficient tumours are
hypermutators (high TMB, often modest aneuploidy), whereas the CN-high / serous-like route
drives high aneuploidy at *low* TMB. If aneuploidy-high and TMB-high were the same tumours,
aneuploidy would add nothing to TMB; the interest is in their being different axes. This part
puts the same question to TCGA.

TMB here is TCGA's `TMB_NONSYNONYMOUS` from `data_clinical_sample.txt`, available on the
exome-sequenced subset, split at `attend_thresholds$tmb` = `r attend_thresholds$tmb` mut/Mb.
It is **not** either of ATTEND's two reported definitions and is never plotted alongside them
— it is a different cohort's single number, and [00-methods](00-methods.html) fixes what the
two ATTEND definitions are. It also needs no ancestry correction: ATTEND recomputes
ancestry-aware TMB because its TMB is *tumour-only*, so residual private germline variants
leak into the count in an ancestry-structured way, whereas TCGA's value comes from
tumour/normal **paired** exomes where the matched normal already removes germline variants
per patient. Applying a Nassar-style correction to it would subtract a bias that has already
been removed.

**POLE is excluded from MMRp, and that is not optional.** POLE-ultramutated tumours have a
median TMB two orders of magnitude above the copy-number groups, so leaving them in MMRp
would make that panel unreadable and misrepresent MMRp as high-TMB. ATTEND has no POLE cases,
so excluding them is also what makes the two cohorts' MMRp groups comparable; the POLE count
and median are printed separately for reference. The aneuploidy class is taken from
`tcga_cuts_df`, i.e. re-derived at each cut by `add_aneuploidy_cuts()` rather than
thresholded here, so the counts below are a partition per cut and the panels that follow can
facet on it. Where ASCETS has not been run on TCGA the cut vector collapses to that cohort's
own median and the page says so, because `r paste(cuts, collapse = " and ")` are ASCETS
[0,1] numbers with no meaning on an arm count.

```{r tcga-tmb-build}
have_tcga_tmb <- nrow(tcga) > 0 && "tmb" %in% names(tcga) && any(is.finite(tcga$tmb))

# aneuploidy_class comes from tcga_cuts_df — already re-derived once per cut by the same
# helper the ATTEND rows go through — so nothing here re-implements the threshold rule.
tcga_tmb <- if (have_tcga_tmb && nrow(tcga_cuts_df) > 0) {
  tcga_cuts_df |>
    mutate(mmr_group = dplyr::case_when(!is.na(pole) & pole ~ "POLE",
                                        mmrd %in% TRUE      ~ "MMRd",
                                        mmrd %in% FALSE     ~ "MMRp",
                                        TRUE                ~ NA_character_)) |>
    filter(is.finite(tmb), !is.na(aneuploidy_class), mmr_group %in% c("MMRp", "MMRd")) |>
    mutate(mmr_group = factor(mmr_group, levels = c("MMRp", "MMRd")))
} else tibble()

if (nrow(tcga_tmb) > 0) {
  pole_n   <- sum(!is.na(tcga$pole) & tcga$pole)
  pole_med <- suppressWarnings(stats::median(tcga$tmb[!is.na(tcga$pole) & tcga$pole], na.rm = TRUE))
  if (!same_cut)
    note_skip("The configured cuts (", paste(cuts, collapse = " and "), ") are ASCETS [0,1] ",
              "numbers and do not transfer to TCGA's arm COUNT, so the panels below are ",
              "drawn at ONE cut — TCGA's own within-cohort median, ",
              signif(tcga_cuts, 3), ". Run `bash code/run_ascets_tcga.sh` for both.")
  cat("patients per cut x MMR group x aneuploidy status:\n")
  print(tcga_tmb |>
          count(as_cut_lab, mmr_group, class = as_label(aneuploidy_class)) |>
          tidyr::pivot_wider(names_from = class, values_from = n, values_fill = 0))
  cat("\nPOLE (excluded from the panels): n = ", pole_n,
      ", median TMB = ", round(pole_med, 1), " mut/Mb\n", sep = "")
} else cat("TMB part skipped — needs the 2013 cohort with TMB_NONSYNONYMOUS.\n")
```

TCGA TMB in aneuploidy-high versus aneuploidy-low tumours, one column per MMR group and one
**row per aneuploidy cut**, boxes filled by aneuploidy status and the dashed line at the
TMB-high cut. `attend_box()` picks the mark per facet cell from the data — box and points
where the cell has at least ten patients, a median crossbar and points below that — so `by`
carries the cut as well as the MMR group, or a cell's mark would be chosen from both cuts'
patients at once. The TCGA MMRd aneuploidy-high cell is small and quartiles drawn from a
handful of values would be an artefact of which two landed in the middle. The layout is
`facet_wrap()` over both variables rather than `facet_grid()` because each panel needs its
**own** y axis: MMRd TMB sits an order of magnitude above MMRp, and `facet_grid()` frees y
by row only, which would put the two MMR groups back on one scale.

```{r tcga-tmb-box, fig.width=9, fig.height=8}
if (nrow(tcga_tmb) > 0) {
  # fill rides the TOP-LEVEL aes and attend_box(fill = NULL) inherits it — the idiom every
  # other semantic-fill box in the pipeline uses. attend_box() applies outlier.shape = NA
  # itself (it always draws the points, so an outlier layer would double-plot patients);
  # it takes no geom_boxplot passthrough, so naming one here is an error, not a no-op.
  ggplot(tcga_tmb, aes(aneuploidy_class, tmb, fill = aneuploidy_class)) +
    attend_box(tcga_tmb, x = "aneuploidy_class", y = "tmb",
               by = c("mmr_group", "as_cut_lab"), fill = NULL) +
    geom_hline(yintercept = attend_thresholds$tmb, linetype = "dashed",
               colour = attend_aneu_high, linewidth = 0.4) +
    scale_fill_manual(values = attend_aneu_cols, labels = as_label, name = attend_as_legend) +
    scale_x_discrete(labels = as_label) +
    facet_wrap(~ as_cut_lab + mmr_group, scales = "free_y",
               nrow = dplyr::n_distinct(tcga_tmb$as_cut_lab)) +
    labs(title = "TCGA UCEC — TMB by aneuploidy status, within MMR strata, at each cut",
         subtitle = paste0("dashed = TMB-high cut (", attend_thresholds$tmb, " mut/Mb); ",
                           "aneuploidy-high = score >= the row's cut (",
                           paste(signif(tcga_cuts, 3), collapse = ", "), "); POLE excluded"),
         x = NULL, y = "TMB (mut/Mb, nonsynonymous)") +
    attend_theme()
} else cat("No TCGA TMB/aneuploidy data to plot.\n")
```

The same contrast as counts: the TMB-high fraction on each side of the aneuploidy line,
within each MMR group **and within each cut**, with a Fisher exact test per cell of that
grid — the cut is a grouping key, never pooled, since pooling would count the same patient
twice under two classifications. The test is withheld — reported as NA — wherever an
aneuploidy group holds fewer than three patients, because a 2×2 with a cell of one or two
produces a p-value that moves entirely with a single reclassification; the higher cut demotes
patients into the low arm, so a row can lose its test even where the lower cut kept one.

```{r tcga-tmb-test}
if (nrow(tcga_tmb) > 0) {
  tab <- tcga_tmb |>
    group_by(as_cut_lab, mmr_group) |>
    group_modify(function(.x, ...) {
      hi    <- .x$tmb >= attend_thresholds$tmb
      is_lo <- .x$aneuploidy_class == "aneuploidy-low"
      n_lo <- sum(is_lo); n_hi <- sum(!is_lo)
      hilo <- sum(is_lo & hi); hihi <- sum(!is_lo & hi)
      p <- if (n_lo >= 3 && n_hi >= 3)
        suppressWarnings(stats::fisher.test(
          matrix(c(hilo, n_lo - hilo, hihi, n_hi - hihi), 2))$p.value) else NA_real_
      tibble(aneu_low_n = n_lo, TMBhi_in_low = hilo,
             aneu_high_n = n_hi, TMBhi_in_high = hihi,
             fisher_p = if (is.na(p)) NA_character_ else signif(p, 3))
    })
  knitr::kable(tab, caption = paste0(
    "TMB-high (>= ", attend_thresholds$tmb, " mut/Mb) fraction by aneuploidy status, within ",
    "each MMR group at each cut. Fisher exact per cut x MMR group, never pooled across ",
    "cuts; NA when an aneuploidy group has < 3 patients."))
} else cat("TMB 2x2 skipped.\n")
```

## Part 5 — the split in TCGA MMRd survival, and why it cannot decide anything

Parts 1–3 ask whether a split exists; this asks whether it carries outcome signal in primary
disease. **Read the event count before the p-value, and treat this part as descriptive
only.** TCGA UCEC is an early-stage, long-follow-up cohort with very few events among MMRd
patients — of the order of a dozen deaths under the `"tcga_exact"` rule. At that event count
the test cannot detect even a large hazard ratio, so a non-significant result here is **not**
evidence that the split is absent in primary disease, and a significant one is fragile,
threshold-dependent noise. The arms that can discriminate are shape, level and prevalence.

The cuts are ATTEND's own cutpoints when the scales match and, otherwise, a single median
taken **within the MMRd subset itself** — this part asks whether aneuploidy substructures
MMRd, so the fallback line has to fall inside that stratum; the cohort-wide median used in
Part 4 would leave an aneuploidy-high arm of a handful of patients here. Labelled either
way, and frozen before the outcome is looked at. Every row of the
table is fitted **within one cut** — one log-rank, one unadjusted Cox and one adjusted Cox
per cut per endpoint — so no model sees a patient under two classifications.
Progression-free survival is the primary endpoint because it carries more events than OS in
this cohort and is the endpoint the paper's own Fig. 1b uses; OS is reported beside it. Each
row carries a log-rank p, an unadjusted Cox HR, and a Cox HR **adjusted for stage and
grade** — TCGA UCEC is heavily Stage I, so an unadjusted split can be reading stage rather
than biology, though the adjustment does not rescue an underpowered comparison, it only stops
an unadjusted one from misleading.

```{r tcga-survival-table}
surv_ok <- have_tcga && requireNamespace("survival", quietly = TRUE)

# Braces required: a top-level `if (x) a` with `else` on the NEXT line is an R syntax error.
cut_lab <- if (same_cut) {
  paste0("ATTEND cutpoints (ASCETS >= ", paste(cuts, collapse = ", "), ")")
} else {
  # MMRd's OWN median, not the cohort's — this is a within-stratum split (see the cohorts
  # chunk), and the label has to say which line the curves were actually cut at.
  paste0("within-MMRd median (>= ", signif(tcga_mmrd_cuts, 3), ")")
}

# MMRd rows of the per-cut frame: aneuploidy_class is already re-derived at each cut, so the
# only thing added here is the stage/grade adjustment model's inputs.
tcga_surv <- if (surv_ok && nrow(tcga_mmrd_cuts_df) > 0) {
  tcga_mmrd_cuts_df |>
    filter(mmrd %in% TRUE, is.finite(aneuploidy)) |>
    mutate(stage_bin = factor(dplyr::case_when(stage %in% c("Stage I", "Stage II")   ~ "I-II",
                                               stage %in% c("Stage III", "Stage IV") ~ "III-IV",
                                               TRUE ~ NA_character_)),
           grade_f   = factor(grade))
} else NULL

# `d` is ONE cut's rows. Taking the frame as an argument rather than reading tcga_surv is
# what keeps a model from pooling the two classifications of the same patient.
fit_one <- function(d, tcol, ecol, label) {
  if (is.null(d) || !all(c(tcol, ecol) %in% names(d))) return(NULL)
  x <- d |> filter(is.finite(.data[[tcol]]), !is.na(.data[[ecol]]),
                   !is.na(aneuploidy_class))
  if (nrow(x) == 0 || dplyr::n_distinct(x$aneuploidy_class) < 2) return(NULL)
  S  <- survival::Surv(x[[tcol]], x[[ecol]])
  sd <- survival::survdiff(S ~ x$aneuploidy_class)
  p  <- 1 - stats::pchisq(sd$chisq, length(sd$n) - 1)
  ci <- summary(survival::coxph(S ~ aneuploidy_class, data = x))$conf.int[1, c(1, 3, 4)]
  # Stage/grade adjustment: NA when a stratum is empty or the model fails to converge.
  adj <- tryCatch(summary(survival::coxph(S ~ aneuploidy_class + stage_bin + grade_f,
                                          data = x))$conf.int[1, c(1, 3, 4)],
                  error = function(e) c(NA_real_, NA_real_, NA_real_))
  ev <- tapply(x[[ecol]], x$aneuploidy_class, sum)
  tibble(endpoint = label,
         n_low = sum(x$aneuploidy_class == "aneuploidy-low"),
         n_high = sum(x$aneuploidy_class == "aneuploidy-high"),
         events_low = unname(ev[1]), events_high = unname(ev[2]),
         events_total = sum(x[[ecol]]),
         logrank_p = signif(p, 3),
         HR = sprintf("%.2f (%.2f-%.2f)", ci[1], ci[2], ci[3]),
         HR_adj_stage_grade = if (is.na(adj[1])) "n/a" else
           sprintf("%.2f (%.2f-%.2f)", adj[1], adj[2], adj[3]))
}

st <- NULL
if (!is.null(tcga_surv) && nrow(tcga_surv) > 0) {
  st <- bind_rows(lapply(split(tcga_surv, droplevels(tcga_surv$as_cut_lab)), function(d)
    bind_rows(fit_one(d, "pfs_months", "pfs_event", "PFS (primary)"),
              fit_one(d, "os_months",  "os_event",  "OS")) |>
      mutate(cut = as.character(d$as_cut_lab[1]), .before = 1)))
}
if (!is.null(st) && nrow(st) > 0) {
  knitr::kable(st, caption = paste0(
    "TCGA UCEC MMRd survival by aneuploidy split, one block per cut — ", cut_lab,
    ". Every model is fitted within its own cut. READ events_total FIRST: below ~30 events ",
    "neither a significant nor a non-significant result is informative."))
} else cat("TCGA survival table skipped — needs data/tcga_2013/ + the survival package.\n")
```

The Kaplan–Meier for whichever endpoint carries the most events at that cut — the least
underpowered picture available, not a separate analysis — drawn **once per cut, in a loop**
rather than as a faceted figure: `ggsurvfit` builds its p-value annotation and its risk table
per plot, and a facet would leave one risk table under two sets of curves. Each cut therefore
gets its own endpoint choice, its own p and its own risk table. Curves take the aneuploidy
palette, so aneuploidy-high is the same red here as in every other figure on the site. The
risk table is the part to read: it shows how quickly both arms run out of patients at risk,
which is the concrete form of the power limit stated above.

```{r tcga-survival-km, fig.height=5}
if (!is.null(tcga_surv) && requireNamespace("ggsurvfit", quietly = TRUE) &&
    !is.null(st) && nrow(st) > 0) {
  for (lv in levels(droplevels(tcga_surv$as_cut_lab))) {
    sl <- st |> filter(cut == lv)
    if (nrow(sl) == 0) { note_skip(lv, ": no fitted endpoint, so no KM."); next }
    best <- sl$endpoint[which.max(sl$events_total)]
    tcol <- if (grepl("^PFS", best)) "pfs_months" else "os_months"
    ecol <- if (grepl("^PFS", best)) "pfs_event"  else "os_event"
    d <- tcga_surv |>
      filter(as.character(as_cut_lab) == lv, is.finite(.data[[tcol]]),
             !is.na(.data[[ecol]]), !is.na(aneuploidy_class)) |>
      mutate(time = .data[[tcol]], event = .data[[ecol]])
    if (nrow(d) > 0 && dplyr::n_distinct(d$aneuploidy_class) == 2) {
      print(
        ggsurvfit::survfit2(survival::Surv(time, event) ~ aneuploidy_class, data = d) |>
          ggsurvfit::ggsurvfit(linewidth = 1) +
          ggsurvfit::add_pvalue() + ggsurvfit::add_risktable() +
          # POSITIONAL, in the factor's own level order, not name-matched: ggsurvfit labels
          # its strata from the formula and the exact spelling it lands on is its business,
          # so a named lookup would fail with "insufficient values in manual scale" at knit
          # time. Same reason km_facet() takes attend_pal(2) positionally. The loop is what
          # keeps this safe across the two cuts: add_molecular_classes() rebuilds
          # aneuploidy_class with the SAME low-then-high level order at every cut
          # (test_class_level_order.R), and the n_distinct() == 2 guard above means both
          # strata are present, so position 1 is always aneuploidy-low.
          ggplot2::scale_colour_manual(
            values = unname(attend_aneu_cols[c("aneuploidy-low", "aneuploidy-high")])) +
          ggplot2::labs(title = paste0("TCGA UCEC MMRd — ", best, " by aneuploidy split (",
                                       lv, ")"),
                        subtitle = paste0("n = ", nrow(d), ", ", sum(d$event),
                                          " events — underpowered; read the CI, not the p")))
    } else note_skip(lv, ": not enough TCGA MMRd rows, or only one class, for the split KM.")
  }
} else cat("TCGA survival KM skipped — needs survival + ggsurvfit.\n")
```

```{r session-info}
sessionInfo()
```
