---
title: "ATTEND — 11. MMRd aneuploidy: metastatic (ATTEND) vs primary (TCGA)"
author: "sceriff0"
date: "2026-08-17"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---
```{r setup, include=FALSE}
# renv: tidyverse, here. Optional and guarded: diptest (Hartigan's dip), survival +
# ggsurvfit (the TCGA KM). If missing:
# renv::install(c("diptest","survival","ggsurvfit")); renv::snapshot()
# The TCGA arm needs data/tcga_2013/ — Rscript code/fetch_tcga_ucec_2013.R — and the
# shared-scale panels additionally need ASCETS run on TCGA's segments: bash code/run_ascets_tcga.sh
library(tidyverse)
library(here)
# Headless-safe raster device. The cluster's compute nodes have no X11, so R's default
# png() fails with "unable to open connection to X11 display" and every figure dies. cairo
# needs no display; the guard keeps this working on a machine built without it.
if (capabilities("cairo")) options(bitmapType = "cairo")

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 5)

source(here("code", "attend_classes.R"))     # mmrd_aneuploidy_split(), dip_bimodality(),
                                             # attend_tcga_ref, attend_tcga_ref_2013, thresholds
source(here("code", "attend_plots.R"))       # attend_theme(), attend_box(), attend_cohort_cols
source(here("code", "attend_io.R"))          # read_intermediate()
source(here("code", "build_master.R"))       # get_master() — builds on demand
source(here("code", "load_wes_results.R"))   # load_tcga_ucec_2013(), load_tcga_2013_ascets(),
                                             # load_tcga_ucec() (legacy fallback)
```
> **Question.** Does aneuploidy substructure MMR-deficient tumours in metastatic disease (ATTEND) in a way that is absent from primary disease (TCGA UCEC)?
> **Cohort.** MMR-deficient only, in both cohorts, side by side.
> **Reads.** The master (built on demand by get_master()) for the ATTEND arm; data/tcga_2013/ plus ASCETS run on its segments for the TCGA arm.
> **Out of scope.** The ATTEND MMRd split itself, against mutations and composition -> [05-aneuploidy-mmrd](05-aneuploidy-mmrd.html). ATTEND TMB bimodality and its own TCGA replication -> [08-tmb](08-tmb.html). Every ATTEND survival model -> [04-survival](04-survival.html). Cluster-level correspondence between the two cohorts -> [10-published-peak-clustering](10-published-peak-clustering.html).

ATTEND is a cohort of **metastatic** endometrial tumours; TCGA UCEC is **primary**.
[05-aneuploidy-mmrd](05-aneuploidy-mmrd.html) works inside ATTEND's MMR-deficient stratum,
where aneuploidy appears to carve the group in two. This report asks whether that is specific
to metastatic disease, on four measurements that fail in different ways, so no single one has
to carry the answer:

* **Shape** — is the aneuploidy distribution within MMRd bimodal in ATTEND and unimodal in
  TCGA? Bimodality is **scale-invariant**, so this works even when the two cohorts' scores
  are on different axes.
* **Level** — is MMRd aneuploidy *absolutely* higher in metastatic disease? This needs a
  shared axis and is skipped without one.
* **Prevalence** — what fraction of MMRd tumours cross the aneuploidy line in each cohort?
  Two integers per cohort and an exact test: no distributional assumption, no survival event,
  and the only arm whose precision improves as the cohorts grow.
* **Orthogonality** — does TMB track aneuploidy *within* an MMR stratum in TCGA? If it did,
  the aneuploidy split would be re-discovering the hypermutator phenotype rather than adding
  an axis.

The shared axis is the load-bearing piece. ATTEND's aneuploidy score is ASCETS (fraction of
evaluable arms altered, 0–1); the cBioPortal column is Taylor et al. 2018, a *count* of
altered arms (0–39). No constant converts one into the other, because ASCETS's denominator is
the per-sample number of evaluable arms. `code/run_ascets_tcga.sh` removes the obstacle by
running **ASCETS on TCGA's own segments**, producing the same statistic on both sides;
without that output every scale-dependent panel below is skipped rather than computed on
incomparable numbers.

Calling stringency is matched where it can be: ATTEND runs `ascets(min_boc = 0.5,
alteration_threshold = 0.7)` and leaves `threshold` and `noise` at their defaults — a fixed
±0.2 log2 cutoff with no LCR noise modelling — and `run_ascets_tcga.sh` uses those same
values. (ATTEND runs ASCETS per sample and the TCGA script runs it pooled; with a fixed
threshold that is equivalent, since the only cohort-level step in ASCETS is the
noise-threshold estimation, which is not in use.) **One difference cannot be removed here:**
ATTEND is tumour-only DRAGEN WES with no purity/ploidy correction, TCGA is purity-corrected
SNP6. Low-purity ATTEND tumours have compressed log2 ratios and will *under*-call arms, so an
ATTEND-higher result is conservative — but the assay difference also changes which arms clear
the coverage cut, so a *small* level difference is suggestive, not decisive.

## Part 1 — is the split present in each cohort?

The ATTEND arm is the MMR-deficient stratum of the master, taken through
`mmrd_aneuploidy_split()`, whose MMRd criterion is MMR-protein loss by **IHC** —
[00-methods](00-methods.html) explains why MSI-high is a correlate and not a criterion. Its
aneuploidy values are the ASCETS scores loaded from `data/ascets/`.

The TCGA arm prefers the **2013 publication cohort** (`data/tcga_2013/`, via
`load_tcga_ucec_2013()`) joined to ASCETS run on its segments (`load_tcga_2013_ascets()`),
which is what creates the shared scale. Its MMRd rule is `attend_tcga_ref_2013$mmrd_rule`,
whose default `"tcga_exact"` makes MMRd **the paper's own MSI (hyper-mutated) integrative
cluster and nothing else**: every other classified tumour is MMRp, and the tumours TCGA never
classified are NA — unclassified, not proficient. That is the only rule whose output can be
quoted as "TCGA's MSI group". If the ASCETS output is absent the report falls back to the
legacy PanCanAtlas export (`load_tcga_ucec()`, Taylor arm count) and drops every
scale-dependent panel, saying so in the table.

```{r cohorts}
master      <- get_master() |> add_molecular_classes()
attend_mmrd <- mmrd_aneuploidy_split(master)
attend_aneu <- if (!is.null(attend_mmrd))
  attend_mmrd$aneu_value[is.finite(attend_mmrd$aneu_value)] else numeric(0)

t13 <- tryCatch(load_tcga_ucec_2013(),   error = function(e) tibble())
tas <- tryCatch(load_tcga_2013_ascets(), error = function(e) tibble())
use_ascets <- nrow(t13) > 0 && nrow(tas) > 0

if (use_ascets) {
  tcga <- t13 |>
    left_join(tas, by = "ID") |>
    distinct(pid, .keep_all = TRUE) |>          # survival is a per-PATIENT analysis
    transmute(pid, mmrd, pole, tmb, aneuploidy = ascets_aneuploidy, cn_cluster_k4,
              os_months, os_event, pfs_months, pfs_event, stage, grade)
  tcga_scale  <- "ASCETS (fraction of evaluable arms, 0-1) — SAME statistic as ATTEND"
  tcga_source <- "data/tcga_2013/ascets/ (code/run_ascets_tcga.sh)"
} else {
  tcga <- tryCatch(load_tcga_ucec(), error = function(e) tibble())
  if (nrow(tcga) > 0) tcga <- tcga |>
    mutate(pfs_months = NA_real_, pfs_event = NA_integer_,
           stage = NA_character_, grade = NA_character_, cn_cluster_k4 = NA_character_,
           pole = NA, tmb = suppressWarnings(as.numeric(tmb)))
  tcga_scale  <- paste0("cBioPortal ", attend_tcga_ref$aneuploidy_col,
                        " (Taylor 2018 arm COUNT, 0-39) — NOT ATTEND's scale")
  tcga_source <- "data/tcga/ (code/fetch_tcga_ucec.R)"
}

tcga_mmrd <- if (nrow(tcga) > 0) filter(tcga, mmrd) else tcga
tcga_aneu <- if (nrow(tcga_mmrd) > 0) tcga_mmrd$aneuploidy[is.finite(tcga_mmrd$aneuploidy)] else numeric(0)

lab_attend <- "ATTEND (metastatic)"
lab_tcga   <- "TCGA UCEC (primary)"
have_attend <- length(attend_aneu) >= 4
have_tcga   <- length(tcga_aneu)   >= 4
# Every LEVEL comparison below requires both cohorts on the ASCETS [0,1] scale. Shape
# comparisons do not, because bimodality is scale-invariant.
comparable  <- use_ascets && have_attend && have_tcga &&
               all(is.finite(range(attend_aneu))) && max(attend_aneu, na.rm = TRUE) <= 1

knitr::kable(tibble(
  cohort  = c(lab_attend, lab_tcga),
  MMRd_n  = c(length(attend_aneu), length(tcga_aneu)),
  scale   = c("ASCETS (fraction of evaluable arms, 0-1)", tcga_scale),
  source  = c("data/ascets/ (DRAGEN .seg)", tcga_source)),
  caption = "MMR-deficient tumours per cohort, and the aneuploidy statistic each uses.")

# cat(), not message(): opts_chunk sets message = FALSE, so a message() is invisible on the
# knitted page and a skipped comparison then looks like nothing happened at all.
if (!have_tcga)
  cat("\n**No TCGA MMRd data** — run `Rscript code/fetch_tcga_ucec_2013.R`, then re-knit.\n")
if (nrow(t13) > 0 && !use_ascets)
  cat("\n**TCGA 2013 clinical present but no ASCETS output** — run `bash code/run_ascets_tcga.sh`",
      "to unlock the shared-scale panels (Parts 2 and 3).\n")
```

Aneuploidy density within MMRd, one panel per cohort, each annotated with **Hartigan's
dip-test** p from `dip_bimodality()` — a small p means the distribution is multimodal, i.e. a
real split rather than a continuum. The panels are **free-scaled on purpose**: this figure
compares shape only, and putting two different statistics on one axis would invite exactly
the level comparison that Part 2 gates behind the shared scale. TCGA is drawn in the
reference grey and only ATTEND carries a hue, because TCGA is the external cohort the ATTEND
result is measured against. The dip test needs at least four finite values and the optional
`diptest` package; where either is missing the annotation says so instead of printing a
number.

```{r density-contrast, fig.height=5}
dip_a <- dip_bimodality(attend_aneu)
dip_t <- dip_bimodality(tcga_aneu)

# Braces required: a function body of the bare form `if (x) a` with `else` on the NEXT line
# is an R syntax error, the same trap the survival-cut label below carries.
dip_lab <- function(d) {
  if (is.finite(d$p)) {
    sprintf("dip p = %.3g%s", d$p,
            ifelse(isTRUE(d$bimodal), " (multimodal)", " (unimodal)"))
  } else "diptest unavailable"
}

dens_df <- bind_rows(
  if (have_attend) tibble(cohort = lab_attend, aneu = attend_aneu) else NULL,
  if (have_tcga)   tibble(cohort = lab_tcga,   aneu = tcga_aneu)   else NULL)

if (nrow(dens_df) > 0) {
  labs_df <- tibble(cohort = c(lab_attend, lab_tcga),
                    lab    = c(dip_lab(dip_a), dip_lab(dip_t))) |>
    filter(cohort %in% dens_df$cohort)

  ggplot(dens_df, aes(aneu)) +
    geom_density(aes(fill = cohort), colour = "grey30", alpha = 0.5, adjust = 0.9) +
    geom_rug(alpha = 0.4) +
    geom_text(data = labs_df, aes(x = Inf, y = Inf, label = lab),
              hjust = 1.05, vjust = 1.5, size = 3.3) +
    facet_wrap(~ cohort, scales = "free") +
    scale_fill_manual(values = attend_cohort_cols, guide = "none") +
    labs(title = "Aneuploidy within MMRd — metastatic vs primary",
         subtitle = "free axes: shape only. Multimodal = a real split; unimodal = a continuum",
         x = "aneuploidy score (cohort-specific scale)", y = "density") +
    attend_theme()
} else cat("No aneuploidy data for either cohort yet.\n")
```

The same dip test as a table, so the statistic and n are legible rather than read off a
panel annotation. Read `n` first: with a handful of tumours per cohort the dip test has very
little power, so a non-significant p is weak evidence of unimodality and not evidence of it.

```{r bimodality-table}
knitr::kable(tibble(
  cohort   = c(lab_attend, lab_tcga),
  n        = c(dip_a$n, dip_t$n),
  dip_stat = round(c(dip_a$dip, dip_t$dip), 4),
  dip_p    = signif(c(dip_a$p, dip_t$p), 3),
  multimodal = c(dip_a$bimodal, dip_t$bimodal)),
  caption = paste0("Hartigan dip test per cohort, on the MMRd aneuploidy values. ",
                   "Scale-invariant, so it is valid even on the arm-count fallback."))
```

## Part 2 — is MMRd aneuploidy absolutely higher in metastatic disease?

Part 1 compared shape; this compares **level**, which the shared ASCETS scale makes
legitimate. Both axes are now "fraction of evaluable chromosome arms altered", so the two
densities can be drawn on one axis and the medians compared directly. The dashed line is
`attend_thresholds$aneuploidy` = `r attend_thresholds$aneuploidy`, ATTEND's aneuploidy-high
cutoff, drawn in the aneuploidy-high colour because that is what it marks — and it now means
the same thing in both cohorts.

The test is a **Mann–Whitney** (rank-based, no distributional assumption) on the two MMRd
aneuploidy vectors, reported with the rank-biserial effect size, because a p-value alone is
uninformative when the group sizes differ by an order of magnitude. The whole part is
**skipped** unless ASCETS has been run on TCGA: on mismatched scales the comparison is
meaningless rather than merely imprecise. Interpret the size of any shift against the
purity/platform asymmetry stated at the top — the two uncorrectable biases run in opposite
directions, so a small shift is not interpretable in either direction.

```{r absolute-level, fig.height=4.5}
if (comparable) {
  both <- bind_rows(tibble(cohort = lab_attend, aneu = attend_aneu),
                    tibble(cohort = lab_tcga,   aneu = tcga_aneu))
  w  <- suppressWarnings(stats::wilcox.test(attend_aneu, tcga_aneu))
  # rank-biserial r = 1 - 2U/(n1*n2); negated below so positive = ATTEND higher.
  U  <- unname(w$statistic); rbc <- 1 - 2 * U / (length(attend_aneu) * length(tcga_aneu))
  md <- both |> group_by(cohort) |>
    summarise(n = n(), median = round(median(aneu), 3),
              IQR = paste0(round(quantile(aneu, .25), 3), "-", round(quantile(aneu, .75), 3)),
              .groups = "drop")
  knitr::kable(md, caption = paste0(
    "MMRd aneuploidy on the SHARED ASCETS scale. Mann-Whitney p = ", signif(w$p.value, 3),
    "; rank-biserial r = ", round(-rbc, 3), " (positive = ATTEND higher)."))

  ggplot(both, aes(aneu, fill = cohort)) +
    geom_density(colour = "grey30", alpha = 0.45, adjust = 0.9) +
    geom_vline(xintercept = attend_thresholds$aneuploidy, linetype = "dashed",
               colour = attend_aneu_high, linewidth = 0.4) +
    scale_fill_manual(values = attend_cohort_cols, name = NULL) +
    labs(title = "MMRd aneuploidy, metastatic vs primary — one shared axis",
         subtitle = "ASCETS fraction of evaluable arms altered; dashed = aneuploidy-high cutoff",
         x = "ASCETS aneuploidy score (0-1)", y = "density") +
    attend_theme()
} else if (!use_ascets) {
  cat("Level comparison skipped — needs ASCETS on TCGA. Run `bash code/run_ascets_tcga.sh`.\n")
} else cat("Level comparison skipped — ATTEND aneuploidy is not on the [0,1] ASCETS scale.\n")
```

## Part 3 — how many MMRd tumours are aneuploidy-high in each cohort?

Part 1 asks whether a split exists and Part 2 whether the whole distribution shifts. This
asks the simplest question of the three: **what fraction of MMRd tumours cross the
aneuploidy line, in each cohort?** A 2×2 and a Fisher exact test, with exact
Clopper–Pearson intervals on each proportion. It needs no distributional assumption and no
survival event, which is why the intervals are the part to read: with a small
aneuploidy-high count on either side the interval stays wide even when the point estimates
differ substantially, and a wide interval is the honest summary.

The cutoff is `attend_thresholds$aneuploidy` = `r attend_thresholds$aneuploidy`, applied
identically to both ASCETS scores, and the arm is skipped on the arm-count fallback because
the criterion is not transferable there. The **fixed** cutoff is used deliberately, not the
GMM-refined boundary `mmrd_aneuploidy_split()` can compute: that boundary is fitted to
ATTEND's own distribution, so applying it to TCGA would beg the question this part asks.

```{r prop-shared}
prop_ci <- function(x, n) {                       # Clopper-Pearson exact interval
  if (n == 0) return("n/a")
  ci <- stats::binom.test(x, n)$conf.int
  sprintf("%.1f%% (%.1f-%.1f)", 100 * x / n, 100 * ci[1], 100 * ci[2])
}

if (comparable) {
  cut_shared <- attend_thresholds$aneuploidy
  a_hi <- sum(attend_aneu >= cut_shared); a_n <- length(attend_aneu)
  t_hi <- sum(tcga_aneu   >= cut_shared); t_n <- length(tcga_aneu)
  ft <- stats::fisher.test(matrix(c(a_hi, a_n - a_hi, t_hi, t_n - t_hi), nrow = 2))
  knitr::kable(tibble(
    cohort          = c(lab_attend, lab_tcga),
    MMRd_n          = c(a_n, t_n),
    aneuploidy_high = c(a_hi, t_hi),
    percent         = c(prop_ci(a_hi, a_n), prop_ci(t_hi, t_n))),
    caption = sprintf(paste0("MMRd tumours at or above the shared ASCETS cutoff (>= %s). ",
                             "Fisher exact p = %.3g; odds ratio %.2f (%.2f-%.2f). ",
                             "Percentages carry exact Clopper-Pearson intervals."),
                      cut_shared, ft$p.value, ft$estimate,
                      ft$conf.int[1], ft$conf.int[2]))
} else cat("Skipped — needs ASCETS on TCGA so both cohorts share one cutoff.",
           "Run `bash code/run_ascets_tcga.sh`.\n")
```

A second reference point taken **entirely from TCGA's published data**: how many of TCGA's
MMRd tumours were assigned to copy-number **cluster 4**, its serous-like SCNA cluster? Both
fields are TCGA's own (integrative cluster and `CNA_CLUSTER_K4`, Suppl. Data File S1.1), so
this proportion involves no cutoff, no ASCETS and no choice of ours. It is **not the same
criterion** as the table above — cluster membership rather than an aneuploidy threshold — so
the two are not interchangeable and only the shared-cutoff table supports a like-for-like
test. This is a sanity anchor: if the shared cutoff puts a wildly different fraction of TCGA
MMRd above the line than TCGA's own clustering did, the cutoff is drawing a different line
than the clustering, and that needs explaining before either number is quoted. The
cross-tabulation is a derived observation — the paper reports cluster 4's histology
composition but never its subtype composition — so it is not quotable from the publication.

```{r prop-published}
if (have_tcga && "cn_cluster_k4" %in% names(tcga_mmrd) && any(!is.na(tcga_mmrd$cn_cluster_k4))) {
  d  <- tcga_mmrd |> filter(!is.na(cn_cluster_k4))
  n4 <- sum(as.character(d$cn_cluster_k4) == "4")
  knitr::kable(tibble(
    definition  = "TCGA MMRd in copy-number cluster 4 (serous-like)",
    MMRd_n      = nrow(d),
    in_cluster4 = n4,
    percent     = prop_ci(n4, nrow(d))),
    caption = "TCGA's own published assignment — no threshold, no ASCETS, no choice of ours.")
} else cat("Published-anchor table skipped — needs the 2013 cohort with CNA_CLUSTER_K4.\n")
```

## Part 4 — does TMB track aneuploidy within an MMR stratum?

[08-tmb](08-tmb.html) Part 2 asks, in ATTEND: within each MMR stratum, does TMB differ
between aneuploidy-high and aneuploidy-low tumours? Splitting by MMR **first** matters
because the two routes to a genomic extreme are distinct — MMR-deficient tumours are
hypermutators (high TMB, often modest aneuploidy), whereas the CN-high / serous-like route
drives high aneuploidy at *low* TMB. If aneuploidy-high and TMB-high were the same tumours,
aneuploidy would add nothing to TMB; the interest is in their being different axes. This part
puts the same question to TCGA.

TMB here is TCGA's `TMB_NONSYNONYMOUS` from `data_clinical_sample.txt`, available on the
exome-sequenced subset, split at `attend_thresholds$tmb` = `r attend_thresholds$tmb` mut/Mb.
It is **not** either of ATTEND's two reported definitions and is never plotted alongside them
— it is a different cohort's single number, and [00-methods](00-methods.html) fixes what the
two ATTEND definitions are. It also needs no ancestry correction: ATTEND recomputes
ancestry-aware TMB because its TMB is *tumour-only*, so residual private germline variants
leak into the count in an ancestry-structured way, whereas TCGA's value comes from
tumour/normal **paired** exomes where the matched normal already removes germline variants
per patient. Applying a Nassar-style correction to it would subtract a bias that has already
been removed.

**POLE is excluded from MMRp, and that is not optional.** POLE-ultramutated tumours have a
median TMB two orders of magnitude above the copy-number groups, so leaving them in MMRp
would make that panel unreadable and misrepresent MMRp as high-TMB. ATTEND has no POLE cases,
so excluding them is also what makes the two cohorts' MMRp groups comparable; the POLE count
and median are printed separately for reference.

```{r tcga-tmb-build}
# The aneuploidy cut is the shared pipeline cutoff when the scales match, else the
# within-cohort median — labelled either way, never tuned to the outcome.
aneu_cut_tmb <- if (comparable) attend_thresholds$aneuploidy else
  stats::median(tcga$aneuploidy, na.rm = TRUE)
have_tcga_tmb <- nrow(tcga) > 0 && "tmb" %in% names(tcga) && any(is.finite(tcga$tmb))

tcga_tmb <- if (have_tcga_tmb) {
  tcga |>
    mutate(
      mmr_group = dplyr::case_when(!is.na(pole) & pole ~ "POLE",
                                   mmrd %in% TRUE      ~ "MMRd",
                                   mmrd %in% FALSE     ~ "MMRp",
                                   TRUE                ~ NA_character_),
      aneuploidy_class = factor(ifelse(aneuploidy >= aneu_cut_tmb,
                                       "aneuploidy-high", "aneuploidy-low"),
                                levels = c("aneuploidy-low", "aneuploidy-high"))) |>
    filter(is.finite(tmb), !is.na(aneuploidy_class), mmr_group %in% c("MMRp", "MMRd")) |>
    mutate(mmr_group = factor(mmr_group, levels = c("MMRp", "MMRd")))
} else tibble()

if (have_tcga_tmb) {
  pole_n   <- sum(!is.na(tcga$pole) & tcga$pole)
  pole_med <- suppressWarnings(stats::median(tcga$tmb[!is.na(tcga$pole) & tcga$pole], na.rm = TRUE))
  cat("patients per MMR group x aneuploidy status:\n")
  print(tcga_tmb |> count(mmr_group, aneuploidy_class) |>
          tidyr::pivot_wider(names_from = aneuploidy_class, values_from = n, values_fill = 0))
  cat("\nPOLE (excluded from the panels): n = ", pole_n,
      ", median TMB = ", round(pole_med, 1), " mut/Mb\n", sep = "")
} else cat("TMB part skipped — needs the 2013 cohort with TMB_NONSYNONYMOUS.\n")
```

TCGA TMB in aneuploidy-high versus aneuploidy-low tumours, faceted by MMR group, boxes
filled by aneuploidy status and the dashed line at the TMB-high cut. `attend_box()` picks the
mark per facet cell from the data — box and points where the cell has at least ten patients,
a median crossbar and points below that — and prints n, because the TCGA MMRd
aneuploidy-high cell is small and quartiles drawn from a handful of values would be an
artefact of which two landed in the middle. The y-axis is free: MMRd TMB sits an order of
magnitude above MMRp, so a shared axis would flatten the MMRp panel.

```{r tcga-tmb-box, fig.height=4.5}
if (nrow(tcga_tmb) > 0) {
  ggplot(tcga_tmb, aes(aneuploidy_class, tmb)) +
    attend_box(tcga_tmb, x = "aneuploidy_class", y = "tmb", by = "mmr_group",
               fill = NULL, outlier.shape = NA) +
    aes(fill = aneuploidy_class) +
    geom_hline(yintercept = attend_thresholds$tmb, linetype = "dashed",
               colour = attend_aneu_high, linewidth = 0.4) +
    scale_fill_manual(values = attend_aneu_cols, name = "aneuploidy status") +
    scale_x_discrete(labels = c("aneuploidy-low" = "low", "aneuploidy-high" = "high")) +
    facet_wrap(~ mmr_group, scales = "free_y") +
    labs(title = "TCGA UCEC — TMB by aneuploidy status, within MMR strata",
         subtitle = paste0("dashed = TMB-high cut (", attend_thresholds$tmb, " mut/Mb); ",
                           "aneuploidy-high = score >= ", signif(aneu_cut_tmb, 3),
                           "; POLE excluded"),
         x = NULL, y = "TMB (mut/Mb, nonsynonymous)") +
    attend_theme()
} else cat("No TCGA TMB/aneuploidy data to plot.\n")
```

The same contrast as counts: the TMB-high fraction on each side of the aneuploidy line,
within each MMR group, with a Fisher exact test per group. The test is withheld — reported as
NA — wherever an aneuploidy group holds fewer than three patients, because a 2×2 with a cell
of one or two produces a p-value that moves entirely with a single reclassification.

```{r tcga-tmb-test}
if (nrow(tcga_tmb) > 0) {
  tab <- tcga_tmb |>
    group_by(mmr_group) |>
    group_modify(function(.x, ...) {
      hi    <- .x$tmb >= attend_thresholds$tmb
      is_lo <- .x$aneuploidy_class == "aneuploidy-low"
      n_lo <- sum(is_lo); n_hi <- sum(!is_lo)
      hilo <- sum(is_lo & hi); hihi <- sum(!is_lo & hi)
      p <- if (n_lo >= 3 && n_hi >= 3)
        suppressWarnings(stats::fisher.test(
          matrix(c(hilo, n_lo - hilo, hihi, n_hi - hihi), 2))$p.value) else NA_real_
      tibble(aneu_low_n = n_lo, TMBhi_in_low = hilo,
             aneu_high_n = n_hi, TMBhi_in_high = hihi,
             fisher_p = if (is.na(p)) NA_character_ else signif(p, 3))
    })
  knitr::kable(tab, caption = paste0(
    "TMB-high (>= ", attend_thresholds$tmb, " mut/Mb) fraction by aneuploidy status, within ",
    "each MMR group. Fisher exact per group; NA when an aneuploidy group has < 3 patients."))
} else cat("TMB 2x2 skipped.\n")
```

## Part 5 — the split in TCGA MMRd survival, and why it cannot decide anything

Parts 1–3 ask whether a split exists; this asks whether it carries outcome signal in primary
disease. **Read the event count before the p-value, and treat this part as descriptive
only.** TCGA UCEC is an early-stage, long-follow-up cohort with very few events among MMRd
patients — of the order of a dozen deaths under the `"tcga_exact"` rule. At that event count
the test cannot detect even a large hazard ratio, so a non-significant result here is **not**
evidence that the split is absent in primary disease, and a significant one is fragile,
threshold-dependent noise. The arms that can discriminate are shape, level and prevalence.

The cut is ATTEND's own cutpoint when the scales match and the within-cohort median
otherwise, labelled either way and frozen before the outcome is looked at.
Progression-free survival is the primary endpoint because it carries more events than OS in
this cohort and is the endpoint the paper's own Fig. 1b uses; OS is reported beside it. Each
row carries a log-rank p, an unadjusted Cox HR, and a Cox HR **adjusted for stage and
grade** — TCGA UCEC is heavily Stage I, so an unadjusted split can be reading stage rather
than biology, though the adjustment does not rescue an underpowered comparison, it only stops
an unadjusted one from misleading.

```{r tcga-survival-table}
surv_ok <- have_tcga && requireNamespace("survival", quietly = TRUE)

cutx    <- if (comparable) attend_thresholds$aneuploidy else stats::median(tcga_aneu, na.rm = TRUE)
# Braces required: a top-level `if (x) a` with `else` on the NEXT line is an R syntax error.
cut_lab <- if (comparable) {
  paste0("ATTEND cutpoint (ASCETS >= ", cutx, ")")
} else {
  paste0("within-cohort median (>= ", signif(cutx, 3), ")")
}

tcga_surv <- if (surv_ok) {
  tcga_mmrd |>
    filter(is.finite(aneuploidy)) |>
    mutate(aneuploidy_class = factor(ifelse(aneuploidy >= cutx,
                                            "aneuploidy-high", "aneuploidy-low"),
                                     levels = c("aneuploidy-low", "aneuploidy-high")),
           stage_bin = factor(dplyr::case_when(stage %in% c("Stage I", "Stage II")   ~ "I-II",
                                               stage %in% c("Stage III", "Stage IV") ~ "III-IV",
                                               TRUE ~ NA_character_)),
           grade_f   = factor(grade))
} else NULL

fit_one <- function(tcol, ecol, label) {
  if (is.null(tcga_surv) || !all(c(tcol, ecol) %in% names(tcga_surv))) return(NULL)
  x <- tcga_surv |> filter(is.finite(.data[[tcol]]), !is.na(.data[[ecol]]),
                           !is.na(aneuploidy_class))
  if (nrow(x) == 0 || dplyr::n_distinct(x$aneuploidy_class) < 2) return(NULL)
  S  <- survival::Surv(x[[tcol]], x[[ecol]])
  sd <- survival::survdiff(S ~ x$aneuploidy_class)
  p  <- 1 - stats::pchisq(sd$chisq, length(sd$n) - 1)
  ci <- summary(survival::coxph(S ~ aneuploidy_class, data = x))$conf.int[1, c(1, 3, 4)]
  # Stage/grade adjustment: NA when a stratum is empty or the model fails to converge.
  adj <- tryCatch(summary(survival::coxph(S ~ aneuploidy_class + stage_bin + grade_f,
                                          data = x))$conf.int[1, c(1, 3, 4)],
                  error = function(e) c(NA_real_, NA_real_, NA_real_))
  ev <- tapply(x[[ecol]], x$aneuploidy_class, sum)
  tibble(endpoint = label,
         n_low = sum(x$aneuploidy_class == "aneuploidy-low"),
         n_high = sum(x$aneuploidy_class == "aneuploidy-high"),
         events_low = unname(ev[1]), events_high = unname(ev[2]),
         events_total = sum(x[[ecol]]),
         logrank_p = signif(p, 3),
         HR = sprintf("%.2f (%.2f-%.2f)", ci[1], ci[2], ci[3]),
         HR_adj_stage_grade = if (is.na(adj[1])) "n/a" else
           sprintf("%.2f (%.2f-%.2f)", adj[1], adj[2], adj[3]))
}

st <- bind_rows(fit_one("pfs_months", "pfs_event", "PFS (primary)"),
                fit_one("os_months",  "os_event",  "OS"))
if (!is.null(st) && nrow(st) > 0) {
  knitr::kable(st, caption = paste0(
    "TCGA UCEC MMRd survival by aneuploidy split — ", cut_lab,
    ". READ events_total FIRST: below ~30 events neither a significant nor a ",
    "non-significant result is informative."))
} else cat("TCGA survival table skipped — needs data/tcga_2013/ + the survival package.\n")
```

The Kaplan–Meier for whichever endpoint carries the most events — the least underpowered
picture available, not a separate analysis. Curves take the aneuploidy palette, so
aneuploidy-high is the same red here as in every other figure on the site. The risk table is
the part to read: it shows how quickly both arms run out of patients at risk, which is the
concrete form of the power limit stated above.

```{r tcga-survival-km, fig.height=5}
if (!is.null(tcga_surv) && requireNamespace("ggsurvfit", quietly = TRUE) &&
    !is.null(st) && nrow(st) > 0) {
  best <- st$endpoint[which.max(st$events_total)]
  tcol <- if (grepl("^PFS", best)) "pfs_months" else "os_months"
  ecol <- if (grepl("^PFS", best)) "pfs_event"  else "os_event"
  d <- tcga_surv |>
    filter(is.finite(.data[[tcol]]), !is.na(.data[[ecol]]), !is.na(aneuploidy_class)) |>
    mutate(time = .data[[tcol]], event = .data[[ecol]])
  if (nrow(d) > 0 && dplyr::n_distinct(d$aneuploidy_class) == 2) {
    print(
      ggsurvfit::survfit2(survival::Surv(time, event) ~ aneuploidy_class, data = d) |>
        ggsurvfit::ggsurvfit(linewidth = 1) +
        ggsurvfit::add_pvalue() + ggsurvfit::add_risktable() +
        ggplot2::scale_colour_manual(values = attend_aneu_cols) +
        ggplot2::labs(title = paste0("TCGA UCEC MMRd — ", best, " by aneuploidy split (",
                                     cut_lab, ")"),
                      subtitle = paste0("n = ", nrow(d), ", ", sum(d$event),
                                        " events — underpowered; read the CI, not the p")))
  } else cat("Not enough TCGA MMRd rows for the split KM.\n")
} else cat("TCGA survival KM skipped — needs survival + ggsurvfit.\n")
```

```{r session-info}
sessionInfo()
```
