---
title: "ATTEND — 11. Join audit: coverage, unmapped records, value sanity"
author: "sceriff0"
date: "2026-08-11"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
library(here)
library(UpSetR)
options(bitmapType = "cairo")

knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 8, fig.height = 5
)

source(here("code", "attend_harmonise.R"))   # id_cfg, crosswalks, barcode_datasets, norm_id()
source(here("code", "attend_classes.R"))     # attend_cols, have_cols(), numeric_audit()
source(here("code", "attend_plots.R"))       # highlight_points() overlay
source(here("code", "attend_io.R"))          # read_intermediate()

# Report 10 wrote every raw table alongside the master, so the audit reconstructs the join
# from those intermediates rather than re-reading HPC storage. Same inputs, same crosswalk
# code, so the mapping audited here is the mapping the master was built with.
rd <- function(nm) tryCatch(read_intermediate(nm), error = function(e) NULL)
master               <- rd("attend_master_joined")
clinical_data        <- rd("clinical_data")
imaging_data         <- rd("imaging_data")
ihc_data             <- rd("ihc_data")
gianlu_clinical_data <- rd("gianlu_clinical_data")
aneu_data            <- rd("aneu_data")
hrd_data             <- rd("hrd_data")
msi_data             <- rd("msi_data")
tmb_data             <- rd("tmb_data")
maf_data             <- rd("maf_data")
conversion_data      <- rd("conversion_data")

id_cfg         <- make_id_cfg()
cw_barcode_pid <- build_barcode_pid(gianlu_clinical_data, id_cfg)
cw_image_pid   <- build_image_pid(imaging_data, id_cfg)
pid_vector     <- make_pid_vector(cw_barcode_pid, cw_image_pid, recover = TRUE)

tables <- list(
  tmb = tmb_data, msi = msi_data, maf = maf_data, hrd = hrd_data,
  aneu = aneu_data, gianlu = gianlu_clinical_data,
  imaging = imaging_data, ihc = ihc_data
)
sets <- build_sets(tables, id_cfg, pid_vector)
```

> **Question.** What did the join cost, and are the values it produced usable?
> **Cohort.** All `r if (is.null(master)) "?" else nrow(master)` patients on the master; no restriction.
> **Reads.** `attend_master_joined` and the per-dataset intermediates, both written by [10_data_integration](10_data_integration.html).
> **Out of scope.** Building the join -> [10_data_integration](10_data_integration.html). Any biological comparison -> the 2x and 3x reports.

A join that silently drops a third of a modality is worse than no join, so the losses are
counted here rather than discovered downstream. "Unmapped" means three different things —
at record level, at patient level, and at cohort level — and each is reported separately.

Presence is not the same as usable, so the last section checks the *contents* of the table
report 10 wrote. That file is the contract every later report inherits, and a bad value is
cheapest to catch here.

## Records that never reached a patient

Native IDs that failed to resolve to any patient — orphaned records. The first table
counts them per dataset; the second lists them.

```{r unmapped}
mapres <- Map(function(df, cfg) {
  ids <- as.character(df[[cfg$col]])
  pid <- pid_vector(df, cfg)
  ok  <- !is.na(pid)
  list(n_ids      = length(unique(ids)),
       mapped     = unique(pid[ok]),
       unmapped   = unique(ids[!ok]),
       mapped_ids = unique(ids[ok]))
}, tables, id_cfg[names(tables)])

unmapped_tbl <- mapres |>
  lapply(\(x) tibble(unmapped_id = x$unmapped)) |>
  bind_rows(.id = "dataset")

unmapped_tbl |> count(dataset, name = "n_unmapped")
print(unmapped_tbl, n = 50)
```

A raw orphan count overstates real loss, because some orphans differ from a known key
only by case or whitespace. This splits the barcode-keyed datasets' orphans into
recoverable formatting drift and genuine absence.

```{r unmapped-nearmiss}
norm <- function(x) toupper(trimws(as.character(x)))
barcode_pool <- unique(norm(cw_barcode_pid$barcode))

unmapped_tbl |>
  filter(dataset %in% barcode_datasets) |>
  mutate(recoverable_by_norm = norm(unmapped_id) %in% barcode_pool) |>
  count(dataset, recoverable_by_norm)
```

### What safe recovery actually reclaimed

The master was built with `recover = TRUE`, so the near-misses above were acted on, not
just counted. First table: records that are orphans under exact matching but resolve to a
single patient once normalised. Second table: normalised keys that map to more than one
patient, refused rather than guessed, because recovering them could attach a sample to the
wrong patient.

```{r recovery-ledger}
pid_vector_exact <- make_pid_vector(cw_barcode_pid, cw_image_pid, recover = FALSE)

recoverable <- names(tables)[vapply(names(tables),
  function(nm) id_cfg[[nm]]$space %in% c("barcode", "image"), logical(1))]
reclaimed <- lapply(recoverable, function(nm) {
  df  <- tables[[nm]]; cfg <- id_cfg[[nm]]
  if (is.null(df)) return(NULL)
  ex  <- pid_vector_exact(df, cfg)
  rc  <- pid_vector(df, cfg)          # recover = TRUE, the one used for the master
  got <- is.na(ex) & !is.na(rc)
  if (!any(got)) return(NULL)
  tibble(dataset = nm,
         space         = cfg$space,
         orphaned_id   = as.character(df[[cfg$col]])[got],
         norm_key      = norm_id(as.character(df[[cfg$col]])[got]),
         recovered_pid = rc[got])
}) |> bind_rows()

cat("Patients reclaimed by safe recovery:",
    dplyr::n_distinct(reclaimed$recovered_pid), "distinct pid(s) across",
    nrow(reclaimed), "reclaimed ID row(s)\n")
if (nrow(reclaimed)) print(reclaimed, n = 50)

collisions_of <- function(cw, key) cw |>
  mutate(nkey = norm_id(.data[[key]])) |>
  group_by(nkey) |>
  summarise(n_pid = dplyr::n_distinct(pid),
            pids  = paste(unique(pid), collapse = "; "), .groups = "drop") |>
  filter(n_pid > 1)
collisions_bc  <- collisions_of(cw_barcode_pid, "barcode")
collisions_img <- collisions_of(cw_image_pid,   "image")
cat("\nAmbiguous keys REFUSED (would risk the wrong patient) — barcode:",
    nrow(collisions_bc), "| image:", nrow(collisions_img), "\n")
if (nrow(collisions_bc))  { cat("-- barcode --\n"); print(collisions_bc,  n = 50) }
if (nrow(collisions_img)) { cat("-- image --\n");   print(collisions_img, n = 50) }
```

## Coverage per dataset

Per dataset: native IDs held, mapped, and lost, plus distinct patients reached in a
separate column because it is a different unit. A patient with three samples is three IDs
and one patient, so the two columns must not be read against each other.

```{r loss-table}
loss <- tibble(
  dataset         = names(mapres),
  n_native_ids    = vapply(mapres, `[[`, integer(1), "n_ids"),
  n_ids_mapped    = vapply(mapres, function(x) length(x$mapped_ids), integer(1)),
  n_ids_unmapped  = vapply(mapres, function(x) length(x$unmapped),  integer(1)),
  n_distinct_pids = vapply(mapres, function(x) length(x$mapped),     integer(1))
)
loss

cat("Patients present in ALL datasets:", length(Reduce(intersect, sets)), "\n")
cat("Patients present in ANY dataset:", length(Reduce(union, sets)), "\n")
```

The same table as a figure. Each bar is one dataset's **native IDs**, split into mapped
(green) and unmapped (orange) — both in the same unit, so the bar is genuinely additive
and its full length is the dataset's raw size. Orange is where linkage is failing.

```{r plot-loss}
loss_long <- loss |>
  transmute(dataset, mapped = n_ids_mapped, unmapped = n_ids_unmapped) |>
  pivot_longer(-dataset, names_to = "status", values_to = "n")

ggplot(loss_long, aes(reorder(dataset, n), n, fill = status)) +
  geom_col() + coord_flip() +
  scale_fill_manual(values = c(mapped = "#1D9E75", unmapped = "#D85A30")) +
  labs(x = NULL, y = "native IDs", fill = NULL,
       title    = "Per-dataset coverage and ID-conversion loss",
       subtitle = "Native-ID units; mapped + unmapped is additive") +
  theme_minimal(base_size = 12)
```

## Which patients have which modalities

Patients per combination of datasets, from the `sets` built in setup by `build_sets()`:
horizontal bars are each dataset's patient count, the dot matrix names a combination, and
the top bars count it. The intersections are **exclusive** — a column with dots on only
`{imaging, ihc}` is patients with imaging and IHC and nothing else — so every patient falls
in exactly one column and the top bars sum to the distinct-patient total.

```{r plot-upset, fig.width=10, fig.height=6}
upset(fromList(sets),
      nsets = length(sets), nintersects = NA,
      order.by = "freq",
      mainbar.y.label = "patients in this EXACT combination",
      sets.x.label    = "patients per dataset",
      text.scale = c(1.3, 1.2, 1.1, 1.0, 1.2, 1.1))
```

UpSet answers *which* combination; this answers *how complete*. For each **k**, the
number of patients appearing in exactly *k* of the eight datasets. The right-most bar is
the fully complete cohort and the left-most are single-modality patients — usually the
number people actually want.

```{r plot-degree}
all_pids   <- Reduce(union, sets)
membership <- sapply(sets, function(s) all_pids %in% s)
degree_tbl <- tibble(pid = all_pids, k = rowSums(membership))

ggplot(degree_tbl, aes(factor(k))) +
  geom_bar(fill = "#4C72B0") +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.3, size = 3.5) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(title = "Cohort completeness",
       x = "number of datasets a patient appears in (k)", y = "patients") +
  theme_minimal(base_size = 12)
```

The per-patient view of the same information, read off the master's `in_<dataset>` flags.
One row per patient, ordered by how many modalities they have (least complete at the
bottom); blue = present, beige = absent. Vertical beige stripes are the bottleneck
modalities; the solid blue block at the top is the complete cohort.

```{r plot-availability, fig.width=8, fig.height=7}
presence <- master |>
  select(pid, starts_with("in_")) |>
  pivot_longer(-pid, names_to = "dataset", values_to = "present") |>
  mutate(dataset = sub("^in_", "", dataset))

pid_order <- presence |> group_by(pid) |> summarise(k = sum(present), .groups = "drop") |>
  arrange(k) |> pull(pid)

ggplot(presence, aes(dataset, factor(pid, levels = pid_order), fill = present)) +
  geom_tile() +
  scale_fill_manual(values = c(`TRUE` = "#185FA5", `FALSE` = "#F1EFE8")) +
  labs(title = "Patient x dataset availability", x = NULL, y = "patients", fill = "present") +
  theme_minimal(base_size = 11) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1), panel.grid = element_blank())
```

### Two datasets that are loaded but never joined

`clinical_data` and `conversion_data` are saved but are not in `tables`, so neither is
harmonised into the master — only `gianlu_clinical_data` supplies clinical fields. This
checks whether `clinical_data`'s IDs are even in `pid` space, which decides how much work
joining it would be.

```{r clinical-overlap}
clin_ids <- unique(as.character(clinical_data$ID))
cat("clinical_data IDs that are in the pid union:",
    sum(clin_ids %in% Reduce(union, sets)), "of", length(clin_ids), "\n")
```

## Are the joined values sane?

A patient can be flagged `in_tmb == TRUE` and still have a missing or impossible TMB
value, so presence and usability are checked separately.

Fraction of missing values per column of the master, worst first. A payload column near
1.0 is either an unset placeholder name or a modality that mostly failed to join.

```{r master-missingness}
col_missing <- master |>
  summarise(across(everything(), ~ mean(is.na(.)))) |>
  pivot_longer(everything(), names_to = "column", values_to = "frac_NA") |>
  arrange(desc(frac_NA))

print(col_missing, n = Inf)
```

The same picture with a ranked plot (`naniar`) and a typed per-column summary (`skimr`).
Both optional.

```{r eda-naniar-skimr}
if (requireNamespace("naniar", quietly = TRUE)) {
  print(naniar::miss_var_summary(master), n = 25)
  print(naniar::gg_miss_var(master) +
          ggplot2::labs(title = "Missingness per column (master)"))
} else message("naniar not installed — skipping missingness summary/plot.")

if (requireNamespace("skimr", quietly = TRUE)) skimr::skim(master) else
  message("skimr not installed — skipping typed EDA summary.")
```

Range sanity on the four genomic score tables (`data/`-side, read back from the
intermediates), computed *before* collapsing so a bad input value is not masked by
averaging. Per numeric column: missing/zero/negative rates, observed range, distinct-value
count, and a robust outlier count at median ± 5·MAD. The MAF is excluded — its numeric
columns are genomic coordinates, where "range" means nothing.

```{r value-audit}
score_tables <- list(tmb = tmb_data, msi = msi_data, hrd = hrd_data, aneu = aneu_data)

value_audit <- imap(score_tables, ~ numeric_audit(.x, .y)) |> bind_rows()
print(value_audit, n = Inf)
```

Optional **hard domain bounds** — anything outside `[lower, upper]` is flagged. Left
empty so it runs as-is; the commented rows are a template with plausible ranges to
confirm against the real definitions.

```{r value-bounds}
score_bounds <- tribble(
  ~dataset, ~column,       ~lower, ~upper,
  # "tmb",  "TMB",          0,      Inf,   # mutations/Mb, non-negative
  # "msi",  "msi_score",    0,      1,     # fraction of unstable loci
  # "hrd",  "HRD_score",    0,      100,   # composite HRD score
  # "aneu", "aneuploidy",   0,      1,     # fraction of genome altered
)

if (nrow(score_bounds) > 0) {
  pmap_dfr(score_bounds, function(dataset, column, lower, upper) {
    df <- score_tables[[dataset]]
    if (is.null(df) || !column %in% names(df))
      return(tibble(dataset, column, status = "column not found",
                    n_violations = NA_integer_))
    x <- suppressWarnings(as.numeric(df[[column]]))
    tibble(dataset, column, status = "checked",
           n_violations = sum(x < lower | x > upper, na.rm = TRUE))
  })
} else {
  message("score_bounds empty — fill in column names to enable hard-bound checks.")
}
```

One histogram per numeric score column of the master, so these are the post-collapse
values every downstream report sees. Look for impossible values, a spike at zero (a failed
measurement coded as 0 rather than missing), and bimodality, which the mean collapse would
blur.

```{r plot-score-dist, fig.width=9, fig.height=5}
score_long <- master |>
  select(pid, matches("^(tmb|msi|hrd|aneu)__")) |>
  select(pid, where(is.numeric)) |>
  pivot_longer(-pid, names_to = "metric", values_to = "value") |>
  filter(!is.na(value))

if (nrow(score_long) > 0) {
  ggplot(score_long, aes(value)) +
    geom_histogram(bins = 30, fill = "#4C72B0") +
    facet_wrap(~ metric, scales = "free", ncol = 3) +
    labs(title = "Distributions of the genomic score columns (master)",
         subtitle = "Check for impossible values, spikes at 0, and bimodality",
         x = NULL, y = "patients") +
    theme_minimal(base_size = 11)
} else {
  message("No numeric score columns matched in master — check column prefixes.")
}
```

The continuous WES MSI score (`msi__MSI_SCORE`, percent unstable sites, from
`data/msi.20250831.csv`) against the clinical MSI status the clinical table calls
independently. The two arrive through different ID spaces — score through the barcode
crosswalk, status through the patient id — so clean separation is positive evidence that
the barcode-to-pid mapping is right, and overlap points at a linkage fault rather than
biology. Wilcoxon rank-sum on the two groups.

```{r msi-score-vs-status}
msi_score_col <- "msi__MSI_SCORE"   # continuous WES score; categorical = attend_cols$msi_status

if (have_cols(master, msi_score_col, attend_cols$msi_status)) {
  msi_df <- master |>
    transmute(pid,
              status = .data[[attend_cols$msi_status]],
              score  = suppressWarnings(as.numeric(.data[[msi_score_col]]))) |>
    filter(!is.na(status), !is.na(score))

  p_lab <- tryCatch({
    grp <- split(msi_df$score, msi_df$status)
    if (length(grp) == 2)
      sprintf("Wilcoxon p = %s",
              format.pval(wilcox.test(grp[[1]], grp[[2]])$p.value, digits = 2, eps = 1e-4))
    else NULL
  }, error = function(e) NULL)

  ggplot(msi_df, aes(status, score)) +
    geom_boxplot(outlier.shape = NA, fill = "#A6CEE3", width = 0.6) +
    highlight_points(msi_df, "status", "score", base_size = 0.7, base_alpha = 0.4, base_const = "black") +
    labs(title = "WES MSI score by clinical MSI status",
         subtitle = p_lab, x = "clinical MSI status",
         y = "WES MSI score (% unstable sites)") +
    theme_minimal(base_size = 11)
} else {
  message("Both msi__MSI_SCORE and ", attend_cols$msi_status,
          " must be in the master — skipping MSI concordance plot.")
}
```

```{r session-info}
sessionInfo()
```
