---
title: "ATTEND — 12. Aneuploidy, immune infiltrate and PD-L1"
author: "sceriff0"
date: "2026-09-16"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
# renv: tidyverse, here, fs, data.table (+ optional bioc::ComplexHeatmap, circlize for the
# Fig-1a-style heatmap). If missing: renv::install(c("bioc::ComplexHeatmap", "circlize"));
# renv::snapshot(). Without them the heatmap section says so and every other figure runs.
library(tidyverse)
library(here)
# Headless-safe raster device. The cluster's compute nodes have no X11, so R's default
# png() fails with "unable to open connection to X11 display" and every figure dies. cairo
# needs no display; the guard keeps this working on a machine built without it.
if (capabilities("cairo")) options(bitmapType = "cairo")

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 6)

source(here("code", "attend_harmonise.R"))   # make_id_cfg(), build_barcode_pid()
source(here("code", "attend_classes.R"))     # add_aneuploidy_cuts(), add_scna_group()
source(here("code", "attend_plots.R"))       # fig1a_heatmap(), feature_positions(), attend_box()
source(here("code", "attend_io.R"))          # read_intermediate()
source(here("code", "build_master.R"))       # get_master() — builds on demand
source(here("code", "load_clinical.R"))      # load_gianlu_clinical_data(), load_imaging_data()
source(here("code", "load_wes_results.R"))   # load_gistic_continuous()

master <- get_master() |> add_molecular_classes() |> add_scna_group()

# The same patients once per aneuploidy cut, with aneuploidy_class and scna_group re-derived
# at each. Every cut-dependent figure below facets on as_cut_lab rather than being written
# twice — see add_aneuploidy_cuts() for why that is a helper and not two call sites.
master_cuts <- add_aneuploidy_cuts(master, scna = TRUE)
cuts        <- aneu_cuts()
```

> **Question.** How is copy number laid out across the four MMR × aneuploidy groups, does aneuploidy track the pathologist's immune infiltrate and PD-L1 calls, and does either answer depend on where the aneuploidy cut is put?
> **Cohort.** All `r nrow(master)` patients on the master, with MMR-deficient patients beside them in the PD-L1 panels; the heatmap needs GISTIC output and the immune panels need a pathologist imaging call — each figure drops only its own missing values and prints how many.
> **Reads.** GISTIC all_data_by_genes.txt in data/gistic/ (runbooks/gistic.md), the "Imaging (2)" sheet of data/attend_data.xlsx, and attend_master_joined.
> **Out of scope.** Data-driven copy-number clusters -> [07-molecular-classification](07-molecular-classification.html). Which peaks recur in each group -> [10-recurrent-cna](10-recurrent-cna.html). FlowPath composition by aneuploidy, MMRd only -> [04-aneuploidy-mmrd](04-aneuploidy-mmrd.html). The aneuploidy classes, the two cuts and why both are reported -> [00-methods](00-methods.html).

## Data

The four groups are add_scna_group(): MMR status from MMR IHC (is_mmrd(), MSI not a
criterion) crossed with aneuploidy class. A patient missing either input has no group.
**Every aneuploidy-keyed figure in this report is drawn at both cuts**,
`r paste(cuts, collapse = " and ")` — attend_thresholds$aneuploidy and $aneuploidy_alt. The
first is the pipeline's primary and the one the master's own aneuploidy_class carries; the
second had never been measured against it, which is why both are shown rather than one
replacing the other. The table below is the cost of the choice in patients.

```{r data}
cw <- tryCatch(
  build_barcode_pid(load_gianlu_clinical_data(), make_id_cfg()),
  error = function(e) { note_skip("crosswalk unavailable: ", conditionMessage(e)); NULL }
)

master_cuts |>
  count(as_cut_lab,
        scna_group = replace_na(as.character(scna_group), "no group (MMR or AS missing)")) |>
  pivot_wider(names_from = as_cut_lab, values_from = n, values_fill = 0)
```

Which patients change class between the cuts, and in which direction. A patient counted off
the diagonal is AS High at `r min(cuts)` and AS Low at `r max(cuts)`; nothing can move the
other way, because raising a cut can only demote. Where that count is a large fraction of
the AS High arm, the two halves of every figure below are not two views of one result.

```{r cut-movers}
lo_lab <- paste0("AS cut ", min(cuts)); hi_lab <- paste0("AS cut ", max(cuts))

master_cuts |>
  filter(!is.na(aneuploidy_class)) |>
  select(pid, as_cut, aneuploidy_class) |>
  mutate(as_cut = ifelse(as_cut == min(cuts), "lo", "hi")) |>
  pivot_wider(names_from = as_cut, values_from = aneuploidy_class) |>
  transmute(!!lo_lab := as_label(lo), !!hi_lab := as_label(hi)) |>
  count(.data[[lo_lab]], .data[[hi_lab]], name = "patients")
```

## Part 1 — copy number across the four MMR × aneuploidy groups

The layout of Kandoth et al. Fig. 1a — tumours as columns, genes as rows in chromosomal
order, red gain and blue loss — with one change: the column blocks are **not** copy-number
clusters but the four pre-defined groups, left to right MMRp AS Low, MMRp AS High, MMRd AS
Low, MMRd AS High. The body is GISTIC's all_data_by_genes.txt (genome-wide, continuous, not
peak-restricted; data/gistic/, from the DRAGEN .seg), the same display matrix report 07
draws. Nothing is clustered into groups: within a block, tumours are only **ordered** by the
pipeline's configured pair (`r cnv_dist` distance, `r cnv_method` linkage) so similar
profiles sit together. One heatmap per cut — same tumours, same matrix, only the column
split and the aneuploidy annotation bar move. The cut is passed in explicitly (aneu_cut =),
because the bar derives its own high/low split and would otherwise take the configured
primary cut in both figures and contradict the blocks beneath it. Since aneuploidy class
defines the blocks and comes from arm-level ASCETS calls on these same tumours, the AS High
blocks looking more altered is true by construction and is not evidence of anything; what
the figure shows is **where** on the genome each group's burden sits. Colour limits are
±98th percentile of |value|.

```{r scna-group-heatmap, fig.width=11, fig.height=9}
disp_mat <- tryCatch(load_gistic_continuous(), error = function(e) NULL)

if (is.null(disp_mat)) {
  note_skip("No GISTIC all_data_by_genes.txt under data/gistic/ — run GISTIC per ",
            "runbooks/gistic.md. The heatmaps are skipped; Part 2 does not need them.")
} else if (is.null(cw)) {
  note_skip("No barcode -> pid crosswalk, so GISTIC columns cannot be given a group.")
} else {
  hm0 <- as.matrix(disp_mat[, setdiff(names(disp_mat), "ID"), drop = FALSE])
  rownames(hm0) <- disp_mat$ID

  pid_v <- cw$pid[match(rownames(hm0), cw$barcode)]
  fpos  <- attr(disp_mat, "feature_pos")
  cyto  <- if (!is.null(fpos)) setNames(fpos$cytoband, fpos$feature) else NULL

  cat("ComplexHeatmap installed:", requireNamespace("ComplexHeatmap", quietly = TRUE),
      "| circlize installed:", requireNamespace("circlize", quietly = TRUE), "\n")

  for (k in cuts) {
    mc   <- master_cuts |> filter(as_cut == k)
    grab <- function(col) mc[[col]][match(pid_v, mc$pid)]   # keyed on the matrix's barcodes
    grp  <- factor(as.character(grab("scna_group")), levels = attend_scna$group_levels)
    names(grp) <- rownames(hm0)

    keep <- !is.na(grp)
    cat("\n--- AS cut", k, "---\nGISTIC samples:", nrow(hm0), "| with a group:", sum(keep),
        "| dropped (no pid, or MMR / AS missing):", sum(!keep), "\n")
    print(table(group = grp[keep]))
    hm  <- hm0[keep, , drop = FALSE]
    grp <- droplevels(grp[keep])

    # The same covariate bars report 07 draws, built from the master so they agree with the
    # blocks. The binary AS bar is derived inside .fig1a_top_annotation() at the cut passed
    # below, not at the configured primary one.
    ids  <- rownames(hm)
    mmrd <- is_mmrd(grab("MMR_class")[keep])
    ann  <- data.frame(
      MMR = factor(ifelse(is.na(mmrd), NA, ifelse(mmrd, "MMRd", "MMRp")), levels = c("MMRp", "MMRd")),
      Aneuploidy = suppressWarnings(as.numeric(grab(attend_cols$aneuploidy)[keep])),
      row.names = ids)
    tp <- grab("TP53_pathogenic")[keep]
    if (!all(is.na(tp)))
      ann$TP53 <- factor(ifelse(as.logical(tp), "mut", "wt"), levels = c("wt", "mut"))

    feat_pos <- feature_positions(colnames(hm), cytoband = cyto)
    cat("genes with a position:", sum(feat_pos$feature %in% colnames(hm)), "\n")

    if (nrow(hm) >= 2 && nlevels(grp) >= 1) {
      fig1a_heatmap(hm, feat_pos, grp, ann_col = ann, value_type = "continuous",
                    cluster_cols = attend_scna_cols, cluster_name = "MMR x AS group",
                    column_distance = cnv_dist, column_linkage = cnv_method,
                    aneu_cut = k,
                    main = paste0("SCNAs by tumour x chromosomal location; columns split by ",
                                  "MMR x AS group — AS cut ", k))
    } else {
      note_skip("AS cut ", k, ": fewer than two GISTIC samples carry a group; nothing to draw.")
    }
  }
}
```

The per-group GISTIC runs on disk (data/gistic/mmrd_high/ and its siblings) were prepared
from the **primary** cut. Nothing here reads them — both heatmaps are built from the pooled
genome-wide matrix and split in memory — but [10-recurrent-cna](10-recurrent-cna.html) does,
so its peaks remain a `r min(cuts)`-cut result until those runs are regenerated.

## Part 2 — aneuploidy against the pathologist's immune calls

All three measures come from the "Imaging (2)" sheet of data/attend_data.xlsx, read by
load_imaging_data(), one row per image, keyed by pid. **IC %** is *Immune infiltrating cells
(%)*, continuous, averaged over a patient's images. The two PD-L1 measures are **banded**
calls, not percentages: *Percentage of tumor-infiltrating immune cells (IC) expressing PD-L1*
(IC0 < 1%, IC1 1–5%, IC2 5–10%, …) and *Percentage of tumor cells (TC) expressing PD-L1*
(TC0, TC1, …). They are ordered by the digit after the prefix, a blank cell is **not
measured** (NA, never band 0), and a patient's first non-missing band across images is
used — the same rule reports 04 and 05 apply. Aneuploidy score, class (at both cuts) and MMR
group are joined from the master.

```{r imaging}
first_nonNA <- function(x) { x <- x[!is.na(x) & x != ""]; if (length(x)) x[1] else NA }

# Order banded calls by the number after the letters: IC0 < IC1 < IC2 < IC3. Alphabetical
# order happens to agree up to 9 bands, but an explicit key keeps an unexpected spelling
# from being slotted in by string comparison.
band_factor <- function(x) {
  x   <- as.character(x)
  lv  <- unique(stats::na.omit(x))
  key <- suppressWarnings(as.integer(sub("^[A-Za-z]+\\s*([0-9]+).*$", "\\1", lv)))
  factor(x, levels = lv[order(key, lv, na.last = TRUE)])
}

imaging_raw <- tryCatch(load_imaging_data(), error = function(e) {
  note_skip("imaging sheet unavailable: ", conditionMessage(e)); NULL })

imm <- NULL
if (!is.null(imaging_raw)) {
  imm_pt <- imaging_raw |>
    filter(!is.na(ID)) |>
    group_by(pid = as.character(ID)) |>
    summarise(
      IC_pct   = mean(suppressWarnings(as.numeric(Immune_infiltrate)), na.rm = TRUE),
      PDL1_IC  = first_nonNA(as.character(Immune_infiltrate_PDL1)),
      PDL1_TC  = first_nonNA(as.character(Tumor_cells_PDL1)),
      .groups = "drop") |>
    mutate(IC_pct  = ifelse(is.finite(IC_pct), IC_pct, NA_real_),
           PDL1_IC = band_factor(PDL1_IC),
           PDL1_TC = band_factor(PDL1_TC))

  # Joined to the LONG master, so every patient appears once per cut carrying the class that
  # cut gives them. The score itself is cut-independent and identical in both blocks.
  imm <- imm_pt |>
    inner_join(master_cuts |>
                 transmute(pid, as_cut, as_cut_lab, aneuploidy_class, mmr_group,
                           aneuploidy_score = suppressWarnings(as.numeric(.data[[attend_cols$aneuploidy]]))),
               by = "pid") |>
    filter(!is.na(aneuploidy_score))

  n_pt <- n_distinct(imm$pid)
  cat("Patients on the master with an aneuploidy score and an imaging row:", n_pt,
      "| MMR-deficient among them:",
      n_distinct(imm$pid[!is.na(imm$mmr_group) & imm$mmr_group == "MMRd"]), "\n")
  one <- imm |> filter(as_cut == min(cuts))
  print(tibble(measure  = c("IC %", "PD-L1 on immune cells (IC band)", "PD-L1 on tumour cells (TC band)"),
               measured = c(sum(!is.na(one$IC_pct)), sum(!is.na(one$PDL1_IC)), sum(!is.na(one$PDL1_TC))),
               not_measured = n_pt - measured))
}
have_imm <- !is.null(imm) && n_distinct(imm$pid) > 2

# All patients and the MMRd subset as one factor, so a figure shows them side by side rather
# than being written twice — the shape report 05 uses for responder status. MMRd rows are a
# SUBSET of the all-patient rows, not an independent comparison.
with_cohorts <- function(d) {
  bind_rows(d |> mutate(cohort = "All patients"),
            d |> filter(!is.na(mmr_group), mmr_group == "MMRd") |> mutate(cohort = "MMRd only")) |>
    mutate(cohort = factor(cohort, levels = c("All patients", "MMRd only")))
}
```

### Immune infiltrate by aneuploidy class

IC % by AS class: columns are all patients and then each MMR group, rows are the two cuts.
The within-MMR columns are there because MMR-deficient tumours are both more infiltrated and
less often AS High, so a pooled difference can be MMR status seen twice; a difference that
survives inside both strata cannot be. Wilcoxon rank-sum per panel, uncorrected, tabulated
below the figure with each panel's n — read the two rows against each other before reading
either alone, since they are the same patients partitioned two ways.

```{r ic-by-as-class, fig.width=9, fig.height=7}
if (have_imm && sum(!is.na(imm$IC_pct)) > 2) {
  icd <- bind_rows(
    imm |> mutate(panel = "All patients"),
    imm |> filter(!is.na(mmr_group)) |> mutate(panel = as.character(mmr_group))) |>
    filter(!is.na(IC_pct), !is.na(aneuploidy_class)) |>
    mutate(panel = factor(panel, levels = c("All patients", "MMRp", "MMRd")))

  print(
    ggplot(icd, aes(aneuploidy_class, IC_pct, fill = aneuploidy_class)) +
      attend_box(icd, x = "aneuploidy_class", y = "IC_pct", by = c("panel", "as_cut_lab"),
                 fill = NULL, points = FALSE) +
      highlight_points(icd, "aneuploidy_class", "IC_pct", base_size = 1.2, base_alpha = 0.55,
                       base_const = "black") +
      facet_grid(as_cut_lab ~ panel) +
      scale_x_discrete(labels = as_label) +
      scale_fill_manual(values = attend_aneu_cols, labels = as_label, name = attend_as_legend) +
      labs(title = "Immune infiltrating cells by aneuploidy class, at both cuts",
           x = NULL, y = "Immune infiltrating cells, IC (%)") +
      attend_theme() + theme(legend.position = "none"))

  icd |>
    group_by(as_cut_lab, panel) |>
    summarise(
      n_AS_low  = sum(grepl("low",  aneuploidy_class)),
      n_AS_high = sum(grepl("high", aneuploidy_class)),
      wilcox_p  = if (n_AS_low >= 2 && n_AS_high >= 2)
        signif(wilcox.test(IC_pct[grepl("low",  aneuploidy_class)],
                           IC_pct[grepl("high", aneuploidy_class)], exact = FALSE)$p.value, 3)
        else NA_real_,
      .groups = "drop")
} else {
  note_skip("Fewer than three patients have both an IC % and an aneuploidy score.")
}
```

### Immune infiltrate against the continuous aneuploidy score

The same two measures without the class cut: aneu__aneuploidy_score on x against IC % on y,
one point per patient, all patients and MMRd alone. This panel is what lets either cut be
judged at all — it carries no class, so the two dashed lines
(`r paste(cuts, collapse = " and ")`) are the only thing the choice changes, and a reader can
see whether either falls anywhere the data suggests a break. Spearman's rho, because both
variables are bounded and skewed and IC % is recorded in coarse steps, which makes ties
common.

```{r ic-vs-as-score, fig.width=9, fig.height=4.5}
if (have_imm && sum(!is.na(imm$IC_pct)) > 2) {
  ics <- imm |> filter(as_cut == min(cuts), !is.na(IC_pct)) |> with_cohorts()

  rho_lab <- ics |>
    group_by(cohort) |>
    summarise(n = n(),
              rho = unname(suppressWarnings(cor.test(aneuploidy_score, IC_pct,
                     method = "spearman", exact = FALSE))$estimate),
              p   = suppressWarnings(cor.test(aneuploidy_score, IC_pct,
                     method = "spearman", exact = FALSE))$p.value,
              .groups = "drop") |>
    mutate(strip = sprintf("%s\nrho = %.2f, p = %s, n = %d", cohort, rho,
                           format.pval(p, digits = 2), n))
  ics <- ics |> left_join(rho_lab |> select(cohort, strip), by = "cohort")

  ggplot(ics, aes(aneuploidy_score, IC_pct)) +
    geom_vline(xintercept = cuts, linetype = 2, colour = "grey40") +
    geom_point(size = 1.8, alpha = 0.7, colour = "grey25") +
    facet_wrap(~ strip, nrow = 1) +
    labs(title = "Immune infiltrate against aneuploidy score",
         subtitle = paste("Dashed lines: the two AS cuts,", paste(cuts, collapse = " and ")),
         x = attend_as_axis, y = "Immune infiltrating cells, IC (%)") +
    attend_theme()
} else {
  note_skip("Fewer than three patients have both an IC % and an aneuploidy score.")
}
```

### Aneuploidy score across the PD-L1 bands

The continuous aneuploidy score (y) within each pathologist PD-L1 band (x), for each marker
on all patients and on MMR-deficient patients alone. Like the scatter above, this panel is
**cut-independent** — the score is plotted raw — so the dashed lines are the only trace of
the threshold, and a band whose spread straddles them is one whose class the choice decides.
Because the bands are ordered, the test is a **trend**, not a difference between unordered
groups: Spearman's rho between band rank (0, 1, 2, …) and the score, printed in each panel's
strip. A band holding fewer than five patients is drawn as a median bar rather than a box.

```{r as-by-pdl1-band, fig.width=11, fig.height=7}
pdl1_long <- NULL
if (have_imm) {
  one_cut <- imm |> filter(as_cut == min(cuts))
  pdl1_long <- bind_rows(
    one_cut |> transmute(pid, aneuploidy_score, mmr_group, band = as.character(PDL1_IC),
                         rank = as.integer(PDL1_IC) - 1L, marker = "PD-L1 immune cells (IC)"),
    one_cut |> transmute(pid, aneuploidy_score, mmr_group, band = as.character(PDL1_TC),
                         rank = as.integer(PDL1_TC) - 1L, marker = "PD-L1 tumour cells (TC)")) |>
    filter(!is.na(band))
}

if (!is.null(pdl1_long) && nrow(pdl1_long) > 2) {
  # Band order within each marker, carried into one factor so the x axis reads IC0..ICn then
  # TC0..TCn; free_x drops the other marker's bands from each panel.
  pdl1_long$band <- factor(pdl1_long$band,
                           levels = unique(pdl1_long$band[order(pdl1_long$marker, pdl1_long$rank)]))
  pb <- with_cohorts(pdl1_long)

  trend <- pb |>
    group_by(cohort, marker) |>
    summarise(n = n(), n_bands = n_distinct(rank),
              rho = if (n_bands > 1) unname(suppressWarnings(
                cor.test(rank, aneuploidy_score, method = "spearman", exact = FALSE))$estimate) else NA_real_,
              p   = if (n_bands > 1) suppressWarnings(
                cor.test(rank, aneuploidy_score, method = "spearman", exact = FALSE))$p.value else NA_real_,
              .groups = "drop") |>
    mutate(strip = sprintf("%s — %s\nrho = %.2f, p = %s, n = %d", cohort, marker, rho,
                           format.pval(p, digits = 2), n))
  pb <- pb |> left_join(trend |> select(cohort, marker, strip), by = c("cohort", "marker"))

  ggplot(pb, aes(band, aneuploidy_score)) +
    attend_box(pb, x = "band", y = "aneuploidy_score", by = "strip", points = FALSE) +
    highlight_points(pb, "band", "aneuploidy_score", base_size = 1.2, base_alpha = 0.55,
                     base_const = "black") +
    geom_hline(yintercept = cuts, linetype = 2, colour = "grey40") +
    facet_wrap(~ strip, scales = "free_x", nrow = 2) +
    labs(title = "Aneuploidy score by pathologist PD-L1 band",
         subtitle = paste("Dashed lines: the two AS cuts,", paste(cuts, collapse = " and ")),
         x = "PD-L1 band (pathologist)", y = attend_as_axis) +
    attend_theme() +
    theme(axis.text.x = element_text(angle = 20, hjust = 1))
} else {
  note_skip("Fewer than three patients have a PD-L1 band and an aneuploidy score.")
}
```

### Aneuploidy class across the PD-L1 bands

Both variables as categories, and the one PD-L1 figure the cut does move: within each band,
the fraction of patients that are AS Low and AS High, at each cut (rows) for each cohort and
marker (columns). Bars are proportions, so a band of two patients and one of twenty look the
same height — the counts are in the table below, with Fisher's exact test of AS class × band
per panel (exact on the whole r × 2 table, no band merging). These panels are not independent
evidence: IC and TC bands are called on the same slides of the same patients, and the MMRd
columns are a subset of the all-patient ones.

```{r as-class-by-pdl1-band, fig.width=11, fig.height=7}
if (!is.null(pdl1_long) && nrow(pdl1_long) > 2) {
  # Re-joined to the LONG master so each band carries the class from BOTH cuts.
  pc <- pdl1_long |>
    select(pid, band, marker) |>
    # many-to-many DECLARED, not silenced by accident: two marker rows per patient meet two
    # cut rows per patient, and 2 x 2 is exactly the intended shape.
    inner_join(master_cuts |> select(pid, as_cut, as_cut_lab, aneuploidy_class, mmr_group),
               by = "pid", relationship = "many-to-many") |>
    filter(!is.na(aneuploidy_class)) |>
    with_cohorts()

  print(
    ggplot(pc, aes(band, fill = aneuploidy_class)) +
      geom_bar(position = "fill", width = 0.7) +
      facet_grid(as_cut_lab ~ cohort + marker, scales = "free_x") +
      scale_fill_manual(values = attend_aneu_cols, labels = as_label, name = attend_as_legend) +
      labs(title = "Aneuploidy class within each PD-L1 band, at both cuts",
           x = "PD-L1 band (pathologist)", y = "fraction of patients in the band") +
      attend_theme() +
      theme(axis.text.x = element_text(angle = 20, hjust = 1)))

  fisher_p <- pc |>
    group_by(as_cut_lab, cohort, marker) |>
    summarise(fisher_p = {
      tb <- table(droplevels(band), droplevels(aneuploidy_class))
      if (all(dim(tb) >= 2)) signif(fisher.test(tb)$p.value, 3) else NA_real_
    }, .groups = "drop")

  pc |>
    count(as_cut_lab, cohort, marker, band, aneuploidy_class) |>
    mutate(aneuploidy_class = as_label(aneuploidy_class)) |>
    pivot_wider(names_from = aneuploidy_class, values_from = n, values_fill = 0) |>
    left_join(fisher_p, by = c("as_cut_lab", "cohort", "marker"))
} else {
  note_skip("Fewer than three patients have a PD-L1 band and an aneuploidy class.")
}
```

```{r session-info}
sessionInfo()
```
