---
title: "ATTEND — 13. Recurrent copy number and immune-evasion gene function"
author: "sceriff0"
date: "2026-09-24"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
# renv: tidyverse, here, fs, data.table. No new package — the catalogue is a tracked TSV
# and the test is base-R resampling.
library(tidyverse)
library(here)
# Headless-safe raster device: the cluster's compute nodes have no X11, so the default
# png() dies with "unable to open connection to X11 display" and every figure with it.
if (capabilities("cairo")) options(bitmapType = "cairo")

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 6)

source(here("code", "attend_harmonise.R"))   # make_id_cfg(), build_barcode_pid()
source(here("code", "attend_classes.R"))     # attend_crispr, attend_cnv, add_scna_group()
source(here("code", "attend_scna.R"))        # run_peak_blocks(), crispr_peak_overlap()
source(here("code", "attend_plots.R"))       # note_skip(), attend_theme(), attend_pal()
source(here("code", "attend_io.R"))          # read_intermediate()
source(here("code", "build_master.R"))       # get_master() — builds on demand
source(here("code", "load_crispr.R"))        # load_crispr_screens(), crispr_counts()
source(here("code", "load_wes_results.R"))   # find_gistic_files(), load_gistic_continuous_at()

master <- tryCatch(get_master() |> add_molecular_classes() |> add_scna_group(),
                   error = function(e) NULL)
have_master <- !is.null(master) && nrow(master) > 0

crispr    <- load_crispr_screens()
have_cat  <- nrow(crispr) > 0

gx_dir   <- here("data", attend_cnv$gistic$dir)
run_dirs <- c(list(all = file.path(gx_dir, attend_cnv$gistic$pooled_dir)),
              setNames(as.list(file.path(gx_dir, scna_group_token(attend_scna$group_levels))),
                       scna_group_token(attend_scna$group_levels)))
have_gistic <- dir.exists(run_dirs[["mmrd_high"]])

# The genes GISTIC actually measured, in GENOMIC order. Both facts matter: the order is what
# the block-shift null slides peaks along, and the membership is half of the universe. GISTIC
# writes all_data_by_genes.txt in genomic order, and feature_pos preserves that row order.
gene_order <- local({
  g <- tryCatch(load_gistic_continuous_at(run_dirs[["mmrd_high"]], attend_cnv),
                error = function(e) NULL)
  fp <- if (is.null(g)) NULL else attr(g, "feature_pos")
  if (is.null(fp)) character(0) else as.character(fp$feature)
})
have_order <- length(gene_order) > 0
```

> **Question.** Are the loci GISTIC calls recurrent inside a stratum enriched for genes that CRISPR screens show control tumour-cell killing by T cells — losses of resisters, gains of sensitizers?
> **Cohort.** MMRd aneuploidy-high first, the target of [10-recurrent-cna](10-recurrent-cna.html)'s Question 1, then all four MMR × aneuploidy strata side by side. Peaks come from each stratum's own GISTIC run at the primary aneuploidy cut.
> **Reads.** data/crispr_t_cell_screens_wu2025.tsv (code/fetch_crispr_screens.R, from Wu et al. Table S3), the amp_genes/del_genes/all_data_by_genes files of each run in data/gistic/ (runbooks/gistic.md), and attend_master_joined.
> **Out of scope.** Whether a locus recurs at all, and whether the same loci recur in MMRp → [10-recurrent-cna](10-recurrent-cna.html). Immune infiltrate and PD-L1 as measured phenotypes → [12-aneuploidy-immune](12-aneuploidy-immune.html). The aneuploidy classes and the two cuts → [00-methods](00-methods.html).

## What this report does and does not claim

Wu et al. (*Immunity* 58:2864–2877, 2025) intersect recurrently altered genes with a pooled
CRISPR catalogue of T-cell-killing **resisters** and **sensitizers**, direction-matched. That
filter is what this report borrows. Three differences from their analysis are load-bearing and
are stated here rather than left for a reader to infer.

**Their recurrence is a within-patient subtraction; this one is not.** Wu et al. compare each
patient's progression tumour against that patient's own pre-treatment baseline and call a gene
recurrent when it changed in at least three patients. ATTEND has one tumour per patient and
`collapse_pid()` averages to one row per patient, so there is no baseline to subtract. Nothing
in this report is evidence about what a tumour acquired under treatment, and the phrase they
use for that axis is deliberately absent below.

**The two recurrence definitions are different constructs.** Theirs is a count with no
background model, which on their cohort returns about a quarter of the exome. This report's is
GISTIC's background-corrected q, `r attend_cnv$gistic$fdr_cutoff`, on the stratum's own run.
A count and a peak statistic are not comparable, and no number here is compared to one of theirs.

**The catalogue is melanoma- and ICI-derived.** It was assembled from screens of tumour-cell
killing by T cells, pooled across mouse and human models. Applying it to endometrial carcinoma
is an assumption this report makes, not a result it establishes.

## The catalogue

`load_crispr_screens()` reads the tracked TSV and applies the evidence cutoff as an argument,
`attend_crispr$min_measurements` = `r attend_crispr$min_measurements` — Wu et al.'s own
"concordant evidence in ≥ two independent measurements", which reproduces their reported 519
resisters and 877 sensitizers exactly. The unfiltered blocks stay in the file so the cutoff is
a choice that can be varied, not a property of the data. Symbols that qualify as **both** a
resister and a sensitizer carry no direction, so they are dropped and counted here rather than
resolved by whichever row a join kept.

```{r catalogue}
if (!have_cat) {
  note_skip("no ", attend_crispr$file, " under data/ — run `Rscript code/fetch_crispr_screens.R`.")
} else {
  print(knitr::kable(crispr_counts(crispr), row.names = FALSE,
                     caption = paste0("The CRISPR T-cell-killing catalogue at the configured ",
                                      "evidence cutoff (Wu et al., Table S3)")))
}
```

The direction map is the method in one line, and it is not symmetric: a **resister** matters
when it is **lost**, a **sensitizer** when it is **gained**. Both moves push a tumour towards
immune escape, and swapping either entry would invert every statement below while leaving all
the counts unchanged — which is why it is configuration (`attend_crispr$direction_of`) and not
an argument at a call site.

```{r direction-map}
print(knitr::kable(
  data.frame(role = names(attend_crispr$direction_of),
             `alteration that promotes escape` = unname(attend_crispr$direction_of),
             check.names = FALSE),
  row.names = FALSE, caption = "attend_crispr$direction_of — the whole method"))
```

## Part 1 — MMRd aneuploidy-high

### What was measured, and what could not be

The partition below is the denominator, printed before any test. Two of its lines exist
because an absence must not read as a negative result. A peak gene that the 23 pooled screens
never covered is **unscreened**, not "not a resister": the screens do not tile the genome, so
scoring an unscreened gene as a measured negative would inflate every p-value on the page.
`attend_crispr$universe` is therefore `r attend_crispr$universe` — the genes this GISTIC run
assayed **intersected with** the genes the catalogue screened. Genes sitting in both an
amplification and a deletion peak carry no direction either, and are dropped on the same rule
`gistic_feature_direction()` already applies.

```{r p1-partition}
ov1 <- NULL
if (!have_gistic) {
  note_skip("no MMRd-high GISTIC run under ", run_dirs[["mmrd_high"]],
            " — run `MODE=groups Rscript code/prep_gistic_group_segs.R`, then ",
            "`MODE=groups bash code/run_gistic.sh`.")
} else if (!have_cat) {
  note_skip("the catalogue is absent, so no overlap can be computed — see above.")
} else {
  ov1 <- crispr_peak_overlap(run_dirs[["mmrd_high"]], crispr = crispr,
                             assayed = if (have_order) gene_order else NULL)
  if (is.null(ov1$partition)) {
    note_skip("the MMRd-high run has no amp_genes/del_genes peak lists to read.")
  } else {
    print(knitr::kable(ov1$partition, row.names = FALSE,
                       caption = "MMRd aneuploidy-high: what entered the overlap, and what was excluded"))
  }
}
```

### The direction-matched hits

Every row is a gene that sits inside a significant peak of the MMRd-high run **and** is a
catalogue member **in the matching direction** — a resister inside a deletion peak, or a
sensitizer inside an amplification peak. `n_measurements` is the screen support behind the
functional call, from the catalogue, and carries no information about this cohort.
`wu_overlap` is Wu et al.'s **own** result on their melanoma patients, carried for provenance:
it says the gene was in their recurrent set, and says nothing whatever about ATTEND. It must
never be joined on.

```{r p1-hits}
if (is.null(ov1) || is.null(ov1$hits)) {
  note_skip("no direction-matched hits to list — see the partition above for why.")
} else {
  h <- ov1$hits[order(ov1$hits$role, -ov1$hits$n_measurements, ov1$hits$gene), ]
  cat("**", nrow(h), " direction-matched gene(s)**: ",
      sum(h$role == "resister"), " resister(s) inside deletion peaks, ",
      sum(h$role == "sensitizer"), " sensitizer(s) inside amplification peaks.\n\n", sep = "")
  print(knitr::kable(utils::head(h, 40), row.names = FALSE, digits = 3,
                     caption = "MMRd aneuploidy-high: peak genes with a matching functional call"))
}
```

### Is that more overlap than chance?

A count of hits is not a result until it is read against what chance produces, and the null
here has to respect two kinds of clustering at once. GISTIC peaks are contiguous blocks — an
amplicon spanning 300 genes is one event, not 300 independent successes — and the catalogue is
**also** positionally clustered, because immune genes sit in families: *TAP1*, *TAP2*, *PSMB8*
and *TAPBP* across the MHC region, the *IFNA* cluster on 9p21 immediately beside *CDKN2A*, a
classic deletion peak. So the peaks are slid to random positions along the genomically ordered
universe while the catalogue is held exactly where it is. Measured on simulated null data with
both kinds of clustering, a hypergeometric test runs at roughly five times its nominal type-I
rate and so does a test that holds the peaks and reshuffles the gene labels; the block shift is
calibrated. The three measured type-I rates, and why each alternative fails, are in
[00-methods](00-methods.html#why-the-peaks-move-and-the-catalogue-stays-put); the calibration
that pins them is `code/tests/test_crispr_overlap.R`. This is **Family B**, corrected with
`r attend_scna$fdr_method`: the catalogue is external and fixed in advance, but the peaks come
from ATTEND's own group labels, so no result here can be promoted to a confirmatory one.

```{r p1-test}
t1 <- NULL
if (is.null(ov1) || is.null(ov1$hits)) {
  note_skip("no overlap to test.")
} else if (!have_order) {
  note_skip("the MMRd-high run has no all_data_by_genes.txt, so genes cannot be put in ",
            "genomic order and the block-shift null is not defined. No p-value is reported ",
            "rather than one computed on an arbitrary ordering.")
} else {
  blocks1 <- run_peak_blocks(run_dirs[["mmrd_high"]])
  t1 <- crispr_overlap_test(ov1, blocks1, gene_order = gene_order, crispr = crispr)
  if (is.null(t1)) {
    note_skip("too few genes survive the universe restriction for a resampling test.")
  } else {
    print(knitr::kable(t1, row.names = FALSE, digits = c(0, 0, 2, 4, 4, 0),
                       caption = paste0("MMRd aneuploidy-high: block-shift permutation, B = ",
                                        t1$B[1], ", Family B (", attend_scna$fdr_method, ")")))
  }
}
```

The observed and expected counts are the two numbers to read together: `expected` is the mean
of the null, so a hit count close to it is the answer "no more than the genome's layout already
predicts", whatever the catalogue membership looks like gene by gene. The combined row adds the
two directional rows and is a summary, not a third question, so it carries no adjusted p.

### How much of this rests on one patient

Report 10 reports, for each recurrent locus, the fraction of leave-one-out MMRd-high GISTIC
re-runs that retain it. That column qualifies this one too: the peak set is the input here, so
a functional enrichment computed on peaks that turn over when a single tumour is dropped
inherits exactly that instability. The table below re-reads the leave-one-out runs and reports
how many of them retain each gene's peak, so the enrichment above can be read against the
stability of the thing it was computed on rather than separately from it.

```{r p1-stability}
if (is.null(ov1) || is.null(ov1$hits)) {
  note_skip("no hits whose stability could be reported.")
} else {
  loo_dirs <- { d <- file.path(gx_dir, "loo")
                if (dir.exists(d)) list.dirs(d, full.names = TRUE, recursive = FALSE)
                else character(0) }
  if (!length(loo_dirs)) {
    note_skip("no leave-one-out runs under ", file.path(gx_dir, "loo"),
              " — run `MODE=loo Rscript code/prep_gistic_group_segs.R`, then ",
              "`MODE=loo bash code/run_gistic.sh`. Without them the hits above have no ",
              "stability column, and at this n that is the column that decides whether a ",
              "hit is a finding.")
  } else {
    keep <- vapply(loo_dirs, function(d) {
      pg <- tryCatch(run_peak_genes(d), error = function(e) character(0))
      if (!length(pg)) return(rep(NA_real_, nrow(ov1$hits)))
      as.numeric(ov1$hits$gene %in% names(pg)[pg == ov1$hits$peak_dir])
    }, numeric(nrow(ov1$hits)))
    st <- data.frame(gene = ov1$hits$gene, role = ov1$hits$role,
                     peak_dir = ov1$hits$peak_dir,
                     loo_retained_frac = rowMeans(matrix(keep, nrow = nrow(ov1$hits)),
                                                  na.rm = TRUE),
                     stringsAsFactors = FALSE)
    st <- st[order(-st$loo_retained_frac, st$gene), ]
    cat("Leave-one-out re-runs read: ", length(loo_dirs), ".\n\n", sep = "")
    print(knitr::kable(utils::head(st, 40), row.names = FALSE, digits = 3,
                       caption = paste0("Each hit's peak, across the leave-one-out MMRd-high ",
                                        "re-runs that retain it")))
  }
}
```

## Part 2 — the same question in all four strata

The four MMR × aneuploidy strata answered side by side, each on its **own** GISTIC run, so a
stratum with more peaks is not being compared against another stratum's peak list. The
universe is rebuilt per run for the same reason: a run that assayed fewer genes has a smaller
denominator, and sharing one denominator across runs would move hits between strata without
any of them changing. Strata whose GISTIC run is missing are listed as such rather than
dropped silently, because an absent row and a null row are different statements.

```{r p2-strata}
if (!have_cat) {
  note_skip("the catalogue is absent, so no stratum can be scored.")
} else {
  rows <- lapply(scna_group_token(attend_scna$group_levels), function(tok) {
    d <- run_dirs[[tok]]
    if (!dir.exists(d))
      return(data.frame(stratum = tok, status = "no GISTIC run", peak_genes = NA_integer_,
                        universe = NA_integer_, hits = NA_integer_, p_del = NA_real_,
                        p_amp = NA_real_, stringsAsFactors = FALSE))
    go <- local({
      g <- tryCatch(load_gistic_continuous_at(d, attend_cnv), error = function(e) NULL)
      fp <- if (is.null(g)) NULL else attr(g, "feature_pos")
      if (is.null(fp)) character(0) else as.character(fp$feature)
    })
    ov <- crispr_peak_overlap(d, crispr = crispr, assayed = if (length(go)) go else NULL)
    if (is.null(ov$partition))
      return(data.frame(stratum = tok, status = "no peak lists", peak_genes = NA_integer_,
                        universe = NA_integer_, hits = NA_integer_, p_del = NA_real_,
                        p_amp = NA_real_, stringsAsFactors = FALSE))
    tt <- if (length(go)) crispr_overlap_test(ov, run_peak_blocks(d), gene_order = go,
                                              crispr = crispr) else NULL
    data.frame(stratum = tok,
               status = if (is.null(tt)) "no genomic order — untested" else "tested",
               peak_genes = length(ov$peak_dir), universe = length(ov$universe),
               hits = if (is.null(ov$hits)) 0L else nrow(ov$hits),
               p_del = if (is.null(tt)) NA_real_ else tt$p[1],
               p_amp = if (is.null(tt)) NA_real_ else tt$p[2],
               stringsAsFactors = FALSE)
  })
  tab <- do.call(rbind, rows)
  print(knitr::kable(tab, row.names = FALSE, digits = 4,
                     caption = paste0("All four MMR x aneuploidy strata, each on its own ",
                                      "GISTIC run and its own universe")))
}
```

Each stratum's two p-values were corrected within that stratum, not across the four: the strata
are four separate questions asked of four different patient sets, and pooling them into one
family would penalise each for the others having been asked. Reading down a column is
descriptive comparison, not a test of difference between strata — that question is
[10-recurrent-cna](10-recurrent-cna.html)'s Question 2, which asks it of the loci themselves.

```{r session-info}
sessionInfo()
```
