---
title: "ATTEND — 31. Aneuploidy vs mutation status"
author: "sceriff0"
date: "2026-08-11"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
# renv: needs ggpubr (+ tidyverse, here). Mutation status also uses attend_harmonise.R.
# If missing: renv::install("ggpubr"); renv::snapshot()
library(tidyverse)
library(here)
library(ggpubr)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 10, fig.height = 6)

source(here("code", "attend_classes.R"))     # gene panels, add_molecular_classes(), is_mmrd()
source(here("code", "attend_harmonise.R"))   # build_barcode_pid(), make_id_cfg()
source(here("code", "attend_plots.R"))       # highlight_points(), attend_theme()
source(here("code", "attend_io.R"))          # read_intermediate()
source(here("code", "load_clinical.R"))      # load_gianlu_clinical_data()
source(here("code", "load_wes_results.R"))   # MAF loaders

master <- read_intermediate("attend_master_joined") |> add_molecular_classes()
n_mmrd <- sum(is_mmrd(master[[attend_cols$mmr_status]]), na.rm = TRUE)
```

> **Question.** Among tumours that are already mismatch-repair deficient, does aneuploidy vary with which gene is mutated?
> **Cohort.** MMR-deficient only (n = `r n_mmrd`).
> **Reads.** `attend_master_joined` for the aneuploidy score, and the long MAF for per-gene status.
> **Out of scope.** Aneuploidy against TMB -> [37_tmb_by_aneuploidy_and_mmr](37_tmb_by_aneuploidy_and_mmr.html), which draws both MMR strata instead of only this one. Cohort-wide mutation frequency -> [33_mutation_landscape](33_mutation_landscape.html).

The restriction is the point. MMR deficiency is itself a route to genomic instability, so
in the whole cohort any mutation that tracks MMR status will also appear to track
aneuploidy. Conditioning on MMRd removes that shortcut. The outcome variable in every
figure is the same continuous aneuploidy score; what changes is the grouping on the x-axis.

```{r child='_explainers/cohort-mmrd.Rmd'}
```

```{r child='_explainers/aneuploidy-classes.Rmd'}
```

```{r child='_explainers/highlight-groups.Rmd'}
```

### The gene panel, spelled out

Genes are tested one at a time, each in its own panel, so a signal is attributable to a
specific gene rather than to a union that could be driven by whichever gene is most
frequently mutated. Two groups, kept deliberately separate:

| Group | Genes | Why they are grouped this way |
|---|---|---|
| **MMR panel** | `r paste(attend_gene_panel, collapse = ", ")` | Loss of any of these causes mismatch-repair deficiency. MSH3 pairs with MSH2 to form MutSb. |
| **Wnt / b-catenin** | `r paste(attend_gene_panel_wnt, collapse = ", ")` | An independent oncogenic pathway, not an MMR mechanism. |
| **TP53** | TP53 | Handled separately: it is already a first-class column on the master. |

The MMR union (`panel-altered` = at least one pathogenic variant in any MMR panel gene) is
shown last, after the per-gene panels. `CTNNB1` and `APC` are deliberately excluded from
that union, because folding them in would change what "panel-altered" means everywhere else
in the pipeline. There is no Wnt union: with two genes the per-gene panels already show
everything a union would.

P-values come from `stat_compare_means()`, Wilcoxon rank-sum for two groups and
Kruskal-Wallis for more, uncorrected across panels. With one test per gene, treat them as
descriptive.

## Data

Two sources, joined on `pid`. The aneuploidy score and its derived classes come from
`attend_master_joined`. The mutation status is recomputed here directly from the long MAF
rather than taken from the master's collapsed columns, because the average of a variant row
is meaningless; `pathogenic_by_patient()` and `mutation_status_long()` are the same helpers
[10_data_integration](10_data_integration.html) used, so the two cannot drift. Crossing the
crosswalk against the full gene list guarantees every gene gets a column even when it never
appears in the MAF, with unmutated patients `FALSE` rather than `NA`.

```{r data}
cw  <- build_barcode_pid(load_gianlu_clinical_data(), make_id_cfg())
maf <- read_intermediate("maf_data")

mut <- pathogenic_by_patient(maf, cw)   # pid, TP53_pathogenic, panel_pathogenic (MMR union)

# Per-gene status for the one-gene-at-a-time panels: MMR panel + Wnt genes.
gene_status <- cw |> distinct(pid) |>
  tidyr::crossing(gene = attend_genes_per_gene) |>
  left_join(mutation_status_long(maf, cw), by = c("pid", "gene")) |>
  mutate(mutated = replace_na(mutated, FALSE),
         class   = if_else(mutated, paste0(gene, "-mut"), paste0(gene, "-wt"))) |>
  select(pid, gene, class) |>
  pivot_wider(names_from = gene, values_from = class, names_glue = "{gene}_class")

dat_all <- master |>
  select(-any_of(c("TP53_pathogenic", "panel_pathogenic"))) |>  # recompute here
  left_join(mut, by = "pid") |>
  left_join(gene_status, by = "pid") |>
  mutate(
    aneuploidy_value = suppressWarnings(as.numeric(.data[[attend_cols$aneuploidy]])),
    TP53_class  = if_else(TP53_pathogenic,  "TP53-mut",     "TP53-wt"),
    panel_class = if_else(panel_pathogenic, "MMR-panel-altered", "MMR-panel-wt")
  )
```
### Restricting to MMR-deficient

The single filter the whole report rests on. The counts below are the cohort every
figure that follows is drawn from — if the MMRd group is small, every p-value in this
report inherits that.

```{r restrict-mmrd}
have_mmr <- have_cols(dat_all, "MMR_class", "aneuploidy_value")

dat <- if (have_mmr) dat_all |> filter(MMR_class == attend_levels$mmr_deficient) else dat_all[0, ]

cat("Whole cohort              :", nrow(dat_all), "patients\n")
cat("MMR-deficient (this report):", nrow(dat), "patients\n")
if (nrow(dat) > 0)
  cat("  with an aneuploidy score :", sum(!is.na(dat$aneuploidy_value)), "patients\n")

if (have_mmr && nrow(dat) == 0)
  message("No MMR-deficient patients (MMR_class == '", attend_levels$mmr_deficient,
          "') — every figure below is skipped.")
```

Which of the manually highlighted patients actually resolve to a `pid` in this restricted
cohort. Printed so that a silent ID mismatch shows up as a number here rather than as
"the gold points just aren't drawn".

```{r highlight-coverage}
if (nrow(dat) > 0) cat(highlight_coverage(dat))
```

```{r box-helper, include=FALSE}
# Shared boxplot: aneuploidy score on y, one facet per grouping variable. Boxes are
# neutral; points are coloured by aneuploidy class, with the highlight overlay drawn
# on top (partitioned, so no patient is plotted twice). highlight_points() owns both
# colour scales, so no scale_colour_manual is added here.
aneu_box <- function(d, vars, title, subtitle = NULL, ncol = 3) {
  long <- d |>
    select(pid, aneuploidy_value, aneuploidy_class, any_of(vars)) |>
    pivot_longer(any_of(vars), names_to = "comparison", values_to = "group") |>
    filter(!is.na(aneuploidy_value), !is.na(group)) |>
    mutate(comparison = factor(comparison, levels = vars))
  if (nrow(long) == 0) return(NULL)

  ggplot(long, aes(group, aneuploidy_value)) +
    geom_boxplot(outlier.shape = NA, fill = "#A6CEE3", width = 0.6) +
    highlight_points(long, "group", "aneuploidy_value", base_colour = "aneuploidy_class",
                     base_palette = attend_aneu_cols, base_name = "aneuploidy",
                     base_size = 0.7, base_alpha = 0.6) +
    facet_wrap(~ comparison, scales = "free_x", ncol = ncol) +
    stat_compare_means(label = "p.format", size = 3) +
    labs(title = title, subtitle = subtitle, x = NULL, y = "aneuploidy score") +
    attend_theme() +
    theme(axis.text.x = element_text(angle = 20, hjust = 1))
}

# Group sizes behind each panel — a two-group Wilcoxon on 3 vs 40 patients looks
# identical to one on 20 vs 23 unless the counts are printed.
group_counts <- function(d, vars) {
  d |>
    select(pid, aneuploidy_value, any_of(vars)) |>
    pivot_longer(any_of(vars), names_to = "comparison", values_to = "group") |>
    filter(!is.na(group), !is.na(aneuploidy_value)) |>
    count(comparison, group, name = "n") |>
    pivot_wider(names_from = group, values_from = n, values_fill = 0)
}
```

## 1. Aneuploidy by gene — MMR panel, one gene at a time

Each panel is one MMR gene: patients carrying a pathogenic variant in that gene
(`<GENE>-mut`) versus those who do not (`<GENE>-wt`), within the MMRd cohort only.
Pathogenicity is decided by `attend_pathogenic_classes` on the MAF's
`Variant_Classification`.

Read the group sizes first. Some of these genes are mutated in very few patients, and a
panel with a handful of `-mut` patients will show a wide box and an unstable p-value.

```{r mmr-gene-counts}
mmr_gene_vars <- paste0(attend_gene_panel, "_class")
if (nrow(dat) > 0)
  knitr::kable(group_counts(dat, mmr_gene_vars),
               caption = "Patients per group, MMR panel genes (MMR-deficient cohort)")
```

Aneuploidy score (`aneu__aneuploidy_score`, ASCETS arm-level burden from `data/ascets/`)
in mutated versus wild-type patients, one panel per MMR gene, over the `dat` cohort above.

```{r mmr-gene-box, fig.height=7}
if (nrow(dat) > 0)
  aneu_box(dat, mmr_gene_vars,
           "Aneuploidy score by MMR panel gene (MMR-deficient only)",
           paste0("One panel per gene; pathogenic vs wild-type. n = ", nrow(dat),
                  " MMR-deficient patients"))
```

## 2. Aneuploidy by gene — Wnt / β-catenin pathway

The same comparison for `r paste(attend_gene_panel_wnt, collapse = " and ")`. These are
shown apart from the MMR panel because they are a different mechanism: they are not a
cause of mismatch-repair deficiency, so a signal here would mean something biologically
different from a signal in the MMR panel above.

```{r wnt-gene-counts}
wnt_gene_vars <- paste0(attend_gene_panel_wnt, "_class")
if (nrow(dat) > 0)
  knitr::kable(group_counts(dat, wnt_gene_vars),
               caption = "Patients per group, Wnt pathway genes (MMR-deficient cohort)")
```

The same aneuploidy score against Wnt-pathway mutation status, two panels.

```{r wnt-gene-box, fig.height=4.5}
if (nrow(dat) > 0)
  aneu_box(dat, wnt_gene_vars,
           "Aneuploidy score by Wnt pathway gene (MMR-deficient only)",
           "CTNNB1 and APC — shown separately from the MMR panel; not part of the MMR union",
           ncol = 2)
```

## 3. TP53 and the MMR union

`TP53` on its own, then the MMR panel **union** — at least one pathogenic variant in any
of `r paste(attend_gene_panel, collapse = ", ")`. The union is the last mutation view
rather than the first: it is the least specific, and the per-gene panels above are what
say which gene any union signal came from.

```{r tp53-union-counts}
summary_vars <- c("TP53_class", "panel_class")
if (nrow(dat) > 0)
  knitr::kable(group_counts(dat, summary_vars),
               caption = "Patients per group, TP53 and the MMR union (MMR-deficient cohort)")
```

Aneuploidy score by TP53 status and by the MMR union, side by side, so a union signal can
be read against the per-gene panels above that would have to explain it.

```{r tp53-union-box, fig.height=4.5}
if (nrow(dat) > 0)
  aneu_box(dat, summary_vars,
           "Aneuploidy score by TP53 status and by the MMR panel union (MMR-deficient only)",
           ncol = 2)
```

## 4. Aneuploidy by the other molecular classes

The non-mutation axes, still within MMRd: TMB, HRD, and MSI status. `MMR_class` is
omitted because it is constant here by construction — every patient in this report is
MMR-deficient.

MSI is worth reading carefully: MMR loss is the usual *cause* of microsatellite
instability, so within an MMRd cohort the MSI-stable group is both small and unusual.

```{r other-classes-box, fig.height=4.5}
other_vars <- c("TMB_class", "HRD_class", "MSI_class")
if (nrow(dat) > 0) {
  knitr::kable(group_counts(dat, other_vars),
               caption = "Patients per group, score-derived classes (MMR-deficient cohort)")
  aneu_box(dat, other_vars,
           "Aneuploidy score by TMB / HRD / MSI class (MMR-deficient only)",
           ncol = 3)
}
```


```{r session-info}
sessionInfo()
```
