---
title: "ATTEND — 32. Cell-type composition vs aneuploidy"
author: "sceriff0"
date: "2026-08-11"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
# renv: tidyverse, here, ggpubr (boxplot p-values).
#   renv::install("ggpubr"); renv::snapshot()
#   # richer spatial stack, if/when point-pattern statistics are added (research/ §5):
#   renv::install(c("bioc::SpatialExperiment","bioc::imcRtools","spatstat.geom","sf","dbscan"))
library(tidyverse)
library(here)
library(ggpubr)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 6)

source(here("code", "load_phenotypes.R"))    # load_ihc_celltypes()
source(here("code", "load_clinical.R"))      # load_imaging_data()
source(here("code", "attend_classes.R"))     # add_molecular_classes(), add_scna_group()
source(here("code", "attend_plots.R"))       # highlight_points() overlay
source(here("code", "attend_io.R"))          # read_intermediate()
source(here("code", "attend_ihc.R"))         # ihc_celltype_metrics()

master <- read_intermediate("attend_master_joined") |>
  add_molecular_classes() |>
  mutate(aneuploidy_class = factor(aneuploidy_class,
                                   levels = c("aneuploidy-high", "aneuploidy-low")))
mmrd_pids <- add_scna_group(master) |> filter(mmr_group == "MMRd") |> pull(pid)
```

> **Question.** Does the mix of cells inside the annotated tumour region differ by aneuploidy class?
> **Cohort.** MMR-deficient only (n = `r length(mmrd_pids)`), with one marked exception: the interaction model, which cannot be estimated inside a single stratum.
> **Reads.** `attend_master_joined` plus the FlowPath annotation tree, via `load_ihc_celltypes()`.
> **Out of scope.** Whether FlowPath agrees with the pathologist -> [20_flowpath_concordance](20_flowpath_concordance.html). Survival by immune content -> [30_survival](30_survival.html).

```{r child='_explainers/cohort-mmrd.Rmd'}
```

```{r child='_explainers/aneuploidy-classes.Rmd'}
```

```{r child='_explainers/highlight-groups.Rmd'}
```

### Cell types and normalization

"Cell type" is `phenotype_clean`, the parenthetical phenotype label in the FlowPath export
(changeable in `code/load_phenotypes.R`). Every count is taken inside the annotated region
and expressed as a fraction, but the denominator changes the question, so all three are
shown:

| Normalization | Denominator | The question it answers |
|---|---|---|
| **all cells inside** | every cell in the annotation | what fraction of the tissue is this cell type? |
| **tumour cells inside** | tumour cells only | how much of this cell type per unit of tumour? |
| **CD45+ cells inside** | leukocytes only | what is the composition *of the immune compartment*? |

Each is drawn twice: the raw fraction with a Wilcoxon test, and the
arcsine-square-root-transformed fraction with a two-sample t-test. Agreement between the
two is reassurance; disagreement means the result depends on the test rather than the data.
No multiple-testing correction is applied anywhere in this report, and with one test per
cell type per normalization that is many comparisons, so every p-value here is descriptive.

## Building the restricted cohort

Three things are joined on `pid`: the aneuploidy class from the master, the per-patient
cell-type composition read from the FlowPath tree by `ihc_celltype_metrics()`, and the
continuous covariates. MMR status comes from `add_scna_group()`, so it is the same
`is_mmrd()` call the copy-number and classification reports use. The counts printed here
are the cohort for every figure below.

```{r data}
master_grp <- add_scna_group(master)        # adds mmr_group ("MMRd"/"MMRp") + scna_group

# per-patient cell-type composition inside the annotation (reads the FlowPath tree)
ct_all <- ihc_celltype_metrics(load_ihc_celltypes(), load_imaging_data()) |>
  left_join(master |> select(pid, aneuploidy_class), by = "pid") |>
  filter(!is.na(aneuploidy_class))
ct <- ct_all |> filter(pid %in% mmrd_pids)          # <- the MMRd restriction

# per-patient continuous covariates + TP53 status, restricted the same way
numcol <- function(col) if (col %in% names(master)) suppressWarnings(as.numeric(master[[col]])) else NA_real_
pt_all <- tibble(
  pid              = master$pid,
  aneuploidy_class = master$aneuploidy_class,
  aneuploidy_score = numcol(attend_cols$aneuploidy),
  HRD              = numcol(attend_cols$hrd),
  MSI              = numcol("msi__MSI_SCORE"),
  TP53             = if ("TP53_pathogenic" %in% names(master))
                       if_else(master$TP53_pathogenic, "TP53-abnormal", "TP53-normal") else NA_character_
)

# The two TMB definitions, added under their reader-facing names so every facet reads
# "TMB (normal)" / "TMB (ancestry corrected)" rather than an internal column name.
tmb_defs04 <- tmb_defs_present(master)
tmb_labs04 <- tmb_def_labels(tmb_defs04)
for (i in seq_along(tmb_defs04)) pt_all[[tmb_labs04[i]]] <- numcol(tmb_defs04[[i]])
covar_cols04 <- c(tmb_labs04, "HRD", "MSI")

pt <- pt_all |> filter(pid %in% mmrd_pids)          # <- the MMRd restriction

cat("MMR-deficient patients                       :", length(mmrd_pids), "\n")
cat("  ... with cell-type metrics + aneuploidy class:", n_distinct(ct$pid), "\n")
cat("  ... cell types measured                      :", n_distinct(ct$cell_type), "\n")
cat("TMB definitions plotted:", paste(tmb_labs04, collapse = " | "), "\n\n")
if (nrow(ct) > 0) print(table(aneuploidy = droplevels(factor(
  ct |> distinct(pid, aneuploidy_class) |> pull(aneuploidy_class)))))
```

```{r box-helpers, include=FALSE}
# One boxplot per cell type: aneuploidy-high vs -low. Boxes are filled by aneuploidy
# class; the highlight overlay is drawn on top in its own colour scale (colour != fill,
# so the two legends do not collide), partitioned so no patient is plotted twice.
box_by_aneu <- function(df, frac_col, denom_label) {
  d <- df |> filter(is.finite(.data[[frac_col]]))
  if (nrow(d) == 0) return(NULL)
  ggplot(d, aes(aneuploidy_class, .data[[frac_col]], fill = aneuploidy_class)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.5) +
    highlight_points(d, "aneuploidy_class", frac_col, base_size = 0.5, base_alpha = 0.35,
                     base_const = "black") +
    facet_wrap(~ cell_type, scales = "free_y") +
    stat_compare_means(method = "wilcox.test", label = "p.format", size = 3) +
    scale_fill_manual(values = attend_aneu_cols) +
    labs(title = paste0("Cell-type presence by aneuploidy (MMRd only) — normalized by ", denom_label),
         x = NULL, y = paste0("fraction of ", denom_label)) +
    attend_theme() +
    theme(legend.position = "none", axis.text.x = element_text(angle = 20, hjust = 1))
}

# Variance-stabilised companion: same panels, arcsine-square-root fraction + t-test.
box_by_aneu_arcsin <- function(df, frac_col, denom_label) {
  d <- df |> filter(is.finite(.data[[frac_col]])) |>
    mutate(.y = arcsin_sqrt(.data[[frac_col]]))
  if (nrow(d) == 0) return(NULL)
  ggplot(d, aes(aneuploidy_class, .y, fill = aneuploidy_class)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.5) +
    highlight_points(d, "aneuploidy_class", ".y", base_size = 0.5, base_alpha = 0.35,
                     base_const = "black") +
    facet_wrap(~ cell_type, scales = "free_y") +
    stat_compare_means(method = "t.test", label = "p.format", size = 3) +
    scale_fill_manual(values = attend_aneu_cols) +
    labs(title = paste0("Cell-type presence by aneuploidy (MMRd only) — arcsin(sqrt), by ", denom_label),
         subtitle = "arcsine-square-root transformed fraction; two-sample t-test",
         x = NULL, y = paste0("arcsin(sqrt(fraction of ", denom_label, "))")) +
    attend_theme() +
    theme(legend.position = "none", axis.text.x = element_text(angle = 20, hjust = 1))
}
```

## Cell-type composition by aneuploidy class

One panel per cell type, aneuploidy-high vs -low, within MMRd, from the `ct` table built
above. Each normalization is shown as a raw-fraction/Wilcoxon figure followed by its
arcsine/t-test companion.

### Normalized by all cells inside

Each cell type as a fraction of every cell in the annotation, so a panel reads "how much
of the tissue is this". Raw fractions, Wilcoxon.

```{r box-inside, fig.width=15, fig.height=11, out.width="100%"}
box_by_aneu(ct, "frac_inside", "all cells inside")
```

The same panels on the arcsine-square-root scale with a two-sample t-test, which is the
variance-stabilised companion rather than an alternative answer.

```{r box-inside-arcsin, fig.width=15, fig.height=11, out.width="100%"}
box_by_aneu_arcsin(ct, "frac_inside", "all cells inside")
```

### Normalized by tumour cells inside

Each cell type per unit of tumour: the denominator is tumour cells only, so a shift here
survives a change in how much tumour the annotation caught. Raw fractions, Wilcoxon.

```{r box-tumor, fig.width=15, fig.height=11, out.width="100%"}
box_by_aneu(ct, "frac_tumor", "tumour cells inside")
```

The arcsine/t-test companion of the panels above.

```{r box-tumor-arcsin, fig.width=15, fig.height=11, out.width="100%"}
box_by_aneu_arcsin(ct, "frac_tumor", "tumour cells inside")
```

### Normalized by CD45+ cells inside

Composition *of the immune compartment*: the denominator is leukocytes, so tumour-content
differences between patients cannot drive the contrast. Raw fractions, Wilcoxon.

```{r box-cd45, fig.width=15, fig.height=11, out.width="100%"}
box_by_aneu(ct, "frac_cd45", "CD45+ cells inside")
```

The arcsine/t-test companion of the panels above.

```{r box-cd45-arcsin, fig.width=15, fig.height=11, out.width="100%"}
box_by_aneu_arcsin(ct, "frac_cd45", "CD45+ cells inside")
```

## Immune content — the headline contrast

The question this report exists for: among MMR-deficient tumours, is aneuploidy-high
associated with an immune-cold phenotype?

Two leukocyte metrics counted inside the annotation — **CD45+ total** (all leukocytes)
and **CD3+CD45+** (double-positive T cells) — under each normalization. CD3+CD45+
requires a `CD3_sign` column in the FlowPath export; without it those rows are `NA` and
their facets drop out on their own.

```{r immune-data}
imm_all <- ihc_immune_metrics(load_ihc_celltypes(), load_imaging_data()) |>
  left_join(master |> select(pid, aneuploidy_class), by = "pid") |>
  filter(!is.na(aneuploidy_class))
imm_metrics <- imm_all |> filter(pid %in% mmrd_pids)      # <- the MMRd restriction

cat("MMR-deficient patients with immune metrics:", n_distinct(imm_metrics$pid),
    "| metrics present:", paste(sort(unique(imm_metrics$metric)), collapse = ", "), "\n")
```

```{r immune-box-helper, include=FALSE}
box_immune <- function(df, arcsin = FALSE) {
  d <- df |> filter(is.finite(frac))
  if (nrow(d) == 0) return(NULL)
  if (arcsin) d <- d |> mutate(frac = arcsin_sqrt(frac))
  method <- if (arcsin) "t.test"                 else "wilcox.test"
  ylab   <- if (arcsin) "arcsin(sqrt(fraction))" else "fraction"
  ttl    <- if (arcsin) "arcsin(sqrt) + t-test"  else "raw fraction + Wilcoxon"
  ggplot(d, aes(aneuploidy_class, frac, fill = aneuploidy_class)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.5) +
    highlight_points(d, "aneuploidy_class", "frac", base_size = 0.6, base_alpha = 0.4,
                     base_const = "black") +
    facet_wrap(~ metric + denom_label, scales = "free_y") +
    stat_compare_means(method = method, label = "p.format", size = 3) +
    scale_fill_manual(values = attend_aneu_cols) +
    labs(title = paste0("Immune content by aneuploidy (MMRd only) — ", ttl),
         x = NULL, y = ylab) +
    attend_theme() +
    theme(legend.position = "none", axis.text.x = element_text(angle = 20, hjust = 1))
}
```

CD45+ total and CD3+CD45+ fractions from `imm_metrics`, by aneuploidy class, one facet per
normalization, within MMRd.

```{r immune-box, fig.width=12, fig.height=8, out.width="100%"}
if (nrow(filter(imm_metrics, is.finite(frac))) > 2) {
  print(box_immune(imm_metrics, arcsin = FALSE))
  print(box_immune(imm_metrics, arcsin = TRUE))
} else message("No finite immune metrics in the MMRd cohort (CD3/CD45 counts missing) — skipping.")
```

### The pre-specified primary contrast

Of everything above, one comparison is designated **primary**: **CD45+ / all cells
inside**, tested with the **arcsine-square-root transform and a two-sample t-test**. The
Wilcoxon on the raw fraction is reported next to it as a companion, not a substitute.
The other two normalizations, and CD3+CD45+, are **secondary** — they can and do degrade
to a skip when counts are missing.

Naming one primary endpoint in advance is what keeps this from being a search across six
panels for the smallest p-value.

```{r immune-mmrd-test}
mmrd_arcsin_t_wilcox <- function(df, metric_, denom_) {
  d <- df |> dplyr::filter(metric == metric_, denom_label == denom_, is.finite(frac))
  tab <- table(d$aneuploidy_class)
  if (length(tab) < 2 || any(tab < 2)) {
    message("MMRd ", metric_, " / ", denom_, ": insufficient n per aneuploidy arm — skipping.")
    return(invisible(NULL))
  }
  t_res <- t.test(arcsin_sqrt(frac) ~ aneuploidy_class, data = d)     # primary
  w_res <- wilcox.test(frac ~ aneuploidy_class, data = d)             # companion
  cat(sprintf("MMRd %-10s / %-20s  n=%d (high=%d, low=%d)  arcsin-t p=%.3g  Wilcoxon p=%.3g\n",
      metric_, denom_, nrow(d), tab[["aneuploidy-high"]], tab[["aneuploidy-low"]],
      t_res$p.value, w_res$p.value))
  list(d = d, t = t_res, w = w_res)
}
res_primary <- mmrd_arcsin_t_wilcox(imm_metrics, "CD45+", "all cells inside")      # PRIMARY
res_tumor   <- mmrd_arcsin_t_wilcox(imm_metrics, "CD45+", "tumour cells inside")   # secondary
res_cd45n   <- mmrd_arcsin_t_wilcox(imm_metrics, "CD3+CD45+", "CD45+ cells inside") # secondary
```

The primary contrast on its own. The subtitle carries both p-values and the group size —
at this n a single patient moving between groups can change the result, which is why the
count is on the figure rather than in a caption.

```{r immune-mmrd-box, fig.width=6, fig.height=5}
if (!is.null(res_primary)) {
  d <- res_primary$d
  ggplot(d, aes(aneuploidy_class, frac, fill = aneuploidy_class)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.5) +
    highlight_points(d, "aneuploidy_class", "frac", base_size = 1.2, base_alpha = 0.5, base_const = "black") +
    stat_compare_means(method = "wilcox.test", label = "p.format", size = 3) +
    scale_fill_manual(values = attend_aneu_cols) +
    labs(title = "PRIMARY: CD45+ immune fraction (all cells inside) — MMR-deficient only",
         subtitle = paste0("arcsin(sqrt) t p = ", signif(res_primary$t$p.value, 3),
                           " | Wilcoxon p = ", signif(res_primary$w$p.value, 3),
                           " (n = ", nrow(d), "; small-n — suggestive only)"),
         x = NULL, y = "CD45+ fraction of all cells inside") +
    attend_theme() + theme(legend.position = "none")
} else message("Primary within-MMRd immune contrast skipped — see message above.")
```

### The one model that needs MMR-proficient patients

The single exception to this report's MMRd restriction, and it has to be: an interaction
asks whether the aneuploidy effect *differs between* MMRd and MMRp, which cannot be
estimated inside one stratum. The model is
`arcsin(sqrt(CD45+/all cells inside)) ~ aneuploidy_class * mmr_group`, fitted on both
groups from `imm_all`. The interaction row is the one of interest: a large coefficient
there is what justifies restricting everything else to MMRd. No correction is applied.

```{r immune-interaction}
imm_int <- imm_all |>
  dplyr::filter(metric == "CD45+", denom_label == "all cells inside", is.finite(frac)) |>
  dplyr::left_join(master_grp |> dplyr::select(pid, mmr_group), by = "pid") |>
  dplyr::filter(!is.na(mmr_group), !is.na(aneuploidy_class)) |>
  dplyr::mutate(y = arcsin_sqrt(frac))

if (nrow(imm_int) > 4 && dplyr::n_distinct(imm_int$aneuploidy_class) == 2 &&
    dplyr::n_distinct(imm_int$mmr_group) == 2) {
  fit <- lm(y ~ aneuploidy_class * mmr_group, data = imm_int)
  coef_tbl <- if (requireNamespace("broom", quietly = TRUE)) broom::tidy(fit) else
    as.data.frame(summary(fit)$coefficients)
  print(knitr::kable(coef_tbl, digits = 3,
        caption = "arcsin(sqrt(CD45+/all-cells-inside)) ~ aneuploidy_class * mmr_group — BOTH MMR groups, no correction"))
} else message("Not enough patients across both MMR groups for the interaction model.")
```

## Aneuploidy against the continuous covariates

```{r child='_explainers/tmb-two-definitions.Rmd'}
```

Aneuploidy score against TMB (both definitions), HRD, and MSI, within MMRd. One
point per patient, coloured by aneuploidy class, with Spearman's rho per panel. Spearman
rather than Pearson because these scores are bounded and skewed.

The two TMB panels are the same patients under the two definitions, so a visible
difference between them is the germline correction moving values, not new data.

```{r aneu-scatter, fig.width=12, fig.height=7}
scat <- pt |>
  pivot_longer(all_of(covar_cols04), names_to = "covariate", values_to = "value") |>
  filter(!is.na(aneuploidy_score), !is.na(value)) |>
  mutate(covariate = factor(covariate, levels = covar_cols04))

if (nrow(scat) > 2) {
  rho <- scat |>
    group_by(covariate) |>
    summarise(label = paste0("rho = ",
                round(suppressWarnings(cor(aneuploidy_score, value, method = "spearman",
                                           use = "complete.obs")), 2)), .groups = "drop")
  ggplot(scat, aes(aneuploidy_score, value, colour = aneuploidy_class)) +
    highlight_points(scat, "aneuploidy_score", "value", base_colour = "aneuploidy_class",
                     base_palette = attend_aneu_cols,
                     base_name = "aneuploidy", jitter_width = 0, base_size = 1.5, base_alpha = 0.6) +
    geom_smooth(method = "lm", se = FALSE, colour = "grey40", linewidth = 0.6) +
    geom_text(data = rho, aes(x = Inf, y = Inf, label = label),
              inherit.aes = FALSE, hjust = 1.05, vjust = 1.5, size = 3) +
    facet_wrap(~ covariate, scales = "free_y", ncol = 3) +
    labs(title = "Aneuploidy score vs continuous covariates (MMR-deficient only)",
         x = "aneuploidy score", y = "covariate value", colour = NULL) +
    attend_theme() + theme(legend.position = "top")
} else message("Not enough MMR-deficient patients with aneuploidy score + a covariate.")
```

The same covariates against the **binarised** aneuploidy class rather than the continuous
score — the categorical counterpart to the scatter above, with a per-covariate Wilcoxon.
Comparing the two is a check on the `r attend_thresholds$aneuploidy` cutpoint: a
correlation visible in the scatter but absent here means the cutpoint is discarding it.

```{r aneu-class-box, fig.width=12, fig.height=7}
cb <- pt |>
  pivot_longer(all_of(covar_cols04), names_to = "covariate", values_to = "value") |>
  filter(!is.na(aneuploidy_class), !is.na(value)) |>
  mutate(covariate = factor(covariate, levels = covar_cols04))

if (nrow(cb) > 2) {
  ggplot(cb, aes(aneuploidy_class, value, fill = aneuploidy_class)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.5) +
    highlight_points(cb, "aneuploidy_class", "value", base_size = 0.6, base_alpha = 0.4,
                     base_const = "black") +
    stat_compare_means(method = "wilcox.test", label = "p.format", size = 3) +
    facet_wrap(~ covariate, scales = "free_y", ncol = 3) +
    scale_fill_manual(values = attend_aneu_cols) +
    labs(title = "Continuous covariates by aneuploidy class (MMR-deficient only)",
         x = NULL, y = "covariate value") +
    attend_theme() + theme(legend.position = "none")
} else message("Not enough MMR-deficient patients with aneuploidy class + a covariate.")
```

## Aneuploidy by TP53 status

Within MMRd, aneuploidy score in TP53-abnormal versus TP53-normal tumours. TP53 status is
`TP53_pathogenic` from the master, derived from the MAF in report 10. TP53 loss is the
canonical driver of chromosomal instability, so this is the expected-positive control for
the aneuploidy score itself: if aneuploidy does not separate on TP53 here, the score is
suspect before any of the immune results are interpreted.

```{r mmrd-tp53, fig.width=5, fig.height=5}
tp <- pt |> filter(!is.na(TP53), !is.na(aneuploidy_score))

if (nrow(tp) > 2 && dplyr::n_distinct(tp$TP53) == 2) {
  ggplot(tp, aes(TP53, aneuploidy_score, fill = TP53)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.5) +
    highlight_points(tp, "TP53", "aneuploidy_score", base_size = 1.5, base_alpha = 0.5,
                     base_const = "black") +
    stat_compare_means(method = "wilcox.test", label = "p.format") +
    scale_fill_manual(values = attend_tp53_cols) +
    labs(title = "Aneuploidy by TP53 status (MMR-deficient only)",
         x = NULL, y = "aneuploidy score") +
    attend_theme() + theme(legend.position = "none")
} else {
  message("Not enough MMR-deficient patients with both TP53 classes, or TP53_pathogenic ",
          "is missing from the master (rebuild report 10).")
}
```

## Aneuploidy against the pathologist's imaging calls

The pathologist's per-patient scores compared between aneuploidy high and low, within
MMRd. These come from the imaging sheet keyed by image → pid, averaged over a patient's
images for the continuous measures and taking the first non-missing value for the
categorical calls.

```{r imaging-covars}
first_nonNA <- function(x) { x <- x[!is.na(x)]; if (length(x)) x[1] else NA }

imaging_pt <- load_imaging_data() |>
  filter(!is.na(ID)) |>
  group_by(pid = as.character(ID)) |>
  summarise(
    Immune_infiltrate      = mean(suppressWarnings(as.numeric(Immune_infiltrate)),     na.rm = TRUE),
    Immune_infiltrate_PD1  = mean(suppressWarnings(as.numeric(Immune_infiltrate_PD1)), na.rm = TRUE),
    ARID1A                 = first_nonNA(ARID1A),
    Tumor_cells_PDL1       = first_nonNA(Tumor_cells_PDL1),
    Immune_infiltrate_PDL1 = first_nonNA(Immune_infiltrate_PDL1),
    .groups = "drop"
  ) |>
  left_join(master |> select(pid, aneuploidy_class), by = "pid") |>
  filter(!is.na(aneuploidy_class), pid %in% mmrd_pids)   # <- the MMRd restriction

cat("MMR-deficient patients with imaging covariates:", n_distinct(imaging_pt$pid), "\n")
```

Continuous pathologist scores (immune infiltrate and PD-1 infiltrate, as percentages) by
aneuploidy class, Wilcoxon per panel. These are the pathologist's independent read of the
same immune question the FlowPath contrast above asked.

```{r imaging-covars-cont, fig.width=8, fig.height=4}
icc <- imaging_pt |>
  pivot_longer(c(Immune_infiltrate, Immune_infiltrate_PD1),
               names_to = "covariate", values_to = "value") |>
  filter(is.finite(value))

if (nrow(icc) > 2) {
  ggplot(icc, aes(aneuploidy_class, value, fill = aneuploidy_class)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.5) +
    highlight_points(icc, "aneuploidy_class", "value", base_size = 0.6, base_alpha = 0.4,
                     base_const = "black") +
    stat_compare_means(method = "wilcox.test", label = "p.format", size = 3) +
    facet_wrap(~ covariate, scales = "free_y") +
    scale_fill_manual(values = attend_aneu_cols) +
    labs(title = "Pathologist immune scores by aneuploidy class (MMR-deficient only)",
         x = NULL, y = "pathologist %") +
    attend_theme() + theme(legend.position = "none")
}
```

Categorical calls as stacked proportions — how each call distributes within aneuploidy
high vs low. Bars are proportions, so the **group sizes are not visible in this figure**;
read them from the count printed above. `MMR_Status` is omitted because it is constant
here by construction.

```{r imaging-covars-cat, fig.width=11, fig.height=4}
icat <- imaging_pt |>
  select(aneuploidy_class, ARID1A, Tumor_cells_PDL1, Immune_infiltrate_PDL1) |>
  pivot_longer(-aneuploidy_class, names_to = "covariate", values_to = "level") |>
  filter(!is.na(level), level != "")

if (nrow(icat) > 0) {
  ggplot(icat, aes(aneuploidy_class, fill = level)) +
    geom_bar(position = "fill") +
    facet_wrap(~ covariate, scales = "free_x", nrow = 1) +
    labs(title = "Pathologist categorical calls by aneuploidy class (MMR-deficient only)",
         x = NULL, y = "proportion of patients", fill = NULL) +
    attend_theme() +
    theme(axis.text.x = element_text(angle = 20, hjust = 1))
}
```

```{r session-info}
sessionInfo()
```
