---
title: "ATTEND — 34. Recurrent copy number"
author: "sceriff0"
date: "2026-08-11"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
# renv: tidyverse, here. No maftools — both figures here are copy-number only.
library(tidyverse)
library(here)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 11, fig.height = 8)

source(here("code", "load_wes_results.R"))   # load_ascets_calls(), load_cnv_data()
source(here("code", "attend_classes.R"))     # recurrent_arm_calls(), segment_pileup(), attend_cnv
source(here("code", "attend_plots.R"))       # attend_theme()
```

> **Question.** Which chromosome arms and which genomic regions are recurrently gained or lost across the cohort?
> **Cohort.** Every sample with copy-number data; no molecular restriction.
> **Reads.** ASCETS arm-level calls in `data/ascets/` and ClassifyCNV altered segments in `data/cnv_annotations/`.
> **Out of scope.** Whether recurrence differs between subgroups -> [35_subgroup_cna_contrast](35_subgroup_cna_contrast.html). Mutations -> [33_mutation_landscape](33_mutation_landscape.html).

```{r child='_explainers/aneuploidy-classes.Rmd'}
```

Two resolutions of the same question, coarsest first. Both are **descriptive frequencies
with no significance testing**: GISTIC 2.0 q-values are the field's standard for
recurrence significance and are not applied to these figures.

## Arm level

Cohort frequency of whole-arm gains and losses from the **ASCETS arm-level calls** — the
recurrent events that define endometrial subtypes, such as 1q and 8q gain in serous-like
tumours. Gains extend right, losses left, and arms are shown in **genome order**
(`order_arms_genomic()`), not sorted by frequency. Each arm's fraction is out of its own
**evaluable** samples — calls of LOWCOV/NC/NA are excluded from that arm's denominator,
not counted as neutral — and that count is printed on the axis label (`arm (n=…)`) so a
reduced denominator is visible rather than hidden. The bars below show the
`r attend_cnv$arm_calls$n_barplot` most-altered *eligible* arms out of roughly 44 in the
genome — not every arm — where eligible means evaluable in at least
`r 100 * attend_cnv$arm_calls$min_evaluable_frac`% of samples, the identical rule
`top_arms` applies to the arm annotation tracks on the oncoplots in
[33_mutation_landscape](33_mutation_landscape.html), so neither figure can rank a
poorly-covered arm ahead of a well-covered one. An uncallable arm's `CNV_<arm>` value on
those oncoplots is `NA`, not `"neutral"`: missing data is not evidence of no alteration.

```{r recurrent-cnv, fig.width=9, fig.height=6}
arm_freq <- recurrent_arm_calls(load_ascets_calls(), n_top = attend_cnv$oncoplot_arms)

if (is.null(arm_freq)) {
  message("No ASCETS arm-level calls (data/ascets/) — skipping recurrent-CNV summary.")
} else {
  # Same eligibility gate recurrent_arm_calls() applies to top_arms (freq$eligible), so
  # the barplot and the oncoplot annotation tracks can never disagree about which arms
  # are well-covered enough to rank. Applied BEFORE slice_head(), so a poorly-covered arm
  # (e.g. an acrocentric p-arm ASCETS marks LOWCOV in most samples) cannot be promoted into
  # the figure just because it scores 100% altered on a 2-sample denominator.
  fr <- arm_freq$freq |>
    filter(eligible) |>
    slice_head(n = attend_cnv$arm_calls$n_barplot) |>
    mutate(arm = factor(arm, levels = rev(order_arms_genomic(arm))))

  # Axis labels carry each arm's evaluable-sample count, so a reader can see when a
  # frequency rests on a reduced denominator (matched by factor level, not row order).
  arm_labels <- setNames(paste0(as.character(fr$arm), " (n=", fr$n_evaluable, ")"),
                         as.character(fr$arm))

  fr_long <- fr |>
    pivot_longer(c(gain, loss), names_to = "direction", values_to = "frac") |>
    mutate(frac = if_else(direction == "loss", -frac, frac))

  ggplot(fr_long, aes(arm, frac, fill = direction)) +
    geom_col(width = 0.8) +
    geom_hline(yintercept = 0, colour = "grey40", linewidth = 0.3) +
    coord_flip() +
    scale_fill_manual(values = c(gain = "firebrick", loss = "steelblue")) +
    scale_x_discrete(labels = arm_labels) +
    scale_y_continuous(labels = function(x) paste0(abs(round(100 * x)), "%")) +
    labs(title = "Recurrent arm-level SCNAs across the cohort",
         subtitle = "gain (right) vs loss (left), by fraction of evaluable samples; arms in genome order",
         x = NULL, y = "fraction of evaluable samples", fill = NULL) +
    attend_theme()
}
```

## Genome-wide, by chromosomal location

The TCGA Fig. 1a convention: alterations plotted **by position along the genome** rather
than summarised per arm, split into two amplitude panels — **low-level** and **high-level**
(`|log2 ratio| >= ` `r attend_cnv$pileup$high_abs_logratio`, the conventional
amplification/homozygous-deletion cutoff) — gain up/red, loss down/blue. Grey vertical
lines are chromosome boundaries; dashed lines mark centromeres, read from the arm-boundary
table. ChrY is dropped from the axis — the cohort is uterine and therefore all female, so
it carries no calls and only an empty tail.

Built from the ClassifyCNV altered segments (`data/cnv_annotations/`) in
`r attend_cnv$pileup$bin / 1e6` Mb bins; the figure caption states the sample denominator
and whether it came from an explicit count, the profiled-sample attribute, or the segments
observed. **These are descriptive frequencies with no significance testing** — GISTIC
q-values are the field's standard for recurrence significance and are not applied here.
This is the figure that shows *focal* recurrence — a narrow tall spike is a focal peak, a
broad plateau is an arm-level event and also appears in the arm barplot above.

```{r segment-pileup, fig.width=12, fig.height=6}
arm_bounds <- load_arm_boundaries()
pile <- segment_pileup(load_cnv_data(), arm_bounds)

if (is.null(pile)) {
  message("No ClassifyCNV segments (data/cnv_annotations/) or arm-boundary table — ",
          "skipping the genome-wide pileup.")
} else {
  # Capture the denominator + its source before any filter()/mutate() that could drop
  # these custom attributes.
  n_samples        <- attr(pile, "n_samples")
  n_samples_source <- attr(pile, "n_samples_source")
  offs             <- attr(pile, "chrom_offsets")

  # Readable phrase for the subtitle, not the internal token (argument/profiled/observed)
  # segment_pileup() records in n_samples_source — a subtitle should read as English.
  n_samples_desc <- c(argument = "an explicitly supplied cohort size",
                      profiled = "the number of samples profiled",
                      observed = "the samples present in the segment data")[[n_samples_source]]

  # ChrY dropped: the cohort is uterine and therefore all female, so chrY carries no
  # calls and would only draw an empty axis tail. ChrX is kept.
  pile <- pile |> filter(chrom != "chrY")
  offs <- offs  |> filter(chrom != "chrY")

  # Centromere = each chromosome's p/q boundary, read from the arm-boundary table
  # (never hard-coded) and converted to genome coordinates with the same offsets used
  # for the chromosome-boundary lines and axis breaks.
  centromeres <- arm_bounds |>
    filter(str_detect(arm, "p$"), chrom != "chrY") |>
    mutate(chrom = as.character(chrom)) |>
    inner_join(offs |> select(chrom, offset), by = "chrom") |>
    transmute(gpos = offset + end)

  pl <- pile |>
    select(gpos, gain_low, gain_high, loss_low, loss_high) |>
    pivot_longer(-gpos, names_to = c("direction", "tier"), names_sep = "_",
                values_to = "frac") |>
    mutate(frac = if_else(direction == "loss", -frac, frac),
           tier = factor(tier, levels = c("low", "high"),
                        labels = c("low-level", "high-level")))

  ggplot(pl, aes(gpos, frac, fill = direction)) +
    geom_col(width = attend_cnv$pileup$bin) +
    geom_hline(yintercept = 0, colour = "grey40", linewidth = 0.3) +
    geom_vline(xintercept = offs$offset[-1], colour = "grey85", linewidth = 0.2) +
    geom_vline(data = centromeres, aes(xintercept = gpos), colour = "grey60",
              linewidth = 0.3, linetype = "dashed") +
    facet_grid(tier ~ ., scales = "free_y") +
    scale_fill_manual(values = c(gain = "firebrick", loss = "steelblue")) +
    scale_x_continuous(breaks = offs$mid, labels = str_remove(offs$chrom, "^chr"),
                       expand = c(0, 0)) +
    scale_y_continuous(labels = function(x) paste0(abs(round(100 * x)), "%")) +
    labs(title = "Genome-wide recurrent SCNA pileup (TCGA Fig. 1a convention)",
         subtitle = paste0("gain (up) vs loss (down), fraction of ", n_samples,
                           " samples (", n_samples_desc, "); ",
                           attend_cnv$pileup$bin / 1e6,
                           " Mb bins; dashed lines = centromeres; no significance testing applied"),
         x = "chromosome", y = "fraction of samples", fill = NULL) +
    attend_theme() +
    theme(panel.grid.major.x = element_blank(),
          panel.grid.minor.x = element_blank())
}
```

---

```{r session-info}
sessionInfo()
```
