---
title: "ATTEND — 35. Aneuploidy x MMR copy-number contrast"
author: "sceriff0"
date: "2026-08-11"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
# renv: tidyverse, here, fs.
library(tidyverse)
library(here)
library(fs)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 11, fig.height = 8)

source(here("code", "load_clinical.R"))      # load_gianlu_clinical_data()
source(here("code", "load_wes_results.R"))   # GISTIC + ASCETS + seg loaders
source(here("code", "attend_harmonise.R"))   # build_barcode_pid(), make_id_cfg()
source(here("code", "attend_classes.R"))     # add_molecular_classes(), add_scna_group(), attend_scna
source(here("code", "attend_scna.R"))        # peak union, panel score, LOO stability
source(here("code", "attend_plots.R"))       # scna_mirror_plot(), panel_score_plot()
source(here("code", "attend_io.R"))          # read_intermediate()

master <- tryCatch(read_intermediate("attend_master_joined") |>
                     add_molecular_classes() |> add_scna_group(),
                   error = function(e) NULL)
n_mmrd_high <- if (is.null(master)) NA_integer_ else
  sum(master$scna_group == "MMRd-high", na.rm = TRUE)
```

> **Question.** Do aneuploidy-high MMR-deficient tumours carry the same recurrent copy-number peaks as aneuploidy-high MMR-proficient tumours, or do they reach aneuploidy by a different route?
> **Cohort.** A contrast, not a stratum: MMRd-high (n = `r n_mmrd_high`) against MMRp-high, both drawn from the full cohort.
> **Reads.** GISTIC peaks in `data/gistic/`, the `.seg` segments, and `attend_master_joined` for the grouping.
> **Out of scope.** Cohort-wide recurrence -> [34_recurrent_cna](34_recurrent_cna.html). Mutations -> [33_mutation_landscape](33_mutation_landscape.html).

```{r child='_explainers/aneuploidy-classes.Rmd'}
```

Either the MMRd-high group reaches aneuploidy the same way MMR-proficient serous-like
tumours do, in which case it should carry the same recurrent peaks, or it gets there by a
different route and the peak profile should differ. With n = `r n_mmrd_high` in the
MMRd-high group, the endpoint is a single pre-specified composite score rather than a
per-peak test.

## Setup and the two ID spaces

GISTIC and `.seg` sample ids are sequencing **barcodes**; the master is keyed on **pid**.
Joining them directly gives an all-`NA` result *with no error*, so every lookup below goes
through `group_of_barcode()`, built on the same crosswalk as everything else.

The grouping variable `scna_group` composes `is_mmrd()` (IHC-based, MSI deliberately not a
criterion) with `aneuploidy_class` at the pre-registered cut
`r attend_thresholds$aneuploidy`, the same definition
[33_mutation_landscape](33_mutation_landscape.html)'s frequency comparison uses.

```{r scna-gate}
have_master <- exists("master") && !is.null(master) && nrow(master) > 0
if (have_master && !"scna_group" %in% names(master)) master <- add_scna_group(master)

cw_scna <- tryCatch(build_barcode_pid(load_gianlu_clinical_data(), make_id_cfg()),
                    error = function(e) NULL)
have_cw <- !is.null(cw_scna) && nrow(cw_scna) > 0

group_of_barcode <- function(ids) {
  if (!have_master || !have_cw)
    return(factor(rep(NA_character_, length(ids)), levels = attend_scna$group_levels))
  barcode_scna_group(ids, master, cw_scna)
}

gx_dir   <- here("data", attend_cnv$gistic$dir)
run_dirs <- c(all       = file.path(gx_dir, "all"),
              mmrp_high = file.path(gx_dir, "mmrp_high"),
              mmrp_low  = file.path(gx_dir, "mmrp_low"),
              mmrd_low  = file.path(gx_dir, "mmrd_low"),
              mmrd_high = file.path(gx_dir, "mmrd_high"))
have_gistic <- dir.exists(run_dirs[["all"]])

if (!have_master) {
  cat("Master table absent — run report 10 first.\n")
} else if (!have_gistic) {
  cat("**GISTIC outputs absent** (`", run_dirs[["all"]], "`).\n\n",
      "Run `MODE=groups sbatch code/run_gistic.sh`, then `MODE=loo`. ",
      "this report is skipped, not silently empty.\n", sep = "")
}
```

Peaks come from five runs; per-sample calls come from one. Five GISTIC2 runs feed this
report: the pooled cohort and the four `scna_group` strata. GISTIC's background model is
cohort-specific, so the same tumour gets a different thresholded call depending on which
cohort it was run in, and comparing calls across runs would compare five different null
models. The two roles are therefore kept apart:

- **Peak definition** uses the union of significant peaks across all five runs
  (`gistic_peak_union()`), so a peak private to the small MMRd-high stratum, which the
  pooled run's genome-wide threshold would bury, is still discoverable.
- **Per-sample calls** are read only from the pooled run's `all_lesions.conf_*.txt`. No
  per-sample call here comes from a stratified run.

The cohort this contrast conditions on. Fisher tests whether MMR status and aneuploidy class are
associated at all — if they are strongly associated, the MMRd-high cell is small by
construction and everything below is underpowered.

```{r scna-contingency, eval=have_master}
if (all(c("mmr_group", "aneuploidy_class") %in% names(master))) {
  tab2 <- table(MMR = master$mmr_group, Aneuploidy = master$aneuploidy_class)
  print(knitr::kable(tab2, caption = "MMR x aneuploidy-class contingency (patient level)"))
  ft2 <- tryCatch(fisher.test(tab2), error = function(e) NULL)
  if (!is.null(ft2)) cat("Fisher p =", signif(ft2$p.value, 3),
                        ", OR =", signif(unname(ft2$estimate), 3), "\n\n")
}
```

## The primary endpoint — a pre-specified serous-like panel score

**Why a composite score rather than testing peaks one at a time.** At n ≈ 9 no individual
peak has the power to show a difference, and testing hundreds of peaks in a group that
small produces false positives, not findings. A single composite score over a *fixed,
externally-specified* set of loci concentrates the evidence into one test.

**Pre-specification is what makes this confirmatory.** The 12-locus panel
(`attend_scna$panel`) is fixed *a priori* from the TCGA UCEC recurrent-peak list
(Kandoth et al. 2013). No locus was added, removed, or re-directioned after looking at
ATTEND's own GISTIC output. Each locus is **directional** — a deletion at *MYC* does not
count as serous-like — matched by `match_panel_peaks()`.

Loci with no matching pooled-run peak of the right direction **leave the denominator**
rather than counting as unaltered, so the score is a fraction of *resolvable* loci.

The test is a **label permutation** (B = `r attend_scna$perm_B`), which makes no
distributional assumption — appropriate at this n. A Wilcoxon is reported alongside.
A permutation p is bounded below by 1/(B+1), so a floor-level result is reported as
`p < …` rather than as an exact tiny value.

```{r scna-panel-score, eval=have_master && have_gistic, fig.height=5}
pooled <- load_gistic_lesions_at(run_dirs[["all"]])
if (is.null(pooled)) {
  cat("Pooled GISTIC run (`", run_dirs[["all"]], "`) not readable — panel score skipped.\n", sep = "")
} else {
  scores <- panel_score(pooled$mat, pooled$peaks)
  grp    <- group_of_barcode(rownames(pooled$mat))

  cat("Panel resolved", attr(scores, "n_loci"), "of 12 pre-specified loci to a pooled-run ",
      "peak of the matching direction",
      if (attr(scores, "n_loci") > 0) paste0(" (", paste(attr(scores, "loci"), collapse = ", "), ").")
      else ".", "\n\n")

  p1 <- panel_score_plot(scores, grp,
                         title = "Composite TCGA UCEC serous-like panel score, by subgroup")
  if (!is.null(p1)) print(p1)

  keep2 <- !is.na(grp) & grp %in% c("MMRd-high", "MMRp-high")
  if (sum(keep2) >= 3 && dplyr::n_distinct(droplevels(grp[keep2])) == 2) {
    g2 <- droplevels(grp[keep2])
    pt <- perm_test_two_group(scores[keep2], g2, B = attend_scna$perm_B, seed = 1)
    wt <- tryCatch(stats::wilcox.test(scores[keep2] ~ g2), error = function(e) NULL)

    floor_p <- 1 / (pt$B + 1)
    p_txt <- if (pt$p <= floor_p + 1e-12) {
      sprintf("p < %s (bounded at the B = %s resolution floor)",
              format(floor_p, scientific = TRUE, digits = 1), format(pt$B, big.mark = ","))
    } else sprintf("p = %.4g", pt$p)

    cat("**MMRd-high (n =", sum(g2 == "MMRd-high"), ") vs MMRp-high (n =", sum(g2 == "MMRp-high"),
        ")**: observed difference in mean panel score =", signif(pt$stat, 3),
        "; label-permutation ", p_txt, ".",
        if (!is.null(wt)) paste0(" Wilcoxon p = ", signif(wt$p.value, 3), ".") else "",
        "\n\nIf the 9 reach aneuploidy by a different route, the score separates from ",
        "MMRp-high; if they phenocopy copy-number-high tumours, it does not.\n\n")
  } else {
    cat("Fewer than 3 samples resolved to both MMRd-high and MMRp-high — permutation test skipped.\n\n")
  }
}
```

### Which loci drive the score

The composite above is the primary endpoint; this breaks it into its 12 parts as a
**secondary, more granular** view. Each locus is tested one at a time and **Holm-corrected
within the 12-locus family**.

Holm is applied over **all 12 pre-specified loci**, including any that did not resolve to
a pooled-run peak (`p = NA`) — not just the testable ones. Shrinking the family to the
loci that happened to resolve would inflate significance on the survivors.

```{r scna-locus-table, eval=have_master && have_gistic}
if (exists("pooled") && !is.null(pooled)) {
  mp <- match_panel_peaks(pooled$peaks, attend_scna$panel)
  ft_all <- scna_freq_table(pooled$mat, grp)
  rows <- lapply(seq_len(nrow(mp)), function(i) {
    pk <- mp$peak_id[i]
    if (is.na(pk)) return(data.frame(locus = mp$locus[i], direction = mp$direction[i],
                                     freq_mmrd_high = NA_real_, freq_mmrp_high = NA_real_,
                                     p = NA_real_, stringsAsFactors = FALSE))
    a <- ft_all[ft_all$peak_id == pk & ft_all$scna_group == "MMRd-high", ]
    b <- ft_all[ft_all$peak_id == pk & ft_all$scna_group == "MMRp-high", ]
    p <- NA_real_
    if (nrow(a) && nrow(b)) {
      tt <- matrix(c(a$n_altered, a$n - a$n_altered, b$n_altered, b$n - b$n_altered), nrow = 2)
      p <- tryCatch(stats::fisher.test(tt)$p.value, error = function(e) NA_real_)
    }
    data.frame(locus = mp$locus[i], direction = mp$direction[i],
               freq_mmrd_high = if (nrow(a)) a$freq else NA_real_,
               freq_mmrp_high = if (nrow(b)) b$freq else NA_real_,
               p = p, stringsAsFactors = FALSE)
  })
  locus_tbl <- do.call(rbind, rows)
  locus_tbl$p_holm <- family_adjust(locus_tbl$p, "A", n = nrow(locus_tbl))
  knitr::kable(locus_tbl, digits = 3,
               caption = "The 12 pre-specified TCGA UCEC loci, Holm-adjusted within family (secondary to the composite score)")
}
```

## Where each subgroup's calls sit on the genome

Peak-call frequency from the pooled run, amplifications above the axis and deletions
below, split by subgroup and chromosome. The visual counterpart to the composite score:
if MMRd-high really phenocopies MMRp-high, their two panels should have peaks in the same
places.

```{r scna-mirror, eval=have_master && have_gistic}
if (exists("pooled") && !is.null(pooled)) {
  grp3      <- group_of_barcode(rownames(pooled$mat))
  freq_tbl3 <- scna_freq_table(pooled$mat, grp3)
  p3 <- scna_mirror_plot(freq_tbl3, pooled$peaks,
                         title = "Pooled-run peak-call frequency by subgroup (amp above / del below)")
  if (!is.null(p3)) print(p3)
}
```

How much the five runs' significant-peak sets overlap, as pairwise Jaccard. Low overlap
between a stratum and the pooled run means that stratum has genuinely private peaks —
which is the case the union-of-five design exists to catch. Peaks private to a single run
are listed separately.

```{r scna-peakset, eval=have_master && have_gistic}
union_peaks <- gistic_peak_union(run_dirs)
if (is.null(union_peaks)) {
  cat("No GISTIC lesions parsed from any of the five run folders — skipped.\n")
} else {
  pss <- peak_set_structure(union_peaks)
  print(knitr::kable(round(pss$jaccard, 2),
                     caption = "Jaccard overlap of significant-peak sets between the five GISTIC runs"))
  if (nrow(pss$private))
    print(knitr::kable(pss$private,
                       caption = "Peaks private to a single run (no overlapping peak in any other)"))
}
```

## Leave-one-out stability of the MMRd-high peaks

**The asymmetric rule this rests on.** A peak *present* in the MMRd-high list, supported by
several of the 9 tumours and stable to leave-one-out, is **informative**. A peak *absent*
from that list is **uninformative** — at n = 9 absence is indistinguishable from low power.
Presence carries evidence; absence does not. That is why the table reports support counts
and retention fractions for MMRd-high peaks rather than a between-group presence/absence
verdict.

The retention fraction is the proportion of the 9 leave-one-out GISTIC re-runs in which a
peak survives. A peak retained in 9/9 does not depend on any single tumour; a peak
retained in 3/9 is one or two tumours' worth of signal.

**This is a descriptive robustness diagnostic, not a calibrated error bound** — informal
stability selection (Meinshausen & Bühlmann 2010) adapted to n = 9. Canonical stability
selection uses ⌊n/2⌋ subsamples and an exchangeability assumption that leave-one-out at
this n only approximates. It is not an FDR statement.

Support counts carry Wilson 95% confidence intervals, which behave sensibly at small n
and at frequencies near 0 or 1 where a normal-approximation interval does not.

```{r scna-loo, eval=have_master && have_gistic}
if (exists("union_peaks") && !is.null(union_peaks)) {
  mmrd_high <- load_gistic_lesions_at(run_dirs[["mmrd_high"]])
  if (is.null(mmrd_high)) {
    cat("MMRd-high GISTIC run (`", run_dirs[["mmrd_high"]], "`) absent — cannot list its peaks.\n", sep = "")
  } else {
    # Soft check: the GISTIC run's sample composition may predate a later threshold
    # change on the master, so a mismatch is a warning, not an error.
    gb_supp <- group_of_barcode(rownames(mmrd_high$mat))
    if (any(!is.na(gb_supp) & gb_supp != "MMRd-high")) {
      warning("MMRd-high GISTIC run contains ", sum(!is.na(gb_supp) & gb_supp != "MMRd-high"),
              " sample(s) no longer classified MMRd-high under the current master — ",
              "the GISTIC run may predate a threshold change")
    }

    supp <- scna_freq_table(mmrd_high$mat,
                            factor(rep("MMRd-high", nrow(mmrd_high$mat)),
                                   levels = attend_scna$group_levels))
    supp <- supp[supp$scna_group == "MMRd-high", ]

    loo_dir <- file.path(gx_dir, "loo")
    loo_folders <- if (dir.exists(loo_dir))
      list.dirs(loo_dir, full.names = TRUE, recursive = FALSE) else character(0)
    loo <- if (length(loo_folders)) loo_stability(mmrd_high$peaks, loo_folders) else NULL

    tbl <- merge(supp[, c("peak_id", "n_altered", "n", "freq", "ci_lo", "ci_hi")],
                 mmrd_high$peaks[, c("peak_id", "descriptor", "chrom", "direction")],
                 by = "peak_id")
    if (!is.null(loo)) {
      tbl <- merge(tbl, loo, by = "peak_id", all.x = TRUE)
    } else {
      tbl$n_loo <- NA_integer_; tbl$n_retained <- NA_integer_; tbl$retained_frac <- NA_real_
    }
    tbl <- tbl[order(-tbl$freq), ]
    print(knitr::kable(tbl, digits = 2,
          caption = "Every MMRd-high peak: support count with Wilson 95% CI, and leave-one-out retention fraction (descriptive, not a calibrated bound)"))
    if (is.null(loo)) cat("\nnote: no leave-one-out GISTIC output under `", loo_dir,
                          "` — retention columns are NA; the support counts stand alone.\n\n", sep = "")
  }
}
```


```{r session-info}
sessionInfo()
```
