---
title: "ATTEND — 36. Is the TMB distribution split?"
author: "bolt3x"
date: "2026-08-11"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
library(here)

# renv: tidyverse, here. Optional (requireNamespace-guarded, like the rest of the
# pipeline): diptest (Hartigan's dip test) + mclust (Gaussian mixture). If missing:
#   renv::install(c("diptest","mclust")); renv::snapshot()
# The report still knits without them — those panels degrade to a scaffolding note.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 10, fig.height = 7)

source(here("code", "attend_classes.R"))     # tmb_defs_present(), dip_bimodality(), is_mmrd()
source(here("code", "attend_bimodality.R"))  # .bimodality_coef(), .gmm_modes(), have_* flags
source(here("code", "attend_plots.R"))       # attend_theme()
source(here("code", "attend_io.R"))          # read_intermediate(), write_intermediate()

master <- read_intermediate("attend_master_joined")
```

> **Question.** Does TMB split the cohort into two populations, or is it one continuous distribution with a long tail?
> **Cohort.** All `r nrow(master)` patients on the master; no restriction.
> **Reads.** `attend_master_joined`, the `tmb__*` columns only.
> **Out of scope.** Whether the split reproduces externally -> [41_tcga_tmb_replication](41_tcga_tmb_replication.html). TMB against aneuploidy -> [37_tmb_by_aneuploidy_and_mmr](37_tmb_by_aneuploidy_and_mmr.html).

A genuine second mode at high TMB *is* the hypermutator population: MMR-deficient tumours
whose failed mismatch repair produces an order of magnitude more mutations. POLE-
ultramutated tumours are the other hypermutator route, and ATTEND has none, so the question
is really whether TMB by itself cleanly identifies the hypermutators.

Four steps: describe each distribution, test for bimodality under three independent
criteria, validate any second mode against MMR status, and — in
[41_tcga_tmb_replication](41_tcga_tmb_replication.html) — reproduce the whole thing in
TCGA UCEC, where the answer is externally known.

```{r child='_explainers/master-provenance.Rmd'}
```

```{r child='_explainers/tmb-two-definitions.Rmd'}
```

Comparing the two definitions is a second question this report answers. Tumour-only TMB
carries ancestry-dependent germline inflation, which is noise added to the signal, so
removing it should deepen the valley between the modes. If the corrected definition is
more clearly bimodal than the uncorrected one, the correction is doing something real; if
not, it is not helping the split, and the battery below says so either way.

Every distribution and every test is on the raw mut/Mb scale, with no log transform. TMB is
heavily right-skewed, and on that scale a long tail and a genuine second mode can look
alike, which is why three independent criteria are triangulated rather than trusting any
one of them.

## Data

The `tmb__*` columns of the master, reshaped to one row per patient and definition.
Definitions with fewer than 10 finite values are dropped: the ancestry-corrected column is
all-`NA` until `attend_ancestry$clinical_col` is set, and a distribution test needs values.

```{r data}
master <- read_intermediate("attend_master_joined")

# The two TMB definitions the pipeline plots: normal first, then ancestry corrected.
# Same helper every other report uses, so this one cannot drift from them.
def_cols <- tmb_defs_present(master)

# Drop definitions that are absent or effectively empty (the ancestry-corrected column
# stays all-NA until attend_ancestry$clinical_col is set) — fewer than 10 finite values
# cannot support a distribution test.
finite_n <- vapply(def_cols,
                   function(cc) sum(is.finite(suppressWarnings(as.numeric(master[[cc]])))),
                   integer(1))
def_cols <- def_cols[finite_n >= 10]

# Reader-facing labels from the one shared map, so "nassar" never reaches the page.
def_labels <- tmb_def_labels(def_cols)

# One long row per (patient, definition), raw mut/Mb, NA dropped. The factor keeps
# correction order so every facet and axis reads normal -> ancestry corrected.
long <- master |>
  select(pid, all_of(unname(def_cols))) |>
  pivot_longer(all_of(unname(def_cols)), names_to = "col", values_to = "tmb") |>
  mutate(
    tmb        = suppressWarnings(as.numeric(tmb)),
    definition = factor(names(def_cols)[match(col, def_cols)],
                        levels = names(def_cols), labels = unname(def_labels))
  ) |>
  filter(is.finite(tmb)) |>
  select(pid, definition, tmb)

cat("TMB definitions analysed:",
    if (length(def_cols)) paste(def_labels, collapse = ", ") else "none present", "\n")
cat("finite values per definition (full master):\n")
print(finite_n)
```

Both the long table and the battery are written to intermediates, because
[41_tcga_tmb_replication](41_tcga_tmb_replication.html) puts the TCGA reference alongside
these exact rows and must not rebuild them by a second route.

```{r save-tmb-long}
invisible(write_intermediate(long, "attend_tmb_long"))
```

## 1. Distribution overview

Per-definition summary of the raw TMB. `%high` is the fraction at or above the
pipeline TMB-high cut (`attend_thresholds$tmb` = `r attend_thresholds$tmb` mut/Mb),
via the shared `tmb_class_of()` — the same high/low split every TMB-keyed figure uses.

```{r summary-table}
if (nrow(long) > 0) {
  summ <- long |>
    group_by(definition) |>
    summarise(
      n       = dplyr::n(),
      median  = median(tmb),
      IQR     = IQR(tmb),
      mean    = mean(tmb),
      sd      = sd(tmb),
      min     = min(tmb),
      max     = max(tmb),
      skew    = unname(.skew_kurt(tmb)["skew"]),
      exkurt  = unname(.skew_kurt(tmb)["kurt"]),
      `%high` = 100 * mean(tmb_class_of(tmb) == "TMB-high"),
      .groups = "drop"
    )
  knitr::kable(summ, digits = 2,
               caption = "Raw TMB (mut/Mb) by definition. High right-skew (large positive `skew`) is expected and is itself a hint of a heavy hypermutator tail.")
} else {
  cat("No TMB values present — nothing to summarise.\n")
}
```

Kernel density of the raw mut/Mb values in `long`, one curve per definition, with the
`r attend_thresholds$tmb` mut/Mb cutoff marked. A valley between two humps is the shape the
battery below puts numbers on.

```{r density, fig.height=6}
if (nrow(long) > 0) {
  ggplot(long, aes(tmb)) +
    geom_histogram(aes(y = after_stat(density)), bins = 40,
                   fill = "grey85", colour = "grey65", linewidth = 0.2) +
    geom_density(colour = "steelblue", linewidth = 0.8, adjust = 0.9, na.rm = TRUE) +
    geom_rug(alpha = 0.25) +
    geom_vline(xintercept = attend_thresholds$tmb, linetype = "dashed",
               colour = "firebrick") +
    facet_wrap(~ definition, scales = "free") +
    labs(title = "TMB distribution by definition (raw mut/Mb)",
         subtitle = paste0("dashed red = TMB-high cut (", attend_thresholds$tmb,
                           " mut/Mb). A distinct high-TMB hump = hypermutator mode."),
         x = "TMB (mut/Mb)", y = "density") +
    attend_theme()
} else {
  message("No TMB values to plot.")
}
```

## 2. Bimodality battery

Three independent criteria per definition, because on the raw scale no single test is
decisive:

- **Hartigan's dip test** (`dip_bimodality()`, reused from `attend_classes.R`) — the
  scale-invariant departure-from-unimodality statistic; small **p** ⇒ bimodal.
- **Sarle's bimodality coefficient (BC)** — from skewness/kurtosis; **b > 0.555** leans
  bimodal. (Inflated by a heavy tail, so corroborative only.)
- **Gaussian mixture (GMM)** — `mclust::Mclust(G = 1:3)`, BIC-selected component count;
  **≥ 2** components ⇒ the data prefer more than one Gaussian.

The **verdict** calls a definition *bimodal* when **≥ 2 of the 3** criteria agree —
a deliberately conservative majority vote.

```{r battery}
if (nrow(long) > 0) {
  battery <- long |>
    group_by(definition) |>
    group_modify(function(.x, ...) {
      x   <- .x$tmb
      dp  <- dip_bimodality(x)          # list(dip, p, n, bimodal)
      bc  <- as.numeric(.bimodality_coef(x))
      gmm <- .gmm_modes(x)
      tibble(
        n           = length(x),
        dip_D       = dp$dip,
        dip_p       = dp$p,
        dip_bimodal = isTRUE(dp$bimodal),
        BC          = bc,
        BC_bimodal  = is.finite(bc) && bc > 0.555,
        gmm_G       = gmm$G,
        gmm_bimodal = is.finite(gmm$G) && gmm$G >= 2,
        gmm_gap_sd  = gmm$gap_sd
      )
    }) |>
    ungroup() |>
    mutate(
      votes   = dip_bimodal + BC_bimodal + gmm_bimodal,
      verdict = if_else(votes >= 2, "bimodal", "unimodal")
    )

  knitr::kable(
    battery |>
      transmute(definition, n,
                dip_D = round(dip_D, 4), dip_p = signif(dip_p, 3), dip_bimodal,
                BC = round(BC, 3), BC_bimodal,
                gmm_G, gmm_bimodal,
                votes, verdict),
    caption = "Bimodality battery per TMB definition. verdict = bimodal when ≥ 2 of {dip, BC, GMM} agree."
  )
} else {
  battery <- tibble()
  cat("No TMB values — bimodality battery skipped.\n")
}
```

The same table read back as sentences, one line per definition, so the vote and its
verdict are stated rather than inferred from the columns.

```{r battery-note, results='asis'}
if (exists("battery") && nrow(battery) > 0) {
  bm <- battery |> filter(verdict == "bimodal") |> pull(definition) |> as.character()
  if (length(bm)) {
    cat("**Bimodal by majority vote:** ", paste(bm, collapse = ", "), ".\n\n", sep = "")
  } else {
    cat("**No definition reaches the ≥2-of-3 bimodality threshold** on the raw scale. ",
        "The high-TMB signal may be a heavy tail rather than a resolved second mode — ",
        "revisit on a log scale if a hypermutator subpopulation is expected.\n\n", sep = "")
  }
}
```

## 3. Mode validation — does the high-TMB mode capture MMR-deficient tumours?

A statistical second mode is only interesting if it is the **biological** one. For every
definition the mixture split into ≥ 2 components, we assign each patient to a component by
posterior, label the **highest-mean component** as the high-TMB mode, and cross-tabulate
that membership against **MMR status** (`gianlu__MMR_STATUS`, via the shared `is_mmrd()`).
A Fisher test asks whether the high-TMB mode is enriched for MMR-deficient tumours — the
expected hypermutators.

```{r mode-validation, results='asis'}
mmr_col <- attend_cols$mmr_status   # gianlu__MMR_STATUS

validate_mode <- function(def_key, def_lab) {
  col <- def_cols[[def_key]]
  d <- master |>
    transmute(pid,
              tmb  = suppressWarnings(as.numeric(.data[[col]])),
              mmrd = is_mmrd(.data[[mmr_col]])) |>
    filter(is.finite(tmb))
  if (nrow(d) < 6) return(NULL)
  mc <- tryCatch(mclust::Mclust(d$tmb, G = 1:3, verbose = FALSE), error = function(e) NULL)
  if (is.null(mc) || mc$G < 2) return(NULL)
  hi_comp  <- which.max(as.numeric(mc$parameters$mean))         # highest-mean component
  d$hi_mode <- factor(if_else(mc$classification == hi_comp, "high-TMB mode", "bulk"),
                      levels = c("bulk", "high-TMB mode"))
  d2 <- d |> filter(!is.na(mmrd))
  if (nrow(d2) < 4 || dplyr::n_distinct(d2$hi_mode) < 2 || dplyr::n_distinct(d2$mmrd) < 2)
    return(list(lab = def_lab, tab = table(d$hi_mode), p = NA_real_))
  tab <- table(MMR = if_else(d2$mmrd, "MMRd", "MMRp"), mode = d2$hi_mode)
  p   <- tryCatch(fisher.test(tab)$p.value, error = function(e) NA_real_)
  list(lab = def_lab, tab = tab, p = p)
}

if (!have_mclust) {
  cat("_`mclust` not installed — mode validation skipped._\n\n")
} else if (!mmr_col %in% names(master)) {
  cat("_MMR status column (`", mmr_col, "`) absent from the master — mode validation skipped._\n\n", sep = "")
} else if (!exists("battery") || nrow(battery) == 0) {
  cat("_No definitions to validate._\n\n")
} else {
  qualifying <- battery |> filter(is.finite(gmm_G) & gmm_G >= 2) |> pull(definition) |> as.character()
  key_by_lab <- setNames(names(def_cols), unname(def_labels))
  if (!length(qualifying)) {
    cat("_No definition produced a ≥2-component mixture — no high-TMB mode to validate._\n\n")
  } else {
    for (lab in qualifying) {
      res <- validate_mode(key_by_lab[[lab]], lab)
      if (is.null(res)) next
      cat("### ", res$lab, "\n\n", sep = "")
      print(knitr::kable(as.data.frame.matrix(res$tab),
                         caption = paste0("High-TMB mode membership vs MMR status — ", res$lab)))
      cat("\n\nFisher exact p = ",
          if (is.finite(res$p)) signif(res$p, 3) else "NA (a margin was degenerate)",
          if (is.finite(res$p) && res$p < 0.05) " — the high-TMB mode is significantly enriched for MMRd." else "",
          "\n\n", sep = "")
    }
  }
}
```

## 4. Cross-definition synthesis — does correction sharpen the split?

Removing ancestry-dependent germline inflation should make the hypermutator mode *stand
out more*: as
ancestry-inflated germline counts are stripped, the bulk tightens and the valley
deepens. Two comparators, read left-to-right along the chain:

- **Dip statistic D** (scale-invariant) — higher = further from unimodal.
- **SD-normalised mixture gap** — the distance between the extreme component means in
  units of the definition's own SD (comparable across definitions; the raw mut/Mb gap
  is not, because the definitions live on shifted scales).

```{r synthesis, fig.height=4.5}
have_syn <- exists("battery") && nrow(battery) > 0 &&
  (any(is.finite(battery$dip_D)) || any(is.finite(battery$gmm_gap_sd)))
if (have_syn) {
  syn <- battery |>
    select(definition, `dip statistic D` = dip_D, `mixture gap (÷SD)` = gmm_gap_sd) |>
    pivot_longer(-definition, names_to = "metric", values_to = "value") |>
    filter(is.finite(value))
  if (nrow(syn) > 0) {
    ggplot(syn, aes(definition, value, group = 1)) +
      geom_line(colour = "grey60") +
      geom_point(size = 3, colour = "steelblue") +
      facet_wrap(~ metric, scales = "free_y") +
      labs(title = "Separation strength along the germline-correction chain",
           subtitle = "left to right = normal -> ancestry corrected; up = more clearly bimodal",
           x = NULL, y = NULL) +
      attend_theme() +
      theme(axis.text.x = element_text(angle = 20, hjust = 1))
  } else {
    message("No finite separation metrics to plot.")
  }
} else {
  message("Battery unavailable — synthesis skipped.")
}
```

The cross-definition comparison in words: whether moving from the delivered score to the
ancestry-corrected one deepens the valley, holds it, or closes it.

```{r synthesis-note, results='asis'}
if (exists("battery") && nrow(battery) > 0 && sum(is.finite(battery$dip_D)) >= 2) {
  b <- battery |> filter(is.finite(dip_D))
  first_lab <- as.character(b$definition[1]); last_lab <- as.character(b$definition[nrow(b)])
  d_first <- b$dip_D[1]; d_last <- b$dip_D[nrow(b)]
  cat("Dip D moves from ", signif(d_first, 3), " (", first_lab, ") to ",
      signif(d_last, 3), " (", last_lab, "). ",
      if (d_last > d_first)
        "Separation **increases** after correction — consistent with the correction sharpening the hypermutator mode."
      else if (d_last < d_first)
        "Separation **decreases** after correction — it does not visibly sharpen the split on the raw scale (or the corrected column is too sparse to resolve it)."
      else "Separation is unchanged after correction.",
      "\n\n", sep = "")
}
```

The battery is written to an intermediate for
[41_tcga_tmb_replication](41_tcga_tmb_replication.html), which appends TCGA as one more
row of this same table.

```{r save-battery}
if (exists("battery") && nrow(battery) > 0)
  invisible(write_intermediate(dplyr::mutate(battery, definition = as.character(definition)),
                               "attend_tmb_battery"))
```

```{r session-info}
sessionInfo()
```
