---
title: "ATTEND — 41. TCGA UCEC replication of the TMB split"
author: "bolt3x"
date: "2026-08-11"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
library(here)

# renv: tidyverse, here. Optional: diptest, mclust (same guarded pair as report 36).
# Needs data/tcga_2013/ — Rscript code/fetch_tcga_ucec_2013.R
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 10, fig.height = 7)

source(here("code", "attend_classes.R"))     # dip_bimodality(), attend_thresholds
source(here("code", "attend_bimodality.R"))  # .bimodality_coef(), .gmm_modes(), have_mclust
source(here("code", "attend_plots.R"))       # attend_theme()
source(here("code", "attend_io.R"))          # read_intermediate()
source(here("code", "load_wes_results.R"))   # load_tcga_ucec_2013()

# The ATTEND side is not rebuilt here: report 36 wrote both tables, so the rows compared
# against TCGA are byte-for-byte the rows report 36 tested.
long    <- tryCatch(read_intermediate("attend_tmb_long"),    error = function(e) tibble())
battery <- tryCatch(read_intermediate("attend_tmb_battery"), error = function(e) tibble())
```

> **Question.** Does the TMB split ATTEND shows reproduce in a paired-sequencing cohort of the same disease?
> **Cohort.** TCGA UCEC 2013, exome-sequenced tumours, POLE-ultramutated excluded; compared against all ATTEND patients.
> **Reads.** `data/tcga_2013/data_clinical_sample.txt` via `load_tcga_ucec_2013()`, plus the `attend_tmb_long` and `attend_tmb_battery` intermediates written by [36_tmb_distribution](36_tmb_distribution.html).
> **Out of scope.** The ATTEND-only battery -> [36_tmb_distribution](36_tmb_distribution.html).

TCGA UCEC 2013 (Kandoth et al., *Nature* 2013) is the same disease sequenced tumour with
matched normal, so germline variants were subtracted at source and its TMB carries none of
the tumour-only inflation the ancestry correction is trying to undo. That makes it a
reference rather than just another cohort: if TCGA is clearly bimodal and ATTEND is not,
the missing split is a measurement problem in tumour-only TMB; and if the ancestry-corrected
definition looks more TCGA-like than the uncorrected one, the correction is moving ATTEND
toward the paired-sequencing ground truth.

TCGA UCEC is **primary** disease and ATTEND is **metastatic**, so any difference between
the cohorts is confounded with disease setting and cannot be attributed to sequencing
strategy alone. TCGA is a reference for what a paired-sequencing TMB distribution looks
like, not a matched control.

```{r child='_explainers/tmb-two-definitions.Rmd'}
```

## Loading the reference cohort

TCGA's `TMB_NONSYNONYMOUS` (non-synonymous mutations per Mb, available on the 248 of 373
tumours that were exome-sequenced) on the same raw mut/Mb scale as ATTEND. POLE-ultramutated
tumours are excluded: TCGA has 17 at roughly 232 mut/Mb, an order of magnitude above even
the MSI hypermutators, and ATTEND structurally cannot have them, so leaving them in would
make every comparison detect that rather than the MMRd split. TCGA's MMRd group is its own
MSI (Hyper-mutated) integrative cluster (`mmrd_rule = "tcga_exact"`), i.e. the paper's
published classification rather than a marker call.

```{r tcga-load}
tcga_raw <- tryCatch(load_tcga_ucec_2013(), error = function(e) tibble())
have_tcga <- nrow(tcga_raw) > 0 && "tmb" %in% names(tcga_raw) &&
             sum(is.finite(tcga_raw$tmb)) >= 10

if (!have_tcga) {
  cat("**TCGA UCEC 2013 not available** (`data/tcga_2013/`) — run",
      "`Rscript code/fetch_tcga_ucec_2013.R`. This report is skipped, not silently empty.\n")
} else {
  n_all  <- sum(is.finite(tcga_raw$tmb))
  tcga <- tcga_raw |> filter(is.finite(tmb), !pole)      # <- POLE excluded
  cat("TCGA tumours with a TMB value :", n_all, "\n")
  cat("POLE-ultramutated, excluded   :", n_all - nrow(tcga), "\n")
  cat("TCGA cohort used here         :", nrow(tcga), "\n")
  cat("  of which MMRd (MSI cluster) :", sum(tcga$mmrd, na.rm = TRUE),
      " | MMRp:", sum(!tcga$mmrd, na.rm = TRUE),
      " | unclassified:", sum(is.na(tcga$mmrd)), "\n")
}
```

Densities of the ATTEND definitions and the TCGA reference on one axis. The ATTEND
distributions are tumour-only, TCGA's is paired — so a systematic rightward shift of the
ATTEND curves relative to TCGA is the germline inflation, visible directly. The dashed
line is the pipeline's `r attend_thresholds$tmb` mut/Mb TMB-high cutoff.

The x-axis is trimmed at the 99th percentile so the shape near the modes stays readable;
the extreme tail is still in every test below, just not in the picture.

```{r tcga-density, fig.height=5}
if (have_tcga && nrow(long) > 0) {
  cmp <- bind_rows(
    long |> transmute(cohort = paste0("ATTEND — ", as.character(definition)), tmb),
    tibble(cohort = "TCGA UCEC (paired, POLE excluded)", tmb = tcga$tmb)
  ) |> filter(is.finite(tmb))

  xmax <- quantile(cmp$tmb, 0.99, na.rm = TRUE)

  ggplot(cmp, aes(tmb, colour = cohort)) +
    geom_density(linewidth = 0.9, na.rm = TRUE) +
    geom_vline(xintercept = attend_thresholds$tmb, linetype = 2, colour = "grey50") +
    coord_cartesian(xlim = c(0, xmax)) +
    labs(title = "TMB distribution: ATTEND (tumour-only) vs TCGA UCEC (tumour-normal paired)",
         subtitle = paste0("raw mut/Mb, x-axis trimmed at the 99th percentile; dashed line = ",
                           attend_thresholds$tmb, " mut/Mb TMB-high cutoff"),
         x = "TMB (mut/Mb)", y = "density", colour = NULL) +
    attend_theme() +
    theme(legend.position = "top") +
    guides(colour = guide_legend(ncol = 1))
}
```

The same three-criterion battery [36_tmb_distribution](36_tmb_distribution.html) runs, now
with TCGA as an extra row, read back from the `attend_tmb_battery` intermediate. Read the
TCGA row first: it is the benchmark. If TCGA is bimodal and an ATTEND definition is not,
the split exists in this disease and that definition is failing to resolve it.

```{r tcga-battery}
if (have_tcga) {
  tcga_row <- local({
    x   <- tcga$tmb
    dp  <- dip_bimodality(x)
    bc  <- as.numeric(.bimodality_coef(x))
    gmm <- .gmm_modes(x)
    tibble(definition  = "TCGA UCEC (paired, POLE excl.)",
           n           = length(x),
           dip_D       = dp$dip,
           dip_p       = dp$p,
           dip_bimodal = isTRUE(dp$bimodal),
           BC          = bc,
           BC_bimodal  = is.finite(bc) && bc > 0.555,
           gmm_G       = gmm$G,
           gmm_bimodal = is.finite(gmm$G) && gmm$G >= 2,
           gmm_gap_sd  = gmm$gap_sd) |>
      mutate(votes   = dip_bimodal + BC_bimodal + gmm_bimodal,
             verdict = if_else(votes >= 2, "bimodal", "unimodal"))
  })

  combined <- if (exists("battery") && nrow(battery) > 0)
    bind_rows(battery |> mutate(definition = as.character(definition)), tcga_row) else tcga_row

  knitr::kable(
    combined |>
      transmute(definition, n,
                dip_D = round(dip_D, 4), dip_p = signif(dip_p, 3), dip_bimodal,
                BC = round(BC, 3), BC_bimodal,
                gmm_G, gmm_bimodal,
                gmm_gap_sd = round(gmm_gap_sd, 2),
                votes, verdict),
    caption = "Bimodality battery — ATTEND definitions plus the TCGA paired-sequencing reference. verdict = bimodal when >= 2 of {dip, BC, GMM} agree."
  )
}
```

Does TCGA's high-TMB mode contain TCGA's MMR-deficient tumours? The same mode validation
[36_tmb_distribution](36_tmb_distribution.html) runs on ATTEND, applied to the reference
cohort. A strong enrichment
here establishes that the high mode really is the hypermutator population *in this
disease* — which is what licenses reading ATTEND's high mode the same way.

```{r tcga-mode-validation, results='asis'}
if (have_tcga && have_mclust) {
  mc <- tryCatch(mclust::Mclust(tcga$tmb, G = 1:3, verbose = FALSE), error = function(e) NULL)
  if (is.null(mc) || mc$G < 2) {
    cat("_TCGA TMB resolves to a single mixture component — no high mode to validate._\n\n")
  } else {
    hi_comp <- which.max(as.numeric(mc$parameters$mean))
    d <- tcga |>
      mutate(hi_mode = factor(if_else(mc$classification == hi_comp, "high-TMB mode", "bulk"),
                              levels = c("bulk", "high-TMB mode"))) |>
      filter(!is.na(mmrd))
    if (nrow(d) >= 4 && dplyr::n_distinct(d$hi_mode) == 2 && dplyr::n_distinct(d$mmrd) == 2) {
      tab <- table(MMR = if_else(d$mmrd, "MMRd", "MMRp"), mode = d$hi_mode)
      pv  <- tryCatch(fisher.test(tab)$p.value, error = function(e) NA_real_)
      cat("**TCGA — high-TMB mode vs MMR status** (", mc$G, " mixture components):\n\n", sep = "")
      print(knitr::kable(tab))
      cat("\nFisher p = ", signif(pv, 3), "\n\n", sep = "")
    } else {
      cat("_Not enough classified TCGA tumours across both mode and MMR groups._\n\n")
    }
  }
} else if (have_tcga) {
  cat("_`mclust` not installed — TCGA mode validation skipped._\n\n")
}
```

How each cohort's TMB separates MMR-deficient from MMR-proficient tumours, side by side.
This is the comparison the whole section is for: the **effect size** of MMR status on TMB
in a paired cohort versus in ATTEND under each definition. If the ancestry correction is
working, ATTEND's corrected separation should look more like TCGA's than the uncorrected
one does.

Wilcoxon rather than t-test, because TMB is heavily skewed on the raw scale.

```{r tcga-mmr-contrast, fig.height=5}
if (have_tcga) {
  mmr_col_l <- attend_cols$mmr_status
  att <- if (mmr_col_l %in% names(master)) {
    master |>
      select(pid, all_of(unname(def_cols)), all_of(mmr_col_l)) |>
      pivot_longer(all_of(unname(def_cols)), names_to = "col", values_to = "tmb") |>
      mutate(cohort = paste0("ATTEND — ", def_labels[match(col, def_cols)]),
             tmb    = suppressWarnings(as.numeric(tmb)),
             mmrd   = is_mmrd(.data[[mmr_col_l]])) |>
      filter(is.finite(tmb), !is.na(mmrd)) |>
      select(cohort, tmb, mmrd)
  } else tibble()

  both <- bind_rows(att,
    tcga |> filter(!is.na(mmrd)) |>
      transmute(cohort = "TCGA UCEC (paired)", tmb, mmrd))

  if (nrow(both) > 0) {
    both <- both |> mutate(MMR = if_else(mmrd, "MMRd", "MMRp"))

    print(knitr::kable(
      both |> group_by(cohort, MMR) |>
        summarise(n = n(), median = round(median(tmb), 2),
                  IQR = paste0(round(quantile(tmb, .25), 2), "-", round(quantile(tmb, .75), 2)),
                  .groups = "drop"),
      caption = "TMB by MMR status in each cohort / definition"))

    pv <- both |> group_by(cohort) |>
      summarise(p = tryCatch(signif(wilcox.test(tmb ~ MMR)$p.value, 3),
                             error = function(e) NA_real_), .groups = "drop")
    print(knitr::kable(pv, caption = "MMRd vs MMRp Wilcoxon p, per cohort / definition"))

    ggplot(both, aes(MMR, tmb, fill = MMR)) +
      geom_boxplot(outlier.shape = NA, alpha = 0.6) +
      geom_jitter(width = 0.15, size = 0.5, alpha = 0.35) +
      geom_hline(yintercept = attend_thresholds$tmb, linetype = 2, colour = "grey50") +
      scale_y_log10() +
      facet_wrap(~ cohort, nrow = 1) +
      scale_fill_manual(values = attend_mmr_cols) +
      labs(title = "TMB by MMR status — ATTEND definitions vs the TCGA paired reference",
           subtitle = "log10 y-axis for display only; all tests above are on the raw scale",
           x = NULL, y = "TMB (mut/Mb, log10)") +
      attend_theme() + theme(legend.position = "none")
  }
}
```

```{r session-info}
sessionInfo()
```
